### 2020-08-04 doc/manuscript/Dow_manuscript.Rmd
upload manuscript template, organizational files

---
title: ''
output: word_document
bibliography: references.bib
csl: global-change-biology.csl #(this will change depending on journal we choose. GCB would be a great choice)
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\raggedright

**Title:**

**Authors (not yet complete, just alphabetical for now--final order TBD):**

Cameron Dow^1^
Erika B. Gonzalez-Akre^1^,
Ryan Helcoski^1^,
Valentine Herrmann^1^,
Ian McGregor^1^,
Sean McMahon,
Albert Kim^1^,
Kristina J. Anderson-Teixeira^1,n^*

**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA

n. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama

*corresponding author: teixeirak@si.edu; +1 540 635 6546

```{r eval=FALSE, echo=FALSE}
## This is a chunk that won't be evaluated or displayed. I use these for notes.
## Target journal: ??? Global Change Biology????
## Instructions for authors: 
#[insert guidance on word count, n figures/ tables allowed, and then some rough planning of what goes in this manuscript]

```

\newpage

### Abstract

**Keywords**:

\newpage

### Introduction

*Here, I'll insert a reference to Sean's paper [@mcmahon_general_2015]. This is pulled from references.bib.*

### Materials and Methods

### Results

### Discussion

### Acknowledgements
ForestGEO

### Authors' contributions

### References

### 2020-08-04 doc/manuscript/Dow_manuscript.Rmd
transferred content from previous Word doc

@@
 
-**Title:**
+**Title:** Effect of spring temperatures on tree growth phenology in a temperate deciduous forest
 
@@
 
-*Here, I'll insert a reference to Sean's paper [@mcmahon_general_2015]. This is pulled from references.bib.*
+**With climate warming, we can expect earlier springs in temperate forests.** 
+These are likely to affect tree growth, woody productivity, and ecosystem C cycling.
+However, we know little about the growth phenology of temperate deciduous species (see D’Orangeville et al., in revision for more info/ citations)—let alone how these are affected by an early spring.
+
+Here, we ...
+- use 10 years of biweekly dendro band measurements for # species to characterize seasonal growth phenology
+
+- combine these records with 110 years of tree-ring data for 12 species to examine how spring temperatures impact annual tree growth
+
+- examine the effects of an extremely early spring (2020) seasonal growth phenology and annual growth
 
 ### Materials and Methods
+
+#### Study site and data
+SCBI ForestGEO plot
+
+10 years biweekly dendrometer measurements
+
+leaf phenology data?
+
+- leaf phenology data from NEON and/or satellite-based (leaf phenology network: https://www.usanpn.org/news/spring) ?
+
+- NDVI or PRI? https://onlinelibrary.wiley.com/doi/10.1111/gcb.15112 (from Ian)
+
+- From Ian:  I noticed how the growth patterns observed by remote sensing pretty much mirror what Sean was finding with his dendro R package (btw is that functional on CRAN yet?) for both SERC and SCBI dendroband data.
+
+perhaps bring in cores?
+
+#### Analysis
+
+*Here, I'll insert a reference to Sean's paper [@mcmahon_general_2015]. This is pulled from references.bib.*
 
@@
 ### Discussion
+
+*Content to incorporate:
+N.
+American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts.
+Thus, climate change is having less impact [@zohner_late-spring_2020]*
 

### 2020-08-08 doc/manuscript/Dow_manuscript.Rmd
updated manuscript
began drafting methods

@@
 
-10 years biweekly dendrometer measurements
+-SCBI met-tower
 
+Climate data was obtained from a meteorological tower in a clearing adjacent to, and within elevation range of the SCBI ForestGEO plot.
+This tower is part of the ForestGEO meteorological monitoring program.
+Temperature readings are taken every 5 minutes using a CR1000 datalogger.
+
+9 years biweekly dendrometer measurements
+
+  Dendrometer band measurements were taken every two weeks for 9 years (After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees ).
+Data manually cleaned by visual inspection.
+Three classes of mistakes: 1.
+Error in measurement - weekly measurement was drastically different from previous week and following week.
+Mistakes were removed from the raw data before modeling.
+(## cases) 2.
+Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern.
+Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year.
+In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level.
+(## cases) 3.
+Other - cases where data was clearly wrong but with unknown causes.
+This seemed to happen mostly in 2011, when the program was just beginning.
+In cases where several measurements were an issue, they were simply removed.
+If there were no clear solutions to 'fix' the data, the entire year was removed from the analysis.
 leaf phenology data?
@@
 
+- Sean's model
+
+- Climwin
+
+The period where the phenology milestone(s) were most affected by Spring time temperature was determined using the R package Climwin.
+This package tests the correlation between climate variables (daily averaged temperature) and biological variables (DOY) within a specified time-frame, reporting the window with the highest correlation as the 'best-model'.
+We instructed Climwin to search for the best window within ### to 0 days before the phenology milestone.
+This was done to find corresponding best-windows for both of our wood-types; ring-porous and diffuse-porous.
+
 #### Analysis
@@
 *Here, I'll insert a reference to Sean's paper [@mcmahon_general_2015]. This is pulled from references.bib.*
+
+- Mixed effect model
+
+A mixed effect model was used to test the response of wood phenology variables (25% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, and random effects of species and tag.
+We ran two separate models for each of our major wood-types *and one combined model to use for comparison during the bayesian analysis.* (talking to Albert about possibly doing this)
+
+- Bayesian heirarchal model
 

### 2020-08-24 RScripts/wood_phenology_analysis.R
Added joint models

@@ -232,9 +232,17 @@ seventyfiveDP <- subset(seventyfive, wood_type == "diffuse-porous")
 seventyfiveDP <- merge(seventyfiveDP, dpmeans, by = "year")
 
 #Check out relationship with plot
-dpdoys <- aggregate(twentyfiveDP$DOY, by = list(twentyfiveDP$year), FUN = mean)
-plot(dpdoys$x~dpmeans$dptemp)
-summary(lm(dpdoys$x~dpmeans$dptemp))
+dpdoys25 <- aggregate(twentyfiveDP$DOY, by = list(twentyfiveDP$year), FUN = mean)
+plot(dpdoys25$x~dpmeans$dptemp, xlab = "Average max temp", ylab = "DOY 25% growth acheived", main = "Diffuse-porous preseason relationship")
+summary(lm(dpdoys25$x~dpmeans$dptemp))
+
+dpdoys50 <- aggregate(fiftyDP$DOY, by = list(fiftyDP$year), FUN = mean)
+plot(dpdoys50$x~dpmeans$dptemp, xlab = "Average max temp", ylab = "DOY 50% growth acheived", main = "Diffuse-porous preseason relationship")
+summary(lm(dpdoys50$x~dpmeans$dptemp))
+
+dpdoys75 <- aggregate(seventyfiveDP$DOY, by = list(seventyfiveDP$year), FUN = mean)
+plot(dpdoys75$x~dpmeans$dptemp, xlab = "Average max temp", ylab = "DOY 75% growth acheived", main = "Diffuse-porous preseason relationship")
+summary(lm(dpdoys75$x~dpmeans$dptemp))
 
 rpdoys <- aggregate(twentyfiveRP$DOY, by = list(twentyfiveRP$year), FUN = mean)
 plot(rpdoys$x~rpmeans$rptemp)
@@ -412,6 +420,17 @@ maxrateDOY_formulaRP <- "max_rate_DOY ~ rptemp + (1|sp) + (1|sp:tag)" %>% as.for
 
 maxrate_formulaRP <- "max_rate ~ rptemp + (1|sp) + (1|sp:tag)" %>% as.formula()
 
+#Multivariate model
+jointmodelRP <- stan_mvmer(
+  formula = list(
+    DOY ~ rptemp + (1 | sp) + (1|tag),
+    DOY ~ rptemp + (1 | sp) + (1|tag),
[…]

### 2020-08-24 doc/manuscript/Dow_manuscript.Rmd
add authors
Update Dow_manuscript.Rmd
Jotting down general thoughts / outline of ideas to talk about

--- 2020-08-04 doc/manuscript/Dow_manuscript.Rmd
@@
 Cameron Dow^1^
+Loïc D’Orangeville
 Erika B. Gonzalez-Akre^1^,
@@
 Albert Kim^1^,
+David A. Orwig
 Kristina J. Anderson-Teixeira^1,n^*
@@
 
-*Here, I'll insert a reference to Sean's paper [@mcmahon_general_2015]. This is pulled from references.bib.*
+As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented to combat this, the planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
+This will have potentially negative implications for nearly every ecosystem on Earth, including forests.
+-importantce of forests Deciduous forests in the Northern Hemisphere have the potential to remove up to ## tons of carbon in the next ## years, and currently represent %% of the worlds carbon reserves.
+Trees remove CO2 from the atmosphere, and fix it into structural components like wood, roots, or leaves, through the process of photosynthesis.
+Current theory suggests that tree growth could benefit from global warming, as warmer early Spring time allows leaves to emerge earlier and increased CO2 concentrations increase plant water use efficiency.
+This increase in growth can slow the rate of global warming as CO2 sequestration is sped up, making forests key natural combatants in the fight against greenhouse gasses.
+
+In recent decades, Spring has been arriving earlier as Spring temperatures are increasing.
+This leads to earlier leaf out in deciduous trees, meaning a potentially longer growing season ().
+Earlier Spring emergence has been shown to cause both positive and negative lagged effects on subsequent Autumn or Spring growth in European broad leaf deciduous and mixed forests, respectively (Crabbe et al, 2016).
+Even more, the timing of Spring warming seems to have a significant effect on leaf-out response, where warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014).
+These shifts in phenological timing are likely to affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests.
+However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood.
+Many predictions of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale observations of phenological timing, and do little to discern differences between individual species.
+In reality, we know little about the growth phenology of temperate deciduous species (see D’Orangeville et al., in revision for more info/ citations)—let alone how these are affected by an early spring.
+
+To date, very little research has been done on the phenology of tree growth.
+While observations of leaf phenology have shown a decent relationship with NPP, understanding intraannual wood-growth phenology allows a finer-scale look at differences in climate sensitivity and growth responses between individual species.
+Recent results from D’Orangeville et al.
+(in prep) have shown that wood-type plays a significant role in the timing of leaf and wood phnology, as ring-porous begin growth earlier in the season compared to diffuse-porous, but both reached 75% total annual growth around the same time.
+This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stressors.
+
+Temperate trees are projected to increase NPP and sequester greater amounts of CO2 as climate warms, to a point, but the contribution of individual species, and the implications of potentially chanign species compositions is less well known.
+This simplification is necessary at larger scales, as generalizations are needed to ensure models are not overly complex.
+However, the results here indicate that all tree species may not be responding equally to an earlier Spring.
+We suggest that efforts be made to examine regions on a smaller scale, investigating indiviudual species or wood-types’ reaction to historic warming to forecast future NPP.
+For example, at SCBI, the loss of ring-porous Fraxinus spp., the high prolification of the diffuse-porous Lirodendron Tulipera, and Decline of mature individuals along with browsing pressure on Quercus saplings could lead to a population boom of diffuse-porous species, making our forest less likely to increase growth in response to a warming early-Spring and thus, lessening the impact our forest has on the global carbon balance.
+
+Dendrometer bands allow a look at indivudal trees’ intraannual growth.
+Using many measurements taken within a year allows the creation of an idealized growth curve using a 5 paramter logistic model (Sean’s model).
+From this, growth milestones and variables can be extracted from each year and compared to other years, effectivley simulating a controlled experiment where climate variables were manipulated.
+
+To our knowledge, no research has been done exploring the relationship between intraannual growth patterns and early Spring temperatures.
+Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual growth phenology within the FOrestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
+These variables were then compared to pre-season temperatures over the 9 year period of 2011-2019 to search for correlation between early-spring temps and growth phenology.
+
+We hypothesized that (1) ring-porous and diffuse-porous trees would both reach 25, and 50% total annual growth earlier and (2) both wood types would increase total growth in response to warmer early-Spring temperatures.
+
+Here, we … - use 10 years of biweekly dendro band measurements for # species to characterize seasonal growth phenology
+•	combine these records with 110 years of tree-ring data for 12 species to examine how spring temperatures impact annual tree growth
+•	examine the effects of an extremely early spring (2020) seasonal growth phenology and annual growth
 
@@
 
+Study site and data
+SCBI ForestGEO plot
+-SCBI met-tower
+Climate data was obtained from a meteorological tower in a clearing adjacent to, and within elevation range of the SCBI ForestGEO plot.
+This tower is part of the ForestGEO meteorological monitoring program.
+Temperature readings are taken every 5 minutes using a CR1000 datalogger.
+9 years biweekly dendrometer measurements
+Dendrometer band measurements were taken every two weeks for 9 years (After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees ).
+Data manually cleaned by visual inspection.
+Three classes of mistakes: 1.
+Error in measurement - weekly measurement was drastically different from previous week and following week.
+Mistakes were removed from the raw data before modeling.
+(## cases) 2.
+Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern.
+Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year.
+In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level.
+(## cases) 3.
+Other - cases where data was clearly wrong but with unknown causes.
+This seemed to happen mostly in 2011, when the program was just beginning.
+In cases where several measurements were an issue, they were simply removed.
+If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.
+
+leaf phenology data?
+
+•	leaf phenology data from NEON and/or satellite-based (leaf phenology network: https://www.usanpn.org/news/spring) ?
+
+•	NDVI or PRI? https://onlinelibrary.wiley.com/doi/10.1111/gcb.15112 (from Ian)
+
+•	From Ian: I noticed how the growth patterns observed by remote sensing pretty much mirror what Sean was finding with his dendro R package (btw is that functional on CRAN yet?) for both SERC and SCBI dendroband data.
+
+perhaps bring in cores?
+
+•	Sean’s model
+
+•	Climwin
+
+The period where the phenology milestone(s) were most affected by Spring time temperature was determined using the R package Climwin.
+This package tests the correlation between climate variables (daily averaged temperature) and biological variables (DOY) within a specified time-frame, reporting the window with the highest correlation as the ‘best-model’.
+We instructed Climwin to search for the best window within ### to 0 days before the phenology milestone.
+This was done to find corresponding best-windows for both of our wood-types; ring-porous and diffuse-porous.
+
+### Analysis
+
+Here, I’ll insert a reference to Sean’s paper (McMahon & Parker, 2015). This is pulled from references.bib.
+
+-Mixed effect model
+
+A mixed effect model was used to test the response of wood phenology variables (25% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, and random effects of species and tag.
+We ran two separate models for each of our major wood-types and one combined model to use for comparison during the bayesian analysis.
+(talking to Albert about possibly doing this)
+
+•	Bayesian heirarchal model
+
 ### Results
 
+•	variable averages
+
+Pooled across all species and all years the average DOY where max growth rate was achieved is DOY: 158, the DOY where 25% was achieved is DOY: 129, 50% - DOY: 157, 75% - DOY:184.
+
+Ring-porous (1395 obs) max growth rate – DOY: 147, max growth rate = .048 (find units in Sean’s function), 25% - DOY: 113, 50% DOY: 147, 75% - DOY: 181…68 days
+
+Diffuse-porous (993 obs) max growth rate – DOY: 172, max growth rate = .068 (find units in Sean’s function), 25% - DOY: 153, 50% - DOY: 171, 75% - DOY: 190… 37 days
+
+•	Climwin results
+
+•	Accounting for random effects of species and tag, pre-season temperatures had a greater effect on ring-porous species than diffuse-porous.
+
+•	Bayesian variable results (PRE-JOINT MODEL, updating ASAP)
+
+Ring-porous 25% DOY estimate: -2.23 (credible interval: -2.94 - -1.55)
+
+Ring-porous 50% DOY estimate: - 2.10 (credible interval: -2.71 - -1.42)
+
+Ring-porous 75% DOY estimate: -1.89 (credible interval: -2.53 - -1.28)
+
+Ring-porous total growth estimate: 0.0552 (credible interval: 0.00346 - 0.108)
+
+Ring-porous max-rate DOY estimate: -2.20 (credible interval: -2.77 - -1.62)
+
+Ring-porous max-rate estimate: 0.00113 (credible interval: 0.000151 – 0.00208)
+
+Diffuse-porous 25% DOY estimate: -1.83 (credible interval: -2.86 - -0.822)
+
+Diffuse-porous 50% DOY estimate: -2.39 (credible interval: -3.35 - -1.43)
+
+Diffuse-porous 75% DOY estimate: -2.96 (credible interval: -4.03 - -1.92)
+
+Diffuse-porous total growth estimate: -0.0642 (credible interval: -0.172 – 0.0460)
+
+Diffuse-porous max-rate DOY estimate: -2.54 (credible interval: -3.41 - -1.69)
+
+Diffuse-porous max-rate estimate: 0.00209 (credible interval: 0.000118 – 0.00413)
+
 ### Discussion
+
+Diffuse-porous 25-75% growth period shrinking leaving trees more vulnerable to drought while ring-porous 25-75% growing, potentially making effect of drought less damaging?
+But RP still prone to cavitation + longer growth period could mean exposure to higher number of droughts resulting in serious damage to water transport system?
+
+DIfferences in total relationship could be because of response difference in wood types OR could be because the diffuse-porous’ growth occuring during the later summer means it is exposed to drought/heat and on average, not feeling the effects of the earlier spring.
+differences in wood porosity growth timing means exposure to different periods of climate stress ring porous adaptable to spring temps, diffuse not so much leads to higher growth in ring however, diffuse still more resistant to cavitation potential for more frequent late frost events having large impacts as spring warms?
+offset of chilling requirements and spring leaf out requirements
+
+much assumed about future forest productivity based on gis different wood types/species may respond differently need to understand at a finer scale what is happening
+
+Content to incorporate: N.
+American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts.
+Thus, climate change is having less impact (Zohner et al., 2020)
 

### 2020-08-30 doc/manuscript/Dow_manuscript.Rmd
more work on introduction

@@
 
-As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented to combat this, the planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
-This will have potentially negative implications for nearly every ecosystem on Earth, including forests.
--importantce of forests Deciduous forests in the Northern Hemisphere have the potential to remove up to ## tons of carbon in the next ## years, and currently represent %% of the worlds carbon reserves.
-Trees remove CO2 from the atmosphere, and fix it into structural components like wood, roots, or leaves, through the process of photosynthesis.
-Current theory suggests that tree growth could benefit from global warming, as warmer early Spring time allows leaves to emerge earlier and increased CO2 concentrations increase plant water use efficiency.
-This increase in growth can slow the rate of global warming as CO2 sequestration is sped up, making forests key natural combatants in the fight against greenhouse gasses.
+As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented to combat this, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
+The absence of human intervention increases our reliance on natural systems to soften the blow of climate change.
 
-In recent decades, Spring has been arriving earlier as Spring temperatures are increasing.
-This leads to earlier leaf out in deciduous trees, meaning a potentially longer growing season ().
-Earlier Spring emergence has been shown to cause both positive and negative lagged effects on subsequent Autumn or Spring growth in European broad leaf deciduous and mixed forests, respectively (Crabbe et al, 2016).
-Even more, the timing of Spring warming seems to have a significant effect on leaf-out response, where warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014).
-These shifts in phenological timing are likely to affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests.
+Forests are an extremely valuable ecosystem for many reasons.
+They house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil.
+Forests sequester about one-third of all anthropogenic CO2 and store ~45% of all terrestrial carbon (Bonan, 2008).
+However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).
+Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019).
+Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO2, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens.
+Understanding how growth is influenced by multiple factors at once is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation.
+CO2 concentrations in the atmosphere are increasing, and in a perfect scenario, trees and other plants would simply speed up growth in response to this, increasing photosynthetic efficiency in a process called CO2 fertilization (Strain and Cure, 1986; Huang et al., 2007).
+However, factors other than CO2 also limit tree growth, and seasonal temperature averages and precipitation patterns are changing concurrently.
+It is anticipated that, throughout much of the United States, future heat and drought may limit tree growth more strongly than it can be enhanced by elevated CO2 (Charney et al, 2016).
+
+In recent decades, an increase in average early Spring temperatures has been attributed to climate change.
+As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, meaning a potentially longer growing season (Ibáñez et al., 2010; Jeong et al., 2011).
+The effects of this extension of the growing season on leaf phenology and NPP have been well studied (Way, 2011; ), but little work has been done to understand the impacts to intraannual stem-growth variables.
+Shifts in leaf phenology timing could be coupled with a change in stem-growth phenology, which are likely to affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests.
+Thus, it is important to understand how the warming of Spring changes stem-growth phenology.
+
+The timing of Spring warming has a significant effect on leaf phenology response.
+Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
+GIS source Ian sent.
 However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood.
-Many predictions of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale observations of phenological timing, and do little to discern differences between individual species.
+Many predictions of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale observations of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
 In reality, we know little about the growth phenology of temperate deciduous species (see D’Orangeville et al., in revision for more info/ citations)—let alone how these are affected by an early spring.
 
-To date, very little research has been done on the phenology of tree growth.
-While observations of leaf phenology have shown a decent relationship with NPP, understanding intraannual wood-growth phenology allows a finer-scale look at differences in climate sensitivity and growth responses between individual species.
+To date, very little research has been done on the phenology of stem-growth.
+Timing of intraannual growth is extremely improtant as it can dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ).
+For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered creating a negative climate feedback loop.
 Recent results from D’Orangeville et al.
-(in prep) have shown that wood-type plays a significant role in the timing of leaf and wood phnology, as ring-porous begin growth earlier in the season compared to diffuse-porous, but both reached 75% total annual growth around the same time.
-This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stressors.
+(in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology, as ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time.
+This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses.
 
-Temperate trees are projected to increase NPP and sequester greater amounts of CO2 as climate warms, to a point, but the contribution of individual species, and the implications of potentially chanign species compositions is less well known.
-This simplification is necessary at larger scales, as generalizations are needed to ensure models are not overly complex.
-However, the results here indicate that all tree species may not be responding equally to an earlier Spring.
-We suggest that efforts be made to examine regions on a smaller scale, investigating indiviudual species or wood-types’ reaction to historic warming to forecast future NPP.
-For example, at SCBI, the loss of ring-porous Fraxinus spp., the high prolification of the diffuse-porous Lirodendron Tulipera, and Decline of mature individuals along with browsing pressure on Quercus saplings could lead to a population boom of diffuse-porous species, making our forest less likely to increase growth in response to a warming early-Spring and thus, lessening the impact our forest has on the global carbon balance.
-
-Dendrometer bands allow a look at indivudal trees’ intraannual growth.
-Using many measurements taken within a year allows the creation of an idealized growth curve using a 5 paramter logistic model (Sean’s model).
-From this, growth milestones and variables can be extracted from each year and compared to other years, effectivley simulating a controlled experiment where climate variables were manipulated.
-
-To our knowledge, no research has been done exploring the relationship between intraannual growth patterns and early Spring temperatures.
-Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual growth phenology within the FOrestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
-These variables were then compared to pre-season temperatures over the 9 year period of 2011-2019 to search for correlation between early-spring temps and growth phenology.
+It is difficult to map intraannual growth, or compare interspecific responses to Spring warming using leaf emergence timing alone.
+A solution to this is the use of dendrometer bands.
+Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale.
+Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014).
+This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions.
+Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
+Correlation between these variables and pre-season temperatures over the 9 year study period (2011-2019) were assessed using a heirarchal bayesian model.
 
 We hypothesized that (1) ring-porous and diffuse-porous trees would both reach 25, and 50% total annual growth earlier and (2) both wood types would increase total growth in response to warmer early-Spring temperatures.
-
-Here, we … - use 10 years of biweekly dendro band measurements for # species to characterize seasonal growth phenology
-•	combine these records with 110 years of tree-ring data for 12 species to examine how spring temperatures impact annual tree growth
-•	examine the effects of an extremely early spring (2020) seasonal growth phenology and annual growth
 
@@
 SCBI ForestGEO plot
--SCBI met-tower
-Climate data was obtained from a meteorological tower in a clearing adjacent to, and within elevation range of the SCBI ForestGEO plot.
-This tower is part of the ForestGEO meteorological monitoring program.
-Temperature readings are taken every 5 minutes using a CR1000 datalogger.
-9 years biweekly dendrometer measurements
+
+ForestGEO is a global network of scientists and 71 forest plots dedicated to the long-term study of the world's forests.
+The ForestGEO plot at SCBI is a 26-ha forest located in Front Royal, Virginia.
+Every stem ≥1 cm is tagged, mapped, identified to species, and measured in diameter at breast height (DBH) every 5 years following standardized ForestGEO procedure (Condit, 1995).
+Within the plot, >500 trees have dendrometer bands which are measured at the start and end of each growing season, while a subset of ~150 are additionally measured every two weeks during the growing season.
+
 Dendrometer band measurements were taken every two weeks for 9 years (After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees ).
-Data manually cleaned by visual inspection.
+The data were manually cleaned by visual inspection.
 Three classes of mistakes: 1.
@@
 If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.
+
+Climate data used was obtained from the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia.
+Average daily maximum temperature for 2011-2019 are from the Daily Summary dataset available on their website (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
 

### 2020-09-03 doc/manuscript/Dow_manuscript.Rmd
Update Dow_manuscript.Rmd
a bit of work on methods

@@
 
-Study site and data
-SCBI ForestGEO plot
+*Study sites and data*
 
-ForestGEO is a global network of scientists and 71 forest plots dedicated to the long-term study of the world's forests.
-The ForestGEO plot at SCBI is a 26-ha forest located in Front Royal, Virginia.
-Every stem ≥1 cm is tagged, mapped, identified to species, and measured in diameter at breast height (DBH) every 5 years following standardized ForestGEO procedure (Condit, 1995).
-Within the plot, >500 trees have dendrometer bands which are measured at the start and end of each growing season, while a subset of ~150 are additionally measured every two weeks during the growing season.
+Study sites included **2/3** temperate forests in the Eastern United states, **both/all** part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al.
+2015] (**Table 1?**).
+**Both/all** sites are dominated by broadleaf deciduous trees.
+*(insert site descriptions, including lat/long, elevation, plot size, climate info, dominant species.
+If we include SERC and if space permits, we may put this in a table, otherwise just text)*
 
-Dendrometer band measurements were taken every two weeks for 9 years (After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees ).
+Within the plot, metal dendrometer bands were installed on *>#* trees and measured approximately every 1-2 weeks within the growing season for *9* years (*list ranges, for each site if different*).
+*(don't bother to mention biannual bands, unless you end up using the data--which you could, for annual growth, if you'd like to increase the sample size.
+You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census *(Table S1)*.
+Across years, the number of bands sampled averaged *~150* at SCBI, # at Harvard Forest*, and # at SERC*.
+*(summary of how bands were distributed across species, focusing on ring/ diffuse porous distinction.
+details in an SI table)*
+
+(After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees *--put this in Table S1*).
+
+Measurements were timed to begin before beginning of spring growth and to continue through the cessation of growth in the fall.
+*(Give specifics, which may vary by site.
+Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)*
+
 The data were manually cleaned by visual inspection.
@@
 
-### Analysis
+*Analysis*
 

### 2020-09-03 RScripts/Climwin_Cameron.R
Update Climwin_Cameron.R

@@ -1,5 +1,6 @@
 #Climwin attempt
 library(climwin)
+library(readr)
 #NCDC_NOAA Front Royal station data
 climate <- read_csv("climate data/NCDC_NOAA_precip_temp.csv")
 climate <- climate[complete.cases(climate$TMAX),]
@@ -7,7 +8,7 @@ climate <- climate[complete.cases(climate$DATE),]
 
 #The data containing the biological responses for testing
 Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V4.csv") #Master datafrmae containing 20%, 50%, and 75% growth milestones
-twentyfive <- subset(Wood_pheno_table, perc == .25 & wood_type == "diffuse-porous") # subset to only 25%
+twentyfive <- subset(Wood_pheno_table, perc == .25 & wood_type == "ring porous") # subset to only 25%
 
 #We would want to test all three percentages (25, 50, 75) but I dont know how to.
 #I would also like to test how $Max_rate, $Max_rate_DOY, and $tot (total growth) are correlated with climate
@@ -19,22 +20,20 @@ for(i in c(2011:2019)){ #Assigns dates in the proper format for Climwin analysis
   biodata <- rbind(biodata, df)
 }
 
+round(mean(biodata$DOY))
 
-#biodata <- biodata[biodata$date %in% climate$DATE,] #Removes any biodata where climate data is missing. Is this necessary?
 
-#trythis <- subset(biodata, year != 2011)
 MassWin <- slidingwin(xvar = list(Temp = climate$TMAX),
                       cdate = climate$DATE,
                       bdate = biodata$date,
                       baseline = lm(DOY ~ 1, data = biodata),
                       cinterval = "day",
-                      range = c(153, 0),
+                      range = c(113, 0),
                       type = "relative",
                       stat = "mean",
                       func = "lin",
                       cmissing = "method2")
 
-#Best window is 37-0 days before 25% DOY?
 
 MassRand <- randwin(repeats = 5,
                     xvar = list(Temp = climate$TMAX),
@@ -42,10 +41,34 @@ MassRand <- randwin(repeats = 5,
                     bdate = biodata$date,
                     baseline = lm(DOY ~ 1, data = biodata),
                     cinterval = "day",
-                    range = c(50, 0), #make range the reference day to January 1st
+                    range = c(113, 0), #make range the reference day to January 1st
                     type = "relative", #make reference day the mean DOY
                     stat = "mean",
[…]

### 2020-09-09 doc/manuscript/Dow_manuscript.Rmd
MVP of Fig1

@@
 
-Here, I’ll insert a reference to Sean’s paper (McMahon & Parker, 2015). This is pulled from references.bib.
+Here, I’ll insert a reference to Sean’s paper [@mcmahon_general_2015]. This is pulled from references.bib.
 
@@
 
+### Figures
+
+```{r logistic-growth-model, out.width="100%", echo=FALSE, fig.align="center", fig.cap="Lorem ipsum"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/fig1.png")
+```
+
 ### References

### 2020-09-09 RScripts/additional_figures.R

@@ -1,0 +1,105 @@
+library(tidyverse)
+
+# Figure numbers from: https://github.com/SCBI-ForestGEO/growth_phenology/issues/13
+
+
+
+# Figure 1: Logistic growth curve and parameter illustration -------------------
+# Set parameters from McMahon & Parker (2014)
+# 1. L: lower asymptotes
+# 2. K: upper asymptotes
+# 3. doy.ip: inflection point
+# 4. r: slope of curve at inflection point
+# 5. theta: theta = 1 indicates pre/post inflection post symmetry
+K <- 13.2
+L <- 13.8
+doy.ip <- 200
+r <- 0.075
+theta <- 2
+params <- c(K, L, doy.ip, r, theta)
+
+# Function for logistic growth model written by Sean. Supposed to be in RDendrom
+# package https://github.com/seanmcm/RDendrom, but function does not seem to be included
+lg5.pred <- function(params, doy) {
+  L <- params[1] # min(dbh, na.rm = T)
+  K <- params[2]
+  doy.ip <- params[3]
+  r <- params[4]
+  theta <- params[5]
+  dbh <- vector(length = length(doy))
+  dbh <- L + ((K - L) / (1 + 1 / theta * exp(-(r * (doy - doy.ip) / theta))^theta))
+  return(dbh)
+}
+
+# Compute true values to plot true growth curve
+true_values <- tibble(
+  doy = seq(from = 100, to = 300),
+  diameter = lg5.pred(params, doy)
+)
+
+# Compute observed values (noise/error added) to plot with points
+set.seed(76)
+observed_values <- tibble(
+  doy = seq(from = 105, to = 295, length = 13),
+  diameter = lg5.pred(params, doy)
+) %>%
+  mutate(diameter = diameter + rnorm(n(), sd = 0.025))
+
+# Compute DOY of diameter quartiles (based on true values?)
+doy_diameter_quartile <- bind_rows(
+  true_values %>% filter(diameter >= quantile(diameter, probs = 0.25)) %>% slice(1),
+  true_values %>% filter(diameter >= quantile(diameter, probs = 0.50)) %>% slice(1),
+  true_values %>% filter(diameter >= quantile(diameter, probs = 0.75)) %>% slice(1),
+  true_values %>% filter(diameter >= quantile(diameter, probs = 1)) %>% slice(1)
+) %>%
+  mutate(
+    growth = c(0.25, 0.5, 0.75, 1)
+  )
+
[…]

### 2020-09-28 doc/manuscript/Dow_manuscript.Rmd
add to methods / reorganize intro

--- 2020-09-03 doc/manuscript/Dow_manuscript.Rmd
@@
 The absence of human intervention increases our reliance on natural systems to soften the blow of climate change.
-
-Forests are an extremely valuable ecosystem for many reasons.
-They house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil.
+Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services.
+For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil.
 Forests sequester about one-third of all anthropogenic CO2 and store ~45% of all terrestrial carbon (Bonan, 2008).
 However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).
+
 Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019).
 Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO2, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens.
-Understanding how growth is influenced by multiple factors at once is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation.
+Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.
+
+#relavant? reworded first line above#
+(Understanding how growth is influenced by multiple factors at once is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation.
 CO2 concentrations in the atmosphere are increasing, and in a perfect scenario, trees and other plants would simply speed up growth in response to this, increasing photosynthetic efficiency in a process called CO2 fertilization (Strain and Cure, 1986; Huang et al., 2007).
 However, factors other than CO2 also limit tree growth, and seasonal temperature averages and precipitation patterns are changing concurrently.
-It is anticipated that, throughout much of the United States, future heat and drought may limit tree growth more strongly than it can be enhanced by elevated CO2 (Charney et al, 2016).
+It is anticipated that, throughout much of the United States, future heat and drought may limit tree growth more strongly than it can be enhanced by elevated CO2 (Charney et al, 2016).)
+#
 
-In recent decades, an increase in average early Spring temperatures has been attributed to climate change.
-As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, meaning a potentially longer growing season (Ibáñez et al., 2010; Jeong et al., 2011).
-The effects of this extension of the growing season on leaf phenology and NPP have been well studied (Way, 2011; ), but little work has been done to understand the impacts to intraannual stem-growth variables.
-Shifts in leaf phenology timing could be coupled with a change in stem-growth phenology, which are likely to affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests.
-Thus, it is important to understand how the warming of Spring changes stem-growth phenology.
+In recent decades, an increase in average Spring temperatures has been attributed to climate change ().
+The effects of this Spring warming on leaf phenology have been extensively studied (Way, 2011; ) leading to many interesting observations.
+As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by earlier leaf senecence, meaning a potential shift in growing season timing (Ibáñez et al., 2010; Jeong et al., 2011).
+The timing of Spring warming also has a significant effect on leaf phenology response.
+Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
 
-The timing of Spring warming has a significant effect on leaf phenology response.
-Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
-GIS source Ian sent.
-However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood.
-Many predictions of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale observations of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
-In reality, we know little about the growth phenology of temperate deciduous species (see D’Orangeville et al., in revision for more info/ citations)—let alone how these are affected by an early spring.
+However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood (source?).
+Many observations of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
+Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests.
+In reality, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by an early spring.
 
-To date, very little research has been done on the phenology of stem-growth.
-Timing of intraannual growth is extremely improtant as it can dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ).
-For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered creating a negative climate feedback loop.
+To date, very little research, has been done on the phenology of stem-growth.
+Timing of intraannual growth is extremely important as it can dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ).
+For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered creating a negative climate feedback loop (source?).
 Recent results from D’Orangeville et al.
-(in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology, as ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time.
-This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses.
+(in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology.
+They showed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time.
+This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses (source for RP vs DP?).
 
-It is difficult to map intraannual growth, or compare interspecific responses to Spring warming using leaf emergence timing alone.
+While not entirely impossible, it is difficult to record intraannual growth, or compare interspecific responses to Spring warming using leaf emergence timing alone.
 A solution to this is the use of dendrometer bands.
-Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale.
+Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?).
 Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014).
@@
 Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
-Correlation between these variables and pre-season temperatures over the 9 year study period (2011-2019) were assessed using a heirarchal bayesian model.
+Correlation between these variables and SPring temperatures over the 9 year study period (2011-2019) were assessed using a heirarchal bayesian model.
 
-We hypothesized that (1) ring-porous and diffuse-porous trees would both reach 25, and 50% total annual growth earlier and (2) both wood types would increase total growth in response to warmer early-Spring temperatures.
+We hypothesized that (1) stem-growth phenology variables would occur earlier in warmer years as compared to cooler years and (2) both wood types would increase total growth and max growth rate in response to warmer Spring temperatures.
 
@@
 
-The data were manually cleaned by visual inspection.
+The dendroband data were manually cleaned by visual inspection.
 Three classes of mistakes: 1.
@@
 
-Climate data used was obtained from the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia.
-Average daily maximum temperature for 2011-2019 are from the Daily Summary dataset available on their website (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
+Daily maximum temperatures (TMAX) were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github).
+The R package climpact was used to plot TMAX for visual inspection (# points removed?).
+Climpact also identifies readings that were greater than 3 standard deviations away from yearly means.
+These points were labeled as outliers and removed from the dataset.
+Gaps in the SCBI meteorological tower data were subsequently filled using TMAX readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
 
@@
 
-The period where the phenology milestone(s) were most affected by Spring time temperature was determined using the R package Climwin.
-This package tests the correlation between climate variables (daily averaged temperature) and biological variables (DOY) within a specified time-frame, reporting the window with the highest correlation as the ‘best-model’.
-We instructed Climwin to search for the best window within ### to 0 days before the phenology milestone.
-This was done to find corresponding best-windows for both of our wood-types; ring-porous and diffuse-porous.
+The period where the phenology milestone(s) were most affected by changes in early Spring temperature was determined using the R package Climwin.
+This package tests the correlation between climate variables (TMAX) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'.
+as the best-model.
+We instructed Climwin to search for the best window beginning at February 1st and ending at the mean DOY for each phenology variable (chart in SI?).
+This was done to find corresponding (median of 95% CI window open and close) start and end dates for the analysis of ring-porous and diffuse-porous trees, as well as all trees pooled together.
 
@@
 
-A mixed effect model was used to test the response of wood phenology variables (25% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, and random effects of species and tag.
-We ran two separate models for each of our major wood-types and one combined model to use for comparison during the bayesian analysis.
-(talking to Albert about possibly doing this)
+A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, along with random effects of species and tag.
+We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison.
 
 •	Bayesian heirarchal model
+This mixed-effect model was run within a heirarchal bayesian framework.
 

### 2020-09-28 RScripts/wood_phenology_analysis.R
add ncdc comparisons for QC

@@ -173,7 +173,13 @@ fifty <- subset(Wood_pheno_table, perc == .50)
 seventyfive <- subset(Wood_pheno_table, perc == .75)
 #Temperature
 #NEON_summary_temp <- read_csv("climate data/NEON_summary_temp.csv")
-#weatherdata <- read_csv("climate data/NCDC_NOAA_precip_temp.csv")
+weatherdatancdc <- read_csv("climate data/NCDC_NOAA_precip_temp.csv")
+weatherdatancdc$DATE <- strptime(as.character(weatherdatancdc$DATE), format = "%d/%m/%Y")
+weatherdatancdc$DATE <- format(weatherdatancdc$DATE, "%Y-%m-%d")
+
+weatherdatancdc$doy <-yday(weatherdatancdc$DATE)
+weatherdatancdc$TMAX <- (weatherdatancdc$TMAX-32)*(5/9)
+weatherdatancdc <- weatherdatancdc[complete.cases(weatherdatancdc$TMAX),]
 #weatherdata$months <- months(as.Date(weatherdata$DATE))
 #weatherdata <- weatherdata[complete.cases(weatherdata$TMAX),]
 #weatherdatarp <- weatherdata[complete.cases(weatherdata$flag),]
@@ -198,27 +204,52 @@ weatherdata$doy <-yday(weatherdata$DATE)
 rp25 <- subset(weekly_climwin_resultsnew, wood_type =="ring porous" & percs == .25)
 rp25seq <- seq(rp25$opendoy, rp25$closedoy, 1)
 weatherdatarp25 <- weatherdata[weatherdata$doy %in% rp25seq,]
-
+#NCDC comparison
+weatherdatarp25_NCDC <- weatherdatancdc[weatherdatancdc$doy %in% rp25seq,]
+aggregate(weatherdatarp25$TMAX, by = list(weatherdatarp25$year), FUN = mean)
+aggregate(weatherdatarp25_NCDC$TMAX, by = list(weatherdatarp25_NCDC$year), FUN = mean)
+###
 rp50 <- subset(weekly_climwin_resultsnew, wood_type =="ring porous" & percs == .50)
 rp50seq <- seq(rp50$opendoy, rp50$closedoy, 1)
 weatherdatarp50 <- weatherdata[weatherdata$doy %in% rp50seq,]
-
+#NCDC comparison
+weatherdatarp50_NCDC <- weatherdatancdc[weatherdatancdc$doy %in% rp50seq,]
[…]

### 2020-10-01 doc/manuscript/Dow_manuscript.Rmd
intro edits

@@
 
-As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented to combat this, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
+As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
 The absence of human intervention increases our reliance on natural systems to soften the blow of climate change.
@@
 
-In recent decades, an increase in average Spring temperatures has been attributed to climate change ().
-The effects of this Spring warming on leaf phenology have been extensively studied (Way, 2011; ) leading to many interesting observations.
-As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by earlier leaf senecence, meaning a potential shift in growing season timing (Ibáñez et al., 2010; Jeong et al., 2011).
+In recent decades, an increase in average Spring temperatures has been attributed to climate change.
+This relatively small increase has already been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).
+The effects of this Spring warming on leaf phenology have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018) leading to many interesting observations.
+As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by earlier leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
 The timing of Spring warming also has a significant effect on leaf phenology response.
 Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
+Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
 
-However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood (source?).
-Many observations of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
-Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests.
-In reality, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by an early spring.
+However, the fine-scale responses of temperate deciduous species to a warming Spring are not commonly researched due to the increased effort required.
+Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
+Many observations of climate induced phenological shifts are based on GIS, eddy-covarience measurements, or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
+To date, very little research, has been done on the phenology of stem-growth.
 
-To date, very little research, has been done on the phenology of stem-growth.
-Timing of intraannual growth is extremely important as it can dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ).
-For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered creating a negative climate feedback loop (source?).
-Recent results from D’Orangeville et al.
+In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns.
+Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ).
+For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered contributing to a negative climate feedback loop (source?).
+Recent results from a paper by D’Orangeville et al.
 (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology.
@@
 
-While not entirely impossible, it is difficult to record intraannual growth, or compare interspecific responses to Spring warming using leaf emergence timing alone.
+While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations.
 A solution to this is the use of dendrometer bands.
@@
 Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
-Correlation between these variables and SPring temperatures over the 9 year study period (2011-2019) were assessed using a heirarchal bayesian model.
+Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
 

### 2020-10-07 RScripts/bert/test.R
cleaned test code

@@ -1,19 +1,103 @@
 # 2020/9/16, 2020/10/5, 2020/10/6 -------
 # Testing plotting credible intervals from http://mjskay.github.io/tidybayes/
+library(tidyverse)
+library(lubridate)
 library(rstanarm)
 library(tidybayes)
-library(plotly)
 library(broom)
 library(modelr)
+library(patchwork)
+
+
+# Visualize relationship between DOY and marchmeans ----------------------------
+# Get growth data
+Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V4.csv") %>%
+  # Keep only RP and DP for now
+  filter(wood_type != "other") %>%
+  # Rename ring porous to not have a space
+  mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
+
+# Create temperature variable
+# 0. Get all weather data
+weatherdata <-
+  read_csv("climate data/NCDC_NOAA_precip_temp.csv") %>%
+  mutate(
+    DATE = dmy(DATE),
+    months = month(DATE, label = TRUE, abbr = FALSE)
+  ) %>%
+  # Remove entries with no tmax data
+  filter(!is.na(TMAX)) %>%
+  # Rename RP flag set by Cam
+  rename(flagrp = flag)
+
+
+# 1. Get mean march daily maximum temperatures
+marchmeans <- weatherdata %>%
+  filter(months == "March") %>%
+  group_by(year) %>%
+  summarize(marchmean = mean(TMAX))
+
+
+# 2.a) EDA of climwin windows
+# RP climwin window is around 3/15 to 4/23
+weatherdata %>%
+  filter(flagrp == "RP") %>%
+  mutate(DOY = yday(DATE)) %>%
+  arrange(DOY) %>%
+  slice(c(1, n()))
+
+# DP climwin window is around 3/27 to 6/2
+weatherdata %>%
+  filter(flagdp == "DP") %>%
+  mutate(DOY = yday(DATE)) %>%
+  arrange(DOY) %>%
+  slice(c(1, n()))
+
+climwin_windows <-
+  tibble(
+    wood_type = c("diffuse-porous", "ring-porous"),
+    window = c("climwin window: 3/27 - 6/2", "climwin window: 3/15 - 4/23")
+  )
+
+
+# 2.b) Get mean climwin daily maximum temperatures
+# RP separately
[…]

### 2020-10-07 doc/manuscript/Dow_manuscript.Rmd
Done Figure 6

--- 2020-09-09 doc/manuscript/Dow_manuscript.Rmd
@@
 title: ''
-output: word_document
+output: 
+  word_document: 
+    fig_caption: yes
 bibliography: references.bib
@@
 
-```{r logistic-growth-model, out.width="100%", echo=FALSE, fig.align="center", fig.cap="Lorem ipsum"}
+**Figure 1** 
+```{r fig1, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Logistic growth model"}
 # Created in RScripts/additional_figures.R
@@
 
+**Figure 6**
+
+Notes for now:
+
+1. Model was fit to a random sample of 20 out of 106 tags to speed up computation.
+1. Climwin mean temperatures are recentered at 65F to help sanity check intercepts and slopes. 65F was chosen because it is a "typical" climwinmean value
+1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
+    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
+    1. Random effects: tag number (no species just yet)
+
+```{r, echo = FALSE}
+fig6_caption <- "Relationship between DOY and climwin mean tempeatures. For each 
+observed climwin mean temperature value, the posterior mean of the fitted day of year
+is represented by the solid blue line and credible intervals are represented by bands."
+```
+
+```{r fig6, out.width="100%", echo=FALSE, fig.align="center", fig.cap=fig6_caption}
+# Created in RScripts/bert/wood_phenology_analysis_bert.R
+knitr::include_graphics("tables_figures/fig6.png")
+```
+
 ### References

### 2020-10-07 RScripts/bert/wood_phenology_analysis_bert.R

@@ -1,10 +1,11 @@
 # Load packages and data ---------------------------------------
 library(tidyverse)
 library(lubridate)
-library(rstanarm)
 library(tidybayes)
 library(patchwork)
-
+library(knitr)
+library(rstanarm)
+options(mc.cores = parallel::detectCores())
 
 
 
@@ -92,156 +93,191 @@ Wood_pheno_table <- Wood_pheno_table %>%
       perc == 0.5 ~ "DOY_50",
       perc == 0.75 ~ "DOY_75"
     )
-  )
+  ) %>%
+  arrange(tag, year)
 View(Wood_pheno_table)
 
 
+# TODO Erase later: Run analysis
+# - Only for subset of tags to speed up computation
+# - Recenter all climwin values at 65 since mean of climwinmeans is 66.310
+climwin_mean <- mean(climwinmeans$climwinmean)
+climwin_mean
 
+set.seed(76)
+sample_tags <- Wood_pheno_table %>%
+  pull(tag) %>%
+  unique() %>%
+  sample(20)
 
-# First model: Fit multivariate model using marchmeans -----------------------
-joint_model_marchmeans <- stan_mvmer(
-  formula = list(
-    DOY_25 ~ wood_type + wood_type:marchmean + (1|sp) + (1|tag),
-    DOY_50 ~ wood_type + wood_type:marchmean + (1|sp) + (1|tag),
-    DOY_75 ~ wood_type + wood_type:marchmean + (1|sp) + (1|tag)
-  ),
-  # Note we transform the data from tall/tidy format to wide format first:
-  data = Wood_pheno_table %>% pivot_wider(names_from = perc, values_from = DOY),
-  seed = 349,
-  # Once we feel good about our results, increase these values. The code will
-  # take longer to run however
-  chains = 2,
-  iter = 1000
-)
-
-# Used to identify fixed effects of interest:
-joint_model_marchmeans %>%
-  get_variables() %>%
-  dput()
-
-# Extract draws from posterior distributions
-# See http://mjskay.github.io/tidybayes/articles/tidy-rstanarm.html
-posterior_draws_marchmeans <- joint_model_marchmeans %>%
-  gather_draws(
-    # This is a hack:
-    `y1|(Intercept)`, `y2|(Intercept)`, `y3|(Intercept)`, `y1|wood_typering-porous`,
-    `y1|wood_typediffuse-porous:marchmean`, `y1|wood_typering-porous:marchmean`,
[…]

### 2020-10-07 RScripts/SCBI_wood_phenology.R
adding all_stems modeled growth data csv

@@ -532,6 +532,9 @@ for (e in 1:length(all_stems_intra)) { # Loop to merge all dendroband surveys in
 }
 all_stems <- all_stems[complete.cases(all_stems$dbh2), ]
 all_stems <- all_stems[, c(1:31, 34, 35)]
+write.csv(all_stems, file = "Data/all_stems.csv", row.names = FALSE)
+
+
 # Objects needed for the following loop
 data <- data.frame(NULL)
 # throwaways <- data.frame(NULL)

### 2020-10-07 RScripts/SCBI_wood_phenology.R
modified script to output LG5 parameter values for each tag-year

@@ -566,6 +566,16 @@ colnames(masterDF) <- c("tot", "perc", "tag", "DOY", "sp", "year", "dbh", "max_r
 # 4. Loop to create a final "master dataframe" -------
 # The following loop will cycle through years, species, and tags to create a final "master dataframe"
 # It includes two of Sean's functions wrapped with some of my code
+
+# Added by bert: track LG5 parameters for each tag/year in this list
+tag_years <- all_stems %>%
+  transmute(tag_year = str_c(tag, year, sep = "-")) %>%
+  pull(tag_year) %>%
+  unique()
+
+LG5_parameters <- vector(mode = "list", length = length(tag_years))
+names(LG5_parameters) <- tag_years
+
 for (q in 2011:2019) {
   skip_to_next <- FALSE
   Stem2 <- subset(all_stems, year == q)
@@ -836,6 +846,15 @@ for (q in 2011:2019) {
       names(Param.df) <- c("L", "K", "doy_ip", "r", "theta", "a", "b")
       # write.csv(Param.df, file = "SCBI_hi_lo_lg5.csv", quote = FALSE, row.names = FALSE)
 
+      # Added by bert: save tag/year + parameter values
+      param_values_to_save <- c(stemstag$tag[1], stemstag$year[1],  Param.df) %>% unlist()
+      names(param_values_to_save) <- c("tag", "year", names(Param.df))
+      tagyear_id <- stemstag %>%
+        slice(1) %>%
+        transmute(tag_year = str_c(tag, year, sep = "-")) %>%
+        pull(tag_year)
+      LG5_parameters[[tagyear_id]] <- param_values_to_save
+
 
       # Max growth rate DOY
       maxDOY <- NULL
@@ -947,7 +966,10 @@ masterDF <- masterDF[-1, ]
 masterDF$wood_type <- ifelse(masterDF$sp == "quru" | masterDF$sp == "qual", "ring porous", ifelse(masterDF$sp == "litu" | masterDF$sp == "fagr", "diffuse-porous", "other"))
 write.csv(masterDF, file = "Data/Wood_pheno_table_V4.csv", row.names = FALSE)
 
-
+# Added by bert: save parameter values
+LG5_parameters %>%
+  bind_rows() %>%
+  write_csv(file = "Data/LG5_parameter_values.csv")
 
 
 

### 2020-10-07 doc/manuscript/Dow_manuscript.Rmd
Done first pass of figures 2 and 3

@@
 **Figure 1**
+
+Based on D'Orangeville Figure 1
+
 ```{r fig1, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Logistic growth model"}
@@
 ```
+
+**Figure 2**
+
+Based on D'Orangeville Figure 3.a), but with relative growth
+
+```{r fig2, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Percent total (modeled) growth"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/fig2.png")
+```
+
+**Figure 3**
+
+Similar to D'Orangeville Figure 4.a), but with cummalative relative growth
+
+```{r fig3, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Cummulative percent total (modeled) growth"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/fig3.png")
+```
+
+**Figure 4**
+
+**Figure 5**
 

### 2020-10-07 RScripts/additional_figures.R

@@ -1,8 +1,8 @@
 library(tidyverse)
+library(scales)
+library(tidybayes)
 
 # Figure numbers from: https://github.com/SCBI-ForestGEO/growth_phenology/issues/13
-
-
 
 # Figure 1: Logistic growth curve and parameter illustration -------------------
 # Set parameters from McMahon & Parker (2014)
@@ -76,7 +76,7 @@ create_growth_polygon <- function(percent){
 
 
 # Output figure
-ggplot() +
+fig1<- ggplot() +
   # Mark 25%, 50%, 75%, 100% DOY
   geom_vline(data = doy_diameter_quartile, mapping = aes(xintercept = doy), linetype = "dashed", col = "grey") +
   geom_polygon(data = create_growth_polygon(percent = 0.25), mapping = aes(x = doy, y = diameter), fill = grey(level = 0.8)) +
@@ -90,6 +90,7 @@ ggplot() +
   geom_point(data = observed_values, mapping = aes(x = doy, y = diameter), shape = 21, colour = "black", fill = "white", size = 3, stroke = 1) +
   # Plot
   labs(x = "Day of year", y = "Diameter (cm)")
+fig1
 # TODO: Add
 # - Explanations of parameters
 # - 25-75 DOY window
@@ -97,9 +98,87 @@ ggplot() +
 # - total annual growth
 
 fig_width <- 8
-ggsave("doc/manuscript/tables_figures/fig1.png", width = fig_width, height = fig_width / 1.52)
+ggsave("doc/manuscript/tables_figures/fig1.png", plot = fig1, width = fig_width, height = fig_width / 1.52)
 
 
 
-# Figure 2:
+
+
+
+
+# Fig2 & Fig3: Percent modeled growth and cummulative percent modeled growth ---
+Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V4.csv") %>%
+  # Keep only RP and DP for now
+  filter(wood_type != "other") %>%
+  # Rename ring porous to not have a space
+  mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
+
+
+# As generated in Rscripts/SCBI_wood_phenology.R
+LG5_parameter_values <- read_csv("Data/LG5_parameter_values.csv")
+
+# Generalized 5-parameter logistic function (modified version of Sean's function)
[…]

### 2020-10-07 RScripts/bert/wood_phenology_analysis_bert.R

@@ -4,8 +4,9 @@ library(lubridate)
 library(tidybayes)
 library(patchwork)
 library(knitr)
-library(rstanarm)
+library(scales)
 options(mc.cores = parallel::detectCores())
+library(rstanarm)
 
 
 
@@ -15,7 +16,6 @@ Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V4.csv") %>%
   filter(wood_type != "other") %>%
   # Rename ring porous to not have a space
   mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
-
 
 
 
@@ -98,6 +98,9 @@ Wood_pheno_table <- Wood_pheno_table %>%
 View(Wood_pheno_table)
 
 
+
+
+# Fit multivariate model using climwinmeans ------------------------------------
 # TODO Erase later: Run analysis
 # - Only for subset of tags to speed up computation
 # - Recenter all climwin values at 65 since mean of climwinmeans is 66.310
@@ -114,12 +117,10 @@ Wood_pheno_table <- Wood_pheno_table %>%
   filter(tag %in% sample_tags) %>%
   mutate(climwinmean = climwinmean - 65)
 
-
-
-
-# Fit multivariate model using climwinmeans ------------------------------------
+# Delete all non-needed columns
 Wood_pheno_table <- Wood_pheno_table %>%
   select(perc, tag, year, wood_type, sp, climwinmean, starts_with("DOY"))
+
 
 # Convert to wide format for use in rstanarm::stan_mvmer()
 Wood_pheno_table_wide <- Wood_pheno_table %>%
@@ -217,11 +218,6 @@ bayesian_regression_table %>%
 
 
 
-# Analyze random effects ---------------------------------
-# Skipped
-
-
-
 
 # Fig6: Plot of regression of DOY over climwinmeans with credible intervals ----
 # Extract predicted DOY_25, DOY_50, DOY_75
@@ -265,19 +261,4 @@ ggsave(filename = "doc/manuscript/tables_figures/fig6.png", width = 14.7*.7, hei
 
 # Sanity check this plot with regression table intercepts and slopes
 posterior_means_fixed_effects
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
 

### 2020-10-07 RScripts/SCBI_wood_phenology.R
added comment

@@ -532,6 +532,8 @@ for (e in 1:length(all_stems_intra)) { # Loop to merge all dendroband surveys in
 }
 all_stems <- all_stems[complete.cases(all_stems$dbh2), ]
 all_stems <- all_stems[, c(1:31, 34, 35)]
+
+# Added by bert: write all_stems observed data to csv
 write.csv(all_stems, file = "Data/all_stems.csv", row.names = FALSE)
 
 

### 2020-10-08 RScripts/wood_phenology_analysis.R
Update wood_phenology_analysis.R

@@ -292,11 +292,11 @@ twentyfiveRP <- merge(rpmeans25, twentyfiveRP, by = "year")##Change rpmeans## he
 
 colnames(rpmeans50) <- c("year", "rptemp")
 fiftyRP <- subset(fifty, wood_type == "ring porous")
-fiftyRP <- merge(fiftyRP, rpmeans25, by = "year")##Change rpmeans## here
+fiftyRP <- merge(fiftyRP, rpmeans50, by = "year")##Change rpmeans## here
 
 colnames(rpmeans75) <- c("year", "rptemp")
 seventyfiveRP <- subset( seventyfive, wood_type == "ring porous")
-seventyfiveRP <- merge(seventyfiveRP, rpmeans25, by = "year")##Change rpmeans## here
+seventyfiveRP <- merge(seventyfiveRP, rpmeans75, by = "year")##Change rpmeans## here
 
 #Diffuse porous - repeat of RP process
 #dpmeans <- subset(tempmaxmeansdp, Group.2 == "DP")

### 2020-10-08 doc/manuscript/Dow_manuscript.Rmd

--- 2020-10-01 doc/manuscript/Dow_manuscript.Rmd
@@
 
+### Figures
+
+**Figure 1**
+
+Based on D'Orangeville Figure 1
+
+```{r fig1, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Logistic growth model"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/fig1.png")
+```
+
+**Figure 2**
+
+Based on D'Orangeville Figure 3.a), but with relative growth
+
+```{r fig2, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Percent total (modeled) growth"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/fig2.png")
+```
+
+**Figure 3**
+
+Similar to D'Orangeville Figure 4.a), but with cummalative relative growth
+
+```{r fig3, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Cummulative percent total (modeled) growth"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/fig3.png")
+```
+
+**Figure 4**
+
+**Figure 5**
+
+**Figure 6**
+
+Notes for now:
+
+1. Model was fit to a random sample of 20 out of 106 tags to speed up computation.
+1. Climwin mean temperatures are recentered at 65F to help sanity check intercepts and slopes. 65F was chosen because it is a "typical" climwinmean value
+1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
+    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
+    1. Random effects: tag number (no species just yet)
+
+```{r, echo = FALSE}
+fig6_caption <- "Relationship between DOY and climwin mean tempeatures. For each 
+observed climwin mean temperature value, the posterior mean of the fitted day of year
+is represented by the solid blue line and credible intervals are represented by bands."
+```
+
+```{r fig6, out.width="100%", echo=FALSE, fig.align="center", fig.cap=fig6_caption}
+# Created in RScripts/bert/wood_phenology_analysis_bert.R
+knitr::include_graphics("tables_figures/fig6.png")
+```
+
 ### References

### 2020-10-10 doc/manuscript/Dow_manuscript.Rmd
brief description of study sites / start of new discussion point

@@
 
+Our first study site was the forest dynamics plot located at the Smithsonian COnservation Biology Institute 5 KM South of Front Royal, Virginia within the Blue Ridge Mountains (38 °53036.6″ N, 78°08043.4″W; elevation 273–338 m).
+The forest here was established in the mid-19th century after conversion from agricultural fields.
+Dominate canopy species within the 25.6 HA forest plot include tulip poplar (Liriodendron tulipifera), hickories (Carya spp.), and oaks (Quercus spp.).
+The climate is humid temperate, with a mean annual precipitation of 998 mm, January temperatures averaging 1 degrees C and July temperatures averaging 24 degrees C (Helcoski et al, 2019).
+The second study site was Harvard Forest in the central Massachusetts town of Petersham (42°53′ N, 72°17′ W, elevation?).
+Harvard forest was established in 1907 as a study site for university students and researchers.
+(Previous use?) Dominate species within the approximately 13 HA plot are hemlock (Tsuga spp.), oak (Quercus spp.) and maple (Acer spp.).
+The climate is temperate continental with a mean annual precipitation of 1120 mm, January temperatures averaging -12 degrees C and July temperatures averaging 19 degrees C (D'Orangeville et al, ).
+
+The third study site was the forest dynamics plot located at the Smithsonian Ecological Research Center in Edgewater, Maryland near a sub estuary of the Chesapeake Bay (38°88′ N, 76°55′ W, elevation?).
+The forest here established roughly 120 years ago, while the plot was established in 2007.
+The 16 HA plot is dominated by tulip poplar (Liriodendron tulipifera), oak (Quercus spp.), beech (Fagus spp.), hickory (Carya spp.), and maple (Acer spp.).
+(Climate summary)
+
 Within the plot, metal dendrometer bands were installed on *>#* trees and measured approximately every 1-2 weeks within the growing season for *9* years (*list ranges, for each site if different*).
@@
 
+This analysis is one of a first generation of intraannual tree growth studies and as such, is the metaphorical tip of the ice berg.
+Here, we focused our analysis purely on early Spring warming.
+This falls short of the reality of our changing climate as every season if affected in some way.
+For example, diffuse porous trees seemed to show little, if any, relationship with the above climate window, but if we shift that window forward 20 days, the effect is much more severe, signalling a greater impact of May/June TMAX on trees of that wood type (figure or table?).
+A paper by Helcoski et al (2019) found that trees at SCBI had a significantly negative correlation between tree growth and May temperatures, agreeing with the results here.
+Other studies have found that Autumn leaf senescence  is being delayed, potentially in response to warmer temperatures in the previous year's Winter or Autumn (), this warming also impacts chilling requirements as well (), total growth is being stunted by drought and high TMAX in the current year's Summer ().
+How these realtionships may impact intraannual growth patterns or variables like thos described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting our forests.
+
+We present a mere stepping stone for future studies to build upon.
+
 Diffuse-porous 25-75% growth period shrinking leaving trees more vulnerable to drought while ring-porous 25-75% growing, potentially making effect of drought less damaging?

### 2020-10-12 doc/manuscript/Dow_manuscript.Rmd
Fleshing out discussion outline based on current results

@@
 ### Discussion
+-Growth timing
+We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures.
+The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here.
+Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,).
+On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al).
+These differences are likely evolutionary hydraulic adaptations to climate().
+The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought ().
+The pre-leaf stem growth occurring in these species is fueled by carbon stores often used for other cell maintenance activities within the plant ().
+The dependence on these carbon stores for cambial activation could prove to be a detriment to the tree, as exhaustion of reserves leads to carbon starvation ().
+Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, focusing on a faster, shorter growing season in the Summer.
+This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer focused growth period.
 
-This analysis is one of a first generation of intraannual tree growth studies and as such, is the metaphorical tip of the ice berg.
+The mechanisms controlling activation of cambium reactivation is still a mystery, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves().
+However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.
+
+Ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while 75% remains statistically unchanged, in response to warmer temperautres during early Spring.
+This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
+The reactivation of cambial cells is likely beginning earlier potentially exposing these trees to higher risk of cavitation or leaf damage during frost events ().
+Damage of this nature would then need to be repaired using carbon reserves which may be in short demand as they are already being used for initial growth.
+Hypothetically, this could lead to carbon starvation in some trees.
+Meanwhile, diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps.
+This disconnect means that diffuse-porous trees will continue tofocus their growing season in the Summer, which is projected to become less condusive to tree growth as temperatures and drought intensity increase.
+However, diffuse-porous trees are well adapted to drought conditions, and may continue to be successful even in future climate scenarios.
+If not, and they begin to suffer without being able to adapt by shifting their growing period earlier, the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.
+(Is mortality of ring-porous species higher in recent decades?
+Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)
+
+This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is the metaphorical tip of the ice berg.
 Here, we focused our analysis purely on early Spring warming.
-This falls short of the reality of our changing climate as every season if affected in some way.
-For example, diffuse porous trees seemed to show little, if any, relationship with the above climate window, but if we shift that window forward 20 days, the effect is much more severe, signalling a greater impact of May/June TMAX on trees of that wood type (figure or table?).
-A paper by Helcoski et al (2019) found that trees at SCBI had a significantly negative correlation between tree growth and May temperatures, agreeing with the results here.
-Other studies have found that Autumn leaf senescence  is being delayed, potentially in response to warmer temperatures in the previous year's Winter or Autumn (), this warming also impacts chilling requirements as well (), total growth is being stunted by drought and high TMAX in the current year's Summer ().
-How these realtionships may impact intraannual growth patterns or variables like thos described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting our forests.
+This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth.
+For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?).
+This revelation agrees with the results from Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year using tree cores.
+Additionally, Other studies have found that Autumn leaf senescence is being delayed potentially in response to warmer temperatures in the previous year's Winter or Autumn (), changes in chilling requirements due to current year Autumn and winter temperatures (), and stunted total productivity induced by drought and high TMAX in the current year's Summer ().
+How these relationships may impact intraannual growth patterns or variables like those described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
 

### 2020-10-13 doc/manuscript/Dow_manuscript.Rmd
adding sources

@@
 We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison.
+Models were run using the LME4 package in R (Bates, D., Mächler, M., Bolker, B.
+& Walker, S.
+(2015).
+Fitting linear mixed‐effects models using lme4.
+J.
+Stat.
+Softw., 67, 1–48.)
 
@@
 
-•	Bayesian variable results (PRE-JOINT MODEL, updating ASAP)
-
-Ring-porous 25% DOY estimate: -2.23 (credible interval: -2.94 - -1.55)
-
-Ring-porous 50% DOY estimate: - 2.10 (credible interval: -2.71 - -1.42)
-
-Ring-porous 75% DOY estimate: -1.89 (credible interval: -2.53 - -1.28)
-
-Ring-porous total growth estimate: 0.0552 (credible interval: 0.00346 - 0.108)
-
-Ring-porous max-rate DOY estimate: -2.20 (credible interval: -2.77 - -1.62)
-
-Ring-porous max-rate estimate: 0.00113 (credible interval: 0.000151 – 0.00208)
-
-Diffuse-porous 25% DOY estimate: -1.83 (credible interval: -2.86 - -0.822)
-
-Diffuse-porous 50% DOY estimate: -2.39 (credible interval: -3.35 - -1.43)
-
-Diffuse-porous 75% DOY estimate: -2.96 (credible interval: -4.03 - -1.92)
-
-Diffuse-porous total growth estimate: -0.0642 (credible interval: -0.172 – 0.0460)
-
-Diffuse-porous max-rate DOY estimate: -2.54 (credible interval: -3.41 - -1.69)
-
-Diffuse-porous max-rate estimate: 0.00209 (credible interval: 0.000118 – 0.00413)
-
 ### Discussion
@@
 These differences are likely evolutionary hydraulic adaptations to climate().
-The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought ().
-The pre-leaf stem growth occurring in these species is fueled by carbon stores often used for other cell maintenance activities within the plant ().
+The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990).
+The pre-leaf stem growth occurring in these species is fueled by carbon stores, often used for other cell maintenance activities within the plant ().
 The dependence on these carbon stores for cambial activation could prove to be a detriment to the tree, as exhaustion of reserves leads to carbon starvation ().
-Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, focusing on a faster, shorter growing season in the Summer.
-This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer focused growth period.
+Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer.
+This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
 
@@
 
-Ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while 75% remains statistically unchanged, in response to warmer temperautres during early Spring.
+Ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while 75% remains statistically unchanged, in response to warmer temperatures during early Spring.
 This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
-The reactivation of cambial cells is likely beginning earlier potentially exposing these trees to higher risk of cavitation or leaf damage during frost events ().
+The reactivation of cambial cells is likely beginning earlier potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
 Damage of this nature would then need to be repaired using carbon reserves which may be in short demand as they are already being used for initial growth.
 Hypothetically, this could lead to carbon starvation in some trees.
+Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.
+In the long term, this could lead to changes in species dominance within temperate forests (talk more aobut this?).
+
 Meanwhile, diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps.
-This disconnect means that diffuse-porous trees will continue tofocus their growing season in the Summer, which is projected to become less condusive to tree growth as temperatures and drought intensity increase.
-However, diffuse-porous trees are well adapted to drought conditions, and may continue to be successful even in future climate scenarios.
-If not, and they begin to suffer without being able to adapt by shifting their growing period earlier, the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.
-(Is mortality of ring-porous species higher in recent decades?
+This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase.
+However, diffuse-porous trees are adapted to survive in drought conditions, and may continue to be successful even in future climate scenarios (RIght?
+seeing conflicting reports.
+Sources.).
+Studies have found that these diffuse-porous trees in the Eastern USA show the largest reduction in ring-width increment during, and following, Summer droughts (Kannenberg et al, 2018) as well as being generally more sensitive to Summer precipitation patterns (Elliot et al, 2015).
+The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition.
+--the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades?
 Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)
@@
 This revelation agrees with the results from Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year using tree cores.
-Additionally, Other studies have found that Autumn leaf senescence is being delayed potentially in response to warmer temperatures in the previous year's Winter or Autumn (), changes in chilling requirements due to current year Autumn and winter temperatures (), and stunted total productivity induced by drought and high TMAX in the current year's Summer ().
+Additionally, Other studies have found that Autumn leaf senescence is being delayed potentially in response to warmer temperatures in the previous year's Winter or Autumn (), changes in chilling requirements due to current year Autumn and winter temperatures (), and stunted total productivity induced by drought and high TMAX in the current year's Summer (), diffuse porous trees increased growth during the following growing season when winter temperatures (TMIN) were lower (Benito & Pederson, 2015; Elliot et al, 2015).
 How these relationships may impact intraannual growth patterns or variables like those described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
 
-We present a mere stepping stone for future studies to build upon.
-
-Diffuse-porous 25-75% growth period shrinking leaving trees more vulnerable to drought while ring-porous 25-75% growing, potentially making effect of drought less damaging?
-But RP still prone to cavitation + longer growth period could mean exposure to higher number of droughts resulting in serious damage to water transport system?
-
-DIfferences in total relationship could be because of response difference in wood types OR could be because the diffuse-porous’ growth occuring during the later summer means it is exposed to drought/heat and on average, not feeling the effects of the earlier spring.
-differences in wood porosity growth timing means exposure to different periods of climate stress ring porous adaptable to spring temps, diffuse not so much leads to higher growth in ring however, diffuse still more resistant to cavitation potential for more frequent late frost events having large impacts as spring warms?
-offset of chilling requirements and spring leaf out requirements
-
-much assumed about future forest productivity based on gis different wood types/species may respond differently need to understand at a finer scale what is happening
+We present a stepping stone for future studies to build upon.
 

### 2020-10-15 RScripts/wood_phenology_analysis_bert_updated.R
add plots to end

@@ -9,11 +9,12 @@ options(mc.cores = parallel::detectCores())
 library(rstanarm)
 
 
-
 # Get growth data ----------------------------------
-Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V4.csv") %>%
+Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V6.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
+  filter(tot >= 1) %>%
+  filter(tot <= 12.06)%>%
   # Rename ring porous to not have a space
   mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
 
@@ -66,7 +67,6 @@ marchmeans <- weatherdata %>%
   group_by(year) %>%
   summarize(marchmean = mean(cleantmax))
 
-
 # 2.a) EDA of climwin windows
 # RP climwin window is around 3/15 to 4/23
 #weatherdata %>%
@@ -85,7 +85,7 @@ marchmeans <- weatherdata %>%
 climwin_windows <-
   tibble(
     wood_type = c("diffuse-porous", "ring-porous"),
-    window = c("climwin window: 3/27 - 6/2", "climwin window: 3/5 - 4/5")
+    window = c("climwin window: 3/8 - 5/14", "climwin window: 3/5 - 4/5")
   )
 
 
@@ -108,7 +108,7 @@ climwinmeans_rp <- weatherdata %>%
 #  summarize(climwinmean = mean(TMAX)) %>%
 #  mutate(wood_type = "diffuse-porous")
 climwinmeans_dp <- weatherdata %>%
-  filter(doy %in% c(climwindows[4,11]:climwindows[4,12])) %>% #86:156
+  filter(doy %in% c(climwindows[4,11]:climwindows[4,12])) %>% #68:135
   group_by(year) %>%
   summarize(climwinmean = mean(cleantmax)) %>%
   mutate(wood_type = "diffuse-porous")
@@ -125,7 +125,7 @@ Wood_pheno_table <- Wood_pheno_table %>%
   left_join(fiftyseventy, by = c("tag", "year")) %>%
   left_join(twofifty, by = c("tag", "year")) %>%
   # Remove other variables
-  select(-c(tot, dbh, max_rate_DOY, max_rate)) %>%
+  #select(-c(tot, dbh, max_rate_DOY, max_rate)) %>%
   mutate(
     perc = case_when(
       perc == 0.25 ~ "DOY_25",
[…]

### 2020-10-15 RScripts/SCBI_wood_phenology.R
add error messages to trycatch's

@@ -535,15 +535,8 @@ all_stems <- all_stems[, c(1:31, 34, 35)]
 
 # Added by bert: write all_stems observed data to csv
 write.csv(all_stems, file = "Data/all_stems.csv", row.names = FALSE)
+all_stems <- read.csv("Data/all_stems.csv")
 
-
-# Objects needed for the following loop
-data <- data.frame(NULL)
-# throwaways <- data.frame(NULL)
-check_list <- NA
-DOY2 <- NULL
-masterDF <- data.frame(1, 2, 3, 4, 5, 6, 7, 8, 9)
-colnames(masterDF) <- c("tot", "perc", "tag", "DOY", "sp", "year", "dbh", "max_rate_DOY", "max_rate")
 
 
 ## Manual check of data ##
@@ -569,6 +562,14 @@ colnames(masterDF) <- c("tot", "perc", "tag", "DOY", "sp", "year", "dbh", "max_r
 # The following loop will cycle through years, species, and tags to create a final "master dataframe"
 # It includes two of Sean's functions wrapped with some of my code
 
+# Objects needed for the following loop
+data <- data.frame(NULL)
+# throwaways <- data.frame(NULL)
+check_list <- NA
+DOY2 <- NULL
+masterDF <- data.frame(1, 2, 3, 4, 5, 6, 7, 8, 9)
+colnames(masterDF) <- c("tot", "perc", "tag", "DOY", "sp", "year", "dbh", "max_rate_DOY", "max_rate")
+
 # Added by bert: track LG5 parameters for each tag/year in this list
 tag_years <- all_stems %>%
   transmute(tag_year = str_c(tag, year, sep = "-")) %>%
@@ -577,19 +578,27 @@ tag_years <- all_stems %>%
 
 LG5_parameters <- vector(mode = "list", length = length(tag_years))
 names(LG5_parameters) <- tag_years
+growth <- NULL
+all_stems$tag_stem <- paste0(all_stems$tag, "_", all_stems$stemtag)
 
 for (q in 2011:2019) {
   skip_to_next <- FALSE
   Stem2 <- subset(all_stems, year == q)
   for (w in unique(Stem2$sp)) { # removes trees with less than 10 measurements in each year
     Stem3 <- subset(Stem2, sp == w)
-    count_df <- count(Stem3, tag)
-    count_df <- subset(count_df, n >= 10)
-    Stem3 <- Stem3[Stem3$tag %in% count_df$tag, ]
[…]

### 2020-10-15 doc/manuscript/Dow_manuscript.Rmd
edits

@@
 
-#relavant? reworded first line above#
-(Understanding how growth is influenced by multiple factors at once is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation.
-CO2 concentrations in the atmosphere are increasing, and in a perfect scenario, trees and other plants would simply speed up growth in response to this, increasing photosynthetic efficiency in a process called CO2 fertilization (Strain and Cure, 1986; Huang et al., 2007).
-However, factors other than CO2 also limit tree growth, and seasonal temperature averages and precipitation patterns are changing concurrently.
-It is anticipated that, throughout much of the United States, future heat and drought may limit tree growth more strongly than it can be enhanced by elevated CO2 (Charney et al, 2016).)
-#
-
-In recent decades, an increase in average Spring temperatures has been attributed to climate change.
-This relatively small increase has already been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).
-The effects of this Spring warming on leaf phenology have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018) leading to many interesting observations.
-As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by earlier leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
+In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).
+The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many interesting observations.
+As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
 The timing of Spring warming also has a significant effect on leaf phenology response.
@@
 
+Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
 However, the fine-scale responses of temperate deciduous species to a warming Spring are not commonly researched due to the increased effort required.
-Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
-Many observations of climate induced phenological shifts are based on GIS, eddy-covarience measurements, or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
+Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
 To date, very little research, has been done on the phenology of stem-growth.
@@
 Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ).
-For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered contributing to a negative climate feedback loop (source?).
+For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered, contributing to a negative climate feedback loop (source?).
 Recent results from a paper by D’Orangeville et al.
 (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology.
-They showed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time.
-This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses (source for RP vs DP?).
+They observed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time.
+This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses.
+
+^More info on wood-type? move some stuff up from discussion?
 
 While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations.
+However, traditional methods of direct observation are labor intensive, nad very time consuming (Source about cell sampling).
 A solution to this is the use of dendrometer bands.
@@
 The climate is humid temperate, with a mean annual precipitation of 998 mm, January temperatures averaging 1 degrees C and July temperatures averaging 24 degrees C (Helcoski et al, 2019).
+
 The second study site was Harvard Forest in the central Massachusetts town of Petersham (42°53′ N, 72°17′ W, elevation?).
@@
 
-Within the plot, metal dendrometer bands were installed on *>#* trees and measured approximately every 1-2 weeks within the growing season for *9* years (*list ranges, for each site if different*).
+Within the plot, metal dendrometer bands were installed on *>#* trees and measured with a digital caliper approximately every 1-2 weeks within the growing season for *9* years (*list ranges, for each site if different*).
 *(don't bother to mention biannual bands, unless you end up using the data--which you could, for annual growth, if you'd like to increase the sample size.
@@
 Across years, the number of bands sampled averaged *~150* at SCBI, # at Harvard Forest*, and # at SERC*.
+Four main species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.
 *(summary of how bands were distributed across species, focusing on ring/ diffuse porous distinction.
@@
 
-(After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees *--put this in Table S1*).
+(Bands After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees *--put this in Table S1*).
 
+(Bands used after minimum growth checks: 
+1  2011    84
+2  2012    84
+3  2013   100
+4  2014   100
+5  2015    96
+6  2016    94
+7  2017    89
+8  2018    88
+9  2019    75) this seems like alot are being thrown out - manually check if WAI
 Measurements were timed to begin before beginning of spring growth and to continue through the cessation of growth in the fall.
 *(Give specifics, which may vary by site.
-Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)*
+Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)* The median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, and the median end date was 10/17, coinciding with late senesence in the plot.
+The earliest start date for intraannual measurements was in 2020 on March 30, and the latest end date was on 2012 on November 26.
 
@@
 ### Discussion
--Growth timing
+
 We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures.
@@
 On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al).
-These differences are likely evolutionary hydraulic adaptations to climate().
-The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990).
-The pre-leaf stem growth occurring in these species is fueled by carbon stores, often used for other cell maintenance activities within the plant ().
-The dependence on these carbon stores for cambial activation could prove to be a detriment to the tree, as exhaustion of reserves leads to carbon starvation ().
-Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer.
+These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011).
+The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
+The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014).
+The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014).
+Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994).
 This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
 
-The mechanisms controlling activation of cambium reactivation is still a mystery, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves().
+The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005).
 However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.
 
-Ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while 75% remains statistically unchanged, in response to warmer temperatures during early Spring.
+Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth remains statistically unchanged.
 This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
-The reactivation of cambial cells is likely beginning earlier potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
-Damage of this nature would then need to be repaired using carbon reserves which may be in short demand as they are already being used for initial growth.
-Hypothetically, this could lead to carbon starvation in some trees.
+In fact, we found a positive correlation between growing season length in ring-porous trees and Spring temperature, while no such relationship was observed in diffuse-porous trees.
+The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
+Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens.
 Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.
-In the long term, this could lead to changes in species dominance within temperate forests (talk more aobut this?).
 
-Meanwhile, diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps.
-This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase.
-However, diffuse-porous trees are adapted to survive in drought conditions, and may continue to be successful even in future climate scenarios (RIght?
-seeing conflicting reports.
-Sources.).
-Studies have found that these diffuse-porous trees in the Eastern USA show the largest reduction in ring-width increment during, and following, Summer droughts (Kannenberg et al, 2018) as well as being generally more sensitive to Summer precipitation patterns (Elliot et al, 2015).
-The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition.
+Diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps.
+This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas.
+---- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015).
+The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.
+
 --the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades?
@@
 
-This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is the metaphorical tip of the ice berg.
+This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg.
 Here, we focused our analysis purely on early Spring warming.
@@
 For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?).
-This revelation agrees with the results from Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year using tree cores.
-Additionally, Other studies have found that Autumn leaf senescence is being delayed potentially in response to warmer temperatures in the previous year's Winter or Autumn (), changes in chilling requirements due to current year Autumn and winter temperatures (), and stunted total productivity induced by drought and high TMAX in the current year's Summer (), diffuse porous trees increased growth during the following growing season when winter temperatures (TMIN) were lower (Benito & Pederson, 2015; Elliot et al, 2015).
-How these relationships may impact intraannual growth patterns or variables like those described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
+This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year.
+Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high TMAX in the current year's Summer (), and an increase in diffuse-porous tree growth during the gorwing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
+How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
 
@@
 American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts.
-Thus, climate change is having less impact (Zohner et al., 2020)
+Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?
 

### 2020-10-16 doc/manuscript/Dow_manuscript.Rmd
intro work

@@
     fig_caption: yes
+  pdf_document: default
 bibliography: references.bib
-csl: global-change-biology.csl #(this will change depending on journal we choose. GCB would be a great choice)
+csl: global-change-biology.csl
 ---
@@
 In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).
-The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many interesting observations.
+The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations.
 As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
-The timing of Spring warming also has a significant effect on leaf phenology response.
+The timing of this Spring warming also has a significant effect on leaf phenology response.
 Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
 Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
+Therefore, timing of seasonal warming could influence when bud burst occurs, and consequently, lead to variations in total CO2 sequestered by deciduous forests.
 
 Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
-However, the fine-scale responses of temperate deciduous species to a warming Spring are not commonly researched due to the increased effort required.
+However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required ().
 Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
-To date, very little research, has been done on the phenology of stem-growth.
+Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology traits are dissimilar among trees of different wood types (ring-porous and diffuse-porous).
+For example, they observed that bud burst was occuring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
 
+The significance of wood type in regards to wood and leaf phenology likely implies evolutionary differences and adaptations to local climates where these trees can be found.
+Ring-porous trees have large xylem cells which are prone to embolism during the winter.
+These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst.
+Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process.
+Results from the paper by D’Orangeville et al.
+(in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during Spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time.
+--Consequently, the Summer focused diffuse-porous trees are exposed to higher temperatures and drought risk than ring-porous trees.-- If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.
+
+Earlier leaf up timings could signal that these trees are shifting their peak growth windows back in response to increasing Spring temperatures, but to date, very little research, has been done on the phenology of stem-growth.
 In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns.
-Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ).
-For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered, contributing to a negative climate feedback loop (source?).
-Recent results from a paper by D’Orangeville et al.
-(in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology.
-They observed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time.
-This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses.
-
-^More info on wood-type? move some stuff up from discussion?
+Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity in temperate forests (Foster, 2014; ).
+If a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop (source?).
 
 While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations.
-However, traditional methods of direct observation are labor intensive, nad very time consuming (Source about cell sampling).
+However, traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling).
 A solution to this is the use of dendrometer bands.
@@
 This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions.
-Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
+Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within a forest dynamics plot in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
 Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
 
-We hypothesized that (1) stem-growth phenology variables would occur earlier in warmer years as compared to cooler years and (2) both wood types would increase total growth and max growth rate in response to warmer Spring temperatures.
+We hypothesized that (1) stem-growth milestones would be reached earlier in warmer years as compared to cooler years and (2) both wood types would increase total growth and max growth rate in response to warmer Spring temperatures.
 
@@
 However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.
+*READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early Spring temperature signals that the mechanism is influenced by temperature?
 
-Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth remains statistically unchanged.
+Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!!
+Changed by constant amount !!!
+.
 This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
-In fact, we found a positive correlation between growing season length in ring-porous trees and Spring temperature, while no such relationship was observed in diffuse-porous trees.
+In fact, we found a positive correlation between ring-porous growing season length and Spring temperature, while no such relationship was observed in diffuse-porous trees.
 The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
@@
 Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?
+- - RP trees BB = 25% and 75% leaf development = 50% growth (D'Orangeville).
+Are DP trees leafing up earlier?
+If yes, how much earlier?
+Similar to 25,50,75% DOY retreat?
 

### 2020-10-16 doc/manuscript/Dow_manuscript_KAT.Rmd
structuring figures in the manuscript, part 1
-renamed latest version of each figure with a content name
- added box near beginning of .Rmd to assign figure numbers and captions
- placed figures in appropriate locations in manuscript
- automated numbering of figures in citations throughout text

--- 2020-10-15 doc/manuscript/Dow_manuscript.Rmd
@@
 #[insert guidance on word count, n figures/ tables allowed, and then some rough planning of what goes in this manuscript]
+
+```
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+#Figures & Tables: listing figures and their number here is helpful for planning, and also allows automatic numbering within text -->easy changing of order.
+
+#Table numbers
+
+#Figure numbers
+schematic=1 #Based on D'Orangeville Figure 1
+growth_pheno_by_xylem = 2
+rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
+cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
+# 3 climwin
+
+pheno_Tsensitivity = 4
+
+#captions
+
+schematic_caption <- "Figure 1.
+Schematic illustrating the parameters considered here.
+Shown are measurements for an example tree (## cm SPECIES), fit with the model of McMahon & Parker (2015), from which phenology and growth rate parameters are obtained.
+...
+(HERE, WE NEED TO ANNOTATE WITH LABELS FOR EACH PARAMETER CONSIDERED IN THE MANUSCRIPT.
+I'D ALSO DROP THE BACKGROUND SHADING--MAKES IT BUSY.)"
+
+rel_growth_caption <- "Figure 2a. (SHOULD BE MERGED INTO SINGLE IMAGE WITH 2B. SEE BELOW FOR CAPTION)"
+
+cum_growth_caption  <- "Figure 2.
+Growth trajectories for ring- and diffuse-porous trees, as both relative (a) and cumulative (b) fractions of total annual growth.
+Each line represents one year's growth for a given tree, fit with McMahon model."
+
+pheno_Tsensitivity_caption <- "Figure 4. Relationship between DOY and climwin mean tempeatures. For each 
+observed climwin mean temperature value, the posterior mean of the fitted day of year
+is represented by the solid blue line and credible intervals are represented by bands."
 
@@
 
-Here, I’ll insert a reference to Sean’s paper [@mcmahon_general_2015]. This is pulled from references.bib.
+We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig. `r schematic`).
+
+```{r schematic, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/schematic.png")
+```
+
+Notes for now:
+
+1. Model was fit to a random sample of 20 out of 106 tags to speed up computation.
+1. Climwin mean temperatures are recentered at 65F to help sanity check intercepts and slopes. 65F was chosen because it is a "typical" climwinmean value
+1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
+    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
+    1. Random effects: tag number (no species just yet)
 
@@
 
-•	variable averages
+*Phenological patterns*
 
-Pooled across all species and all years the average DOY where max growth rate was achieved is DOY: 158, the DOY where 25% was achieved is DOY: 129, 50% - DOY: 157, 75% - DOY:184.
+Across all trees, Pooled across all species and all years the average DOY where max growth rate was achieved is DOY: 158, the DOY where 25% was achieved is DOY: 129, 50% - DOY: 157, 75% - DOY:184.
+
+Growth pheonology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig.
+`r growth_pheno_by_xylem`).
+Specifically,
 
@@
 
+\newpage
+
+```{r rel_growth, out.width="100%", echo=FALSE, fig.align="center", fig.cap= rel_growth_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/rel_growth.png")
+```
+
+```{r cum_growth, out.width="100%", echo=FALSE, fig.align="center", fig.cap= cum_growth_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/cum_growth.png")
+```
+
+*Climate windows*
+
 •	Climwin results
 
-•	Accounting for random effects of species and tag, pre-season temperatures had a greater effect on ring-porous species than diffuse-porous.
+*Climate sensitivity of phenology*
+
+Warmer spring temperatures resulted in an earlier start of growth for ring-porous species and an earlier growing season for diffuse-porous species (Fig. `r pheno_Tsensitivity`).
+
+```{r pheno_Tsensitivity, out.width="100%", echo=FALSE, fig.align="center", fig.cap=pheno_Tsensitivity_caption}
+# Created in RScripts/bert/wood_phenology_analysis_bert.R
+knitr::include_graphics("tables_figures/pheno_Tsensitivity.png")
+```
 
@@
 
-### Figures
-
-**Figure 1**
-
-Based on D'Orangeville Figure 1
-
-```{r fig1, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Logistic growth model"}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/fig1.png")
-```
-
-**Figure 2**
-
-Based on D'Orangeville Figure 3.a), but with relative growth
-
-```{r fig2, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Percent total (modeled) growth"}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/fig2.png")
-```
-
-**Figure 3**
-
-Similar to D'Orangeville Figure 4.a), but with cummalative relative growth
-
-```{r fig3, out.width="100%", echo=FALSE, fig.align="center", fig.cap= "Cummulative percent total (modeled) growth"}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/fig3.png")
-```
-
-**Figure 4**
-
-**Figure 5**
-
-**Figure 6**
-
-Notes for now:
-
-1. Model was fit to a random sample of 20 out of 106 tags to speed up computation.
-1. Climwin mean temperatures are recentered at 65F to help sanity check intercepts and slopes. 65F was chosen because it is a "typical" climwinmean value
-1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
-    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
-    1. Random effects: tag number (no species just yet)
-
-```{r, echo = FALSE}
-fig6_caption <- "Relationship between DOY and climwin mean tempeatures. For each 
-observed climwin mean temperature value, the posterior mean of the fitted day of year
-is represented by the solid blue line and credible intervals are represented by bands."
-```
-
-```{r fig6, out.width="100%", echo=FALSE, fig.align="center", fig.cap=fig6_caption}
-# Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/fig6.png")
-```
-
 ### References

### 2020-10-16 doc/manuscript/Dow_manuscript.Rmd
Includes edits from KAT

@@
     fig_caption: yes
-  pdf_document: default
 bibliography: references.bib
@@
 
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+#Figures & Tables: listing figures and their number here is helpful for planning, and also allows automatic numbering within text -->easy changing of order.
+
+#Table numbers
+
+#Figure numbers
+schematic=1 #Based on D'Orangeville Figure 1
+growth_pheno_by_xylem = 2
+rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
+cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
+# 3 climwin
+
+pheno_Tsensitivity = 4
+
+#captions
+
+schematic_caption <- "Figure 1.
+Schematic illustrating the parameters considered here.
+Shown are measurements for an example tree (## cm SPECIES), fit with the model of McMahon & Parker (2015), from which phenology and growth rate parameters are obtained.
+...
+(HERE, WE NEED TO ANNOTATE WITH LABELS FOR EACH PARAMETER CONSIDERED IN THE MANUSCRIPT.
+I'D ALSO DROP THE BACKGROUND SHADING--MAKES IT BUSY.)"
+
+rel_growth_caption <- "Figure 2a. (SHOULD BE MERGED INTO SINGLE IMAGE WITH 2B. SEE BELOW FOR CAPTION)"
+
+cum_growth_caption  <- "Figure 2.
+Growth trajectories for ring- and diffuse-porous trees, as both relative (a) and cumulative (b) fractions of total annual growth.
+Each line represents one year's growth for a given tree, fit with McMahon model."
+
+pheno_Tsensitivity_caption <- "Figure 4. Relationship between DOY and climwin mean tempeatures. For each 
+observed climwin mean temperature value, the posterior mean of the fitted day of year
+is represented by the solid blue line and credible intervals are represented by bands."
+
+```
+
 \newpage
@@
 
-The significance of wood type in regards to wood and leaf phenology likely implies evolutionary differences and adaptations to local climates where these trees can be found.
+The significance of wood type in regards to wood and leaf phenology likely stems from evolutionary differences and adaptations to local climates where these trees can be found.
 Ring-porous trees have large xylem cells which are prone to embolism during the winter.
@@
 
-Here, I’ll insert a reference to Sean’s paper [@mcmahon_general_2015]. This is pulled from references.bib.
+We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig. `r schematic`).
+
+```{r schematic, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/schematic.png")
+```
+
+Notes for now:
+
+1. Model was fit to a random sample of 20 out of 106 tags to speed up computation. *(This isn't the case anymore, I fit it to all tag years)*
+1.
+Climwin mean temperatures are recentered at 65F to help sanity check intercepts and slopes.
+65F was chosen because it is a "typical" climwinmean value *(Changed to 16 C based on updated climwin windows)*
+1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
+    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
+    1. Random effects: tag number (no species just yet)
 
@@
 
+*Phenological patterns*
+
+Across all trees, Pooled across all species and all years the average DOY where max growth rate was achieved is DOY: 158, the DOY where 25% was achieved is DOY: 129, 50% - DOY: 157, 75% - DOY:184.
+
+Growth pheonology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig.
+`r growth_pheno_by_xylem`).
+Specifically,
+
 Ring-porous (1395 obs) max growth rate – DOY: 147, max growth rate = .048 (find units in Sean’s function), 25% - DOY: 113, 50% DOY: 147, 75% - DOY: 181…68 days

### 2020-10-16 doc/manuscript/Dow_manuscript.Rmd
added climwin figure (sketch), worked on manuscript

@@
 schematic=1 #Based on D'Orangeville Figure 1
-growth_pheno_by_xylem = 2
+growth_pheno_by_xylem = 2  #if we include more sites, we'll make this multi-panel
 rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
 cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
-# 3 climwin
+climwin = 3
+pheno_Tsensitivity = 4
 
-pheno_Tsensitivity = 4
+#SI:
+#tables
+
+#figures
+climwin_rp_DOY25_SCBI = "S1"
+climwin_dp_DOY25_SCBI = "S2"
 
@@
 
+climwin_caption <- "Figure 3.
+Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees.
+Shown is climwin output indicating the change in AIC relative to a null model (first panel) and linear coefficients (second panel).
+Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: DATE; diffuse-porous: DATE).
+Black boxes indicate the optimal window (i.e., that with lowest dAIC)."
+
 pheno_Tsensitivity_caption <- "Figure 4. Relationship between DOY and climwin mean tempeatures. For each 
@@
 
+climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
+climwin_dp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for diffuse porous species at SCBI."
 ```
@@
 
-•	Sean’s model
-
-•	Climwin
-
-The period where the phenology milestone(s) were most affected by changes in early Spring temperature was determined using the R package Climwin.
-This package tests the correlation between climate variables (TMAX) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'.
-as the best-model.
-We instructed Climwin to search for the best window beginning at February 1st and ending at the mean DOY for each phenology variable (chart in SI?).
-This was done to find corresponding (median of 95% CI window open and close) start and end dates for the analysis of ring-porous and diffuse-porous trees, as well as all trees pooled together.
-
-*Analysis*
+*Fitting growth curves*
 
@@
 ```
+
+*Determining optimal climate windows*
+
+The period where the phenology milestone(s) were most affected by changes in early Spring temperature was determined using the R package Climwin.
+This package tests the correlation between climate variables (TMAX) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'.
+as the best-model.
+We instructed Climwin to search for the best window beginning at February 1st and ending at the mean DOY for each phenology variable (chart in SI?).
+This was done to find corresponding (median of 95% CI window open and close) start and end dates for the analysis of ring-porous and diffuse-porous trees, as well as all trees pooled together.
 
@@
 
-•	variable averages
-
-Pooled across all species and all years the average DOY where max growth rate was achieved is DOY: 158, the DOY where 25% was achieved is DOY: 129, 50% - DOY: 157, 75% - DOY:184.
-
 *Phenological patterns*
 
-Across all trees, Pooled across all species and all years the average DOY where max growth rate was achieved is DOY: 158, the DOY where 25% was achieved is DOY: 129, 50% - DOY: 157, 75% - DOY:184.
+*(Numbers in bold will change. If we bring in HF and/or SERC, we should create a table, rather than trying to list everything in the text.)*
+
+Across all trees and years, 25% of annual growth was achieved on **May 9** ($DOY_{25}$=**129**), 50% on **June 6** ($DOY_{50}$=**157**), and 75% on **July 3** ($DOY_{75}$=**184**), for an average peak growing season length of **56** days.
+Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 7** ($DOY_{max}$=**158**).
 
@@
 `r growth_pheno_by_xylem`).
-Specifically,
-
-Ring-porous (1395 obs) max growth rate – DOY: 147, max growth rate = .048 (find units in Sean’s function), 25% - DOY: 113, 50% DOY: 147, 75% - DOY: 181…68 days
-
-Diffuse-porous (993 obs) max growth rate – DOY: 172, max growth rate = .068 (find units in Sean’s function), 25% - DOY: 153, 50% - DOY: 171, 75% - DOY: 190… 37 days
+For ring-porous trees, on average, 25% of annual growth was achieved on **April 23** ($DOY_{25}$=**113**), 50% on **May 27** ($DOY_{50}$=**147**), and 75% on **June 30** ($DOY_{75}$=**181**), for an average peak growing season length of **68** days.
+Maximum growth rate occurred *on $DOY_{50}$* and averaged .048 **mm day^-1^ (?)**.
+For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 2** ($DOY_{25}$=**153**), 50% on **June 20** ($DOY_{50}$=**171**), and 75% on **July 9** ($DOY_{75}$=**190**), for an average peak growing season length of **37** days.
+Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 21** ($DOY_{max}$=**172**) and averaged .068 **mm day^-1^ (?)**.
 
@@
 
-•	Climwin results
+•	Climwin results (Figs. `r climwin`, `r climwin_rp_DOY25_SCBI`, `r climwin_dp_DOY25_SCBI`)
+
+```{r climwin, out.width="100%", echo=FALSE, fig.align="center", fig.cap= cum_growth_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/climwin.png")
+```
 
@@
 
-### Authors' contributions
+### Author Contributions
+
+### Supporting Information
+
+`r climwin_rp_DOY25_SCBI_short_caption`
+
+`r climwin_dp_DOY25_SCBI_short_caption`
 

### 2020-10-16 doc/manuscript/Dow_manuscript.Rmd
worked on results

@@
 
-*Climate windows*
+*Optimal climate windows*
 
-•	Climwin results (Figs. `r climwin`, `r climwin_rp_DOY25_SCBI`, `r climwin_dp_DOY25_SCBI`)
+For the majority of weekly time windows from Feb.
+1 to $DOY_{25}$, there was negative relationship between temperature and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
+`r climwin`, `r climwin_rp_DOY25_SCBI`, `r climwin_dp_DOY25_SCBI`).
+This held true for both ring- and diffuse-porous trees, *albeit* with stronger relationships for ring-porous trees. 
+*(Describe the landscape:)*
+In general, there were significant negative relationships between temperature and $DOY_{25}$ for any time period beginning $\ge$ **5** weeks prior to $DOY_{25}$, ending  $\le$ **6** weeks prior, and covering at least a **3**-week period.
+Relationships tended to be weaker and more stochastic in sign for short time periods, particularly those just prior to $DOY_{25}$.
+
+Optimal windows occurred from **##-##** for ring-porous species and from **##-##** for diffuse-porous species.
 

### 2020-10-17 doc/manuscript/Dow_manuscript.Rmd
small changes

@@
 pheno_Tsensitivity = 4
+model_summary = 5
 
@@
 is represented by the solid blue line and credible intervals are represented by bands."
+
+model_summary_caption <- "Figure 5. (PLACEHOLDER FOR FIGURE WE DISCUSSED AT LAB MEETING)"
 

### 2020-10-26 doc/LG5_errata/LG5_errata.Rmd
Adding LG5 model errata docs

---
title: "LG5 Function Errata"
author: "Albert Y. Kim"
date: "Last updated on `r Sys.Date()`"
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: kable
    highlight: tango
    number_sections: yes
    theme: cosmo
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, echo = FALSE)
library(tidyverse)
```

# LG5 model formula typo

The formula is written as follows in [McMahon and Parker 2015 Eq 1 (page 245)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4314258/). Note how in the numerator the $L$'s cancel out, leaving only $K$ being included in the formula:

$$
dbh = \frac{L + (K-L)}{1 + 1/\theta*\exp(-r(doy - doy_{ip})/\theta)^{\theta}}
$$
The formula is written as follows in the D'Orangeville et al. draft manuscript (page 9). Note the numerator is different than McMahon and Parker's formulation:

$$
dbh = \frac{K-L}{1 + 1/\theta*\exp(-r(doy - doy_{ip})/\theta)^{\theta}}
$$
However the `lg5.pred()` function from the [`RDendrom`
package](https://github.com/seanmcm/RDendrom/blob/master/R/DENDRO_BASE_FUNCTIONS.R#L544) computes DBH as follows:

```{r, echo = TRUE}
lg5.pred <- function(params, doy) {
  L <- params[1] # min(dbh, na.rm = T)
  K <- params[2]
  doy.ip <- params[3]
  r <- params[4]
  theta <- params[5]
  dbh <- vector(length = length(doy))
  dbh <- L + ((K - L) / (1 + 1 / theta * exp(-(r * (doy - doy.ip) / theta))^theta))
  return(dbh)
}
```

which corresponds to the following formula:

$$
dbh = L + \frac{K-L}{1 + 1/\theta*\exp(-r(doy - doy_{ip})/\theta)^{\theta}}
$$

Also note that the denominator of the LG5 function can be simplified since 
$$
\exp(-r(doy - doy_{ip})/\theta)^{\theta} = \exp(-r(doy - doy_{ip}))
$$

# $r$ parameter from LG5 model

I have a suspicion both these interpretations of the $r$ parameter of LG5 model are incorrect:

1. McMahon and Parker 2015 state just above Eq 1 (page 245): "the rate parameter $r$ describes the slope of the curve at the inflection point"
1. D'Orangeville et al. state (page 9): "$r$ represents the maximum growth rate" and "$doy_{ip}$ is the day of year when maximum growth rate occurs"

Rather, I believe that the correct formulation of the slope/the maximum growth rate is instead:

$$
r_{new} = \frac{K-L}{\left(1 + \frac{1}{\theta}\right)^2}\frac{r}{\theta}
$$

## Plot LG5 diameter and growth curves

Let's set arbitrarily chosen parameter values then

1. Use `lg5.pred()` to compute all diameters for all DOY = 1, ..., 365
1. Compute growth by taking differences of diameters

```{r, echo = TRUE}
L <- 13
K <- 14
doy.ip <- 175
r <- 0.05
theta <- 1
```

We then plot both the diameter curve as well as the growth curve, with `doy.ip` marked with a vertical dotted line. Observe how growth peaks at `doy.ip`.

```{r, fig.width=10}
params <- c(L, K, doy.ip, r, theta)
dbh_growth_values <- tibble(doy = seq(from = 1, to = 365)) %>%
  mutate(
    diameter = lg5.pred(params, doy),
    growth = diameter - lag(diameter)
  )

par(mfrow=c(1, 2))
plot(dbh_growth_values$doy, dbh_growth_values$diameter, type = "l", 
     main = "LG5 diameter", xlab = "doy", ylab = "diameter")
abline(v = doy.ip, lty = "dotted")
plot(dbh_growth_values$doy, dbh_growth_values$growth, type = "l", 
     main = "Growth curve", xlab = "doy", ylab = "growth")
abline(v = doy.ip, lty = "dotted")
```

## Original interpretations of $r$

1. McMahon and Parker 2015 state: "the rate parameter $r$ describes the slope of the curve at the inflection point". We mark the appropriate line with slope `r` in orange in the diameter plot. 
1. D'Orangeville et al. state: "$r$ represents the maximum growth rate" and "$doy_{ip}$ is the day of year when maximum growth rate occurs". Let's mark this maximum growth rate with a horizontal orange line in the growth plot.

```{r, fig.width=10}
# Compute diameter at doy.ip, then solve for intercept via y = mx + b 
y <- lg5.pred(params, doy.ip)
intercept <- y - r*doy.ip

par(mfrow=c(1, 2))
plot(dbh_growth_values$doy, dbh_growth_values$diameter, type = "l", 
     main = "LG5 diameter", xlab = "doy", ylab = "diameter")
abline(v = doy.ip, lty = "dotted")
abline(intercept, r, col = "orange")
plot(dbh_growth_values$doy, dbh_growth_values$growth, type = "l", 
     main = "Growth curve", xlab = "doy", ylab = "growth")
abline(v = doy.ip, lty = "dotted")
abline(h = r, col = "orange")
```

Observe

1. In the left plot how the orange line does not appear to be tangent to the LG5 curve at `doy.ip` and hence is not the slope at that point.
1. In the right plot $r$ = `r r` does not correspond to the growth rate at `doy.ip` which should be the max growth rate. Rather it appears to be around 0.012

## Proposed new $r$

Let's repeat the above with our proposed growth rate $r_{new}$, which is $\frac{d}{dx}LG5(x)$ evaluated at $x$ = `doy.ip`. See this [PDF](https://drive.google.com/file/d/1CMT2o1svTfbS7wQ_uJoWjqHbvDopkCZw/view?usp=sharing) for the analytic derivation of the derivative.

$$
r_{new} = \frac{K-L}{\left(1 + \frac{1}{\theta}\right)^2}\frac{r}{\theta}
$$

```{r, fig.width=10}
# Comupte new r, then find intercept
r_new <- ((K-L)/(1 + 1/theta)^2) * (r/theta)
intercept_new <- y - r_new*doy.ip

par(mfrow=c(1, 2))
plot(dbh_growth_values$doy, dbh_growth_values$diameter, type = "l", 
     main = "LG5 diameter", xlab = "doy", ylab = "diameter")
abline(v = doy.ip, lty = "dotted")
abline(intercept_new, r_new, col = "orange")
plot(dbh_growth_values$doy, dbh_growth_values$growth, type = "l", 
     main = "Growth curve", xlab = "doy", ylab = "growth")
abline(v = doy.ip, lty = "dotted")
abline(h = r_new, col = "orange")
```

Observe:

1. In the left plot how the orange line is now tangent to the LG5 curve at `doy.ip` and thus is the correct slope.
1. In the right plot $r_{new}$ = `r r_new` now corresponds to the max growth rate occurring at `doy.ip`.

### 2020-10-26 doc/LG5_errata/LG5_errata.Rmd
fixed broken link

@@
 However the `lg5.pred()` function from the [`RDendrom`
-package](https://github.com/seanmcm/RDendrom/blob/master/R/DENDRO_BASE_FUNCTIONS.R#L544) computes DBH as follows:
+package](https://github.com/seanmcm/RDendrom/blob/master/R/DENDRO_BASE_FUNCTIONS.R) computes DBH as follows:
 

### 2020-10-26 doc/manuscript/Dow_manuscript.Rmd
Small changes to intro. Add DOYtiming figure

@@
 model_summary = 5
-
+DOYtiming = 6
 #SI:
@@
 climwin_dp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for diffuse porous species at SCBI."
+DOYtiming_caption <- "Figure 6. Day of year where growth milestones were achied at the Smithonian Conservation Biology Institute."
 ```
@@
 As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
-The timing of this Spring warming also has a significant effect on leaf phenology response.
-Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
-Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
-Therefore, timing of seasonal warming could influence when bud burst occurs, and consequently, lead to variations in total CO2 sequestered by deciduous forests.
+The timing of this Spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).
 
-Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
+Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
 However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required ().
 Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
-Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology traits are dissimilar among trees of different wood types (ring-porous and diffuse-porous).
-For example, they observed that bud burst was occuring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
+Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous).
+For example, they observed that bud burst was occurring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
 
@@
 
-Earlier leaf up timings could signal that these trees are shifting their peak growth windows back in response to increasing Spring temperatures, but to date, very little research, has been done on the phenology of stem-growth.
+Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing Spring temperatures, but to date, very little research has been done on topics like these.
 In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns.
-Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity in temperate forests (Foster, 2014; ).
-If a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop (source?).
+Warming events in late-Spring were shown to shift leaf emergence timing back more than early-Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
+Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.
 
-While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations.
-However, traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling).
+(May remove...
+not sure where to work in...or if it needs to be in) Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
+
+While the afformentioned broad scale studies have shown a potential relationship between total tree growth and temperature, timing of intraannual stem-growth is extremely important as well.
+The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; ).
+Species have co-evolved to utilize the phenological stages of trees at specific times in the year.
+If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees ().
+Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).
+
+While not entirely impossible, it is difficult to record intraannual growth traits or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations.
+However, some traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling).
 A solution to this is the use of dendrometer bands.
 Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?).
-Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014).
-This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions.
-Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within a forest dynamics plot in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth.
-Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
+Using many measurements taken within a year, we can model the growth of individual trees and extract data on growth milestones and traits for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014).
+This allows the determination of variable responses of key tree species within a forest, meaning a better understanding of future growth under projected climate conditions.
+Here, we use 15 (5 @ Harvard Forest, 10 @ SCBI) years of biweekly dendrometer band measurements for 19 (15 @ Harvard Forest, 4 @ SCBI) species to characterize intraannual stem-growth phenology within two forest dynamics plots in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; the total growth; and season length (75%-25% DOY).
+Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
 
@@
 
+```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/DOYtiming.png")
+```
+
 The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005).

### 2020-10-26 RScripts/additional_figures_HarvardForest.R

@@ -3,7 +3,7 @@ library(scales)
 library(tidybayes)
 
 # Figure numbers from: https://github.com/SCBI-ForestGEO/growth_phenology/issues/13
-
+unique(Wood_pheno_table_HarvardForest_V2$sp)
 # Figure 1: Logistic growth curve and parameter illustration -------------------
 # Set parameters from McMahon & Parker (2014)
 # 1. L: lower asymptotes
@@ -17,7 +17,7 @@ doy.ip <- 175
 r <- 0.05
 theta <- 1
 params <- c(K, L, doy.ip, r, theta)
-
+lg5
 get.lg5.ML <- function(params, doy, dbh, resid.sd) {
   pred.dbh <- lg5.pred(params, doy)
   pred.ML <-  -sum(dnorm(dbh, pred.dbh, resid.sd, log = T))

### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
big-picture structuring of intro

@@
 #Table numbers
+hypothesis_table=1
 
 #Figure numbers
-schematic=1 #Based on D'Orangeville Figure 1
+schematic=1 #illustrates parameters
 growth_pheno_by_xylem = 2  #if we include more sites, we'll make this multi-panel
@@
 
+#### (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
 As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
@@
 For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil.
-Forests sequester about one-third of all anthropogenic CO2 and store ~45% of all terrestrial carbon (Bonan, 2008).
+Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon (Bonan, 2008).
 However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).
@@
 Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019).
-Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO2, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens.
+Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens.
 Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.
 
-In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).
-The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations.
+#### (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
+In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).** The effects of this Spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*.
 As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
@@
 
+#### (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
 Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
@@
 Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
-Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous).
+
+Tree-ring records give some hints that warmer springs may increase annual growth (Helcoski et al.
+2019), but tree-rings tell us nothing about the timing of that growth.
+Timing of intraannual stem-growth is extremely important as well.
+
+Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing Spring temperatures, but to date, very little research has been done on topics like these.
+In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns.
+Warming events in late-Spring were shown to shift leaf emergence timing back more than early-Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
+Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.
+
+#### (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
+A recent publication by D'Orangeville et al () found that leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous).
 For example, they observed that bud burst was occurring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
@@
 (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during Spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time.
---Consequently, the Summer focused diffuse-porous trees are exposed to higher temperatures and drought risk than ring-porous trees.-- If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.
+If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.
 
-Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing Spring temperatures, but to date, very little research has been done on topics like these.
-In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns.
-Warming events in late-Spring were shown to shift leaf emergence timing back more than early-Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
-Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.
-
-(May remove...
-not sure where to work in...or if it needs to be in) Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
-
-While the afformentioned broad scale studies have shown a potential relationship between total tree growth and temperature, timing of intraannual stem-growth is extremely important as well.
-The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; ).
-Species have co-evolved to utilize the phenological stages of trees at specific times in the year.
-If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees ().
-Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).
-
-While not entirely impossible, it is difficult to record intraannual growth traits or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations.
-However, some traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling).
-A solution to this is the use of dendrometer bands.
-Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?).
+#### (5. Here, we...)
 Using many measurements taken within a year, we can model the growth of individual trees and extract data on growth milestones and traits for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014).
 This allows the determination of variable responses of key tree species within a forest, meaning a better understanding of future growth under projected climate conditions.
-Here, we use 15 (5 @ Harvard Forest, 10 @ SCBI) years of biweekly dendrometer band measurements for 19 (15 @ Harvard Forest, 4 @ SCBI) species to characterize intraannual stem-growth phenology within two forest dynamics plots in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; the total growth; and season length (75%-25% DOY).
+Here, we use biweekly dendrometer band measurements (5 years @ Harvard Forest, 10 @ SCBI) for dominant tree species at each site (15 @ Harvard Forest, 4 @ SCBI)  to characterize intraannual stem-growth phenology within two forest dynamics plots in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; the total growth; and season length (75%-25% DOY) (Fig.
+`r schematic`).
 Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
 
-We hypothesized that (1) stem-growth milestones would be reached earlier in warmer years as compared to cooler years and (2) both wood types would increase total growth and max growth rate in response to warmer Spring temperatures.
+We test the hypotheses that (1) stem-growth milestones are reached earlier in years with warmer early springs, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).
+
+```{r schematic, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/schematic.png")
+```
+
+```{r hypothesis_table, eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)  
+hypothesis_table <- read.csv(file = "tables_figures/hypothesis_table.csv")
+knitr::kable(hypothesis_table, caption = "Table 1. Hypothesis table. (NEEDS FORMATTING...obviously!)")
+```
 
@@
 
-```{r schematic, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/schematic.png")
-```
-
 *Determining optimal climate windows*
@@
 
-1. Model was fit to a random sample of 20 out of 106 tags to speed up computation. *(This isn't the case anymore, I fit it to all tag years)*
-1.
-Climwin mean temperatures are recentered at 65F to help sanity check intercepts and slopes.
-65F was chosen because it is a "typical" climwinmean value *(Changed to 16 C based on updated climwin windows)*
 1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
@@
 
-Content to incorporate: N.
+### (Content to incorporate)
+
+N.
 American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts.
@@
 Similar to 25,50,75% DOY retreat?
+
+Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
+
+The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; D’Orangeville et al.
+in review).
+Species have co-evolved to utilize the phenological stages of trees at specific times in the year.
+If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees ().
+Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).
 

### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
structuring/ writing

@@
 DOYtiming = 6
-#SI:
+DOYtiming_SCBI = "6a"
+DOYtiming_Harvard = "6b"
+
+#captions
 #tables
+hypothesis_table_caption <-"Table 1. Hypothesis table. (NEEDS FORMATTING...obviously!)"
 
 #figures
-climwin_rp_DOY25_SCBI = "S1"
-climwin_dp_DOY25_SCBI = "S2"
-
-#captions
-
 schematic_caption <- "Figure 1.
@@
 rel_growth_caption <- "Figure 2a. (SHOULD BE MERGED INTO SINGLE IMAGE WITH 2B. SEE BELOW FOR CAPTION)"
-
 cum_growth_caption  <- "Figure 2.
@@
 climwin_caption <- "Figure 3.
-Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees.
-Shown is climwin output indicating the change in AIC relative to a null model (first panel) and linear coefficients (second panel).
-Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: DATE; diffuse-porous: DATE).
-Black boxes indicate the optimal window (i.e., that with lowest dAIC)."
+Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
+Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: DATE at SCBI, DATE at HF; diffuse-porous: DATE at SCBI, DATE at HF).
+Black boxes indicate the optimal window (i.e., that with lowest dAIC).
+(REPLOT ON THE SAME SCALE, WITH BLUE FOR POSITIVE AND RED FOR NEGATIVE)"
 
@@
 
+DOYtiming_caption <- "Figure 6. Day of year where growth milestones were achied at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest."
+
+# SUPPLEMENTARY INFO
+#SI figure numbers
+climwin_rp_DOY25_SCBI = "S1"
+climwin_dp_DOY25_SCBI = "S2"
+climwin_rp_DOY25_SCBI = "S3"
+climwin_dp_DOY25_SCBI = "S4"
+#SI figure captions
 climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
-climwin_dp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for diffuse porous species at SCBI."
-DOYtiming_caption <- "Figure 6. Day of year where growth milestones were achied at the Smithonian Conservation Biology Institute."
+climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for DOY 25 for diffuse porous species at SCBI."
+climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for DOY 25 for ring porous species at Harvard Forest."
+climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
+
 ```
@@
 #### (5. Here, we...)
-Using many measurements taken within a year, we can model the growth of individual trees and extract data on growth milestones and traits for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014).
-This allows the determination of variable responses of key tree species within a forest, meaning a better understanding of future growth under projected climate conditions.
-Here, we use biweekly dendrometer band measurements (5 years @ Harvard Forest, 10 @ SCBI) for dominant tree species at each site (15 @ Harvard Forest, 4 @ SCBI)  to characterize intraannual stem-growth phenology within two forest dynamics plots in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; the total growth; and season length (75%-25% DOY) (Fig.
-`r schematic`).
-Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
-
-We test the hypotheses that (1) stem-growth milestones are reached earlier in years with warmer early springs, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).
+Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within **two/three** forest dynamics plots in the Eastern USA.
+Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig.
+`r schematic`; @mcmahon_general_2015].
+We test the hypotheses that (1) warmer early springs result in earlier stem growth and a period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).
 
@@
 hypothesis_table <- read.csv(file = "tables_figures/hypothesis_table.csv")
-knitr::kable(hypothesis_table, caption = "Table 1. Hypothesis table. (NEEDS FORMATTING...obviously!)")
+knitr::kable(hypothesis_table, caption = hypothesis_table_caption)
 ```
@@
 
-*Study sites and data*
+#### Study sites and data
 
@@
 
+(5 years @ Harvard Forest, 10 @ SCBI)
+
 (Bands After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees *--put this in Table S1*).
@@
 9  2019    75) this seems like alot are being thrown out - manually check if WAI
+
 Measurements were timed to begin before beginning of spring growth and to continue through the cessation of growth in the fall.
@@
 
-*Fitting growth curves*
+#### Fitting growth curves
 
@@
 
-*Determining optimal climate windows*
+#### Determining optimal climate windows
 
@@
 This was done to find corresponding (median of 95% CI window open and close) start and end dates for the analysis of ring-porous and diffuse-porous trees, as well as all trees pooled together.
+
+#### Hypothesis testing
+
+Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
 
@@
 
-```{r climwin, out.width="100%", echo=FALSE, fig.align="center", fig.cap= cum_growth_caption}
+```{r climwin, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
 # Created in RScripts/additional_figures.R
@@
 knitr::include_graphics("tables_figures/DOYtiming.png")
+```
+
+```{r DOYtiming2, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
+knitr::include_graphics("tables_figures/DOYtiming_HF_PRELIM.png")
 ```
@@
 
+`r climwin_rp_DOY25_HARV_short_caption`
+
+`r climwin_dp_DOY25_HARV_short_caption`
+
 ### References

### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
remove notes on additional analysis ideas
#20

@@
 
-leaf phenology data?
-
-•	leaf phenology data from NEON and/or satellite-based (leaf phenology network: https://www.usanpn.org/news/spring) ?
-
-•	NDVI or PRI? https://onlinelibrary.wiley.com/doi/10.1111/gcb.15112 (from Ian)
-
-•	From Ian: I noticed how the growth patterns observed by remote sensing pretty much mirror what Sean was finding with his dendro R package (btw is that functional on CRAN yet?) for both SERC and SCBI dendroband data.
-
-perhaps bring in cores?
-
 #### Fitting growth curves

### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
draft growth_curves_all figure
#21

@@
 schematic=1 #illustrates parameters
-growth_pheno_by_xylem = 2  #if we include more sites, we'll make this multi-panel
-rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
-cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
+growth_curves_all = 2  
+    rel_growth_SCBI = "2a" 
+    rel_growth_HF = "2b" 
+    cum_growth_SCBI = "2c" 
+    cum_growth_HF = "2d" 
 climwin = 3
@@
 
-rel_growth_caption <- "Figure 2a. (SHOULD BE MERGED INTO SINGLE IMAGE WITH 2B. SEE BELOW FOR CAPTION)"
-cum_growth_caption  <- "Figure 2.
-Growth trajectories for ring- and diffuse-porous trees, as both relative (a) and cumulative (b) fractions of total annual growth.
+growth_curves_all_caption  <- "Figure 2.
+Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
 Each line represents one year's growth for a given tree, fit with McMahon model."
@@
 Growth pheonology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig.
-`r growth_pheno_by_xylem`).
+`r growth_curves_all`).
 For ring-porous trees, on average, 25% of annual growth was achieved on **April 23** ($DOY_{25}$=**113**), 50% on **May 27** ($DOY_{50}$=**147**), and 75% on **June 30** ($DOY_{75}$=**181**), for an average peak growing season length of **68** days.
@@
 
-```{r rel_growth, out.width="100%", echo=FALSE, fig.align="center", fig.cap= rel_growth_caption}
+```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption}
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/rel_growth.png")
-```
-
-```{r cum_growth, out.width="100%", echo=FALSE, fig.align="center", fig.cap= cum_growth_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/cum_growth.png")
+knitr::include_graphics("tables_figures/growth_curves_all.png")
 ```
@@
 This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year.
-Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high TMAX in the current year's Summer (), and an increase in diffuse-porous tree growth during the gorwing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
+Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high TMAX in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
 How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
@@
 ### (Content to incorporate)
+
+Geng et al. 2020 GCB- Climate warming increases spring phenological differences among temperate trees.
 

### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
remove duplicate figure reference

@@
 pheno_Tsensitivity = 4
-model_summary = 5
-DOYtiming = 6
-DOYtiming_SCBI = "6a"
-DOYtiming_Harvard = "6b"
+DOYtiming = 5
+DOYtiming_SCBI = "5a"
+DOYtiming_Harvard = "5b"
 
@@
 
-model_summary_caption <- "Figure 5. (PLACEHOLDER FOR FIGURE WE DISCUSSED AT LAB MEETING)"
-
-DOYtiming_caption <- "Figure 6. Day of year where growth milestones were achied at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest."
+DOYtiming_caption <- "Figure 5. Day of year where growth milestones were achied at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest."
 

### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
table `sample_size_by_year`
#28

@@
 # SUPPLEMENTARY INFO
+#SI table numbers
+sample_size_by_year= "S1"
+
+#SI table captions
+sample_size_by_year_caption <- "Table S1. Sample size by year"
+
 #SI figure numbers
@@
 *(don't bother to mention biannual bands, unless you end up using the data--which you could, for annual growth, if you'd like to increase the sample size.
-You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census *(Table S1)*.
+You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`).
 Across years, the number of bands sampled averaged *~150* at SCBI, # at Harvard Forest*, and # at SERC*.
@@
 (5 years @ Harvard Forest, 10 @ SCBI)
-
-(Bands After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees *--put this in Table S1*).
-
-(Bands used after minimum growth checks: 
-1  2011    84
-2  2012    84
-3  2013   100
-4  2014   100
-5  2015    96
-6  2016    94
-7  2017    89
-8  2018    88
-9  2019    75) this seems like alot are being thrown out - manually check if WAI
 
@@
 
+`r sample_size_by_year_caption`
+
 `r climwin_rp_DOY25_SCBI_short_caption`

### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
bit of work on intro & cleanup

@@
 
-**Title:** Effect of spring temperatures on tree growth phenology in a temperate deciduous forest
+**Title:** Effect of spring temperatures on tree growth phenology in two temperate deciduous forests
 
@@
 Schematic illustrating the parameters considered here.
-Shown are measurements for an example tree (## cm SPECIES), fit with the model of McMahon & Parker (2015), from which phenology and growth rate parameters are obtained.
-...
-(HERE, WE NEED TO ANNOTATE WITH LABELS FOR EACH PARAMETER CONSIDERED IN THE MANUSCRIPT.
-I'D ALSO DROP THE BACKGROUND SHADING--MAKES IT BUSY.)"
+Shown are measurements for an example tree (## cm SPECIES), fit with the model of McMahon & Parker (2015), from which phenology and growth rate parameters are obtained."
 
@@
 Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: DATE at SCBI, DATE at HF; diffuse-porous: DATE at SCBI, DATE at HF).
-Black boxes indicate the optimal window (i.e., that with lowest dAIC).
-(REPLOT ON THE SAME SCALE, WITH BLUE FOR POSITIVE AND RED FOR NEGATIVE)"
+Black boxes indicate the optimal window."
 
@@
 #### (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
-In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).** The effects of this Spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*.
+In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
+Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
+The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*.
 As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
-The timing of this Spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).
+The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).
+
+Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
 
 #### (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
-Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
-However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required ().
-Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
+
+Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research has been done on this topic. 
+In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns.
+*(but check into this....
+I suspect there may be some that we're missing)*
+Warming events in late-spring were shown to shift leaf emergence timing back more than early-spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
+Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.
 
@@
 Timing of intraannual stem-growth is extremely important as well.
-
-Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing Spring temperatures, but to date, very little research has been done on topics like these.
-In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns.
-Warming events in late-Spring were shown to shift leaf emergence timing back more than early-Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
-Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.
 
@@
 Results from the paper by D’Orangeville et al.
-(in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during Spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time.
-If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.
+(in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time.
+If leaf development is sped up when spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.
 
@@
 
-The period where the phenology milestone(s) were most affected by changes in early Spring temperature was determined using the R package Climwin.
+The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package Climwin.
 This package tests the correlation between climate variables (TMAX) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'.
@@
 
-Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
+Correlation between these variables and spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
 
@@
 
-A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, along with random effects of species and tag.
+A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and spring temperature, along with random effects of species and tag.
 We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison.
@@
 
-We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures.
+We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures.
 The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here.
@@
 These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011).
-The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
+The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
 The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014).
 The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014).
-Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994).
+Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994).
 This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
@@
 However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.
-*READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early Spring temperature signals that the mechanism is influenced by temperature?
+*READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early spring temperature signals that the mechanism is influenced by temperature?
 
@@
 .
-This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
-In fact, we found a positive correlation between ring-porous growing season length and Spring temperature, while no such relationship was observed in diffuse-porous trees.
+This implies that the growing season of ring-porous trees expanded due to spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
+In fact, we found a positive correlation between ring-porous growing season length and spring temperature, while no such relationship was observed in diffuse-porous trees.
 The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
@@
 
-Diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps.
+Diffuse-porous growth timing seems relatively unaffected by early spring temperatures, instead showing a higher correlation with late spring or Summer temps.
 This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas.
@@
 This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg.
-Here, we focused our analysis purely on early Spring warming.
+Here, we focused our analysis purely on early spring warming.
 This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth.
@@
 
-Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
+Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
 

### 2020-10-28 RScripts/HF_wood_phenology.R
misc work

@@ -780,14 +780,14 @@ warnings()
 masterDF <- masterDF[-1, ]
 unique(masterDF$sp)
 masterDF$wood_type <- ifelse(masterDF$sp == "betual"| masterDF$sp == "acerru" | masterDF$sp == "fagugr" | masterDF$sp == "betule" | masterDF$sp == "betupa" | masterDF$sp == "acerpe" | masterDF$sp == "betupo" | masterDF$sp == "prunse", "diffuse-porous", ifelse(masterDF$sp == "querru" | masterDF$sp == "querve" | masterDF$sp == "fraxam", "ring-porous", "other"))
-write.csv(masterDF, file = "Data/Wood_pheno_table_HarvardForest_subset.csv", row.names = FALSE)
+write.csv(masterDF, file = "Data/Wood_pheno_table_HarvardForest_V3.csv", row.names = FALSE)
 masterDF$DOY <- as.numeric(masterDF$DOY)
 # Added by bert: save parameter values
 LG5_parameters %>%
   bind_rows() %>%
   write_csv(file = "Data/LG5_parameter_values_HarvardForest_subset.csv")
 
-write.csv(bind_rows(LG5_parameters), file = "Data/LG5_parameter_values_HarvardForest_subset.csv")
+write.csv(bind_rows(LG5_parameters), file = "Data/LG5_parameter_values_HarvardForest_V3.csv")
 lg5 <- bind_rows(LG5_parameters)
 #Clean data
 Wood_pheno_table <- read_csv("Data/Wood_pheno_table_HarvardForest_V1.csv") #Master datafrmae containing 20%, 50%, and 75% growth milestones

### 2020-10-28 RScripts/SCBI_wood_phenology.R

@@ -399,6 +399,11 @@ fit.outer.hull <- function(dbh, doy.full, params, quant = 0.8) {
   out.fit <- outer.hull(params, doy, dbh)
 }
 
+get.lg5.ML <- function(params, doy, dbh, resid.sd) {
+  pred.dbh <- lg5.pred(params, doy)
+  pred.ML <-  -sum(dnorm(dbh, pred.dbh, resid.sd, log = T))
+  return(pred.ML)
+}
 
 
 # 3. Loop to create timeseries of dbh measurements manually -----
@@ -862,10 +867,18 @@ for (q in 2011:2019) {
 
       names(Param.df) <- c("L", "K", "doy_ip", "r", "theta", "a", "b")
       # write.csv(Param.df, file = "SCBI_hi_lo_lg5.csv", quote = FALSE, row.names = FALSE)
-
+      K <- as.numeric(Param.df[1,2])
+      L <- as.numeric(Param.df[1,1])
+      doy.ip <- as.numeric(Param.df[1,3])
+      r <- as.numeric(Param.df[1,4])
+      theta <- as.numeric(Param.df[1,5])
+      params_function <- c(K, L, doy.ip, r, theta)
+      dbh <- dbh
+      doy <- doy
+      ML_value <- get.lg5.ML(params_function, doy,dbh, .5)
       # Added by bert: save tag/year + parameter values
-      param_values_to_save <- c(stemstag$tag[1], stemstag$year[1],  Param.df) %>% unlist()
-      names(param_values_to_save) <- c("tag", "year", names(Param.df))
+      param_values_to_save <- c(stemstag$tag[1], stemstag$year[1],  Param.df, ML_value) %>% unlist()
+      names(param_values_to_save) <- c("tag", "year", names(Param.df), "ML_value")
       tagyear_id <- stemstag %>%
         slice(1) %>%
         transmute(tag_year = str_c(tag, year, sep = "-")) %>%
@@ -992,15 +1005,15 @@ for (q in 2011:2019) {
 warnings()
 masterDF <- masterDF[-1, ]
 masterDF$wood_type <- ifelse(masterDF$sp == "quru" | masterDF$sp == "qual", "ring porous", ifelse(masterDF$sp == "litu" | masterDF$sp == "fagr", "diffuse-porous", "other"))
-write.csv(masterDF, file = "Data/Wood_pheno_table_V6.csv", row.names = FALSE)
[…]

### 2020-10-28 RScripts/additional_figures.R

@@ -146,8 +146,14 @@ ggsave("doc/manuscript/tables_figures/schematic.png", plot = schematic, width = 
 
 
 
-# Fig2 & Fig3: Percent modeled growth and cummulative percent modeled growth ---
-Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V7.csv") %>%
+# Growth_Cruves_all: Percent modeled growth and cummulative percent modeled growth ---
+Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V7.csv") %>%
+  # Keep only RP and DP for now
+  filter(wood_type != "other") %>%
+  # Rename ring porous to not have a space
+  mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
+
+Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HFtemp.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   # Rename ring porous to not have a space
@@ -155,7 +161,7 @@ Wood_pheno_table <- read_csv("Data/Wood_pheno_table_V7.csv") %>%
 
 
 # As generated in Rscripts/SCBI_wood_phenology.R
-LG5_parameter_values <- read_csv("Data/LG5_parameter_values.csv")
+LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values.csv")
 
 # Generalized 5-parameter logistic function (modified version of Sean's function)
 lg5 <- function(L, K, doy_ip, r, theta, doy) {

### 2020-10-28 RScripts/additional_figures_HarvardForest.R
SI figures

@@ -281,3 +281,77 @@ fig_width <- 7
 ggsave(filename = "doc/manuscript/tables_figures/DOYtiming_HF_PRELIM.png",
        plot = doytiming,
        width = fig_width, height = fig_width / 2)
+
+####Tags per Year table ----
+all_stems$tag_year <- paste(all_stems$tag, all_stems$year, sep = "_")
+unique <- unique(all_stems$tag_year)
+unique <- data.frame(unique)
+unique$year <- substr(unique$unique,nchar(as.character(unique$unique))-3, nchar(as.character(unique$unique))) #use this to split cores up into two columns, ID and letter, then use unique id's to pick one of each core?
+tags_per_year <- count(unique$year)
+names(tags_per_year) <- c("Year", "SCBI")
+
+all_stems$tag_year <- paste(all_stems$plot, all_stems$tag, all_stems$year, sep = "_")
+unique <- unique(all_stems$tag_year)
+unique <- data.frame(unique)
+unique$year <- substr(unique$unique,nchar(as.character(unique$unique))-3, nchar(as.character(unique$unique))) #use this to split cores up into two columns, ID and letter, then use unique id's to pick one of each core?
+tags_per_year_HF <- count(unique$year)
+names(tags_per_year_HF) <- c("Year", "Harvard Forest")
+
+tags <- merge(tags_per_year,tags_per_year_HF, by = "Year", all = TRUE)
+tags$Year <- as.character(tags$Year)
+tags$Year <- as.numeric(tags$Year)
+tags <- tags[order(tags$Year),-4]
+
+write.csv(tags, file = "Tags_per_year.csv", row.names = FALSE)
+
+#Species distibution of bands
+stems_unclean <- read.csv("data/Wood_pheno_table_HarvardForest_V2.csv")
+stems_clean <- read_csv("data/wood_pheno_table_HFtemp.csv")
+stems_clean <- subset(stems_clean, perc == .25)
+stems_unclean <- subset(stems_unclean, perc == .25 & wood_type != "other")
+
+clean <- count(stems_clean$sp)
+unclean <-count(stems_unclean$sp)
+
+clean$x
+Species <- c("Striped Maple", "Red Maple", "Yellow Birch", "Black Birch", "White Birch", "Grey Birch",
[…]

### 2020-10-28 doc/manuscript/Dow_manuscript.Rmd
add dendroband numbers

@@
 
-Within the plot, metal dendrometer bands were installed on *>#* trees and measured with a digital caliper approximately every 1-2 weeks within the growing season for *9* years (*list ranges, for each site if different*).
+Metal dendrometer bands installed on *700-755* trees at Harvard Forest and *99-145* trees at SCBI were measured with a digital caliper approximately every 1-2 weeks within the growing season from 1998-2003, and 2011-2020 respectively (*list ranges, for each site if different*).
 *(don't bother to mention biannual bands, unless you end up using the data--which you could, for annual growth, if you'd like to increase the sample size.
-You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`).
-Across years, the number of bands sampled averaged *~150* at SCBI, # at Harvard Forest*, and # at SERC*.
-Four main species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.
+You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`) (tags_per_year_preclean in data folder).
+Across years, the number of bands sampled averaged *~135* at SCBI, *~717* at Harvard Forest*, and # at SERC*.
+At SCBI, 4 species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.
+At Harvard Forest, 11 species were sampled including 8 diffuse-porous and 3 ring-porous (SI table)(bands_bySP csv's in data folder).
 *(summary of how bands were distributed across species, focusing on ring/ diffuse porous distinction.
 details in an SI table)*
-
-(5 years @ Harvard Forest, 10 @ SCBI)
 

### 2020-10-28 doc/manuscript/Dow_manuscript.Rmd
add HF description

@@
 *(Give specifics, which may vary by site.
-Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)* The median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, and the median end date was 10/17, coinciding with late senesence in the plot.
-The earliest start date for intraannual measurements was in 2020 on March 30, and the latest end date was on 2012 on November 26.
+Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)* At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, and the median end date was 10/17, coinciding with late senescence in the plot.
+The earliest start date for intraannual measurements at SCBI was March 30, 2020 and the latest end date was November 26, 2012.
+Harvard Forest measurements took place at a similar time, with the median start date of 4/23 and median end date of 10/30.
+1998 was an anomalous year where initial measurements were taken on 1/5, but not taken again until 4/15.
+The latest end date was November 11, 2002.
 
@@
 Mistakes were removed from the raw data before modeling.
-(## cases) 2.
+2.
 Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern.
@@
 In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level.
-(## cases) 3.
+3.
 Other - cases where data was clearly wrong but with unknown causes.
-This seemed to happen mostly in 2011, when the program was just beginning.
 In cases where several measurements were an issue, they were simply removed.
@@
 
-Daily maximum temperatures (TMAX) were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github).
-The R package climpact was used to plot TMAX for visual inspection (# points removed?).
+Daily maximum temperatures (TMAX) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github).
+Daily maximum temperatures at Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
+The R package climpact () was used to plot TMAX for visual inspection.
 Climpact also identifies readings that were greater than 3 standard deviations away from yearly means.

### 2020-11-01 doc/manuscript/Dow_tables.Rmd
changes to lots of things

---
title: ''
author: ''
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document:
    df_print: paged
bibliography: null
biblio-style: apalike
csl: new-phytologist.csl
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
---
```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table1 <- read.csv("tables_figures/HF_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
kbl(table2, format = "latex", booktabs = T)
```

```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
kbl(table2, format = "latex", booktabs = T)
```

```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
kbl(table2, format = "latex", booktabs = T)
```

### 2020-11-01 RScripts/CMI_analysis.R

@@ -1,0 +1,184 @@
+#Calculate Climate moisture index (CMI: Monthly precip - PET)
+library(readr)
+library(dplyr)
+library(tidyverse)
+library(reshape2)
+pet <- read_csv("C:/Users/world/Desktop/Github/Climatenew/Climate_Data/CRU/CRU_v4_04/pet.1901.2019-ForestGEO_sites-6-03.csv")
+pre <- read_csv("C:/Users/world/Desktop/Github/Climatenew/Climate_Data/CRU/CRU_v4_04/pre.1901.2019-ForestGEO_sites-6-03.csv")
+
+#### HF ----
+HF_pet <- pet[pet$sites.sitename == "Harvard_Forest",]
+HF_pet <- HF_pet %>% select(colnames(HF_pet[grepl("1998|1999|2000|2001|2002|2003", colnames(HF_pet))]))
+HF_pet_melt <- HF_pet %>%
+  melt()
+names(HF_pet_melt) <- c("XDATE", "PET")
+
+HF_pre <- pre[pre$sites.sitename == "Harvard_Forest",]
+HF_pre <- HF_pre %>% select(colnames(HF_pre[grepl("1998|1999|2000|2001|2002|2003", colnames(HF_pre))]))
+HF_pre_melt <- HF_pre %>%
+  melt()
+names(HF_pre_melt) <- c("XDATE", "PRE")
+
+HF_CMI <- left_join(HF_pet_melt, HF_pre_melt)
+HF_CMI$CMI <- HF_CMI$PRE-HF_CMI$PET
+
+HF_CMI$year <- substr(HF_CMI$XDATE,2, 5)
+HF_CMI$month <- substr(HF_CMI$XDATE,7, 8)
+HF_CMI$day <- substr(HF_CMI$XDATE,10,11)
+
+HF_CMI <- HF_CMI[complete.cases(HF_CMI$CMI),]
+
+HF_CMI$DATE <- paste(HF_CMI$day, HF_CMI$month, HF_CMI$year, sep = "/")
+HF_CMI$DATE <- strptime(as.character(HF_CMI$DATE), format = "%d/%m/%Y")
+HF_CMI$DATE <- format(HF_CMI$DATE, "%d/%m/%Y")
+
+write.csv(HF_CMI, file = "HF_CMI.csv", row.names = FALSE)
+
+####SCBI ----
+SCBI_pet <- pet[pet$sites.sitename == "Smithsonian_Conservation_Biology_Institute",]
[…]

### 2020-11-01 RScripts/HF_wood_phenology.R

@@ -20,7 +20,6 @@ all_stems$month <- as.numeric(format(all_stems$date, format = "%m"))
 all_stems$day <- as.numeric(format(all_stems$date, format = "%d"))
 
 
-
 # 2. Load necessary functions ----------------------------------------------------
 # Sean's functions -- I think they are supposed to be in the rdendrom package but I couldn't the package to work correctly.
 ### code chunk number 2: lg5.functions
@@ -330,7 +329,8 @@ plot_tag_years <- all_stems %>%
 LG5_parameters <- vector(mode = "list", length = length(plot_tag_years))
 names(LG5_parameters) <- plot_tag_years
 growth <- NULL
-all_stems$tag_stem <- paste0(all_stems$plot, all_stems$tag, all_stems$year)
+all_stems$tag_stem <- paste0(all_stems$plot, all_stems$tag)#, all_stems$year)
+unique(all_stems$tag_stem)
 all_stems <- all_stems[!(all_stems$tag_stem %in% "A191998"),]
 all_stems <- all_stems[complete.cases(all_stems$dbh2),]
 #tag_years <- unique(all_stems$tag_stem)
@@ -538,6 +538,7 @@ for (q in 1998:2003) {
       if(skip_to_next) { next }
       names(optim.output.df) <- c("Tree.no", "Optim.call", "L", "K", "doy_ip", "r", "theta", "ML", "Best.ML")
       # write.csv(optim.output.df, file = "Optim_output.csv", quote = FALSE, row.names = FALSE)
+      ML_value <- as.character(optim.output.df[optim.output.df$Optim.call == winning.optim.call, 8])
 
 
       winner.tab <- table(winning.optim.call)
@@ -630,18 +631,18 @@ for (q in 1998:2003) {
 
       names(Param.df) <- c("L", "K", "doy_ip", "r", "theta", "a", "b")
       # write.csv(Param.df, file = "SCBI_hi_lo_lg5.csv", quote = FALSE, row.names = FALSE)
-      ML_value <- NA
+      #ML_value <- NA
 
       #for(i in 1:nrow(LG5_parameter_values)){
-        K <- as.numeric(Param.df[1,2])
-        L <- as.numeric(Param.df[1,1])
-        doy.ip <- as.numeric(Param.df[1,3])
-        r <- as.numeric(Param.df[1,4])
-        theta <- as.numeric(Param.df[1,5])
[…]

### 2020-11-01 RScripts/SCBI_wood_phenology.R

@@ -130,6 +130,7 @@ lg5.pred <- function(params, doy) {
   dbh <- L + ((K - L) / (1 + 1 / theta * exp(-(r * (doy - doy.ip) / theta))^theta))
   return(dbh)
 }
+lg5.pred(params, 200)
 
 lg5.ML <- function(params, doy, dbh, resid.sd) {
   pred.dbh <- lg5.pred(params, doy)
@@ -425,7 +426,7 @@ get.lg5.ML <- function(params, doy, dbh, resid.sd) {
 ## 2v. If new.band=1, we have a measure, and there's no new dbh (indicating a new dbh wasn't recorded when the band was changed), dbh2 is the sum of the differences of the previous dbh2's added to the most recent dbh2.
 ## 2vi. UNCOMMON If new.band=1 , measure is NA, and the dbh in the original column is unchanged , dbh2 is the sum of the differences of the previous dbh2's added to the most recent dbh2.
 ## 2vii. UNCOMMON If new.band=1, measure is NA, and dbh is different, dbh2 is the new dbh plus the mean of the differences of the previous dbh2's.
-
+mean(tags_per_year_HF$`Harvard Forest`)
 ## intraannual data
 for (stems in names(all_stems_intra)) {
   tree.n <- all_stems_intra[[stems]]
@@ -566,7 +567,8 @@ all_stems <- read.csv("Data/all_stems.csv")
 # 4. Loop to create a final "master dataframe" -------
 # The following loop will cycle through years, species, and tags to create a final "master dataframe"
 # It includes two of Sean's functions wrapped with some of my code
-
+all_stems <- read_csv("data/all_stems.csv")
+all_stems$dbh2 <- all_stems$dbh2/10
 # Objects needed for the following loop
 data <- data.frame(NULL)
 # throwaways <- data.frame(NULL)
@@ -591,13 +593,13 @@ for (q in 2011:2019) {
   Stem2 <- subset(all_stems, year == q)
   for (w in unique(Stem2$sp)) { # removes trees with less than 10 measurements in each year
     Stem3 <- subset(Stem2, sp == w)
-    count_df <- count(Stem3$tag_stem)
-    count_df <- subset(count_df, freq >= 10)
-    Stem3 <- Stem3[Stem3$tag_stem %in% count_df$x, ]
[…]

### 2020-11-01 RScripts/additional_figures.R

@@ -147,13 +147,13 @@ ggsave("doc/manuscript/tables_figures/schematic.png", plot = schematic, width = 
 
 
 # Growth_Cruves_all: Percent modeled growth and cummulative percent modeled growth ---
-Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V8CLEAN.csv") %>%
+Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V9CLEAN.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   # Rename ring porous to not have a space
   mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
 
-Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HarvardForest_V3CLEAN.csv") %>%
+Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HarvardForest_V4CLEAN.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   # Rename ring porous to not have a space
@@ -161,7 +161,7 @@ Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HarvardForest_V3CLEAN.csv
 
 
 # As generated in Rscripts/SCBI_wood_phenology.R
-LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V8CLEAN.csv")
+LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V9CLEAN.csv")
 LG5_parameter_values_hf <- read_csv("Data/LG5_parameter_values_HarvardForest_V3CLEAN.csv")
 
 # Generalized 5-parameter logistic function (modified version of Sean's function)
@@ -217,7 +217,7 @@ percent_growth_hf <- Wood_pheno_table_hf %>%
   mutate(
     dbh = lg5(L, K, doy_ip, r, theta, doy),
     dbh_growth = dbh - lag(dbh),
-    dbh_total_growth = lg5(L, K, doy_ip, r, theta, 365)- lg5(L, K, doy_ip, r, theta, 1),
+    dbh_total_growth = K-L,
     dbh_growth_percent = dbh_growth/dbh_total_growth
   ) %>%
   filter(!is.na(dbh_growth)) %>%
[…]

### 2020-11-01 RScripts/additional_figures_HarvardForest.R

@@ -168,7 +168,7 @@ LG5_parameter_values <- read_csv("Data/LG5_parameter_values_HarvardForest_V2.csv
 LG5_parameter_values$site_tag <- paste0(LG5_parameter_values$plot, LG5_parameter_values$tag)
 # Generalized 5-parameter logistic function (modified version of Sean's function)
 lg5 <- function(L, K, doy_ip, r, theta, doy) {
- For specified 5 parameters and x = doy, compute y = dbh
+# For specified 5 parameters and x = doy, compute y = dbh
   dbh <- L + ((K - L) / (1 + 1/theta * exp(-(r * (doy - doy_ip) / theta)) ^ theta))
   return(dbh)
 }

### 2020-11-01 RScripts/climwin_HF.R

@@ -44,7 +44,7 @@ climate$DATE <- strptime(as.character(climate$DATE), format = "%d/%m/%Y")
 climate$DATE <- format(climate$DATE, "%d/%m/%Y")
 
 #The data containing the biological responses for testing
-Wood_pheno_table <- read_csv("Data/Wood_pheno_table_HFtemp.csv") #Master datafrmae containing 20%, 50%, and 75% growth milestones
+Wood_pheno_table <- read_csv("Data/Wood_pheno_table_HarvardForest_V3CLEAN.csv") #Master datafrmae containing 20%, 50%, and 75% growth milestones
 ####Percentage DOY climwin all wood types, all percs DAILY ----
 #rangedates <- data.frame("feb1", round(30)) #/7 for week
 #rangedates2 <- data.frame("mar1", round(59)) #/7 for week

### 2020-11-01 RScripts/data_cleaning.R

@@ -3,13 +3,13 @@ library(scales)
 library(tidybayes)
 
 #SCBI ----
-Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V8RAW.csv") %>%
+Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V10RAW.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   # Rename ring porous to not have a space
   mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
 
-LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V8RAW.csv")
+LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V10RAW.csv")
 
 # Generalized 5-parameter logistic function (modified version of Sean's function)
 lg5 <- function(L, K, doy_ip, r, theta, doy) {
@@ -34,9 +34,9 @@ percent_growth <- Wood_pheno_table_scbi %>%
   unnest_longer(doy) %>%
   group_by(tag_year) %>%
   mutate(
-    dbh = lg5(L, K, doy_ip, r, theta, doy),
+    dbh = lg5(a, b, doy_ip, r, theta, doy),
     dbh_growth = dbh - lag(dbh),
-    dbh_total_growth =  lg5(L, K, doy_ip, r, theta, 365)- lg5(L, K, doy_ip, r, theta, 1),
+    dbh_total_growth =  b-a,
     dbh_growth_percent = dbh_growth/dbh_total_growth
   ) %>%
   filter(!is.na(dbh_growth)) %>%
@@ -44,17 +44,22 @@ percent_growth <- Wood_pheno_table_scbi %>%
     dbh_growth_percent_cummulative = cumsum(dbh_growth_percent)
   )
 
+percent_growth_one <- subset(percent_growth, dbh_growth_percent_cummulative == 1)
+percent_growth <- percent_growth[!(percent_growth$tag_year %in% unique(percent_growth_one$tag_year)),]
 maxgrowthrate <- aggregate(percent_growth$dbh_growth_percent, by = list(percent_growth$tag_year), FUN = max)
 names(maxgrowthrate) <- c("tag_year", "rate")
 
+maxgrowthcumul <- aggregate(percent_growth$dbh_growth_percent_cummulative, by = list(percent_growth$tag_year), FUN = max)
[…]

### 2020-11-01 RScripts/wood_pheno_analysis_HarvardForest.R

@@ -10,7 +10,7 @@ library(rstanarm)
 library(broom.mixed)
 
 # Get growth data ----------------------------------
-Wood_pheno_table <- read_csv("Data/Wood_pheno_table_HFtemp.csv") %>%
+Wood_pheno_table <- read_csv("Data/Wood_pheno_table_HarvardForest_V4CLEAN.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   #filter(tot >= 1) %>%
@@ -135,7 +135,7 @@ Wood_pheno_table <- Wood_pheno_table %>%
   ) %>%
   arrange(tag, year)
 View(Wood_pheno_table)
-
+memory.limit(10000)
 
 
 # Fit multivariate model using climwinmeans ------------------------------------
@@ -153,11 +153,11 @@ sample_tags <- Wood_pheno_table %>%
 
 Wood_pheno_table <- Wood_pheno_table %>%
   #filter(tag %in% sample_tags) %>%
-  mutate(climwinmean = climwinmean - 18)
+  mutate(climwinmean = climwinmean - 13)
 
 # Delete all non-needed columns
 Wood_pheno_table <- Wood_pheno_table %>%
-  select(perc, tag, year, wood_type, sp, climwinmean, starts_with("DOY"))
+  select(perc, tag, year, wood_type, sp, climwinmean, starts_with("DOY"), site.x)
 
 
 # Convert to wide format for use in rstanarm::stan_mvmer()
@@ -167,9 +167,9 @@ Wood_pheno_table_wide <- Wood_pheno_table %>%
 # Fit multivariate model
 joint_model_climwinmeans <- stan_mvmer(
   formula = list(
-    DOY_25 ~ wood_type + wood_type:climwinmean + (1|tag),
-    DOY_50 ~ wood_type + wood_type:climwinmean + (1|tag),
-    DOY_75 ~ wood_type + wood_type:climwinmean + (1|tag)
+    DOY_25 ~ wood_type + wood_type:climwinmean + (1|site.x) + (1|tag),
+    DOY_50 ~ wood_type + wood_type:climwinmean + (1|site.x) + (1|tag),
+    DOY_75 ~ wood_type + wood_type:climwinmean + (1|site.x) + (1|tag)
   ),
   # Note we transform the data from tall/tidy format to wide format first:
   data = Wood_pheno_table_wide,

### 2020-11-02 doc/manuscript/Dow_manuscript.Rmd
Closes #8

@@
 
-•	Bayesian heirarchal model
-This mixed-effect model was run within a heirarchal bayesian framework.
+•	Bayesian heirarchical model
+This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language @rstanarm_2020.
+In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults.
+<!--
+TODO added by Bert: Whereever we specify model details, include details about prior from here:
+https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
+-->
 

### 2020-11-03 doc/manuscript/Dow_manuscript.Rmd
updated figure names/legends in manuscript

@@
 
-pheno_Tsensitivity_caption <- "Figure 4. Relationship between DOY and climwin mean tempeatures. For each 
+pheno_Tsensitivity_caption <- "Figure 4. Relationship between DOY and climwin mean temperatures. For each 
 observed climwin mean temperature value, the posterior mean of the fitted day of year
@@
 # Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/pheno_Tsensitivity.png")
+knitr::include_graphics("tables_figures/pheno_Tsensitivity_scbi.png")
+```
+
+```{r , out.width="100%", echo=FALSE, fig.align="center", fig.cap="Harvard. FIGURES WILL NEED TO BE COMBINED"}
+# Created in RScripts/bert/wood_phenology_analysis_bert.R
+knitr::include_graphics("tables_figures/pheno_Tsensitivity_HF.png")
 ```
@@
 knitr::include_graphics("tables_figures/DOYtiming.png")
-```
-
-```{r DOYtiming2, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
-knitr::include_graphics("tables_figures/DOYtiming_HF_PRELIM.png")
 ```

### 2020-11-03 doc/manuscript/Dow_tables.Rmd
quick changes to make tables knit
@camerondow35

@@
 ---
-```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
+```{r, Table HF_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
 library(kableExtra)
-table1 <- read.csv("tables_figures/HF_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table2, format = "latex", booktabs = T)
+table <- read.csv("tables_figures/HF_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
+kbl(table, format = "latex", booktabs = T)
 ```
 
-```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
+\newpage
+```{r, Table SCBI_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
 library(kableExtra)
-table2 <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table2, format = "latex", booktabs = T)
+table <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
+kbl(table, format = "latex", booktabs = T)
 ```
 
-```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
+\newpage
+```{r, Table sample_size_by_year, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
 library(kableExtra)
-table2 <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table2, format = "latex", booktabs = T)
+table <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
+kbl(table, format = "latex", booktabs = T)
 ```

### 2020-11-03 doc/manuscript/Dow_SI.Rmd
rename Dow_tables-->Dow_SI
@camerondow35 , this can be the start of the SI. I believe all tables here are destined to be SI tables

---
title: ''
author: ''
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document:
    df_print: paged
bibliography: null
biblio-style: apalike
csl: new-phytologist.csl
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
---

```{r, Table HF_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table <- read.csv("tables_figures/HF_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
kbl(table, format = "latex", booktabs = T)
```

\newpage
```{r, Table SCBI_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
kbl(table, format = "latex", booktabs = T)
```

\newpage
```{r, Table sample_size_by_year, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
kbl(table, format = "latex", booktabs = T)
```

### 2020-11-03 doc/manuscript/Dow_SI.Rmd
added climwin output figures to SI
had to copy over figures and rename (technical difficulties)

@@
 ---
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+path_to_results <- paste0(dirname(dirname(getwd())), "/results/") #I was trying to read directly from results folder, but couldn't get it working.
+
+# SUPPLEMENTARY INFO (copied from Dow_manuscript.Rmd)
+#SI table numbers
+sample_size_by_year= "S1"
+
+#SI table captions
+sample_size_by_year_caption <- "Table S1. Sample size by year"
+
+#SI figure numbers
+climwin_rp_DOY25_SCBI = "S1"
+climwin_dp_DOY25_SCBI = "S2"
+climwin_rp_DOY25_HARV = "S3"
+climwin_dp_DOY25_HARV = "S4"
+#SI figure captions
+climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
+climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for DOY 25 for diffuse porous species at SCBI."
+climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for DOY 25 for ring porous species at Harvard Forest."
+climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
+
+```
 
@@
 ```
+
+\newpage
+
+```{r climwin_rp_DOY25_SCBI, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_rp_DOY25_SCBI_short_caption}
+knitr::include_graphics("tables_figures/SCBI_mettower_ring-porous_DOY25.png")
+```
+
+\newpage
+
+```{r climwin_dp_DOY25_SCBI, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_dp_DOY25_SCBI_short_caption}
+knitr::include_graphics("tables_figures/SCBI_mettower_diffuse-porous_DOY25.png")
+```
+
+\newpage
+
+```{r climwin_rp_DOY25_HARV, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_rp_DOY25_HARV_short_caption}
+knitr::include_graphics("tables_figures/HF_ring-porous_DOY25.png")
+```
+
+\newpage
+
+```{r climwin_dp_DOY25_HARV, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_dp_DOY25_HARV_short_caption}
+knitr::include_graphics("tables_figures/HF_diffuse-porous_DOY25.png")
+```

### 2020-11-03 doc/manuscript/Dow_manuscript.Rmd
separately formatted hypothesis table, inserted in manuscript

@@
 
-#captions
-#tables
-hypothesis_table_caption <-"Table 1. Hypothesis table. (NEEDS FORMATTING...obviously!)"
+#figure captions
 
-#figures
 schematic_caption <- "Figure 1.
@@
 
-# SUPPLEMENTARY INFO
+# SUPPLEMENTARY INFO (this chunk must be repeated in Dow_SI.Rmd)
 #SI table numbers
@@
 ```{r hypothesis_table, eval=TRUE, echo=FALSE, warning=FALSE}
-library(knitr)  
-hypothesis_table <- read.csv(file = "tables_figures/hypothesis_table.csv")
-knitr::kable(hypothesis_table, caption = hypothesis_table_caption)
+knitr::include_graphics("tables_figures/hypothesis_table.png")
 ```

### 2020-11-03 doc/manuscript/tables_figures/tables.Rmd
separately formatted hypothesis table, inserted in manuscript

---
title: " "

output: 
  pdf_document:
    fig_caption: yes

header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}

---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
options(knitr.table.format = "latex")
```

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv(file = "hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

kable(hypothesis_table, col.names = c("Hypotheses and Specific Predictions", "RP", "DP","RP", "DP"), caption = "Table 1. Hypothesis table", format = "latex", booktabs = TRUE,  escape = F)%>%
  #column_spec(1, width = "4cm") %>%
  add_header_above(c(" ", "SCBI" = 2, "Harvard"=2 )) %>%
  pack_rows("1. Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("2. Maximum growth rates are independent of early spring temperatures.", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("3. Annual stem growth responds positively to warmer spring temperatures.", 7, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

### 2020-11-03 RScripts/data_cleaning.R
Update data_cleaning.R

@@ -3,6 +3,7 @@ library(scales)
 library(tidybayes)
 
 #SCBI ----
+#Load up percent growth DF ----
 Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V10RAW.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
@@ -41,62 +42,48 @@ percent_growth <- Wood_pheno_table_scbi %>%
   mutate(
     dbh_growth_percent_cummulative = cumsum(dbh_growth_percent)
   )
-
+#Clean the data ----
+#Break into RP and DP
 percent_growth_RP <- subset(percent_growth, wood_type == "ring-porous")
+
+#Remove models with single days > 2% growth (this is abnormal for RP trees in our study)
 maxgrowthrate <- aggregate(percent_growth_RP$dbh_growth_percent, by = list(percent_growth_RP$tag_year), FUN = max)
 names(maxgrowthrate) <- c("tag_year", "rate")#Maximum growth rate of each tree
-high_growthday <- subset(maxgrowthrate, rate >= 0.02) #Remove trees where one day had higher than 4% of total growth
+high_growthday <- subset(maxgrowthrate, rate >= 0.02) #Remove trees where one day had higher than 2% of total growth
 percent_growth_RP <- percent_growth_RP[!(percent_growth_RP$tag_year %in% unique(high_growthday$tag_year)),]
-#maxgrowthcumul <- aggregate(percent_growth$dbh_growth_percent_cummulative, by = list(percent_growth$tag_year), FUN = max)
-#maxgrowthcumul_good <- subset(maxgrowthcumul, x >= 0.99)
-#percent_growth <- percent_growth[percent_growth$tag_year %in% maxgrowthcumul_good$Group.1,]
-ratedoy <- percent_growth_RP[percent_growth_RP$dbh_growth_percent %in% maxgrowthrate$rate,]
+
+#Remove models with peak growth outside of expected season. IE, winter growth peaks
+ratedoy <- percent_growth_RP[percent_growth_RP$dbh_growth_percent %in% maxgrowthrate$rate,]#DOY where max growth occured
 ratedoysubset <- subset(ratedoy, doy >= 110 & doy <= 190) #Remove models where peak growth was in winter
 percent_growth_RP <- percent_growth_RP[percent_growth_RP$tag_year %in% ratedoysubset$tag_year,]
 
+#DIffuse porous - Same method
[…]

### 2020-11-10 doc/manuscript/Dow_manuscript.Rmd
adding some content from NEE refs

@@
 For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil.
-Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon (Bonan, 2008).
-However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).
+Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon [@bonan_forests_2008].
+However, their future under the current era of rapid change is uncertain [@friedlingstein_climatecarbon_2006].
 
@@
 The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).
+
+Leaf phenology--as characterized by satellite observations of NDVI or EVI-- is linked to the ecosystem CO~2~ uptake period.
+In turn, longer carbon uptake periods have been linked to greater net annual C sequestration (i.e., net ecosystem exchange, NEE) by ecosystems [@churkina_spatial_2005; @keenan_net_2014], although more of the interannual variation in forest NEE is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@fu_maximum_2019].
 
@@
 
+results align with @fu_maximum_2019, which shows that interannual variation in NEE is more strongly linked to peak C uptake rates than to growing season length.
+Our study demonstrates that this holds true for $ANPP_{woody}$ as well as NEE.
+
 Geng et al. 2020 GCB- Climate warming increases spring phenological differences among temperate trees.

### 2020-11-10 doc/manuscript/Dow_manuscript.Rmd
title change

@@
 
-**Title:** Effect of spring temperatures on tree growth phenology in two temperate deciduous forests
+**Title:** ~~Effect of spring temperatures on tree growth phenology in two temperate deciduous forests~~
+Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity
 

### 2020-11-10 doc/manuscript/Dow_manuscript.Rmd
update figs

--- 2020-11-10 doc/manuscript/Dow_manuscript.Rmd
@@
 
-**Title:** Effect of spring temperatures on tree growth phenology in two temperate deciduous forests
+**Title:** ~~Effect of spring temperatures on tree growth phenology in two temperate deciduous forests~~
+Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little/only modest(?) effect on annual woody productivity
 
@@
 
-pheno_Tsensitivity_caption <- "Figure 4. Relationship between DOY and climwin mean temperatures. For each 
+pheno_Tsensitivity_caption <- "Figure 4. Relationship between growth parameters and climwin mean temperatures. For each 
 observed climwin mean temperature value, the posterior mean of the fitted day of year
@@
 # Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/pheno_Tsensitivity_scbi.png")
-```
-
-```{r , out.width="100%", echo=FALSE, fig.align="center", fig.cap="Harvard. FIGURES WILL NEED TO BE COMBINED"}
-# Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/pheno_Tsensitivity_HF.png")
+knitr::include_graphics("tables_figures/pheno_Tsensitivity_combo.png")
 ```

### 2020-11-10 doc/manuscript/Dow_manuscript.Rmd
incorporating refs
#19

@@
 
-Leaf phenology--as characterized by satellite observations of NDVI or EVI-- is linked to the ecosystem CO~2~ uptake period.
-In turn, longer carbon uptake periods have been linked to greater net annual C sequestration (i.e., net ecosystem exchange, NEE) by ecosystems [@churkina_spatial_2005; @keenan_net_2014], although more of the interannual variation in forest NEE is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@fu_maximum_2019].
+Leaf phenology is linked to the ecosystem CO~2~ uptake period, and in turn longer carbon uptake periods have been linked to greater net annual C sequestration (*i.e.*, net ecosystem exchange, $NEE$; @churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].
+However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019].
 
-Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
+Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, we know little about the effect on tree growth and woody productivity, which can be decoupled from leaf phenology and represent only a modest proportion of $NEE$.
+
+~~,Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
+~~
 

### 2020-11-11 RScripts/additional_figures.R
updated figures

@@ -146,14 +146,14 @@ ggsave("doc/manuscript/tables_figures/schematic.png", plot = schematic, width = 
 
 
 
-# Growth_Cruves_all: Percent modeled growth and cummulative percent modeled growth ---
-Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V9CLEAN.csv") %>%
+# Growth_Cruves_all: Percent modeled growth and cummulative percent modeled growth ----
+Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V13CLEAN.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   # Rename ring porous to not have a space
   mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
 
-Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HarvardForest_V5CLEAN.csv") %>%
+Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HarvardForest_V9CLEAN.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   # Rename ring porous to not have a space
@@ -161,8 +161,8 @@ Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HarvardForest_V5CLEAN.csv
 
 
 # As generated in Rscripts/SCBI_wood_phenology.R
-LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V10CLEAN.csv")
-LG5_parameter_values_hf <- read_csv("Data/LG5_parameter_values_HarvardForest_V5CLEAN.csv")
+LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V13CLEAN.csv")
+LG5_parameter_values_hf <- read_csv("Data/LG5_parameter_values_HarvardForest_V9CLEAN.csv")
 
 # Generalized 5-parameter logistic function (modified version of Sean's function)
 lg5 <- function(L, K, doy_ip, r, theta, doy) {
@@ -272,13 +272,25 @@ ggplot(percent_growth_hf, aes(x = doy, y = dbh_growth_percent_cummulative, group
 as.table = TRUE, nrow=2, ncol=2) ###as.table specifies order if multiple rows
 
 dev.off()
+# ----
+#DOY timing firgure----
[…]

### 2020-11-11 RScripts/wood_pheno_analysis_HarvardForest.R

@@ -10,7 +10,7 @@ library(rstanarm)
 library(broom.mixed)
 
 # Get growth data ----------------------------------
-Wood_pheno_table <- read_csv("Data/Wood_pheno_table_HarvardForest_V6CLEAN.csv") %>%
+Wood_pheno_table <- read_csv("Data/Wood_pheno_table_HarvardForest_V9CLEAN.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
   #filter(tot >= 1) %>%
@@ -25,15 +25,15 @@ sevenfive <- subset(Wood_pheno_table, perc == .75)
 #25-50
 twofifty <- cbind(twofive,fifty$DOY)
 twofifty$twentyfive_to_fifty <- twofifty$`fifty$DOY`-twofifty$DOY
-twofifty <- twofifty[,c(3,6,16)]
+twofifty <- twofifty[,c(3,6,17)]
 #50-75
 fiftyseventy <- cbind(fifty, sevenfive$DOY)
 fiftyseventy$fifty_to_seventy <- fiftyseventy$`sevenfive$DOY`-fiftyseventy$DOY
-fiftyseventy <- fiftyseventy[,c(3,6,16)]
+fiftyseventy <- fiftyseventy[,c(3,6,17)]
 #25-75
 twosevenfive <- cbind(twofive, sevenfive$DOY)
 twosevenfive$seasonlength <- twosevenfive$`sevenfive$DOY`-twosevenfive$DOY
-twosevenfive <- twosevenfive[,c(3,6,16)]
+twosevenfive <- twosevenfive[,c(3,6,17)]
 
 
 # Create temperature variables ----------------------------------
@@ -51,7 +51,7 @@ weatherdata <-
 # Rename RP flag set by Cam
 #rename(flagrp = flag)
 climwindows <-
-  read.csv("results/Climwin_results/Weekly/Harvard Forest/weekly_climwin_results_all_HF.csv") %>%
+  read.csv("results/Climwin_results/Weekly/Harvard Forest/weekly_climwin_results_all_HF_975.csv") %>%
   filter(wood_type != "other") %>%
   mutate(
   median_windowopendate = as.Date(median_windowopendate, format = "%Y-%m-%d"),

### 2020-11-11 doc/manuscript/Dow_manuscript.Rmd
Update Dow_manuscript.Rmd

@@
 
-Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, we know little about the effect on tree growth and woody productivity, which can be decoupled from leaf phenology and represent only a modest proportion of $NEE$.
-
-~~,Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).
-~~
+Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, we know little about the effect on tree growth and woody productivity, which can be decoupled from leaf phenology (D'Orangeville et al, in review) and represent only a modest proportion of total C sequestration (Anderson-Teixeira et al.
+in review).
 
@@
 *(but check into this....
-I suspect there may be some that we're missing)*
+I suspect there may be some that we're missing--perhaps auto dendro bands in Europe.
+Neil did say that there's little on early spring T)*
 Warming events in late-spring were shown to shift leaf emergence timing back more than early-spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
@@
 Tree-ring records give some hints that warmer springs may increase annual growth (Helcoski et al.
-2019), but tree-rings tell us nothing about the timing of that growth.
+2019; probably some more), but tree-rings tell us nothing about the timing of that growth.
 Timing of intraannual stem-growth is extremely important as well.

### 2020-11-12 doc/manuscript/Dow_manuscript.Rmd
study sites

@@
 
-Study sites included **2/3** temperate forests in the Eastern United states, **both/all** part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al.
-2015] (**Table 1?**).
-**Both/all** sites are dominated by broadleaf deciduous trees.
-*(insert site descriptions, including lat/long, elevation, plot size, climate info, dominant species.
-If we include SERC and if space permits, we may put this in a table, otherwise just text)*
+Study sites included two temperate forests in the Eastern United States, both part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al. 2015].
 
-Our first study site was the forest dynamics plot located at the Smithsonian COnservation Biology Institute 5 KM South of Front Royal, Virginia within the Blue Ridge Mountains (38 °53036.6″ N, 78°08043.4″W; elevation 273–338 m).
+Our first study site was the forest dynamics plot located at the Smithsonian Conservation Biology Institute 5 km South of Front Royal, Virginia within the Blue Ridge Mountains (38.8935° N, -78.1454° W; elevation 273–338 m.a.s.l.).
 The forest here was established in the mid-19th century after conversion from agricultural fields.
-Dominate canopy species within the 25.6 HA forest plot include tulip poplar (Liriodendron tulipifera), hickories (Carya spp.), and oaks (Quercus spp.).
+Dominate canopy species within the 25.6 HA forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*).
 The climate is humid temperate, with a mean annual precipitation of 998 mm, January temperatures averaging 1 degrees C and July temperatures averaging 24 degrees C (Helcoski et al, 2019).
 
-The second study site was Harvard Forest in the central Massachusetts town of Petersham (42°53′ N, 72°17′ W, elevation?).
-Harvard forest was established in 1907 as a study site for university students and researchers.
-(Previous use?) Dominate species within the approximately 13 HA plot are hemlock (Tsuga spp.), oak (Quercus spp.) and maple (Acer spp.).
+The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
+*Harvard Forest was established in 1907 as a study site for university students and researchers.
+(Previous use?)* Dominate species within the approximately 13 HA plot are hemlock (*Tsuga spp.*), oak (*Quercus spp.*) and maple (*Acer spp.*).
 The climate is temperate continental with a mean annual precipitation of 1120 mm, January temperatures averaging -12 degrees C and July temperatures averaging 19 degrees C (D'Orangeville et al, ).
-
-The third study site was the forest dynamics plot located at the Smithsonian Ecological Research Center in Edgewater, Maryland near a sub estuary of the Chesapeake Bay (38°88′ N, 76°55′ W, elevation?).
-The forest here established roughly 120 years ago, while the plot was established in 2007.
-The 16 HA plot is dominated by tulip poplar (Liriodendron tulipifera), oak (Quercus spp.), beech (Fagus spp.), hickory (Carya spp.), and maple (Acer spp.).
-(Climate summary)
 

### 2020-11-12 doc/manuscript/Dow_manuscript.Rmd
manuscript work
add species_table, misc clean-up

@@
 hypothesis_table=1
+species_table=2
 
@@
 
+I like the [Nature summary paragraph template](https://cbs.umn.edu/sites/cbs.umn.edu/files/public/downloads/Annotated_Nature_abstract.pdf):
+
+- One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline.
+- Two to three sentences ofmore detailed background, comprehensible to scientists in related disciplines.
+- One sentence clearly stating the general problem being addressed by this particular study.
+- One sentence summarising the main result (with the words “here we show” or their equivalent).
+- Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge.
+- One or two sentences to put the results into a more general context.
+- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline
+
 **Keywords**:
@@
 
-Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research has been done on this topic. 
+Earlier timing of spring leaf-up and initiation of carbon sequestration could signal that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research has been done on this topic.
 In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns.
@@
 #### (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
-A recent publication by D'Orangeville et al () found that leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous).
-For example, they observed that bud burst was occurring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
+~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al.
+YY**).
+For example, ~~they observed that~~ bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
 
@@
 Results from the paper by D’Orangeville et al.
-(in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time.
+(in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time.
 If leaf development is sped up when spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.
@@
 
-Our first study site was the forest dynamics plot located at the Smithsonian Conservation Biology Institute 5 km South of Front Royal, Virginia within the Blue Ridge Mountains (38.8935° N, -78.1454° W; elevation 273–338 m.a.s.l.).
+Our first study site was the forest dynamics plot located at the Smithsonian Conservation Biology Institute 5 km South of Front Royal, Virginia within the Blue Ridge Mountains (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.).
 The forest here was established in the mid-19th century after conversion from agricultural fields.
-Dominate canopy species within the 25.6 HA forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*).
+Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*).
 The climate is humid temperate, with a mean annual precipitation of 998 mm, January temperatures averaging 1 degrees C and July temperatures averaging 24 degrees C (Helcoski et al, 2019).
@@
 
-Metal dendrometer bands installed on *700-755* trees at Harvard Forest and *99-145* trees at SCBI were measured with a digital caliper approximately every 1-2 weeks within the growing season from 1998-2003, and 2011-2020 respectively (*list ranges, for each site if different*).
-*(don't bother to mention biannual bands, unless you end up using the data--which you could, for annual growth, if you'd like to increase the sample size.
-You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`) (tags_per_year_preclean in data folder).
-Across years, the number of bands sampled averaged *~135* at SCBI, *~717* at Harvard Forest*, and # at SERC*.
-At SCBI, 4 species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.
-At Harvard Forest, 11 species were sampled including 8 diffuse-porous and 3 ring-porous (SI table)(bands_bySP csv's in data folder).
-*(summary of how bands were distributed across species, focusing on ring/ diffuse porous distinction.
-details in an SI table)*
+Metal dendrometer bands installed on *99-145* and *700-755* trees within the SCBI and Harvard Forest ForestGEO plots, respectively, were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 and 1998-2003, respectively.
+The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`) *(tags_per_year_preclean in data folder)*.
+Across years, the number of bands sampled averaged *~135* at SCBI and  *~717* at Harvard Forest.
+At SCBI, four species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.
+At Harvard Forest, eleven species were sampled including eight diffuse-porous and three ring-porous (Table `r species_table`)*(bands_bySP csv's in data folder)*.
+
+*(Do we want a table of species here?? Columns would be Species, xylem porosity, n trees, n tree-years (after cleaning). Details will be in an SI table)*
+
+```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+species_table <- read.csv("tables_figures/species_table.csv",stringsAsFactors = FALSE, check.names = FALSE)
+knitr::kable(species_table) 
+```
 
 Measurements were timed to begin before beginning of spring growth and to continue through the cessation of growth in the fall.
-*(Give specifics, which may vary by site.
-Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)* At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, and the median end date was 10/17, coinciding with late senescence in the plot.
+At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, and the median end date was 10/17, coinciding with late senescence in the plot.
 The earliest start date for intraannual measurements at SCBI was March 30, 2020 and the latest end date was November 26, 2012.
@@
 
+(*The following seems out of date, right? Scripted cleaning is not described.*)
 The dendroband data were manually cleaned by visual inspection.
@@
 
-Daily maximum temperatures (TMAX) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github).
+Daily maximum temperatures ($T_{MAX}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github).
 Daily maximum temperatures at Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
-The R package climpact () was used to plot TMAX for visual inspection.
+The R package climpact () was used to plot $T_{MAX}$ for visual inspection.
 Climpact also identifies readings that were greater than 3 standard deviations away from yearly means.
 These points were labeled as outliers and removed from the dataset.
-Gaps in the SCBI meteorological tower data were subsequently filled using TMAX readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
+Gaps in the SCBI meteorological tower data were subsequently filled using $T_{MAX}$ readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
+(*how were gaps filled for Harvard?
+Or was the record good?*)
 
@@
 The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package Climwin.
-This package tests the correlation between climate variables (TMAX) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'.
+This package tests the correlation between climate variables ($T_{MAX}$) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'.
 as the best-model.
@@
 This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth.
-For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?).
+For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June $T_{MAX}$ on trees of that wood type (figure or table?).
 This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year.
-Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high TMAX in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
+Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high $T_{MAX}$ in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
 How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

### 2020-11-12 doc/manuscript/Dow_manuscript.Rmd
add mean growth parameters table

@@
 species_table=2
+mean_growth_parameters_table=3
 
@@
 
+```{r mean_growth_parameters_table, eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+mean_growth_parameters <- read.csv("tables_figures/mean_growth_parameters.csv",stringsAsFactors = FALSE, check.names = FALSE)
+knitr::kable(mean_growth_parameters) 
+```
+
 *Optimal climate windows*

### 2020-11-12 doc/manuscript/Dow_manuscript.Rmd
update title

@@
 **Title:** ~~Effect of spring temperatures on tree growth phenology in two temperate deciduous forests~~
-Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little/only modest(?) effect on annual woody productivity
+Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity
 
@@
 
-Cameron Dow^1^
-Loïc D’Orangeville
+Cameron Dow^1^,
+Loïc D’Orangeville,
 Erika B. Gonzalez-Akre^1^,
@@
 Albert Kim^1^,
-David A. Orwig
+David A. Orwig,
 Kristina J. Anderson-Teixeira^1,n^*

### 2020-11-13 doc/manuscript/Dow_manuscript.Rmd
add tree-rings, format, misc

@@
 output:
-  word_document:
+  bookdown::word_document2: 
+    reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
     fig_caption: yes
@@
 DOYtiming_Harvard = "5b"
+tree_rings = 6
 
@@
 DOYtiming_caption <- "Figure 5. Day of year where growth milestones were achied at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest."
+
+tree_rings_caption <- "Figure 6. Climate sensitiivity of annual growth from tree-rings. THIS IS JUST A PLACEHOLDER TO SHOW THE RESULTS--WE'LL WANT TO RE-MAKE FIGURE."
 
@@
 
-### Abstract
+# Abstract
 
@@
 
-### Introduction
+# Introduction
 
-#### (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
+## (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
 As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
@@
 
-#### (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
+## (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
 In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
@@
 
-#### (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
+## (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
 
@@
 
-#### (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
+## (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
 ~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al.
@@
 
-#### (5. Here, we...)
-Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within **two/three** forest dynamics plots in the Eastern USA.
+## (5. Here, we...)
+Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA.
 Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig.
@@
 
-### Materials and Methods
+# Materials and Methods
 
-#### Study sites and data
+## Study sites and data
 
@@
 
+\newpage
 *(Do we want a table of species here?? Columns would be Species, xylem porosity, n trees, n tree-years (after cleaning). Details will be in an SI table)*
@@
 The latest end date was November 11, 2002.
+
+(*tree rings, if included-- KAT can write this*)
 
@@
 
-#### Fitting growth curves
+## Fitting growth curves
 
@@
 
-#### Determining optimal climate windows
+## Determining optimal climate windows
 
@@
 
-#### Hypothesis testing
+## Hypothesis testing
 
@@
 
-### Results
+# Results
 
@@
 
-### Discussion
+*Climate sensitivity of growth*
+
+```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/tree_rings.tif")
+```
+
+# Discussion
 
@@
 
-### (Content to incorporate)
+# (Content to incorporate)
 
@@
 
-### Acknowledgements
-ForestGEO
+# Acknowledgements
+people: (everyone who helped collect data), Jess Shue (data), Neil Pederson (helpful discussion) 
+Funding: ForestGEO
 
-### Author Contributions
+# Author Contributions
 
-### Supporting Information
+# Supporting Information
 
@@
 
-### References
+# References

### 2020-11-13 doc/manuscript/tables_figures/tables.Rmd
add tree-rings, format, misc

@@
   pack_rows("2. Maximum growth rates are independent of early spring temperatures.", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("3. Annual stem growth responds positively to warmer spring temperatures.", 7, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("3. Annual stem growth responds positively to warmer spring temperatures.", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-11-13 doc/manuscript/Dow_manuscript.Rmd
cleanup, organization

@@
 
-## (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
+## (forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
 As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
@@
 
-## (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
+## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
 In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
@@
 
-## (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
+## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
 
@@
 
-## (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
+## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
 ~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al.
-YY**).
+**).
 For example, ~~they observed that~~ bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
@@
 
-## (5. Here, we...)
+## (Here, we...)
 Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA.
@@
 
+## Data collection and cleaning
+### Dendrometer bands
+
 Metal dendrometer bands installed on *99-145* and *700-755* trees within the SCBI and Harvard Forest ForestGEO plots, respectively, were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 and 1998-2003, respectively.
@@
 
-(*tree rings, if included-- KAT can write this*)
-
 (*The following seems out of date, right? Scripted cleaning is not described.*)
@@
 
+### Tree-rings
+(*tree rings, if included-- KAT can write this*)
+
+### Climate
 Daily maximum temperatures ($T_{MAX}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github).
@@
 
-## Fitting growth curves
+## Data analysis
+### Fitting growth curves
 
@@
 
-## Determining optimal climate windows
+### Determining optimal climate windows
 
@@
 
-## Hypothesis testing
+### Hypothesis testing
 
@@
 
-*Phenological patterns*
+## Phenological patterns
 
-*(Numbers in bold will change. If we bring in HF and/or SERC, we should create a table, rather than trying to list everything in the text.)*
+*(Numbers in bold will change)*
 
-Across all trees and years, 25% of annual growth was achieved on **May 9** ($DOY_{25}$=**129**), 50% on **June 6** ($DOY_{50}$=**157**), and 75% on **July 3** ($DOY_{75}$=**184**), for an average peak growing season length of **56** days.
+Across all trees and years at SCBI, 
+25% of annual growth was achieved on **May 9** ($DOY_{25}$=**129**), 
+50% on **June 6** ($DOY_{50}$=**157**), 
+and 75% on **July 3** ($DOY_{75}$=**184**), 
+for an average peak growing season length of **56** days. 
 Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 7** ($DOY_{max}$=**158**). 
+At Harvard Forest,
+25% of annual growth was achieved on **XX** ($DOY_{25}$=**XX**), 
+50% on **XX** ($DOY_{50}$=**XX**), 
+and 75% on **XX** ($DOY_{75}$=**XX**), 
+for an average peak growing season length of **XX** days. 
+Maximum growth rate occurred on **XX** ($DOY_{max}$=**XX**).
 
-Growth pheonology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig.
-`r growth_curves_all`).
-For ring-porous trees, on average, 25% of annual growth was achieved on **April 23** ($DOY_{25}$=**113**), 50% on **May 27** ($DOY_{50}$=**147**), and 75% on **June 30** ($DOY_{75}$=**181**), for an average peak growing season length of **68** days.
-Maximum growth rate occurred *on $DOY_{50}$* and averaged .048 **mm day^-1^ (?)**.
-For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 2** ($DOY_{25}$=**153**), 50% on **June 20** ($DOY_{50}$=**171**), and 75% on **July 9** ($DOY_{75}$=**190**), for an average peak growing season length of **37** days.
-Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 21** ($DOY_{max}$=**172**) and averaged .068 **mm day^-1^ (?)**.
+At both sites, growth phenology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig.
+`r growth_curves_all`, Table `r mean_growth_parameters_table`).
+
+~~For ring-porous trees, on average, 25% of annual growth was achieved on **April 23** ($DOY_{25}$=**113**), 50% on **May 27** ($DOY_{50}$=**147**), and 75% on **June 30** ($DOY_{75}$=**181**), for an average peak growing season length of **68** days.
+Maximum growth rate occurred *on $DOY_{50}$* and averaged .048 **mm day^-1^ (?)**.~~
+~~For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 2** ($DOY_{25}$=**153**), 50% on **June 20** ($DOY_{50}$=**171**), and 75% on **July 9** ($DOY_{75}$=**190**), for an average peak growing season length of **37** days.
+Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 21** ($DOY_{max}$=**172**) and averaged .068 **mm day^-1^ (?)**.~~
 
@@
 
-*Optimal climate windows*
+## Optimal climate windows
 
@@
 
-*Climate sensitivity of phenology*
+## Climate sensitivity of phenology
 
-Warmer spring temperatures resulted in an earlier start of growth for ring-porous species and an earlier growing season for diffuse-porous species (Fig. `r pheno_Tsensitivity`).
+For both ring- and diffuse- porous species at both sites, warmer spring temperatures resulted in earlier achievement of growth milestones (Tables `r hypothesis_table`, Figs.
+`r pheno_Tsensitivity`, `r DOYtiming`).
+Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ all decreased significantly with increasing $T_{max}$ during the critical period (Table `r mean_growth_parameters_table`).
 
@@
 
-*Climate sensitivity of growth*
+```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/DOYtiming.png")
+```
+
+Because both $DOY_{25}$ and $DOY_{75}$ both shifted forward, peak growing season length was not consistently increased by warmer early spring temperatures (Tables `r hypothesis_table`, Figs.
+`r pheno_Tsensitivity`, `r DOYtiming`).
+At both sites, $DOY_{75}-DOY_{25}$ decreased with temperature for ring-porous species 
+and was neutral (SCBI) or positive (Harvard Forest) for diffuse-porous species.
+
+## Climate sensitivity of growth
+
+Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`). 
+Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer early spring temperatures.
+In contrast, diffuse-porous species exhibited lower $g_{max}$ in years with higher early spring $T_{max}$.
+
+Total annual growth rate, $\Delta DBH$, also was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`).
 
@@
 # Discussion
+
+## (summary of key findings, gap filled)
+We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity...
+
+## ("limitation in filling gap, how to interpret/fix")
+
+## ("limitation in generalization, how to interpret/fix")
+
+## ("contributions beyond")
+
+## ("science is better now")
 
@@
 This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
-
-```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming.png")
-```
 

### 2020-11-13 doc/manuscript/Dow_manuscript.Rmd
update hypothesis table, misc cleanup

@@
 
-**Title:** ~~Effect of spring temperatures on tree growth phenology in two temperate deciduous forests~~
-Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity
+**Title:** Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity
 
@@
 
-Cameron Dow^1^,
-Loïc D’Orangeville,
-Erika B. Gonzalez-Akre^1^,
-Ryan Helcoski^1^,
-Valentine Herrmann^1^,
-Ian McGregor^1^,
-Sean McMahon,
-Albert Kim^1^,
-David A. Orwig,
+Cameron Dow^1^
+
+Loïc D’Orangeville
+
+Erika B. Gonzalez-Akre^1^
+
+Ryan Helcoski^1^
+
+Valentine Herrmann^1^
+
+Ian McGregor^1^
+
+Sean McMahon
+
+Albert Kim^1^
+
+David A. Orwig
+
+Neil Pederson
+
 Kristina J. Anderson-Teixeira^1,n^*
@@
 
-```{r eval=FALSE, echo=FALSE}
-## This is a chunk that won't be evaluated or displayed. I use these for notes.
-## Target journal: ??? Global Change Biology????
-## Instructions for authors: 
-#[insert guidance on word count, n figures/ tables allowed, and then some rough planning of what goes in this manuscript]
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
 
-```
+## Target journal: Global Change Biology
+## Instructions for authors: https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/page/journal/13652486/homepage/forauthors.html
+## Primary Research Articles present the results of a completed research project and are up to 8000 words in length.
+Word limits apply to the main body of the text (Introduction, Materials and Methods, Results, Discussion, and Acknowledgements).
 
-```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+## I don't think GCB has an official limit. I just saw a recently accepted article with 10 figures and 4 tables, so we're fine!
+
 #Figures & Tables: listing figures and their number here is helpful for planning, and also allows automatic numbering within text -->easy changing of order.
@@
 climwin_caption <- "Figure 3.
+THIS IS OUT OF DATE.
 Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
@@
 # Abstract
+```{r eval=FALSE, echo=FALSE}
+## For GCB:
+## Abstract limit for GCB is 300 words. 
+## 6-10 keywords
+
+```
 
@@
 ## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
-~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al.
+Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al.
 **).
-For example, ~~they observed that~~ bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
+For example, bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
 
@@
 
-## Study sites and data
+## Study sites
 
@@
 
-## Data collection and cleaning
+## Data preparation
 ### Dendrometer bands
@@
 
+We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig. `r schematic`). 
+...DETAILS...
+
+After model fitting, we used  a script to remove outliers....
+
 ### Tree-rings
-(*tree rings, if included-- KAT can write this*)
+(*tree rings-- KAT can write this*)
 
@@
 
-## Data analysis
-### Fitting growth curves
+(*climate data for tree rings-- KAT can write this*)
 
-We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig. `r schematic`).
-
-### Determining optimal climate windows
+## (Climwin)
 
@@
 
-### Hypothesis testing
+## Hypothesis testing
 
@@
 # Acknowledgements
-people: (everyone who helped collect data), Jess Shue (data), Neil Pederson (helpful discussion) 
+people: (everyone who helped collect data), Jess Shue (data)
 Funding: ForestGEO

### 2020-11-13 doc/manuscript/tables_figures/tables.Rmd
update hypothesis table, misc cleanup

@@
 
-kable(hypothesis_table, col.names = c("Hypotheses and Specific Predictions", "RP", "DP","RP", "DP"), caption = "Table 1. Hypothesis table", format = "latex", booktabs = TRUE,  escape = F)%>%
+kable(hypothesis_table, col.names = c("Hypotheses and Specific Predictions", "RP", "DP","RP", "DP", "Results"), caption = "Table 1.
+Table of hypotheses and associated specific predictions, whether each was supported ('yes'; signficant at p<0.05), rejected ('no'; opposite trend signficant at p<0.05), or found insigificant ('n.s.'; no significant correlation), and display items showing the results.
+'RP' and 'DP' refer to ring- and diffuse- porous species, respectively.", format = "latex", booktabs = TRUE,  escape = F)%>%
   #column_spec(1, width = "4cm") %>%
-  add_header_above(c(" ", "SCBI" = 2, "Harvard"=2 )) %>%
+  add_header_above(c(" ", "SCBI" = 2, "Harvard"=2, " " )) %>%
   pack_rows("1. Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%

### 2020-11-16 RScripts/additional_figures.R
Done v2 of schematic

@@ -1,5 +1,6 @@
 library(tidyverse)
 library(scales)
+library(ggrepel)
 library(tidybayes)
 
 # Figure numbers from: https://github.com/SCBI-ForestGEO/growth_phenology/issues/13
@@ -9,14 +10,18 @@ library(tidybayes)
 # 1. L: lower asymptotes
 # 2. K: upper asymptotes
 # 3. doy.ip: inflection point
-# 4. r: slope of curve at inflection point
+# 4. r: value that determines slope of curve at inflection point ((K-L)*r/theta )/(1 + 1/theta)^2
 # 5. theta: theta = 1 indicates pre/post inflection post symmetry
-L <- 13.2
-K <- 13.8
-doy.ip <- 175
-r <- 0.05
-theta <- 1
+L <- 13
+K <- 14
+doy.ip <- 182
+r <- 0.025
+theta <- 1.5
 params <- c(L, K, doy.ip, r, theta)
+total_growth <- K - L
+
+start_doy <- 1
+end_doy <- 365
 
 # Function for logistic growth model written by Sean. Supposed to be in RDendrom
 # package https://github.com/seanmcm/RDendrom, but function does not seem to be included
@@ -31,58 +36,57 @@ lg5.pred <- function(params, doy) {
   return(dbh)
 }
 
-# Compute true values to plot true growth curve
+# 1. True growth (logistic curve):
 true_values <- tibble(
-  doy = seq(from = 100, to = 300),
+  doy = seq(from = start_doy, to = end_doy, by = 0.01),
   diameter = lg5.pred(params, doy)
 )
 
-# Max Growth rate
-intercept <- lg5.pred(params, doy.ip) - r *doy.ip
+# 2. Max growth rate/inflection point (tangent line):
+# Incorrect intercept since r is not true slope at inflection point:
+# intercept <- lg5.pred(params, doy.ip) - r *doy.ip
 
+# Correct slope and intercept
 r_true <- ((K-L)*r/theta )/(1 + 1/theta)^2
 intercept_true <- lg5.pred(params, doy.ip) - r_true *doy.ip
 
 offset <- 40
 max_growth_rate <- tibble(
-  doy = c(doy.ip - offset, doy.ip + offset),
+  doy = c(doy.ip - offset, doy.ip + offset)
 ) %>%
-  mutate(
-    diameter = doy * r_true + intercept_true
-  )
+  mutate(diameter = doy * r_true + intercept_true)
 
-
-
-
-
[…]

### 2020-11-16 RScripts/Pheno_tsensitivity_figure.R
correct SCBI total growth measurement

@@ -25,15 +25,15 @@ sevenfive <- subset(Wood_pheno_table, perc == .75)
 #25-50
 twofifty <- cbind(twofive,fifty$DOY)
 twofifty$twentyfive_to_fifty <- twofifty$`fifty$DOY`-twofifty$DOY
-twofifty <- twofifty[,c(3,7,15)]
+twofifty <- twofifty[,c(3,7,16)]
 #50-75
 fiftyseventy <- cbind(fifty, sevenfive$DOY)
 fiftyseventy$fifty_to_seventy <- fiftyseventy$`sevenfive$DOY`-fiftyseventy$DOY
-fiftyseventy <- fiftyseventy[,c(3,7,15)]
+fiftyseventy <- fiftyseventy[,c(3,7,16)]
 #25-75
 twosevenfive <- cbind(twofive, sevenfive$DOY)
 twosevenfive$seasonlength <- twosevenfive$`sevenfive$DOY`-twosevenfive$DOY
-twosevenfive <- twosevenfive[,c(3,7,15)]
+twosevenfive <- twosevenfive[,c(3,7,16)]
 
 
 # Create temperature variables ----------------------------------
@@ -242,9 +242,7 @@ fig6_DP
 
 ###TOTAL GROWTH----
 woodtable <- subset(Wood_pheno_table, perc == "DOY_25")
-total_formulaRP <- "tot ~ wood_type + wood_type:climwinmean + (1|tag)" %>% as.formula()
-
-earlyperiod_formulaRP <- "twentyfive_to_fifty ~ wood_type + wood_type:climwinmean + (1|tag)" %>% as.formula()
+total_formulaRP <- "dbh_total_growth ~ wood_type + wood_type:climwinmean + (1|tag)" %>% as.formula()
 
 mixedmodel_stanlmerRP_total <- stan_lmer(
   formula = total_formulaRP,
@@ -277,7 +275,7 @@ fig6_RP_tot <- ggplot() +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   scale_fill_brewer() +
   #facet_grid(perc) +
-  coord_cartesian(xlim =c(19.5, 23.5), ylim = c(-.4, 1.5))+
+  coord_cartesian(xlim =c(17.5, 22.5), ylim = c(-.4, 1.5))+
   theme(legend.position = "none")+
   labs(x = "", y = "Total (cm)", subtitle  = "Total Growth")
[…]

### 2020-11-16 RScripts/Pheno_tsensitivity_figure.R
fixed typo

@@ -98,7 +98,7 @@ climwin_windows <-
 #  summarize(climwinmean = mean(TMAX)) %>%
 #  mutate(wood_type = "ring-porous")
 climwinmeans_rp <- weatherdata %>%
-  filter(doy %in% c(climwindows[2,11]:climwindows[2,12])) %>%
+  filter(doy %in% c(climwindows[1,11]:climwindows[1,12])) %>%
   group_by(year) %>%
   summarize(climwinmean = mean(cleantmax)) %>%
   mutate(wood_type = "ring-porous")
@@ -221,7 +221,7 @@ fig6_RP <- ggplot() +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   scale_fill_brewer() +
   #facet_grid(perc) +
-  coord_cartesian(xlim =c(19.5, 23.5), ylim = c(80, 240))+
+  coord_cartesian(xlim =c(11.8, 19.7), ylim = c(80, 240))+
   theme(legend.position = "none")+
   labs(x = "", y = "DOY", col = "Percentile", title  = "Ring-porous", subtitle = "Relationship of DOY versus climwin mean temperature")
 #geom_text(data = climwin_windows, aes(label = window), x = -Inf, y = -Inf, hjust = -0.01, vjust = -0.5, family = "Avenir")
@@ -912,7 +912,7 @@ fig6_DP_mrdoy_hf <- ggplot() +
 fig6_DP_mrdoy_hf
 
 #test
-predictions_RP$sig <- ifelse(predictions_RP$perc == "DOY_25", "Not sig", "Sig")
+predictions_RP$sig <- ifelse(predictions_RP$perc == "DOY_75", "Not sig", "Sig")
 predictions_RP_25 <- subset(predictions_RP, perc == "DOY_25")
 predictions_RP_50 <- subset(predictions_RP, perc == "DOY_50")
 predictions_RP_75 <- subset(predictions_RP, perc == "DOY_75")
@@ -926,7 +926,7 @@ ggplot() +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   scale_fill_brewer() +
   #facet_grid(perc) +
-  coord_cartesian(xlim =c(17.5, 22.5), ylim = c(80, 240))+
+  coord_cartesian(xlim =c(11.8, 19.7), ylim = c(80, 240))+
[…]

### 2020-11-16 RScripts/additional_figures.R
v3 of schematic; closes #25

@@ -23,6 +23,14 @@ total_growth <- K - L
 start_doy <- 1
 end_doy <- 365
 
+window_open <- 60
+window_close <- 105
+
+critical_temp_window <- tibble(
+  x = c(window_open, window_close, window_close, window_open, window_open),
+  y = c()
+)
+
 # Function for logistic growth model written by Sean. Supposed to be in RDendrom
 # package https://github.com/seanmcm/RDendrom, but function does not seem to be included
 lg5.pred <- function(params, doy) {
@@ -38,9 +46,11 @@ lg5.pred <- function(params, doy) {
 
 # 1. True growth (logistic curve):
 true_values <- tibble(
-  doy = seq(from = start_doy, to = end_doy, by = 0.01),
+  doy = seq(from = start_doy, to = end_doy, by = 1),
   diameter = lg5.pred(params, doy)
-)
+) %>%
+  mutate(diff = diameter - lag(diameter))
+
 
 # 2. Max growth rate/inflection point (tangent line):
 # Incorrect intercept since r is not true slope at inflection point:
@@ -50,11 +60,6 @@ true_values <- tibble(
 r_true <- ((K-L)*r/theta )/(1 + 1/theta)^2
 intercept_true <- lg5.pred(params, doy.ip) - r_true *doy.ip
 
-offset <- 40
-max_growth_rate <- tibble(
-  doy = c(doy.ip - offset, doy.ip + offset)
-) %>%
-  mutate(diameter = doy * r_true + intercept_true)
 
 # 3. Observed values (noise/error added) (points)
 set.seed(76)
@@ -100,23 +105,29 @@ create_growth_polygon <- function(doy_diameter_quartile, percent){
 
 
 # Output figure
+geom.text.size <- 3.5
+theme.size <- (14/5) * geom.text.size
+
+
 schematic <- ggplot() +
   # Overall theme:
   theme_bw() +
   theme(
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
+    axis.text = element_text(size = theme.size),
+    axis.title = element_text(size = theme.size)
   ) +
   labs(
     x = "Day of year",
-    y = "Diameter at breast height (in cm)",
-    title = "Observed diameters (points) and LG5 fitted diameters (solid line)"
[…]

### 2020-11-16 RScripts/additional_figures_HarvardForest.R

@@ -4,152 +4,6 @@ library(tidybayes)
 
 # Figure numbers from: https://github.com/SCBI-ForestGEO/growth_phenology/issues/13
 unique(Wood_pheno_table_HarvardForest_V2$sp)
-# Figure 1: Logistic growth curve and parameter illustration -------------------
-# Set parameters from McMahon & Parker (2014)
-# 1. L: lower asymptotes
-# 2. K: upper asymptotes
-# 3. doy.ip: inflection point
-# 4. r: slope of curve at inflection point
-# 5. theta: theta = 1 indicates pre/post inflection post symmetry
-K <- 13.2
-L <- 13.8
-doy.ip <- 175
-r <- 0.05
-theta <- 1
-params <- c(K, L, doy.ip, r, theta)
-lg5
-get.lg5.ML <- function(params, doy, dbh, resid.sd) {
-  pred.dbh <- lg5.pred(params, doy)
-  pred.ML <-  -sum(dnorm(dbh, pred.dbh, resid.sd, log = T))
-  return(pred.ML)
-}
-
-# Function for logistic growth model written by Sean. Supposed to be in RDendrom
-# package https://github.com/seanmcm/RDendrom, but function does not seem to be included
-lg5.pred <- function(params, doy) {
-  L <- params[1] # min(dbh, na.rm = T)
-  K <- params[2]
-  doy.ip <- params[3]
-  r <- params[4]
-  theta <- params[5]
-  dbh <- vector(length = length(doy))
-  dbh <- L + ((K - L) / (1 + 1 / theta * exp(-(r * (doy - doy.ip) / theta))^theta))
-  return(dbh)
-}
-
-# Compute true values to plot true growth curve
-true_values <- tibble(
-  doy = seq(from = 100, to = 300),
-  diameter = lg5.pred(params, doy)
-)
-
-# Max Growth rate
-intercept <- lg5.pred(params, doy.ip) - r *doy.ip
-
-r_true <- (-1)*((K-L)*r/theta )/(1 + 1/theta)^2
-intercept_true <- lg5.pred(params, doy.ip) - r_true *doy.ip
-
-offset <- 40
-max_growth_rate <- tibble(
-  doy = c(doy.ip - offset, doy.ip + offset),
-) %>%
-  mutate(
-    diameter = doy * r_true + intercept_true
-  )
-
-
-
-
-
-# Compute observed values (noise/error added) to plot with points
-set.seed(76)
-observed_values <- tibble(
[…]

### 2020-11-16 RScripts/additional_figures.R
minor tweaks to schematic #25

@@ -105,7 +105,7 @@ create_growth_polygon <- function(doy_diameter_quartile, percent){
 
 
 # Output figure
-geom.text.size <- 3.5
+geom.text.size <- 4
 theme.size <- (14/5) * geom.text.size
 
 
@@ -119,10 +119,10 @@ schematic <- ggplot() +
     axis.title = element_text(size = theme.size)
   ) +
   labs(
-    x = "Day of year",
+    x = "Day of year (1 to 365)",
     y = "Diameter at breast height"
   ) +
-  coord_cartesian(xlim = c(1, 300))+
+  coord_cartesian(xlim = c(30, 280))+
   # Mark DOY's on x-axis:
   geom_vline(data = doy_diameter_quartile, aes(xintercept = doy), linetype = "dashed", show.legend = FALSE, col = "grey") +
   scale_x_continuous(
@@ -149,7 +149,7 @@ schematic <- ggplot() +
     "text",
     x = doy.ip,
     y = lg5.pred(params, doy.ip) + 0.025,
-    label = expression(paste(g[max], " = slope of tangent line at ", DOY[ip], "  ")),
+    label = expression(paste(g[max], " = slope of tangent line  ")),
     hjust = 1,
     size = geom.text.size
   ) +

### 2020-11-16 RScripts/additional_figures.R
restored deleted code

@@ -210,6 +210,246 @@ ggsave("doc/manuscript/tables_figures/schematic.png", plot = schematic, width = 
 
 
 
+# Growth_Cruves_all: Percent modeled growth and cummulative percent modeled growth ----
+Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V13CLEAN.csv") %>%
+  # Keep only RP and DP for now
+  filter(wood_type != "other") %>%
+  # Rename ring porous to not have a space
+  mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
+
+Wood_pheno_table_hf <- read_csv("Data/Wood_pheno_table_HarvardForest_V9CLEAN.csv") %>%
+  # Keep only RP and DP for now
+  filter(wood_type != "other") %>%
+  # Rename ring porous to not have a space
+  mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
+
+
+# As generated in Rscripts/SCBI_wood_phenology.R
+LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V13CLEAN.csv")
+LG5_parameter_values_hf <- read_csv("Data/LG5_parameter_values_HarvardForest_V9CLEAN.csv")
+
+# Generalized 5-parameter logistic function (modified version of Sean's function)
+lg5 <- function(L, K, doy_ip, r, theta, doy) {
+  # For specified 5 parameters and x = doy, compute y = dbh
+  dbh <- L + ((K - L) / (1 + 1/theta * exp(-(r * (doy - doy_ip) / theta)) ^ theta))
+  return(dbh)
+}
+Wood_pheno_table_scbi <- Wood_pheno_table_scbi[,-12]
+LG5_parameter_values_scbi <- LG5_parameter_values_scbi[,-11]
+percent_growth_scbi <-Wood_pheno_table_scbi %>%
+  mutate(
+    tag = as.numeric(tag),
+    tag_year = str_c(tag,year)
+  ) %>%
+  select(tag, year, tag_year, wood_type) %>%
+  distinct() %>%
+  left_join(LG5_parameter_values_scbi, by = c("tag", "year")) %>%
+  group_by(tag, year) %>%
+  mutate(doy = list(seq(from = 1, to = 365, by = 1))) %>%
+  unnest_longer(doy) %>%
+  #mutate(
+  #  tag_year = tag_year.x
+  #) %>%
[…]

### 2020-11-16 doc/manuscript/Dow_manuscript.Rmd
add Neil's references
updated name and affiliation
updated hypothesis table, knit in updated figures

@@
 
-Albert Kim^1^
+Albert Y. Kim^1,2^
 
@@
 1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA
+
+2. Statistical & Data Sciences; Smith College; Northampton, MA 01063, USA
 
@@
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming.png")
+knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
 ```
@@
 
+From Neil: @zhang_drought_2021, @zhang_extended_2020
+
+effect of spring temperature on annual growth in New England: @alexander_potential_2019
+
 results align with @fu_maximum_2019, which shows that interannual variation in NEE is more strongly linked to peak C uptake rates than to growing season length.

### 2020-11-16 doc/manuscript/Dow_manuscript.Rmd
DOY_gmax --> DOY_ip

@@
 `r pheno_Tsensitivity`, `r DOYtiming`).
-Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ all decreased significantly with increasing $T_{max}$ during the critical period (Table `r mean_growth_parameters_table`).
+Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{ip}$ all decreased significantly with increasing $T_{max}$ during the critical period (Table `r mean_growth_parameters_table`).
 

### 2020-11-16 doc/LG5_errata/LG5_errata_updated.Rmd
Proof of error with doy_ip

--- 2020-10-26 doc/LG5_errata/LG5_errata.Rmd
@@
 ---
-title: "LG5 Function Errata"
+title: "LG5 Function Errata Updated"
 author: "Albert Y. Kim"
@@
 
-# $r$ parameter from LG5 model
+# $doy_{ip}$ and $r$ parameter from LG5 model
 
-I have a suspicion both these interpretations of the $r$ parameter of LG5 model are incorrect:
+I have a suspicion both these interpretations of the $r$ and $doy_{ip}$ parameters of LG5 model are incorrect:
 
-1. McMahon and Parker 2015 state just above Eq 1 (page 245): "the rate parameter $r$ describes the slope of the curve at the inflection point"
-1. D'Orangeville et al. state (page 9): "$r$ represents the maximum growth rate" and "$doy_{ip}$ is the day of year when maximum growth rate occurs"
+1.
+McMahon and Parker 2015 state just above Eq 1 (page 245): "$doy_{ip}$ marks the day of the year when the inflection of the curve is predicted to occur, and the rate parameter $r$ describes the slope of the curve at the inflection point"
+1.
+D'Orangeville et al.
+state (page 9): "$r$ represents the maximum growth rate" and "$doy_{ip}$ is the day of year when maximum growth rate occurs" and "$doy_{ip}$ is the day of year (doy) when maximum growth rate occurs..."
 
-Rather, I believe that the correct formulation of the slope/the maximum growth rate is instead:
+Rather, I believe that the correct formulations of (1) the slope/the maximum growth rate and (2) inflection points are instead:
 
 $$
-r_{new} = \frac{K-L}{\left(1 + \frac{1}{\theta}\right)^2}\frac{r}{\theta}
+\begin{aligned}
+doy_{ip}^* &= -\frac{\log(\theta)}{r} + doy_{ip}\\
+r^* &= \frac{r(K-L)}{4}
+\end{aligned}
 $$
@@
 
-Let's set arbitrarily chosen parameter values then
+Let's set arbitrarily chosen parameter values **but making sure to use a value other than $\theta$ = 1**:
 
@@
 r <- 0.05
-theta <- 1
+# Use a value other than theta = 1:
+theta <- 1.5
 ```
 
-We then plot both the diameter curve as well as the growth curve, with `doy.ip` marked with a vertical dotted line. Observe how growth peaks at `doy.ip`.
+## Original interpretations of $doy_{ip}$
+
+We then plot both the diameter curve as well as the growth curve, with `doy.ip` marked with a vertical dotted line.
 
@@
 ```
+
+Observe how `doy.ip` is not the day the peak growth occurs.
 
@@
 1. In the left plot how the orange line does not appear to be tangent to the LG5 curve at `doy.ip` and hence is not the slope at that point.
-1. In the right plot $r$ = `r r` does not correspond to the growth rate at `doy.ip` which should be the max growth rate. Rather it appears to be around 0.012
+1. In the right plot $r$ = `r r` does not correspond to the growth rate at the peak of the curve, rather it appears to be around 0.012.
 
-## Proposed new $r$
+## Proposed $doy_{ip}^*$ and $r^*$
 
-Let's repeat the above with our proposed growth rate $r_{new}$, which is $\frac{d}{dx}LG5(x)$ evaluated at $x$ = `doy.ip`.
-See this [PDF](https://drive.google.com/file/d/1CMT2o1svTfbS7wQ_uJoWjqHbvDopkCZw/view?usp=sharing) for the analytic derivation of the derivative.
+Let's repeat the above with our proposed $doy_{ip}^*$, which is the point $x$ at which $\frac{d^2}{dx^2}LG5(x) = 0$ (per calculus on inflection points), and our proposed new growth rate $r^*$, which is $\frac{d}{dx}LG5(x)$ evaluated at $x$ = $doy_{ip}^*$.
+See [this](https://drive.google.com/file/d/1CMT2o1svTfbS7wQ_uJoWjqHbvDopkCZw/view?usp=sharing) and [this](https://drive.google.com/file/d/10MjbohLvOvLxkBpNmC_bnJKnLiE0K-pr/view?usp=sharing) PDF for all analytic derivative derivations.
 
 $$
-r_{new} = \frac{K-L}{\left(1 + \frac{1}{\theta}\right)^2}\frac{r}{\theta}
+\begin{aligned}
+r^* &= \frac{r(K-L)}{4}\\
+doy_{ip}^* &= -\frac{\log(\theta)}{r} + doy_{ip}
+\end{aligned}
 $$
@@
 ```{r, fig.width=10}
-# Comupte new r, then find intercept
-r_new <- ((K-L)/(1 + 1/theta)^2) * (r/theta)
-intercept_new <- y - r_new*doy.ip
+# Comupte new r and doy.ip, then find intercept
+r_new <- r*(K-L)/4
+doy.ip_new <- -log(theta)/r + doy.ip
+y <- lg5.pred(params, doy.ip_new)
+intercept_new <- y - r_new*doy.ip_new
 
@@
      main = "LG5 diameter", xlab = "doy", ylab = "diameter")
-abline(v = doy.ip, lty = "dotted")
+abline(v = doy.ip_new, lty = "dotted")
 abline(intercept_new, r_new, col = "orange")
@@
      main = "Growth curve", xlab = "doy", ylab = "growth")
-abline(v = doy.ip, lty = "dotted")
+abline(v = doy.ip_new, lty = "dotted")
 abline(h = r_new, col = "orange")
@@
 1. In the left plot how the orange line is now tangent to the LG5 curve at `doy.ip` and thus is the correct slope.
-1. In the right plot $r_{new}$ = `r r_new` now corresponds to the max growth rate occurring at `doy.ip`.
+1. In the right plot $r^*$ = `r r_new` now corresponds to the max growth rate correctly occurring at `doy.ip`.
+
+# Previous errata
+
+In the previous errata PDF `LG5_errata.pdf` I suggested that the correct max growth rate was
+
+$$
+r_{new} = \frac{K-L}{\left(1 + \frac{1}{\theta}\right)^2}\frac{r}{\theta}
+$$
+
+However, this result was based on the incorrect assumption that $doy_{ip}$ was the correct inflection point, which we evaluated $\frac{d}{dx}LG5(x)$ to obtain the max growth rate.
+Furthermore, the empirical simulations/plots I showed were for $\theta = 1$.
+If we plug this value into the previous $r_{new}$ we get
+
+$$
+r_{new} = \frac{K-L}{\left(1 + \frac{1}{\theta}\right)^2}\frac{r}{\theta} = \frac{K-L}{\left(1 + \frac{1}{1}\right)^2}\frac{r}{1} = \frac{r(K-L)}{4}
+$$
+which is our updated formulation above.

### 2020-11-16 RScripts/additional_figures.R
plotting updated doy.ip and r

@@ -26,10 +26,6 @@ end_doy <- 365
 window_open <- 60
 window_close <- 105
 
-critical_temp_window <- tibble(
-  x = c(window_open, window_close, window_close, window_open, window_open),
-  y = c()
-)
 
 # Function for logistic growth model written by Sean. Supposed to be in RDendrom
 # package https://github.com/seanmcm/RDendrom, but function does not seem to be included
@@ -56,9 +52,19 @@ true_values <- tibble(
 # Incorrect intercept since r is not true slope at inflection point:
 # intercept <- lg5.pred(params, doy.ip) - r *doy.ip
 
-# Correct slope and intercept
+# Version 1 of correct slope and intercept
 r_true <- ((K-L)*r/theta )/(1 + 1/theta)^2
 intercept_true <- lg5.pred(params, doy.ip) - r_true *doy.ip
+
+# Version 2 of correct doy/ip, slope, and intercept
+doy.ip_true <- -log(theta)/r + doy.ip
+r_true_true <- r*(K-L)/4
+intercept_true_true <- lg5.pred(params, doy.ip_true) - r_true_true *doy.ip_true
+
+ggplot(true_values, aes(x = doy, y = diff)) +
+  geom_line() +
+  geom_vline(xintercept = doy.ip_true) +
+  geom_hline(yintercept = r_true_true)
 
 
 # 3. Observed values (noise/error added) (points)

### 2020-11-17 doc/manuscript/Dow_manuscript.Rmd
work on results

@@
 Across all trees and years at SCBI, 
-25% of annual growth was achieved on **May 9** ($DOY_{25}$=**129**), 
-50% on **June 6** ($DOY_{50}$=**157**), 
-and 75% on **July 3** ($DOY_{75}$=**184**), 
-for an average peak growing season length of **56** days. 
-Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 7** ($DOY_{max}$=**158**). 
+25% of annual growth was achieved on **May 20** ($DOY_{25}$=**139**), 
+50% on **June 12** ($DOY_{50}$=**162**), 
+and 75% on **July 5** ($DOY_{75}$=**185**), 
+for an average peak growing season length of **45** days. 
+Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 13** ($DOY_{max}$=**163**). 
 At Harvard Forest,
-25% of annual growth was achieved on **XX** ($DOY_{25}$=**XX**), 
-50% on **XX** ($DOY_{50}$=**XX**), 
-and 75% on **XX** ($DOY_{75}$=**XX**), 
-for an average peak growing season length of **XX** days. 
-Maximum growth rate occurred on **XX** ($DOY_{max}$=**XX**).
+25% of annual growth was achieved on **June 1** ($DOY_{25}$=**151**), 
+50% on **June 23** ($DOY_{50}$=**173**), 
+and 75% on **July 14** ($DOY_{75}$=**194**), 
+for an average peak growing season length of **43** days. 
+Maximum growth rate occurred on **June 24** ($DOY_{max}$=**174**).
 
@@
 
-~~For ring-porous trees, on average, 25% of annual growth was achieved on **April 23** ($DOY_{25}$=**113**), 50% on **May 27** ($DOY_{50}$=**147**), and 75% on **June 30** ($DOY_{75}$=**181**), for an average peak growing season length of **68** days.
-Maximum growth rate occurred *on $DOY_{50}$* and averaged .048 **mm day^-1^ (?)**.~~
-~~For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 2** ($DOY_{25}$=**153**), 50% on **June 20** ($DOY_{50}$=**171**), and 75% on **July 9** ($DOY_{75}$=**190**), for an average peak growing season length of **37** days.
-Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 21** ($DOY_{max}$=**172**) and averaged .068 **mm day^-1^ (?)**.~~
+~~For ring-porous trees, on average, 25% of annual growth was achieved on **May 4** ($DOY_{25}$=**123**) at SCBI and **May 15** ($DOY_{25}$=**132**) at Harvard FOrest, 50% on **June 2** ($DOY_{50}$=**152**) at SCBI and **June 9** ($DOY_{50}$=**159**) at Harvard Forest, and 75% on **June 30** ($DOY_{75}$=**180**) at SCBI and **July 6** ($DOY_{75}$=**186**) at Harvard Forest, for an average peak growing season length of **56** days at SCBI and **54** days at Harvard Forest.
+Maximum growth rate occurred *on $DOY_{50}$* at SCBI and *June 11* ($DOY_{max}$=**161**) at Harvard Forest and averaged .046 and .030 **mm day^-1^ (?)** respectively.~~
+
+~~For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 4** ($DOY_{25}$=**154**) at SCBI and **June 14** ($DOY_{25}$=**164**) at Harvard Forest, 50% on **June 22** ($DOY_{50}$=**172**) at SCBI and **July 2** ($DOY_{50}$=**182**) at Harvard Forest, and 75% on **July 10** ($DOY_{75}$=**190**) and **July 19** ($DOY_{75}$=**199**) at Harvard forest, for an average peak growing season length of **36** days at SCBI and **35** days at Harvard Forest.
+Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 23** ($DOY_{max}$=**173**) at SCBI and **just one day after $DOY_{50}$** on **July 3** ($DOY_{max}$=**183**) at Harvard Forest and averaged .061 and .025 **mm day^-1^ (?)** respectively.~~
 
@@
 
-For the majority of weekly time windows from Feb.
+For the majority of weekly time windows from Jan.
 1 to $DOY_{25}$, there was negative relationship between temperature and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
 `r climwin`, `r climwin_rp_DOY25_SCBI`, `r climwin_dp_DOY25_SCBI`).
-This held true for both ring- and diffuse-porous trees, *albeit* with stronger relationships for ring-porous trees. 
+This held true for both ring- and diffuse-porous trees at both sites, *albeit* with stronger relationships for diffuse-porous trees. 
 *(Describe the landscape:)*
-In general, there were significant negative relationships between temperature and $DOY_{25}$ for any time period beginning $\ge$ **5** weeks prior to $DOY_{25}$, ending  $\le$ **6** weeks prior, and covering at least a **3**-week period.
+In general, there were significant negative relationships between temperature and $DOY_{25}$ for any time period beginning $\ge$ **5** weeks prior to $DOY_{25}$, ending  $\le$ **8-20** weeks prior depending on wood-type and site, and covering at least a **3**-week period.
 Relationships tended to be weaker and more stochastic in sign for short time periods, particularly those just prior to $DOY_{25}$.
 
-Optimal windows occurred from **##-##** for ring-porous species and from **##-##** for diffuse-porous species.
+Optimal windows occurred from **3/22-5/7** for ring-porous species and from **2/19-5/21** for diffuse-porous species.
 
@@
 `r pheno_Tsensitivity`, `r DOYtiming`).
-Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{ip}$ all decreased significantly with increasing $T_{max}$ during the critical period (Table `r mean_growth_parameters_table`).
+Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical period (Table `r mean_growth_parameters_table`).
 
@@
 
-Because both $DOY_{25}$ and $DOY_{75}$ both shifted forward, peak growing season length was not consistently increased by warmer early spring temperatures (Tables `r hypothesis_table`, Figs.
+At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted back while ${DOY_75}$ did not shift, resulting in an extended peak growing season length when critical window temperatures were warmer (Tables `r hypothesis_table`, Figs.
 `r pheno_Tsensitivity`, `r DOYtiming`).
-At both sites, $DOY_{75}-DOY_{25}$ decreased with temperature for ring-porous species 
-and was neutral (SCBI) or positive (Harvard Forest) for diffuse-porous species.
+Diffuse-porous $DOY_{25}$, $DOY_{50}$, and ${DOY_75}$ shifted back uniformly resulting in no significant change in peak growing season length.
+At Harvard Forest, ring-porous ${DOY_75}$ shifted back faster than ${DOY_25}$, resulting in shorter peak growins season length, while diffuse-porous ${DOY_25}$ shifted back faster than ${DOY_75}$, resulting in an extended peak growing season length in response to warmer critical window temperatures.
 
@@
 Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer early spring temperatures.
-In contrast, diffuse-porous species exhibited lower $g_{max}$ in years with higher early spring $T_{max}$.
+In contrast, diffuse-porous species exhibited no change (SCBI) or lower $g_{max}$ in years with higher early spring $T_{max}$.
 
-Total annual growth rate, $\Delta DBH$, also was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`).
+Total annual growth, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes (Harvard Forest) (Tables `r hypothesis_table`, Fig.
+`r pheno_Tsensitivity`).
 

### 2020-11-18 RScripts/data_cleaning.R
Create cleaning figures

@@ -85,6 +85,9 @@ year <- subset(year, doy <= upper & doy >= lower) #Remove models where peak grow
 goodtags <- unique(year$tag_year)
 keeptags <- append(keeptags, goodtags)
 }
+percent_growth_RP_gone <- percent_growth_RP[!(percent_growth_RP$tag_year %in% keeptags),]
+SCBI_badpeak_RP <- unique(percent_growth_RP_gone$tag_year)
+
 percent_growth_RP <- percent_growth_RP[percent_growth_RP$tag_year %in% keeptags,]
 #ENTRIES REMOVED IN THIS STEP
 #percent_growth_badpeak <- percent_growth_RP[!(percent_growth_RP$tag_year %in% ratedoysubset$tag_year),]
@@ -105,8 +108,8 @@ maxgrowthrate <- aggregate(percent_growth_DP$dbh_growth_percent, by = list(perce
 names(maxgrowthrate) <- c("tag_year", "rate")#Maximum growth rate of each tree
 high_growthday <- subset(maxgrowthrate, rate >= (mean(maxgrowthrate$rate)+(sd(maxgrowthrate$rate)*sd)) & rate <= 0) #Remove trees where one day had higher than 4% of total growth
 #ENTRIES REMOVED IN THIS STEP
-#percent_growth_high <- percent_growth_DP[percent_growth_DP$tag_year %in% high_growthday$tag_year,]
-#SCBI_highgrowth_DP <- unique(percent_growth_high$tag_year)
+percent_growth_high <- percent_growth_DP[percent_growth_DP$tag_year %in% high_growthday$tag_year,]
+SCBI_highgrowth_DP <- unique(percent_growth_high$tag_year)
 #wood_pheno_removed <- Wood_pheno_table_scbi[Wood_pheno_table_scbi$tag_year %in% SCBI_highgrowth_DP,]
 #wood_pheno_removed <- subset(wood_pheno_removed, perc == .25)
 #count(wood_pheno_removed,sp)
@@ -127,6 +130,9 @@ for(i in 2011:2020){
   goodtags <- unique(year$tag_year)
   keeptags <- append(keeptags, goodtags)
 }
+percent_growth_DP_gone <- percent_growth_DP[!(percent_growth_DP$tag_year %in% keeptags),]
+SCBI_badpeak_DP <- unique(percent_growth_DP_gone$tag_year)
+
[…]

### 2020-11-18 doc/manuscript/Dow_manuscript.Rmd
add paragraph on data cleaning

@@
 
-(*The following seems out of date, right? Scripted cleaning is not described.*)
-The dendroband data were manually cleaned by visual inspection.
+(*The following seems out of date, right? Scripted cleaning is not described.* -- *Manually cleaning was done for pre-analysis data. Added some text to clarify.*)
+The dendroband data were manually cleaned by visual inspection before being used for modeling.
 Three classes of mistakes: 1.
@@
 
-After model fitting, we used  a script to remove outliers....
+After model fitting, we used a script to remove poorly fit models.
+Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the acceptable range we determined.
+Modeled tag years were removed if: 1.
+single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, HF = 34) 2.
+DOY where peak growth rate occured was 2 standard deviations away, calculated for each wood-type in each year (aimed to remove models which predicted growth to peak during the winter; SCBI = 62, HF = 106) 3.
+tag years with small or negligible total growth (deltaDBH < 0.02 mm; SCBI = 0, HF = 66), 4.
+models which predicted total yearly growth to take longer than 365 days (SCBI = 181, HF = 199) 5.
+models with unexplained spikes (SCBI = 0, HF = 3).
+At Harvard Forest the tag years removed through this method was proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others *(SI figure?
+data_cleaning_figure_hf in manuscript folder)*.
+At SCBI a higher proportion of oak trees were removed *(data_cleaning_figure_scbi)*.
+The majority (White oak = 91, Red oak = 79) were removed in step 5 above, indicating that the LG5 model struggled to accurately model these these species.
 
@@
 
-*(Numbers in bold will change)*
+*(Numbers in bold will change, updated; 11/18. Should be final results)*
 
@@
 
-~~For ring-porous trees, on average, 25% of annual growth was achieved on **May 4** ($DOY_{25}$=**123**) at SCBI and **May 15** ($DOY_{25}$=**132**) at Harvard FOrest, 50% on **June 2** ($DOY_{50}$=**152**) at SCBI and **June 9** ($DOY_{50}$=**159**) at Harvard Forest, and 75% on **June 30** ($DOY_{75}$=**180**) at SCBI and **July 6** ($DOY_{75}$=**186**) at Harvard Forest, for an average peak growing season length of **56** days at SCBI and **54** days at Harvard Forest.
-Maximum growth rate occurred *on $DOY_{50}$* at SCBI and *June 11* ($DOY_{max}$=**161**) at Harvard Forest and averaged .046 and .030 **mm day^-1^ (?)** respectively.~~
+For ring-porous trees, on average, 25% of annual growth was achieved on **May 4** ($DOY_{25}$=**123**) at SCBI and **May 15** ($DOY_{25}$=**132**) at Harvard FOrest, 50% on **June 2** ($DOY_{50}$=**152**) at SCBI and **June 9** ($DOY_{50}$=**159**) at Harvard Forest, and 75% on **June 30** ($DOY_{75}$=**180**) at SCBI and **July 6** ($DOY_{75}$=**186**) at Harvard Forest, for an average peak growing season length of **56** days at SCBI and **54** days at Harvard Forest.
+Maximum growth rate occurred *on $DOY_{50}$* at SCBI and *June 11* ($DOY_{max}$=**161**) at Harvard Forest and averaged .046 and .030 **mm day^-1^ (?)** respectively.
 
-~~For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 4** ($DOY_{25}$=**154**) at SCBI and **June 14** ($DOY_{25}$=**164**) at Harvard Forest, 50% on **June 22** ($DOY_{50}$=**172**) at SCBI and **July 2** ($DOY_{50}$=**182**) at Harvard Forest, and 75% on **July 10** ($DOY_{75}$=**190**) and **July 19** ($DOY_{75}$=**199**) at Harvard forest, for an average peak growing season length of **36** days at SCBI and **35** days at Harvard Forest.
-Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 23** ($DOY_{max}$=**173**) at SCBI and **just one day after $DOY_{50}$** on **July 3** ($DOY_{max}$=**183**) at Harvard Forest and averaged .061 and .025 **mm day^-1^ (?)** respectively.~~
+For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 4** ($DOY_{25}$=**154**) at SCBI and **June 14** ($DOY_{25}$=**164**) at Harvard Forest, 50% on **June 22** ($DOY_{50}$=**172**) at SCBI and **July 2** ($DOY_{50}$=**182**) at Harvard Forest, and 75% on **July 10** ($DOY_{75}$=**190**) and **July 19** ($DOY_{75}$=**199**) at Harvard forest, for an average peak growing season length of **36** days at SCBI and **35** days at Harvard Forest.
+Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 23** ($DOY_{max}$=**173**) at SCBI and **just one day after $DOY_{50}$** on **July 3** ($DOY_{max}$=**183**) at Harvard Forest and averaged .061 and .025 **mm day^-1^ (?)** respectively.
 

### 2020-11-19 RScripts/dendroband_functions.R
DRY: dendroband functions

@@ -1,0 +1,339 @@
+# 2. Load necessary functions ----------------------------------------------------
+# convert measure to DBH (Use function in Dendrobands/Rscripts/analysis/convert_caliper_meas_to_dbh.R (I pasted it below for now))
+
+# Caliper to DBH function by Ian##
+objectiveFuncDendro <- function(diameter2, diameter1, gap1, gap2) {
+  if (gap1 > diameter1) {
+    return(20)
+  }
+  if (gap2 > diameter2) {
+    return(20)
+  }
+
+  delta <- abs(diameter1 - diameter2 + (1 / pi) * diameter2 * asin(gap2 / diameter2) - (1 / pi) * diameter1 * asin(gap1 / diameter1))
+
+  return(return(delta))
+}
+
+findOneDendroDBH <- function(dbh1, m1, m2, func = objectiveFuncDendro) {
+  if (is.na(dbh1) | is.na(m1) | is.na(m2) | dbh1 <= 0) {
+    return(NA)
+  }
+
+  if (m2 > 0) {
+    upper <- dbh1 + m2
+  } else {
+    upper <- dbh1 + 1
+  }
+  if (m2 < m1) {
+    lower <- 0
+  } else {
+    lower <- dbh1
+  }
+
+  result <- optimize(f = func, interval = c(lower, upper), diameter1 = dbh1, gap1 = m1, gap2 = m2)
+  return(result$minimum)
+}
+
+findDendroDBH <- function(dbh1, m1, m2, func = objectiveFuncDendro) {
+  records <- max(length(dbh1), length(m1), length(m2))
+
+  if (length(dbh1) == 1) dbh1 <- rep(dbh1, records)
+  if (length(m1) == 1) m1 <- rep(m1, records)
+  if (length(m2) == 1) m2 <- rep(m2, records)
+
+  dbh2 <- numeric()
+  for (i in 1:records) dbh2[i] <- findOneDendroDBH(dbh1[i], m1[i], m2[i], func)
+  return(dbh2)
+}
+
+# Sean's functions -- I think they are supposed to be in the rdendrom package but I couldn't the package to work correctly.
+### code chunk number 2: lg5.functions
+# This function predicts a diameter given the day of the year and a vector of parameters for the lg5 model.
+# It is called by the lg5.ss and lg5.plot functions.
+
[…]

### 2020-11-19 RScripts/HF_wood_phenology.R
DRY: dendroband functions

@@ -1,3 +1,9 @@
+# Setup ------------------------------------------------------------------------
+#
+# To get a quick overview of the sections of this code, go to RStudio menu bar ->
+# Edit -> Folding -> Collapse all.
+#
+
 library(data.table)
 library(plyr)
 library(dplyr)
@@ -9,7 +15,13 @@ library(lubridate)
 library(RColorBrewer)
 library(MASS)
 
-# 1. Format dendroband data ----------------------------------------------------
+# Load Sean McMahon's Dendroband functions
+source("RScripts/dendroband_functions.R")
+
+
+
+
+# Format dendroband data ----------------------------------------------------
 all_stems <- read_csv("data/HarvardDendroband_cleaned.csv")
 names(all_stems) <- c("plot", "tag", "sp", "date","dbh2")
 all_stems$DOY <- yday(strptime(all_stems$date, format = "%m/%d/%y"))
@@ -20,298 +32,6 @@ all_stems$month <- as.numeric(format(all_stems$date, format = "%m"))
 all_stems$day <- as.numeric(format(all_stems$date, format = "%d"))
 
 
-# 2. Load necessary functions ----------------------------------------------------
-# Sean's functions -- I think they are supposed to be in the rdendrom package but I couldn't the package to work correctly.
-### code chunk number 2: lg5.functions
-# This function predicts a diameter given the day of the year and a vector of parameters for the lg5 model.
-# It is called by the lg5.ss and lg5.plot functions.
-
-lg5.pred <- function(params, doy) {
-  L <- params[1] # min(dbh, na.rm = T)
-  K <- params[2]
-  doy.ip <- params[3]
-  r <- params[4]
-  theta <- params[5]
-  dbh <- vector(length = length(doy))
-  dbh <- L + ((K - L) / (1 + 1 / theta * exp(-(r * (doy - doy.ip) / theta))^theta))
-  return(dbh)
-}
-
-lg5.ML <- function(params, doy, dbh, resid.sd) {
-  pred.dbh <- lg5.pred(params, doy)
[…]

### 2020-11-19 RScripts/SCBI_wood_phenology.R

@@ -1,14 +1,9 @@
-# Notes for collaboration #
-# 1.) Much of the code behind a '#' is code made obsolete after I found a more efficient way of doing things
-# 2.) To begin, you'll need to set your working directory to your local github dendroband repo.
-# 3.) I am not entirely convinced that this is returning the correct values for the variables i'm asking it for
-# so take everything with a grain of salt
-# 4.) I found it was cleaner to capture the variables I am interested in by doing the trees one at a time.
-# Of you want to get the parameters of each measurement we can create a second script to do them all at once.
+# Setup ------------------------------------------------------------------------
+#
+# To get a quick overview of the sections of this code, go to RStudio menu bar ->
+# Edit -> Folding -> Collapse all.
+#
 
-
-
-# 0. Setup ----------------------------------------------------
 library(data.table)
 library(plyr)
 library(dplyr)
@@ -20,9 +15,13 @@ library(lubridate)
 library(RColorBrewer)
 library(MASS)
 
+# Load Sean McMahon's Dendroband functions
+source("RScripts/dendroband_functions.R")
 
 
-# 1. Format dendroband data ----------------------------------------------------
+
+
+# Format dendroband data ----------------------------------------------------
 # files <- dir("data", pattern="_201[0-9]*.csv")
 files <- dir("data", pattern = "_20[1-2][0-9]*.csv")
 dates <- c(2011:2020)
@@ -67,346 +66,9 @@ make_growth_list(files, dates)
 
 
 
-# 2. Load necessary functions ----------------------------------------------------
-# convert measure to DBH (Use function in Dendrobands/Rscripts/analysis/convert_caliper_meas_to_dbh.R (I pasted it below for now))
-# Caliper to DBH function by Ian##
-objectiveFuncDendro <- function(diameter2, diameter1, gap1, gap2) {
-  if (gap1 > diameter1) {
-    return(20)
-  }
-  if (gap2 > diameter2) {
[…]

### 2020-11-19 RScripts/dendroband_functions.R

@@ -1,0 +1,339 @@
+# Sean McMahon's dendrobands functions. From: https://github.com/seanmcm/RDendrom/
+
+# convert measure to DBH (Use function in Dendrobands/Rscripts/analysis/convert_caliper_meas_to_dbh.R (I pasted it below for now))
+# Caliper to DBH function by Ian##
+objectiveFuncDendro <- function(diameter2, diameter1, gap1, gap2) {
+  if (gap1 > diameter1) {
+    return(20)
+  }
+  if (gap2 > diameter2) {
+    return(20)
+  }
+
+  delta <- abs(diameter1 - diameter2 + (1 / pi) * diameter2 * asin(gap2 / diameter2) - (1 / pi) * diameter1 * asin(gap1 / diameter1))
+
+  return(return(delta))
+}
+
+findOneDendroDBH <- function(dbh1, m1, m2, func = objectiveFuncDendro) {
+  if (is.na(dbh1) | is.na(m1) | is.na(m2) | dbh1 <= 0) {
+    return(NA)
+  }
+
+  if (m2 > 0) {
+    upper <- dbh1 + m2
+  } else {
+    upper <- dbh1 + 1
+  }
+  if (m2 < m1) {
+    lower <- 0
+  } else {
+    lower <- dbh1
+  }
+
+  result <- optimize(f = func, interval = c(lower, upper), diameter1 = dbh1, gap1 = m1, gap2 = m2)
+  return(result$minimum)
+}
+
+findDendroDBH <- function(dbh1, m1, m2, func = objectiveFuncDendro) {
+  records <- max(length(dbh1), length(m1), length(m2))
+
+  if (length(dbh1) == 1) dbh1 <- rep(dbh1, records)
+  if (length(m1) == 1) m1 <- rep(m1, records)
+  if (length(m2) == 1) m2 <- rep(m2, records)
+
+  dbh2 <- numeric()
+  for (i in 1:records) dbh2[i] <- findOneDendroDBH(dbh1[i], m1[i], m2[i], func)
+  return(dbh2)
+}
+
+# Sean's functions -- I think they are supposed to be in the rdendrom package but I couldn't the package to work correctly.
+### code chunk number 2: lg5.functions
+# This function predicts a diameter given the day of the year and a vector of parameters for the lg5 model.
+# It is called by the lg5.ss and lg5.plot functions.
+
[…]

### 2020-11-19 RScripts/HF_wood_phenology.R
Added description of wood phenology scripts

@@ -1,5 +1,10 @@
 # Setup ------------------------------------------------------------------------
 #
+# This script takes the Harvard Forest dendroband measurements and fits Sean McMahon's LG5
+# dendrometer model for each tree-year. The output is a .csv of RAW fitted
+# parameters and related values (max rate, max rate DOY, etc.), which we will
+# then "clean" for outliers in data_cleaning.R.
+#
 # To get a quick overview of the sections of this code, go to RStudio menu bar ->
 # Edit -> Folding -> Collapse all.
 #

### 2020-11-19 RScripts/SCBI_wood_phenology.R

@@ -1,5 +1,10 @@
 # Setup ------------------------------------------------------------------------
 #
+# This script takes the SCBI dendroband measurements and fits Sean McMahon's LG5
+# dendrometer model for each tree-year. The output is a .csv of RAW fitted
+# parameters and related values (max rate, max rate DOY, etc.), which we will
+# then "clean" for outliers in data_cleaning.R.
+#
 # To get a quick overview of the sections of this code, go to RStudio menu bar ->
 # Edit -> Folding -> Collapse all.
 #

### 2020-11-19 RScripts/HF_wood_phenology.R
Stated purpose of and cleaned data_cleaning.R

@@ -1,9 +1,12 @@
 # Setup ------------------------------------------------------------------------
 #
-# This script takes the Harvard Forest dendroband measurements and fits Sean McMahon's LG5
-# dendrometer model for each tree-year. The output is a .csv of RAW fitted
-# parameters and related values (max rate, max rate DOY, etc.), which we will
-# then "clean" for outliers in data_cleaning.R.
+# This script takes as input the Harvard Forest dendroband measurements and fits Sean McMahon's LG5
+# dendrometer model for each tree-year. The outputs are two .csv files of RAW
+#
+# 1. wood phenology DOY25, DOY50, DOY75 data
+# 2. fitted parameters and related values (max rate, max rate DOY, etc.)
+#
+# which we will then "clean" for outliers in data_cleaning.R.
 #
 # To get a quick overview of the sections of this code, go to RStudio menu bar ->
 # Edit -> Folding -> Collapse all.
@@ -501,13 +504,20 @@ for (q in 1998:2003) {
   }
 }
 
-# Write master data frame
+
+
+
+
+# Write master data frames ----
+## Wood phenology data ----
 warnings()
 masterDF <- masterDF[-1, ]
 unique(masterDF$sp)
 masterDF$wood_type <- ifelse(masterDF$sp == "betual"| masterDF$sp == "acerru" | masterDF$sp == "fagugr" | masterDF$sp == "betule" | masterDF$sp == "betupa" | masterDF$sp == "acerpe" | masterDF$sp == "betupo" | masterDF$sp == "prunse", "diffuse-porous", ifelse(masterDF$sp == "querru" | masterDF$sp == "querve" | masterDF$sp == "fraxam", "ring-porous", "other"))
 write.csv(masterDF, file = "Data/Wood_pheno_table_HarvardForest_V4.csv", row.names = FALSE)
 masterDF$DOY <- as.numeric(masterDF$DOY)
+
+## Fitted parameter data ----
 # Added by bert: save parameter values
 LG5_parameters %>%
   bind_rows() %>%

### 2020-11-19 RScripts/SCBI_wood_phenology.R

@@ -1,9 +1,12 @@
 # Setup ------------------------------------------------------------------------
 #
-# This script takes the SCBI dendroband measurements and fits Sean McMahon's LG5
-# dendrometer model for each tree-year. The output is a .csv of RAW fitted
-# parameters and related values (max rate, max rate DOY, etc.), which we will
-# then "clean" for outliers in data_cleaning.R.
+# This script takes as input the SCBI dendroband measurements and fits Sean McMahon's LG5
+# dendrometer model for each tree-year. The outputs are two .csv files of RAW
+#
+# 1. wood phenology DOY25, DOY50, DOY75 data
+# 2. raw fitted parameters and related values (max rate, max rate DOY, etc.)
+#
+# which we will then "clean" for outliers in data_cleaning.R.
 #
 # To get a quick overview of the sections of this code, go to RStudio menu bar ->
 # Edit -> Folding -> Collapse all.
@@ -668,12 +671,18 @@ for (q in 2020) {
   }
 }
 
-# Write master data frame
+
+
+
+# Write master data frames ----
+## Wood phenology data ----
 warnings()
 masterDF <- masterDF[-1, ]
 masterDF$wood_type <- ifelse(masterDF$sp == "quru" | masterDF$sp == "qual", "ring porous", ifelse(masterDF$sp == "litu" | masterDF$sp == "fagr", "diffuse-porous", "other"))
 write.csv(masterDF, file = "Data/Wood_pheno_table_2020.csv", row.names = FALSE)
 masterDF$DOY <- as.numeric(masterDF$DOY)
+
+## Fitted parameter data ----
 # Added by bert: save parameter values
 LG5_parameters %>%
   bind_rows() %>%

### 2020-11-19 RScripts/data_cleaning.R

@@ -1,9 +1,30 @@
+# Setup ------------------------------------------------------------------------
+#
+# This script takes as inputs the SCBI and Harvard Forest RAW
+#
+# 1. wood phenology DOY25, DOY50, DOY75 data
+# 2. fitted parameters and related values (max rate, max rate DOY, etc.)
+#
+# and removes "outlier" tree-years based on whether their fitted growth curves
+# make sense. These CLEAN data sets are then output for use in our main analysis
+# script Pheno_tsensitivity_figure.R
+#
+# To get a quick overview of the sections of this code, go to RStudio menu bar ->
+# Edit -> Folding -> Collapse all.
+#
 library(tidyverse)
 library(scales)
 library(tidybayes)
-sd = 2
-#SCBI ----
-#Load up percent growth DF----
+
+# Number of standard deviations used to identify "outliers"
+sd <- 2
+
+# Load Sean McMahon's Dendroband functions
+source("RScripts/dendroband_functions.R")
+
+
+# SCBI -------------------------------------------------------------------------
+## Load up percent growth DF ----
 Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_V10RAW.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
@@ -15,12 +36,6 @@ unitag <- unique(Wood_pheno_table_scbi)
 
 LG5_parameter_values_scbi <- read_csv("Data/LG5_parameter_values_V10RAW.csv")
 
-# Generalized 5-parameter logistic function (modified version of Sean's function)
-lg5 <- function(L, K, doy_ip, r, theta, doy) {
-  # For specified 5 parameters and x = doy, compute y = dbh
-  dbh <- L + ((K - L) / (1 + 1/theta * exp(-(r * (doy - doy_ip) / theta)) ^ theta))
-  return(dbh)
-}
 percent_growth <- Wood_pheno_table_scbi %>%
   separate(tag, into = c("tag", "stem"), sep = "_") %>%
   mutate(
@@ -37,195 +52,199 @@ percent_growth <- Wood_pheno_table_scbi %>%
   mutate(
     dbh = lg5(L, K, doy_ip, r, theta, doy),
     dbh_growth = dbh - lag(dbh),
-    dbh_total_growth =  K-L,
[…]

### 2020-11-19 doc/manuscript/Dow_manuscript.Rmd

@@
 
-We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig. `r schematic`). 
-...DETAILS...
+We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig.
+`r schematic`).
+This 5 parameter logistic model was fitted to each tag year's dendroband measurements in R: L + ((K - L) / (1 + 1/theta * exp(-(r * (doy - doy_ip) / theta)) ^ theta))
+Where L and K are lower and upper asymptotes of the model, theta is a tuning parameter controlling the slope of the curve toward the upper asymptote, r is the slope of the curve at the inflection point, doy_ip is the day of year where the inflection point occurs, and doy are the modeled days of year, in this case, 1-365.
+*We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow.* Using the optimized parameters for each tree year, we could then determine the DBH on every day within that year, giving us the ability to extract the intra-annual growth variables we were interested in.
 

### 2020-11-19 RScripts/HF_Climwin_Cameron.R
Went over climwin scripts

@@ -1,25 +1,37 @@
-#Climwin analysis script
+# Setup ------------------------------------------------------------------------
+#
+# This script generates Harvard Forest climwin critical temperature windows 6 different
+# types of outcome variables y and outputs them to csv
+#
+# 1. wood phenology DOY25, DOY50, DOY75 data
+# 2. raw fitted parameters and related values (max rate, max rate DOY, etc.)
+#
+# which we will then "clean" for outliers in data_cleaning.R.
+#
+# To get a quick overview of the sections of this code, go to RStudio menu bar ->
+# Edit -> Folding -> Collapse all.
+#
 library(climwin)
 library(readr)
 library(lubridate)
 library(tidyverse)
 library(dplyr)
-#Merge two weather DF's into one ----
+# Merge two weather DF's into one ----
 HF_1964to2002 <- read_csv("climate data/HF_1964to2002.csv")
 HF_2001toPresent <- read_csv("climate data/HF_2001toPresent.csv")
 
-HF_1964to2002$DOY <- yday(HF_1964to2002$date)# format = "%y-%m-%d")
+HF_1964to2002$DOY <- yday(HF_1964to2002$date) # format = "%y-%m-%d")
 
 HF_1964to2002$year <- as.numeric(format(HF_1964to2002$date, format = "%Y"))
-HF_1998to2002 <- subset(HF_1964to2002, year >=1998)
-HF_1998to2002 <- HF_1998to2002[,c(1,6,10,11)]
+HF_1998to2002 <- subset(HF_1964to2002, year >= 1998)
+HF_1998to2002 <- HF_1998to2002[, c(1, 6, 10, 11)]
 HF_1998to2002$month <- as.numeric(format(HF_1998to2002$date, format = "%m"))
 HF_1998to2002$day <- as.numeric(format(HF_1998to2002$date, format = "%d"))
 
 HF_2001toPresent$DOY <- yday(HF_2001toPresent$date)
 HF_2001toPresent$year <- as.numeric(format(HF_2001toPresent$date, format = "%Y"))
-HF_2001toPresent <- HF_2001toPresent[,c(1,7,47,48)]
-HF_2001toPresent <- HF_2001toPresent[c(-1:-505),]
+HF_2001toPresent <- HF_2001toPresent[, c(1, 7, 47, 48)]
+HF_2001toPresent <- HF_2001toPresent[c(-1:-505), ]
[…]

### 2020-11-19 RScripts/SCBI_Climwin_Cameron.R

@@ -1,479 +1,524 @@
-#Climwin analysis script
+# Setup ------------------------------------------------------------------------
+#
+# This script generates SCBI climwin critical temperature windows 6 different
+# types of outcome variables y and outputs them to csv
+#
+# To get a quick overview of the sections of this code, go to RStudio menu bar ->
+# Edit -> Folding -> Collapse all.
+#
 library(climwin)
 library(readr)
 library(lubridate)
 library(tidyverse)
 
-####Read in climate and biological data for climwin analysis ----
-#SCBI met tower
-#climate <- read_csv("climate data/SCBI_mettower_data_sensor2.csv", col_names = FALSE) #only goes to October 2019, fix?
-climate <- read_csv("climate data/met_tower_data_sensor2_ncdc_supplemented.csv", col_names = TRUE) #only goes to October 2019, fix?
+# Read in climate and biological data for climwin analysis ----
+# SCBI met tower
+# climate <- read_csv("climate data/SCBI_mettower_data_sensor2.csv", col_names = FALSE) #only goes to October 2019, fix?
+climate <- read_csv("climate data/met_tower_data_sensor2_ncdc_supplemented.csv", col_names = TRUE) # only goes to October 2019, fix?
 
-#colnames(climate) <- c("year","month", "day", "prcip", "TMAX", "TMIN")
+# colnames(climate) <- c("year","month", "day", "prcip", "TMAX", "TMIN")
 
-colnames(climate) <- c("date","year", "month", "day", "doy", "TMAX")
-climate[climate$TMAX == -99.9,] <- NA
+colnames(climate) <- c("date", "year", "month", "day", "doy", "TMAX")
+climate[climate$TMAX == -99.9, ] <- NA
 
-climate <- climate[complete.cases(climate$TMAX),]
+climate <- climate[complete.cases(climate$TMAX), ]
 
 climate$DATE <- paste(climate$day, climate$month, climate$year, sep = "/")
 climate$DATE <- strptime(as.character(climate$DATE), format = "%d/%m/%Y")
 climate$DATE <- format(climate$DATE, "%d/%m/%Y")
-climate <- distinct(climate,DATE,.keep_all = TRUE)
-#The data containing the biological responses for testing
[…]

### 2020-11-30 doc/manuscript/Dow_manuscript.Rmd
writing- results
@camerondow35 , note that I deleted a couple paragraphs that presented numbers given in the table.

@@
 
-*(Numbers in bold will change, updated; 11/18. Should be final results)*
+Across all trees and years at SCBI, 
+25% of annual growth was achieved on May 20 ($DOY_{25}$=139), 
+50% on June 12 ($DOY_{50}$=162), 
+and 75% on July 5 ($DOY_{75}$=185), 
+for an average peak growing season length of 45 days. 
+Maximum growth rate occurred just one day after $DOY_{50}$ on June 13 ($DOY_{max}$=163). 
+At Harvard Forest,
+25% of annual growth was achieved on June 1 ($DOY_{25}$=151), 
+50% on June 23 ($DOY_{50}$=173), 
+and 75% on July 14 ($DOY_{75}$=194), 
+for an average peak growing season length of 43 days. 
+Maximum growth rate occurred on June 24 ($DOY_{max}$=174).
 
-Across all trees and years at SCBI, 
-25% of annual growth was achieved on **May 20** ($DOY_{25}$=**139**), 
-50% on **June 12** ($DOY_{50}$=**162**), 
-and 75% on **July 5** ($DOY_{75}$=**185**), 
-for an average peak growing season length of **45** days. 
-Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 13** ($DOY_{max}$=**163**). 
-At Harvard Forest,
-25% of annual growth was achieved on **June 1** ($DOY_{25}$=**151**), 
-50% on **June 23** ($DOY_{50}$=**173**), 
-and 75% on **July 14** ($DOY_{75}$=**194**), 
-for an average peak growing season length of **43** days. 
-Maximum growth rate occurred on **June 24** ($DOY_{max}$=**174**).
-
-At both sites, growth phenology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig.
-`r growth_curves_all`, Table `r mean_growth_parameters_table`).
-
-For ring-porous trees, on average, 25% of annual growth was achieved on **May 4** ($DOY_{25}$=**123**) at SCBI and **May 15** ($DOY_{25}$=**132**) at Harvard FOrest, 50% on **June 2** ($DOY_{50}$=**152**) at SCBI and **June 9** ($DOY_{50}$=**159**) at Harvard Forest, and 75% on **June 30** ($DOY_{75}$=**180**) at SCBI and **July 6** ($DOY_{75}$=**186**) at Harvard Forest, for an average peak growing season length of **56** days at SCBI and **54** days at Harvard Forest.
-Maximum growth rate occurred *on $DOY_{50}$* at SCBI and *June 11* ($DOY_{max}$=**161**) at Harvard Forest and averaged .046 and .030 **mm day^-1^ (?)** respectively.
-
-For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 4** ($DOY_{25}$=**154**) at SCBI and **June 14** ($DOY_{25}$=**164**) at Harvard Forest, 50% on **June 22** ($DOY_{50}$=**172**) at SCBI and **July 2** ($DOY_{50}$=**182**) at Harvard Forest, and 75% on **July 10** ($DOY_{75}$=**190**) and **July 19** ($DOY_{75}$=**199**) at Harvard forest, for an average peak growing season length of **36** days at SCBI and **35** days at Harvard Forest.
-Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 23** ($DOY_{max}$=**173**) at SCBI and **just one day after $DOY_{50}$** on **July 3** ($DOY_{max}$=**183**) at Harvard Forest and averaged .061 and .025 **mm day^-1^ (?)** respectively.
+At both sites, growth phenology differed between ring- and diffuse- porous species (Fig. `r growth_curves_all`, Table `r mean_growth_parameters_table`). 
+Specifically, ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark `r 154-123` days earlier at SCBI and `r 164-132` days earlier at Harvard Forest. 
+Growth of ring-porous species was spread out over a longer growing season, with average $L_{PGS}$ of ring-porous species being **21** days longer at SCBI and **19** days longer at Harvard Forest.
+*(These numbers are calculated as based on `mean DOY75- mean DOY25`, whereas we want `mean [DOY75- DOY25]`.
+Need to fix this, but I filled in approximate numbers for now.)*
 
@@
 
+new Science paper: @zani_increased_2020
+
 From Neil: @zhang_drought_2021, @zhang_extended_2020

### 2020-11-30 doc/manuscript/Dow_manuscript.Rmd
results - climwin
knit in new figure, deleted old draft version,
issues #26, #34

@@
 climwin_caption <- "Figure 3.
-THIS IS OUT OF DATE.
-Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
-Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: DATE at SCBI, DATE at HF; diffuse-porous: DATE at SCBI, DATE at HF).
-Black boxes indicate the optimal window."
+Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- (1ST ROW) and diffuse-porous (2ND ROW) trees at SCBI (FIRST COLUMN) and Harvard Forest (SECOND COLUMN).
+Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: May 4 at SCBI, May 15 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
+Black boxes indicate the optimal window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
+19- May 21 at SCBI, March 19 - May 07 at HF)."
 
@@
 #SI figure numbers
+all_climwin_SI_figs = " S1-S4"
 climwin_rp_DOY25_SCBI = "S1"
@@
 For the majority of weekly time windows from Jan.
-1 to $DOY_{25}$, there was negative relationship between temperature and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
-`r climwin`, `r climwin_rp_DOY25_SCBI`, `r climwin_dp_DOY25_SCBI`).
-This held true for both ring- and diffuse-porous trees at both sites, *albeit* with stronger relationships for diffuse-porous trees. 
-*(Describe the landscape:)*
-In general, there were significant negative relationships between temperature and $DOY_{25}$ for any time period beginning $\ge$ **5** weeks prior to $DOY_{25}$, ending  $\le$ **8-20** weeks prior depending on wood-type and site, and covering at least a **3**-week period.
-Relationships tended to be weaker and more stochastic in sign for short time periods, particularly those just prior to $DOY_{25}$.
-
-Optimal windows occurred from **3/22-5/7** for ring-porous species and from **2/19-5/21** for diffuse-porous species.
+1 to $DOY_{25}$, there was neutral negative relationship between temperature and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
+`r climwin`, `r all_climwin_SI_figs`).
+This held true for both ring- and diffuse-porous trees at both sites. 
+The strongest relationships, including but not limited to the optimal time windows (Table `r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$. 
+(In general, there were significant negative relationships between temperature and $DOY_{25}$ for any time period beginning $\ge$ **5** weeks prior to $DOY_{25}$, ending  $\le$ **8-20** weeks prior depending on wood-type and site, and covering at least a **3**-week period (`r all_climwin_SI_figs`).
+Relationships tended to be weaker and more stochastic in sign for short time periods, particularly those just prior to $DOY_{25}$.)
+(**Cameron, I'm not sure if the above text is up-to-date... Also, see [issue 34](https://github.com/EcoClimLab/growth_phenology/issues/34)** )
 
@@
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/climwin.png")
+knitr::include_graphics("tables_figures/climwin_figure.png")
 ```

### 2020-11-30 doc/manuscript/tables_figures/tables.Rmd
minor update to hypothesis table

@@
   add_header_above(c(" ", "SCBI" = 2, "Harvard"=2, " " )) %>%
-  pack_rows("1. Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("2. Maximum growth rates are independent of early spring temperatures.", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("3. Annual stem growth responds positively to warmer spring temperatures.", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Maximum growth rates are independent of early spring temperatures.", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Annual stem growth responds positively to warmer spring temperatures.", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-11-30 doc/manuscript/Dow_manuscript.Rmd
bit of work on manuscript

@@
 
-\newpage
-
 ```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption}
@@
 
+\newpage
 ```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
@@
 
-At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted back while ${DOY_75}$ did not shift, resulting in an extended peak growing season length when critical window temperatures were warmer (Tables `r hypothesis_table`, Figs.
+At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier while ${DOY_75}$ did not shift, resulting in longer peak growing season length when critical window temperatures were warmer (Tables `r hypothesis_table`, Figs.
 `r pheno_Tsensitivity`, `r DOYtiming`).
-Diffuse-porous $DOY_{25}$, $DOY_{50}$, and ${DOY_75}$ shifted back uniformly resulting in no significant change in peak growing season length.
-At Harvard Forest, ring-porous ${DOY_75}$ shifted back faster than ${DOY_25}$, resulting in shorter peak growins season length, while diffuse-porous ${DOY_25}$ shifted back faster than ${DOY_75}$, resulting in an extended peak growing season length in response to warmer critical window temperatures.
+Diffuse-porous $DOY_{25}$, $DOY_{50}$, and ${DOY_75}$ shifted earlier uniformly resulting in no significant change in peak growing season length. 
+At Harvard Forest, ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length, while diffuse-porous ${DOY_25}$ shifted more than ${DOY_75}$, resulting in an extended peak growing season length in response to warmer critical window temperatures.
 
@@
 
-Total annual growth, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes (Harvard Forest) (Tables `r hypothesis_table`, Fig.
+Total annual growth measured with dendrometer bands, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes (Harvard Forest) (Tables `r hypothesis_table`, Fig.
 `r pheno_Tsensitivity`).
+
+Annual growth recorded in tree-ring records, $RW$, ...
 

### 2020-11-30 RScripts/Pheno_tsensitivity_figure.R
Done n_iter = 2500

@@ -20,7 +20,7 @@ library(rstanarm)
 library(patchwork)
 # Number of MCMC chains & number of simulations per chain.
 # Need to increase this at the end
-n_iter <- 1000
+n_iter <- 2500
 n_chains <- 2
 
 # Objects to keep during memory management clean-up

### 2020-11-30 doc/manuscript/Dow_manuscript.Rmd
added another reference to incorporate
update climwin fig caption
#26

@@
 climwin_caption <- "Figure 3.
-Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- (1ST ROW) and diffuse-porous (2ND ROW) trees at SCBI (FIRST COLUMN) and Harvard Forest (SECOND COLUMN).
+Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
 Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: May 4 at SCBI, May 15 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
@@
 
+@delgado_differences_2020
+
 From Neil: @zhang_drought_2021, @zhang_extended_2020

### 2020-11-30 RScripts/Pheno_tsensitivity_figure.R
Done n_iter = 5000

@@ -20,7 +20,7 @@ library(rstanarm)
 library(patchwork)
 # Number of MCMC chains & number of simulations per chain.
 # Need to increase this at the end
-n_iter <- 2500
+n_iter <- 5000
 n_chains <- 2
 
 # Objects to keep during memory management clean-up

### 2020-11-30 RScripts/Pheno_tsensitivity_figure.R
Removed all mid-script visual outputs of plots to save memory

@@ -220,7 +220,6 @@ fig6_RP <- ggplot() +
   coord_cartesian(xlim = c(11.8, 19.7), ylim = c(80, 240)) +
   theme(legend.position = "none") +
   labs(x = "", y = "DOY", col = "Percentile", title = "SCBI", subtitle = "Ring-porous")
-fig6_RP
 
 fig6_DP <- ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
@@ -232,7 +231,6 @@ fig6_DP <- ggplot() +
   # facet_grid(perc) +
   coord_cartesian(xlim = c(14.7, 19.2), ylim = c(80, 240)) +
   labs(x = "", y = "", col = "Percentile", title = "SCBI", subtitle = "Diffuse-porous")
-fig6_DP
 
 woodtable <- subset(Wood_pheno_table, perc == "DOY_25")
 
@@ -280,7 +278,6 @@ fig6_RP_tot <-  ggplot() +
   coord_cartesian(xlim = c(11.8, 19.7), ylim = c(-.4, 1.5)) +
   theme(legend.position = "none") +
   labs(x = "Temperature (c) 3/22-4/9", y = "Total Growth (cm)")
-fig6_RP_tot
 
 fig6_DP_tot <-  ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
@@ -292,7 +289,6 @@ fig6_DP_tot <-  ggplot() +
   coord_cartesian(xlim = c(14.7, 19.2), ylim = c(-.4, 1.5)) +
   theme(legend.position = "none") +
   labs(x = "Temperature (c) 2/19-5/21", y = "")
-fig6_DP_tot
 
 # Clean-up
 rm(list = setdiff(ls(), objects_to_keep))
@@ -339,7 +335,6 @@ fig6_RP_sl <- ggplot() +
   coord_cartesian(xlim = c(11.8, 19.7), ylim = c(3, 106)) +
   theme(legend.position = "none") +
   labs(x = "", y = "Season Length (# of Days)")
-fig6_RP_sl
 
 fig6_DP_sl <- ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
@@ -351,7 +346,6 @@ fig6_DP_sl <- ggplot() +
   coord_cartesian(xlim = c(14.7, 19.2), ylim = c(3, 90)) +
[…]

### 2020-12-03 doc/manuscript/Dow_manuscript.Rmd
working on text... slower than hoped!
FYI, @camerondow35 , I'm tyring to make some progress now,

@@
 ## ("limitation in filling gap, how to interpret/fix")
+Although the number of years analyzed here is limited (# years at SCBI, # years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (*CITE RELEVANT FIGS & TABLES* ).
+This indicates that this is not a subtle effect.
+However, the effect on L_PGS, r_max, and $\Delta DBH$, if any, is subtle, and a much longer record would be needed to detect such effects. 
+Fortunately, we were able to get at annual growth using tree-ring data (Fig. `r tree_rings`).
+These confirm that the effect of spring temperatures on annual growth is modest and generally not significant, while revealing a slightly more complex picture.
+Although spring temperatures prior to May rarely had a significant influence on
 

### 2020-12-03 doc/manuscript/Dow_manuscript.Rmd
rough work on discussion
@camerondow35 , this is quite different from what you had. (sorry!) Your text is all quite interesting and well-written, but I'd focus more on the themes in my text, as opposed to differences between ring- and diffuse-porous species. D'Orangeville et al covers those differences, and our novel contribution is the spring temperatures.

This still ndeds work, but hopefully helps. I'll do more when time allows.

@@
 These confirm that the effect of spring temperatures on annual growth is modest and generally not significant, while revealing a slightly more complex picture.
-Although spring temperatures prior to May rarely had a significant influence on
+Although spring temperatures prior to May rarely had a significant influence on ...
 
 ## ("limitation in generalization, how to interpret/fix")
+THe tree-rings also reveal that the strongest temperature sensitivity aligned with peak growing season.
+This pattern is common across tree-ring studies in temperate deciduous forests.
+(*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
+Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree grwoth will hold generall.y.
 
 ## ("contributions beyond")
+Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
+Warm spring temperatures result in earlier leaf-out (REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (*CITE RELEVANT FIGS & TABLES* ).
+Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypotheses`, Figs `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
+Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
 
 ## ("science is better now")
+
+~~~~
 

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
updated hypothesis table, knit in new tree-ring figure

@@
 ```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+## potential reviewers: Zani
 
@@
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/tree_rings.tif")
+knitr::include_graphics("tables_figures/tree_cores_tmax.png")
 ```
@@
 Warm spring temperatures result in earlier leaf-out (REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (*CITE RELEVANT FIGS & TABLES* ).
-Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypotheses`, Figs `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
+Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
 Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
+If sink limitation controls the end of woody growing season,
 
 ## ("science is better now")
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models [Table `r hypothesis_table`; Figs.
+`r pheno_Tsensitivity`, `r tree_rings`; @zani_increased_2020].
+The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.
 

### 2020-12-04 doc/manuscript/tables_figures/tables.Rmd
updated hypothesis table, knit in new tree-ring figure

@@
 
+#add footnote markers:
+hypothesis_table[8,2] <- paste0(hypothesis_table[8,2], footnote_marker_number(1)) 
+hypothesis_table[8,3] <- paste0(hypothesis_table[8,3], footnote_marker_number(2)) 
+hypothesis_table[8,5] <- paste0(hypothesis_table[8,5], footnote_marker_number(3))
+
 kable(hypothesis_table, col.names = c("Hypotheses and Specific Predictions", "RP", "DP","RP", "DP", "Results"), caption = "Table 1.
@@
   pack_rows("Annual stem growth responds positively to warmer spring temperatures.", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) %>%
+
+#add footnote: 
+kableExtra::footnote(number = c("One of nine species analyzed had significant positive response to April $T_{max}$; two had significant negative response to March $T_{max}$", "One of two species analyzed had signficant positive response to April $T_{max}$, both had negative response to May $T_{max}$", "The one species analyzed had a signficant negative response to April $T_{max}$."), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
 ```

### 2020-12-04 RScripts/SCBI_quiltplot.R
change SCBI SSS to 0.8
@camerondow35 , this fixes issue #38. You'll just need to re-run

@@ -47,7 +47,7 @@ type.of.start.date <- c("1901_2009", "1920_1949", "1950_1979", "1980_2009")
 
 
 ## Define sss threshold ####
-sss.threshold = 0.75
+sss.threshold = 0.80
 ## Define start and end month for anlaysis ####
 start <- 1 # April of previous year
 end <- 8 # August of current year

### 2020-12-04 doc/manuscript/tables_figures/tables.Rmd
fix up species table, format in tables.Rmd
#36

@@
 #add footnote: 
-kableExtra::footnote(number = c("One of nine species analyzed had significant positive response to April $T_{max}$; two had significant negative response to March $T_{max}$", "One of two species analyzed had signficant positive response to April $T_{max}$, both had negative response to May $T_{max}$", "The one species analyzed had a signficant negative response to April $T_{max}$."), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
+kableExtra::footnote(number = c("One of nine species analyzed had significant positive response to April $T_{max}$; one had significant negative response to March $T_{max}$", "One of two species analyzed had signficant positive response to April $T_{max}$, both had negative response to May $T_{max}$", "The one species analyzed had a signficant negative response to April $T_{max}$."), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
 ```
+
+\newpage
+```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+species_table <- read.csv(file = "species_table.csv", col.names = c("site", "xylem porosity","species","n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
+kable(species_table, caption = "Table 2.
+Dominant ring- and diffuse-porous species at SCBI and Harvard Forest analyzed here, along with sample sizes.",  format = "latex", booktabs = TRUE,  escape = F) %>%
+add_header_above(c(" ", " ", " ", "Dendrometer Bands" = 2, "Tree Cores"=2 )) %>%
+kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
replace word table with .png, insert tree-ring methods, misc

@@
 
-\newpage
-*(Do we want a table of species here?? Columns would be Species, xylem porosity, n trees, n tree-years (after cleaning). Details will be in an SI table)*
-
 ```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-species_table <- read.csv("tables_figures/species_table.csv",stringsAsFactors = FALSE, check.names = FALSE)
-knitr::kable(species_table) 
+knitr::include_graphics("tables_figures/species_table.png")
 ```
@@
 
-(*The following seems out of date, right? Scripted cleaning is not described.* -- *Manually cleaning was done for pre-analysis data. Added some text to clarify.*)
-The dendroband data were manually cleaned by visual inspection before being used for modeling.
+The dendroband data were manually cleaned by visual inspection before analysis.
 Three classes of mistakes: 1.
@@
 `r schematic`).
-This 5 parameter logistic model was fitted to each tag year's dendroband measurements in R: L + ((K - L) / (1 + 1/theta * exp(-(r * (doy - doy_ip) / theta)) ^ theta))
-Where L and K are lower and upper asymptotes of the model, theta is a tuning parameter controlling the slope of the curve toward the upper asymptote, r is the slope of the curve at the inflection point, doy_ip is the day of year where the inflection point occurs, and doy are the modeled days of year, in this case, 1-365.
+This 5 parameter logistic model was fitted to each tag year's dendroband measurements in R:
+$$L + ((K - L) / (1 + 1/\theta * exp(-(r * (doy - doy_{ip}) / \theta)) ^ {\theta}))$$  **(Cameron, please check formatting. It should be possible to write this more clearly with latex formatting. )**
+Here, $L$ and $K$ are lower and upper asymptotes of the model, theta is a tuning parameter controlling the slope of the curve toward the upper asymptote, $r$ is the slope of the curve at the inflection point, $doy_{ip}$ is the day of year where the inflection point occurs, and $doy$ are the modeled days of year, in this case, 1-365.
 *We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow.* Using the optimized parameters for each tree year, we could then determine the DBH on every day within that year, giving us the ability to extract the intra-annual growth variables we were interested in.
@@
 ### Tree-rings
-(*tree rings-- KAT can write this*)
+Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
+We analyzed records for the dominant ring- and diffuse- porous species at each site (Table `r species_table`), excluding species with other xylem architectures (*Juglans nigra* L.
+at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
+At SCBI,  were collected as described in @helcoski_growing_2019. 
+Briefly, ...  (*summarize / reword from Helcoski*). 
+In total, we analyzed # cores representing eleven of the twelve species contributing most to aboveground woody productivity within the SCBI ForestGEO plot.
+At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
+In total, we analyzed # cores from three species.
+
+All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
+(*describe chronology building-- as in @helcoski_growing_2019-- Neil can do this quickly*)
+Start year was defined at SSS=0.8.
 
@@
 (*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
-Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree grwoth will hold generall.y.
+Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree growth will hold generally.
 
@@
 These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011).
-The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
+The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likelihood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
 The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014).

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
Formatted mean growth parameters table
moved  mean growth parameters table to tables. Rmd, formatted nicely

@@
 At SCBI, four species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.
-At Harvard Forest, eleven species were sampled including eight diffuse-porous and three ring-porous (Table `r species_table`)*(bands_bySP csv's in data folder)*.
+At Harvard Forest, eleven species were sampled including eight diffuse-porous and three ring-porous (Table `r species_table`).
 
@@
 
-\newpage
-```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-mean_growth_parameters <- read.csv("tables_figures/mean_growth_parameters.csv",stringsAsFactors = FALSE, check.names = FALSE)
-knitr::kable(mean_growth_parameters) 
+```{r mean_growth_parameters_table, out.width="100%", echo=FALSE, fig.align="center"}
+knitr::include_graphics("tables_figures/mean_growth_parameters_table.png")
 ```

### 2020-12-04 doc/manuscript/tables_figures/tables.Rmd
Formatted mean growth parameters table
moved  mean growth parameters table to tables. Rmd, formatted nicely

@@
   #column_spec(1, width = "4cm") %>%
-  add_header_above(c(" ", "SCBI" = 2, "Harvard"=2, " " )) %>%
+  add_header_above(c(" ", "SCBI" = 2, "Harvard Forest"=2, " " )) %>%
   pack_rows("Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
@@
 ```
+
+\newpage
+```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+mean_growth_parameters <- read.csv("mean_growth_parameters.csv",  col.names = c("", "ring","diffuse","ring", "diffuse"), stringsAsFactors = FALSE, check.names = FALSE)
+kable(mean_growth_parameters, caption = "Table 3.
+Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest.
+",  format = "latex", booktabs = TRUE,  escape = F) %>%
+add_header_above(c(" ",  "SCBI" = 2, "Harvard Forest"=2 )) %>%
+    column_spec(1, width = "4cm") %>%
+kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```

### 2020-12-04 doc/manuscript/tables_figures/tables.Rmd
add species code to species table
needed for interpretation of tree-ring figure

#36

@@
 library(kableExtra)
-species_table <- read.csv(file = "species_table.csv", col.names = c("site", "xylem porosity","species","n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
+species_table <- read.csv(file = "species_table.csv", col.names = c("site", "xylem porosity","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
 kable(species_table, caption = "Table 2.
 Dominant ring- and diffuse-porous species at SCBI and Harvard Forest analyzed here, along with sample sizes.",  format = "latex", booktabs = TRUE,  escape = F) %>%
-add_header_above(c(" ", " ", " ", "Dendrometer Bands" = 2, "Tree Cores"=2 )) %>%
+add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
tree rings caption

@@
 
-tree_rings_caption <- "Figure 6. Climate sensitiivity of annual growth from tree-rings. THIS IS JUST A PLACEHOLDER TO SHOW THE RESULTS--WE'LL WANT TO RE-MAKE FIGURE."
+tree_rings_caption <- "Figure 6.
+Climate sensitiivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithonian Conservation Biology Institute and Harvard Forest.
+Shown are Pearson correlations between ring-width index and and monthly Tmax.
+Species codes are as given in Table 2."
 

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
switch to regular word formatting
ugly, but solves section/ figure numbering issues

@@
 output:
-  bookdown::word_document2: 
-    reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
-    fig_caption: yes
+  word_document
+  #bookdown::word_document2: 
+    #reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+    #fig_caption: yes
 bibliography: references.bib

### 2020-12-04 RScripts/HF_quiltplot.R
updated qulit plots

@@ -164,7 +164,7 @@ save.plots = TRUE
 for(v in climate_variables) {
   print(v)
 
-  X <- all.dcc.output[all.dcc.output$variable %in% "tmn", ]
+  X <- all.dcc.output[all.dcc.output$variable %in% v, ]
 
   x <- data.frame(reshape(X[, c("month", "Species", "coef")], idvar = "month", timevar = "Species", direction = "wide"))
   rownames(x) <- ifelse(grepl("curr",  rownames(x)), toupper(rownames(x)), tolower( rownames(x)))
@@ -193,9 +193,9 @@ for(v in climate_variables) {
 
   v <-  toupper(v)
   v <- gsub("PDSI_PREWHITEN" , "PDSI", v)
-  x <- x[,c(3,2,1)]
-  x.sig <- x.sig[,c(3,2,1)]
-  x.sig2 <- x.sig2[,c(3,2,1)]
+  x <- x[,c(2,1,3)]
+  x.sig <- x.sig[,c(2,1,3)]
+  x.sig2 <- x.sig2[,c(2,1,3)]
   png(paste0("results/", "monthly_", "correlation", "Harvard", v, ".png"), res = 150, width = 169, height = 169, units = "mm", pointsize = 10)
 
   my.dccplot(x = as.data.frame(t(x)), sig = as.data.frame(t(x.sig)), sig2 = as.data.frame(t(x.sig2)),  main = ifelse(v %in% "PETminusPRE", "PET-PRE", v), method = "correlation")

### 2020-12-04 RScripts/SCBI_quiltplot.R

@@ -265,7 +265,7 @@ for(f in species) {
 all.dcc.output$variable <- substr(paste(row.names(all.dcc.output)), 1, 3)#get variable from row name
 all.dcc.output$month <- substr(paste(row.names(all.dcc.output)), 5, 12)#get month from row name
 
-write.csv(all.dcc.output, file = "results/SCBI_core_corr.csv", row.names = FALSE)
+#write.csv(all.dcc.output, file = "results/SCBI_core_corr.csv", row.names = FALSE)
 #} # for(f in species)
 
 #############################################
@@ -305,9 +305,9 @@ for(v in climate_variables) {
   v <-  toupper(v)
   v <- gsub("PDSI_PREWHITEN" , "PDSI", v)
   #x <- x[,c(5,9,1,2,3,4,6,7,11,12,13,14)]
-  x <- x[,c(14,13,12,11,7,6,4,3,2,1,9,5)]
-  x.sig <- x.sig[,c(14,13,12,11,7,6,4,3,2,1,9,5)]
-  x.sig2 <- x.sig2[,c(14,13,12,11,7,6,4,3,2,1,9,5)]
+  x <- x[,c(9,5,14,13,12,11,6,4,3,2,1)]
+  x.sig <- x.sig[,c(9,5,14,13,12,11,6,4,3,2,1)]
+  x.sig2 <- x.sig2[,c(9,5,14,13,12,11,6,4,3,2,1)]
 
   png(paste0("results/", "monthly_", "correlation", "SCBI", v, ".png"), res = 150, width = 169, height = 169, units = "mm", pointsize = 10)
 

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
some work on methods, auto fill some statistics

@@
 hypothesis_table=1
-species_table=2
+table_species=2
 mean_growth_parameters_table=3
@@
 climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
+
+## some stats/ 
+species_table <- read.csv(file = "tables_figures/species_table.csv")
+n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
+n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
+n_cores <- sum(na.omit(species_table[, "n.cores"]))
 
@@
 
-Metal dendrometer bands installed on *99-145* and *700-755* trees within the SCBI and Harvard Forest ForestGEO plots, respectively, were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 and 1998-2003, respectively.
+Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (Table `r table_species`). 
+Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
+Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
 The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`) *(tags_per_year_preclean in data folder)*.
-Across years, the number of bands sampled averaged *~135* at SCBI and  *~717* at Harvard Forest.
-At SCBI, four species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.
-At Harvard Forest, eleven species were sampled including eight diffuse-porous and three ring-porous (Table `r species_table`).
+Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
+In total, our analysis included `r n_tree_years` tree-years (Table `r table_species`).
 
-```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
+```{r species, eval=TRUE, echo=FALSE, warning=FALSE}
 knitr::include_graphics("tables_figures/species_table.png")
@@
 Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
-We analyzed records for the dominant ring- and diffuse- porous species at each site (Table `r species_table`), excluding species with other xylem architectures (*Juglans nigra* L.
+We analyzed records for the dominant ring- and diffuse- porous species at each site (Table `r table_species`), excluding species with other xylem architectures (*Juglans nigra* L.
 at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
-At SCBI,  were collected as described in @helcoski_growing_2019. 
-Briefly, ...  (*summarize / reword from Helcoski*). 
-In total, we analyzed # cores representing eleven of the twelve species contributing most to aboveground woody productivity within the SCBI ForestGEO plot.
+At SCBI, cores were collected as described in @helcoski_growing_2019. 
+Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
 At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
-In total, we analyzed # cores from three species.
+In total, we analyzed `r n_cores` cores from eleven species at SCBI and three species at Harvard Forest (Table `r table_species`).
 
@@
 ### Climate
-Daily maximum temperatures ($T_{MAX}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github).
-Daily maximum temperatures at Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
-The R package climpact () was used to plot $T_{MAX}$ for visual inspection.
-Climpact also identifies readings that were greater than 3 standard deviations away from yearly means.
-These points were labeled as outliers and removed from the dataset.
-Gaps in the SCBI meteorological tower data were subsequently filled using $T_{MAX}$ readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
-(*how were gaps filled for Harvard?
-Or was the record good?*)
+For the dendrometer band analysis, we used climate data from local weather stations. 
+Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeoclimate_2020].
+Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
+The R package climpact () was used to plot temperatures for visual inspection. 
+Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
+Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
+(*how were gaps filled for Harvard? Or was the record good?*)
 
-(*climate data for tree rings-- KAT can write this*)
+For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020], again via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeoclimate_2020].
 
@@
 
-```{r climwin, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
+```{r climwin, out.width="80%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
 # Created in RScripts/additional_figures.R
@@
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/tree_cores_tmax.png")
+knitr::include_graphics("tables_figures/tree_cores_tmax.tif")
 ```
@@
 
-## (summary of key findings, gap filled)
 We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity...
 
-## ("limitation in filling gap, how to interpret/fix")
 Although the number of years analyzed here is limited (# years at SCBI, # years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (*CITE RELEVANT FIGS & TABLES* ).
@@
 
-## ("limitation in generalization, how to interpret/fix")
 THe tree-rings also reveal that the strongest temperature sensitivity aligned with peak growing season.
@@
 
-## ("contributions beyond")
 Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
@@
 Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
-If sink limitation controls the end of woody growing season,
+As sink limitation controls the end of woody growing season [@zani_increased_2020], we might expect that an earlier start to growth buffers against unfavorable growth conditions during the peak growing season (e.g., drought).
+*Do we see any evidence of this???*
 
-## ("science is better now")
 As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models [Table `r hypothesis_table`; Figs.
@@
 ~~~~
+## Cam's previous content
 

### 2020-12-04 doc/manuscript/tables_figures/tables.Rmd
update hypothesis table
https://github.com/EcoClimLab/growth_phenology/commit/571dd68a58006b25a32cc256c3e5673dc4243d12#commitcomment-44814447

@@
 #add footnote: 
-kableExtra::footnote(number = c("One of nine species analyzed had significant positive response to April $T_{max}$; one had significant negative response to March $T_{max}$", "One of two species analyzed had signficant positive response to April $T_{max}$, both had negative response to May $T_{max}$", "The one species analyzed had a signficant negative response to April $T_{max}$."), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
+kableExtra::footnote(number = c("One of nine species analyzed had significant positive response to April $T_{max}$; one had significant negative response to March $T_{max}$", "One of two species analyzed had signficant positive response to April $T_{max}$, both had negative response to May $T_{max}$", "One of the two species was negatively correlated with April $T_{max}$, and the other positively correlated with May $T_{max}$."), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
 ```

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
update authors/affiliations
working on methods.
@camerondow35 , most of this is minor wording/ formatting with new line per sentence (I find easier to track).

I did format the equation; please check.

@@
 
-Loïc D’Orangeville
+Albert Y. Kim^1,2^
+
+Loïc D’Orangeville^3,4^
 
@@
 
-Ian McGregor^1^
+Ian McGregor^1,5^
 
-Sean McMahon
+William McShea^1^
 
-Albert Y. Kim^1,2^
+Sean McMahon^6,7^
 
-David A. Orwig
+David A. Orwig^3^
 
-Neil Pederson
+Neil Pederson^3^
 
-Kristina J. Anderson-Teixeira^1,n^*
+Kristina J. Anderson-Teixeira^1,7^*
 
@@
 
-n. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+3. Harvard Forest, Petersham, MA 01366, USA
+
+4. Faculty of Forestry and Environmental Management, University of New Brunswick, Fredericton, NB, E3B 5A3, Canada.
+
+5. NCSU
+
+6. SERC
+
+7. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
 
@@
 
-Study sites included two temperate forests in the Eastern United States, both part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al. 2015].
+Study sites included two temperate forests in the eastern United States, both part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al. 2015].
 
@@
 
-Measurements were timed to begin before beginning of spring growth and to continue through the cessation of growth in the fall.
-At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, and the median end date was 10/17, coinciding with late senescence in the plot.
-The earliest start date for intraannual measurements at SCBI was March 30, 2020 and the latest end date was November 26, 2012.
-Harvard Forest measurements took place at a similar time, with the median start date of 4/23 and median end date of 10/30.
+Measurements were timed to begin before the beginning of spring growth and to continue through the cessation of growth in the fall. 
+At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, with the earliest start date being March 30, 2020.
+Measurements were continued through to fall leaf senescence, with the median end date being 10/17 and the latest end date was November 26, 2012.
+Timing of measurements at Harvard Forest was similar, with the median start date of 4/23 and median end date of 10/30.
 1998 was an anomalous year where initial measurements were taken on 1/5, but not taken again until 4/15.
@@
 
-The dendroband data were manually cleaned by visual inspection before analysis.
-Three classes of mistakes: 1.
-Error in measurement - weekly measurement was drastically different from previous week and following week.
-Mistakes were removed from the raw data before modeling.
-2.
-Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern.
-Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year.
-In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level.
-3.
-Other - cases where data was clearly wrong but with unknown causes.
+The dendrometer band data were manually cleaned by visual inspection before analysis. 
+We screened for three classes of mistakes. 
+First, when a weekly measurement was drastically different from previous week and following weeks, this was assumed to be a measurement error and the record was removed.
+Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump followed by normal growth pattern, this was assumed to be a case where the band was stuck and then released.
+In these cases, the full annual record for the tree was removed.
+Third, there were cases where data were clearly wrong but with unknown causes.
 In cases where several measurements were an issue, they were simply removed.
@@
 `r schematic`).
-This 5 parameter logistic model was fitted to each tag year's dendroband measurements in R:
-$$L + ((K - L) / (1 + 1/\theta * exp(-(r * (doy - doy_{ip}) / \theta)) ^ {\theta}))$$  **(Cameron, please check formatting. It should be possible to write this more clearly with latex formatting. )**
-Here, $L$ and $K$ are lower and upper asymptotes of the model, theta is a tuning parameter controlling the slope of the curve toward the upper asymptote, $r$ is the slope of the curve at the inflection point, $doy_{ip}$ is the day of year where the inflection point occurs, and $doy$ are the modeled days of year, in this case, 1-365.
-*We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow.* Using the optimized parameters for each tree year, we could then determine the DBH on every day within that year, giving us the ability to extract the intra-annual growth variables we were interested in.
+This 5 parameter logistic model was fitted to each tag year's dendrometer band measurements in R:
+$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$  **(Cameron, please check equation following formatting. )**
+Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to $DBH$ at the beginning and end of the year, respectively. 
+$\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote, $r$ is the slope of the curve at the inflection point, $doy_{ip}$ is the day of year where the inflection point occurs, and $doy$ are the modeled days of year, in this case, 1-365.
+*We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow.* 
+Using the optimized parameters for each tree year, we could then determine the $DBH$ on every day within that year, giving us the ability to extract the intra-annual growth variables we were interested in.
 
@@
 Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the acceptable range we determined. 
-Modeled tag years were removed if: 1.
-single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, HF = 34) 2.
-DOY where peak growth rate occured was 2 standard deviations away, calculated for each wood-type in each year (aimed to remove models which predicted growth to peak during the winter; SCBI = 62, HF = 106) 3.
-tag years with small or negligible total growth (deltaDBH < 0.02 mm; SCBI = 0, HF = 66), 4.
-models which predicted total yearly growth to take longer than 365 days (SCBI = 181, HF = 199) 5.
-models with unexplained spikes (SCBI = 0, HF = 3).
+Modeled tag years were removed under five conditions:  
+(1) single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, HF = 34); 
+(2) DOY where peak growth rate occurred was 2 standard deviations away, calculated for each wood-type in each year (aimed to remove models which predicted growth to peak during the winter; SCBI = 62, HF = 106);
+(3) tree-years with small or negligible total growth ($\Delta DBH$ < 0.02 mm; SCBI = 0, HF = 66); 
+(4) model fit predicted total yearly growth to take longer than 365 days (SCBI = 181, HF = 199); 
+and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, HF = 3). 
 At Harvard Forest the tag years removed through this method was proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others *(SI figure?

### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
method clean-up

@@
 
-We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig.
-`r schematic`).
-This 5 parameter logistic model was fitted to each tag year's dendrometer band measurements in R:
+We fit the model of @mcmahon_general_2015 to define phenological dates and growth rates (Fig. `r schematic`). 
+This five-parameter logistic model was fitted to dendrometer band data from each tree-year in R: 
 $$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$  **(Cameron, please check equation following formatting. )**
 Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to $DBH$ at the beginning and end of the year, respectively. 
-$\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote, $r$ is the slope of the curve at the inflection point, $doy_{ip}$ is the day of year where the inflection point occurs, and $doy$ are the modeled days of year, in this case, 1-365.
-*We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow.* 
-Using the optimized parameters for each tree year, we could then determine the $DBH$ on every day within that year, giving us the ability to extract the intra-annual growth variables we were interested in.
+$\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote, $r$ is the slope of the curve at the inflection point, $DOY_{ip}$ is the day of year where the inflection point in growth rate occurs, and $DOY$ are the modeled days of year (*i.e.*, julian days).
+*We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow. (???)* 
+Using the optimized parameters for each tree-year, we could then determine the $DBH$ on every day within that year, giving us the ability to extract the intra-annual growth variables of interest (Fig.
+`r schematic`).
 
-After model fitting, we used a script to remove poorly fit models. 
+After fitting the  @mcmahon_general_2015 model, we used a script to remove poorly fit models. 
 Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the acceptable range we determined. 
-Modeled tag years were removed under five conditions:  
+Modeled fits for tree-years were removed under five conditions:  
 (1) single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, HF = 34); 
-(2) DOY where peak growth rate occurred was 2 standard deviations away, calculated for each wood-type in each year (aimed to remove models which predicted growth to peak during the winter; SCBI = 62, HF = 106);
+(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for it's xylem architecture group, year, and site (SCBI = 62, HF = 106); 
 (3) tree-years with small or negligible total growth ($\Delta DBH$ < 0.02 mm; SCBI = 0, HF = 66); 
-(4) model fit predicted total yearly growth to take longer than 365 days (SCBI = 181, HF = 199); 
+(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 181, HF = 199); 
 and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, HF = 3). 
@@
 data_cleaning_figure_hf in manuscript folder)*.
-At SCBI a higher proportion of oak trees were removed *(data_cleaning_figure_scbi)*. 
-The majority (White oak = 91, Red oak = 79) were removed in step 5 above, indicating that the LG5 model struggled to accurately model these these species.
+At SCBI a higher proportion of RP trees were removed--the majority in step 4 *(data_cleaning_figure_scbi)*.
 

### 2020-12-06 doc/manuscript/tables_figures/tables.Rmd
table updates

@@
 kable(species_table, caption = "Table 2.
-Dominant ring- and diffuse-porous species at SCBI and Harvard Forest analyzed here, along with sample sizes.",  format = "latex", booktabs = TRUE,  escape = F) %>%
+Dominant ring- and diffuse-porous species at SCBI and Harvard Forest, along with sample sizes analyzed here.",  format = "latex", booktabs = TRUE,  escape = F) %>%
 add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
+  pack_rows("", 1, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 10, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 12, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 15, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-12-06 doc/manuscript/Dow_manuscript.Rmd
work on methods

@@
 
-Study sites included two temperate forests in the eastern United States, both part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al. 2015].
+Study sites included two temperate forests in the eastern United States, both part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al. 2015; Davies et al. 2021].
 
@@
 Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*). 
-The climate is humid temperate, with a mean annual precipitation of 998 mm, January temperatures averaging 1 degrees C and July temperatures averaging 24 degrees C (Helcoski et al, 2019).
+The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July (Helcoski et al, 2019).
 
 The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
-*Harvard Forest was established in 1907 as a study site for university students and researchers.
-(Previous use?)* Dominate species within the approximately 13 HA plot are hemlock (*Tsuga spp.*), oak (*Quercus spp.*) and maple (*Acer spp.*).
-The climate is temperate continental with a mean annual precipitation of 1120 mm, January temperatures averaging -12 degrees C and July temperatures averaging 19 degrees C (D'Orangeville et al, ).
+Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and signficant hurricane damage in *YEAR*. 
+Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga spp.*), oak (*Quercus spp.*) and maple (*Acer spp.*). 
+The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July (D'Orangeville et al, ).
 
-## Data preparation
+## Data collection and preparation
 ### Dendrometer bands
@@
 This five-parameter logistic model was fitted to dendrometer band data from each tree-year in R: 
-$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$  **(Cameron, please check equation following formatting. )**
+$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
 Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to $DBH$ at the beginning and end of the year, respectively. 
@@
 All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
-(*describe chronology building-- as in @helcoski_growing_2019-- Neil can do this quickly*)
-Start year was defined at SSS=0.8.
+We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
+*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time [@cook_calculating_1997].
+Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology [@jones_estimating_1997].
+The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
+the inner 30–50 yr of each record depending on full chronology length) dropped below three trees.
+Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices [@cook_time_1985].
+We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].*
+We defined chronology start year (Table `r table_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
 
@@
 
-## (Climwin)
+## Determining critical temperature windows
 
-The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package Climwin.
-This package tests the correlation between climate variables ($T_{MAX}$) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'.
-as the best-model.
-We instructed Climwin to search for the best window beginning at February 1st and ending at the mean DOY for each phenology variable (chart in SI?).
-This was done to find corresponding (median of 95% CI window open and close) start and end dates for the analysis of ring-porous and diffuse-porous trees, as well as all trees pooled together.
+The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@van_de_pol_identifying_2016]. 
+This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window.
+Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from February 1 (Julian day = 32) and the mean $DOY_{25}$ for the species group and site (Table `r mean_growth_parameters_table`).
+*The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
+Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic`), because we were interested in the influence of temperature on the initiation of woody growth.
+We note that other growth parameters would be controlled by different critical temperature windows; for example, at SCBI $\Delta DBH$ responds most strongly to May-July climate [@helcoski_growing_2019].
+Determining the climate variables that most strongly control all parameters is beyond the scope of this analysis.
 
@@
 
-Correlation between these variables and spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
+(*This section needs more detail.*)
 
-Notes for now:
+Correlation between these variables and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
 
-1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
-    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
-    1. Random effects: tag number (no species just yet)
+Mixed effects models were used to test the response of growth phenology variables ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{ip}$, $r_{max}$, and $\Delta DBH$) to fixed effects of xylem porosity and mean temperature ($T_{max}$, $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
+We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ and $T_{min}$. 
+Models were run using the LME4 package in R [@bates_fitting_2015].
 
--Mixed effect model
-
-A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and spring temperature, along with random effects of species and tag.
-We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison.
-Models were run using the LME4 package in R (Bates, D., Mächler, M., Bolker, B.
-& Walker, S.
-(2015).
-Fitting linear mixed‐effects models using lme4.
-J.
-Stat.
-Softw., 67, 1–48.)
-
-•	Bayesian heirarchical model
 This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language @rstanarm_2020.
@@
 Specifically, ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark `r 154-123` days earlier at SCBI and `r 164-132` days earlier at Harvard Forest. 
-Growth of ring-porous species was spread out over a longer growing season, with average $L_{PGS}$ of ring-porous species being **21** days longer at SCBI and **19** days longer at Harvard Forest.
-*(These numbers are calculated as based on `mean DOY75- mean DOY25`, whereas we want `mean [DOY75- DOY25]`.
-Need to fix this, but I filled in approximate numbers for now.)*
+Growth of ring-porous species was spread out over a longer growing season, with average $L_{PGS}$ of ring-porous species being 21 days longer at SCBI and 19 days longer at Harvard Forest. 
+Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.
 
@@
 
-## Optimal climate windows
+## Critical temperature windows
 

### 2020-12-06 doc/manuscript/Dow_SI.Rmd
started putting climwin results in SI
SCBI done, HF still needs to be done.

@camerondow35, it would help to auto-save climwin results in tables_figures folder.

@@
 
-```{r climwin_rp_DOY25_SCBI, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_rp_DOY25_SCBI_short_caption}
-knitr::include_graphics("tables_figures/SCBI_mettower_ring-porous_DOY25.png")
+**(a) $T_{max}$**
+```{r climwin_rp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left"}
+knitr::include_graphics("tables_figures/SCBI_ring-porous_TMAX.png")
+```
+
+**(b) $T_{min}$**
+```{r climwin_rp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_SCBI_short_caption}
+knitr::include_graphics("tables_figures/SCBI_ring-porous_TMIN.png")
 ```
@@
 
-```{r climwin_dp_DOY25_SCBI, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_dp_DOY25_SCBI_short_caption}
-knitr::include_graphics("tables_figures/SCBI_mettower_diffuse-porous_DOY25.png")
+**(a) $T_{max}$**
+```{r climwin_dp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left"}
+knitr::include_graphics("tables_figures/SCBI_diffuse-porous_TMAX.png")
+```
+
+**(b) $T_{min}$**
+```{r climwin_dp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_SCBI_short_caption}
+knitr::include_graphics("tables_figures/SCBI_diffuse-porous_TMIN.png")
 ```
@@
 
-```{r climwin_rp_DOY25_HARV, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_rp_DOY25_HARV_short_caption}
+**(a) $T_{max}$**
+```{r climwin_rp_tmax_HARV, out.width="100%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_HARV_short_caption}
 knitr::include_graphics("tables_figures/HF_ring-porous_DOY25.png")
@@
 
+**(b) $T_{min}$**
+
 \newpage
 
-```{r climwin_dp_DOY25_HARV, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_dp_DOY25_HARV_short_caption}
+**(a) $T_{max}$**
+```{r climwin_dp_tmin_HARV, out.width="100%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_HARV_short_caption}
 knitr::include_graphics("tables_figures/HF_diffuse-porous_DOY25.png")
 ```
+
+**(b) $T_{min}$**

### 2020-12-06 doc/manuscript/Dow_manuscript.Rmd
bot of work on methods/ results

@@
 This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window.
-Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from February 1 (Julian day = 32) and the mean $DOY_{25}$ for the species group and site (Table `r mean_growth_parameters_table`).
+Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (Table `r mean_growth_parameters_table`).
 *The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
@@
 For the majority of weekly time windows from Jan.
-1 to $DOY_{25}$, there was neutral negative relationship between temperature and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
+1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
 `r climwin`, `r all_climwin_SI_figs`).
@@
 Relationships tended to be weaker and more stochastic in sign for short time periods, particularly those just prior to $DOY_{25}$.)
-(**Cameron, I'm not sure if the above text is up-to-date... Also, see [issue 34](https://github.com/EcoClimLab/growth_phenology/issues/34)** )
+(**Cameron, I'm not sure if the above text is up-to-date... ** )
 
@@
 ```
+
+*$DOY_{25}$ was generally less strongly correlated with $T_{min}$ than with $T_{max}$.-- not sure this is true*
+(*give some details, cite SI tables.*)
+
+Henceforth, we focus on $T_{max}$.
 

### 2020-12-06 doc/manuscript/Dow_SI.Rmd
finished putting in climwin plots

@@
 **(a) $T_{max}$**
-```{r climwin_rp_tmax_HARV, out.width="100%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_HARV_short_caption}
-knitr::include_graphics("tables_figures/HF_ring-porous_DOY25.png")
+```{r climwin_rp_tmax_HARV, out.width="70%", echo=FALSE, fig.align="left"}
+knitr::include_graphics("tables_figures/HF_ring-porous_TMAX.png")
 ```
@@
 **(b) $T_{min}$**
+```{r climwin_rp_tmin_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_HARV_short_caption}
+knitr::include_graphics("tables_figures/HF_ring-porous_TMIN.png")
+```
 
@@
 **(a) $T_{max}$**
-```{r climwin_dp_tmin_HARV, out.width="100%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_HARV_short_caption}
-knitr::include_graphics("tables_figures/HF_diffuse-porous_DOY25.png")
+```{r climwin_dp_tmax_HARV, out.width="70%", echo=FALSE, fig.align="left"}
+knitr::include_graphics("tables_figures/HF_diffuse-porous_TMAX.png")
 ```
@@
 **(b) $T_{min}$**
+```{r climwin_dp_tmin_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_HARV_short_caption}
+knitr::include_graphics("tables_figures/HF_diffuse-porous_TMIN.png")
+```

### 2020-12-07 doc/manuscript/Dow_SI.Rmd
working on results

@@
 climwin_dp_DOY25_HARV = "S4"
+tree_rings_TMIN = "S5"
+
 #SI figure captions
@@
 climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
-
+tree_rings_TMIN_short_caption <- "Figure S5. ..."
 ```
@@
 ```
+
+\newpage
+```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption}
+#knitr::include_graphics("tables_figures/tree_cores_tmin.tif")
+```

### 2020-12-07 doc/manuscript/Dow_manuscript.Rmd
working on results

@@
 climwin_dp_DOY25_SCBI = "S2"
-climwin_rp_DOY25_SCBI = "S3"
-climwin_dp_DOY25_SCBI = "S4"
+climwin_rp_DOY25_HARV = "S3"
+climwin_dp_DOY25_HARV = "S4"
+tree_rings_TMIN = "S5"
+
 #SI figure captions
@@
 climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
+tree_rings_TMIN_short_caption <- "Figure S5. ..."
 
@@
 `r climwin`, `r all_climwin_SI_figs`).
-This held true for both ring- and diffuse-porous trees at both sites. 
+This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
 The strongest relationships, including but not limited to the optimal time windows (Table `r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.
-(In general, there were significant negative relationships between temperature and $DOY_{25}$ for any time period beginning $\ge$ **5** weeks prior to $DOY_{25}$, ending  $\le$ **8-20** weeks prior depending on wood-type and site, and covering at least a **3**-week period (`r all_climwin_SI_figs`).
-Relationships tended to be weaker and more stochastic in sign for short time periods, particularly those just prior to $DOY_{25}$.)
-(**Cameron, I'm not sure if the above text is up-to-date... ** )
 
@@
 
-*$DOY_{25}$ was generally less strongly correlated with $T_{min}$ than with $T_{max}$.-- not sure this is true*
-(*give some details, cite SI tables.*)
+The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (Figs. `r climwin`, `r all_climwin_SI_figs`). 
+At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (Figs.
+` r climwin_rp_DOY25_SCBI` - `climwin_dp_DOY25_SCBI`).
+In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows, both > 5 weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
+However, for both ring- and diffuse- porous species, average $T_{min}$ of the selected period was below freezing (<0° C) in all years, suggesting that these are sporadic correlations that are unlikely to be biologically meaningful.
 
@@
 
-For both ring- and diffuse- porous species at both sites, warmer spring temperatures resulted in earlier achievement of growth milestones (Tables `r hypothesis_table`, Figs.
+For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Tables `r hypothesis_table`, Figs.
 `r pheno_Tsensitivity`, `r DOYtiming`).
-Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical period (Table `r mean_growth_parameters_table`).
+Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical temperature window (Table `r hypothesis_table`, `r pheno_Tsensitivity`, `r DOYtiming`).
 
-```{r pheno_Tsensitivity, out.width="100%", echo=FALSE, fig.align="center", fig.cap=pheno_Tsensitivity_caption}
+```{r pheno_Tsensitivity, out.width="80%", echo=FALSE, fig.align="center", fig.cap=pheno_Tsensitivity_caption}
 # Created in RScripts/bert/wood_phenology_analysis_bert.R
@@
 
-At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier while ${DOY_75}$ did not shift, resulting in longer peak growing season length when critical window temperatures were warmer (Tables `r hypothesis_table`, Figs.
+For three of the four species group-site combinations (exception: ring-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Tables `r hypothesis_table`, Figs.
 `r pheno_Tsensitivity`, `r DOYtiming`).
-Diffuse-porous $DOY_{25}$, $DOY_{50}$, and ${DOY_75}$ shifted earlier uniformly resulting in no significant change in peak growing season length. 
-At Harvard Forest, ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length, while diffuse-porous ${DOY_25}$ shifted more than ${DOY_75}$, resulting in an extended peak growing season length in response to warmer critical window temperatures.
+At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier as a result of warmer springs whereas ${DOY_75}$ did not shift significantly, resulting in a positive correlation between $L_{PGS}$ and $T_{max}$ during the critical temperature.
+For diffuse-porous species at both SCBI and Harvard Forest, all growth milestones occurred earlier in warmer springs, but $DOY_{25}$ was more strongly affected than ${DOY_75}$, resulting in increased $L_{PGS}$.
+At Harvard Forest,  all growth milestones occurred earlier in warmer springs, but ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length.
 
@@
 Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`). 
-Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer early spring temperatures.
-In contrast, diffuse-porous species exhibited no change (SCBI) or lower $g_{max}$ in years with higher early spring $T_{max}$.
+Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer $T_{max}$ during the critical temperature window.
+In contrast, diffuse-porous species at both sites exhibited lower $g_{max}$ in years with higher early spring $T_{max}$.
 
-Total annual growth measured with dendrometer bands, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes (Harvard Forest) (Tables `r hypothesis_table`, Fig.
+Total annual growth was not consistently linked to spring temperatures (Table `r hypothesis_table`).
+Annual growth measured with dendrometer bands, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest) (Fig.
 `r pheno_Tsensitivity`).
 
-Annual growth recorded in tree-ring records, $RW$, ...
+Similarly, annual growth recorded in tree-ring records, $RW$, rarely displayed a significant correlation to spring  $T_{max}$ (Fig. `r tree_rings` or  $T_{min}$ (Fig. `r tree_rings_TMIN`).
 
 ```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
-# Created in RScripts/additional_figures.R
 knitr::include_graphics("tables_figures/tree_cores_tmax.tif")

### 2020-12-07 doc/manuscript/Dow_manuscript.Rmd
writing- results

@@
 
-Similarly, annual growth recorded in tree-ring records, $RW$, rarely displayed a significant correlation to spring  $T_{max}$ (Fig. `r tree_rings` or  $T_{min}$ (Fig. `r tree_rings_TMIN`).
+Similarly, between species-level residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Figs.
+`r tree_rings`, `r tree_rings_TMIN`).
+Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
+responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig.
+`r tree_rings`).
+In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
+In the prior months, the generally negative correlations to $T_{max}$ were not significant, excepting significant correlations of one species at each site  (ring-porous *Carya tomentosa* in March at SCBI; diffuse-porous *Acer rubrum* in April at Harvard Forest).
+Starting in May at SCBI, significant correlations to $T_{max}$ were common, occurring for nine of the eleven species in May and 3-7 species for each month June through August.
+At Harvard Forest, there was only one statistically significant negative correlation to post-May $T_{max}$ (*Quercus rubra* in June).
 
@@
 ```
+
+Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (Fig. `r tree_rings_TMIN`). 
+There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.
 
@@
 # Acknowledgements
-people: (everyone who helped collect data), Jess Shue (data)
+people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data)
 Funding: ForestGEO

### 2020-12-07 doc/manuscript/tables_figures/tables.Rmd
writing- results

--- 2020-12-04 doc/manuscript/tables_figures/tables.Rmd
@@
 kable(species_table, caption = "Table 2.
-Dominant ring- and diffuse-porous species at SCBI and Harvard Forest analyzed here, along with sample sizes.",  format = "latex", booktabs = TRUE,  escape = F) %>%
+Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes analyzed here.",  format = "latex", booktabs = TRUE,  escape = F) %>%
 add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
+  pack_rows("", 1, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 10, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 12, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 15, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-12-07 doc/manuscript/Dow_manuscript.Rmd
some clean-up of parameters
#37

@@
 Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to $DBH$ at the beginning and end of the year, respectively. 
-$\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote, $r$ is the slope of the curve at the inflection point, $DOY_{ip}$ is the day of year where the inflection point in growth rate occurs, and $DOY$ are the modeled days of year (*i.e.*, julian days).
-*We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow. (???)* 
-Using the optimized parameters for each tree-year, we could then determine the $DBH$ on every day within that year, giving us the ability to extract the intra-annual growth variables of interest (Fig.
-`r schematic`).
+$DOY$ are the modeled days of year (*i.e.*, julian days), 
+$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
+$r$ *shapes the slope of the curve at the inflection point*,
+and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote. 
+*We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic`) occurs on $DOY_{ip}$ only when $\theta = 1$.*
+Using the optimized parameters for each tree-year, we then modeled $DBH$ throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic`).
 
@@
 
-Mixed effects models were used to test the response of growth phenology variables ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{ip}$, $r_{max}$, and $\Delta DBH$) to fixed effects of xylem porosity and mean temperature ($T_{max}$, $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
+Mixed effects models were used to test the response of growth phenology variables ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_max}}$, $g_{max}$, and $\Delta DBH$) to fixed effects of xylem porosity and mean temperature ($T_{max}$, $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
 We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ and $T_{min}$. 

### 2020-12-07 RScripts/Pheno_tsensitivity_figure.R
plotting three separate credible envelopes

@@ -211,11 +211,11 @@ Wood_pheno_table_DP <- subset(Wood_pheno_table, wood_type == "diffuse-porous")
 predictions_RP$sig <- ifelse(predictions_RP$perc == "DOY_75", 1, 0)
 fig6_RP <- ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
-  stat_lineribbon(data = predictions_RP, aes(x = climwinmean, y = predictions_rstanarm, group = perc, col = perc, linetype = as.factor(sig)), .width = c(.99, .95)) +
+  stat_lineribbon(data = predictions_RP, aes(x = climwinmean, y = predictions_rstanarm, col = perc, fill = perc, linetype = as.factor(sig)), .width = .95, alpha = 0.5) +
   geom_point(data = Wood_pheno_table_RP, aes(x = climwinmean, y = DOY, col = perc)) +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   scale_linetype_manual(values = c("solid", "dashed")) +
-  scale_fill_brewer() +
+  # scale_fill_brewer() +
   # facet_grid(perc) +
   coord_cartesian(xlim = c(11.8, 19.7), ylim = c(80, 240)) +
   theme_bw()+

### 2020-12-07 doc/manuscript/Dow_manuscript.Rmd
work on discussion

@@
 
-We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity...
+We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity.
 
-Although the number of years analyzed here is limited (# years at SCBI, # years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (*CITE RELEVANT FIGS & TABLES* ).
-This indicates that this is not a subtle effect.
-However, the effect on L_PGS, r_max, and $\Delta DBH$, if any, is subtle, and a much longer record would be needed to detect such effects. 
-Fortunately, we were able to get at annual growth using tree-ring data (Fig. `r tree_rings`).
-These confirm that the effect of spring temperatures on annual growth is modest and generally not significant, while revealing a slightly more complex picture.
-Although spring temperatures prior to May rarely had a significant influence on ...
+Although the number of years analyzed here is limited (# years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs.
+`r  climwin`-`r DOYtiming`).
+This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
+indeed, growth milestones advanced by #-# days per degree Celsius during the critical temperature window.
+This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*
 
-THe tree-rings also reveal that the strongest temperature sensitivity aligned with peak growing season.
-This pattern is common across tree-ring studies in temperate deciduous forests.
+In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Table `r hypothesis_table`, Fig.
+`r pheno_Tsensitivity`), such that a much longer record would be needed to clarify what, if any, effect spring temperatures have on maximum daily and annual growth rates.
+Fortunately, we were able to use tree-ring data to understand centennial-scale temperature sensitivity of annual growth (Fig. `r tree_rings`).
+This analysis confirms that the effect of spring temperatures on annual growth is modest and not statistically significant for most species, while revealing a slightly more complex picture.
+It appears that temperatures--particularly $T_{max}$--during one spring month at each site (April at SCBI, May at Harvard Forest) consistently have a weakly positive (*albiet* rarely significant) effect on annual growth, in striking contrast to the predominantly negative correlations of growth rates with $T_{max}$ in preceding and subsequent months.
+This suggests that spring $T_{max}$, though it's pronounced influence on the start of woody growth and subtle influence on length of the peak growing season, has a modestly positive influence on annual tree growth, but that this influence is generally of secondary importance to other factors.
+
+Indeed, the tree-ring records reveal that at SCBI the strongest temperature sensitivity aligned with peak growing season [@helcoski_growing_2019], whereas there were fewer significant climate correlations for the species at Harvard Forest (Fig.
+`r tree_rings`).
+Relative to other temperate deciduous forest sites, the pattern at SCBI tends to be more typical: annual growth is most commonly controlled by climate conditions during the peak growing season.
 (*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
@@
 Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
-Warm spring temperatures result in earlier leaf-out (REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (*CITE RELEVANT FIGS & TABLES* ).
-Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
+Warm spring temperatures result in earlier leaf-out (REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (Figs. `r climwin`- `r DOYtiming`).
+Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs. `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
 Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
-As sink limitation controls the end of woody growing season [@zani_increased_2020], we might expect that an earlier start to growth buffers against unfavorable growth conditions during the peak growing season (e.g., drought).
-*Do we see any evidence of this???*
 
@@
 ~~~~
+
 ## Cam's previous content

### 2020-12-08 doc/manuscript/Dow_manuscript.Rmd
update refs
drafted abstract

@@
 
-I like the [Nature summary paragraph template](https://cbs.umn.edu/sites/cbs.umn.edu/files/public/downloads/Annotated_Nature_abstract.pdf):
-
-- One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline.
-- Two to three sentences ofmore detailed background, comprehensible to scientists in related disciplines.
-- One sentence clearly stating the general problem being addressed by this particular study.
-- One sentence summarising the main result (with the words “here we show” or their equivalent).
-- Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge.
-- One or two sentences to put the results into a more general context.
-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline
+As the climate changes, warmer spring temperatures are causing earlier leaf-out in temperate deciduous forests, along with earlier initiation of the annual period where these ecosystems are net sinks of carbon dioxide (CO~2~).
+As a result, there is a tendency towards increased growing season length and annual CO~2~ sequestration.
+However, little is known about how spring temperatures affect tree stem growth, which consumes a modest portion of the carbon fixed annually but is critical in that wood has a long residence time in the ecosystem.
+We need to know how the timing and rate of woody growth are affected by warming spring temperatures.
+Here, we show that in two temperate deciduous forests warmer maximum temperatures in the spring shift growth milestones earlier in the year, but have only a modest effect on peak growing season length and little effect on annual growth.
+This contrasts with the expectation *in most models* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
+It is, however, consistent with most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring, with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length, and with evidence that forest CO~2~ sequestration is sink-limited.
+Thus, our findings indicate that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.
 
@@
 At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (Figs.
-` r climwin_rp_DOY25_SCBI` - `climwin_dp_DOY25_SCBI`).
+`r climwin_rp_DOY25_SCBI` - `r climwin_dp_DOY25_SCBI`).
 In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows, both > 5 weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.

### 2020-12-09 RScripts/Pheno_tsensitivity_figure.R
adjust x-axis limits

@@ -273,7 +273,7 @@ fig6_RP <- ggplot() +
   scale_color_manual(values = c("red","dark green","blue", "purple"))+
   #scale_fill_brewer() +
   # facet_grid(perc) +
-  coord_cartesian(xlim = c(11.8, 19.2), ylim = c(min(Wood_pheno_table_RP$DOY)-3,max(Wood_pheno_table_RP$DOY)+3)) +
+  coord_cartesian(xlim = c(15.2, 25.5), ylim = c(min(Wood_pheno_table_RP$DOY)-3,max(Wood_pheno_table_RP$DOY)+3)) +
   theme_bw()+
   theme(legend.position = "none",
         axis.text.x = element_blank(),
@@ -293,7 +293,7 @@ fig6_DP <- ggplot() +
         axis.text.x = element_blank(),
         text = element_text(size = 20)) +
   # facet_grid(perc) +
-  coord_cartesian(xlim = c(14.7, 19.2), ylim = c(min(Wood_pheno_table_DP$DOY)-3,max(Wood_pheno_table_DP$DOY)+3)) +
+  coord_cartesian(xlim = c(17.3, 20.1), ylim = c(min(Wood_pheno_table_DP$DOY)-3,max(Wood_pheno_table_DP$DOY)+3)) +
   labs(x = "", y = "", col = "Percentile", title = "SCBI", subtitle = "Diffuse-porous")
 
 # Clean-up
@@ -337,7 +337,7 @@ fig6_RP_tot <-  ggplot() +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   scale_fill_brewer() +
   # facet_grid(perc) +
-  coord_cartesian(xlim = c(11.8, 19.7), c(min(Wood_pheno_table_RP_tot$tot),max(Wood_pheno_table_RP_tot$tot))) +
+  coord_cartesian(xlim = c(15.2, 25.5), c(min(Wood_pheno_table_RP_tot$tot),max(Wood_pheno_table_RP_tot$tot))) +
   theme_bw()+
   theme(legend.position = "none",
         text = element_text(size = 20)) +
@@ -350,7 +350,7 @@ fig6_DP_tot <-  ggplot() +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   scale_fill_brewer() +
   # facet_grid(perc) +
[…]

### 2020-12-09 doc/manuscript/Dow_manuscript.Rmd
intro work
rearranging for clarity / more focus on stem-phenology instead of leaf phenology

@@
 climwin_caption <- "Figure 3.
-Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
-Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: May 4 at SCBI, May 15 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
-Black boxes indicate the optimal window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
+Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
+Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
+Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
 19- May 21 at SCBI, March 19 - May 07 at HF)."
@@
 observed climwin mean temperature value, the posterior mean of the fitted day of year
-is represented by the solid blue line and credible intervals are represented by bands."
+is represented by the solid blue line and 95% credible intervals are represented by bands."
 
-DOYtiming_caption <- "Figure 5. Day of year where growth milestones were achied at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest."
+DOYtiming_caption <- "Figure 5.
+Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest.
+Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
+Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."
 
 tree_rings_caption <- "Figure 6.
-Climate sensitiivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithonian Conservation Biology Institute and Harvard Forest.
+Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
 Shown are Pearson correlations between ring-width index and and monthly Tmax.
@@
 #SI figure captions
-climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
-climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for DOY 25 for diffuse porous species at SCBI."
-climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for DOY 25 for ring porous species at Harvard Forest."
-climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
+climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
+climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for $DOY_{25}$ for diffuse porous species at SCBI."
+climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for $DOY_{25}$ for ring porous species at Harvard Forest."
+climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
 tree_rings_TMIN_short_caption <- "Figure S5. ..."
@@
 
-Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019).
-Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens.
+Reducing this uncertainty requires better understanding of the drivers of tree growth [@bugmann_tree_2019].
+Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season [@Helcoski_growing_2019]; [@Ciais_europe-wide_2005]; [@Goulden_exchange_1996], as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain [@mathias_disentangling_2018], and invasion by pests or pathogens.
 Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.
@@
 ## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
-In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
-Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;).
-The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*.
-As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011).
-The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).
+In recent decades, an increase in average spring temperature has been impacting many natural systems including forests ([@parmesan_globally_2003]; [@zhizong_fingerprinting_2006]). 
+The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*.
+For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010];@jeong_phenology_2011];@crabbe_extreme_2016].
+The emergence of leaves signals the beginning of carbon sequestration or in other words, the growing season of a tree.
+It is believed that the longer the tree has leaves - or as the growing season is extended - the net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$; @churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].)
 
-Leaf phenology is linked to the ecosystem CO~2~ uptake period, and in turn longer carbon uptake periods have been linked to greater net annual C sequestration (*i.e.*, net ecosystem exchange, $NEE$; @churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].
 However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019].
+This implies that earlier leafout will lead to smaller increases in NEE than changes in other phenological traits like the maximum C uptake rate.
+This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@Helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE.
+A paper by @zani_increased_2020 found that earlier leaf out led to earlier leaf senescence within the deciduous trees in their study, reducing the ability of the trees to sequester significantly higher levels of C.
+Instead they posit that senescence is induced when seasonal carbon-sink capacity of the tree is reached, no matter how extended the potential growing season is.
+This result directly contradicts those previous predictions that warmer Spring temperatures and earlier leaf out will lead to drastically increased C sequestration.
 
-Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, we know little about the effect on tree growth and woody productivity, which can be decoupled from leaf phenology (D'Orangeville et al, in review) and represent only a modest proportion of total C sequestration (Anderson-Teixeira et al.
-in review).
+Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, indirect observations of these effects - like those involving leaf phenology - are not enough to draw firm conclusions on the future of forests.
+Instead, direct observations of intraannual stem-growth should be compared between years with a variety of spring climate conditions so a clearer relationship between climate and stem-growth can be deciphered.
+Looking directly at the stem-growth is advantageous as it allows examination of individual trees as opposed to larger landscape scale patterns obtained through eddy-covariance or remote sensing studies.
+A direct, fine scale, individual based study would allow a look into the effect of warming springtime on stem-growth phenology and the related shifts in growing season length and $NEE$.
 
 ## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
-
-Earlier timing of spring leaf-up and initiation of carbon sequestration could signal that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research has been done on this topic.
+Earlier timing of spring leaf-up and initiation of carbon sequestration signals that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research on the characteristics of stem-growth during this shift has been done.
 In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns.
@@
 I suspect there may be some that we're missing--perhaps auto dendro bands in Europe.
-Neil did say that there's little on early spring T)*
-Warming events in late-spring were shown to shift leaf emergence timing back more than early-spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).
-Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.
+Neil did say that there's little on early spring T)*.
 
-Tree-ring records give some hints that warmer springs may increase annual growth (Helcoski et al.
-2019; probably some more), but tree-rings tell us nothing about the timing of that growth.
-Timing of intraannual stem-growth is extremely important as well.
+Tree-ring records give some hints that warmer springs may increase annual growth ([@Helcoski_growing_2019]; probably some more), but tree-rings tell us nothing about the timing of that growth.
+Similarly, large scale observations of leaf-phenolgy and eddy-covariance also hint that warmer springs increase annual growth but tell us little about the maximum C uptake rates, or interspecific differences in total carbon sequestered.
 
@@
 **).
-For example, bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.
-
-The significance of wood type in regards to wood and leaf phenology likely stems from evolutionary differences and adaptations to local climates where these trees can be found.
+Interspecific variation of this sort likely stems from evolutionary differences and adaptations to local climates where these trees can be found.
 Ring-porous trees have large xylem cells which are prone to embolism during the winter.
-These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst.
+These cells are repaired before leaf development in the Spring, corresponding with an earlier initiaiton of stem-growth.
 Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process.
@@
 (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time.
-If leaf development is sped up when spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.
+If leaf development is sped up when spring temperatures are warmer, we should seek to understand how this shift influences stem-growth timing and other phenological variables.
 
@@
 Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA.
-Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig.
+Using data from dendrometer bands measured throughout the growing season, we fit a growth model to the time series of individual trees to determine the day of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$); maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$); total annual growth ($\Delta DBH$); and peak growing season length ($DOY_{75}$-$DOY_{25}$; $L_{pgs}$) [Fig.
 `r schematic`; @mcmahon_general_2015].
-We test the hypotheses that (1) warmer early springs result in earlier stem growth and a period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).
+We test the hypotheses that (1) warmer early springs result in earlier stem growth and a longer period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).
 
@@
 Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*). 
-The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July (Helcoski et al, 2019).
+The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July ([@Helcoski_growing_2019]).
 
@@
 
-Although the number of years analyzed here is limited (# years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs.
+Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs.
 `r  climwin`-`r DOYtiming`).

### 2020-12-09 RScripts/Pheno_tsensitivity_figure.R
corrected linetype for sig

@@ -20,7 +20,7 @@ library(rstanarm)
 library(patchwork)
 # Number of MCMC chains & number of simulations per chain.
 # Need to increase this at the end
-n_iter <- 10000
+n_iter <- 100
 n_chains <- 2
 
 # Objects to keep during memory management clean-up
@@ -215,7 +215,7 @@ predictions_RP <- filter(predictions, wood_type == "ring-porous")
 predictions_DP <- filter(predictions, wood_type == "diffuse-porous")
 Wood_pheno_table_RP <- filter(Wood_pheno_table, wood_type == "ring-porous")
 Wood_pheno_table_DP <- filter(Wood_pheno_table, wood_type == "diffuse-porous")
-predictions_RP$sig <- ifelse(predictions_RP$perc == "DOY_75", 1, 0)
+predictions_RP$sig <- ifelse(predictions_RP$perc == "DOY_50", 1, 0)
 
 woodtable <- filter(Wood_pheno_table, perc == "DOY_25")
 
@@ -246,7 +246,7 @@ predictions_mrdoy <- woodtable %>%
 
 predictions_mrdoy_RP <- filter(predictions_mrdoy, wood_type == "ring-porous")
 predictions_mrdoy_RP$perc <- "Max Rate DOY"
-predictions_mrdoy_RP$sig <- 0
+predictions_mrdoy_RP$sig <- 1
 predictions_RP <- bind_rows(predictions_RP, predictions_mrdoy_RP)
 predictions_mrdoy_DP <- filter(predictions_mrdoy, wood_type == "diffuse-porous")
 predictions_mrdoy_DP$perc <- "Max Rate DOY"
@@ -407,7 +407,7 @@ fig6_RP_sl <- ggplot() +
 
 fig6_DP_sl <- ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
-  stat_lineribbon(data = predictions_sl_DP, aes(x = climwinmean, y = predictions_rstanarm), .width = .95, linetype = "solid") +
+  stat_lineribbon(data = predictions_sl_DP, aes(x = climwinmean, y = predictions_rstanarm), .width = .95, linetype = "dashed") +
   geom_point(data = Wood_pheno_table_DP_sl, aes(x = climwinmean, y = seasonlength)) +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   scale_fill_brewer() +
[…]

### 2020-12-10 doc/manuscript/Dow_manuscript.Rmd
Update Dow_manuscript.Rmd

@@
 We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity.
+Using dendroband data, we found that the growing period is generally shifting back in the season as the critical srping windows are warming, but having little, or negative effect on peak growing season duration.
+Maximum growth rates and total seasonal growth were virtually unchanged when this period warmed, but the Day of year where the maximum growth rate occured shitfed back almost in unison with the DOY_50.
+The lack of influence Spring TMAX exudes on total season growth found with the dendroband data agrees with patterns observed in tree cores.
+Annual tree growth depends more on peak growing season conditions than early season advancement.
 
@@
 This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
-indeed, growth milestones advanced by #-# days per degree Celsius during the critical temperature window.
+indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
 This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*

### 2020-12-10 doc/manuscript/Dow_manuscript_Nature.Rmd
start reformatting for Nature
#41
@camerondow35 , I created a new version rather than overwriting the GCB version

--- 2020-12-10 doc/manuscript/Dow_manuscript.Rmd
@@
 bibliography: references.bib
-csl: global-change-biology.csl
+csl: nature.csl
 ---
@@
 
-## Target journal: Global Change Biology
-## Instructions for authors: https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/page/journal/13652486/homepage/forauthors.html
-## Primary Research Articles present the results of a completed research project and are up to 8000 words in length.
-Word limits apply to the main body of the text (Introduction, Materials and Methods, Results, Discussion, and Acknowledgements).
+## Target journal: Nature
 
-## I don't think GCB has an official limit. I just saw a recently accepted article with 10 figures and 4 tables, so we're fine!
+## Nature articles: 3-4 modest display items
 
@@
 hypothesis_table=1
-table_species=2
+table_species=2  #goes to SI
 mean_growth_parameters_table=3
@@
 schematic=1 #illustrates parameters
-growth_curves_all = 2  
+growth_curves_all = 2  #goes to SI
     rel_growth_SCBI = "2a" 
@@
     cum_growth_HF = "2d" 
-climwin = 3
+climwin = 3 # probably stays?
 pheno_Tsensitivity = 4 
@@
 DOYtiming_Harvard = "5b"
-tree_rings = 6
+tree_rings = 6  #stays, hopefully adding a lot more sites
 
@@
 Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
-Each line represents one year's growth for a given tree, fit with McMahon model."
+Each line represents one year's growth for a given tree, fit with McMahon model.
+(THIS FIGURE GOES TO SI.)"
 
@@
 Shown are Pearson correlations between ring-width index and and monthly Tmax.
-Species codes are as given in Table 2."
+Species codes are as given in Table 2.(THIS FIGURE STAYS, PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
 
@@
 #SI table captions
-sample_size_by_year_caption <- "Table S1. Sample size by year"
+sample_size_by_year_caption <- "Table S1. Sample size by year (THIS TABLE GOES TO SI)"
 
@@
 
-# Abstract
+# (Summary paragraph)
 ```{r eval=FALSE, echo=FALSE}
-## For GCB:
-## Abstract limit for GCB is 300 words. 
-## 6-10 keywords
+## For Nature:
+# Articles start with a fully referenced summary paragraph, ideally of no more than 200 words, which is separate from the main text and avoids numbers, abbreviations, acronyms or measurements unless essential.
+It is aimed at readers outside the discipline.
+This summary paragraph should be structured as follows: 2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
+Please refer to our annotated example  to see how the summary paragraph should be constructed.
 
 ```
+
+(*word limit 200; this is currently 266*)
 
@@
 This contrasts with the expectation *in most models* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-It is, however, consistent with most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring, with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length, and with evidence that forest CO~2~ sequestration is sink-limited.
+It is, however, consistent with most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[@helcoski_growing_2019], with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
 Thus, our findings indicate that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.
-
-**Keywords**:
 
@@
 
-# Introduction
+# (Introduction)
+
+(*this will need to be shortened signficantly*)
 
@@
 Reducing this uncertainty requires better understanding of the drivers of tree growth [@bugmann_tree_2019].
-Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season [@Helcoski_growing_2019]; [@Ciais_europe-wide_2005]; [@Goulden_exchange_1996], as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain [@mathias_disentangling_2018], and invasion by pests or pathogens.
+Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season [@helcoski_growing_2019]; [@Ciais_europe-wide_2005]; [@Goulden_exchange_1996], as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain [@mathias_disentangling_2018], and invasion by pests or pathogens.
 Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.
@@
 ## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
-In recent decades, an increase in average spring temperature has been impacting many natural systems including forests ([@parmesan_globally_2003]; [@zhizong_fingerprinting_2006]). 
+In recent decades, an increase in average spring temperature has been impacting many natural systems including forests [@parmesan_globally_2003]; [@zhizong_fingerprinting_2006]. 
 The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*.
@@
 This implies that earlier leafout will lead to smaller increases in NEE than changes in other phenological traits like the maximum C uptake rate.
-This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@Helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE.
+This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE.
 A paper by @zani_increased_2020 found that earlier leaf out led to earlier leaf senescence within the deciduous trees in their study, reducing the ability of the trees to sequester significantly higher levels of C.
@@
 
-# Materials and Methods
+# (Results)
+
+## (Phenological patterns)
+
+(*most of this is too detailed for Nature*)
+
+Across all trees and years at SCBI, 
+25% of annual growth was achieved on May 20 ($DOY_{25}$=139), 
+50% on June 12 ($DOY_{50}$=162), 
+and 75% on July 5 ($DOY_{75}$=185), 
+for an average peak growing season length of 45 days. 
+Maximum growth rate occurred just one day after $DOY_{50}$ on June 13 ($DOY_{max}$=163). 
+At Harvard Forest,
+25% of annual growth was achieved on June 1 ($DOY_{25}$=151), 
+50% on June 23 ($DOY_{50}$=173), 
+and 75% on July 14 ($DOY_{75}$=194), 
+for an average peak growing season length of 43 days. 
+Maximum growth rate occurred on June 24 ($DOY_{max}$=174).
+
+At both sites, growth phenology differed between ring- and diffuse- porous species (Fig. `r growth_curves_all`, Table `r mean_growth_parameters_table`). 
+Specifically, ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark `r 154-123` days earlier at SCBI and `r 164-132` days earlier at Harvard Forest. 
+Growth of ring-porous species was spread out over a longer growing season, with average $L_{PGS}$ of ring-porous species being 21 days longer at SCBI and 19 days longer at Harvard Forest. 
+Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.
+
+```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/growth_curves_all.png")
+```
+
+```{r mean_growth_parameters_table, out.width="100%", echo=FALSE, fig.align="center"}
+knitr::include_graphics("tables_figures/mean_growth_parameters_table.png")
+```
+
+## (Critical temperature windows)
+
+For the majority of weekly time windows from Jan.
+1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
+`r climwin`, `r all_climwin_SI_figs`).
+This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
+The strongest relationships, including but not limited to the optimal time windows (Table `r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.
+
+```{r climwin, out.width="80%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/climwin_figure.png")
+```
+
+The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (Figs. `r climwin`, `r all_climwin_SI_figs`). 
+At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (Figs.
+`r climwin_rp_DOY25_SCBI` - `r climwin_dp_DOY25_SCBI`).
+In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows, both > 5 weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
+However, for both ring- and diffuse- porous species, average $T_{min}$ of the selected period was below freezing (<0° C) in all years, suggesting that these are sporadic correlations that are unlikely to be biologically meaningful.
+
+Henceforth, we focus on $T_{max}$.
+
+## (Climate sensitivity of phenology )
+
+For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Tables `r hypothesis_table`, Figs.
+`r pheno_Tsensitivity`, `r DOYtiming`).
+Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical temperature window (Table `r hypothesis_table`, `r pheno_Tsensitivity`, `r DOYtiming`).
+
+```{r pheno_Tsensitivity, out.width="80%", echo=FALSE, fig.align="center", fig.cap=pheno_Tsensitivity_caption}
+# Created in RScripts/bert/wood_phenology_analysis_bert.R
+knitr::include_graphics("tables_figures/pheno_Tsensitivity_combo.png")
+```
+
+```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
+```
+
+For three of the four species group-site combinations (exception: ring-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Tables `r hypothesis_table`, Figs.
+`r pheno_Tsensitivity`, `r DOYtiming`).
+At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier as a result of warmer springs whereas ${DOY_75}$ did not shift significantly, resulting in a positive correlation between $L_{PGS}$ and $T_{max}$ during the critical temperature.
+For diffuse-porous species at both SCBI and Harvard Forest, all growth milestones occurred earlier in warmer springs, but $DOY_{25}$ was more strongly affected than ${DOY_75}$, resulting in increased $L_{PGS}$.
+At Harvard Forest,  all growth milestones occurred earlier in warmer springs, but ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length.
+
+## (Climate sensitivity of growth)
+
+Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`). 
+Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer $T_{max}$ during the critical temperature window.
+In contrast, diffuse-porous species at both sites exhibited lower $g_{max}$ in years with higher early spring $T_{max}$.
+
+Total annual growth was not consistently linked to spring temperatures (Table `r hypothesis_table`).
+Annual growth measured with dendrometer bands, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest) (Fig.
+`r pheno_Tsensitivity`).
+
+Similarly, between species-level residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Figs.
+`r tree_rings`, `r tree_rings_TMIN`).
+Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
+responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig.
+`r tree_rings`).
+In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
+In the prior months, the generally negative correlations to $T_{max}$ were not significant, excepting significant correlations of one species at each site  (ring-porous *Carya tomentosa* in March at SCBI; diffuse-porous *Acer rubrum* in April at Harvard Forest).
+Starting in May at SCBI, significant correlations to $T_{max}$ were common, occurring for nine of the eleven species in May and 3-7 species for each month June through August.
+At Harvard Forest, there was only one statistically significant negative correlation to post-May $T_{max}$ (*Quercus rubra* in June).
+
+```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
+knitr::include_graphics("tables_figures/tree_cores_tmax.tif")
+```
+
+Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (Fig. `r tree_rings_TMIN`). 
+There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.
+
+# (Discussion)
+
+(*length may be okay, shouldn't be lengthened*)
+
+We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity.
+Using dendroband data, we found that the growing period is generally shifting back in the season as the critical srping windows are warming, but having little, or negative effect on peak growing season duration.
+Maximum growth rates and total seasonal growth were virtually unchanged when this period warmed, but the Day of year where the maximum growth rate occured shitfed back almost in unison with the DOY_50.
+The lack of influence Spring TMAX exudes on total season growth found with the dendroband data agrees with patterns observed in tree cores.
+Annual tree growth depends more on peak growing season conditions than early season advancement.
+
+Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs.
+`r  climwin`-`r DOYtiming`).
+This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
+indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
+This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*
+
+In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Table `r hypothesis_table`, Fig.
+`r pheno_Tsensitivity`), such that a much longer record would be needed to clarify what, if any, effect spring temperatures have on maximum daily and annual growth rates.
+Fortunately, we were able to use tree-ring data to understand centennial-scale temperature sensitivity of annual growth (Fig. `r tree_rings`).
+This analysis confirms that the effect of spring temperatures on annual growth is modest and not statistically significant for most species, while revealing a slightly more complex picture.
+It appears that temperatures--particularly $T_{max}$--during one spring month at each site (April at SCBI, May at Harvard Forest) consistently have a weakly positive (*albiet* rarely significant) effect on annual growth, in striking contrast to the predominantly negative correlations of growth rates with $T_{max}$ in preceding and subsequent months.
+This suggests that spring $T_{max}$, though it's pronounced influence on the start of woody growth and subtle influence on length of the peak growing season, has a modestly positive influence on annual tree growth, but that this influence is generally of secondary importance to other factors.
+
+Indeed, the tree-ring records reveal that *annual growth is most commonly controlled by climate conditions during the peak growing season*  [Fig. `r tree_rings`; @helcoski_growing_2019], .
+(*1-2 sentences reviewing our regional analysis results (HOPEFULLY) generalizing about climate sensitivity of temperate deciduous forest growth.*)
+Thus, although our dendrometer band analysis is limited to two sites, the finding that spring temperatures have limited effect on annual tree growth holds generally.
+
+Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
+Warm spring temperatures result in earlier leaf-out (fig., REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (Figs.
+`r climwin`- `r DOYtiming`).
+Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs. `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
+Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
+
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models [Table `r hypothesis_table`; Figs.
+`r pheno_Tsensitivity`, `r tree_rings`; @zani_increased_2020].
+The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.
+
+~~~~
+
+## Cam's previous content
+
+We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures.
+The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here.
+Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,).
+On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al).
+These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011).
+The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likelihood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
+The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014).
+The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014).
+Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994).
+This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
+
+The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005).
+However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.
+*READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early spring temperature signals that the mechanism is influenced by temperature?
+
+Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!!
+Changed by constant amount !!!
+.
+This implies that the growing season of ring-porous trees expanded due to spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
+In fact, we found a positive correlation between ring-porous growing season length and spring temperature, while no such relationship was observed in diffuse-porous trees.
+The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
+Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens.
+Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.
+
+Diffuse-porous growth timing seems relatively unaffected by early spring temperatures, instead showing a higher correlation with late spring or Summer temps.
+This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas.
+---- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015).
+The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.
+
+--the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades?
+Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)
+
+This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg.
+Here, we focused our analysis purely on early spring warming.
+This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth.
+For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June $T_{MAX}$ on trees of that wood type (figure or table?).
+This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year.
+Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high $T_{MAX}$ in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
+How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
+
+We present a stepping stone for future studies to build upon.
+
+# (Content to incorporate)
+
+new Science paper: @zani_increased_2020
+
+@delgado_differences_2020
+
+From Neil: @zhang_drought_2021, @zhang_extended_2020
+
+effect of spring temperature on annual growth in New England: @alexander_potential_2019
+
+results align with @fu_maximum_2019, which shows that interannual variation in NEE is more strongly linked to peak C uptake rates than to growing season length.
+Our study demonstrates that this holds true for $ANPP_{woody}$ as well as NEE.
+
+Geng et al. 2020 GCB- Climate warming increases spring phenological differences among temperate trees.
+
+N.
+American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts.
+Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?
+- - RP trees BB = 25% and 75% leaf development = 50% growth (D'Orangeville).
+Are DP trees leafing up earlier?
+If yes, how much earlier?
+Similar to 25,50,75% DOY retreat?
+
+Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
+
+The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; D’Orangeville et al.
+in review).
+Species have co-evolved to utilize the phenological stages of trees at specific times in the year.
+If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees ().
+Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).
+
+# Methods
 
@@
 
-# Results
-
-## Phenological patterns
-
-Across all trees and years at SCBI, 
-25% of annual growth was achieved on May 20 ($DOY_{25}$=139), 
-50% on June 12 ($DOY_{50}$=162), 
-and 75% on July 5 ($DOY_{75}$=185), 
-for an average peak growing season length of 45 days. 
-Maximum growth rate occurred just one day after $DOY_{50}$ on June 13 ($DOY_{max}$=163). 
-At Harvard Forest,
-25% of annual growth was achieved on June 1 ($DOY_{25}$=151), 
-50% on June 23 ($DOY_{50}$=173), 
-and 75% on July 14 ($DOY_{75}$=194), 
-for an average peak growing season length of 43 days. 
-Maximum growth rate occurred on June 24 ($DOY_{max}$=174).
-
-At both sites, growth phenology differed between ring- and diffuse- porous species (Fig. `r growth_curves_all`, Table `r mean_growth_parameters_table`). 
-Specifically, ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark `r 154-123` days earlier at SCBI and `r 164-132` days earlier at Harvard Forest. 
-Growth of ring-porous species was spread out over a longer growing season, with average $L_{PGS}$ of ring-porous species being 21 days longer at SCBI and 19 days longer at Harvard Forest. 
-Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.
-
-```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/growth_curves_all.png")
-```
-
-```{r mean_growth_parameters_table, out.width="100%", echo=FALSE, fig.align="center"}
-knitr::include_graphics("tables_figures/mean_growth_parameters_table.png")
-```
-
-## Critical temperature windows
-
-For the majority of weekly time windows from Jan.
-1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
-`r climwin`, `r all_climwin_SI_figs`).
-This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
-The strongest relationships, including but not limited to the optimal time windows (Table `r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.
-
-```{r climwin, out.width="80%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/climwin_figure.png")
-```
-
-The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (Figs. `r climwin`, `r all_climwin_SI_figs`). 
-At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (Figs.
-`r climwin_rp_DOY25_SCBI` - `r climwin_dp_DOY25_SCBI`).
-In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows, both > 5 weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
-However, for both ring- and diffuse- porous species, average $T_{min}$ of the selected period was below freezing (<0° C) in all years, suggesting that these are sporadic correlations that are unlikely to be biologically meaningful.
-
-Henceforth, we focus on $T_{max}$.
-
-## Climate sensitivity of phenology
-
-For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Tables `r hypothesis_table`, Figs.
-`r pheno_Tsensitivity`, `r DOYtiming`).
-Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical temperature window (Table `r hypothesis_table`, `r pheno_Tsensitivity`, `r DOYtiming`).
-
-```{r pheno_Tsensitivity, out.width="80%", echo=FALSE, fig.align="center", fig.cap=pheno_Tsensitivity_caption}
-# Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/pheno_Tsensitivity_combo.png")
-```
-
-```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
-```
-
-For three of the four species group-site combinations (exception: ring-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Tables `r hypothesis_table`, Figs.
-`r pheno_Tsensitivity`, `r DOYtiming`).
-At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier as a result of warmer springs whereas ${DOY_75}$ did not shift significantly, resulting in a positive correlation between $L_{PGS}$ and $T_{max}$ during the critical temperature.
-For diffuse-porous species at both SCBI and Harvard Forest, all growth milestones occurred earlier in warmer springs, but $DOY_{25}$ was more strongly affected than ${DOY_75}$, resulting in increased $L_{PGS}$.
-At Harvard Forest,  all growth milestones occurred earlier in warmer springs, but ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length.
-
-## Climate sensitivity of growth
-
-Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`). 
-Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer $T_{max}$ during the critical temperature window.
-In contrast, diffuse-porous species at both sites exhibited lower $g_{max}$ in years with higher early spring $T_{max}$.
-
-Total annual growth was not consistently linked to spring temperatures (Table `r hypothesis_table`).
-Annual growth measured with dendrometer bands, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest) (Fig.
-`r pheno_Tsensitivity`).
-
-Similarly, between species-level residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Figs.
-`r tree_rings`, `r tree_rings_TMIN`).
-Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
-responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig.
-`r tree_rings`).
-In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
-In the prior months, the generally negative correlations to $T_{max}$ were not significant, excepting significant correlations of one species at each site  (ring-porous *Carya tomentosa* in March at SCBI; diffuse-porous *Acer rubrum* in April at Harvard Forest).
-Starting in May at SCBI, significant correlations to $T_{max}$ were common, occurring for nine of the eleven species in May and 3-7 species for each month June through August.
-At Harvard Forest, there was only one statistically significant negative correlation to post-May $T_{max}$ (*Quercus rubra* in June).
-
-```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
-knitr::include_graphics("tables_figures/tree_cores_tmax.tif")
-```
-
-Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (Fig. `r tree_rings_TMIN`). 
-There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.
-
-# Discussion
-
-We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity.
-Using dendroband data, we found that the growing period is generally shifting back in the season as the critical srping windows are warming, but having little, or negative effect on peak growing season duration.
-Maximum growth rates and total seasonal growth were virtually unchanged when this period warmed, but the Day of year where the maximum growth rate occured shitfed back almost in unison with the DOY_50.
-The lack of influence Spring TMAX exudes on total season growth found with the dendroband data agrees with patterns observed in tree cores.
-Annual tree growth depends more on peak growing season conditions than early season advancement.
-
-Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs.
-`r  climwin`-`r DOYtiming`).
-This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
-indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
-This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*
-
-In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Table `r hypothesis_table`, Fig.
-`r pheno_Tsensitivity`), such that a much longer record would be needed to clarify what, if any, effect spring temperatures have on maximum daily and annual growth rates.
-Fortunately, we were able to use tree-ring data to understand centennial-scale temperature sensitivity of annual growth (Fig. `r tree_rings`).
-This analysis confirms that the effect of spring temperatures on annual growth is modest and not statistically significant for most species, while revealing a slightly more complex picture.
-It appears that temperatures--particularly $T_{max}$--during one spring month at each site (April at SCBI, May at Harvard Forest) consistently have a weakly positive (*albiet* rarely significant) effect on annual growth, in striking contrast to the predominantly negative correlations of growth rates with $T_{max}$ in preceding and subsequent months.
-This suggests that spring $T_{max}$, though it's pronounced influence on the start of woody growth and subtle influence on length of the peak growing season, has a modestly positive influence on annual tree growth, but that this influence is generally of secondary importance to other factors.
-
-Indeed, the tree-ring records reveal that at SCBI the strongest temperature sensitivity aligned with peak growing season [@helcoski_growing_2019], whereas there were fewer significant climate correlations for the species at Harvard Forest (Fig.
-`r tree_rings`).
-Relative to other temperate deciduous forest sites, the pattern at SCBI tends to be more typical: annual growth is most commonly controlled by climate conditions during the peak growing season.
-(*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
-Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree growth will hold generally.
-
-Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
-Warm spring temperatures result in earlier leaf-out (REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (Figs. `r climwin`- `r DOYtiming`).
-Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs. `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
-Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
-
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models [Table `r hypothesis_table`; Figs.
-`r pheno_Tsensitivity`, `r tree_rings`; @zani_increased_2020].
-The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.
-
-~~~~
-
-## Cam's previous content
-
-We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures.
-The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here.
-Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,).
-On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al).
-These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011).
-The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likelihood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
-The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014).
-The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014).
-Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994).
-This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
-
-The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005).
-However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.
-*READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early spring temperature signals that the mechanism is influenced by temperature?
-
-Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!!
-Changed by constant amount !!!
-.
-This implies that the growing season of ring-porous trees expanded due to spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
-In fact, we found a positive correlation between ring-porous growing season length and spring temperature, while no such relationship was observed in diffuse-porous trees.
-The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
-Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens.
-Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.
-
-Diffuse-porous growth timing seems relatively unaffected by early spring temperatures, instead showing a higher correlation with late spring or Summer temps.
-This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas.
----- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015).
-The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.
-
---the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades?
-Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)
-
-This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg.
-Here, we focused our analysis purely on early spring warming.
-This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth.
-For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June $T_{MAX}$ on trees of that wood type (figure or table?).
-This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year.
-Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high $T_{MAX}$ in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
-How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
-
-We present a stepping stone for future studies to build upon.
-
-# (Content to incorporate)
-
-new Science paper: @zani_increased_2020
-
-@delgado_differences_2020
-
-From Neil: @zhang_drought_2021, @zhang_extended_2020
-
-effect of spring temperature on annual growth in New England: @alexander_potential_2019
-
-results align with @fu_maximum_2019, which shows that interannual variation in NEE is more strongly linked to peak C uptake rates than to growing season length.
-Our study demonstrates that this holds true for $ANPP_{woody}$ as well as NEE.
-
-Geng et al. 2020 GCB- Climate warming increases spring phenological differences among temperate trees.
-
-N.
-American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts.
-Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?
-- - RP trees BB = 25% and 75% leaf development = 50% growth (D'Orangeville).
-Are DP trees leafing up earlier?
-If yes, how much earlier?
-Similar to 25,50,75% DOY retreat?
-
-Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
-
-The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; D’Orangeville et al.
-in review).
-Species have co-evolved to utilize the phenological stages of trees at specific times in the year.
-If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees ().
-Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).
-
 # Acknowledgements

### 2020-12-10 doc/manuscript/Dow_manuscript_Nature.Rmd
update

@@
 table_species=2  #goes to SI
-mean_growth_parameters_table=3
+mean_growth_parameters_table=3  #probably goes to SI
 
@@
 
-(*most of this is too detailed for Nature*)
+(*most or all of this should be cut-- first paragraph too detailed for Nature, and we want minimal comparison of ring vs diffuse porous*)
 

### 2020-12-10 doc/manuscript/Dow_manuscript_Nature.Rmd
small changes

@@
 
-Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, indirect observations of these effects - like those involving leaf phenology - are not enough to draw firm conclusions on the future of forests.
-Instead, direct observations of intraannual stem-growth should be compared between years with a variety of spring climate conditions so a clearer relationship between climate and stem-growth can be deciphered.
-Looking directly at the stem-growth is advantageous as it allows examination of individual trees as opposed to larger landscape scale patterns obtained through eddy-covariance or remote sensing studies.
-A direct, fine scale, individual based study would allow a look into the effect of warming springtime on stem-growth phenology and the related shifts in growing season length and $NEE$.
-
 ## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
@@
 Tree-ring records give some hints that warmer springs may increase annual growth ([@Helcoski_growing_2019]; probably some more), but tree-rings tell us nothing about the timing of that growth.
-Similarly, large scale observations of leaf-phenolgy and eddy-covariance also hint that warmer springs increase annual growth but tell us little about the maximum C uptake rates, or interspecific differences in total carbon sequestered.
+On the other hand, earlier leaf development shows us how the growing season is advancing, but can't tell us about the total carbon sequestered, while neither can be used to map out intra-annual growth patterns.
+
+Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on the future of forests.A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals.
+A mixture of these three would produce the most complete picture of spring climate influence on forests.
 
 ## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
+*(Too much detail here? how about just keeping some variation of the first and last sentences)*
 Leaf and wood phenology traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al.
@@
 We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity.
-Using dendroband data, we found that the growing period is generally shifting back in the season as the critical srping windows are warming, but having little, or negative effect on peak growing season duration.
-Maximum growth rates and total seasonal growth were virtually unchanged when this period warmed, but the Day of year where the maximum growth rate occured shitfed back almost in unison with the DOY_50.
-The lack of influence Spring TMAX exudes on total season growth found with the dendroband data agrees with patterns observed in tree cores.
+Using dendroband data, we found that the peak growing season is generally advancing as the critical spring windows are warming but having little, or negative effect, on $L_{pgs}$.
+$g_{max}$ and $\Delta DBH$ were virtually unchanged when this period warmed, but $DOY_{g_{max}}$ shifted back almost in unison with $DOY_{50}$.
+The lack of influence Spring TMAX exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores.
 Annual tree growth depends more on peak growing season conditions than early season advancement.
@@
 The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.
-
-~~~~
-
-## Cam's previous content
-
-We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures.
-The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here.
-Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,).
-On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al).
-These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011).
-The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likelihood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994).
-The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014).
-The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014).
-Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994).
-This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
-
-The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005).
-However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.
-*READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early spring temperature signals that the mechanism is influenced by temperature?
-
-Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!!
-Changed by constant amount !!!
-.
-This implies that the growing season of ring-porous trees expanded due to spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies.
-In fact, we found a positive correlation between ring-porous growing season length and spring temperature, while no such relationship was observed in diffuse-porous trees.
-The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990).
-Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens.
-Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.
-
-Diffuse-porous growth timing seems relatively unaffected by early spring temperatures, instead showing a higher correlation with late spring or Summer temps.
-This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas.
----- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015).
-The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.
-
---the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades?
-Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)
-
-This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg.
-Here, we focused our analysis purely on early spring warming.
-This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth.
-For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June $T_{MAX}$ on trees of that wood type (figure or table?).
-This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year.
-Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high $T_{MAX}$ in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015).
-How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.
-
-We present a stepping stone for future studies to build upon.
 

### 2020-12-10 doc/manuscript/Dow_SI.Rmd
rearranged display items
moved several figures and tables to SI, auto numbering

#41

--- 2020-12-06 doc/manuscript/Dow_SI.Rmd
@@
 - \usepackage{xcolor}
+- \usepackage{float}
+- \floatplacement{figure}{H}
 ---
@@
 #SI table numbers
-sample_size_by_year= "S1"
+table_species="S1"  
+sample_size_by_year= "S2"
+mean_growth_parameters_table="S3"
 
 #SI table captions
-sample_size_by_year_caption <- "Table S1. Sample size by year"
+sample_size_by_year_caption <- "Table S1. Sample size by year "
 
 #SI figure numbers
-climwin_rp_DOY25_SCBI = "S1"
-climwin_dp_DOY25_SCBI = "S2"
-climwin_rp_DOY25_HARV = "S3"
-climwin_dp_DOY25_HARV = "S4"
+growth_curves_all = "S1"  
+all_climwin_SI_figs = " S2-S5"
+  climwin_rp_DOY25_SCBI = "S2"
+  climwin_dp_DOY25_SCBI = "S3"
+  climwin_rp_DOY25_HARV = "S4"
+  climwin_dp_DOY25_HARV = "S5"
+DOYtiming = "S6"
+  DOYtiming_SCBI = "S6a"
+  DOYtiming_Harvard = "S6b"
+tree_rings_TMIN = "S7"
+
 #SI figure captions
-climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
-climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for DOY 25 for diffuse porous species at SCBI."
-climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for DOY 25 for ring porous species at Harvard Forest."
-climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
+growth_curves_all_caption  <- "Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
+Each line represents one year's growth for a given tree, fit with McMahon model.
+"
+
+climwin_rp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
+climwin_dp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at SCBI."
+climwin_rp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at Harvard Forest."
+climwin_dp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
+DOYtiming_caption <- "Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest.
+Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
+Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."
+tree_rings_TMIN_short_caption <- "Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
+Shown are Pearson correlations between ring-width index and and monthly Tmin.
+Species codes are as given in Table 2.(THIS FIGURE WILL PROBABLY BE PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
 
 ```
+
+\newpage
+
+```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+species_table <- read.csv(file = "tables_figures/species_table.csv", col.names = c("site", "xylem porosity","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
+kable(species_table, caption = "Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes analyzed here.",  format = "latex", booktabs = TRUE,  escape = F) %>%
+add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
+  pack_rows("", 1, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 10, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 12, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 15, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
+
+\newpage
+
+```{r, Table SCBI_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
+library(knitr)
+library(kableExtra)
+table <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
+kbl(table, caption = "SCBI bands by species", format = "latex", booktabs = T) %>%
+  kable_styling(latex_options = "hold_position")
+```
+
+\newpage
 
@@
 table <- read.csv("tables_figures/HF_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, format = "latex", booktabs = T)
+kbl(table, caption = "HF bands by species", format = "latex", booktabs = T) %>%
+  kable_styling(latex_options = "hold_position")
 ```
@@
 \newpage
-```{r, Table SCBI_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
+
+```{r, Table sample_size_by_year, eval=TRUE, echo=FALSE, message=FALSE, fig.pos="H"}
 library(knitr)
 library(kableExtra)
-table <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, format = "latex", booktabs = T)
+table <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
+kbl(table, caption = "sample size by year", format = "latex", booktabs = T) %>%
+  kable_styling(latex_options = "hold_position")
 ```
@@
 \newpage
-```{r, Table sample_size_by_year, eval=TRUE, echo=FALSE, message=FALSE}
+
+```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
 library(kableExtra)
-table <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, format = "latex", booktabs = T)
+mean_growth_parameters <- read.csv("tables_figures/mean_growth_parameters.csv",  col.names = c("", "ring","diffuse","ring", "diffuse"), stringsAsFactors = FALSE, check.names = FALSE)
+kable(mean_growth_parameters, caption = "Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest.
+",  format = "latex", booktabs = TRUE,  escape = F) %>%
+add_header_above(c(" ",  "SCBI" = 2, "Harvard Forest"=2 )) %>%
+    column_spec(1, width = "4cm") %>%
+kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
+
+\newpage
+
+```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption, fig.pos="H"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/growth_curves_all.png")
 ```
@@
 **(a) $T_{max}$**
-```{r climwin_rp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left"}
+```{r climwin_rp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
 knitr::include_graphics("tables_figures/SCBI_ring-porous_TMAX.png")
@@
 **(b) $T_{min}$**
-```{r climwin_rp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_SCBI_short_caption}
+```{r climwin_rp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_SCBI_short_caption, fig.pos="H"}
 knitr::include_graphics("tables_figures/SCBI_ring-porous_TMIN.png")
@@
 **(a) $T_{max}$**
-```{r climwin_dp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left"}
+```{r climwin_dp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
 knitr::include_graphics("tables_figures/SCBI_diffuse-porous_TMAX.png")
@@
 **(b) $T_{min}$**
-```{r climwin_dp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_SCBI_short_caption}
+```{r climwin_dp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_SCBI_short_caption, fig.pos="H"}
 knitr::include_graphics("tables_figures/SCBI_diffuse-porous_TMIN.png")
@@
 **(a) $T_{max}$**
-```{r climwin_rp_tmax_HARV, out.width="100%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_HARV_short_caption}
-knitr::include_graphics("tables_figures/HF_ring-porous_DOY25.png")
+```{r climwin_rp_tmax_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
+knitr::include_graphics("tables_figures/HF_ring-porous_TMAX.png")
 ```
@@
 **(b) $T_{min}$**
+```{r climwin_rp_tmin_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_HARV_short_caption, fig.pos="H"}
+knitr::include_graphics("tables_figures/HF_ring-porous_TMIN.png")
+```
 
@@
 **(a) $T_{max}$**
-```{r climwin_dp_tmin_HARV, out.width="100%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_HARV_short_caption}
-knitr::include_graphics("tables_figures/HF_diffuse-porous_DOY25.png")
+```{r climwin_dp_tmax_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
+knitr::include_graphics("tables_figures/HF_diffuse-porous_TMAX.png")
 ```
@@
 **(b) $T_{min}$**
+```{r climwin_dp_tmin_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_HARV_short_caption, fig.pos="H"}
+knitr::include_graphics("tables_figures/HF_diffuse-porous_TMIN.png")
+```
+
+\newpage
+
+```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption, fig.pos="H"}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
+```
+
+\newpage
+```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption, fig.pos="H"}
+#knitr::include_graphics("tables_figures/tree_cores_tmin.tif")
+```

### 2020-12-10 doc/manuscript/Dow_manuscript_Nature.Rmd
rearranged display items
moved several figures and tables to SI, auto numbering

#41

@@
 
-#Figures & Tables: listing figures and their number here is helpful for planning, and also allows automatic numbering within text -->easy changing of order.
-
-#Table numbers
-hypothesis_table=1
-table_species=2  #goes to SI
-mean_growth_parameters_table=3  #probably goes to SI
+#Display items: listing figures and their number here is helpful for planning, and also allows automatic numbering within text -->easy changing of order.
 
 #Figure numbers
-schematic=1 #illustrates parameters
-growth_curves_all = 2  #goes to SI
-    rel_growth_SCBI = "2a" 
-    rel_growth_HF = "2b" 
-    cum_growth_SCBI = "2c" 
-    cum_growth_HF = "2d" 
-climwin = 3 # probably stays?
-pheno_Tsensitivity = 4 
-DOYtiming = 5
-DOYtiming_SCBI = "5a"
-DOYtiming_Harvard = "5b"
-tree_rings = 6  #stays, hopefully adding a lot more sites
+schematic_summary=1 #illustrates parameters
+climwin = 2 # probably stays?
+pheno_Tsensitivity = 3 
+tree_rings = 4  #stays, hopefully adding a lot more sites
 
@@
 
-schematic_caption <- "Figure 1.
-Schematic illustrating the parameters considered here.
-Shown are measurements for an example tree (## cm SPECIES), fit with the model of McMahon & Parker (2015), from which phenology and growth rate parameters are obtained."
-
-growth_curves_all_caption  <- "Figure 2.
-Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
-Each line represents one year's growth for a given tree, fit with McMahon model.
-(THIS FIGURE GOES TO SI.)"
+schematic_summary_caption <- "Figure 1. Summary of results."
 
@@
 
-DOYtiming_caption <- "Figure 5.
-Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest.
-Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
-Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."
-
 tree_rings_caption <- "Figure 6.
@@
 Shown are Pearson correlations between ring-width index and and monthly Tmax.
-Species codes are as given in Table 2.(THIS FIGURE STAYS, PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
+Species codes are as given in Table 2.(THIS FIGURE WILL PROBABLY BE PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
 
 # SUPPLEMENTARY INFO (this chunk must be repeated in Dow_SI.Rmd)
-#SI table numbers
-sample_size_by_year= "S1"
+##SI table numbers
+table_species="S1"  
+sample_size_by_year= "S2"
+mean_growth_parameters_table="S3"
 
 #SI table captions
-sample_size_by_year_caption <- "Table S1. Sample size by year (THIS TABLE GOES TO SI)"
+sample_size_by_year_caption <- "Table S1. Sample size by year "
 
 #SI figure numbers
-all_climwin_SI_figs = " S1-S4"
-climwin_rp_DOY25_SCBI = "S1"
-climwin_dp_DOY25_SCBI = "S2"
-climwin_rp_DOY25_HARV = "S3"
-climwin_dp_DOY25_HARV = "S4"
-tree_rings_TMIN = "S5"
+growth_curves_all = "S1"  
+all_climwin_SI_figs = " S2-S5"
+  climwin_rp_DOY25_SCBI = "S2"
+  climwin_dp_DOY25_SCBI = "S3"
+  climwin_rp_DOY25_HARV = "S4"
+  climwin_dp_DOY25_HARV = "S5"
+DOYtiming = "S6"
+  DOYtiming_SCBI = "S6a"
+  DOYtiming_Harvard = "S6b"
+tree_rings_TMIN = "S7"
 
 #SI figure captions
-climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
-climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for $DOY_{25}$ for diffuse porous species at SCBI."
-climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for $DOY_{25}$ for ring porous species at Harvard Forest."
-climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
-tree_rings_TMIN_short_caption <- "Figure S5. ..."
+growth_curves_all_caption  <- "Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
+Each line represents one year's growth for a given tree, fit with McMahon model.
+"
 
+climwin_rp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
+climwin_dp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at SCBI."
+climwin_rp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at Harvard Forest."
+climwin_dp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
+DOYtiming_caption <- "Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest.
+Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
+Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."
+tree_rings_TMIN_short_caption <- "Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
+Shown are Pearson correlations between ring-width index and and monthly Tmin.
+Species codes are as given in Table 2.(THIS FIGURE WILL PROBABLY BE PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
+
+###~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 ## some stats/ 
@@
 Using data from dendrometer bands measured throughout the growing season, we fit a growth model to the time series of individual trees to determine the day of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$); maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$); total annual growth ($\Delta DBH$); and peak growing season length ($DOY_{75}$-$DOY_{25}$; $L_{pgs}$) [Fig.
-`r schematic`; @mcmahon_general_2015].
-We test the hypotheses that (1) warmer early springs result in earlier stem growth and a longer period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).
+`r schematic_summary`; @mcmahon_general_2015].
+We test the hypotheses that (1) warmer early springs result in earlier stem growth and a longer period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Fig.
+`r schematic_summary`).
 
-```{r schematic, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_caption}
+```{r schematic_summary, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_summary_caption}
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/schematic.png")
-```
-
-```{r hypothesis_table, eval=TRUE, echo=FALSE, warning=FALSE}
-knitr::include_graphics("tables_figures/hypothesis_table.png")
+knitr::include_graphics("tables_figures/schematic_summary.png")
 ```
@@
 
-```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/growth_curves_all.png")
-```
-
-```{r mean_growth_parameters_table, out.width="100%", echo=FALSE, fig.align="center"}
-knitr::include_graphics("tables_figures/mean_growth_parameters_table.png")
-```
-
 ## (Critical temperature windows)
@@
 
-For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Tables `r hypothesis_table`, Figs.
-`r pheno_Tsensitivity`, `r DOYtiming`).
-Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical temperature window (Table `r hypothesis_table`, `r pheno_Tsensitivity`, `r DOYtiming`).
+For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Fig. `r schematic_summary`, Figs. `r pheno_Tsensitivity`).
+Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical temperature window (Figs.
+`r schematic_summary`, `r pheno_Tsensitivity`).
 
@@
 
-```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
-```
-
-For three of the four species group-site combinations (exception: ring-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Tables `r hypothesis_table`, Figs.
-`r pheno_Tsensitivity`, `r DOYtiming`).
+For three of the four species group-site combinations (exception: ring-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Figs.
+`r schematic_summary`, `r pheno_Tsensitivity`).
 At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier as a result of warmer springs whereas ${DOY_75}$ did not shift significantly, resulting in a positive correlation between $L_{PGS}$ and $T_{max}$ during the critical temperature.
@@
 
-Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`). 
+Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Figs. `r schematic_summary`, `r pheno_Tsensitivity`). 
 Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer $T_{max}$ during the critical temperature window.
@@
 
-Total annual growth was not consistently linked to spring temperatures (Table `r hypothesis_table`).
+Total annual growth was not consistently linked to spring temperatures (Fig. `r schematic_summary`).
 Annual growth measured with dendrometer bands, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest) (Fig.
@@
 
-Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs.
-`r  climwin`-`r DOYtiming`).
+Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Figs.
+`r schematic_summary`-`r pheno_Tsensitivity`).
 This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
@@
 
-In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Table `r hypothesis_table`, Fig.
-`r pheno_Tsensitivity`), such that a much longer record would be needed to clarify what, if any, effect spring temperatures have on maximum daily and annual growth rates.
+In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs.
+`r schematic_summary`, `r pheno_Tsensitivity`), such that a much longer record would be needed to clarify what, if any, effect spring temperatures have on maximum daily and annual growth rates.
 Fortunately, we were able to use tree-ring data to understand centennial-scale temperature sensitivity of annual growth (Fig. `r tree_rings`).
@@
 Warm spring temperatures result in earlier leaf-out (fig., REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (Figs.
-`r climwin`- `r DOYtiming`).
-Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs. `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
+`r climwin`- `r pheno_Tsensitivity`).
+Yet, an earlier start to the growing season does not imply a longer growing season (Figs. `r schematic_summary`, `r pheno_Tsensitivity`) [@zani_increased_2020].
 Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models [Table `r hypothesis_table`; Figs.
-`r pheno_Tsensitivity`, `r tree_rings`; @zani_increased_2020].
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models (Figs.
+`r schematic_summary`, `r pheno_Tsensitivity`, `r tree_rings`)[@zani_increased_2020].
 The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.
@@
 
-```{r species, eval=TRUE, echo=FALSE, warning=FALSE}
-knitr::include_graphics("tables_figures/species_table.png")
-```
-
 Measurements were timed to begin before the beginning of spring growth and to continue through the cessation of growth in the fall. 
@@
 
-We fit the model of @mcmahon_general_2015 to define phenological dates and growth rates (Fig. `r schematic`). 
+We fit the model of @mcmahon_general_2015 to define phenological dates and growth rates (Fig. `r schematic_summary`). 
 This five-parameter logistic model was fitted to dendrometer band data from each tree-year in R: 
@@
 and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote. 
-*We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic`) occurs on $DOY_{ip}$ only when $\theta = 1$.*
-Using the optimized parameters for each tree-year, we then modeled $DBH$ throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic`).
+*We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`) occurs on $DOY_{ip}$ only when $\theta = 1$.*
+Using the optimized parameters for each tree-year, we then modeled $DBH$ throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
 
@@
 *The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
-Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic`), because we were interested in the influence of temperature on the initiation of woody growth.
+Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic_summary`), because we were interested in the influence of temperature on the initiation of woody growth.
 We note that other growth parameters would be controlled by different critical temperature windows; for example, at SCBI $\Delta DBH$ responds most strongly to May-July climate [@helcoski_growing_2019].

### 2020-12-10 doc/manuscript/tables_figures/tables.Rmd
rearranged display items
moved several figures and tables to SI, auto numbering

#41

--- 2020-12-04 doc/manuscript/tables_figures/tables.Rmd
@@
 ```
-
-\newpage
-```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-species_table <- read.csv(file = "species_table.csv", col.names = c("site", "xylem porosity","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
-kable(species_table, caption = "Table 2.
-Dominant ring- and diffuse-porous species at SCBI and Harvard Forest analyzed here, along with sample sizes.",  format = "latex", booktabs = TRUE,  escape = F) %>%
-add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
-kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-\newpage
-```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-mean_growth_parameters <- read.csv("mean_growth_parameters.csv",  col.names = c("", "ring","diffuse","ring", "diffuse"), stringsAsFactors = FALSE, check.names = FALSE)
-kable(mean_growth_parameters, caption = "Table 3.
-Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest.
-",  format = "latex", booktabs = TRUE,  escape = F) %>%
-add_header_above(c(" ",  "SCBI" = 2, "Harvard Forest"=2 )) %>%
-    column_spec(1, width = "4cm") %>%
-kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```

### 2020-12-10 doc/manuscript/Dow_manuscript_Nature.Rmd
minor
include DOYtiming figure instead of climwin
#41

put DOY timing figure back in main manuscript

removed climwin altogether, as its repetitive with existing SI figures

Co-Authored-By: Cameron Dow <55721987+camerondow35@users.noreply.github.com>

@@
 schematic_summary=1 #illustrates parameters
-climwin = 2 # probably stays?
-pheno_Tsensitivity = 3 
+pheno_Tsensitivity = 2 
+DOYtiming = 3
+  DOYtiming_SCBI = "3a"
+  DOYtiming_Harvard = "3b"
 tree_rings = 4  #stays, hopefully adding a lot more sites
@@
 
-climwin_caption <- "Figure 3.
-Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
-Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
-Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
-19- May 21 at SCBI, March 19 - May 07 at HF)."
-
 pheno_Tsensitivity_caption <- "Figure 4. Relationship between growth parameters and climwin mean temperatures. For each 
@@
 is represented by the solid blue line and 95% credible intervals are represented by bands."
+
+DOYtiming_caption <- "Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest.
+Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
+Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+(WE WILL ADD LEAF PHENOLOGY HERE)"
 
@@
   climwin_dp_DOY25_HARV = "S5"
-DOYtiming = "S6"
-  DOYtiming_SCBI = "S6a"
-  DOYtiming_Harvard = "S6b"
+
 tree_rings_TMIN = "S7"
@@
 "
-
 climwin_rp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
@@
 climwin_dp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
-DOYtiming_caption <- "Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest.
-Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
-Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."
+
 tree_rings_TMIN_short_caption <- "Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
@@
 1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
-`r climwin`, `r all_climwin_SI_figs`).
+`r all_climwin_SI_figs`).
 This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
@@
 
-```{r climwin, out.width="80%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/climwin_figure.png")
-```
-
-The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (Figs. `r climwin`, `r all_climwin_SI_figs`). 
+The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (Figs. `r all_climwin_SI_figs`). 
 At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (Figs.
@@
 Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical temperature window (Figs.
-`r schematic_summary`, `r pheno_Tsensitivity`).
+`r schematic_summary`, `r pheno_Tsensitivity`, `r DOYtiming`).
 
@@
 knitr::include_graphics("tables_figures/pheno_Tsensitivity_combo.png")
+```
+
+```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
+# Created in RScripts/additional_figures.R
+knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
 ```
@@
 Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Figs.
-`r schematic_summary`-`r pheno_Tsensitivity`).
+`r schematic_summary`-`r `r DOYtiming``).
 This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
@@
 Warm spring temperatures result in earlier leaf-out (fig., REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (Figs.
-`r climwin`- `r pheno_Tsensitivity`).
+`r schematic_summary`- `r DOYtiming`).
 Yet, an earlier start to the growing season does not imply a longer growing season (Figs. `r schematic_summary`, `r pheno_Tsensitivity`) [@zani_increased_2020].
@@
 # (Content to incorporate)
-
-new Science paper: @zani_increased_2020
 

### 2020-12-10 doc/manuscript/Dow_SI.Rmd

@@
 # SUPPLEMENTARY INFO (copied from Dow_manuscript.Rmd)
-#SI table numbers
+##SI table numbers
 table_species="S1"  
@@
 \newpage
-
-```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption, fig.pos="H"}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
-```
-
-\newpage
 ```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption, fig.pos="H"}

### 2020-12-10 doc/manuscript/tables_figures/tables.Rmd

@@
 library(kableExtra)
+summary_table <- read.csv(file = "summary_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+#add footnote markers:
+summary_table[0,3] <- paste0(summary_table[0,3], footnote_marker_number(1)) 
+summary_table[0,3] <- paste0(summary_table[0,3], footnote_marker_number(2)) 
+#summary_table[8,3] <- paste0(summary_table[8,3], footnote_marker_number(3)) 
+#summary_table[8,4] <- paste0(summary_table[8,4], footnote_marker_number(4)) 
+#summary_table[8,6] <- paste0(summary_table[8,6], footnote_marker_number(5))
+
+kable(summary_table, col.names = c("Variable", "Definition", "RP", "DP","RP", "DP"), format = "latex", booktabs = TRUE,  escape = F)%>%
+  add_header_above(c(" ", " " , "SCBI" = 2, "Harvard Forest"=2)) %>%
+  add_header_above(c(" ", " " , "Response to warmer spring T"=4)) %>%
+  column_spec(3:6, width = "1cm") %>%
+  pack_rows("Timing of growth", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Growth rate", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Annual growth", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) %>%
+
+#add footnote: 
+kableExtra::footnote(number = c("'RP' and 'DP' refer to ring- and diffuse- porous species, respectively", "'-' and '+' = significant negative or positive reponse, respectively; 'n.s.'= no significant correlation"), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
+```
+
+\newpage
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
 hypothesis_table <- read.csv(file = "hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

### 2020-12-11 doc/manuscript/tables_figures/tables.Rmd
update Table 1, fix up schematic summary

--- 2020-12-10 doc/manuscript/tables_figures/tables.Rmd
@@
 library(kableExtra)
+summary_table <- read.csv(file = "summary_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+#add footnote markers:
+#summary_table[0,3] <- paste0(summary_table[0,3], footnote_marker_number(1)) 
+#summary_table[0,3] <- paste0(summary_table[0,3], footnote_marker_number(2)) 
+#summary_table[8,3] <- paste0(summary_table[8,3], footnote_marker_number(3)) 
+#summary_table[8,4] <- paste0(summary_table[8,4], footnote_marker_number(4)) 
+#summary_table[8,6] <- paste0(summary_table[8,6], footnote_marker_number(5))
+
+kable(summary_table, col.names = c("Variable", "Definition", "RP", "DP","RP", "DP"), format = "latex", booktabs = TRUE,  escape = F)%>%
+  add_header_above(c(" ", " " , "SCBI" = 2, "Harvard Forest"=2)) %>%
+  add_header_above(c(" ", " " , "Response to warmer spring T"=4), bold= TRUE) %>%
+  column_spec(3:6, width = "1cm") %>%
+  pack_rows("Timing of growth", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Daily growth rate", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Annual growth", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T)
+
+#add footnote: 
+#kableExtra::footnote(number = c("'RP' and 'DP' refer to ring- and diffuse- porous species, respectively", "'-' and '+' = significant negative or positive response, respectively; 'n.s.'= no significant correlation"), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
+```
+
+\newpage
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
 hypothesis_table <- read.csv(file = "hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

### 2020-12-11 doc/manuscript/Dow_manuscript_Nature.Rmd
update figures

@@
 
-schematic_summary_caption <- "Figure 1. Summary of results."
+schematic_summary_caption <- "Figure 1 | Summary of tree growth changes under warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous).
+(a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over DOY25; (b) Variable definitions and summary responses to warmer temperatures, where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
 
-pheno_Tsensitivity_caption <- "Figure 4. Relationship between growth parameters and climwin mean temperatures. For each 
+pheno_Tsensitivity_caption <- "Figure 2 | Relationship between growth parameters and climwin mean temperatures. For each 
 observed climwin mean temperature value, the posterior mean of the fitted day of year
@@
 
-DOYtiming_caption <- "Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest.
+DOYtiming_caption <- "Figure 3 | Day of year where growth milestones were achieved at (a) the Smithsonian Conservation Biology Institute and (b) Harvard Forest.
 Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
@@
 
-tree_rings_caption <- "Figure 6.
-Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
+tree_rings_caption <- "Figure 4 | Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
 Shown are Pearson correlations between ring-width index and and monthly Tmax.
@@
 # Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/pheno_Tsensitivity_combo.png")
+knitr::include_graphics("tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png")
 ```

### 2020-12-11 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
condensed phenological results

@@
 (*most or all of this should be cut-- first paragraph too detailed for Nature, and we want minimal comparison of ring vs diffuse porous*)
-
-Across all trees and years at SCBI, 
-25% of annual growth was achieved on May 20 ($DOY_{25}$=139), 
-50% on June 12 ($DOY_{50}$=162), 
-and 75% on July 5 ($DOY_{75}$=185), 
-for an average peak growing season length of 45 days. 
-Maximum growth rate occurred just one day after $DOY_{50}$ on June 13 ($DOY_{max}$=163). 
-At Harvard Forest,
-25% of annual growth was achieved on June 1 ($DOY_{25}$=151), 
-50% on June 23 ($DOY_{50}$=173), 
-and 75% on July 14 ($DOY_{75}$=194), 
-for an average peak growing season length of 43 days. 
-Maximum growth rate occurred on June 24 ($DOY_{max}$=174).
-
-At both sites, growth phenology differed between ring- and diffuse- porous species (Fig. `r growth_curves_all`, Table `r mean_growth_parameters_table`). 
-Specifically, ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark `r 154-123` days earlier at SCBI and `r 164-132` days earlier at Harvard Forest. 
-Growth of ring-porous species was spread out over a longer growing season, with average $L_{PGS}$ of ring-porous species being 21 days longer at SCBI and 19 days longer at Harvard Forest. 
+Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: #-# days) and differed between ring- and diffuse-porous species (Fig.
+`r growth_curves_all`, Table `r mean_growth_parameters_table`).
+Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest, while $DOY_{75}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest.
+Ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively).
 Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.

### 2020-12-11 doc/manuscript/Dow_manuscript_Nature.Rmd
improved abstract

@@
 
-(*word limit 200; this is currently 266*)
+(*word limit 200. italic=tentative*)
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out in temperate deciduous forests, along with earlier initiation of the annual period where these ecosystems are net sinks of carbon dioxide (CO~2~).
-As a result, there is a tendency towards increased growing season length and annual CO~2~ sequestration.
-However, little is known about how spring temperatures affect tree stem growth, which consumes a modest portion of the carbon fixed annually but is critical in that wood has a long residence time in the ecosystem.
-We need to know how the timing and rate of woody growth are affected by warming spring temperatures.
-Here, we show that in two temperate deciduous forests warmer maximum temperatures in the spring shift growth milestones earlier in the year, but have only a modest effect on peak growing season length and little effect on annual growth.
-This contrasts with the expectation *in most models* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-It is, however, consistent with most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[@helcoski_growing_2019], with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
-Thus, our findings indicate that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[REFS] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[REFS] and annual CO~2~ sequestration[REFS].
+However, little is known about how spring temperatures affect tree stem growth, which consumes a modest portion of the carbon fixed annually[REFS] but is the longest-lived carbon sink[REFS].
+Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted *both leaf phenology* and woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
+*The latter finding is confirmed on the centennial scale by tree-ring records across # forests.*
+These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019,@zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests, implying that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.
 
@@
 
-(*most or all of this should be cut-- first paragraph too detailed for Nature, and we want minimal comparison of ring vs diffuse porous*)
-Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: #-# days) and differed between ring- and diffuse-porous species (Fig.
-`r growth_curves_all`, Table `r mean_growth_parameters_table`).
-Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest, while $DOY_{75}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest.
+*(we will need to add leaf phenology to this paragraph)*
+
+**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: #-# days) and differed between ring- and diffuse-porous species (Fig.
+`r growth_curves_all`, Table `r mean_growth_parameters_table`).**
+Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 194 (July 14) at Harvard Forest.
 Ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively).

### 2020-12-11 doc/manuscript/Dow_manuscript_Nature.Rmd
update hypotheses & schematic_summary

@@
 *The latter finding is confirmed on the centennial scale by tree-ring records across # forests.*
-These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019,@zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019,@zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests, implying that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.
@@
 ## (Here, we...)
-Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA.
-Using data from dendrometer bands measured throughout the growing season, we fit a growth model to the time series of individual trees to determine the day of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$); maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$); total annual growth ($\Delta DBH$); and peak growing season length ($DOY_{75}$-$DOY_{25}$; $L_{pgs}$) [Fig.
-`r schematic_summary`; @mcmahon_general_2015].
-We test the hypotheses that (1) warmer early springs result in earlier stem growth and a longer period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Fig.
+Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings based on their differing growth phenology[*D'Orangeville*].
+First, using data from dendrometer bands measured throughout the growing season in two forests in the Eastern USA (n=`r n_tree_years` tree-years), we fit a growth model[@mcmahon_general_2015]to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
+*(We also look at leaf phenology.)*
+Second, at these focal sites and *#* additional sites distributed *xxxx*, we analyze *#* tree-ring chronologies to understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale.
 
@@
 In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows, both > 5 weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
-However, for both ring- and diffuse- porous species, average $T_{min}$ of the selected period was below freezing (<0° C) in all years, suggesting that these are sporadic correlations that are unlikely to be biologically meaningful.
+However, for both ring- and diffuse- porous species, average $T_{min}$ of the selected period was below freezing (<0°C) in all years, suggesting that these are sporadic correlations that are unlikely to be biologically meaningful.
 
-Henceforth, we focus on $T_{max}$.
+Henceforth, we focus on $T_{max}$ assessed over *climwin*-selected critical temperature windows, *noting that similar results are obtained using alternate temperature window ([GitHub issue #42](https://github.com/EcoClimLab/growth_phenology/issues/42)).*
 

### 2020-12-11 doc/manuscript/Dow_manuscript_Nature.Rmd
working on intro

@@
 *The latter finding is confirmed on the centennial scale by tree-ring records across # forests.*
-These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019,@zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests, implying that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.
@@
 
-## (forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
+## (ftemperate deciduous forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
+
+(*this part needs to be re-written (very concisely, and focused on the temperate forest C sink and its uncertainty*)
+
 As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
@@
 Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services.
-For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil.
-Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon [@bonan_forests_2008].
-However, their future under the current era of rapid change is uncertain [@friedlingstein_climatecarbon_2006].
-
-Reducing this uncertainty requires better understanding of the drivers of tree growth [@bugmann_tree_2019].
-Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season [@helcoski_growing_2019]; [@Ciais_europe-wide_2005]; [@Goulden_exchange_1996], as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain [@mathias_disentangling_2018], and invasion by pests or pathogens.
-Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.
+For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO~2~ from the atmosphere and storing it in the trees or soil.
+Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon[@bonan_forests_2008].
+However, their future under the current era of rapid change is uncertain[@friedlingstein_climatecarbon_2006].
 
 ## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
-In recent decades, an increase in average spring temperature has been impacting many natural systems including forests [@parmesan_globally_2003]; [@zhizong_fingerprinting_2006]. 
-The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*.
-For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010];@jeong_phenology_2011];@crabbe_extreme_2016].
-The emergence of leaves signals the beginning of carbon sequestration or in other words, the growing season of a tree.
-It is believed that the longer the tree has leaves - or as the growing season is extended - the net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$; @churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].)
+In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003] and (*often later autumn*[REFS].
+In temperate deciduous forests, spring leaf-out has been occurring earlier[REFS], as has the beginning of the annual period of net CO~2~ sequestration[REFS].
+The timing of autumn senescence is less straightforward (*need to do more lit review on this*).
+In some regions earlier leaf out has been matched or exceeded by delays in leaf senescence *when springtime temperatures are above average(??)*[@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016].
+(*is this just spring temperatures?
+Or autumn?
+how widespread is this?*)
+Experimental evidence demonstrates that warmer spring temperatures lead to earlier leaf-out and earlier leaf senescence[@zohner_ongoing_2019], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020], whereas warmer summer and autumn temperatures delay leaf senescence[@zohner_ongoing_2019].
+Whatever the mechanisms controlling autumn senescence, the observed pattern is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[REFS] and the ecosystem is a net CO~2~ sink[REFS].
 
-However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019].
-This implies that earlier leafout will lead to smaller increases in NEE than changes in other phenological traits like the maximum C uptake rate.
-This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE.
-A paper by @zani_increased_2020 found that earlier leaf out led to earlier leaf senescence within the deciduous trees in their study, reducing the ability of the trees to sequester significantly higher levels of C.
-Instead they posit that senescence is induced when seasonal carbon-sink capacity of the tree is reached, no matter how extended the potential growing season is.
-This result directly contradicts those previous predictions that warmer Spring temperatures and earlier leaf out will lead to drastically increased C sequestration.
+The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].
+*This expectation is represented in models [@zani_increased_2020].*
+However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out will not lead to drastically increased C sequestration[@zani_increased_2020].
 
@@
 Earlier timing of spring leaf-up and initiation of carbon sequestration signals that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research on the characteristics of stem-growth during this shift has been done.
-In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns.
-*(but check into this....
-I suspect there may be some that we're missing--perhaps auto dendro bands in Europe.
-Neil did say that there's little on early spring T)*.
+In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns (but see Ref.
+[D'Orangeville], *and we need to make sure we're not missing anything*).
 
-Tree-ring records give some hints that warmer springs may increase annual growth ([@Helcoski_growing_2019]; probably some more), but tree-rings tell us nothing about the timing of that growth.
+Tree-ring records give some hints that warmer springs may increase annual growth (e.g., Refs.[@helcoski_growing_2019] [MORE]), but tree-rings tell us nothing about the timing of that growth.
 On the other hand, earlier leaf development shows us how the growing season is advancing, but can't tell us about the total carbon sequestered, while neither can be used to map out intra-annual growth patterns.
 
-Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on the future of forests.A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals.
+Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on the future of forests.
+A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals.
 A mixture of these three would produce the most complete picture of spring climate influence on forests.
 
-## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
-*(Too much detail here? how about just keeping some variation of the first and last sentences)*
-Leaf and wood phenology traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al.
-**).
-Interspecific variation of this sort likely stems from evolutionary differences and adaptations to local climates where these trees can be found.
-Ring-porous trees have large xylem cells which are prone to embolism during the winter.
-These cells are repaired before leaf development in the Spring, corresponding with an earlier initiaiton of stem-growth.
-Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process.
-Results from the paper by D’Orangeville et al.
-(in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time.
-If leaf development is sped up when spring temperatures are warmer, we should seek to understand how this shift influences stem-growth timing and other phenological variables.
+##(temporarily removed intro content)
+
+(*Cameron, this is content --with associated refs--that I don't want to lose, but that needs to be worked into reworded text above*)
+The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*.
+For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016].
 
 ## (Here, we...)
-Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings based on their differing growth phenology[*D'Orangeville*].
-First, using data from dendrometer bands measured throughout the growing season in two forests in the Eastern USA (n=`r n_tree_years` tree-years), we fit a growth model[@mcmahon_general_2015]to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings, which differ in growth phenology[*D'Orangeville*].
+First, using data from dendrometer bands measured throughout the growing season in two forests in the Eastern USA (n=`r n_tree_years` tree-years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 194 (July 14) at Harvard Forest.
-Ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively).
+Consistent with the results of Ref[D'Orangeville], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively).
 Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.
@@
 # (Content to incorporate)
+
+However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019].
+This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE.
 

### 2020-12-11 doc/manuscript/Dow_manuscript_Nature.Rmd
worked on intro

@@
 
-(*this will need to be shortened signficantly*)
-
-## (ftemperate deciduous forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
-
-(*this part needs to be re-written (very concisely, and focused on the temperate forest C sink and its uncertainty*)
+(*this first paragraph needs to be re-written (very concisely, and focused on the temperate forest C sink and its uncertainty*)
 
@@
 
-## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
-In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003] and (*often later autumn*[REFS].
+**In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003] and *often later autumn*[REFS].**
 In temperate deciduous forests, spring leaf-out has been occurring earlier[REFS], as has the beginning of the annual period of net CO~2~ sequestration[REFS].
-The timing of autumn senescence is less straightforward (*need to do more lit review on this*).
+Controls on the timing of autumn senescence are less clearly understood (*need to do more lit review on this*).
 In some regions earlier leaf out has been matched or exceeded by delays in leaf senescence *when springtime temperatures are above average(??)*[@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016].
@@
 
-The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].
-*This expectation is represented in models [@zani_increased_2020].*
-However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out will not lead to drastically increased C sequestration[@zani_increased_2020].
+~~~~~~~~~~
+##(content --with associated refs--that we don't want to lose, but that needs to be worked into reworded text above)
 
-## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
-Earlier timing of spring leaf-up and initiation of carbon sequestration signals that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research on the characteristics of stem-growth during this shift has been done.
-In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns (but see Ref.
-[D'Orangeville], *and we need to make sure we're not missing anything*).
-
-Tree-ring records give some hints that warmer springs may increase annual growth (e.g., Refs.[@helcoski_growing_2019] [MORE]), but tree-rings tell us nothing about the timing of that growth.
-On the other hand, earlier leaf development shows us how the growing season is advancing, but can't tell us about the total carbon sequestered, while neither can be used to map out intra-annual growth patterns.
-
-Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on the future of forests.
-A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals.
-A mixture of these three would produce the most complete picture of spring climate influence on forests.
-
-##(temporarily removed intro content)
-
-(*Cameron, this is content --with associated refs--that I don't want to lose, but that needs to be worked into reworded text above*)
 The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*.
 For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016].
+~~~~~~~~~~~~~~~~
 
-## (Here, we...)
-Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings, which differ in growth phenology[*D'Orangeville*].
+**The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].**
+*This expectation is represented in models [@zani_increased_2020].*
+However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].
+
+**We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures.** 
+In fact, we know little about the stem-growth phenology of temperate deciduous species [but see Ref.
+D'Orangeville].
+(*Here, we need to review the literature to make sure we're not missing anything.
+We also want to set up the fact that our dataset is unique/justify just 2 sites*).
+Tree-ring records give some hints that warmer springs may increase annual growth (e.g., Refs.[@helcoski_growing_2019] [MORE]), but rely purely on correlations of annual growth to monthly temperatures and can tell us nothing about the timing of that growth.
+
+*Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on how forest biomass growth is likely to respond to warming spring temperatures.
+A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals.
+A mixture of these three would produce the most complete picture of spring climate influence on forests.*
+
+**Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings, which differ in growth phenology[*D'Orangeville*].**
 First, using data from dendrometer bands measured throughout the growing season in two forests in the Eastern USA (n=`r n_tree_years` tree-years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
@@
 
-## (Phenological patterns)
+## (overview of phenological patterns)
 

### 2021-01-16 RScripts/Leaf_phenology.R
first pass at leaf pheno

@@ -1,0 +1,66 @@
+library(readr)
+library(lubridate)
+library(tidyverse)
+
+pheno <- read_csv("Data/Leaf phenology/scbiPhen.csv")
+pheno_hf <- subset(pheno, pointid == "HARV")
+pheno_scbi <- subset(pheno, pointid == "SCBI")
+pheno_events <- data.frame(1,1,1,1,1,1)
+names(pheno_events) <- c("site","year","greenup","midgreenup", "peak","senescence")
+
+for(i in c(1:nrow(pheno_scbi))){
+  pheno_events[i,1] <- as.character("SCBI")
+  pheno_events[i,2] <- 2000+i
+  pheno_events[i,3] <- as.numeric(pheno_scbi[i,6])
+  pheno_events[i,4] <- as.numeric(pheno_scbi[i,8])
+  pheno_events[i,5] <- as.numeric(pheno_scbi[i,10])
+  pheno_events[i,6] <- as.numeric(pheno_scbi[i,14])
+
+}
+
+for(i in c(1:nrow(pheno_hf))){
+  pheno_events[i+18,1] <- as.character("HF")
+  pheno_events[i+18,2] <- 2000+i
+  pheno_events[i+18,3] <- as.numeric(pheno_hf[i,6])
+  pheno_events[i+18,4] <- as.numeric(pheno_hf[i,8])
+  pheno_events[i+18,5] <- as.numeric(pheno_hf[i,10])
+  pheno_events[i+18,6] <- as.numeric(pheno_hf[i,14])
+
+}
+
+pheno_events$los <- pheno_events$senescence-pheno_events$greenup
+
+#Add climate variable
+scbi <- read_csv("climate data/scbi_2001to2018.csv")
+scbi <- scbi[,c(3,4)]
+scbi <- scbi[complete.cases(scbi$TMAX),]
+scbi$month <- month(scbi$DATE)
+scbi$year <- year(scbi$DATE)
+scbi$doy <- yday(scbi$DATE)
+scbi_sub <- subset(scbi, doy >= 78 & doy <= 140)
+scbi_agg <- aggregate(scbi_sub$TMAX, by = list(scbi_sub$year), FUN = mean)
+scbi_agg$site <- "SCBI"
+names(scbi_agg) <- c("year","tmp", "site")
+
+HF <- read_csv("climate data/HF_2001toPresent.csv")
+HF <- HF[,c(1,5)]
+
+HF$month <- month(HF$date)
[…]

### 2021-01-16 doc/manuscript/Dow_manuscript_Nature.Rmd
small changes

@@
 Controls on the timing of autumn senescence are less clearly understood (*need to do more lit review on this*).
-In some regions earlier leaf out has been matched or exceeded by delays in leaf senescence *when springtime temperatures are above average(??)*[@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016].
-(*is this just spring temperatures?
-Or autumn?
-how widespread is this?*)
-Experimental evidence demonstrates that warmer spring temperatures lead to earlier leaf-out and earlier leaf senescence[@zohner_ongoing_2019], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020], whereas warmer summer and autumn temperatures delay leaf senescence[@zohner_ongoing_2019].
+Experimental evidence demonstrates that warmer spring temperatures lead to earlier leaf-out and earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020], whereas warmer summer and autumn temperatures delay leaf senescence[@zohner_ongoing_2019;@crabbe_extreme_2016;@dragoni_evidence_2011;@jeong_phenology_2011;@ibanez_forecasting_2010;@vitasse_responses_2009].
 Whatever the mechanisms controlling autumn senescence, the observed pattern is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[REFS] and the ecosystem is a net CO~2~ sink[REFS].
@@
 *Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on how forest biomass growth is likely to respond to warming spring temperatures.
-A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals.
-A mixture of these three would produce the most complete picture of spring climate influence on forests.*
-
+To produce the most complete picture of spring influence on forests, a combination of established data types - tree cores, for inter-annual stem growth, and remote sensing, for leaf phenology - should be used in conjunction with a recently developed intra-annual modeling technique [@mcmahon_general_2015].
+*
 **Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings, which differ in growth phenology[*D'Orangeville*].**
@@
 
-**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: #-# days) and differed between ring- and diffuse-porous species (Fig.
+**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (Fig.
 `r growth_curves_all`, Table `r mean_growth_parameters_table`).**
-Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 194 (July 14) at Harvard Forest.
+Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
 Consistent with the results of Ref[D'Orangeville], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively).
@@
 For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Fig. `r schematic_summary`, Figs. `r pheno_Tsensitivity`).
-Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{g.max}$ at SCBI and $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at Harvard Forest all decreased significantly with increasing $T_{max}$ during the critical temperature window (Figs.
+Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with increasing $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r pheno_Tsensitivity`, `r DOYtiming`).
@@
 
-For three of the four species group-site combinations (exception: ring-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Figs.
+For two of the four species group-site combinations (ring-porous at SCBI and diffuse-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Figs.
 `r schematic_summary`, `r pheno_Tsensitivity`).
-At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier as a result of warmer springs whereas ${DOY_75}$ did not shift significantly, resulting in a positive correlation between $L_{PGS}$ and $T_{max}$ during the critical temperature.
-For diffuse-porous species at both SCBI and Harvard Forest, all growth milestones occurred earlier in warmer springs, but $DOY_{25}$ was more strongly affected than ${DOY_75}$, resulting in increased $L_{PGS}$.
+*At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier as a result of warmer springs whereas ${DOY_75}$ did not shift significantly, resulting in a positive correlation between $L_{PGS}$ and $T_{max}$ during the critical temperature.*
+*For diffuse-porous species* At SCBI, all growth milestones occurred earlier in warmer springs, but $DOY_{25}$ was more strongly affected than ${DOY_75}$, resulting in increased $L_{PGS}$.
 At Harvard Forest,  all growth milestones occurred earlier in warmer springs, but ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length.

### 2021-01-25 doc/manuscript/Dow_manuscript_Nature.Rmd
add some refs

@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[REFS] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[REFS] and annual CO~2~ sequestration[REFS].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[REFS] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
 However, little is known about how spring temperatures affect tree stem growth, which consumes a modest portion of the carbon fixed annually[REFS] but is the longest-lived carbon sink[REFS].

### 2021-01-26 doc/manuscript/tables_figures/tables.Rmd
update schematic summary with new figure from @rudeboybert

@@
 ```
+(a)
+
+![](schematic_v2.png)
+
+(b)
 
@@
 ```
-
-\newpage
-
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-hypothesis_table <- read.csv(file = "hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-
-#add footnote markers:
-hypothesis_table[8,2] <- paste0(hypothesis_table[8,2], footnote_marker_number(1)) 
-hypothesis_table[8,3] <- paste0(hypothesis_table[8,3], footnote_marker_number(2)) 
-hypothesis_table[8,5] <- paste0(hypothesis_table[8,5], footnote_marker_number(3))
-
-kable(hypothesis_table, col.names = c("Hypotheses and Specific Predictions", "RP", "DP","RP", "DP", "Results"), caption = "Table 1.
-Table of hypotheses and associated specific predictions, whether each was supported ('yes'; signficant at p<0.05), rejected ('no'; opposite trend signficant at p<0.05), or found insigificant ('n.s.'; no significant correlation), and display items showing the results.
-'RP' and 'DP' refer to ring- and diffuse- porous species, respectively.", format = "latex", booktabs = TRUE,  escape = F)%>%
-  #column_spec(1, width = "4cm") %>%
-  add_header_above(c(" ", "SCBI" = 2, "Harvard Forest"=2, " " )) %>%
-  pack_rows("Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Maximum growth rates are independent of early spring temperatures.", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Annual stem growth responds positively to warmer spring temperatures.", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) %>%
-
-#add footnote: 
-kableExtra::footnote(number = c("One of nine species analyzed had significant positive response to April $T_{max}$; one had significant negative response to March $T_{max}$", "One of two species analyzed had signficant positive response to April $T_{max}$, both had negative response to May $T_{max}$", "One of the two species was negatively correlated with April $T_{max}$, and the other positively correlated with May $T_{max}$."), footnote_as_chunk = FALSE, threeparttable=TRUE, escape = FALSE)
-```

### 2021-01-26 doc/manuscript/Dow_manuscript_Nature.Rmd
a few updates

@@
 Valentine Herrmann^1^
+
+Justin Maxwell
 
@@
 
-schematic_summary_caption <- "Figure 1 | Summary of tree growth changes under warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous).
-(a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over DOY25; (b) Variable definitions and summary responses to warmer temperatures, where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
+schematic_summary_caption <- "Figure 1 | Summary of tree growth changes under warmer spring temperatures.
+(a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
 
@@
 
-Mixed effects models were used to test the response of growth phenology variables ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_max}}$, $g_{max}$, and $\Delta DBH$) to fixed effects of xylem porosity and mean temperature ($T_{max}$, $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
+Mixed effects models were used to test the response of growth phenology variables ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_max}$, $g_{max}$, and $\Delta DBH$) to fixed effects of xylem porosity and mean temperature ($T_{max}$, $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
 We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ and $T_{min}$. 

### 2021-02-01 doc/manuscript/Dow_manuscript.Rmd
add Babst citation

@@
 `r tree_rings`).
-Relative to other temperate deciduous forest sites, the pattern at SCBI tends to be more typical: annual growth is most commonly controlled by climate conditions during the peak growing season.
+Relative to other temperate deciduous forest sites, the pattern at SCBI tends to be more typical: annual growth is most commonly controlled most strongly by climate conditions during the peak growing season [@babst_twentieth_2019].
+There is, however, evidence that growth of some trees in humid temperate regions increases in response to warmer late winter/ early spring conditions [e.g., Anderson-Teixeira et al.
+in prep; @babst_twentieth_2019].
 (*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)

### 2021-03-02 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 
+*Grant Harley (ASSUMING WE USE SMOKIES TREE RING RECORDS)*
+
 Justin Maxwell
@@
 
-7. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+7. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
 
@@
 
+The tree-ring records from our focal sites were complemented with a much larger collection spanning **#** deciduous and mixed forest sites in Eastern North America.
+Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).
+
 All cores were mounted, sanded, cross-dated, and measured using standard methodology. 

### 2021-03-14 doc/manuscript/Dow_manuscript_Nature.Rmd
tweaks to results and discussion

@@
 `r growth_curves_all`, Table `r mean_growth_parameters_table`).**
-Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
+Across all trees and years, the mean $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
 Consistent with the results of Ref[D'Orangeville], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively).
-Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.
+Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species.
 
@@
 `r climwin_rp_DOY25_SCBI` - `r climwin_dp_DOY25_SCBI`).
-In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows, both > 5 weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
+In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows greater than five weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
 However, for both ring- and diffuse- porous species, average $T_{min}$ of the selected period was below freezing (<0°C) in all years, suggesting that these are sporadic correlations that are unlikely to be biologically meaningful.
@@
 For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Fig. `r schematic_summary`, Figs. `r pheno_Tsensitivity`).
-Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with increasing $T_{max}$ during the critical temperature window (Figs.
+Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g.max}$, leaf greenup, and mid-greenup, at both sites all decreased significantly with increasing $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r pheno_Tsensitivity`, `r DOYtiming`).
@@
 At Harvard Forest,  all growth milestones occurred earlier in warmer springs, but ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length.
+
+Ring- and diffuse- porous species were pooled together at each site for leaf phenology measures.
+Day of year of greenup occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the year with the warmest and coldest critical windows in the analysis (Fig.
+3).
+Mid-greenup timing also shifted forward at both sites when spring temperatures were higher but less so than greenup.
+Senescence occurred earlier at Harvard forest by 12 days in the warm year while SCBI saw a 5 day delay of senescence during the warm year.
 
@@
 We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity.
-Using dendroband data, we found that the peak growing season is generally advancing as the critical spring windows are warming but having little, or negative effect, on $L_{pgs}$.
+Using dendroband data, we found that the peak growing season is generally advancing as the critical spring windows are warming.
+Despite this shift, growth variables $L_{pgs}$.
 $g_{max}$ and $\Delta DBH$ were virtually unchanged when this period warmed, but $DOY_{g_{max}}$ shifted back almost in unison with $DOY_{50}$.
-The lack of influence Spring TMAX exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores.
-Annual tree growth depends more on peak growing season conditions than early season advancement.
+The lack of influence Spring TMAX exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores, which show that annual tree growth depends more on peak growing season conditions than early season advancement.
 
@@
 `r schematic_summary`-`r `r DOYtiming``).
-This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
-indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
+This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle. 
+Indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
 This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*
@@
 Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
-Warm spring temperatures result in earlier leaf-out (fig., REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (Figs.
+Warm spring temperatures result in earlier leaf-out (fig., REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
 `r schematic_summary`- `r DOYtiming`).
-Yet, an earlier start to the growing season does not imply a longer growing season (Figs. `r schematic_summary`, `r pheno_Tsensitivity`) [@zani_increased_2020].
-Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
+Yet, an earlier start to the growing season is not associated with a longer growing season (Figs. `r schematic_summary`, `r pheno_Tsensitivity`) [@zani_increased_2020]
+because growth finishes sooner under these conditions, presumably because of *C sink limitation* [@zani_increased_2020].
 
@@
 `r schematic_summary`, `r pheno_Tsensitivity`, `r tree_rings`)[@zani_increased_2020].
-The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.
+Because of this, temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that these models are underestimating future overall atmospheric CO~2~.
 

### 2021-03-19 doc/manuscript/Dow_manuscript_Nature.Rmd
a bit of work on text and refs

@@
 As the climate changes, warmer spring temperatures are causing earlier leaf-out[REFS] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
-However, little is known about how spring temperatures affect tree stem growth, which consumes a modest portion of the carbon fixed annually[REFS] but is the longest-lived carbon sink[REFS].
-Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted *both leaf phenology* and woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
-*The latter finding is confirmed on the centennial scale by tree-ring records across # forests.*
-These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
-They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests, implying that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.
+However, little is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood, which has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
+Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
+The latter finding is confirmed on the centennial scale by **#** tree-ring chronologies from **#** forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
+These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
+Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] must be allocated primarily to relatively short-lived C pools and therefore unlikely to have a lasting impact on the forest C balance.*
+Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
 
@@
 
-As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid.
-The absence of human intervention increases our reliance on natural systems to soften the blow of climate change.
-Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services.
-For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO~2~ from the atmosphere and storing it in the trees or soil.
-Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon[@bonan_forests_2008].
-However, their future under the current era of rapid change is uncertain[@friedlingstein_climatecarbon_2006].
+As global atmospheric CO~2~ concentrations increase and Earth's temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@bonan_forests_2008; @friedlingstein_climate_2006].
+Temperate deciduous forests, including those in eastern North America, *(have been net carbon sinks over the past several decades)*... 
+*[Discuss this sink, mechanisms, uncertainties, ending with the point that one meaningful uncertainty is how the C sink will respond to warming springs]*
 
@@
 Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*). 
-The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July ([@Helcoski_growing_2019]).
+The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July ([helcoski_growing_2019]).
 
@@
 For the dendrometer band analysis, we used climate data from local weather stations. 
-Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeoclimate_2020].
+Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020].
 Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
@@
 
-For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020], again via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeoclimate_2020].
+For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020], again via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
 
@@
 
-The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@van_de_pol_identifying_2016]. 
+The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@vandepol_identifying_2016]. 
 This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window.

### 2021-03-19 doc/manuscript/Dow_manuscript_Nature.Rmd
compiled draft DOY timing figure with stem growth + foliage
#50

@@
 
-DOYtiming_caption <- "Figure 3 | Day of year where growth milestones were achieved at (a) the Smithsonian Conservation Biology Institute and (b) Harvard Forest.
-Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site.
-Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-(WE WILL ADD LEAF PHENOLOGY HERE)"
+DOYtiming_caption <- "Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).
+**DRAFT FIGURE-- REAL DATA BUT NEEDS FORMATTING.** Panels (a-b) show leaf phenology...
+Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site.
+Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."
 
@@
 
-As global atmospheric CO~2~ concentrations increase and Earth's temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@bonan_forests_2008; @friedlingstein_climate_2006].
+As global atmospheric CO~2~ concentrations increase and Earth's temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
 Temperate deciduous forests, including those in eastern North America, *(have been net carbon sinks over the past several decades)*... 
@@
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
+knitr::include_graphics("tables_figures/DOYtiming_stemgrowth+foliage.png")
 ```

### 2021-04-16 doc/manuscript/Dow_SI.Rmd
update hypotheses & schematic_summary
add draft tree rings figure to manuscript
#49

@@
 mean_growth_parameters_table="S3" 
+chronologies_table="S4"
 
@@
 \newpage
+# (placeholder for Table S4-- chronologies table.)
+
+\newpage
 

### 2021-04-16 doc/manuscript/Dow_manuscript_Nature.Rmd
knit in latest fig
add draft tree rings figure to manuscript
#49

@@
 
-*Grant Harley (ASSUMING WE USE SMOKIES TREE RING RECORDS)*
+*Grant Harley*
 
@@
 DOYtiming_caption <- "Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).
-**DRAFT FIGURE-- REAL DATA BUT NEEDS FORMATTING.** Panels (a-b) show leaf phenology...
+Panels (a-b) show leaf phenology...
 Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site.
@@
 
-tree_rings_caption <- "Figure 4 | Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
-Shown are Pearson correlations between ring-width index and and monthly Tmax.
-Species codes are as given in Table 2.(THIS FIGURE WILL PROBABLY BE PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
+tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for ## chronologies from ## sites across eastern North America.
+Chronologies are grouped by xylem porosity and ordered by mean April T_max.
+Chronology details are given in Table S4.
+(NOTE: Figure still needs some work.
+See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"
 
@@
 ```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
-knitr::include_graphics("tables_figures/tree_cores_tmax.tif")
+knitr::include_graphics("tables_figures/tree_rings_tmax.png")
 ```

### 2021-04-16 doc/manuscript/Dow_SI.Rmd
add leaf phenology to Table S3

@@
     column_spec(1, width = "4cm") %>%
+     pack_rows("Stem Growth", 1, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    pack_rows("Leaf Phenology (ecosystem level)", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    #add_indent(9:12)
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2021-04-17 doc/manuscript/Dow_manuscript_Nature.Rmd
a bit of work on writing

@@
 observed climwin mean temperature value, the posterior mean of the fitted day of year
-is represented by the solid blue line and 95% credible intervals are represented by bands."
+is represented by the solid blue line and 95% credible intervals are represented by bands.
+(**SHOULD WE SHOW ONLY THE DOY TRENDS (top row), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)"
 
@@
 Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site.
-Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."
+Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+(**I THINK WE SHOULD COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50).
+)"
 
@@
 n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
-n_cores <- sum(na.omit(species_table[, "n.cores"]))
-
+n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
+n_chronologies <- NA
+n_cores_sites <- NA
+n_chronologies_SCBI <- NA
+n_chronologies_HF <- NA
 ```
@@
 Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
-The latter finding is confirmed on the centennial scale by **#** tree-ring chronologies from **#** forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
+The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
 These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
@@
 
-As global atmospheric CO~2~ concentrations increase and Earth's temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
+As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
 Temperate deciduous forests, including those in eastern North America, *(have been net carbon sinks over the past several decades)*... 
@@
 **We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures.** 
-In fact, we know little about the stem-growth phenology of temperate deciduous species [but see Ref.
-D'Orangeville].
-(*Here, we need to review the literature to make sure we're not missing anything.
-We also want to set up the fact that our dataset is unique/justify just 2 sites*).
-Tree-ring records give some hints that warmer springs may increase annual growth (e.g., Refs.[@helcoski_growing_2019] [MORE]), but rely purely on correlations of annual growth to monthly temperatures and can tell us nothing about the timing of that growth.
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [D'Orangeville].
+*(We need to make sure we're not missing anything.)*
+Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
+Thus, tree-ring records do not necessarily corroborate findings that warming spring temperatures lengthen the growing season and increase annual CO~2~ sequestration. 
+Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
 
-*Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on how forest biomass growth is likely to respond to warming spring temperatures.
-To produce the most complete picture of spring influence on forests, a combination of established data types - tree cores, for inter-annual stem growth, and remote sensing, for leaf phenology - should be used in conjunction with a recently developed intra-annual modeling technique [@mcmahon_general_2015].
-*
-**Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings, which differ in growth phenology[*D'Orangeville*].**
-First, using data from dendrometer bands measured throughout the growing season in two forests in the Eastern USA (n=`r n_tree_years` tree-years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+**Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests (Fig.
+`r schematic_summary`) and annual growth across these and an additional `r n_cores_sites-2` forests spanning eastern North America.**
+First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **##** trees over 10 years) and Harvard Forest (Massachusetts; n = **##** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
-*(We also look at leaf phenology.)*
-Second, at these focal sites and *#* additional sites distributed *xxxx*, we analyze *#* tree-ring chronologies to understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale.
+This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[*D'Orangeville*], and set in the context of leaf phenology at these sites.
 
@@
 
-Similarly, between species-level residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Figs.
+**To understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies for `r n_chronologies_SCBI` species at SCBI, `r n_chronologies_HF` species at Harvard Forest.**
+*(The following should be condensed:)*
+Residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Figs.
 `r tree_rings`, `r tree_rings_TMIN`).
@@
 There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.
+
+**To set these findings in a broader scale context, we analyzed an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2`.** 
+In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## degree range in April T.** 
+*(describe results across all sites)*
+(Fig. `r tree_rings`)
 
@@
 At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
-In total, we analyzed `r n_cores` cores from eleven species at SCBI and three species at Harvard Forest (Table `r table_species`).
+In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (Table `r table_species`).
 

### 2021-04-26 doc/manuscript/Dow_manuscript_Nature.Rmd
add some refs, made some changes

@@
 is represented by the solid blue line and 95% credible intervals are represented by bands.
-(**SHOULD WE SHOW ONLY THE DOY TRENDS (top row), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)"
+(**SHOULD WE SHOW ONLY THE DOY TRENDS and Lpgs (top 2 rows), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)"
 
@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[REFS] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
 However, little is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood, which has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 
-**In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003] and *often later autumn*[REFS].**
-In temperate deciduous forests, spring leaf-out has been occurring earlier[REFS], as has the beginning of the annual period of net CO~2~ sequestration[REFS].
+**In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and *often later autumn*[@ibanez_forecasting_2010,@jeong_phenology_2011].**
+In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[REFS].
 Controls on the timing of autumn senescence are less clearly understood (*need to do more lit review on this*).
 Experimental evidence demonstrates that warmer spring temperatures lead to earlier leaf-out and earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020], whereas warmer summer and autumn temperatures delay leaf senescence[@zohner_ongoing_2019;@crabbe_extreme_2016;@dragoni_evidence_2011;@jeong_phenology_2011;@ibanez_forecasting_2010;@vitasse_responses_2009].
-Whatever the mechanisms controlling autumn senescence, the observed pattern is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[REFS] and the ecosystem is a net CO~2~ sink[REFS].
+Whatever the mechanisms controlling autumn senescence, the observed pattern is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
 
@@
 `r schematic_summary`) and annual growth across these and an additional `r n_cores_sites-2` forests spanning eastern North America.**
-First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **##** trees over 10 years) and Harvard Forest (Massachusetts; n = **##** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 
-We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity.
-Using dendroband data, we found that the peak growing season is generally advancing as the critical spring windows are warming.
+We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity ( `r schematic_summary`- `r DOYtiming`).
+Using dendroband data, we found that the peak growing season generally advanced as the critical spring windows warmed (Fig.
+``r DOYtiming``).
 Despite this shift, growth variables $L_{pgs}$.
-$g_{max}$ and $\Delta DBH$ were virtually unchanged when this period warmed, but $DOY_{g_{max}}$ shifted back almost in unison with $DOY_{50}$.
-The lack of influence Spring TMAX exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores, which show that annual tree growth depends more on peak growing season conditions than early season advancement.
+$g_{max}$ and $\Delta DBH$ were virtually unchanged when this period warmed, but $DOY_{g_{max}}$ shifted back almost in unison with $DOY_{50}$ (Fig 2).
+The lack of influence Spring $T_{max}$ exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores, which show that annual tree growth depends more on peak growing season conditions than early season advancement.
 
@@
 Indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
-This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*
+This is consistent with previous studies showing that ...
+*(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T--- This could be tough.
+It looks like that varies alot by region and species.)*
 
@@
 Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
-Warm spring temperatures result in earlier leaf-out (fig., REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
+Warm spring temperatures result in earlier leaf-out (Fig.
+`r doytiming` , REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
 `r schematic_summary`- `r DOYtiming`).

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
update tree-ring figure
#49

@@
   DOYtiming_Harvard = "3b"
-tree_rings = 4  #stays, hopefully adding a lot more sites
+tree_rings = 4
 
@@
 Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-(**I THINK WE SHOULD COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50).
+(**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50).
 )"
@@
 (NOTE: Figure still needs some work.
-See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"
+**COLOR SCALE IS  A BIT OFF.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"
 
@@
 # Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming_stemgrowth+foliage.png")
+knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
 ```
@@
 Using dendroband data, we found that the peak growing season generally advanced as the critical spring windows warmed (Fig.
-``r DOYtiming``).
+`r DOYtiming`).
 Despite this shift, growth variables $L_{pgs}$.
@@
 Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Figs.
-`r schematic_summary`-`r `r DOYtiming``).
+`r schematic_summary`-`r DOYtiming`).
 This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle. 
@@
 Warm spring temperatures result in earlier leaf-out (Fig.
-`r doytiming` , REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
+`r DOYtiming` , REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
 `r schematic_summary`- `r DOYtiming`).

### 2021-04-27 doc/manuscript/Dow_SI.Rmd
worked on SI contents

--- 2020-12-10 doc/manuscript/Dow_SI.Rmd
@@
 ---
-title: ''
-author: ''
+title: "Supplementary Information"
+author:
 output: 
   pdf_document:
-    citation_package: natbib
     fig_caption: yes
-  html_document:
-    df_print: paged
-bibliography: null
-biblio-style: apalike
-csl: new-phytologist.csl
-header-includes:
-- \usepackage{booktabs}
-- \usepackage{longtable}
-- \usepackage{array}
-- \usepackage{multirow}
-- \usepackage{wrapfig}
-- \usepackage{float}
-- \usepackage{colortbl}
-- \usepackage{pdflscape}
-- \usepackage{tabu}
-- \usepackage{threeparttable}
-- \usepackage{threeparttablex}
-- \usepackage[normalem]{ulem}
-- \usepackage{makecell}
-- \usepackage{xcolor}
-- \usepackage{float}
-- \floatplacement{figure}{H}
+    #toc: true # table of content true
+    #toc_depth: 2  
+header-includes:  \usepackage{float}
+                  \usepackage{caption}
+                  \captionsetup[table]{font=footnotesize}
+                  \captionsetup[figure]{font=footnotesize}
+                  \captionsetup[figure]{labelformat=empty}
+                  \captionsetup[table]{labelformat=empty}
+                  \usepackage{pdflscape}
+                  \newcommand{\blandscape}{\begin{landscape}}
+                  \newcommand{\elandscape}{\end{landscape}}
 ---
@@
 table_species="S1"  
+sample_size_by_species= "S2"
 sample_size_by_year= "S2"
 mean_growth_parameters_table="S3" 
-
-#SI table captions
-sample_size_by_year_caption <- "Table S1. Sample size by year "
+chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
 
@@
 growth_curves_all = "S1"  
-all_climwin_SI_figs = " S2-S5"
-  climwin_rp_DOY25_SCBI = "S2"
-  climwin_dp_DOY25_SCBI = "S3"
-  climwin_rp_DOY25_HARV = "S4"
-  climwin_dp_DOY25_HARV = "S5"
+climwin = "S2"
+tree_rings_TMIN = "S3"
 
-tree_rings_TMIN = "S7"
+#SI table captions
+species_table_caption <-"Table S1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes analyzed here."
+sample_size_by_species_caption <-"Table EXTRA | Dendrometer band sample size by species at SCBI and HF. TO BE REMOVED"
+sample_size_by_year_caption <- "Table S2 | Sample size by year "
+mean_growth_parameters_caption = "Table S3 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "
 
 #SI figure captions
-growth_curves_all_caption  <- "Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
+growth_curves_all_caption  <- " Figure S1 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
 Each line represents one year's growth for a given tree, fit with McMahon model.
 "
-climwin_rp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
-climwin_dp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at SCBI."
-climwin_rp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at Harvard Forest."
-climwin_dp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
 
-tree_rings_TMIN_short_caption <- "Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
-Shown are Pearson correlations between ring-width index and and monthly Tmin.
-Species codes are as given in Table 2.(THIS FIGURE WILL PROBABLY BE PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
+climwin_caption <- "Figure S2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
+Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
+Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
+19- May 21 at SCBI, March 19 - May 07 at HF)."
+
+tree_rings_TMIN_short_caption <- "Figure S3 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for ## chronologies from ## sites across eastern North America.
+Chronologies are grouped by xylem porosity and ordered by mean April T_max.
+Chronology details are given in Table S4.
+(**NOTE: Figure still needs some work.
+Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"
 
@@
 species_table <- read.csv(file = "tables_figures/species_table.csv", col.names = c("site", "xylem porosity","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
-kable(species_table, caption = "Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes analyzed here.",  format = "latex", booktabs = TRUE,  escape = F) %>%
+kable(species_table, caption = species_table_caption ,  format = "latex", booktabs = TRUE,  escape = F) %>%
 add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
@@
 table <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, caption = "SCBI bands by species", format = "latex", booktabs = T) %>%
+kbl(table, caption = sample_size_by_species_caption , format = "latex", booktabs = T) %>%
   kable_styling(latex_options = "hold_position")
 ```
-
-\newpage
 
@@
 table <- read.csv("tables_figures/HF_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, caption = "HF bands by species", format = "latex", booktabs = T) %>%
+kbl(table, caption = sample_size_by_species_caption, format = "latex", booktabs = T) %>%
   kable_styling(latex_options = "hold_position")
 ```
+
+*These tables overlap with the species table, and probably is not necessary. They also do not match-- probably out of date.*
 
@@
 table <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, caption = "sample size by year", format = "latex", booktabs = T) %>%
+kbl(table, caption = sample_size_by_year_caption, format = "latex", booktabs = T) %>%
   kable_styling(latex_options = "hold_position")
@@
 mean_growth_parameters <- read.csv("tables_figures/mean_growth_parameters.csv",  col.names = c("", "ring","diffuse","ring", "diffuse"), stringsAsFactors = FALSE, check.names = FALSE)
-kable(mean_growth_parameters, caption = "Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest.
-",  format = "latex", booktabs = TRUE,  escape = F) %>%
+kable(mean_growth_parameters, caption = mean_growth_parameters_caption ,  format = "latex", booktabs = TRUE,  escape = F) %>%
 add_header_above(c(" ",  "SCBI" = 2, "Harvard Forest"=2 )) %>%
     column_spec(1, width = "4cm") %>%
+     pack_rows("Stem Growth", 1, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    pack_rows("Leaf Phenology (ecosystem level)", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    #add_indent(9:12)
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
@@
 \newpage
+# (placeholder for chronologies table.)
 
-```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption, fig.pos="H"}
-# Created in RScripts/additional_figures.R
+\newpage
+
+```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption, fig.pos="H", warning=FALSE}
 knitr::include_graphics("tables_figures/growth_curves_all.png")
@@
 \newpage
-
-**(a) $T_{max}$**
-```{r climwin_rp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
-knitr::include_graphics("tables_figures/SCBI_ring-porous_TMAX.png")
-```
-
-**(b) $T_{min}$**
-```{r climwin_rp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_SCBI_short_caption, fig.pos="H"}
-knitr::include_graphics("tables_figures/SCBI_ring-porous_TMIN.png")
+```{r climwin, out.width="70%", echo=FALSE, fig.align="center", fig.cap= climwin_caption, warning=FALSE}
+knitr::include_graphics("tables_figures/climwin_figure.png")
 ```
@@
 \newpage
-
-**(a) $T_{max}$**
-```{r climwin_dp_tmax_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
-knitr::include_graphics("tables_figures/SCBI_diffuse-porous_TMAX.png")
+```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption, fig.pos="H", warning=FALSE}
+knitr::include_graphics("tables_figures/tree_rings_tmin.png")
 ```
-
-**(b) $T_{min}$**
-```{r climwin_dp_tmin_DOY25_SCBI, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_SCBI_short_caption, fig.pos="H"}
-knitr::include_graphics("tables_figures/SCBI_diffuse-porous_TMIN.png")
-```
-
-\newpage
-
-**(a) $T_{max}$**
-```{r climwin_rp_tmax_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
-knitr::include_graphics("tables_figures/HF_ring-porous_TMAX.png")
-```
-
-**(b) $T_{min}$**
-```{r climwin_rp_tmin_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_rp_DOY25_HARV_short_caption, fig.pos="H"}
-knitr::include_graphics("tables_figures/HF_ring-porous_TMIN.png")
-```
-
-\newpage
-
-**(a) $T_{max}$**
-```{r climwin_dp_tmax_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.pos="H"}
-knitr::include_graphics("tables_figures/HF_diffuse-porous_TMAX.png")
-```
-
-**(b) $T_{min}$**
-```{r climwin_dp_tmin_HARV, out.width="70%", echo=FALSE, fig.align="left", fig.cap= climwin_dp_DOY25_HARV_short_caption, fig.pos="H"}
-knitr::include_graphics("tables_figures/HF_diffuse-porous_TMIN.png")
-```
-
-\newpage
-```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption, fig.pos="H"}
-#knitr::include_graphics("tables_figures/tree_cores_tmin.tif")
-```

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
worked on SI contents

@@
 table_species="S1"  
+sample_size_by_species= "S2"
 sample_size_by_year= "S2"
 mean_growth_parameters_table="S3" 
-chronologies_table="S4"
-
-#SI table captions
-sample_size_by_year_caption <- "Table S1. Sample size by year "
+chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
 
@@
 growth_curves_all = "S1"  
-all_climwin_SI_figs = " S2-S5"
-  climwin_rp_DOY25_SCBI = "S2"
-  climwin_dp_DOY25_SCBI = "S3"
-  climwin_rp_DOY25_HARV = "S4"
-  climwin_dp_DOY25_HARV = "S5"
-
-tree_rings_TMIN = "S7"
-
-#SI figure captions
-growth_curves_all_caption  <- "Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
-Each line represents one year's growth for a given tree, fit with McMahon model.
-"
-climwin_rp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
-climwin_dp_DOY25_SCBI_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at SCBI."
-climwin_rp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for ring porous species at Harvard Forest."
-climwin_dp_DOY25_HARV_short_caption <- "Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
-
-tree_rings_TMIN_short_caption <- "Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest.
-Shown are Pearson correlations between ring-width index and and monthly Tmin.
-Species codes are as given in Table 2.(THIS FIGURE WILL PROBABLY BE PROBABLY MODIFIED SIGNFICANTLY TO INCLUDE LOTS MORE SITES.)"
+climwin = "S2"
+tree_rings_TMIN = "S3"
 
@@
 For the majority of weekly time windows from Jan.
-1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Figs.
-`r all_climwin_SI_figs`).
+1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Fig.
+`r climwin`).
 This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
@@
 
-The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (Figs. `r all_climwin_SI_figs`). 
-At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (Figs.
-`r climwin_rp_DOY25_SCBI` - `r climwin_dp_DOY25_SCBI`).
+The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (**NEEDS SUPPORTING FIG**). 
+At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (**NEEDS SUPPORTING FIG**).
 In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows greater than five weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
@@
 
-Indeed, the tree-ring records reveal that *annual growth is most commonly controlled by climate conditions during the peak growing season*  [Fig. `r tree_rings`; @helcoski_growing_2019], .
-(*1-2 sentences reviewing our regional analysis results (HOPEFULLY) generalizing about climate sensitivity of temperate deciduous forest growth.*)
+Indeed, the tree-ring records reveal that annual growth is most commonly controlled by climate conditions during the peak growing season [Fig. `r tree_rings`; @helcoski_growing_2019], .
+(*1-2 sentences reviewing our regional analysis results /generalizing about climate sensitivity of temperate deciduous forest growth.*)
 Thus, although our dendrometer band analysis is limited to two sites, the finding that spring temperatures have limited effect on annual tree growth holds generally.
@@
 
-`r sample_size_by_year_caption`
-
-`r climwin_rp_DOY25_SCBI_short_caption`
-
-`r climwin_dp_DOY25_SCBI_short_caption`
-
-`r climwin_rp_DOY25_HARV_short_caption`
-
-`r climwin_dp_DOY25_HARV_short_caption`
-
 # References

### 2021-04-27 doc/manuscript/Dow_SI.Rmd
remove sample size by species tables
#51

@@
 #SI table captions
-species_table_caption <-"Table S1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes analyzed here."
-sample_size_by_species_caption <-"Table EXTRA | Dendrometer band sample size by species at SCBI and HF. TO BE REMOVED"
+species_table_caption <-"Table S1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
 sample_size_by_year_caption <- "Table S2 | Sample size by year "
@@
 ```
-
-\newpage
-
-```{r, Table SCBI_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
-library(knitr)
-library(kableExtra)
-table <- read.csv("tables_figures/SCBI_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, caption = sample_size_by_species_caption , format = "latex", booktabs = T) %>%
-  kable_styling(latex_options = "hold_position")
-```
-
-```{r, Table HF_bands_bySP, eval=TRUE, echo=FALSE, message=FALSE}
-library(knitr)
-library(kableExtra)
-table <- read.csv("tables_figures/HF_bands_bySP.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, caption = sample_size_by_species_caption, format = "latex", booktabs = T) %>%
-  kable_styling(latex_options = "hold_position")
-```
-
-*These tables overlap with the species table, and probably is not necessary. They also do not match-- probably out of date.*
 

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
remove sample size by species tables
#51

@@
 table_species="S1"  
-sample_size_by_species= "S2"
 sample_size_by_year= "S2"

### 2021-04-27 doc/manuscript/archive/Dow_manuscript.Rmd
archive previous version

--- 2021-02-01 doc/manuscript/Dow_manuscript.Rmd
@@
 in prep; @babst_twentieth_2019].
-(*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
 Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree growth will hold generally.

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
archived old version

@@
 n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
-n_chronologies <- NA
+
+#copy over chronologies table
+chronologies_table <- read.csv(file ="tables_figures/chronology_table.csv")
+n_chronologies <- nrow(chronologies_table) 
 n_cores_sites <- NA
@@
 
-# Supporting Information
+# Additional Information
+
+*Supplementary Information* is available for this paper.
+
+Correspondence and requests for materials should be addressed to Kristina Anderson-Teixeira (teixeirak@si.edu).
 

### 2021-04-27 doc/manuscript/Dow_Extended_Data.Rmd
SI --> Extended Data
#52

--- 2021-04-27 doc/manuscript/Dow_SI.Rmd
@@
 ---
-title: "Supplementary Information"
+title: "Extended Data"
 author:
@@
 
-# SUPPLEMENTARY INFO (copied from Dow_manuscript.Rmd)
-##SI table numbers
-table_species="S1"  
-sample_size_by_species= "S2"
-sample_size_by_year= "S2"
-mean_growth_parameters_table="S3" 
+# EXTENDED DATA 
+##tables
+table_species="Extended Data Table 1"  
+sample_size_by_year= "Extended Data Table 2"
+mean_growth_parameters_table="Extended Data Table 3" 
 chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
 
-#SI figure numbers
-growth_curves_all = "S1"  
-climwin = "S2"
-tree_rings_TMIN = "S3"
+##figures
+growth_curves_all = "Extended Data Figure 1"  
+climwin = "Extended Data Figure 2"
+tree_rings_TMIN = "Extended Data Figure 3"
 
-#SI table captions
-species_table_caption <-"Table S1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
-sample_size_by_year_caption <- "Table S2 | Sample size by year "
-mean_growth_parameters_caption = "Table S3 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "
+## table captions
+species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
+sample_size_by_year_caption <- "Extended Data Table 2 | Sample size by year "
+mean_growth_parameters_caption = "Extended Data Table 3 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "
 
-#SI figure captions
-growth_curves_all_caption  <- " Figure S1 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
+## figure captions
+growth_curves_all_caption  <- " Extended Data Figure 1 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
 Each line represents one year's growth for a given tree, fit with McMahon model.
@@
 
-climwin_caption <- "Figure S2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
+climwin_caption <- "Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
 Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
@@
 
-tree_rings_TMIN_short_caption <- "Figure S3 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for ## chronologies from ## sites across eastern North America.
+tree_rings_TMIN_short_caption <- "Extended Data Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for ## chronologies from ## sites across eastern North America.
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in Table S4.
+Chronology details are given in the Supplementary Information.
 (**NOTE: Figure still needs some work.
@@
 
+*This figure needs some work if we choose to include it. It would also probably make sense to include years for which leaf phenology is available.*
+
 \newpage
@@
 \newpage
-# (placeholder for chronologies table.)
-
-\newpage
 
@@
 \newpage
-```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption, fig.pos="H", warning=FALSE}
+```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption, warning=FALSE}
 knitr::include_graphics("tables_figures/tree_rings_tmin.png")

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
SI --> Extended Data
#52

--- 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in Table S4.
+Chronology details are given in the Supplementary Information.
 (NOTE: Figure still needs some work.
@@
 
-# SUPPLEMENTARY INFO (this chunk must be repeated in Dow_SI.Rmd)
-##SI table numbers
-table_species="S1"  
-sample_size_by_year= "S2"
-mean_growth_parameters_table="S3" 
+# EXTENDED DATA  (this chunk must be repeated in Dow_Extended_Data.Rmd)
+##tables
+table_species="Extended Data Table 1"  
+sample_size_by_year= "Extended Data Table 2"
+mean_growth_parameters_table="Extended Data Table 3" 
 chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
 
-#SI figure numbers
-growth_curves_all = "S1"  
-climwin = "S2"
-tree_rings_TMIN = "S3"
+##figures
+growth_curves_all = "Extended Data Figure 1"  
+climwin = "Extended Data Figure 2"
+tree_rings_TMIN = "Extended Data Figure 3"
 
@@
 
-**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (Fig.
-`r growth_curves_all`, Table `r mean_growth_parameters_table`).**
+**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (`r growth_curves_all`, `r mean_growth_parameters_table`).**
 Across all trees and years, the mean $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
@@
 For the majority of weekly time windows from Jan.
-1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (Fig.
-`r climwin`).
+1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (`r climwin`).
 This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
-The strongest relationships, including but not limited to the optimal time windows (Table `r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.
+The strongest relationships, including but not limited to the optimal time windows (`r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.
 
@@
 *(The following should be condensed:)*
-Residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Figs.
+Residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Fig.
 `r tree_rings`, `r tree_rings_TMIN`).
@@
 
-Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (Fig. `r tree_rings_TMIN`). 
+Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r tree_rings_TMIN`). 
 There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.
@@
 
-Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (Table `r table_species`). 
+Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r table_species`). 
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
 Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
-The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`) *(tags_per_year_preclean in data folder)*.
+The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (`r sample_size_by_year`) *(tags_per_year_preclean in data folder)*.
 Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
-In total, our analysis included `r n_tree_years` tree-years (Table `r table_species`).
+In total, our analysis included `r n_tree_years` tree-years (`r table_species`).
 
@@
 Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
-We analyzed records for the dominant ring- and diffuse- porous species at each site (Table `r table_species`), excluding species with other xylem architectures (*Juglans nigra* L.
+We analyzed records for the dominant ring- and diffuse- porous species at each site (`r table_species`), excluding species with other xylem architectures (*Juglans nigra* L.
 at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
@@
 At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
-In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (Table `r table_species`).
+In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r table_species`).
 
@@
 We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].*
-We defined chronology start year (Table `r table_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
+We defined chronology start year (`r table_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
 
@@
 This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window.
-Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (Table `r mean_growth_parameters_table`).
+Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r mean_growth_parameters_table`).
 *The critical time window is determined based on the subset of models that fall within the 95% confidence set.*

### 2021-04-27 doc/manuscript/Dow_Extended_Data.Rmd
added chronology counts

@@
 
-tree_rings_TMIN_short_caption <- "Extended Data Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for ## chronologies from ## sites across eastern North America.
+tree_rings_TMIN_short_caption <- "Extended Data Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 207 chronologies from 114 sites across eastern North America.
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
added chronology counts

@@
 
-tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for ## chronologies from ## sites across eastern North America.
+tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America.
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
@@
 n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
-n_chronologies <- NA
-n_cores_sites <- NA
-n_chronologies_SCBI <- NA
-n_chronologies_HF <- NA
+chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
+n_chronologies <- nrow(chronology_table)  #!!! entered by hand in figure captions
+n_cores_sites <- length(unique(chronology_table[, "Location"])) #!!! entered by hand in figure captions
+n_chronologies_SCBI <- 12
+n_chronologies_HF <- 3
 ```
@@
 
-# Supporting Information
+# Additional Information
+
+**Supplementary Information** is available for this paper.
+
+Correspondence and requests for materials should be addressed to Kristina Anderson-Teixeira (teixeirak@si.edu).
 

### 2021-04-27 doc/manuscript/Dow_Extended_Data.Rmd
clean up

@@
 ---
-title: "Extended Data"
+title: ""
 author:
@@
 
-```{r growth_curves_all, out.width="100%", echo=FALSE, fig.align="center", fig.cap= growth_curves_all_caption, fig.pos="H", warning=FALSE}
-knitr::include_graphics("tables_figures/growth_curves_all.png")
-```
+![`r growth_curves_all_caption`](tables_figures/growth_curves_all.png){ width=100% }
 
 \newpage
-```{r climwin, out.width="70%", echo=FALSE, fig.align="center", fig.cap= climwin_caption, warning=FALSE}
-knitr::include_graphics("tables_figures/climwin_figure.png")
-```
+
+![`r climwin_caption`](tables_figures/climwin_figure.png){ width=70% }
 
 \newpage
-```{r tree_rings_TMIN, out.width="120%", echo=FALSE, fig.align="center", fig.cap= tree_rings_TMIN_short_caption, warning=FALSE}
-knitr::include_graphics("tables_figures/tree_rings_tmin.png")
-```
+
+![`r tree_rings_TMIN_short_caption`](tables_figures/tree_rings_tmin.png){ width=90% }

### 2021-04-27 doc/manuscript/Dow_Extended_Data.Rmd
implement tentative Extended Data items
#52

@@
 ##tables
-table_species="Extended Data Table 1"  
-sample_size_by_year= "Extended Data Table 2"
-mean_growth_parameters_table="Extended Data Table 3" 
-chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
+EDT_species="Extended Data Table 1"  
+EDT_mean_growth_parameters="Extended Data Table 2" 
+EDT_tree_ring_summary="Extended Data Table 3"
+
+sample_size_by_year= "XX"
+#chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
 
 ##figures
-growth_curves_all = "Extended Data Figure 1"  
-climwin = "Extended Data Figure 2"
-tree_rings_TMIN = "Extended Data Figure 3"
+EDF_map = "Extended Data Figure 1" 
+EDF_growth_curves_all = "Extended Data Figure 2"  
+EDF_climwin = "Extended Data Figure 3"
+EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
+EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
+EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
+EDF_tree_rings_TMIN = "Extended Data Figure 7"
 
@@
 species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
-sample_size_by_year_caption <- "Extended Data Table 2 | Sample size by year "
-mean_growth_parameters_caption = "Extended Data Table 3 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "
+sample_size_by_year_caption <- "TO BE REMOVED | Sample size by year "
+mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "
+tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring data and analysis ( including species, n chronologies (= n sites), n positively correlated with April T, n sig.)"
 
 ## figure captions
-growth_curves_all_caption  <- " Extended Data Figure 1 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
+growth_curves_all_caption  <- " Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
 Each line represents one year's growth for a given tree, fit with McMahon model.
-"
+" #needs wordsmithing!!!
 
-climwin_caption <- "Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
+climwin_caption <- "Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
 Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
 Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
-19- May 21 at SCBI, March 19 - May 07 at HF)."
+19- May 21 at SCBI, March 19 - May 07 at HF)." #info subject to change!!!
 
-tree_rings_TMIN_short_caption <- "Extended Data Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 207 chronologies from 114 sites across eastern North America.
+pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Relationship between growth parameters and mean maximum temperature within the critical temperature window (CTW).** CTW was defined using *climwin*.
+For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!
+
+pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!
+
+pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!
+
+tree_rings_TMIN_caption <- "Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 207 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
@@
 (**NOTE: Figure still needs some work.
-Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"
+Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
+#info subject to change!!!
 
@@
 
-*This figure needs some work if we choose to include it. It would also probably make sense to include years for which leaf phenology is available.*
+*This figure PROBABLY NEEDS TO GO. It needs some work if we choose to include it. *
 
@@
 
+## (PLACEHOLDER FOR `r tree_ring_summary_caption` )
+
+\newpage
+
+## (PLACEHOLDER FOR MAP OF STUDY SITES -- [GITHUB ISSUE #54](https://github.com/EcoClimLab/growth_phenology/issues/54))
+
+\newpage
+
 ![`r growth_curves_all_caption`](tables_figures/growth_curves_all.png){ width=100% }
@@
 
-![`r tree_rings_TMIN_short_caption`](tables_figures/tree_rings_tmin.png){ width=90% }
+![`r pheno_Tmax_climwin_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }
+
+\newpage
+
+![`r pheno_Tmax_AprilMay_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may.png){ width=90% }
+
+\newpage
+
+![`r pheno_Tmin_AprilMay_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may_tmin.png){ width=90% }
+
+\newpage
+
+![`r tree_rings_TMIN_caption`](tables_figures/tree_rings_tmin.png){ width=90% }

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
implement tentative Extended Data items
#52

@@
 is represented by the solid blue line and 95% credible intervals are represented by bands.
-(**SHOULD WE SHOW ONLY THE DOY TRENDS and Lpgs (top 2 rows), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)"
+(**SHOULD WE SHOW ONLY THE DOY TRENDS and Lpgs (top 2 rows), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)" #needs attention!!!
 
@@
 (**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50).
-)"
+)" #needs attention!!!
 
-tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America.
+tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
@@
 (NOTE: Figure still needs some work.
-**COLOR SCALE IS  A BIT OFF.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"
+**COLOR SCALE IS  A BIT OFF.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
+#info subject to change!!!
 
@@
 ##tables
-table_species="Extended Data Table 1"  
-sample_size_by_year= "Extended Data Table 2"
-mean_growth_parameters_table="Extended Data Table 3" 
-chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
+EDT_species="Extended Data Table 1"  
+EDT_mean_growth_parameters="Extended Data Table 2" 
+EDT_tree_ring_summary="Extended Data Table 3"
 
 ##figures
-growth_curves_all = "Extended Data Figure 1"  
-climwin = "Extended Data Figure 2"
-tree_rings_TMIN = "Extended Data Figure 3"
+EDF_map = "Extended Data Figure 1" 
+EDF_growth_curves_all = "Extended Data Figure 2"  
+EDF_climwin = "Extended Data Figure 3"
+EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
+EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
+EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
+EDF_tree_rings_TMIN = "Extended Data Figure 7"
 
@@
 The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
-These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] must be allocated primarily to relatively short-lived C pools and therefore unlikely to have a lasting impact on the forest C balance.*
+Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
 Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
@@
 
-**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (`r growth_curves_all`, `r mean_growth_parameters_table`).**
+**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (`r  EDF_growth_curves_all`, `r  EDT_mean_growth_parameters`).**
 Across all trees and years, the mean $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
@@
 For the majority of weekly time windows from Jan.
-1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (`r climwin`).
+1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (`r EDF_climwin`).
 This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
-The strongest relationships, including but not limited to the optimal time windows (`r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.
+The strongest relationships, including but not limited to the optimal time windows (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.
 
@@
 Residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Fig.
-`r tree_rings`, `r tree_rings_TMIN`).
+`r tree_rings`, `r EDF_tree_rings_TMIN`).
 Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
@@
 
-Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r tree_rings_TMIN`). 
+Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`). 
 There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.
@@
 
-Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r table_species`). 
+Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
 Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
-The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (`r sample_size_by_year`) *(tags_per_year_preclean in data folder)*.
+The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
 Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
-In total, our analysis included `r n_tree_years` tree-years (`r table_species`).
+In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).
 
@@
 Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
-We analyzed records for the dominant ring- and diffuse- porous species at each site (`r table_species`), excluding species with other xylem architectures (*Juglans nigra* L.
+We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L.
 at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
@@
 At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
-In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r table_species`).
+In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r  EDT_species`).
 
@@
 We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].*
-We defined chronology start year (`r table_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
+We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
 
@@
 This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window.
-Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r mean_growth_parameters_table`).
+Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r  EDT_mean_growth_parameters`).
 *The critical time window is determined based on the subset of models that fall within the 95% confidence set.*

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
remove content on climwin results for Tmin
@camerondow35 , FYI, and let me know if you disagree. This is not central, and we don't have space for supporting figures/ tables (#52). We'll still show Tmin results for the April_May analysis and for the tree-rings.

@@
 The strongest relationships, including but not limited to the optimal time windows (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$. 
-
-The correlation of $DOY_{25}$ to $T_{min}$ was less consistent than that of $T_{max}$ (**NEEDS SUPPORTING FIG**). 
-At SCBI, patterns were quite similar to those described above for $T_{max}$, with similar timing and strengths of correlations across the two temperature variables (**NEEDS SUPPORTING FIG**).
-In contrast, at Harvard Forest, where we had only six years of data, *climwin* selected windows greater than five weeks prior to $DOY_{25}$, had positive correlations between $DOY_{25}$ and $T_{min}$.
-However, for both ring- and diffuse- porous species, average $T_{min}$ of the selected period was below freezing (<0°C) in all years, suggesting that these are sporadic correlations that are unlikely to be biologically meaningful.
-
 Henceforth, we focus on $T_{max}$ assessed over *climwin*-selected critical temperature windows, *noting that similar results are obtained using alternate temperature window ([GitHub issue #42](https://github.com/EcoClimLab/growth_phenology/issues/42)).*

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
focus T sensitivity figure on DOY
#23

@@
 
-schematic_summary_caption <- "Figure 1 | Summary of tree growth changes under warmer spring temperatures.
-(a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
+schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
 
-pheno_Tsensitivity_caption <- "Figure 2 | Relationship between growth parameters and climwin mean temperatures. For each 
-observed climwin mean temperature value, the posterior mean of the fitted day of year
-is represented by the solid blue line and 95% credible intervals are represented by bands.
-(**SHOULD WE SHOW ONLY THE DOY TRENDS and Lpgs (top 2 rows), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)" #needs attention!!!
+DOY_Tsensitivity_caption <- "**Figure 2 | Relationship between growth parameters and climwin mean temperatures.** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.
+(**NOTE THAT THIS PARTIALLY REPLICATES EXTENDED DATA FIG 4**)" #needs attention!!!
 
-DOYtiming_caption <- "Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).
-Panels (a-b) show leaf phenology...
+DOYtiming_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show leaf phenology...
 Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site.
@@
 
-tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+tree_rings_caption <- "**Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
@@
 (NOTE: Figure still needs some work.
-**COLOR SCALE IS  A BIT OFF.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
+**COLOR SCALE IS  A BIT OFF IN THIS COMPOSITE FIGURE (scales not identical across panels).**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
 #info subject to change!!!
@@
 
-For the majority of weekly time windows from Jan.
+For the majority of consecutive weekly time windows from Jan.
 1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (`r EDF_climwin`).
-This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
+This held true for both ring- and diffuse-porous trees at both sites, *albeit* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
 The strongest relationships, including but not limited to the optimal time windows (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$. 
@@
 
-```{r pheno_Tsensitivity, out.width="80%", echo=FALSE, fig.align="center", fig.cap=pheno_Tsensitivity_caption}
+```{r pheno_Tsensitivity, out.width="80%", echo=FALSE, fig.align="center", fig.cap=DOY_Tsensitivity_caption}
 # Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png")
+knitr::include_graphics("tables_figures/DOY_Tsensitivity.png")
 ```

### 2021-04-28 doc/manuscript/Dow_manuscript_Nature.Rmd
authors/ affiliations, formatting cleanup

@@
 output:
-  word_document
-  #bookdown::word_document2: 
-    #reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
-    #fig_caption: yes
+  bookdown::word_document2: 
+    reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+    fig_caption: yes
 bibliography: references.bib
 csl: nature.csl
+
 ---
@@
 
-*Grant Harley*
+Grant Harley
 
@@
 
-Ian McGregor^1,5^
+Ian McGregor^1,6^
 
@@
 
-Sean McMahon^6,7^
+Sean McMahon^7,9^
 
@@
 
-Kristina J. Anderson-Teixeira^1,7^*
+Alan Tepley^1,8^
+
+Kristina J. Anderson-Teixeira^1,9^*
 
@@
 
-5. NCSU
+5. Department of Geography & Geological Sciences, University of Idaho, ID, USA
 
-6. SERC
+6. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
 
-7. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+7. Smithsonian Environmental Research Center, Edgewater, MD, USA
+
+8. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+9. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
 
@@
 As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
-However, little is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood, which has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
+However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
 Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
@@
 **Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests (Fig.
-`r schematic_summary`) and annual growth across these and an additional `r n_cores_sites-2` forests spanning eastern North America.**
+`r schematic_summary`) and annual growth across these and an additional `r n_cores_sites-2` forests spanning eastern North America (`r EDF_map`).**
 First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
@@
 
-```{r schematic_summary, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_summary_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/schematic_summary.png")
-```
+![`r schematic_summary_caption`](tables_figures/schematic_summary.png){ width=100% }
 
@@
 
-```{r pheno_Tsensitivity, out.width="80%", echo=FALSE, fig.align="center", fig.cap=DOY_Tsensitivity_caption}
-# Created in RScripts/bert/wood_phenology_analysis_bert.R
-knitr::include_graphics("tables_figures/DOY_Tsensitivity.png")
-```
+![`r DOY_Tsensitivity_caption`](tables_figures/DOY_Tsensitivity.png){ width=50% }
 
-```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
-# Created in RScripts/additional_figures.R
-knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
-```
+![`r DOYtiming_caption`](tables_figures/DOYtiming_allyears.png){ width=50% }
 
@@
 
-```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
-knitr::include_graphics("tables_figures/tree_rings_tmax.png")
-```
+![`r tree_rings_caption`](tables_figures/tree_rings_tmax.png){ width=100% }
 
@@
 
-**To set these findings in a broader scale context, we analyzed an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2`.** 
+**To set these findings in a broader scale context, we analyzed an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (`r EDF_map`).** 
 In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## degree range in April T.** 
@@
 
-# Methods
+# Methods {-#Methods}
 
-## Study sites
+## Study sites {-#Study}
 
@@
 
-## Data collection and preparation
-### Dendrometer bands
+## Data collection and preparation {-#Data} 
+### Dendrometer bands {-#Dendrometer}
 
@@
 
-### Tree-rings
+### Tree rings {-#Tree} 
 Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
@@
 
-### Climate
+### Climate {-#Climate} 
 For the dendrometer band analysis, we used climate data from local weather stations. 
@@
 
-## Determining critical temperature windows
+## Determining critical temperature windows {-#Determining}
 
@@
 
-## Hypothesis testing
+## Hypothesis testing {-#Hypothesis}
 
@@
 
-# Acknowledgements
+# Acknowledgements {-#Acknowledgements} 
 people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data)
@@
 
-# Author Contributions
+# Author Contributions {-#Author}
 
-# Additional Information
+# Additional Information {-#Additional}
 
@@
 
-# References
+# References {-#References}

### 2021-04-28 doc/manuscript/Dow_manuscript_Nature.Rmd
writing, mainly condensing results

@@
 ```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
-## potential reviewers: Zani
+## potential reviewers: Zani, Keenan
 
@@
 schematic_summary=1 #illustrates parameters
-pheno_Tsensitivity = 2 
-DOYtiming = 3
-  DOYtiming_SCBI = "3a"
-  DOYtiming_Harvard = "3b"
+DOY_timing = 2
+  DOY_timing_SCBI = "2a"  #currently 4 panel!!!!
+  DOY_timing_Harvard = "2b"  #currently 4 panel!!!!
+pheno_Tsensitivity = 3 
 tree_rings = 4
@@
 
-DOYtiming_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show leaf phenology...
+DOY_timing_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show leaf phenology...
 Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site.
@@
 tree_rings_caption <- "**Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
-Chronologies are grouped by xylem porosity and ordered by mean April T_max.
+Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature.
 Chronology details are given in the Supplementary Information.
@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
 However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 
-(*this first paragraph needs to be re-written (very concisely, and focused on the temperate forest C sink and its uncertainty*)
-
 As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
@@
 In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[REFS].
-Controls on the timing of autumn senescence are less clearly understood (*need to do more lit review on this*).
-Experimental evidence demonstrates that warmer spring temperatures lead to earlier leaf-out and earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020], whereas warmer summer and autumn temperatures delay leaf senescence[@zohner_ongoing_2019;@crabbe_extreme_2016;@dragoni_evidence_2011;@jeong_phenology_2011;@ibanez_forecasting_2010;@vitasse_responses_2009].
-Whatever the mechanisms controlling autumn senescence, the observed pattern is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
-
-~~~~~~~~~~
-##(content --with associated refs--that we don't want to lose, but that needs to be worked into reworded text above)
-
-The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*.
-For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016].
-~~~~~~~~~~~~~~~~
+Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
+However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019;@crabbe_extreme_2016;@dragoni_evidence_2011;@jeong_phenology_2011;@ibanez_forecasting_2010;@vitasse_responses_2009].
+The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
 
@@
 **We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures.** 
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [D'Orangeville].
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview].
 *(We need to make sure we're not missing anything.)*
@@
 
-**Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests (Fig.
-`r schematic_summary`) and annual growth across these and an additional `r n_cores_sites-2` forests spanning eastern North America (`r EDF_map`).**
+**Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.**
 First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
-This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[*D'Orangeville*], and set in the context of leaf phenology at these sites.
+This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@doranngeville_radial_inreview], and set in the context of leaf phenology at these sites.
 
-![`r schematic_summary_caption`](tables_figures/schematic_summary.png){ width=100% }
+![`r schematic_summary_caption`](tables_figures/schematic_summary.png){ width=70% }
 
@@
 
-## (overview of phenological patterns)
+Both leaf phenology and stem growth milestones occurred earlier at SCBI than at Harvard Forest, with stem growth milestones achieved on average 6-10 days earlier at SCBI (Fig.
+`r DOY_timing`, `r  EDT_mean_growth_parameters`).
+Consistent with the results of Ref[@doranngeville_radial_inreview], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
+`r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_mean_growth_parameters`).
+Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_mean_growth_parameters`).
 
-*(we will need to add leaf phenology to this paragraph)*
+![`r DOY_timing_caption`](tables_figures/DOYtiming_allyears.png){ width=50% }
 
-**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (`r  EDF_growth_curves_all`, `r  EDT_mean_growth_parameters`).**
-Across all trees and years, the mean $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
-Consistent with the results of Ref[D'Orangeville], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively).
-Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species.
-
-## (Critical temperature windows)
-
-For the majority of consecutive weekly time windows from Jan.
-1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (`r EDF_climwin`).
-This held true for both ring- and diffuse-porous trees at both sites, *albeit* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
-The strongest relationships, including but not limited to the optimal time windows (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$. 
-Henceforth, we focus on $T_{max}$ assessed over *climwin*-selected critical temperature windows, *noting that similar results are obtained using alternate temperature window ([GitHub issue #42](https://github.com/EcoClimLab/growth_phenology/issues/42)).*
-
-## (Climate sensitivity of phenology )
-
-For both ring- and diffuse- porous species at both sites, warmer spring $T_{max}$ resulted in earlier achievement of growth milestones (Fig. `r schematic_summary`, Figs. `r pheno_Tsensitivity`).
-Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g.max}$, leaf greenup, and mid-greenup, at both sites all decreased significantly with increasing $T_{max}$ during the critical temperature window (Figs.
-`r schematic_summary`, `r pheno_Tsensitivity`, `r DOYtiming`).
+Both leaf phenology and stem growth were sensitive to spring temperatures (Figs. `r DOY_timing` -`r pheno_Tsensitivity`). 
+While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-7), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
+1 and $DOY_{25}$  for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window, CTW; Fig.
+`r schematic_summary`).
+The strongest relationships, including but not limited to the CTW (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).
 
@@
 
-![`r DOYtiming_caption`](tables_figures/DOYtiming_allyears.png){ width=50% }
+For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
+Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW.
+Specifically, greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$, and mid-greenup also advanced, but to a lesser extent (Fig.
+`r DOY_timing`).
+Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$ (Figs.
+`r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).
 
-For two of the four species group-site combinations (ring-porous at SCBI and diffuse-porous at Harvard Forest), warmer $T_{max}$ during the critical temperature window resulted in increased $L_{PGS}$ (Figs.
-`r schematic_summary`, `r pheno_Tsensitivity`).
-*At SCBI, ring-porous $DOY_{25}$ and $DOY_{50}$ shifted earlier as a result of warmer springs whereas ${DOY_75}$ did not shift significantly, resulting in a positive correlation between $L_{PGS}$ and $T_{max}$ during the critical temperature.*
-*For diffuse-porous species* At SCBI, all growth milestones occurred earlier in warmer springs, but $DOY_{25}$ was more strongly affected than ${DOY_75}$, resulting in increased $L_{PGS}$.
-At Harvard Forest,  all growth milestones occurred earlier in warmer springs, but ring-porous ${DOY_75}$ shifted more than ${DOY_25}$, resulting in shorter peak growing season length.
-
-Ring- and diffuse- porous species were pooled together at each site for leaf phenology measures.
-Day of year of greenup occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the year with the warmest and coldest critical windows in the analysis (Fig.
-3).
-Mid-greenup timing also shifted forward at both sites when spring temperatures were higher but less so than greenup.
-Senescence occurred earlier at Harvard forest by 12 days in the warm year while SCBI saw a 5 day delay of senescence during the warm year.
+Whereas the length of time between leaf greenup and senescence (i.e., **DEFINE**) increased in years with warmer springs (Fig.
+`r DOY_timing`), there was no consistent lengthening of the period of peak stem growth.
+Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig.
+`r schematic_summary`, Extended Data Figures **4-6**).
 
@@
 
-We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity ( `r schematic_summary`- `r DOYtiming`).
+We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity ( `r schematic_summary`- `r DOY_timing`).
 Using dendroband data, we found that the peak growing season generally advanced as the critical spring windows warmed (Fig.
-`r DOYtiming`).
+`r DOY_timing`).
 Despite this shift, growth variables $L_{pgs}$.
@@
 Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Figs.
-`r schematic_summary`-`r DOYtiming`).
+`r schematic_summary`-`r DOY_timing`).
 This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle. 
@@
 
-Indeed, the tree-ring records reveal that annual growth is most commonly controlled by climate conditions during the peak growing season [Fig. `r tree_rings`; @helcoski_growing_2019], .
+Indeed, the tree-ring records reveal that annual growth is most commonly controlled by climate conditions during the peak growing season (Fig. `r tree_rings`).
 (*1-2 sentences reviewing our regional analysis results /generalizing about climate sensitivity of temperate deciduous forest growth.*)
-Thus, although our dendrometer band analysis is limited to two sites, the finding that spring temperatures have limited effect on annual tree growth holds generally.
+These results align with other studies showing that while warmer spring temperatures can have a positive effect on annual growth for some species, most species show no significant response [@alexander_potential_2019; @helcoski_growing_2019].
+These results align with @fu_maximum_2019, which shows that interannual variation in NEE is more strongly linked to peak C uptake rates than to growing season length.
+Our study demonstrates that this holds true for $ANPP_{woody}$ as well as NEE.
 
@@
 Warm spring temperatures result in earlier leaf-out (Fig.
-`r DOYtiming` , REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
-`r schematic_summary`- `r DOYtiming`).
+`r DOY_timing` , REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
+`r schematic_summary`- `r DOY_timing`).
 Yet, an earlier start to the growing season is not associated with a longer growing season (Figs. `r schematic_summary`, `r pheno_Tsensitivity`) [@zani_increased_2020]
@@
 
-effect of spring temperature on annual growth in New England: @alexander_potential_2019
-
-results align with @fu_maximum_2019, which shows that interannual variation in NEE is more strongly linked to peak C uptake rates than to growing season length.
-Our study demonstrates that this holds true for $ANPP_{woody}$ as well as NEE.
-
 Geng et al. 2020 GCB- Climate warming increases spring phenological differences among temperate trees.
@@
 Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?
-- - RP trees BB = 25% and 75% leaf development = 50% growth (D'Orangeville).
+- - RP trees BB = 25% and 75% leaf development = 50% growth [@doranngeville_radial_inreview].
 Are DP trees leafing up earlier?

### 2021-04-28 doc/manuscript/Dow_manuscript_Nature.Rmd
writing-- results

@@
 
-## (Climate sensitivity of growth)
+**Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig.
+`r schematic_summary`, Extended Data Figures **4-6**).**
+Specifically, $g_{max}$ displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
+while $\Delta DBH$ displayed no relationship CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`).
 
-Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Figs. `r schematic_summary`, `r pheno_Tsensitivity`). 
-Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer $T_{max}$ during the critical temperature window.
-In contrast, diffuse-porous species at both sites exhibited lower $g_{max}$ in years with higher early spring $T_{max}$.
-
-Total annual growth was not consistently linked to spring temperatures (Fig. `r schematic_summary`).
-Annual growth measured with dendrometer bands, $\Delta DBH$, also was not consistently linked to early spring temperatures, displaying either no relationship (SCBI) or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest) (Fig.
-`r pheno_Tsensitivity`).
-
-**To understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies for `r n_chronologies_SCBI` species at SCBI, `r n_chronologies_HF` species at Harvard Forest.**
-*(The following should be condensed:)*
-Residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Fig.
+**To understand how annual growth increments, as registered in ring width indices (RWI), have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig.
+`r tree_rings`, `r EDF_map`).**
+For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig.
 `r tree_rings`, `r EDF_tree_rings_TMIN`).
@@
 In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
-In the prior months, the generally negative correlations to $T_{max}$ were not significant, excepting significant correlations of one species at each site  (ring-porous *Carya tomentosa* in March at SCBI; diffuse-porous *Acer rubrum* in April at Harvard Forest).
-Starting in May at SCBI, significant correlations to $T_{max}$ were common, occurring for nine of the eleven species in May and 3-7 species for each month June through August.
-At Harvard Forest, there was only one statistically significant negative correlation to post-May $T_{max}$ (*Quercus rubra* in June).
+Starting in May at SCBI and June at Harvard Forest, negative correlations of RWI to $T_{max}$ were predominant, being statistically significant in at least one growing season month for all `r n_chronologies_SCBI` species at SCBI and one of `r n_chronologies_HF` species at Harvard Forest.
+Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`).
 
@@
 
-Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`). 
-There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.
-
-**To set these findings in a broader scale context, we analyzed an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (`r EDF_map`).** 
+**The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (`r EDF_map`).** 
 In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## degree range in April T.** 

### 2021-04-28 doc/manuscript/Dow_manuscript_Nature.Rmd
writing- results

@@
 
-**The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (`r EDF_map`).** 
-In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## degree range in April T.** 
-*(describe results across all sites)*
-(Fig. `r tree_rings`)
+**The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (Fig.
+`r tree_rings`; `r EDF_map`).**
+In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
+Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only #% of records: # of # ring-porous and # or # diffuse-porous species-site combinations.
+In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for #% and #% of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
 

### 2021-04-28 doc/manuscript/Dow_manuscript_Nature.Rmd
writing: discussion/ results

@@
 
-DOY_timing_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show leaf phenology...
+DOY_timing_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site.
@@
 For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
-Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW.
+Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively).
 Specifically, greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$, and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
-Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$ (Figs.
+Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing 1-8 days per degree Celsius during the CTW (Figs.
 `r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).
 
-Whereas the length of time between leaf greenup and senescence (i.e., **DEFINE**) increased in years with warmer springs (Fig.
+Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer springs (Fig.
 `r DOY_timing`), there was no consistent lengthening of the period of peak stem growth.
@@
 
-**Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig.
-`r schematic_summary`, Extended Data Figures **4-6**).**
-Specifically, $g_{max}$ displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
+In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs.
+`r schematic_summary`, `r pheno_Tsensitivity`).
+Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig.
+`r schematic_summary`, Extended Data Figures **4-6**).
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
 while $\Delta DBH$ displayed no relationship CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
+Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth rates.
 
-**To understand how annual growth increments, as registered in ring width indices (RWI), have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig.
-`r tree_rings`, `r EDF_map`).**
-For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig.
+To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig.
+`r tree_rings`, `r EDF_map`).
+For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, ring width indices (RWI) rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig.
 `r tree_rings`, `r EDF_tree_rings_TMIN`).
@@
 
-**The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (Fig.
-`r tree_rings`; `r EDF_map`).**
+The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (Fig.
+`r tree_rings`; `r EDF_map`).
 In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
@@
 In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for #% and #% of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
+Again, $T_{min}$ generally exerted less influence over annual growth than $T_{max}$ (`r EDF_tree_rings_TMIN`), with few significant correlations between spring $T_{min}$ and RWI.
 
@@
 
-(*length may be okay, shouldn't be lengthened*)
+Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs.
+`r schematic_summary`- `r tree_rings`).
+The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig.
+`r DOY_timing`)[**REFS**] and NEE [**REFS**].
+*It would be interesting to compare the magnitude of these advances (days per degree C)*
+However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs.
+This suggests that stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
 
-We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity ( `r schematic_summary`- `r DOY_timing`).
-Using dendroband data, we found that the peak growing season generally advanced as the critical spring windows warmed (Fig.
-`r DOY_timing`).
-Despite this shift, growth variables $L_{pgs}$.
-$g_{max}$ and $\Delta DBH$ were virtually unchanged when this period warmed, but $DOY_{g_{max}}$ shifted back almost in unison with $DOY_{50}$ (Fig 2).
-The lack of influence Spring $T_{max}$ exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores, which show that annual tree growth depends more on peak growing season conditions than early season advancement.
+The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on tree growth is modest and not statistically significant for most species (Fig.
+`r tree_rings`, `r EDF_tree_rings_TMIN`).
 
-Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Figs.
-`r schematic_summary`-`r DOY_timing`).
-This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle. 
-Indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
-This is consistent with previous studies showing that ...
-*(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T--- This could be tough.
-It looks like that varies alot by region and species.)*
+*(KAT edits end here.)*
 
-In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs.
-`r schematic_summary`, `r pheno_Tsensitivity`), such that a much longer record would be needed to clarify what, if any, effect spring temperatures have on maximum daily and annual growth rates.
-Fortunately, we were able to use tree-ring data to understand centennial-scale temperature sensitivity of annual growth (Fig. `r tree_rings`).
 This analysis confirms that the effect of spring temperatures on annual growth is modest and not statistically significant for most species, while revealing a slightly more complex picture.
@@
 # Methods {-#Methods}
+
+*We have not yet re-written the methods for Nature. Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.*
 
@@
 At SCBI a higher proportion of RP trees were removed--the majority in step 4 *(data_cleaning_figure_scbi)*.
+
+### Leaf phenology
+description here: https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description
+
+MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019]
 

### 2021-04-29 doc/manuscript/Dow_manuscript_Nature.Rmd
writing: discussion

@@
 
-The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on tree growth is modest and not statistically significant for most species (Fig.
-`r tree_rings`, `r EDF_tree_rings_TMIN`).
+The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig.
+`r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
+Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress driven by increased vapor pressure deficit[**REFS**].
+*It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, effectively reducing drought sensitivity, but this hypothesis remains to be evaluated.*
+Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@fu_maximum_2019].
 
-*(KAT edits end here.)*
+However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases with spring temperatures.
+We show that the extra C does not go into woody growth, raising the question of where this C goes.
 
-This analysis confirms that the effect of spring temperatures on annual growth is modest and not statistically significant for most species, while revealing a slightly more complex picture.
-It appears that temperatures--particularly $T_{max}$--during one spring month at each site (April at SCBI, May at Harvard Forest) consistently have a weakly positive (*albiet* rarely significant) effect on annual growth, in striking contrast to the predominantly negative correlations of growth rates with $T_{max}$ in preceding and subsequent months.
-This suggests that spring $T_{max}$, though it's pronounced influence on the start of woody growth and subtle influence on length of the peak growing season, has a modestly positive influence on annual tree growth, but that this influence is generally of secondary importance to other factors.
-
-Indeed, the tree-ring records reveal that annual growth is most commonly controlled by climate conditions during the peak growing season (Fig. `r tree_rings`).
-(*1-2 sentences reviewing our regional analysis results /generalizing about climate sensitivity of temperate deciduous forest growth.*)
-These results align with other studies showing that while warmer spring temperatures can have a positive effect on annual growth for some species, most species show no significant response [@alexander_potential_2019; @helcoski_growing_2019].
-These results align with @fu_maximum_2019, which shows that interannual variation in NEE is more strongly linked to peak C uptake rates than to growing season length.
-Our study demonstrates that this holds true for $ANPP_{woody}$ as well as NEE.
+*(KAT edits end here)*
 

### 2021-04-29 doc/manuscript/Dow_manuscript_Nature.Rmd
writing; discussion
also, #55

@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[*@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016*; @zhou_explaining_2016;  @fu_maximum_2019].
 However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 
-**The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].**
+**The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019].**
 *This expectation is represented in models [@zani_increased_2020].*
@@
 Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress driven by increased vapor pressure deficit[**REFS**].
-*It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, effectively reducing drought sensitivity, but this hypothesis remains to be evaluated.*
-Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@fu_maximum_2019].
+It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, effectively reducing drought sensitivity, but this hypothesis remains to be evaluated.
 
-However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases with spring temperatures.
+Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
+However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
 We show that the extra C does not go into woody growth, raising the question of where this C goes.
+There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
+First, C that is fixed in late summer or fall may be used for woody growth that is not apparent in the current year's diameter increment: either used to provide extra thickening of cell walls, a process that lags behind stem expansion[REF], or saved within trees as non-structural carbohydrates and used towards growth the following year.
+Indeed, tree-ring evidence from SCBI indicates that favorable growing conditions during this time enhance the next year's growth[@helcoski_growing_2019].
+Second, the additional C may be allocated, likely in the following growing season, to short-lived C pools such as reproduction, foliage, fine roots, or root exudates.
+Most of this carbon would have a relatively low residence time in the ecosystem, although a small fraction would be incorporated into soil organic carbon pools with residence times rivaling that of wood [REFS].
+Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@keenan_net_2014;@zhou_explaining_2016; @fu_maximum_2019].
 
 *(KAT edits end here)*
+
+*There is a difference between interannual variation and directional change: the latter is likely to result in acclimatization, and eventually changes in forest composition/ adaptation.* 
+If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
 
@@
 # (Content to incorporate)
-
-However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019].
-This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE.
 
@@
 
-Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
-
 The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; D’Orangeville et al.
 in review).
-Species have co-evolved to utilize the phenological stages of trees at specific times in the year.
-If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees ().
-Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).
 

### 2021-04-29 doc/manuscript/Dow_manuscript_Nature.Rmd
writing: discussion

@@
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
-First, C that is fixed in late summer or fall may be used for woody growth that is not apparent in the current year's diameter increment: either used to provide extra thickening of cell walls, a process that lags behind stem expansion[REF], or saved within trees as non-structural carbohydrates and used towards growth the following year.
-Indeed, tree-ring evidence from SCBI indicates that favorable growing conditions during this time enhance the next year's growth[@helcoski_growing_2019].
-Second, the additional C may be allocated, likely in the following growing season, to short-lived C pools such as reproduction, foliage, fine roots, or root exudates.
-Most of this carbon would have a relatively low residence time in the ecosystem, although a small fraction would be incorporated into soil organic carbon pools with residence times rivaling that of wood [REFS].
-Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@keenan_net_2014;@zhou_explaining_2016; @fu_maximum_2019].
+First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
+Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, and that this dynamic is likely complex[@helcoski_growing_2019].
+Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
+Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
+Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
+A study examining long-term trends in growing season length and NEE[@keenan_net_2014] -- as opposed to their interannual variation -- showed increasing trends in both within the temperate deciduous biome, suggesting that the C that is not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
+
+Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
+In addition to the existence of temporal lags between C uptake and release, which introduce a lot of uncertainty[@trumbore_quantifying_2002], trees are likely to  acclimate *to some extent* to warming temperatures[**REFS?**], and on even longer time scales, we are likely to see changes in forest composition and adaptation.
+If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
+*Future research will be important to sortitng this out.*
 
@@
 
-*There is a difference between interannual variation and directional change: the latter is likely to result in acclimatization, and eventually changes in forest composition/ adaptation.* 
-If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
-
-Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
+Despite remaining uncertainties, our findings contribute substantially to a broader picture of the influence of spring temperatures on forest ecosystem function. 
 Warm spring temperatures result in earlier leaf-out (Fig.
-`r DOY_timing` , REFS), an earlier start to net C sequestration (REFS), and, as shown here, earlier initiation of stem growth (Figs.
-`r schematic_summary`- `r DOY_timing`).
-Yet, an earlier start to the growing season is not associated with a longer growing season (Figs. `r schematic_summary`, `r pheno_Tsensitivity`) [@zani_increased_2020]
-because growth finishes sooner under these conditions, presumably because of *C sink limitation* [@zani_increased_2020].
+`r DOY_timing` , REFS), an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs.
+`r schematic_summary`- `r pheno_Tsensitivity`).
+Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs.
+`r schematic_summary`- `r tree_ring`), presumably because stem diameter increases are curtailed by other factors such as...., supporting the concept of *C sink limitation* [@zani_increased_2020].
 

### 2021-04-29 doc/manuscript/Dow_manuscript_Nature.Rmd
writing + clean-up

@@
 
-**Authors (not yet complete, just alphabetical for now--final order TBD):**
+**Authors:**
 
@@
 
-Grant Harley
+Grant Harley^5^
 
-Justin Maxwell
+Justin Maxwell^6^
 
-Ian McGregor^1,6^
+Ian McGregor^1,7^
 
@@
 
-Sean McMahon^7,9^
+Sean McMahon^8,10^
 
@@
 
-Alan Tepley^1,8^
+Alan Tepley^1,9^
 
-Kristina J. Anderson-Teixeira^1,9^*
+Kristina J. Anderson-Teixeira^1,10^*
 
@@
 
-6. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
+6. Department of Geography, Indiana University, Bloomington, Indiana, USA
 
-7. Smithsonian Environmental Research Center, Edgewater, MD, USA
+7. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
 
-8. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+8. Smithsonian Environmental Research Center, Edgewater, MD, USA
 
-9. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+9. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+10. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
 
@@
 
-(*word limit 200. italic=tentative*)
+(*word limit 200 - this is currently way too long.*)
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[*@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016*; @zhou_explaining_2016;  @fu_maximum_2019].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
 However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 
-The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (Fig.
-`r tree_rings`; `r EDF_map`).
+The patterns observed at our focal sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
+`r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
 In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
@@
 `r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
-Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress driven by increased vapor pressure deficit[**REFS**].
-It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, effectively reducing drought sensitivity, but this hypothesis remains to be evaluated.
+Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
+It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019], but this hypothesis remains to be evaluated.
 
@@
 First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
-Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, and that this dynamic is likely complex[@helcoski_growing_2019].
+Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
 Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
@@
 Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
-A study examining long-term trends in growing season length and NEE[@keenan_net_2014] -- as opposed to their interannual variation -- showed increasing trends in both within the temperate deciduous biome, suggesting that the C that is not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
+Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome  showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
 
@@
 If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
-*Future research will be important to sortitng this out.*
-
-*(KAT edits end here)*
+*Future research will be important to sorting this out.*
 
@@
 Warm spring temperatures result in earlier leaf-out (Fig.
-`r DOY_timing` , REFS), an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs.
+`r DOY_timing`)[@jeong_phenology_2011], an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs.
 `r schematic_summary`- `r pheno_Tsensitivity`).
 Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs.
-`r schematic_summary`- `r tree_ring`), presumably because stem diameter increases are curtailed by other factors such as...., supporting the concept of *C sink limitation* [@zani_increased_2020].
+`r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@dorangeville_drought_2018;@helcoski_growing_2019].
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models (Figs.
-`r schematic_summary`, `r pheno_Tsensitivity`, `r tree_rings`)[@zani_increased_2020].
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the sustained increase in carbon sequestration that is anticipated in most models (Fig.
+`r schematic_summary`)[@zani_increased_2020].
 Because of this, temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that these models are underestimating future overall atmospheric CO~2~.
+(*see [GitHub issue 56 on representation in models](https://github.com/EcoClimLab/growth_phenology/issues/56)*)
 
@@
 
-From Neil: @zhang_drought_2021, @zhang_extended_2020
+From Neil: @zhang_drought_2021; @zhang_extended_2020
 
@@
 Similar to 25,50,75% DOY retreat?
-
-The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; D’Orangeville et al.
-in review).
 
@@
 ### Leaf phenology
+
+Leaf phenology data for the years ### - ### were extracted for SCBI and Harvard forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
+
 description here: https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description
-
-MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019]
 
@@
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning **#** deciduous and mixed forest sites in Eastern North America.
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@dorangeville_drought_2018] (**OTHER REFS??**).
 Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).
@@
 
-## Determining critical temperature windows {-#Determining}
+## Analyses
+### Determining critical temperature windows {-#Determining}
 
 The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@vandepol_identifying_2016]. 
-This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window.
+This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and $\Delta AIC$ of each window.
 Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r  EDT_mean_growth_parameters`).
@@
 
-## Hypothesis testing {-#Hypothesis}
+### Hypothesis testing {-#Hypothesis}
 
-(*This section needs more detail.*)
+(*This section needs work/ more detail.*)
 

### 2021-04-30 doc/manuscript/Dow_manuscript_Nature.Rmd
misc cleanup, refs

@@
 
-DOY_Tsensitivity_caption <- "**Figure 2 | Relationship between growth parameters and climwin mean temperatures.** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.
-(**NOTE THAT THIS PARTIALLY REPLICATES EXTENDED DATA FIG 4**)" #needs attention!!!
-
-DOY_timing_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
-Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site.
+DOY_timing_caption <- "**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product(**YEARS**), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
 Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
@@
 )" #needs attention!!!
+
+DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25, 50 and 75% of annual growth is acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) decline significantly with mean $T_{max}$ during the critical temperature window (CTW).
+For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands.
+(**DOYrmax to be dropped.
+NOTE THAT THIS PARTIALLY REPLICATES EXTENDED DATA FIG 4**)" #needs attention!!!
 
@@
 
+*NOTE: In general, this needs to be condensed.
+The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
+We're currently at ~2700 words (or ~3000 with fig captions) and 2 modest + 2 large display items.
+Articles typically have 30-50 references (currently ok there)*
+
 # (Summary paragraph)
@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
 However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
-These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+These findings align with findings that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
@@
 
-# (Introduction)
-
+***(DROP THIS FIRST PARAGRAPH?)***
 As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
@@
 
-**In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and *often later autumn*[@ibanez_forecasting_2010,@jeong_phenology_2011].**
-In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[REFS].
+***(CONDENSE THE FOLLOWING 2 PARAGRAPHS INTO 1, AND REMOVE REPETITION WITH ABSTRACT:)***
+
+**In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].**
+In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
 Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
-However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019;@crabbe_extreme_2016;@dragoni_evidence_2011;@jeong_phenology_2011;@ibanez_forecasting_2010;@vitasse_responses_2009].
+However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009].
 The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
 
-**The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019].**
+The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019].
 *This expectation is represented in models [@zani_increased_2020].*
@@
 
-# (Results)
+## Dendrometer band analysis {-#Dendrometer}
 
@@
 
-![`r DOY_timing_caption`](tables_figures/DOYtiming_allyears.png){ width=50% }
+![`r DOY_timing_caption`](tables_figures/DOYtiming_allyears.png){ width=25% }
 
@@
 
+## Tree ring analysis {-#Tree}
+
 To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig.
@@
 
-# (Discussion)
+## Discussion {-#Discussion}
 
@@
 The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig.
-`r DOY_timing`)[**REFS**] and NEE [**REFS**].
+`r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
 *It would be interesting to compare the magnitude of these advances (days per degree C)*
@@
 Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
-It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019], but this hypothesis remains to be evaluated.
+It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].
 
@@
 Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
+*(we can easily test this: [GitHub issue 57](https://github.com/EcoClimLab/growth_phenology/issues/57))*
 Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].

### 2021-04-30 doc/manuscript/Dow_Extended_Data.Rmd
remove sample_size_by_year table
#52

@@
 
-```{r, Table sample_size_by_year, eval=TRUE, echo=FALSE, message=FALSE, fig.pos="H"}
-library(knitr)
-library(kableExtra)
-table <- read.csv("tables_figures/sample_size_by_year.csv",stringsAsFactors = FALSE, check.names = FALSE)
-kbl(table, caption = sample_size_by_year_caption, format = "latex", booktabs = T) %>%
-  kable_styling(latex_options = "hold_position")
-```
-
-*This figure PROBABLY NEEDS TO GO. It needs some work if we choose to include it. *
-
-\newpage
-
 ```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}

### 2021-04-30 doc/manuscript/Dow_Extended_Data.Rmd
version sent to coauthors for initial review

@@
 pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Relationship between growth parameters and mean maximum temperature within the critical temperature window (CTW).** CTW was defined using *climwin*.
-For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!
+For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
+CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
 
-pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!
+pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
+CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
 
-pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!
+pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
+CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
 

### 2021-04-30 doc/manuscript/Dow_manuscript_Nature.Rmd
version sent to coauthors for initial review

@@
 
-**Title:** Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity
+**Title:** Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity */annual increments*
 
@@
 
-DOY_timing_caption <- "**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product(**YEARS**), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+DOY_timing_caption <- "**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
 Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-(**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50).
+(**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?
+(KAT thinks so)** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50).
 )" #needs attention!!!
 
-DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25, 50 and 75% of annual growth is acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) decline significantly with mean $T_{max}$ during the critical temperature window (CTW).
+DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declinde significantly with mean $T_{max}$ during the critical temperature window (CTW).
 For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands.
-(**DOYrmax to be dropped.
-NOTE THAT THIS PARTIALLY REPLICATES EXTENDED DATA FIG 4**)" #needs attention!!!
+Relationsips for additional growth parameters (Fig.
+1a) aare given in Extended Data Figure 4.
+(**DOYrmax will be dropped.** [GitHub Issue 23](https://github.com/EcoClimLab/growth_phenology/issues/23))" #needs attention!!!
 
@@
 
+*The introduction and summary paragraph need work, and removal of overlap. Organizational/content suggestions are welcome, but we're not at the stage of word-smithing.*
+
 ***(DROP THIS FIRST PARAGRAPH?)***
@@
 
-***(CONDENSE THE FOLLOWING 2 PARAGRAPHS INTO 1, AND REMOVE REPETITION WITH ABSTRACT:)***
-
-**In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].**
+In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
 In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
 Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
-However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009].
+However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence [REFS].
 The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
-
-The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019].
-*This expectation is represented in models [@zani_increased_2020].*
+The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@zani_increased_2020].
 However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].
 
-**We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures.** 
+We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures. 
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview].
@@
 
-**Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.**
+Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
+
+## Dendrometer band analysis {-#Dendrometer}
+
 First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
@@
 ![`r schematic_summary_caption`](tables_figures/schematic_summary.png){ width=70% }
-
-## Dendrometer band analysis {-#Dendrometer}
 
@@
 Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively).
-Specifically, greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$, and mid-greenup also advanced, but to a lesser extent (Fig.
+Specifically, [greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
-Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing 1-8 days per degree Celsius during the CTW (Figs.
+Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing [1-8 days per degree Celsius](https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs.
 `r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).
@@
 (*see [GitHub issue 56 on representation in models](https://github.com/EcoClimLab/growth_phenology/issues/56)*)
-
-# (Content to incorporate)
-
-@delgado_differences_2020
-
-From Neil: @zhang_drought_2021; @zhang_extended_2020
-
-Geng et al. 2020 GCB- Climate warming increases spring phenological differences among temperate trees.
-
-N.
-American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts.
-Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?
-- - RP trees BB = 25% and 75% leaf development = 50% growth [@doranngeville_radial_inreview].
-Are DP trees leafing up earlier?
-If yes, how much earlier?
-Similar to 25,50,75% DOY retreat?
+m
 
@@
 
-Leaf phenology data for the years ### - ### were extracted for SCBI and Harvard forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
+Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
 

### 2021-06-06 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
Create Dow_manuscript_Nature_cd.Rmd
Creating my own temp version of this to edit because i'm more comfortable having two versions to compare.

--- 2021-04-30 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 
-Grant Harley^5^
+Grant L. Harley^5^
 
-Justin Maxwell^6^
+Justin T. Maxwell^6^
 
-Ian McGregor^1,7^
+Ian R. McGregor^1,7^
 
@@
 
-Alan Tepley^1,9^
+Alan J. Tepley^1,9^
 
@@
 
-5. Department of Geography & Geological Sciences, University of Idaho, ID, USA
+5. Department of Geography & Geological Sciences, University of Idaho, ID 83844, USA
 
-6. Department of Geography, Indiana University, Bloomington, Indiana, USA
+6. Department of Geography, Indiana University, Bloomington, IN 47405, USA
 
@@
 
-schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
+schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window (CTW)', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
 
@@
 
-DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declinde significantly with mean $T_{max}$ during the critical temperature window (CTW).
+DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW).
 For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands.
-Relationsips for additional growth parameters (Fig.
-1a) aare given in Extended Data Figure 4.
+Relationships for additional growth parameters (Fig.
+1a) are given in Extended Data Figure 4.
 (**DOYrmax will be dropped.** [GitHub Issue 23](https://github.com/EcoClimLab/growth_phenology/issues/23))" #needs attention!!!
@@
 Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
-The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
-These findings align with findings that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+
+*The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.*
+*These findings align with previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019] and with observations that warm springs result in earlier leaf senescence because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].* (David Orwig asked if this is true in both evergreen and deciduous forest.
+Combining the last two points into one sentence may better represent what Zani found.
+Remove starting from "The Latter" to here?)
+
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
+Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.* (Add a line suggeesting this may be because of water stress?)
 Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
@@
 
-In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
+Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
 In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
@@
 Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
-Thus, tree-ring records do not necessarily corroborate findings that warming spring temperatures lengthen the growing season and increase annual CO~2~ sequestration. 
-Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
+Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
+Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)
 
@@
 
-First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees from 2011-2020) and Harvard Forest (Massachusetts; n = **755** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 
-To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig.
+To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` locations (Fig.
 `r tree_rings`, `r EDF_map`).
@@
 The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
-Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and signficant hurricane damage in *YEAR*. 
-Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga spp.*), oak (*Quercus spp.*) and maple (*Acer spp.*). 
+Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and significant hurricane damage in 1938. 
+Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
 The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July (D'Orangeville et al, ).
@@
 
-Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
+Metal dendrometer bands were installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
@@
 We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L.
-at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
+at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest)
 At SCBI, cores were collected as described in @helcoski_growing_2019. 

### 2021-06-15 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
Update Dow_manuscript_Nature_cd.Rmd
mainly text related to Buermann, #61

@camerondow35 , I didn't notice that you'd made a 2nd version and edited this one. Do we now have 2 conflicted copies, or can we treat this as the new master? I find 2 versions very confusing... can we please have just one?

@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
 However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
-However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence [REFS].
+However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
 The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
-The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@zani_increased_2020].
-However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].
+The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
+However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.
 
@@
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview].
-*(We need to make sure we're not missing anything.)*
+*(We need to make sure we're not missing anything.
+See European study cited in A-T et al in review)*
 Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
@@
 However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
-This suggests that stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
+This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possibile mechanisms.
+First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
+Alternatively, warm springs may be associated with greater summer drought stress... [@buermann_widespread_2018]
 
@@
 It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].
+*(ADD MORE ON DROUGHT HERE)*
 

### 2021-06-19 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
some work on draft

@@
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.* (Add a line suggeesting this may be because of water stress?)
+Rather, our findings imply that the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.
+A larger build-up of drought stress due to earlier leaf up increasing seasonal water demand may prevent extra C from being stored in woody biomass.
+(Add a line suggeesting this may be because of water stress?)
 Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
@@
 First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
-Alternatively, warm springs may be associated with greater summer drought stress... [@buermann_widespread_2018]
+Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018]
 
@@
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
-First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
+First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].(Could it also be used to repair drought-related damage in trees?
+Like the cavitation of xylem in ring-porous species?)
 Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].

### 2021-06-19 RScripts/Pheno_tsensitivity_figure_tmax.R
remove doy gmax; increase font size; change colors; add cmi for potential precip analysis

@@ -20,7 +20,7 @@ library(rstanarm)
 library(patchwork)
 # Number of MCMC chains & number of simulations per chain.
 # Need to increase this at the end
-n_iter <- 10000
+n_iter <- 100
 n_chains <- 2
 
 # Objects to keep during memory management clean-up
@@ -182,6 +182,29 @@ climwinmeans_dp <- weatherdata %>%
 # Combine
 climwinmeans <- bind_rows(climwinmeans_rp, climwinmeans_dp)
 
+
+SCBI_CMI <- read_csv("climate data/SCBI_CMI.csv")
+#Summer precipitation totals ----
+summer_precip <- SCBI_CMI[SCBI_CMI$month %in% c(6,7,8),]
+summer_precip <- summer_precip[!(is.na(summer_precip$PRE)),]
+summer_pre <- aggregate(as.numeric(summer_precip$PRE), by = list(summer_precip$year), FUN = sum)
+names(summer_pre) <- c("year", "summer_precip")
+summer_pre <- summer_pre[c(2:11),]
+
+#Previous Autumn precipitation totals ----
+fall_precip <- SCBI_CMI[SCBI_CMI$month %in% c(9,10,11),]
+fall_precip <- fall_precip[!(is.na(fall_precip$PRE)),]
+fall_pre <- aggregate(as.numeric(fall_precip$PRE), by = list(fall_precip$year), FUN = sum)
+names(fall_pre) <- c("year", "fall_precip")
+fall_pre$year <- fall_pre$year+1
+fall_pre <- fall_pre[c(1:10),]
+
+
+climwinmeans <- left_join(climwinmeans, summer_pre, by = "year")
+climwinmeans <- left_join(climwinmeans, fall_pre, by = "year")
+
+
+
 #TMAX-April Means
 #climwinmeans <- weatherdata %>%
 #  filter(month == 4) %>%
@@ -220,7 +243,7 @@ Wood_pheno_table <- distinct(Wood_pheno_table, tag_year_perc, .keep_all = TRUE)
 ### Model Fit 1 (multivariate): (DOY_25, DOY_50, DOY_75) using climwinmeans ----
 # Convert to wide format for use in rstanarm::stan_mvmer()
 Wood_pheno_table_wide <- Wood_pheno_table %>%
-  select(perc, tag, year, wood_type, sp, climwinmean, starts_with("DOY")) %>%
[…]

### 2021-06-30 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
add comment on dendroband year drought analysis

@@
 Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018]
+Further analysis of the dendroband years in this study indicated that the warmer springs did not increase summer water stress, meaning the lack of increased growth may not be attributable to this mechanism (put SPEI plots here?).
 
@@
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
-First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].(Could it also be used to repair drought-related damage in trees?
-Like the cavitation of xylem in ring-porous species?)
+First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].*(Could it also be used to repair drought-related damage in trees?
+Like the cavitation of xylem in ring-porous species?)*
 Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].

### 2021-08-23 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
Update Dow_manuscript_Nature_cd.Rmd

@@
 
-*The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.*
-*These findings align with previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019] and with observations that warm springs result in earlier leaf senescence because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].* (David Orwig asked if this is true in both evergreen and deciduous forest.
-Combining the last two points into one sentence may better represent what Zani found.
-Remove starting from "The Latter" to here?)
+The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.
+These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
 
@@
 A larger build-up of drought stress due to earlier leaf up increasing seasonal water demand may prevent extra C from being stored in woody biomass.
-(Add a line suggeesting this may be because of water stress?)
 Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
@@
 
-***(DROP THIS FIRST PARAGRAPH?)***
-As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
-Temperate deciduous forests, including those in eastern North America, *(have been net carbon sinks over the past several decades)*... 
-*[Discuss this sink, mechanisms, uncertainties, ending with the point that one meaningful uncertainty is how the C sink will respond to warming springs]*
-
 Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
-In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
+In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
 Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
 However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
-The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
+The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
 The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
 However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.
+
+Temperate deciduous forests, including those in eastern North America, have historically been net carbon sinks.
+As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
+Whether the primary effect of global warming will be an increase in forest sink capability though earlier spring leaf up [@keenan_net_2014] or if increased respiration following warmer autumns will reduce the effectiveness of spring warming [@piao_net_2008; @barichivich_large-scale_2013] is still debated.
+In both scenarios, earlier leaf up corresponds to increased C02 sequestration while the sensitivity of photosynthesis compared to respiration in autumn, and the role seasonal water stress plays in this process is the main point of contention.
 
@@
 For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
-Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively).
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively).
 Specifically, [greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig.
@@
 Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
-while $\Delta DBH$ displayed no relationship CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
+while $\Delta DBH$ displayed no relationship with CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
 Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth rates.
@@
 `r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
-In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
-Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only #% of records: # of # ring-porous and # or # diffuse-porous species-site combinations.
-In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for #% and #% of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
+In total, our analysis included 208 species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in CTW.
+Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 or 66 diffuse-porous species-site combinations.
+In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (Aug = 57, Jul = 90, Jun = 108, May = 47) and 46% (Aug = 22, Jul = 37, Jun = 53, May = 11) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
 Again, $T_{min}$ generally exerted less influence over annual growth than $T_{max}$ (`r EDF_tree_rings_TMIN`), with few significant correlations between spring $T_{min}$ and RWI.
@@
 Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018]
-Further analysis of the dendroband years in this study indicated that the warmer springs did not increase summer water stress, meaning the lack of increased growth may not be attributable to this mechanism (put SPEI plots here?).
+
+However, spring warming at SCBI in the study years used here did not result in increased summer water stress (SPEI; Supp(?) SPEI Figure).
+This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails.
 
@@
 Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
+Linear models were run to examine the relationship between spring, summer and an interaction effect with ring-width for each of our 208 chronologies.
+The results indicated that, across all latitudes, the interacting effect of warmer springs and summers resulted in negligable increase of ring-width (supp (?) rw_drought_plot).
 It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].

### 2021-08-23 doc/manuscript/Dow_manuscript_Nature.Rmd
delete extra version

@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
 However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
 Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
-The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
-These findings align with findings that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
+
+The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.
+These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
+
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
+Rather, our findings imply that the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.
+A larger build-up of drought stress due to earlier leaf up increasing seasonal water demand may prevent extra C from being stored in woody biomass.
 Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
@@
 
-***(DROP THIS FIRST PARAGRAPH?)***
+Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
+In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
+Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
+However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
+The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
+The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
+However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.
+
+Temperate deciduous forests, including those in eastern North America, have historically been net carbon sinks.
 As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
-Temperate deciduous forests, including those in eastern North America, *(have been net carbon sinks over the past several decades)*... 
-*[Discuss this sink, mechanisms, uncertainties, ending with the point that one meaningful uncertainty is how the C sink will respond to warming springs]*
-
-In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
-In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
-Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
-However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence [REFS].
-The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
-The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@zani_increased_2020].
-However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].
+Whether the primary effect of global warming will be an increase in forest sink capability though earlier spring leaf up [@keenan_net_2014] or if increased respiration following warmer autumns will reduce the effectiveness of spring warming [@piao_net_2008; @barichivich_large-scale_2013] is still debated.
+In both scenarios, earlier leaf up corresponds to increased C02 sequestration while the sensitivity of photosynthesis compared to respiration in autumn, and the role seasonal water stress plays in this process is the main point of contention.
 
@@
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview].
-*(We need to make sure we're not missing anything.)*
+*(We need to make sure we're not missing anything.
+See European study cited in A-T et al in review)*
 Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
-Thus, tree-ring records do not necessarily corroborate findings that warming spring temperatures lengthen the growing season and increase annual CO~2~ sequestration. 
-Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
+Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
+Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)
 
@@
 However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
-This suggests that stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
+This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possibile mechanisms.
+First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
+Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018]
+
+However, spring warming at SCBI in the study years used here did not result in increased summer water stress (SPEI; Supp(?) SPEI Figure).
+This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails.
 
@@
 Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
+Linear models were run to examine the relationship between spring, summer and an interaction effect with ring-width for each of our 208 chronologies.
+The results indicated that, across all latitudes, the interacting effect of warmer springs and summers resulted in negligable increase of ring-width (supp (?) rw_drought_plot).
 It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].
+*(ADD MORE ON DROUGHT HERE)*
 

### 2021-08-23 doc/manuscript/Dow_Extended_Data.Rmd
remove Fig. 3 from main manuscript
(as planned in our meeting)

Co-Authored-By: Cameron Dow <55721987+camerondow35@users.noreply.github.com>

#64

@@
 
-pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Relationship between growth parameters and mean maximum temperature within the critical temperature window (CTW).** CTW was defined using *climwin*.
-For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
+pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW).
+For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands.
+(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
 CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

### 2021-08-23 doc/manuscript/Dow_manuscript_Nature.Rmd
remove Fig. 3 from main manuscript
(as planned in our meeting)

Co-Authored-By: Cameron Dow <55721987+camerondow35@users.noreply.github.com>

#64

@@
   DOY_timing_Harvard = "2b"  #currently 4 panel!!!!
-pheno_Tsensitivity = 3 
-tree_rings = 4
+tree_rings = 3
 
@@
 
-DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW).
-For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands.
-Relationships for additional growth parameters (Fig.
-1a) are given in Extended Data Figure 4.
-(**DOYrmax will be dropped.** [GitHub Issue 23](https://github.com/EcoClimLab/growth_phenology/issues/23))" #needs attention!!!
-
-tree_rings_caption <- "**Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
+tree_rings_caption <- "**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature.
@@
 
-Both leaf phenology and stem growth were sensitive to spring temperatures (Figs. `r DOY_timing` -`r pheno_Tsensitivity`). 
+Both leaf phenology and stem growth were sensitive to spring temperatures (Figs. `r DOY_timing`, `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`). 
 While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-7), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
@@
 
-![`r DOY_Tsensitivity_caption`](tables_figures/DOY_Tsensitivity.png){ width=50% }
-
-For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
+For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively).
@@
 Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing [1-8 days per degree Celsius](https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs.
-`r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).
+`r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
 
@@
 Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig.
-`r schematic_summary`, Extended Data Figures **4-6**).
+`r schematic_summary`, Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
 
 In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs.
-`r schematic_summary`, `r pheno_Tsensitivity`).
+`r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
 Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig.
@@
 `r DOY_timing`)[@jeong_phenology_2011], an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs.
-`r schematic_summary`- `r pheno_Tsensitivity`).
+`r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
 Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs.

### 2021-08-23 doc/manuscript/Dow_manuscript_Nature.Rmd
edit summary paragraph
(still a bit too long)

@@
 Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
+The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.
+These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.
+Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
 
-The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.
+*removed from abstract; to integrate elsewhere:*
+
 These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
-
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-Rather, our findings imply that the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.
-A larger build-up of drought stress due to earlier leaf up increasing seasonal water demand may prevent extra C from being stored in woody biomass.
-Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
 

### 2021-08-23 doc/manuscript/Dow_manuscript_Nature.Rmd
cut summary paragraph down to the right length

@@
 
-(*word limit 200 - this is currently way too long.*)
-
 As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
-However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
-Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
-The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.
-These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.
+However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
+Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth milestones of deciduous trees earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, or annual growth.
+The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
+These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
 Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
@@
 *removed from abstract; to integrate elsewhere:*
+
+tree stem growth consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021]
 

### 2021-08-23 RScripts/DOYtiming_all_years.R
Remove empty space between rows

@@ -2,6 +2,7 @@ library(ggplot2)
 library(readr)
 library(tidyverse)
 library(lubridate)
+library(patchwork)
 Wood_pheno_table_scbi <- read_csv("Data/Wood_pheno_table_SCBI_CLEAN.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%
@@ -484,7 +485,8 @@ p1 <- ggplot(leaf_phenology_melt, aes(x=value, y = as.character(variable), group
         #axis.text.y = element_text(angle = 90, hjust = 0.5),
         axis.text.x = element_blank(),
         text = element_text(size = 20),
-        plot.title = element_text(hjust = 0.5))+
+        plot.title = element_text(hjust = 0.5),
+        plot.margin = unit(c(-1,-1,-1,-1), "cm"))+
   labs(x = "", y = "Leaf Stage", title = "SCBI", color = "Temp Ratio (1= Warmest pre-season)", linetype = "")+
   scale_colour_gradient(low = "blue", high = "red")
 
@@ -498,8 +500,9 @@ p2 <- ggplot(hf_leaf_phenology_melt, aes(x=value, y = as.character(variable), gr
         axis.text.y = element_blank(),
         axis.text.x = element_blank(),
         text = element_text(size = 20),
-        plot.title = element_text(hjust = 0.5))+
-  labs(x = "", y = "", title = "(b) Harvard Forest", color = "Temp Ratio (1= Warmest pre-season)", linetype = "")+
+        plot.title = element_text(hjust = 0.5),
+        plot.margin = unit(c(-1,-1,-1,-1), "cm"))+
+  labs(x = "", y = "", title = "Harvard Forest", color = "Temp Ratio (1= Warmest pre-season)", linetype = "")+
   scale_colour_gradient(low = "blue", high = "red")
 
 p3 <- ggplot(aggregates_scbi, aes(x=x, y = Group.2, group = interaction(Group.1, year), color = stanT, linetype = Group.1))+
@@ -510,12 +513,13 @@ p3 <- ggplot(aggregates_scbi, aes(x=x, y = Group.2, group = interaction(Group.1,
   theme(legend.position = c(.85,.5),
         text = element_text(size = 20),
         legend.text=element_text(size=10),
[…]

### 2021-08-23 RScripts/additional_figures.R
first pass at new fig 1

@@ -9,211 +9,6 @@ library(lubridate)
 
 
 
-# Figure 1: Logistic growth curve and parameter illustration -------------------
-# # Set parameters from McMahon & Parker (2014)
-# # 1. L: lower asymptotes
-# # 2. K: upper asymptotes
-# # 3. doy.ip: inflection point
-# # 4. r: value that determines slope of curve at inflection point ((K-L)*r/theta )/(1 + 1/theta)^2
-# # 5. theta: theta = 1 indicates pre/post inflection post symmetry
-# L <- 13
-# K <- 14
-# doy.ip <- 182
-# r <- 0.025
-# theta <- 1.5
-# params <- c(L, K, doy.ip, r, theta)
-# total_growth <- K - L
-#
-# start_doy <- 1
-# end_doy <- 365
-#
-# window_open <- 60
-# window_close <- 105
-#
-#
-# # Function for logistic growth model written by Sean. Supposed to be in RDendrom
-# # package https://github.com/seanmcm/RDendrom, but function does not seem to be included
-# lg5.pred <- function(params, doy) {
-#   L <- params[1] # min(dbh, na.rm = T)
-#   K <- params[2]
-#   doy.ip <- params[3]
-#   r <- params[4]
-#   theta <- params[5]
-#   dbh <- vector(length = length(doy))
-#   dbh <- L + ((K - L) / (1 + 1 / theta * exp(-(r * (doy - doy.ip) / theta))^theta))
-#   return(dbh)
-# }
-#
-# # 1. True growth (logistic curve):
-# true_values <- tibble(
-#   doy = seq(from = start_doy, to = end_doy, by = 1),
-#   diameter = lg5.pred(params, doy)
-# ) %>%
-#   mutate(diff = diameter - lag(diameter))
-#
-#
-# # 2. Max growth rate/inflection point (tangent line):
-# # Incorrect intercept since r is not true slope at inflection point:
-# # intercept <- lg5.pred(params, doy.ip) - r *doy.ip
-#
-# # Version 1 of correct slope and intercept
-# r_true <- ((K-L)*r/theta )/(1 + 1/theta)^2
-# intercept_true <- lg5.pred(params, doy.ip) - r_true *doy.ip
-#
-# # Version 2 of correct doy/ip, slope, and intercept
-# doy.ip_true <- -log(theta)/r + doy.ip
[…]

### 2021-08-23 RScripts/additional_figures.R
Done latest iteration

@@ -34,24 +34,6 @@ lg5.pred <- function(params, doy) {
 
 
 ## 1. Get LG5 parameter values ----
-# # Set SCBI values
-# # Identify cold and hot aprils
-# read_csv("climate data/NCDC_NOAA_precip_temp.csv") %>%
-#   mutate(month = month(DATE)) %>%
-#   dplyr::filter(month == 4) %>%
-#   group_by(year) %>%
-#   summarize(TMAX = mean(TMAX, na.rm = TRUE)) %>%
-#   arrange(desc(TMAX))
-#
-# # Real red/hot curve values:
-# SCBI_hot_LG5_values <- read_csv("Data/LG5_parameter_values_SCBI_CLEAN.csv") %>%
-#   dplyr::filter(year == 2019) %>%
-#   group_by(tag_year) %>%
-#   summarize(L = mean(L), K = mean(K), doy_ip = mean(doy_ip), r = mean(r), theta = mean(theta)) %>%
-#   summarize(L = mean(L), K = mean(K), doy_ip = mean(doy_ip), r = mean(r), theta = mean(theta))
-
-
-### Get HF values ----
 # Identify cold and hot aprils
 read_csv("climate data/HF_weatherdata.csv") %>%
   dplyr::filter(month == 4) %>%
@@ -135,11 +117,6 @@ HF_doy_diameter_quartile_hot <- bind_rows(
   ) %>%
   select(-perc) %>%
   rename(perc = growth)
-  # mutate(perc = case_when(
-  #   growth == 0 ~ 0,
-  #   growth == 1 ~ 1,
-  #   TRUE ~ perc
-  # ))
 
 HF_doy_diameter_quartile_cold <- bind_rows(
   tibble(doy = 1, diameter = HF_L_cold),
@@ -156,11 +133,6 @@ HF_doy_diameter_quartile_cold <- bind_rows(
   ) %>%
   select(-perc) %>%
   rename(perc = growth)
-  # mutate(perc = case_when(
-  #   growth == 0 ~ 0,
-  #   growth == 1 ~ 1,
-  #   TRUE ~ perc
-  # ))
 
 HF_doy_diameter_quartile <- bind_rows(
   HF_doy_diameter_quartile_hot %>% mutate(year = "Hot"),
@@ -197,12 +169,7 @@ HF_significant_perc <-
       select(doy_cold = doy, perc),
     by = "perc"
   ) %>%
-  # bind_cols(
[…]

### 2021-09-08 doc/manuscript/Dow_manuscript_Nature.Rmd
some rewording

@@
 As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
-Whether the primary effect of global warming will be an increase in forest sink capability though earlier spring leaf up [@keenan_net_2014] or if increased respiration following warmer autumns will reduce the effectiveness of spring warming [@piao_net_2008; @barichivich_large-scale_2013] is still debated.
-In both scenarios, earlier leaf up corresponds to increased C02 sequestration while the sensitivity of photosynthesis compared to respiration in autumn, and the role seasonal water stress plays in this process is the main point of contention.
+Whether forest sink capability will increase as spring leaf up advances[@keenan_net_2014] or if sink capacity is limited by some mechanism is unknown.
+In the former case, earlier leaf up appears to correspond with increased NEE, but little work has been done to understand where the extra sequestered carbon is being stored[].
 
@@
 However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
-This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possibile mechanisms.
+This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possible mechanisms.
 First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
-Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018]
+Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018; ]
 
@@
 This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails.
+(Is SPEI a good measure of soil moisture?)
 
@@
 Linear models were run to examine the relationship between spring, summer and an interaction effect with ring-width for each of our 208 chronologies.
-The results indicated that, across all latitudes, the interacting effect of warmer springs and summers resulted in negligable increase of ring-width (supp (?) rw_drought_plot).
+The results indicated that, across all latitudes, the interacting effect of warmer springs and summers resulted in negligible increase of ring-width (supp (?) rw_drought_plot).
 It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].

### 2021-09-08 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
incorporated muller_water_2011 and zohner_how_2021

@@
 Whether forest sink capability will increase as spring leaf up advances[@keenan_net_2014] or if sink capacity is limited by some mechanism is unknown.
-In the former case, earlier leaf up appears to correspond with increased NEE, but little work has been done to understand where the extra sequestered carbon is being stored[].
+(sentence about where carbon goes in the case of the former?)
 
@@
 Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
+Stem growth is more sensitive to water deficit than photosynthesis is[@muller_water_2011].
+This seems to promote the investment of sequestered carbon into root growth in times of drought stress.
+Zoehner et al 2021[@zohner_how_2021], found that the proportion of root-to-stem growth increased under spring warming, suggesting that increased ecosystem sequestration resulting from spring warming may be promoting the growth of below-ground biomass.
 Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.

### 2021-09-13 doc/manuscript/Dow_manuscript_Nature.Rmd
refocus on sink-limitation

@@
 Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-(**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?
-(KAT thinks so)** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50).
-)" #needs attention!!!
+#needs attention!!!"
 
@@
 (NOTE: Figure still needs some work.
-**COLOR SCALE IS  A BIT OFF IN THIS COMPOSITE FIGURE (scales not identical across panels).**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
+See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
 #info subject to change!!!
@@
 However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
-Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth milestones of deciduous trees earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, or annual growth.
+Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
 The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
@@
 In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
-Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
-However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
-The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
+Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
+However, warmer summer and autumn temperatures and drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018].
+As a result, warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
+The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;;@ibanez_forecasting_2010;;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
 The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
-However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.
+However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors bcome limiting [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.
 
 Temperate deciduous forests, including those in eastern North America, have historically been net carbon sinks.
+Sink limitation means that resources or temperature effects on the sink (the trees) will become limiting before the source (CO2) is limiting.
+The implication is that the benefits of CO2 fertilization will eventually reach an asymptote as other resources necessary for plant growth run out.
+Currently, our understanding of when and where these limits will be reached is incomplete, so forest models tend to assume the increase in photosynthesis as atmospheric CO2 levels increase is unlimited.(Maybe not true, i'm not 100% familiar with the modeling still.)
+
 As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
-Whether forest sink capability will increase as spring leaf up advances[@keenan_net_2014] or if sink capacity is limited by some mechanism is unknown.
-(sentence about where carbon goes in the case of the former?)
-
-We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures. 
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview].
+Understanding how trees are responding to warming temperatures and rising CO2 levels, and how growth may become limited because of interactions between the two is essential to improve our predictions.
+We know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few site [@doranngeville_radial_inreview; @martin_hydrometeorology_2017].
 *(We need to make sure we're not missing anything.
 See European study cited in A-T et al in review)*
-Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
+Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019;@] [MORE]).
 Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
@@
 For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively).
-Specifically, [greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig.
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the CTW (p=0.001 and 0.139, respectively).
+Specifically, [greenup DOY occurred 21 days earlier (-4.535 days/$^\circ$C) at SCBI and 7 days earlier (-2.437 days/ $^\circ$C) at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
-Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing [1-8 days per degree Celsius](https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs.
+Similarly, $DOY_{25}$ (SCBI: RP = -1.91 days/ $^\circ$C, DP = -3.47 days/$^\circ$C , HF: RP = -2.79 days/$^\circ$C, DP = -7.90 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.54 days/ $^\circ$C, DP = -3.52 days/$^\circ$C, HF: RP =  -4.98 days/$^\circ$C, DP = -7.27 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.128 days/$^\circ$C, DP = -3.56 days/$^\circ$C , HF: RP = -7.17 days/$^\circ$C, DP= -6.62 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with CTW $T_{max}$ (https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs.
 `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
 
-Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer springs (Fig.
+Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer CTW's compared to those with cooler CTW's (Fig.
 `r DOY_timing`), there was no consistent lengthening of the period of peak stem growth.
@@
 while $\Delta DBH$ displayed no relationship with CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
-Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth rates.
+Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth.
 
@@
 Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
-responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig.
+responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral response to $T_{max}$ (Fig.
 `r tree_rings`).
@@
 `r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
-In total, our analysis included 208 species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in CTW.
-Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 or 66 diffuse-porous species-site combinations.
+In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in CTW.
+Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.
 In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (Aug = 57, Jul = 90, Jun = 108, May = 47) and 46% (Aug = 22, Jul = 37, Jun = 53, May = 11) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
@@
 This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possible mechanisms.
-First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
-Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018; ]
+First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) would lead to an earlier depletion of resources needed for growth resulting in an earlier end to the season of rapid growth ($DOY_{75}$).
+Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018; ] or inducing the cessation of stem growth when a critical water deficit is reached[@lempereur_growth_2015].
 
-However, spring warming at SCBI in the study years used here did not result in increased summer water stress (SPEI; Supp(?) SPEI Figure).
-This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails.
-(Is SPEI a good measure of soil moisture?)
+However, spring warming at SCBI in the study years used here did not result in increased deltaDBH or summer water stress (SPEI; Supp(?) SPEI Figure).
+This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at this site.
+This could be because the woody growth at this site is already sink-limited.
 
@@
 The results indicated that, across all latitudes, the interacting effect of warmer springs and summers resulted in negligible increase of ring-width (supp (?) rw_drought_plot).
-It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].
+*It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019; @zhang_drought_2021].
+(Is this relevent?)*(Neil suggested that the zhang paper would be good here but i'm not sure I see how...)
 *(ADD MORE ON DROUGHT HERE)*
@@
 Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
-Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
+Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one *(This study only considered a time frame of a year or less?)*, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
 Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
@@
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
-In addition to the existence of temporal lags between C uptake and release, which introduce a lot of uncertainty[@trumbore_quantifying_2002], trees are likely to  acclimate *to some extent* to warming temperatures[**REFS?**], and on even longer time scales, we are likely to see changes in forest composition and adaptation.
+In addition to the existence of temporal lags between C uptake and release, which introduce a lot of uncertainty[@trumbore_quantifying_2002], trees are likely to  acclimate *to some extent* to warming temperatures[@zhang_drought_2021; @gessler_way_2020], and on even longer time scales, we are likely to see changes in forest composition and adaptation.
 If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
@@
 Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs.
-`r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@dorangeville_drought_2018;@helcoski_growing_2019].
+`r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@zhang_drought_2021;@dorangeville_drought_2018;@helcoski_growing_2019].
 
@@
 The forest here was established in the mid-19th century after conversion from agricultural fields. 
-Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*). 
+Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), and hickories (*Carya spp.*). 
 The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July ([helcoski_growing_2019]).
@@
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
-Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
+Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
 The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
@@
 Measurements were timed to begin before the beginning of spring growth and to continue through the cessation of growth in the fall. 
-At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, with the earliest start date being March 30, 2020.
-Measurements were continued through to fall leaf senescence, with the median end date being 10/17 and the latest end date was November 26, 2012.
-Timing of measurements at Harvard Forest was similar, with the median start date of 4/23 and median end date of 10/30.
+At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory vegetation was observed, with the earliest start date being March 30, 2020.
+Measurements were continued through to fall leaf senescence, with the median end date being 10/17 and the latest end date November 26, 2012.
+Timing of measurements at Harvard Forest were similar, with the median start date of 4/23 and median end date of 10/30.
 1998 was an anomalous year where initial measurements were taken on 1/5, but not taken again until 4/15.
@@
 
-The dendrometer band data were manually cleaned by visual inspection before analysis. 
+The raw dendrometer band data were manually cleaned by visual inspection before analysis. 
 We screened for three classes of mistakes. 
 First, when a weekly measurement was drastically different from previous week and following weeks, this was assumed to be a measurement error and the record was removed.
-Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump followed by normal growth pattern, this was assumed to be a case where the band was stuck and then released.
+Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck and then released.
 In these cases, the full annual record for the tree was removed.
-Third, there were cases where data were clearly wrong but with unknown causes.
-In cases where several measurements were an issue, they were simply removed.
-If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.
+Third, cases where data were clearly wrong but with unknown causes were removed. If a majority of the data points fell into this class the entire year was removed from the analysis.
 
@@
 
-After fitting the @mcmahon_general_2015 model, we used a script to remove poorly fit models. 
+After fitting the @mcmahon_general_2015 model, we removed poorly fit models. 
 Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the acceptable range we determined. 
@@
 and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, HF = 3). 
-At Harvard Forest the tag years removed through this method was proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others *(SI figure?
+At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others *(SI figure?
 data_cleaning_figure_hf in manuscript folder)*.
@@
 At SCBI, cores were collected as described in @helcoski_growing_2019. 
-Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
+Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019]*(If we're citing a method from the helcoski paper, do we need to do this 'briefly' section?
+I would remove.)*.
 At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
@@
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@dorangeville_drought_2018] (**OTHER REFS??**).
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018] (**OTHER REFS??**).
 Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).

### 2021-09-15 doc/manuscript/Dow_manuscript_Nature.Rmd
incorporate zohner_2021

@@
 As a result, warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
-The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;;@ibanez_forecasting_2010;;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
+
+Spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;;@ibanez_forecasting_2010;;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
 The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
-However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors bcome limiting [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.
+
+However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors become limiting [@zani_increased_2020; @zohner_how_2021] suggests that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
 
@@
 Understanding how trees are responding to warming temperatures and rising CO2 levels, and how growth may become limited because of interactions between the two is essential to improve our predictions.
+Zohner et al, (2020)[@zohner_how_2021] found that under controled conditions, spring warming increases biomass accumulation in trees, with a larger proportion of root compared to shoot growth.
+The authors also found evidence of sink limitation in autumn, even with controlled water and fertilizer application to the study trees.
+To our knowledge, the link between phenology and growth has never been studied in-situ.
+
 We know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
@@
 
-However, spring warming at SCBI in the study years used here did not result in increased deltaDBH or summer water stress (SPEI; Supp(?) SPEI Figure).
-This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at this site.
-This could be because the woody growth at this site is already sink-limited.
+Spring warming at SCBI and HF in the study years used here did not result in increased $\Delta DBH$ or summer water stress (SPEI; Supp(?) SPEI Figure).
+This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at these sites.
+This could be because the woody growth at these sites are already sink-limited.
 
@@
 We show that the extra C does not go into woody growth, raising the question of where this C goes.
+
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
@@
 Zoehner et al 2021[@zohner_how_2021], found that the proportion of root-to-stem growth increased under spring warming, suggesting that increased ecosystem sequestration resulting from spring warming may be promoting the growth of below-ground biomass.
+However, this study was done under tightly controlled conditions within potted plants and does not represent the environmental conditions of forests like our study does.
 Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.

### 2021-09-17 doc/manuscript/archive/Dow_manuscript_Nature_cd.Rmd
move _cd version to archive
(avoid confusion)

--- 2021-08-23 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
@@
 As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
-However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
+However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
 Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
-
 The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.
 These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
-
 They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.

### 2021-09-17 doc/manuscript/Dow_Extended_Data.Rmd
knit in map

@@
 ## figure captions
+
+map_caption <- "**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
+Sampling details are provided in **SI TBALE NAME**" #needs attention!!!
+
 growth_curves_all_caption  <- " Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
@@
 
-## (PLACEHOLDER FOR MAP OF STUDY SITES -- [GITHUB ISSUE #54](https://github.com/EcoClimLab/growth_phenology/issues/54))
+![`r map_caption`](tables_figures/chronology_map.png)
 

### 2021-09-17 doc/manuscript/Dow_Extended_Data.Rmd
minor

@@
 
-growth_curves_all_caption  <- " Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.
-Each line represents one year's growth for a given tree, fit with McMahon model.
+growth_curves_all_caption  <- " **Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.** Each line represents one year's growth for a given tree, fit with McMahon model.
 " #needs wordsmithing!!!
 
-climwin_caption <- "Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.
-Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
+climwin_caption <- "**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.** Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
 Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.

### 2021-09-17 doc/manuscript/Dow_Extended_Data.Rmd
cleanup

@@
 species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
-sample_size_by_year_caption <- "TO BE REMOVED | Sample size by year "
-mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "
+mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest."
 tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring data and analysis ( including species, n chronologies (= n sites), n positively correlated with April T, n sig.)"

### 2021-09-17 doc/manuscript/Dow_manuscript_Nature.Rmd
brought table back into schematic summary figure
remove CTW abbreviation

@@
 
-schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window (CTW)', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
+schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
 
@@
 
-![`r schematic_summary_caption`](tables_figures/schematic_summary.png){ width=70% }
+![`r schematic_summary_caption`](tables_figures/schematic_summary_with_table.png){ width=70% }
 
@@
 While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-7), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
-1 and $DOY_{25}$  for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window, CTW; Fig.
+1 and $DOY_{25}$  for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig.
 `r schematic_summary`).
-The strongest relationships, including but not limited to the CTW (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).
+The strongest relationships, including but not limited to the critical temperature window (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).
 
-For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the CTW (p=0.001 and 0.139, respectively).
-Specifically, [greenup DOY occurred 21 days earlier (-4.535 days/$^\circ$C) at SCBI and 7 days earlier (-2.437 days/ $^\circ$C) at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig.
+For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs.
+`r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the critical temperature window (p=0.001 and 0.139, respectively).
+Specifically, [greenup DOY occurred 21 days earlier (-4.535 days/$^\circ$C) at SCBI and 7 days earlier (-2.437 days/ $^\circ$C) at Harvard Forest when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
-Similarly, $DOY_{25}$ (SCBI: RP = -1.91 days/ $^\circ$C, DP = -3.47 days/$^\circ$C , HF: RP = -2.79 days/$^\circ$C, DP = -7.90 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.54 days/ $^\circ$C, DP = -3.52 days/$^\circ$C, HF: RP =  -4.98 days/$^\circ$C, DP = -7.27 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.128 days/$^\circ$C, DP = -3.56 days/$^\circ$C , HF: RP = -7.17 days/$^\circ$C, DP= -6.62 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with CTW $T_{max}$ (https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs.
+Similarly, $DOY_{25}$ (SCBI: RP = -1.91 days/ $^\circ$C, DP = -3.47 days/$^\circ$C , HF: RP = -2.79 days/$^\circ$C, DP = -7.90 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.54 days/ $^\circ$C, DP = -3.52 days/$^\circ$C, HF: RP =  -4.98 days/$^\circ$C, DP = -7.27 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.128 days/$^\circ$C, DP = -3.56 days/$^\circ$C , HF: RP = -7.17 days/$^\circ$C, DP= -6.62 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with mean $T_{max}$ during the critical temperature window (https://github.com/EcoClimLab/growth_phenology/issues/58) during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
 
-Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer CTW's compared to those with cooler CTW's (Fig.
+Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig.
 `r DOY_timing`), there was no consistent lengthening of the period of peak stem growth.
@@
 `r schematic_summary`, Extended Data Figures **4-6**).
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
-while $\Delta DBH$ displayed no relationship with CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
+while $\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`). 
 Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth.
@@
 `r tree_rings`, `r EDF_map`).
-For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, ring width indices (RWI) rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig.
+For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, ring width indices (RWI) rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Fig.
 `r tree_rings`, `r EDF_tree_rings_TMIN`).
@@
 `r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
-In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in CTW.
+In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in ~CTW~ April $T_{max}$ (**Cam, please confirm this change**).
 Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.

### 2021-09-17 doc/manuscript/Dow_manuscript_Nature.Rmd
done with issue #58

@@
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the critical temperature window (p=0.001 and 0.139, respectively).
-Specifically, [greenup DOY occurred 21 days earlier (-4.535 days/$^\circ$C) at SCBI and 7 days earlier (-2.437 days/ $^\circ$C) at Harvard Forest when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig.
+Specifically, greenup DOY advanced 4.535 days/$^\circ$C at SCBI and 2.437 days/ $^\circ$C at Harvard Forest when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window, and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
-Similarly, $DOY_{25}$ (SCBI: RP = -1.91 days/ $^\circ$C, DP = -3.47 days/$^\circ$C , HF: RP = -2.79 days/$^\circ$C, DP = -7.90 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.54 days/ $^\circ$C, DP = -3.52 days/$^\circ$C, HF: RP =  -4.98 days/$^\circ$C, DP = -7.27 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.128 days/$^\circ$C, DP = -3.56 days/$^\circ$C , HF: RP = -7.17 days/$^\circ$C, DP= -6.62 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with mean $T_{max}$ during the critical temperature window (https://github.com/EcoClimLab/growth_phenology/issues/58) during the critical temperature window (Figs.
+Similarly, $DOY_{25}$ (SCBI: RP = -1.91 days/ $^\circ$C, DP = -3.47 days/$^\circ$C , HF: RP = -2.79 days/$^\circ$C, DP = -7.90 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.54 days/ $^\circ$C, DP = -3.52 days/$^\circ$C, HF: RP =  -4.98 days/$^\circ$C, DP = -7.27 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.128 days/$^\circ$C, DP = -3.56 days/$^\circ$C , HF: RP = -7.17 days/$^\circ$C, DP= -6.62 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).

### 2021-09-17 doc/manuscript/Dow_manuscript_Nature.Rmd
reduce n decimal points reported
@camerondow35 , FYI, one digit after the decimal point is sufficient here

@@
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the critical temperature window (p=0.001 and 0.139, respectively).
-Specifically, greenup DOY advanced 4.535 days/$^\circ$C at SCBI and 2.437 days/ $^\circ$C at Harvard Forest when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window, and mid-greenup also advanced, but to a lesser extent (Fig.
+Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window, and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
-Similarly, $DOY_{25}$ (SCBI: RP = -1.91 days/ $^\circ$C, DP = -3.47 days/$^\circ$C , HF: RP = -2.79 days/$^\circ$C, DP = -7.90 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.54 days/ $^\circ$C, DP = -3.52 days/$^\circ$C, HF: RP =  -4.98 days/$^\circ$C, DP = -7.27 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.128 days/$^\circ$C, DP = -3.56 days/$^\circ$C , HF: RP = -7.17 days/$^\circ$C, DP= -6.62 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
+Similarly, $DOY_{25}$ (SCBI: RP = -1.9 days/ $^\circ$C, DP = -3.5 days/$^\circ$C , HF: RP = -2.8 days/$^\circ$C, DP = -7.9 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.5 days/ $^\circ$C, DP = -3.5 days/$^\circ$C, HF: RP =  -5.0 days/$^\circ$C, DP = -7.3 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.1 days/$^\circ$C, DP = -3.6 days/$^\circ$C , HF: RP = -7.2 days/$^\circ$C, DP= -6.6 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).

### 2021-09-17 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@camerondow35 , removed mid-greenup here because p was not sig.

Also removed DOY_g.max.

@@
 `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the critical temperature window (p=0.001 and 0.139, respectively).
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOY of leaf greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the critical temperature window (p=0.001).
 Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window, and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
-Similarly, $DOY_{25}$ (SCBI: RP = -1.9 days/ $^\circ$C, DP = -3.5 days/$^\circ$C , HF: RP = -2.8 days/$^\circ$C, DP = -7.9 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.5 days/ $^\circ$C, DP = -3.5 days/$^\circ$C, HF: RP =  -5.0 days/$^\circ$C, DP = -7.3 days/$^\circ$C), $DOY_{75}$ (SCBI: RP = -1.1 days/$^\circ$C, DP = -3.6 days/$^\circ$C , HF: RP = -7.2 days/$^\circ$C, DP= -6.6 days/$^\circ$C), and $DOY_{g.max}$ at both sites all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
+Similarly, $DOY_{25}$ (SCBI: RP = -1.9 days/ $^\circ$C, DP = -3.5 days/$^\circ$C , HF: RP = -2.8 days/$^\circ$C, DP = -7.9 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.5 days/ $^\circ$C, DP = -3.5 days/$^\circ$C, HF: RP =  -5.0 days/$^\circ$C, DP = -7.3 days/$^\circ$C), and $DOY_{75}$ (SCBI: RP = -1.1 days/$^\circ$C, DP = -3.6 days/$^\circ$C , HF: RP = -7.2 days/$^\circ$C, DP= -6.6 days/$^\circ$C) at both sites all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).

### 2021-09-17 doc/manuscript/Dow_manuscript_Nature.Rmd
reword results content so it's easier to read

@@
 `r DOY_timing`).
-Similarly, $DOY_{25}$ (SCBI: RP = -1.9 days/ $^\circ$C, DP = -3.5 days/$^\circ$C , HF: RP = -2.8 days/$^\circ$C, DP = -7.9 days/$^\circ$C ), $DOY_{50}$ (SCBI: RP = -1.5 days/ $^\circ$C, DP = -3.5 days/$^\circ$C, HF: RP =  -5.0 days/$^\circ$C, DP = -7.3 days/$^\circ$C), and $DOY_{75}$ (SCBI: RP = -1.1 days/$^\circ$C, DP = -3.6 days/$^\circ$C , HF: RP = -7.2 days/$^\circ$C, DP= -6.6 days/$^\circ$C) at both sites all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
+Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
+Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
+At HF, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
 
@@
 `r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
-In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in ~CTW~ April $T_{max}$ (**Cam, please confirm this change**).
+In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in ~~CTW~~ April $T_{max}$ (**Cam, please confirm this change**).
 Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.

### 2021-09-17 doc/manuscript/Dow_manuscript_Nature.Rmd
discussioon
adjust discussion according to issue #57

raise issue #72

@@
 First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
-Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
-*(we can easily test this: [GitHub issue 57](https://github.com/EcoClimLab/growth_phenology/issues/57))*
-Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
-Stem growth is more sensitive to water deficit than photosynthesis is[@muller_water_2011].
-This seems to promote the investment of sequestered carbon into root growth in times of drought stress.
-Zoehner et al 2021[@zohner_how_2021], found that the proportion of root-to-stem growth increased under spring warming, suggesting that increased ecosystem sequestration resulting from spring warming may be promoting the growth of below-ground biomass.
-However, this study was done under tightly controlled conditions within potted plants and does not represent the environmental conditions of forests like our study does.
+We cannot rule out the possibility that more carbon-dense cell walls form in years with warmer springs. (**coauthors, do any of you know literature related to this?**)
+However, [extension of our tree-ring analysis](https://github.com/EcoClimLab/growth_phenology/issues/57) showed very little correlation between April $T_{max}$ and growth the following year (**#** of 142 RP and **#** of 66 DP species-site combinations), refuting the latter possibility and suggesting that, barring formation of much denser wood in years with warm springs, little of the extra C that may be fixed in these years goes into wood formation.
+
+A second possibility is that any additional C fixed during years with warm springs that is not rapidly respired may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
+Stem growth is more sensitive to water deficit than is photosynthesis[@muller_water_2011].
+[This seems to promote the investment of sequestered carbon into root growth in times of drought stress.
+Zoehner et al 2021[@zohner_how_2021] found that the proportion of root-to-stem growth increased under spring warming, suggesting that increased ecosystem sequestration resulting from spring warming may be promoting the growth of below-ground biomass.
+~~However, this study was done under tightly controlled conditions within potted plants and does not represent the environmental conditions of forests.~~](https://github.com/EcoClimLab/growth_phenology/issues/72)
 Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.

### 2021-09-17 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion

@@
 First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
-We cannot rule out the possibility that more carbon-dense cell walls form in years with warmer springs. (**coauthors, do any of you know literature related to this?**)
-However, [extension of our tree-ring analysis](https://github.com/EcoClimLab/growth_phenology/issues/57) showed very little correlation between April $T_{max}$ and growth the following year (**#** of 142 RP and **#** of 66 DP species-site combinations), refuting the latter possibility and suggesting that, barring formation of much denser wood in years with warm springs, little of the extra C that may be fixed in these years goes into wood formation.
+However, [extension of our tree-ring analysis](https://github.com/EcoClimLab/growth_phenology/issues/57) showed very little correlation between April $T_{max}$ and growth the following year (**#** of 142 RP and **#** of 66 DP species-site combinations), refuting the latter possibility.
+We cannot rule out the possibility that warm springs promote formation of more carbon-dense cell walls in the current of following year (**coauthors, do any of you know literature related to this?**); however, this possibility, our results suggest that little of the extra C that may be fixed in these years goes into wood formation.
 
@@
 Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
-Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one *(This study only considered a time frame of a year or less?)*, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
-Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
+Notably, turnover of extra C fixed during years with warm springs would often occur with sufficient temporal lag that it would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
+Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a *fairly long* residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
 
@@
 (*see [GitHub issue 56 on representation in models](https://github.com/EcoClimLab/growth_phenology/issues/56)*)
-m
 

### 2021-09-28 doc/manuscript/Dow_manuscript_Nature.Rmd
intro work

@@
 Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
-However, warmer summer and autumn temperatures and drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018].
-As a result, warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
+Additionally, Warmer summer and autumn temperatures coupled with drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018].
+As a result, warming may advance or delay autumn senescence depending on timing of warming and water availability, with delays more common across eastern North America [@buermann_widespread_2018].
 
@@
 Temperate deciduous forests, including those in eastern North America, have historically been net carbon sinks.
-Sink limitation means that resources or temperature effects on the sink (the trees) will become limiting before the source (CO2) is limiting.
-The implication is that the benefits of CO2 fertilization will eventually reach an asymptote as other resources necessary for plant growth run out.
+
+Resources or temperature effects on the sink (the tree) can limit growth before the source (CO2) is limiting.
+The implication is that the benefits of CO2 fertilization will eventually reach a maximum as other resources necessary for plant growthare depleted.
 Currently, our understanding of when and where these limits will be reached is incomplete, so forest models tend to assume the increase in photosynthesis as atmospheric CO2 levels increase is unlimited.(Maybe not true, i'm not 100% familiar with the modeling still.)
@@
 Understanding how trees are responding to warming temperatures and rising CO2 levels, and how growth may become limited because of interactions between the two is essential to improve our predictions.
-Zohner et al, (2020)[@zohner_how_2021] found that under controled conditions, spring warming increases biomass accumulation in trees, with a larger proportion of root compared to shoot growth.
-The authors also found evidence of sink limitation in autumn, even with controlled water and fertilizer application to the study trees.
-To our knowledge, the link between phenology and growth has never been studied in-situ.
+Under controlled conditions, spring warming increases biomass accumulation in trees, with a larger proportion of root growth compared to shoot growth[@zohner_how_2021].
+To our knowledge, the link between stem phenology and growth has never been studied in-situ.
 
@@
 `r DOY_timing`).
-Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window during the critical temperature window (Figs.
+Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
 Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
-At HF, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
+At Harvard Forest, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
 
 Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig.
-`r DOY_timing`), there was no consistent lengthening of the period of peak stem growth.
+`r DOY_timing`), there was no consistent lengthening of $L_{PGS}$.
 Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig.
@@
 `r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
-In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in ~~CTW~~ April $T_{max}$ (**Cam, please confirm this change**).
+In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
 Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.
@@
 
-Spring warming at SCBI and HF in the study years used here did not result in increased $\Delta DBH$ or summer water stress (SPEI; Supp(?) SPEI Figure).
+Spring warming at SCBI and Harvard Forest in the study years used here did not result in increased $\Delta DBH$ or summer water stress (SPEI; Supp(?) SPEI Figure).
 This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at these sites.
-This could be because the woody growth at these sites are already sink-limited.
+This could be because the woody growth at these sites are already limited by some mechanism.
 
@@
 First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
-However, [extension of our tree-ring analysis](https://github.com/EcoClimLab/growth_phenology/issues/57) showed very little correlation between April $T_{max}$ and growth the following year (**#** of 142 RP and **#** of 66 DP species-site combinations), refuting the latter possibility.
-We cannot rule out the possibility that warm springs promote formation of more carbon-dense cell walls in the current of following year (**coauthors, do any of you know literature related to this?**); however, this possibility, our results suggest that little of the extra C that may be fixed in these years goes into wood formation.
+In fact, [extension of our tree-ring analysis](https://github.com/EcoClimLab/growth_phenology/issues/57) showed a generally positive, although often non-significant, correlation between ring-width and previous year growing season $T_{max}$ in RP trees, indicating this may be a possibility.
+*very little correlation between April $T_{max}$ and growth the following year (5 of 142 RP and 3 of 66 DP species-site combinations), refuting the latter possibility*.
+We also cannot rule out the possibility that warm springs promote formation of more carbon-dense cell walls in the current of following year (**coauthors, do any of you know literature related to this?**); however, our results suggest that little of the extra C that may be fixed in these years goes into wood formation.
 
@@
 `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs.
+Yet, an earlier start to the growing season is not associated with a longer peak growing season or increased growth (Figs.
 `r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@zhang_drought_2021;@dorangeville_drought_2018;@helcoski_growing_2019].
@@
 Modeled fits for tree-years were removed under five conditions:
-(1) single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, HF = 34); 
-(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for it's xylem architecture group, year, and site (SCBI = 62, HF = 106); 
-(3) tree-years with small or negligible total growth ($\Delta DBH$ < 0.02 mm; SCBI = 0, HF = 66); 
-(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 181, HF = 199); 
-and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, HF = 3). 
+(1) single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
+(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for it's xylem architecture group, year, and site (SCBI = 62, Harvard Forest = 106); 
+(3) tree-years with small or negligible total growth ($\Delta DBH$ < 0.02 mm; SCBI = 0, Harvard Forest = 66); 
+(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 181, Harvard Forest = 199); 
+and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3). 
 At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others *(SI figure?

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
add "consistently" for clarity
set title
@camerondow35 , I think we should go with "annual woody productivity" rather than "annual increments" in the title. The latter is more precise, but I think the former is more meaningful.

@@
 
-**Title:** Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity */annual increments*
+**Title:** Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity
 
@@
 `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Yet, an earlier start to the growing season is not associated with a longer peak growing season or increased growth (Figs.
+Yet, an earlier start to the growing season is not consistently associated with a longer peak growing season or increased growth (Figs.
 `r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@zhang_drought_2021;@dorangeville_drought_2018;@helcoski_growing_2019].

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
misc clean-up

@@
 
-*The introduction and summary paragraph need work, and removal of overlap. Organizational/content suggestions are welcome, but we're not at the stage of word-smithing.*
-
 Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
@@
 Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig.
-`r schematic_summary`, Extended Data Figures **4-6**).
+`r schematic_summary`, Extended Data Figures `r EDF_pheno_Tmax_climwin` - `r EDF_pheno_Tmin_AprilMay`).
 Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
@@
 Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.
-In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (Aug = 57, Jul = 90, Jun = 108, May = 47) and 46% (Aug = 22, Jul = 37, Jun = 53, May = 11) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
+In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
 Again, $T_{min}$ generally exerted less influence over annual growth than $T_{max}$ (`r EDF_tree_rings_TMIN`), with few significant correlations between spring $T_{min}$ and RWI.

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
misc clean-up
included moving figure captions from header material to figure locations (so that numbers could be auto-populated)

- fixed n chronologies at Harvard

@@
 #Figure numbers
-schematic_summary=1 #illustrates parameters
+schematic_summary=1 
 DOY_timing = 2
-  DOY_timing_SCBI = "2a"  #currently 4 panel!!!!
-  DOY_timing_Harvard = "2b"  #currently 4 panel!!!!
 tree_rings = 3
-
-#figure captions
-
-schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."
-
-DOY_timing_caption <- "**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
-Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
-Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-#needs attention!!!"
-
-tree_rings_caption <- "**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
-Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature.
-Chronology details are given in the Supplementary Information.
-(NOTE: Figure still needs some work.
-See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
-#info subject to change!!!
 
@@
 chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
-n_chronologies <- nrow(chronology_table)  #!!! entered by hand in figure captions
-n_cores_sites <- length(unique(chronology_table[, "Location"])) #!!! entered by hand in figure captions
+n_chronologies <- nrow(chronology_table)  
+n_cores_sites <- length(unique(chronology_table[, "Location"]))
 n_chronologies_SCBI <- 12
-n_chronologies_HF <- 3
+n_chronologies_HF <- 4
 ```
@@
 However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
-Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
+Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
 The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
@@
 
-![`r schematic_summary_caption`](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 
@@
 
-![`r DOY_timing_caption`](tables_figures/DOYtiming_allyears.png){ width=25% }
+![**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
+Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }
 
@@
 
-![`r tree_rings_caption`](tables_figures/tree_rings_tmax.png){ width=100% }
+![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature.
+Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmax.png){ width=100% }
 
@@
 `r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
-In total, our analysis included `r_n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
+In total, our analysis included `r n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
 Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
update word count
@camerondow35 , we'll need to shorten this : currently at 3300 words body text, vs target of 2000-2500.
Update Dow_manuscript_Nature.Rmd

@@
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~2700 words (or ~3000 with fig captions) and 2 modest + 2 large display items.
+We're currently at ~3333 words (or ~3690 with fig captions) and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently ok there)*
 
-# (Summary paragraph)
+# (Summary paragraph - 200 words)
 ```{r eval=FALSE, echo=FALSE}
@@
 
-*removed from abstract; to integrate elsewhere:*
-
-tree stem growth consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021]
-
-These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
-They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-
-\newpage
+# (Body)
 
@@
 
+# content to incorporate or delete:
+
+These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
+They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
+
 # Methods {-#Methods}

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
worked on intro
@camerondow35 , FYI -
Since this needs dramatic condensing, I removed all content that repeated the summary paragraph. I also moved some content down for integration into the discussion.

@@
 
-Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
-In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
-Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
-Additionally, Warmer summer and autumn temperatures coupled with drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018].
-As a result, warming may advance or delay autumn senescence depending on timing of warming and water availability, with delays more common across eastern North America [@buermann_widespread_2018].
+In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
+The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
+However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors become limiting[@zani_increased_2020; @zohner_how_2021] suggests that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
 
-Spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;;@ibanez_forecasting_2010;;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
-The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
-
-However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors become limiting [@zani_increased_2020; @zohner_how_2021] suggests that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
-
-Temperate deciduous forests, including those in eastern North America, have historically been net carbon sinks.
-
-Resources or temperature effects on the sink (the tree) can limit growth before the source (CO2) is limiting.
-The implication is that the benefits of CO2 fertilization will eventually reach a maximum as other resources necessary for plant growthare depleted.
-Currently, our understanding of when and where these limits will be reached is incomplete, so forest models tend to assume the increase in photosynthesis as atmospheric CO2 levels increase is unlimited.(Maybe not true, i'm not 100% familiar with the modeling still.)
+This would be concerning in that the C sink of temperate deciduous forests, including those in eastern North America, is important for reducing atmospheric CO~2~ concentrations(**REFS**).
 
@@
 See European study cited in A-T et al in review)*
-Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019;@] [MORE]).
+Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (**MORE REFS--preferably some that are already cited**), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [**MORE REFS--but only those that are already cited**]).
 Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
-Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)
+Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importance of stem growth as a carbon sink.)
 
@@
 Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a *fairly long* residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
+*BUT SEE @piao_net_2008!! and @barichivich_large-scale_2013 - these are focused on higher latitudes, but shows greater release of C in the fall!*
 
@@
 # content to incorporate or delete:
+
+Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
+Additionally, Warmer summer and autumn temperatures coupled with drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018].
+As a result, warming may advance or delay autumn senescence depending on timing of warming and water availability, with delays more common across eastern North America [@buermann_widespread_2018].
 

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
big-picture rearrangement of intro

@@
 
+As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
+Temperate deciduous forests, including those in eastern North America, are important for reducing atmospheric CO~2~ concentrations(**REFS**).
+*yet models remain uncertain.*
+
 In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
@@
 However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors become limiting[@zani_increased_2020; @zohner_how_2021] suggests that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
-
-This would be concerning in that the C sink of temperate deciduous forests, including those in eastern North America, is important for reducing atmospheric CO~2~ concentrations(**REFS**).
-
-As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
-Understanding how trees are responding to warming temperatures and rising CO2 levels, and how growth may become limited because of interactions between the two is essential to improve our predictions.
-Under controlled conditions, spring warming increases biomass accumulation in trees, with a larger proportion of root growth compared to shoot growth[@zohner_how_2021]. 
-To our knowledge, the link between stem phenology and growth has never been studied in-situ.
 
@@
 See European study cited in A-T et al in review)*
+To our knowledge, the link between stem phenology and growth has never been studied *in-situ*.
 Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (**MORE REFS--preferably some that are already cited**), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [**MORE REFS--but only those that are already cited**]).
@@
 
+Under controlled conditions, spring warming increases biomass accumulation in trees, with a larger proportion of root growth compared to shoot growth[@zohner_how_2021].
+
 # Methods {-#Methods}

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
worked on intro
@camerondow35 , FYI. I think this is 99% there now. (needs a few refs)

@@
 
-As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
-Temperate deciduous forests, including those in eastern North America, are important for reducing atmospheric CO~2~ concentrations(**REFS**).
-*yet models remain uncertain.*
+In recent decades, forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby substantially slowing the pace of atmospheric CO~2~ accumulation and climate change(**REF**).
+A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes [@pugh_role_2019].
+As Earth's atmospheric CO~2~ concentrations and temperatures continue to increase, the behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
+It remains difficult to predict how temperate deciduous forests will respond to the combined influences of multiple global change drivers, including warming temperatures(**REF**).
 
@@
 
-We know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few site [@doranngeville_radial_inreview; @martin_hydrometeorology_2017].
-*(We need to make sure we're not missing anything.
-See European study cited in A-T et al in review)*
-To our knowledge, the link between stem phenology and growth has never been studied *in-situ*.
-Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (**MORE REFS--preferably some that are already cited**), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [**MORE REFS--but only those that are already cited**]).
-Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
-Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importance of stem growth as a carbon sink.)
+While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few sites [e.g., Ref.
+@dorangeville_peak_2021].
+<!-- one example will suffice here. others include one cited in Anderson-Teixeira et al. GCB tree rings. martin_hydrometeorology_2017 does this, but that is for conifers.-->
+To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a sapling experiment).
+
+Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
+<!-- dorangeville_drought_2018 looks at PET.-->
+with some hints that warmer springs may have a modest positive effect on growth (e.g., Ref.[@helcoski_growing_2019]).
+<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
+Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual CO~2~ sequestration, thereby providing a negative feedback to climate change[@keenan_net_2014].
+Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
 
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@doranngeville_radial_inreview], and set in the context of leaf phenology at these sites.
+This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology at these sites.
 
@@
 `r DOY_timing`, `r  EDT_mean_growth_parameters`).
-Consistent with the results of Ref[@doranngeville_radial_inreview], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
+Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_mean_growth_parameters`).
@@
 Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), and hickories (*Carya spp.*). 
-The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July ([helcoski_growing_2019]).
+The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July (Ref.[@helcoski_growing_2019]).
 
@@
 Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
-The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July (D'Orangeville et al, ).
+The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July [@dorangeville_peak_2021].
+<!-- Is this the right reference? -->
 

### 2021-10-05 doc/manuscript/Dow_manuscript_Nature.Rmd
shortened intro, misc cleanup of intro/methods/results

@@
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~3333 words (or ~3690 with fig captions) and 1 modest + 2 large display items.
+We're currently at ~3116 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently ok there)*
 
-# (Summary paragraph - 200 words)
+# (Summary paragraph)
 ```{r eval=FALSE, echo=FALSE}
@@
 
-In recent decades, forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby substantially slowing the pace of atmospheric CO~2~ accumulation and climate change(**REF**).
+In recent decades, forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
 A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes [@pugh_role_2019].
-As Earth's atmospheric CO~2~ concentrations and temperatures continue to increase, the behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
-It remains difficult to predict how temperate deciduous forests will respond to the combined influences of multiple global change drivers, including warming temperatures(**REF**).
+The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
 
 In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
-However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors become limiting[@zani_increased_2020; @zohner_how_2021] suggests that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
+The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, _NEE_)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
+However, recent findings that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that <!--leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] (lempereur_growth_2015 studied an evergreen)--> sink factors can limit annual growth[@zani_increased_2020; @zohner_how_2021] suggest that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
 
 While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few sites [e.g., Ref.
-@dorangeville_peak_2021].
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few sites (e.g., Ref.[@dorangeville_peak_2021]).
 <!-- one example will suffice here. others include one cited in Anderson-Teixeira et al. GCB tree rings. martin_hydrometeorology_2017 does this, but that is for conifers.-->
-To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a sapling experiment).
+To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
 
@@
 <!-- dorangeville_drought_2018 looks at PET.-->
-with some hints that warmer springs may have a modest positive effect on growth (e.g., Ref.[@helcoski_growing_2019]).
+with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].
 <!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
@@
 
-First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees from 2011-2020) and Harvard Forest (Massachusetts; n = **755** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **146** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **755** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
-This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology at these sites.
+This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology (measured at ecosystem level via remote sensing).
 
@@
 
-For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs.
+For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs.
 `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOY of leaf greenup (measured at ecosystem level via remote sensing) advanced with increasing $T_{max}$ during the critical temperature window (p=0.001).
-Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window, and mid-greenup also advanced, but to a lesser extent (Fig.
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001).
+Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest *when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window*<!--Cameron, was this based on comparing just the warmest and coldest years, or a linear regression.
+We want the latter, and to remove the italic text -->, and mid-greenup also advanced, but to a lesser extent (Fig.
 `r DOY_timing`).
@@
 Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018; ] or inducing the cessation of stem growth when a critical water deficit is reached[@lempereur_growth_2015].
+<!-- lempereur_growth_2015 studied an evergreen oak-->
 
@@
 Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
+<!-- lempereur_growth_2015 studied an evergreen-->
 Additionally, Warmer summer and autumn temperatures coupled with drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018].

### 2021-10-06 doc/manuscript/Dow_manuscript_Nature.Rmd
edit Dendrobands section

@@
 EDF_climwin = "Extended Data Figure 3"
+EDF_pheno_all = "Extended Data Figures 4-6"
 EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
@@
 While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few sites (e.g., Ref.[@dorangeville_peak_2021]).
-<!-- one example will suffice here. others include one cited in Anderson-Teixeira et al. GCB tree rings. martin_hydrometeorology_2017 does this, but that is for conifers.-->
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with *multi-year records of growth phenology having been published for only one site[@dorangeville_peak_2021]*. 
+<!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
 To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
@@
 
-Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **146** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **755** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **146** <!--Cameron, please verify number--> trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **755** <!--Cameron, please verify number--> trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 
-Both leaf phenology and stem growth were sensitive to spring temperatures (Figs. `r DOY_timing`, `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`). 
-While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-7), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
+Both leaf phenology and stem growth were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
+While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-6), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig.
@@
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001).
-Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest *when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window*<!--Cameron, was this based on comparing just the warmest and coldest years, or a linear regression.
+Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest ~~when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window~~<!--Cameron, was this based on comparing just the warmest and coldest years, or a linear regression.
 We want the latter, and to remove the italic text -->, and mid-greenup also advanced, but to a lesser extent (Fig.
@@
 Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
-`r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
+`r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
 Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
@@
 `r DOY_timing`), there was no consistent lengthening of $L_{PGS}$.
-Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig.
-`r schematic_summary`, Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
+Specifically, because both $DOY_{25}$ and $DOY_{75}$ advanced in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, `r EDF_pheno_all`).
 
-In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs.
-`r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig.
-`r schematic_summary`, Extended Data Figures `r EDF_pheno_Tmax_climwin` - `r EDF_pheno_Tmin_AprilMay`).
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
-while $\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`). 
+In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest).
+$\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`). 
 Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth.

### 2021-10-06 doc/manuscript/Dow_manuscript_Nature.Rmd
edited tree-ring section
@camerondow35 , I dramatically shortened this, removing the separate detailed discussion of patterns at SCBI and HF.

I also started a short paragraph on the second analysis, which will need some details filled in.

@@
 
-To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` locations (Fig.
-`r tree_rings`, `r EDF_map`).
-For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, ring width indices (RWI) rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Fig.
-`r tree_rings`, `r EDF_tree_rings_TMIN`).
-Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
-responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral response to $T_{max}$ (Fig.
-`r tree_rings`).
-In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
-Starting in May at SCBI and June at Harvard Forest, negative correlations of RWI to $T_{max}$ were predominant, being statistically significant in at least one growing season month for all `r n_chronologies_SCBI` species at SCBI and one of `r n_chronologies_HF` species at Harvard Forest.
-Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`).
+To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), and an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
+`r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
+In total, our analysis included `r n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
+Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.
+In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
+$T_{min}$ generally exerted less influence over annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
+
+To test whether a positive effect of April temperatures might be offset by a negative effect of summer temperatures, we tested for their joint and interactive effects.
+Results were qualitatively similar to the univariate correlations (Fig.
+`r tree_rings`), with a positive effect of April $T_{max}$ for only **#**% of records and significant negative correlations with summer $T_{max}$ for **##-##**% of records, depending on the set of summer months included in the analysis (**(*Appendix Table*)**).
+Significant interactive effects between spring and summer temperatures were rare (**DETAILS**).
 
@@
 Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmax.png){ width=100% }
-
-The patterns observed at our focal sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
-`r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
-In total, our analysis included `r n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
-Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.
-In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
-Again, $T_{min}$ generally exerted less influence over annual growth than $T_{max}$ (`r EDF_tree_rings_TMIN`), with few significant correlations between spring $T_{min}$ and RWI.
 

### 2021-10-06 doc/manuscript/Dow_Extended_Data.Rmd
create Extended Data Table 3
#75

@@
 mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest."
-tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring data and analysis ( including species, n chronologies (= n sites), n positively correlated with April T, n sig.)"
+tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring data and analysis"
 
@@
 
-## (PLACEHOLDER FOR `r tree_ring_summary_caption` )
+## `r tree_ring_summary_caption`
+
+```{r tree_ring_summary,  eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+TreeRingResultsSummary <- read.csv("tables_figures/TreeRingResultsSummary.csv",  
+                                   col.names = c("", "species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
+                                   stringsAsFactors = FALSE, check.names = FALSE)
+TreeRingResultsSummary <- TreeRingResultsSummary[c(2:19)]
+kable(TreeRingResultsSummary, 
+      #caption = tree_ring_summary_caption ,  
+      format = "latex", booktabs = TRUE,  escape = F) %>%
+add_header_above(c(" " = 2,  "April" = 2, "May" = 2,"June" = 2,"July" = 2,"Aug" = 2,"April" = 2,"June-July" = 2,"A:JJ" = 2)) %>%
+add_header_above(c(" " = 1,  "n sig. correlations in univariate analysis" = 10,"n sig. correlations in multivariate analysis" = 6)) %>%
+    #column_spec(3:18, width = ".3cm") %>%
+pack_rows("Ring Porous", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+pack_rows("Diffuse Porous", 17, 27, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+row_spec(28, bold = T) %>%
+kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
 

### 2021-10-06 doc/manuscript/Dow_manuscript_Nature.Rmd
reference Extended Data Table 3
#75

@@
 To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), and an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
-`r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
+`r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
 In total, our analysis included `r n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
-Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations.
+Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
 In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
@@
 
-To test whether a positive effect of April temperatures might be offset by a negative effect of summer temperatures, we tested for their joint and interactive effects.
+To test whether a positive effect of spring temperatures might be offset by a negative effect of summer temperatures, we tested for the joint and interactive effects of April and June-July $T_{max}$ on RWI.
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with a positive effect of April $T_{max}$ for only **#**% of records and significant negative correlations with summer $T_{max}$ for **##-##**% of records, depending on the set of summer months included in the analysis (**(*Appendix Table*)**).
-Significant interactive effects between spring and summer temperatures were rare (**DETAILS**).
+`r tree_rings`), with a positive effect of April $T_{max}$ for only **#**% of records and significant negative correlations with summer $T_{max}$ for **##**% of chronologies (`r EDT_tree_ring_summary`).
+Significant interactive effects between spring and summer temperatures were rare (**##**% of chronologies).
 

### 2021-10-06 doc/manuscript/Dow_Extended_Data.Rmd
update ED Table 3
#75

--- 2021-09-17 doc/manuscript/Dow_Extended_Data.Rmd
@@
 mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest."
-tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring data and analysis ( including species, n chronologies (= n sites), n positively correlated with April T, n sig.)"
+tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations to monthly $T_{max}$ in univariate and multivariate analyses."
 
@@
 
-## (PLACEHOLDER FOR `r tree_ring_summary_caption` )
+```{r tree_ring_summary,  eval=TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+TreeRingResultsSummary <- read.csv("tables_figures/TreeRingResultsSummary.csv",  
+                                   col.names = c("", "species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
+                                   stringsAsFactors = FALSE, check.names = FALSE)
+TreeRingResultsSummary <- TreeRingResultsSummary[c(2:19)]
+kable(TreeRingResultsSummary, 
+      caption = tree_ring_summary_caption,  
+      format = "latex", booktabs = TRUE,  escape = F) %>%
+add_header_above(c(" " = 2,  "April" = 2, "May" = 2,"June" = 2,"July" = 2,"Aug" = 2,"April" = 2,"June-July" = 2,"A:JJ" = 2)) %>%
+add_header_above(c(" " = 2,  "univariate analysis: n sig." = 10,"multivariate analysis: n sig." = 6)) %>%
+    #column_spec(3:18, width = ".3cm") %>%
+pack_rows("Ring Porous", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+pack_rows("Diffuse Porous", 17, 27, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+row_spec(28, bold = T) %>%
+kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
 

### 2021-10-06 doc/manuscript/Dow_Extended_Data.Rmd
remove numbers from column names
#75

@@
 TreeRingResultsSummary <- read.csv("tables_figures/TreeRingResultsSummary.csv",  
-                                   col.names = c("", "species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
                                    stringsAsFactors = FALSE, check.names = FALSE)
@@
       caption = tree_ring_summary_caption,  
+      col.names = c("species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
       format = "latex", booktabs = TRUE,  escape = F) %>%

### 2021-10-07 doc/manuscript/Dow_manuscript_Nature.Rmd
word count
edits to intro and discussion

@@
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~3116 words and 1 modest + 2 large display items.
+We're currently at ~2922 words (1613 intro + results, discussion 1319--too long!) and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently ok there)*
@@
 These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
-Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
+Thus, in contrast to global carbon cycle model expectations[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014], warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
 
@@
 In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, _NEE_)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
+The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, _NEE_)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and this expectation is represented in models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
 However, recent findings that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that <!--leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] (lempereur_growth_2015 studied an evergreen)--> sink factors can limit annual growth[@zani_increased_2020; @zohner_how_2021] suggest that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
@@
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with *multi-year records of growth phenology having been published for only one site[@dorangeville_peak_2021]*.
-<!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
+<!-- I'm not aware of any others, but we do want to make sure we're not missing anything.
+klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
 To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
@@
 
+<!--1309 words before Krista's edits. Goal: 700-800-->
+
 Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs.
 `r schematic_summary`- `r tree_rings`).
-The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig.
+
+The observed advance in the timing of stem growth milestones under warmer springs parallels phenological advances observed for leaf development (Fig.
 `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
-*It would be interesting to compare the magnitude of these advances (days per degree C)*
 However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
@@
 First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) would lead to an earlier depletion of resources needed for growth resulting in an earlier end to the season of rapid growth ($DOY_{75}$).
-Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018; ] or inducing the cessation of stem growth when a critical water deficit is reached[@lempereur_growth_2015].
-<!-- lempereur_growth_2015 studied an evergreen oak-->
+Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018].
 
@@
 The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig.
-`r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
+`r tree_rings`, `r EDF_tree_rings_TMIN`, `r EDT_tree_ring_summary`)[@alexander_potential_2019; @helcoski_growing_2019].
 Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
-Linear models were run to examine the relationship between spring, summer and an interaction effect with ring-width for each of our 208 chronologies.
-The results indicated that, across all latitudes, the interacting effect of warmer springs and summers resulted in negligible increase of ring-width (supp (?) rw_drought_plot).
 *It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019; @zhang_drought_2021].
-(Is this relevent?)*(Neil suggested that the zhang paper would be good here but i'm not sure I see how...)
-*(ADD MORE ON DROUGHT HERE)*
+(Is this relevent?)*<!-- (Neil suggested that the zhang paper would be good here but i'm not sure I see how...)-->
+<!-- *(ADD MORE ON DROUGHT HERE)*-->
 
@@
 We also cannot rule out the possibility that warm springs promote formation of more carbon-dense cell walls in the current of following year (**coauthors, do any of you know literature related to this?**); however, our results suggest that little of the extra C that may be fixed in these years goes into wood formation.
+
+*There may be a shift towards greater belowground allocation[@zohner_how_2021].*
 
@@
 These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
-They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
+They contrast, however, with model expectations[@zani_increased_2020;@buermann_widespread_2018;@keenan_net_2014] that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
 

### 2021-10-12 doc/manuscript/Dow_Extended_Data.Rmd
put multivariate model without the interaction in Table S3
#78

@@
                                    stringsAsFactors = FALSE, check.names = FALSE)
-TreeRingResultsSummary <- TreeRingResultsSummary[c(2:19)]
+TreeRingResultsSummary <- TreeRingResultsSummary[c(2:17)]
 kable(TreeRingResultsSummary, 
       caption = tree_ring_summary_caption,  
-      col.names = c("species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
+      col.names = c("species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
       format = "latex", booktabs = TRUE,  escape = F) %>%
-add_header_above(c(" " = 2,  "April" = 2, "May" = 2,"June" = 2,"July" = 2,"Aug" = 2,"April" = 2,"June-July" = 2,"A:JJ" = 2)) %>%
-add_header_above(c(" " = 2,  "univariate analysis: n sig." = 10,"multivariate analysis: n sig." = 6)) %>%
+add_header_above(c(" " = 2,  "April" = 2, "May" = 2,"June" = 2,"July" = 2,"Aug" = 2,"April" = 2,"June-July" = 2)) %>%
+add_header_above(c(" " = 2,  "univariate analysis" = 10,"multivariate analysis" = 4)) %>%
+add_header_above(c(" " = 2,  "n. correlations significant at p=0.05" = 14)) %>%
     #column_spec(3:18, width = ".3cm") %>%
 pack_rows("Ring Porous", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-pack_rows("Diffuse Porous", 17, 27, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-row_spec(28, bold = T) %>%
+pack_rows("Diffuse Porous", 17, 26, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+row_spec(27, bold = T) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2021-10-12 doc/manuscript/Dow_manuscript_Nature.Rmd
update manuscript with new multivariate analysis
#78

@@
 
-To test whether a positive effect of spring temperatures might be offset by a negative effect of summer temperatures, we tested for the joint and interactive effects of April and June-July $T_{max}$ on RWI.
+To test whether a positive effect of spring temperatures might be offset by a negative effect of summer temperatures, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with a positive effect of April $T_{max}$ for only **#**% of records and significant negative correlations with summer $T_{max}$ for **##**% of chronologies (`r EDT_tree_ring_summary`).
-Significant interactive effects between spring and summer temperatures were rare (**##**% of chronologies).
+`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies (`r EDT_tree_ring_summary`).
 

### 2021-10-12 RScripts/tree_core_drought_interaction.R
use no-interaction model for chrono table and table s3
#78

@@ -150,6 +150,13 @@ jun_apr_jun <- a[["coefficients"]][3,4]
 jun_apr_jun_est <- a[["coefficients"]][3,1]
 apr_jun_r <- a[["adj.r.squared"]]
 
+# #Estaban's way!
+# a <- lm(ring_width ~ april + june + april:june, data = crns_sub)
+# coefficients (a) #provides the coefficients of the model (these are conditional coefficients: focus on one variable holding the others constant)
+# summary(a) #provides the F, df, and P value for the whole model (check the last line of the output)
+# Anova(a, type =3) #provides the F, df, and P value for each independent factor
+# anova_stats(a, digits =3) #provides the effect size (we are reporting partial Omega squared)
+
 # ab <- summary(lm(ring_width ~ june, data = crns_sub))
 # jun <- ab[["coefficients"]][2,4]
 # jun_est <- ab[["coefficients"]][2,1]
@@ -165,6 +172,13 @@ jj_apr_jun_jul_est <- b[["coefficients"]][3,1]
 apr_jun_jul <- b[["coefficients"]][4,4]
 apr_jun_jul_est <- b[["coefficients"]][4,1]
 apr_jun_jul_r <- b[["adj.r.squared"]]
+
+#Estaban's way!
+# b <- lm(ring_width ~ april + june_july + april:june_july, data = crns_sub)
+# coefficients (b) #provides the coefficients of the model (these are conditional coefficients: focus on one variable holding the others constant)
+# summary(b) #provides the F, df, and P value for the whole model (check the last line of the output)
+# Anova(b, type =3) #provides the F, df, and P value for each independent factor
+# anova_stats(b, digits =3) #provides the effect size (we are reporting partial Omega squared)
 
 # bb <- summary(lm(ring_width ~ june_july, data = crns_sub))
 # jun_jul <- bb[["coefficients"]][2,4]
@@ -281,8 +295,21 @@ names(sp_codes) <- c("sp", "Species")
 #read in chronology table csv which we are appending the lm results to
 chronology_table <- read.csv("doc/manuscript/tables_figures/chronology_table.csv")
 chronology_table <- chronology_table[,1:8]
-chronology_table$Latitude <- round(chronology_table$Latitude, digits = 7)
[…]

### 2021-10-12 RScripts/tree_core_drought_no_interactions.R

@@ -1,0 +1,421 @@
+#Script to analyze the joint effect of spring T and summer drought on tree growth using tree cores#
+#Written by: Cameron Dow###########################################################################
+#Last edit: 8/16/2021##############################################################################
+
+rm(list = ls())
+
+library(tidyverse)
+library(readr)
+library(lme4)
+library(lmerTest)
+library(emmeans)
+library(reshape2)
+library(readxl)
+
+#Create crns_long ----
+crns <- read_csv("Data/tree_rings/Other/all_crns_res_1901.csv") %>%
+  # clean up
+  select(-c(BearIs, OH_Gol_QUAL_1, PineMt_QUMO)) %>%
+  rename(IL_Fer_LITU = IL_Fer_LTU,
+         IL_Bea_QURU = IL_BeaQURU)
+
+crns_long <- crns %>%
+  # convert to long format
+  pivot_longer(-Year, names_to = "site_sp", values_to = "ring_width") %>%
+  # drop years with missing data
+  filter(!is.na(ring_width))
+
+crns_long$Location <- substr(crns_long$site_sp, 1,nchar(crns_long$site_sp)-5)
+
+#Add lat / lon
+TRW_coord <- read_excel("Data/tree_rings/Other/TRW_coord2.xlsx")
+TRW_coord <- TRW_coord[,c(1,2,3)]
+#Add original two locations to include in final quilt plot
+originals <- data.frame(42.5388, -72.18, "HF") #Lat for HF = 42.5388
+originals <- rbind(originals, c(38.8935, -78.1454, "SCBI")) #Lat for scbi = 38.8935
+names(originals) <- c("Latitude","Longitude", "Location")
+TRW_coord <- rbind(TRW_coord, originals)
+TRW_coord <- TRW_coord[!duplicated(TRW_coord$Location),]
+crns_long_try <- left_join(crns_long, TRW_coord, by = "Location")
+
+#replace na's with correct values
+crns_long_try[crns_long_try$Location %in% "Fiddler<U+393C><U+3E32>s_Green,_VA_CAOV", "Latitude"] <- "37.7707167"
+crns_long_try[crns_long_try$Location %in% "Fiddler<U+393C><U+3E32>s_Green,_VA_CAOV", "Longitude"] <- "-79.2416"
+
[…]

### 2021-10-12 doc/manuscript/Dow_Extended_Data.Rmd
knit in updated table S3
#78

@@
 add_header_above(c(" " = 2,  "univariate analysis" = 10,"multivariate analysis" = 4)) %>%
-add_header_above(c(" " = 2,  "n. correlations significant at p=0.05" = 14)) %>%
+add_header_above(c(" " = 2,  "n. correlations to monthly max. T significant at p=0.05" = 14)) %>%
     #column_spec(3:18, width = ".3cm") %>%

### 2021-10-13 RScripts/additional_figures.R
add EVI2 plot
#48

@@ -6,7 +6,7 @@ library(tidybayes)
 library(ggnewscale)
 library(patchwork)
 library(lubridate)
-
+library(gridExtra)
 
 
 # Figure 1 v2: Schematic illustrating parameters and general sense of results ----
@@ -456,19 +456,43 @@ percent_growth_hf <- Wood_pheno_table_hf %>%
     dbh_growth_percent_cummulative = cumsum(dbh_growth_percent)
   )
 
+#EVI2 plots
+scbiEVI2 <- read_csv("Data/Leaf phenology/scbiEVI2.csv")
+scbiEVI2$year <- year(scbiEVI2$date)
+scbiEVI2$doy <- yday(scbiEVI2$date)
+#SCBI plot
+scbiEVI <- scbiEVI2[scbiEVI2$pointid %in% "SCBI",]
+scbiEVI <- scbiEVI[scbiEVI$BRDF_Albedo_Band_Mandatory_Quality_Band1 %in% 0,]
+#HF plot
+hfEVI <- scbiEVI2[scbiEVI2$pointid %in% "HARV",]
+hfEVI <- hfEVI[hfEVI$BRDF_Albedo_Band_Mandatory_Quality_Band1 %in% 0,]
 
-
-png(filename = "doc/manuscript/tables_figures/growth_curves_all.png", width=10, height=5,
+png(filename = "doc/manuscript/tables_figures/growth_curves_all.png", width=10, height=10,
     pointsize=12, bg="transparent", units="in", res=600,
     restoreConsole=FALSE)
 
 grid.arrange(
+  ggplot(scbiEVI, aes(x = doy, y = EVI2, group = year)) +
+    geom_smooth(se = FALSE, size = .5) +
+    theme_bw() +
+    #scale_color_viridis_d("", end = 2/3)+
+    theme(legend.position = "none") +
+    labs(x = "", title = "SCBI") +
+    ylim(0, 0.8),
+
+  ggplot(hfEVI, aes(x = doy, y = EVI2, group = year)) +
+    geom_smooth(se = FALSE, size = 0.5) +
+    theme_bw() +
+    #scale_color_viridis_d("", end = 2/3)+
+    theme(legend.position = "none") +
+    labs(x = "",y = "", title = "Harvard Forest") +
+    ylim(0,0.8),
 
   ggplot(percent_growth_scbi, aes(x = doy, y = dbh_growth_percent, group = tag_year)) +
     coord_cartesian(ylim = c(0, 0.05)) +
     scale_y_continuous(labels = percent) +
[…]

### 2021-10-13 doc/manuscript/Dow_manuscript_Nature.Rmd
tweak intro

@@
 
-In recent decades, forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
-A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes [@pugh_role_2019].
+In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
+A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019].
 The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
@@
 In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, _NEE_)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and this expectation is represented in models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent findings that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that <!--leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] (lempereur_growth_2015 studied an evergreen)--> sink factors can limit annual growth[@zani_increased_2020; @zohner_how_2021] suggest that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.
+The expectation is that as growing season length increases, net annual carbon sequestered (*i.e.*, net ecosystem exchange, _NEE_) will increase as well[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and this expectation is represented in models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+However, recent findings that sink factors can limit annual growth[@zani_increased_2020; @zohner_how_2021] and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018] suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
 
 While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with *multi-year records of growth phenology having been published for only one site[@dorangeville_peak_2021]*.
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021].
 <!-- I'm not aware of any others, but we do want to make sure we're not missing anything.
 klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
-To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
+The climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
 
@@
 <!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
-Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual CO~2~ sequestration, thereby providing a negative feedback to climate change[@keenan_net_2014].
+Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ sequestration[@keenan_net_2014]. 
 Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
 
-Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
+Here, we describe how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
 

### 2021-10-13 doc/manuscript/Dow_Extended_Data.Rmd
knit in new Extended Data Fig. 2
@camerondow35 , FYI, this and other figure captions still need some details filled in/ word-smithing

@@
 
-growth_curves_all_caption  <- " **Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.** Each line represents one year's growth for a given tree, fit with McMahon model.
+growth_curves_all_caption  <- " **Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at our two focal study sites** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2) ..., with each line representing a year between 2001 and 2018.
+For stem growth, each line represents one year's growth for a given tree, fit with McMahon model.
 " #needs wordsmithing!!!

### 2021-10-13 doc/manuscript/Dow_manuscript_Nature.Rmd
work on structuring discussion

@@
 
+<!-- Spring warming at SCBI and Harvard Forest in the study years used here did not result in increased $\Delta DBH$ or summer water stress (SPEI; Supp(?) SPEI Figure).
+This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at these sites.
+This could be because the woody growth at these sites are already limited by some mechanism.
+-->
+
 ## Tree ring analysis {-#Tree}
@@
 
+<!-- how our findings fill knowledge gap (1 par, maybe 2)-->
+
 Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs.
@@
 
-Spring warming at SCBI and Harvard Forest in the study years used here did not result in increased $\Delta DBH$ or summer water stress (SPEI; Supp(?) SPEI Figure).
-This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at these sites.
-This could be because the woody growth at these sites are already limited by some mechanism.
-
 The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig.
@@
 
+<!-- it remains unclear where extra carbon goes (1 par, maybe 2)-->
 Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
@@
 
+<!-- interannual variation ≠ directional change (1 par)-->
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
@@
 
+<!--"contributions beyond" (1 par)-->
 Despite remaining uncertainties, our findings contribute substantially to a broader picture of the influence of spring temperatures on forest ecosystem function. 
@@
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the sustained increase in carbon sequestration that is anticipated in most models (Fig.
-`r schematic_summary`)[@zani_increased_2020].
-Because of this, temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that these models are underestimating future overall atmospheric CO~2~.
-(*see [GitHub issue 56 on representation in models](https://github.com/EcoClimLab/growth_phenology/issues/56)*)
+<!--conclusion (1 par)-->
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models (Fig.
+`r schematic_summary`)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig.
+`r tree_rings`), which constitutes a positive feedback to climate change.
 

### 2021-10-13 doc/manuscript/Dow_Extended_Data.Rmd
answer some comments / add info to extended figure captions

--- 2021-10-12 doc/manuscript/Dow_Extended_Data.Rmd
@@
 
-growth_curves_all_caption  <- " **Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.** Each line represents one year's growth for a given tree, fit with McMahon model.
+growth_curves_all_caption  <- " **Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at our two focal study sites** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2) ..., with each line representing a year between 2000 and 2018.
+For stem growth, each line represents one year's growth for a given tree, fit with McMahon model.
 " #needs wordsmithing!!!
@@
 
-pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW).
-For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands.
+pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+95% credible intervals are represented by bands.
+The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective CTW.
+For each CTW $T_{max}$.
+Modeled values from dendrometer band measurements are represented by dots.
 (**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
@@
 
-pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
+pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).
+Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+95% credible intervals are represented by bands.
+(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
 CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
 
-pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
+pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).
+Relationships between critical temperature window (CTW) $T_min$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+95% credible intervals are represented by bands.
+(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
 CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
 
-tree_rings_TMIN_caption <- "Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 207 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+tree_rings_TMIN_caption <- "Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.

### 2021-10-13 doc/manuscript/Dow_manuscript_Nature.Rmd
answer some comments / add info to extended figure captions

@@
 
-Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **146** <!--Cameron, please verify number--> trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **755** <!--Cameron, please verify number--> trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001).
-Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest ~~when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window~~<!--Cameron, was this based on comparing just the warmest and coldest years, or a linear regression.
-We want the latter, and to remove the italic text -->, and mid-greenup also advanced, but to a lesser extent (Fig.
-`r DOY_timing`).
+Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
 Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
@@
 The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July [@dorangeville_peak_2021].
-<!-- Is this the right reference? -->
+<!-- Is this the right reference? Yes, I got this info from the draft they sent. -->
 
@@
 The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
-Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
+Across years, the number of bands sampled averaged *~124* (range: 91-138) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
 In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).
@@
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
-(*how were gaps filled for Harvard? Or was the record good?*)
+(*how were gaps filled for Harvard? Or was the record good? --- Data came "pre-cleaned" as far as I could tell*)
 

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
small edits to results

@@
 n_cores_sites <- length(unique(chronology_table[, "Location"]))
+n_cores_species <- length(unique(chronology_table[, "Species"]))
 n_chronologies_SCBI <- 12
@@
 
-To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), and an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
+To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
 `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
-In total, our analysis included `r n_chronologies` species at sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
+In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
 Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
-In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
+In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
 $T_{min}$ generally exerted less influence over annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
worked on discussion

@@
 `r schematic_summary`- `r tree_rings`).
+The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for leaf development (Fig.
+`r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
+However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig.
+`r tree_rings`).
 
-The observed advance in the timing of stem growth milestones under warmer springs parallels phenological advances observed for leaf development (Fig.
-`r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
-However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
-This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possible mechanisms.
-First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) would lead to an earlier depletion of resources needed for growth resulting in an earlier end to the season of rapid growth ($DOY_{75}$).
-Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018].
+This suggests that any growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possible mechanisms.
+First, the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`), is likely driven by cues other than carbon limitation, such as daylength or sink limitation[@zani_increased_2020].
+<!--potentially expand on sink limitation here-->
 
-The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig.
-`r tree_rings`, `r EDF_tree_rings_TMIN`, `r EDT_tree_ring_summary`)[@alexander_potential_2019; @helcoski_growing_2019].
-Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
-*It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019; @zhang_drought_2021].
-(Is this relevent?)*<!-- (Neil suggested that the zhang paper would be good here but i'm not sure I see how...)-->
-<!-- *(ADD MORE ON DROUGHT HERE)*-->
+*First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) would lead to an earlier depletion of resources needed for growth resulting in an earlier end to the season of rapid growth ($DOY_{75}$).*
+*Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018].*
+
+*Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].*
 

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion

@@
 `r tree_rings`).
-
-This suggests that any growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possible mechanisms.
-First, the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`), is likely driven by cues other than carbon limitation, such as daylength or sink limitation[@zani_increased_2020].
-<!--potentially expand on sink limitation here-->
-
-*First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) would lead to an earlier depletion of resources needed for growth resulting in an earlier end to the season of rapid growth ($DOY_{75}$).*
-*Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018].*
-
-*Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].*
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`), is likely driven by cues other than carbon limitation, such as daylength<!--REF?--> or sink limitation[@zani_increased_2020].
+Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with of drought stress[@dorangeville_drought_2018].
+Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analaysis.
 
 <!-- it remains unclear where extra carbon goes (1 par, maybe 2)-->
-Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
-However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
-We show that the extra C does not go into woody growth, raising the question of where this C goes.
+Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig.
+`r tree_rings`) than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
+However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
+We show that this extra C sequestration in years with warm springs does not substantially augment woody growth, raising the question of where it goes.
 

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion

@@
 Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with of drought stress[@dorangeville_drought_2018].
-Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analaysis.
+Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analysis.
 
@@
 *very little correlation between April $T_{max}$ and growth the following year (5 of 142 RP and 3 of 66 DP species-site combinations), refuting the latter possibility*.
-We also cannot rule out the possibility that warm springs promote formation of more carbon-dense cell walls in the current of following year (**coauthors, do any of you know literature related to this?**); however, our results suggest that little of the extra C that may be fixed in these years goes into wood formation.
+We also cannot rule out the possibility that warm springs promote formation of more carbon-dense cell walls in the current of following year (**coauthors, do any of you know literature related to this?**).
 

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion
- cite extended tree-ring figure (issue #79 )
- adjust text on previous spring temperatures (issue #73 )

@@
 EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
-EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
+EDF_tree_rings_TMAX_extended = "Extended Data Figure 6"
 EDF_tree_rings_TMIN = "Extended Data Figure 7"
@@
 We show that this extra C sequestration in years with warm springs does not substantially augment woody growth, raising the question of where it goes.
+There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
 
-There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
-First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
-In fact, [extension of our tree-ring analysis](https://github.com/EcoClimLab/growth_phenology/issues/57) showed a generally positive, although often non-significant, correlation between ring-width and previous year growing season $T_{max}$ in RP trees, indicating this may be a possibility.
-*very little correlation between April $T_{max}$ and growth the following year (5 of 142 RP and 3 of 66 DP species-site combinations), refuting the latter possibility*.
-We also cannot rule out the possibility that warm springs promote formation of more carbon-dense cell walls in the current of following year (**coauthors, do any of you know literature related to this?**).
+First, extra C production in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
+It is possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], resulting in formation of more carbon-dense cell walls in years with warm springs (**coauthors, do any of you know studies that have looked at this?**).
+Extra carbon could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013]. 
+Extension of our tree-ring analysis revealed little correlation between April $T_{max}$ and growth the following year (sig.
+pos.
+correlations for 5/142 RP and 3/66 DP species-site combinations, Fig.
+`r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
+Finally, extra C may be allocated to growth of coarse roots[@zohner_how_2021] or branches<!-- REF?--> and therefore be unapparent in the basal portions of trunks.
+Warm springs could increase woody productivity through any of these mechanisms, yet none are likely to provide substantial, sustained C sinks under warming spring temperatures.
 
-*There may be a shift towards greater belowground allocation[@zohner_how_2021].*
-
-A second possibility is that any additional C fixed during years with warm springs that is not rapidly respired may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
-Stem growth is more sensitive to water deficit than is photosynthesis[@muller_water_2011].
-[This seems to promote the investment of sequestered carbon into root growth in times of drought stress.
-Zoehner et al 2021[@zohner_how_2021] found that the proportion of root-to-stem growth increased under spring warming, suggesting that increased ecosystem sequestration resulting from spring warming may be promoting the growth of below-ground biomass.
-~~However, this study was done under tightly controlled conditions within potted plants and does not represent the environmental conditions of forests.~~](https://github.com/EcoClimLab/growth_phenology/issues/72)
+A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
+Stem growth is more sensitive to the water deficits that may be associated with high temperatures than is photosynthesis[@muller_water_2011], and hot or dry conditions have been observed to shift allocation towards roots[@zohner_how_2021], reproduction[**REFS**], or respiration[**REFS**].
 Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion

@@
 A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
-Stem growth is more sensitive to the water deficits that may be associated with high temperatures than is photosynthesis[@muller_water_2011], and hot or dry conditions have been observed to shift allocation towards roots[@zohner_how_2021], reproduction[**REFS**], or respiration[**REFS**].
-Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
-Notably, turnover of extra C fixed during years with warm springs would often occur with sufficient temporal lag that it would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
-Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a *fairly long* residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
-*BUT SEE @piao_net_2008!! and @barichivich_large-scale_2013 - these are focused on higher latitudes, but shows greater release of C in the fall!*
+Stem growth is more sensitive to the water deficits that may be associated with high temperatures than is photosynthesis[@muller_water_2011], and hot or dry conditions have been observed to shift allocation towards roots[@zohner_how_2021], reproduction[@hacket-pain_drought_2017], or respiration[**REFS**].
+Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], consistent with studies focused on higher-latitude forests that show enhanced C sequestration in years with warm springs to be offset by higher C losses in the autumn[@piao_net_2008; @barichivich_large-scale_2013].
+Notably, though, turnover of extra C fixed during years with warm springs may occur with sufficient temporal lag that it would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
+Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth probably has a residence time of more than a year within the ecosystem; however, the amount and fate of this carbon remains unknown.
 

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion

@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
 However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for leaf development (Fig.
-`r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
+`r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020] and NEE [@keenan_net_2014; @finzi_carbon_2020].
 However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig.
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`), is likely driven by cues other than carbon limitation, such as daylength<!--REF?--> or sink limitation[@zani_increased_2020].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength<!--REF?--> or sink limitation[@zani_increased_2020].
 Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with of drought stress[@dorangeville_drought_2018].
@@
 `r tree_rings`) than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
-However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
+However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
 We show that this extra C sequestration in years with warm springs does not substantially augment woody growth, raising the question of where it goes.
@@
 
-A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002].
-Stem growth is more sensitive to the water deficits that may be associated with high temperatures than is photosynthesis[@muller_water_2011], and hot or dry conditions have been observed to shift allocation towards roots[@zohner_how_2021], reproduction[@hacket-pain_drought_2017], or respiration[**REFS**].
-Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], consistent with studies focused on higher-latitude forests that show enhanced C sequestration in years with warm springs to be offset by higher C losses in the autumn[@piao_net_2008; @barichivich_large-scale_2013].
-Notably, though, turnover of extra C fixed during years with warm springs may occur with sufficient temporal lag that it would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
-Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth probably has a residence time of more than a year within the ecosystem; however, the amount and fate of this carbon remains unknown.
+A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, and production of foliage, fine roots[@zohner_how_2021], or root exudates.
+Most of this carbon would have a shorter residence time within the ecosystem than C allocated to woody productivity[**REF**], <!-- [@trumbore_quantifying_2002].?-->but would likely remain in the ecosystem for more than a year<!--ref-->, such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
+Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth has a multi-year residence time within the ecosystem.
+However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.
 

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
word count

@@
 EDF_climwin = "Extended Data Figure 3"
-EDF_pheno_all = "Extended Data Figures 4-6"
+EDF_pheno_all = "Extended Data Figures 4-5"
 EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
@@
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~2922 words (1613 intro + results, discussion 1319--too long!) and 1 modest + 2 large display items.
+We're currently at ~2562 words (1609 intro + results, discussion 953) and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently ok there)*
@@
 
-<!--1309 words before Krista's edits. Goal: 700-800-->
+<!--Currently 953. Goal: 700-800-->
 

### 2021-10-15 doc/manuscript/Dow_Extended_Data.Rmd
started rearranging SI figures
+ moved captions

@@
 EDF_climwin = "Extended Data Figure 3"
+EDF_pheno_all = "Extended Data Figures 4-5"
 EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
 EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
-EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
-EDF_tree_rings_TMIN = "Extended Data Figure 7"
+EDF_tree_rings_TMIN = "Extended Data Figure 6"
+EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"
 
@@
 tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations to monthly $T_{max}$ in univariate and multivariate analyses."
-
-## figure captions
-
-map_caption <- "**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
-Sampling details are provided in **SI TBALE NAME**" #needs attention!!!
-
-growth_curves_all_caption  <- " **Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at our two focal study sites** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2) ..., with each line representing a year between 2000 and 2018.
-For stem growth, each line represents one year's growth for a given tree, fit with McMahon model.
-" #needs wordsmithing!!!
-
-climwin_caption <- "**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.** Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
-Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
-19- May 21 at SCBI, March 19 - May 07 at HF)." #info subject to change!!!
-
-pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
-Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
-95% credible intervals are represented by bands.
-The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective CTW.
-For each CTW $T_{max}$.
-Modeled values from dendrometer band measurements are represented by dots.
-(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
-CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
-
-pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).
-Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
-Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
-95% credible intervals are represented by bands.
-(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
-CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
-
-pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).
-Relationships between critical temperature window (CTW) $T_min$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
-Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
-95% credible intervals are represented by bands.
-(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59).
-CAPTION NEEDS MORE DETAIL**)" #needs attention!!!
-
-tree_rings_TMIN_caption <- "Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
-Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in the Supplementary Information.
-(**NOTE: Figure still needs some work.
-Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!!
-#info subject to change!!!
 
@@
 
-![`r map_caption`](tables_figures/chronology_map.png)
+![**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
+Sampling details are provided in **SI TBALE NAME**](tables_figures/chronology_map.png)
 
@@
 
-![`r growth_curves_all_caption`](tables_figures/growth_curves_all.png){ width=100% }
+![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at our two focal study sites** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2) ..., with each line representing a year between 2000 and 2018.
+For stem growth, each line represents one year's growth for a given tree, fit with McMahon model.](tables_figures/growth_curves_all.png){ width=100% }
 
@@
 
-![`r climwin_caption`](tables_figures/climwin_figure.png){ width=70% }
+![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.** Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
+Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
+19- May 21 at SCBI, March 19 - May 07 at HF).](tables_figures/climwin_figure.png){ width=70% }
 
@@
 
-![`r pheno_Tmax_climwin_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }
+![**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+95% credible intervals are represented by bands.
+The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective CTW.
+For each CTW $T_{max}$.
+Modeled values from dendrometer band measurements are represented by dots.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }
 
@@
 
-![`r pheno_Tmax_AprilMay_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may.png){ width=90% }
+![**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).
+Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+95% credible intervals are represented by bands.](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may.png){ width=90% }
 
@@
 
-![`r pheno_Tmin_AprilMay_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may_tmin.png){ width=90% }
+![Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures of the current year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+Chronologies are grouped by xylem porosity and ordered by mean April T_max.
+Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmin.png){ width=90% }
 
@@
 
-![`r tree_rings_TMIN_caption`](tables_figures/tree_rings_tmin.png){ width=90% }
+**THIS IS JUST A PLACEHOLDER! Figure needs to be made**
+
+![Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+Chronologies are grouped by xylem porosity and ordered by mean April T_max.
+Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmax.png){ width=50% }

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
started rearranging SI figures
+ moved captions

@@
 EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
-EDF_tree_rings_TMAX_extended = "Extended Data Figure 6"
-EDF_tree_rings_TMIN = "Extended Data Figure 7"
+EDF_tree_rings_TMIN = "Extended Data Figure 6"
+EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"
 

### 2021-10-16 doc/manuscript/Dow_manuscript_Nature.Rmd
work on methods

@@
 
-*We have not yet re-written the methods for Nature. Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.*
+*We have not yet re-written the methods for Nature. Nature guidelines: Full methods, typically not exceeding 3000 words (currently at 1800 words). May be subdivided by bold headings.*
 
@@
 
-Metal dendrometer bands were installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
+Metal dendrometer bands were installed on a total of 941 trees within the SCBI and Harvard Forest ForestGEO plots.
+(Changed to raw # of bands.
+Removed EDT reference because we use # bands after cleaning - CD)
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
@@
 We screened for three classes of mistakes. 
-First, when a weekly measurement was drastically different from previous week and following weeks, this was assumed to be a measurement error and the record was removed.
+First, when a measurement was drastically different from previous and following measurements, this was assumed to be an error and the datapoint was removed.
 Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck and then released.
 In these cases, the full annual record for the tree was removed.
-Third, cases where data were clearly wrong but with unknown causes were removed. If a majority of the data points fell into this class the entire year was removed from the analysis.
+Third, when data points were clearly wrong but with unknown causes they were removed.
+If a majority of the data points fell into this class within a tree-year the entire year was removed from the analysis.
 
@@
 
-After fitting the @mcmahon_general_2015 model, we removed poorly fit models. 
-Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the acceptable range we determined. 
+After fitting the model, we removed tree-years with poor fits. 
+Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the logical range. 
 Modeled fits for tree-years were removed under five conditions:
@@
 and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3). 
-At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others *(SI figure?
-data_cleaning_figure_hf in manuscript folder)*.
-At SCBI a higher proportion of RP trees were removed--the majority in step 4 *(data_cleaning_figure_scbi)*.
+At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
+At SCBI a higher proportion of RP trees were removed--the majority falling under condition 4.
 
@@
 Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
-
-description here: https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description
+This dataset was used to determine timing of greenup, midgreenup, senescence, along with enhanced vegetation index values for each year.
 
@@
 At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
-In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r  EDT_species`).
+In total, we analyzed >900 cores from fourteen species at SCBI and four species at Harvard Forest (`r  EDT_species`).
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018] (**OTHER REFS??**).
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018] (Strange et al.
+2019 (QUAL); Au et al.
+2020 (ACSA); LeBlanc et al Accepted (LITU) ; Maxwell et al.
+2020 (QUAL) ; Matheus and Maxwell 2018 ; Schmidt 2015 (Thesis), Maxwell et al.
+(2020); Maxwell and Harley 2017; Wilde and Maxwell 2018; Maxwell et al.
+2019; Maxwell et al.
+2015 --- *FIND REFS*).
 Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).
@@
 We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990].
-*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time [@cook_calculating_1997].
-Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology [@jones_estimating_1997].
-The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
-the inner 30–50 yr of each record depending on full chronology length) dropped below three trees.
-Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices [@cook_time_1985].
-We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].*
+Chronologies were built using the methods described in Helcoski et al.
+2019[@helcoski_growing_2019].
 We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
@@
 ### Climate {-#Climate} 
-For the dendrometer band analysis, we used climate data from local weather stations. 
-Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020].
+For the dendrometer band analysis we used climate data from local weather stations. 
+Daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020].
 Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
-The R package climpact () was used to plot temperatures for visual inspection. 
+The R package climpact (no publication?) was used to plot temperatures for visual inspection. 
 Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
-(*how were gaps filled for Harvard? Or was the record good? --- Data came "pre-cleaned" as far as I could tell*)
 

### 2021-10-16 RScripts/Create_final_quiltplot.R
Remove chronology that came w/ extra coauthor
@teixeirak can you remake this figure? I realized I left in a chrono which would have required another coauthor

@@ -22,6 +22,7 @@ all_dcc_output_hf$site <- paste0("HF_", all_dcc_output_hf$Species)
 all_dcc_output_other <- read_csv("Results/other_core_corr.csv")#created in Other_quiltplot.R
 all_dcc_output_other$site <- paste0(all_dcc_output_other$Site,"_", all_dcc_output_other$Species)
 all_dcc_output_other <- all_dcc_output_other[,-1]
+all_dcc_output_other <- all_dcc_output_other[!(all_dcc_output_other$site %in% "MO_Flu_CAOV"),]
 all_dcc_output_scbi <- read_csv("Results/scbi_core_corr.csv")#Created in scbi_quiltplot.R
 all_dcc_output_scbi$site <- paste0("SCBI_", all_dcc_output_scbi$Species)
 
@@ -71,9 +72,9 @@ save.plots <- TRUE
 #Forloop to create plots. Cycles through wood types (RP, DP, SP) & clim variables (tmx, tmn)
 #Sorts by average April temp across all years of core data at each site
 #First, create numbering scheme and prepare data
-#NEED TO MANUALLY CHANGE TMX AND TMN IN LINES 76 AND 116. SORTING WILL NOT WORK CORRECTLY IF DONE AUTOMATICALLY
+#NEED TO MANUALLY CHANGE TMX AND TMN IN LINES 77 AND 117. SORTING WILL NOT WORK CORRECTLY IF DONE AUTOMATICALLY
 for(v in climate_variables) {
-  v <- "tmn"
+  v <- "tmx"
   print(v)
   TRW_coord <- TRW_coord[!(duplicated(TRW_coord$Location)),]#removes duplicate locations added by original author of TRW_coord excel sheet
   X <- all.dcc.output_all[all.dcc.output_all$variable %in% v, ]#subset core by clim variable
@@ -113,7 +114,7 @@ X <- X[X$wood_type != "SP",]
 
   X <- X %>%
     ungroup()%>%
-    arrange( desc(wood_type), tmn, site, month_new)#CHANGE THIS IF SWITCHING BETWEEN TMX/TMN
+    arrange( desc(wood_type), tmx, site, month_new)#CHANGE THIS IF SWITCHING BETWEEN TMX/TMN
 
   X$site <- as.factor(X$site)
   X$month <- as.factor(X$month)
@@ -156,6 +157,7 @@ tmx_data_table <- tmx_data_table[!duplicated(tmx_data_table$group),]
[…]

### 2021-10-16 doc/manuscript/Dow_Extended_Data.Rmd
proper knitting of SI

@@
 
-**THIS IS JUST A PLACEHOLDER! Figure needs to be made**
-
-![Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+![Extended Data Figure 7 | **PLACEHOLDER!!!** Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.

### 2021-10-16 doc/manuscript/Dow_manuscript_Nature.Rmd
added ORCIDs
just copied over the ones that are authors on tree-rings paper
discussion, new word count

@@
 
-Cameron Dow^1^
+Cameron Dow^1^ ( Orcid ID : 0000-0002-8365-598X)
 
@@
 
-Erika B. Gonzalez-Akre^1^
+Erika B. Gonzalez-Akre^1^ ( Orcid ID : 0000-0001-8305-6672)
 
@@
 
-Valentine Herrmann^1^
+Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)
 
@@
 
-Sean McMahon^8,10^
+Sean McMahon^8,10^ (Orcid ID : 0000-0001-8302-6908)
 
@@
 
-Neil Pederson^3^
+Neil Pederson^3^ (Orcid ID : 0000-0003-3830-263X)
 
-Alan J. Tepley^1,9^
+Alan J. Tepley^1,9^ ( Orcid ID : 0000-0002-5701-9613)
 
-Kristina J. Anderson-Teixeira^1,10^*
+Kristina J. Anderson-Teixeira^1,10^* ( Orcid ID : 0000-0001-8461-9713)
 
@@
 
-*NOTE: In general, this needs to be condensed.
+*NOTE: We're probably pretty close to the upper length limit.
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~2562 words (1609 intro + results, discussion 953) and 1 modest + 2 large display items.
+We're currently at ~2512 words (1609 intro + results, discussion 903) and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently ok there)*
@@
 
-<!--Currently 953. Goal: 700-800-->
+<!--Currently 903. Goal: 700-800-->
 
@@
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
-In addition to the existence of temporal lags between C uptake and release, which introduce a lot of uncertainty[@trumbore_quantifying_2002], trees are likely to  acclimate *to some extent* to warming temperatures[@zhang_drought_2021; @gessler_way_2020], and on even longer time scales, we are likely to see changes in forest composition and adaptation.
-If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
-*Future research will be important to sorting this out.*
-
-<!--"contributions beyond" (1 par)-->
-Despite remaining uncertainties, our findings contribute substantially to a broader picture of the influence of spring temperatures on forest ecosystem function. 
-Warm spring temperatures result in earlier leaf-out (Fig.
-`r DOY_timing`)[@jeong_phenology_2011], an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs.
-`r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Yet, an earlier start to the growing season is not consistently associated with a longer peak growing season or increased growth (Figs.
-`r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@zhang_drought_2021;@dorangeville_drought_2018;@helcoski_growing_2019].
+As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis).
+Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]--> and, on longer time scales, species adaptations and shifts in forest composition[**REFS**] are
+If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
+Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.
 
 <!--conclusion (1 par)-->
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models (Fig.
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models (Fig.
 `r schematic_summary`)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].

### 2021-10-16 doc/manuscript/Dow_manuscript_Nature.Rmd
incorporated/ deleted "content to incorporate or delete"

@@
 The expectation is that as growing season length increases, net annual carbon sequestered (*i.e.*, net ecosystem exchange, _NEE_) will increase as well[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and this expectation is represented in models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent findings that sink factors can limit annual growth[@zani_increased_2020; @zohner_how_2021] and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018] suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
+However, recent findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021] and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018] suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
 
@@
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength<!--REF?--> or sink limitation[@zani_increased_2020].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015].
 Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with of drought stress[@dorangeville_drought_2018].
@@
 However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-We show that this extra C sequestration in years with warm springs does not substantially augment woody growth, raising the question of where it goes.
+Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009;@xie_predicting_2018]<!--we can cut some of these references-->, with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are resulting in a lengthening of the period between the slowing of stem growth and the cessation of ecosystem C uptake.
+As the extra C sequestration in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
@@
 
-Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
-<!-- lempereur_growth_2015 studied an evergreen-->
-Additionally, Warmer summer and autumn temperatures coupled with drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018].
-As a result, warming may advance or delay autumn senescence depending on timing of warming and water availability, with delays more common across eastern North America [@buermann_widespread_2018].
-
-These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
-They contrast, however, with model expectations[@zani_increased_2020;@buermann_widespread_2018;@keenan_net_2014] that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
-
-Under controlled conditions, spring warming increases biomass accumulation in trees, with a larger proportion of root growth compared to shoot growth[@zohner_how_2021].
+ @fu_variation_2014
 

### 2021-10-16 doc/manuscript/Dow_manuscript_Nature.Rmd
delete outline notes

@@
 
-<!-- how our findings fill knowledge gap (1 par, maybe 2)-->
-
 Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs.
@@
 
-<!-- it remains unclear where extra carbon goes (1 par, maybe 2)-->
 Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig.
@@
 
-<!-- interannual variation ≠ directional change (1 par)-->
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
@@
 
-<!--conclusion (1 par)-->
 As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models (Fig.

### 2021-10-16 doc/manuscript/Dow_manuscript_Nature.Rmd
macroscopic rearrangement of methods
changed order to parallel presentation in main article

@@
 
-## Study sites {-#Study}
+## Dendrometer bands {-#Dendrometer} 
+### Study sites {-#Study}
 
@@
 
-## Data collection and preparation {-#Data} 
-### Dendrometer bands {-#Dendrometer}
+### Data collection and preparation {-#Data}
 
@@
 
-### Leaf phenology
-
-Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
-This dataset was used to determine timing of greenup, midgreenup, senescence, along with enhanced vegetation index values for each year.
-
-### Tree rings {-#Tree} 
-Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
-We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L.
-at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest)
-At SCBI, cores were collected as described in @helcoski_growing_2019. 
-Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019]*(If we're citing a method from the helcoski paper, do we need to do this 'briefly' section?
-I would remove.)*.
-At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
-In total, we analyzed >900 cores from fourteen species at SCBI and four species at Harvard Forest (`r  EDT_species`).
-
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018] (Strange et al.
-2019 (QUAL); Au et al.
-2020 (ACSA); LeBlanc et al Accepted (LITU) ; Maxwell et al.
-2020 (QUAL) ; Matheus and Maxwell 2018 ; Schmidt 2015 (Thesis), Maxwell et al.
-(2020); Maxwell and Harley 2017; Wilde and Maxwell 2018; Maxwell et al.
-2019; Maxwell et al.
-2015 --- *FIND REFS*).
-Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).
-
-All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
-We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990].
-Chronologies were built using the methods described in Helcoski et al.
-2019[@helcoski_growing_2019].
-We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
-
 ### Climate {-#Climate} 
@@
 
-For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020], again via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
+### Leaf phenology
 
-## Analyses
+Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
+This dataset was used to determine timing of greenup, midgreenup, senescence, along with enhanced vegetation index values for each year.
+
 ### Determining critical temperature windows {-#Determining}
@@
 
+## Tree rings {-#Tree} 
+Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
+We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L.
+at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest)
+At SCBI, cores were collected as described in @helcoski_growing_2019. 
+Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019]*(If we're citing a method from the helcoski paper, do we need to do this 'briefly' section?
+I would remove.)*.
+At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
+In total, we analyzed >900 cores from fourteen species at SCBI and four species at Harvard Forest (`r  EDT_species`).
+
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018] (Strange et al.
+2019 (QUAL); Au et al.
+2020 (ACSA); LeBlanc et al Accepted (LITU) ; Maxwell et al.
+2020 (QUAL) ; Matheus and Maxwell 2018 ; Schmidt 2015 (Thesis), Maxwell et al.
+(2020); Maxwell and Harley 2017; Wilde and Maxwell 2018; Maxwell et al.
+2019; Maxwell et al.
+2015 --- *FIND REFS*).
+Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).
+
+All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
+We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990].
+Chronologies were built using the methods described in Helcoski et al.
+2019[@helcoski_growing_2019].
+We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
+
+For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020], again via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
+
 # Acknowledgements {-#Acknowledgements} 

### 2021-10-16 RScripts/Create_final_quiltplot_extended.R
remove curr oct
#79

@@ -22,11 +22,12 @@ all_dcc_output_hf$site <- paste0("HF_", all_dcc_output_hf$Species)
 all_dcc_output_other <- read_csv("Results/other_core_corr_EXTENDED.csv")#created in Other_quiltplot.R
 all_dcc_output_other$site <- paste0(all_dcc_output_other$Site,"_", all_dcc_output_other$Species)
 all_dcc_output_other <- all_dcc_output_other[,-1]
+all_dcc_output_other <- all_dcc_output_other[!(all_dcc_output_other$site %in% "MO_Flu_CAOV"),]
 all_dcc_output_scbi <- read_csv("Results/scbi_core_corr_EXTENDED.csv")#Created in scbi_quiltplot.R
 all_dcc_output_scbi$site <- paste0("SCBI_", all_dcc_output_scbi$Species)
 
 all.dcc.output_all <- rbind(all_dcc_output_other,all_dcc_output_scbi,all_dcc_output_hf)
-
+all.dcc.output_all <- all.dcc.output_all[!(all.dcc.output_all$month %in% "curr.oct"),]
 #Load in clim means
 clim_means <- read_csv("Results/clim_means_all.csv")#Created in other_quiltplot.R
 #Create porosity lists
@@ -70,7 +71,7 @@ save.plots <- TRUE
 #Sorts by average April temp across all years of core data at each site
 #First, create numbering scheme and prepare data
 for(v in climate_variables) {
-  #v <- "tmx"
+  v <- "tmn"
   print(v)
   TRW_coord <- TRW_coord[!(duplicated(TRW_coord$Location)),]#removes duplicate locations added by original author of TRW_coord excel sheet
   X <- all.dcc.output_all[all.dcc.output_all$variable %in% v, ]#subset core by clim variable
@@ -123,7 +124,7 @@ for(v in climate_variables) {
 X <- X[X$wood_type != "SP",]
   X <- X %>%
     ungroup()%>%
-    arrange( desc(wood_type), tmx, site, month_new)
+    arrange( desc(wood_type), tmn, site, month_new)
 
   X$site <- as.factor(X$site)
   X$month <- as.factor(X$month)
@@ -162,12 +163,13 @@ tmx_data_table$Location <- ifelse(tmx_data_table$Location == "Fiddler<U+393C><U+
 
 tmx_data_table <- left_join(tmx_data_table, TRW_coord, by = "Location")
 
[…]

### 2021-10-16 RScripts/HF_quiltplot_extended.R

@@ -15,7 +15,7 @@ path_to_sp_res_chrons <- "C:/Users/world/Documents/GitHub/growth_phenology/Data/
 
 
 #species <-  c("vector_species_you_have", "as they are names in Neils files")
-species <-  c("ACRU", "BEAL", "QURU", "TSCA")
+species <-  c("ACRU", "BEAL","FRAM", "QURU", "TSCA")
 
 climate_variables <- c("tmx", "tmn")
 
@@ -26,6 +26,7 @@ climate_variables <- c("tmx", "tmn")
 
 start.years.sss <-c(ACRU = 1930,#
                     BEAL = 1952,
+                    FRAM = 1901,
                     QURU = 1898,
                     TSCA = 1930) # enter the name in species and the corresponding year the analysis should start at # these dates may have been given by email by Neil Pederson when he created the psecies chrionologies, it is a date at which sss passes a certain threshold sss (know what the threshold is .75, or .8?)
 
@@ -33,6 +34,7 @@ start.years.sss <-c(ACRU = 1930,#
 #                               Species2 = 2010) # enter the name in species and the corresponding year the analysis should stop at
 full.time.frame.end.years <- c(ACRU = 2013,#
                                BEAL = 2012,
+                               FRAM = 2008,
                                QURU = 2013,
                                TSCA = 2013) # enter the name in species and the corresponding year the analysis should stop at
 

### 2021-10-17 doc/manuscript/Dow_manuscript_Nature.Rmd
minor edits

@@
 `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001).
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF).
 Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
@@
 At Harvard Forest, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
+*(This might be better as a table.
+Add to an existing one?
+May be able to add to mean_growth_parameters table)*
 
@@
 Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015].
-Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with of drought stress[@dorangeville_drought_2018].
+Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with drought stress[@dorangeville_drought_2018].
 Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analysis.
@@
 As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis).
-Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]--> and, on longer time scales, species adaptations and shifts in forest composition[**REFS**] are
+Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]--> and, on longer time scales, species adaptations and shifts are likely to alter forest composition[**REFS**].
 If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].

### 2021-10-17 doc/manuscript/Dow_manuscript_Nature.Rmd
minor results edits

@@
 Modeled fits for tree-years were removed under five conditions:
-(1) single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
+(1) single day growth rates were $\ge 2$ 2 standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
 (2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for it's xylem architecture group, year, and site (SCBI = 62, Harvard Forest = 106); 
-(3) tree-years with small or negligible total growth ($\Delta DBH$ < 0.02 mm; SCBI = 0, Harvard Forest = 66); 
+(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.02$ mm; SCBI = 0, Harvard Forest = 66); 
 (4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 181, Harvard Forest = 199); 
@@
 Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
-The R package climpact (no publication?) was used to plot temperatures for visual inspection. 
+The R package climpact <!--(no publication?)--> was used to plot temperatures for visual inspection. 
 Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
@@
 Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
-This dataset was used to determine timing of greenup, midgreenup, senescence, along with enhanced vegetation index values for each year.
+This dataset was used to determine timing of greenup, midgreenup, senescence, and enhanced vegetation index values for each year.
 
@@
 2019; Maxwell et al.
-2015 --- *FIND REFS*).
+2015 --- *FIND REFS - sent by Justin with tree cores*).
 Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
Methods
- removed 3rd-level headings in methods,
- worked on site descriptions
- gave climate numbers from a single ref with source specified

@@
 ## Dendrometer bands {-#Dendrometer} 
-### Study sites {-#Study}
+<!--Study sites -->
 
-Study sites included two temperate forests in the eastern United States, both part of the Forest Global Earth Observatory [ForestGEO; Anderson-Teixeira et al. 2015; Davies et al. 2021].
+Dendrometer band were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
+SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km South of Front Royal, Virginia. 
+The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
+Dominate canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
+The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1° C in January and 24° C in July[@anderson-teixeira_joint_2021].
 
-Our first study site was the forest dynamics plot located at the Smithsonian Conservation Biology Institute 5 km South of Front Royal, Virginia within the Blue Ridge Mountains (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.).
-The forest here was established in the mid-19th century after conversion from agricultural fields. 
-Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), and hickories (*Carya spp.*). 
-The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July (Ref.[@helcoski_growing_2019]).
+Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.) is located near the central Massachusetts town of Petersham. 
+The forest is secondary and mixed age, having re-established around the beginning of the 20th century following agricultural use and significant hurricane damage in 1938. 
+Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
+The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2021].
 
-The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
-Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and significant hurricane damage in 1938. 
-Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
-The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures averaging -12° C in January and 19° C in July [@dorangeville_peak_2021].
-<!-- Is this the right reference? Yes, I got this info from the draft they sent. -->
-
-### Data collection and preparation {-#Data}
+<!--Data collection and preparation -->
 
 Metal dendrometer bands were installed on a total of 941 trees within the SCBI and Harvard Forest ForestGEO plots.
-(Changed to raw # of bands.
-Removed EDT reference because we use # bands after cleaning - CD)
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
@@
 
-### Climate {-#Climate} 
+ <!--Climate -->
 For the dendrometer band analysis we used climate data from local weather stations. 
@@
 
-### Leaf phenology
+<!-- Leaf phenology -->
 
@@
 
-### Determining critical temperature windows {-#Determining}
+<!-- Determining critical temperature windows -->
 
@@
 
-### Hypothesis testing {-#Hypothesis}
+<!--Hypothesis testing -->
 
@@
 
-For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020], again via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
+For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020] via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
 

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
Methods

@@
 Metal dendrometer bands were installed on a total of 941 trees within the SCBI and Harvard Forest ForestGEO plots.
-Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
+Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest (`r EDT_species`).
 Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
 The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
-Across years, the number of bands sampled averaged *~124* (range: 91-138) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
+Across years, the number of bands sampled averaged **~124** (range: 91-138) at SCBI and  **~717** (range: **700-755**) at Harvard Forest. <!--Cam, values in bold need to be verified-->
 In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).
@@
 Measurements were timed to begin before the beginning of spring growth and to continue through the cessation of growth in the fall. 
-At SCBI, the median start date was 4/14, which was adjusted when early leaf-out of understory vegetation was observed, with the earliest start date being March 30, 2020.
-Measurements were continued through to fall leaf senescence, with the median end date being 10/17 and the latest end date November 26, 2012.
-Timing of measurements at Harvard Forest were similar, with the median start date of 4/23 and median end date of 10/30.
-1998 was an anomalous year where initial measurements were taken on 1/5, but not taken again until 4/15.
+At SCBI, the median start date was April 14, which was adjusted when early leaf-out of understory vegetation was observed, with the earliest start date being March 30 (in 2020). 
+Measurements were continued through to fall leaf senescence, with the median end date being October 17 and the latest end date November 26 (2012). 
+Timing of measurements at Harvard Forest were similar, with the median start date of April 23 and median end date of October 30. 
+1998 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15. 
 The latest end date was November 11, 2002.
@@
 In these cases, the full annual record for the tree was removed.
-Third, when data points were clearly wrong but with unknown causes they were removed.
-If a majority of the data points fell into this class within a tree-year the entire year was removed from the analysis.
+Third, when data points deviated substantially from normal growth patterns, but with unknown causes, they were removed.
+If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
 
-We fit the model of @mcmahon_general_2015 to define phenological dates and growth rates (Fig. `r schematic_summary`). 
-This five-parameter logistic model was fitted to dendrometer band data from each tree-year in R: 
+We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define phenological dates and growth rates (Fig. `r schematic_summary`): 
 $$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
@@
 *We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`) occurs on $DOY_{ip}$ only when $\theta = 1$.*
+The model was fit in R **v XX** using the script developed in Ref.l[@mcmahon_general_2015]. <!--Cam, please fill in details/proper citions.-->
 Using the optimized parameters for each tree-year, we then modeled $DBH$ throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
 
-After fitting the model, we removed tree-years with poor fits. 
+After fitting the growth model, we removed tree-years with poor fits. 
 Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the logical range. 
@@
 At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
-At SCBI a higher proportion of RP trees were removed--the majority falling under condition 4.
-
- <!--Climate -->
-For the dendrometer band analysis we used climate data from local weather stations. 
-Daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020].
-Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
-The R package climpact <!--(no publication?)--> was used to plot temperatures for visual inspection. 
-Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
-Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
+At SCBI a higher proportion of RP trees were removed, the majority falling under condition 4.
 
@@
 
-Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
-This dataset was used to determine timing of greenup, midgreenup, senescence, and enhanced vegetation index values for each year.
+Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
+MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019].
+These data were used to determine timing of greenup, midgreenup, senescence, and enhanced vegetation index (EVI2) values for each year. 
+**Ian, I think this could use a bit more detail**
+
+ <!--Climate -->
+Climate data corresponding to the measurement periods at both sites were obtained from local weather stations. 
+Daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
+Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station **(Boose & Gould, 2004; Boose, 2018)** ([GitHub issue #81](https://github.com/EcoClimLab/growth_phenology/issues/81)).
+The R package *climpact* [**REF?**]<!--(no publication?)--> was used to plot temperatures for visual inspection. 
+*Climpact* also identifies readings that were greater than 3 standard deviations away from yearly means. 
+These points were labeled as outliers and removed from the dataset. 
+Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
 

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
flag where I stopped
Methods

@@
 Climate data corresponding to the measurement periods at both sites were obtained from local weather stations. 
-Daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
+For SCBI, weather data were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
 Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station **(Boose & Gould, 2004; Boose, 2018)** ([GitHub issue #81](https://github.com/EcoClimLab/growth_phenology/issues/81)).
@@
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
+For each site, we extracted daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
 
@@
 
-The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@vandepol_identifying_2016]. 
-This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and $\Delta AIC$ of each window.
-Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r  EDT_mean_growth_parameters`).
-*The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
-Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic_summary`), because we were interested in the influence of temperature on the initiation of woody growth.
-We note that other growth parameters would be controlled by different critical temperature windows; for example, at SCBI $\Delta DBH$ responds most strongly to May-July climate [@helcoski_growing_2019].
-Determining the climate variables that most strongly control all parameters is beyond the scope of this analysis.
+The critical temperature window (Fig.
+`r schematic_summary`), defined as the period over which $DOY_{25}$ was most strongly affected by early spring $T_{max}$, was determined using the R package *climwin*[@vandepol_identifying_2016].
+This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and difference in Akaike information criterion corrected for small sample size relative to the null model, $\Delta AICc$, of each window.
+Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_mean_growth_parameters`), selecting the critical temperature window as that with the lowest $\Delta AICc$.
+Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable, as opposed to $T_{min}$.
+We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth,
+
+<!-- KAT stopped review here. -->
 

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
methods

@@
 Both leaf phenology and stem growth were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
-While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-6), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
+While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-5), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig.
@@
 Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable, as opposed to $T_{min}$.
-We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth,
-
-<!-- KAT stopped review here. -->
+We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth.
 
@@
 
-(*This section needs work/ more detail.*)
-
-Correlation between these variables and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
-
-Mixed effects models were used to test the response of growth phenology variables ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_max}$, $g_{max}$, and $\Delta DBH$) to fixed effects of xylem porosity and mean temperature ($T_{max}$, $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
-We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ and $T_{min}$. 
+Correlation between growth parameters ($DOY_{25}$,$DOY_{50}$,$DOY_{75}$,$DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
+`r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
+Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
+Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
+We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
 Models were run using the LME4 package in R [@bates_fitting_2015]. 
-
 This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language @rstanarm_2020.
@@
 -->
+
+<!-- KAT stopped review here. -->
 

### 2021-10-18 doc/manuscript/Dow_Extended_Data.Rmd
Update Dow_Extended_Data.Rmd

@@
 EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
+EDF_tree_ring_figs= "Extended Data Figure 6-7"
 EDF_tree_rings_TMIN = "Extended Data Figure 6"

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
methods: use of April as alternative critical temperature window
#82

@@
 
+To ensure that patterns were robust under an alternative definition of critical temperature window, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
+`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses defining the critical temperature window as the month of April.
+This was consistent with the periods identified by *climwin* for ring- and diffuse-porous species groups at both sites, all of which included all or part of April (`r  EDT_mean_growth_parameters`).
+
 <!--Hypothesis testing -->

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
methods

@@
 
-Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 $$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
-Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to $DBH$ at the beginning and end of the year, respectively. 
+Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to stem diameter at breast height (DBH) at the beginning and end of the year, respectively. 
 $DOY$ are the modeled days of year (*i.e.*, julian days), 
@@
 The model was fit in R **v XX** using the script developed in Ref.l[@mcmahon_general_2015]. <!--Cam, please fill in details/proper citions.-->
-Using the optimized parameters for each tree-year, we then modeled $DBH$ throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
+Using the optimized parameters for each tree-year, we then modeled DBH throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
 
@@
 
-<!-- KAT stopped review here. -->
+## Tree rings {-#Tree} 
+We analyzed tree-ring records for `r n_cores_sites`, including our focal sites.
+All cores had been previously collected, cross-dated, and measured using standard methodology[@stokes_introduction_1968].
 
-## Tree rings {-#Tree} 
-Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
+Dominant tree species were cored at both SCBI[@helcoski_growing_2019;@bourg_initial_2013] and Harvard Forest[@alexander_potential_2019; @dye_comparing_2016; @finzi_carbon_2020] following sampling designs that covered a broad range of DBH.
 We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L.
-at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest)
-At SCBI, cores were collected as described in @helcoski_growing_2019. 
-Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019]*(If we're citing a method from the helcoski paper, do we need to do this 'briefly' section?
-I would remove.)*.
-At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
-In total, we analyzed >900 cores from fourteen species at SCBI and four species at Harvard Forest (`r  EDT_species`).
+at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest).
+In total, we analyzed `r n_cores_focal_sites` cores from `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r  EDT_species`).
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018] (Strange et al.
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018; @maxwell_increased_2017] (**Strange et al.
 2019 (QUAL); Au et al.
@@
 2019; Maxwell et al.
-2015 --- *FIND REFS - sent by Justin with tree cores*).
-Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).
+2015** --- *if someone can give all the DOIs, KAT can easily add to .bib file*).
+Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).
 
-All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
 We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
-Chronologies were built using the methods described in Helcoski et al.
-2019[@helcoski_growing_2019].
+*Chronologies were built using the methods described in Ref.[@helcoski_growing_2019].* 
 We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
 
-For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020] via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
+We obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020] via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
+
+**QUILT PLOT METHODS**
+
+**MULTIVARIATE ANALYSIS METHODS**
 

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
added tree-ring chronology references

@@
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018; @maxwell_increased_2017] (**Strange et al.
-2019 (QUAL); Au et al.
-2020 (ACSA); LeBlanc et al Accepted (LITU) ; Maxwell et al.
-2020 (QUAL) ; Matheus and Maxwell 2018 ; Schmidt 2015 (Thesis), Maxwell et al.
-(2020); Maxwell and Harley 2017; Wilde and Maxwell 2018; Maxwell et al.
-2019; Maxwell et al.
-2015** --- *if someone can give all the DOIs, KAT can easily add to .bib file*).
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018; @maxwell_increased_2017; @wilde_comparing_2018;@strange_comparing_2019;@au_demographic_2020;@matheus_placing_2018;@leblanc_radial_2020;@maxwell_higher_2019;@maxwell_dendroclimatic_2015;@schmidt_325year_2015] (**JUSTIN, please confirm that we got all the appropriate references here.**).
 Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
methods

@@
 
-## Dendrometer bands {-#Dendrometer} 
+## Dendrometer band analysis {-#Dendrometer} 
 <!--Study sites -->
@@
 These data were used to determine timing of greenup, midgreenup, senescence, and enhanced vegetation index (EVI2) values for each year. 
-**Ian, I think this could use a bit more detail**
+(**IAN, I think this could use a bit more detail.**)
 
@@
 Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station **(Boose & Gould, 2004; Boose, 2018)** ([GitHub issue #81](https://github.com/EcoClimLab/growth_phenology/issues/81)).
-The R package *climpact* [**REF?**]<!--(no publication?)--> was used to plot temperatures for visual inspection. 
-*Climpact* also identifies readings that were greater than 3 standard deviations away from yearly means. 
-These points were labeled as outliers and removed from the dataset. 
+The R package *clim.pact*[@benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
@@
 
-Correlation between growth parameters ($DOY_{25}$,$DOY_{50}$,$DOY_{75}$,$DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
+Correlation between growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
 `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
@@
 
-## Tree rings {-#Tree} 
+## Tree-ring analysis {-#Tree} 
 We analyzed tree-ring records for `r n_cores_sites`, including our focal sites.
@@
 We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
-*Chronologies were built using the methods described in Ref.[@helcoski_growing_2019].* 
+*Chronologies were built using the methods described in Ref.[@helcoski_growing_2019].* (**(NEIL/JUSTIN, is this accurate, and do you think it's sufficient?
+If not, could you please provide a brief summary?**)
 We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
@@
 
-**QUILT PLOT METHODS**
+<!--QUILT PLOT METHODS-->
+Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v.**##** (R Core Team, **YEAR**), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
+We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
+To test for potential lag effects effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
 
+<!--MULTIVARIATE ANALYSIS METHODS-->
 **MULTIVARIATE ANALYSIS METHODS**

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
methods, misc

@@
 To test for potential lag effects effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
+Correlations and significance levels for months April-August are given in **APPENDIX**.
 
@@
 **MULTIVARIATE ANALYSIS METHODS**
+We used a multivariate model to test for joint effects of April and summer $T_{max}$ on RWI.
+We began by testing univariate correlations of $T_{max}$ over three summer windows: June, June-July, and May-August. 
+Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using **Function** in R.
+Slopes and p-values for each chronology are given in **APPENDIX**.
 
@@
 Funding: ForestGEO
+**COAUTHORS, PLEASE ADD.**
 
 # Author Contributions {-#Author}
+
+**CAM, please draft this before sending around.**
 

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
fill in references, misc cleanup

--- 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models (Fig.
-`r schematic_summary`)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig.
-`r tree_rings`), which constitutes a positive feedback to climate change.
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures, which constitutes a positive feedback to climate change.
 
@@
 
-*We have not yet re-written the methods for Nature. Nature guidelines: Full methods, typically not exceeding 3000 words (currently at 1800 words). May be subdivided by bold headings.*
+<!--Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.-->
+<!--We are currently at ~2000 words -->
 
@@
 For SCBI, weather data were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
-Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station **(Boose & Gould, 2004; Boose, 2018)** ([GitHub issue #81](https://github.com/EcoClimLab/growth_phenology/issues/81)).
-The R package *clim.pact*[@benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
+Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station **(Boose & Gould, 2004; Boose, 2018)** (**CAM, we need these references.
+And see [GitHub issue #81](https://github.com/EcoClimLab/growth_phenology/issues/81)**).
+The R package *clim.pact*[@climpact_REF; @benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
@@
 Models were run using the LME4 package in R [@bates_fitting_2015]. 
-This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language @rstanarm_2020.
+This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020].
 In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults.
@@
 
-We obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020] via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeo_2020].
+We obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
 
@@
 To test for potential lag effects effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
+Correlations and significance levels for months April-August are given in **APPENDIX**.
 
 <!--MULTIVARIATE ANALYSIS METHODS-->
-**MULTIVARIATE ANALYSIS METHODS**
+We used a multivariate model to test for joint effects of April and summer $T_{max}$ on RWI.
+We began by testing univariate correlations of $T_{max}$ over three summer windows: June, June-July, and May-August. 
+Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using **(Function)** in R.
+Slopes and p-values for each chronology are given in **APPENDIX**.
 
 # Acknowledgements {-#Acknowledgements} 
-people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data)
-Funding: ForestGEO
+- people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data)
+
+- Funding: ForestGEO
+
+**COAUTHORS, PLEASE ADD.**
 
 # Author Contributions {-#Author}
+
+**(CAM, please draft this before sending around.)**
 

### 2021-10-19 doc/manuscript/Dow_manuscript_Nature.Rmd
First go at author contributions

@@
 For SCBI, weather data were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
-Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station **(Boose & Gould, 2004; Boose, 2018)** (**CAM, we need these references.
-And see [GitHub issue #81](https://github.com/EcoClimLab/growth_phenology/issues/81)**).
+Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station[@boose_shaler_2021-1; @boose_fisher_2021-2].
+Climate data for Harvard Forest were complete and used as is.
 The R package *clim.pact*[@climpact_REF; @benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
@@
 <!--QUILT PLOT METHODS-->
-Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v.**##** (R Core Team, **YEAR**), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
+Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v.1.3.1093 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
 We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
@@
 We began by testing univariate correlations of $T_{max}$ over three summer windows: June, June-July, and May-August. 
-Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using **(Function)** in R.
+Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using the base lm() function in R.
 Slopes and p-values for each chronology are given in **APPENDIX**.
@@
 
-**(CAM, please draft this before sending around.)**
+All authors gave suggestions on improvements to manuscript during development.
+
+Cameron Dow^1^ ( Orcid ID : 0000-0002-8365-598X)
+
+Albert Y. Kim - Developed and assisted in coding the statistical framework for analyses used.
+
+Loïc D’Orangeville - Provided tree core chronologies for additional sites, gave feedback on draft of manuscript.
+
+Erika B. Gonzalez-Akre - Oversees and manages data collection at SCBI.
+
+Ryan Helcoski - Collected and measured cores at SCBI.
+
+Valentine Herrmann - Assisted with coding and interpretation of statistical results.
+
+Grant L. Harley - Provided tree core chronologies for additional sites and assisted with interpretation of results.
+
+Justin T. Maxwell - Provided tree core chronologies for additional sites, detrended crns, and assisted with interpretation of results.
+
+Ian R. McGregor - Assisted with the acquisition and processing of leaf phenology data.
+
+William McShea -
+
+Sean McMahon -
+
+David A. Orwig -
+
+Neil Pederson - Provided tree core chronologies for additional sites and gave feedback on tree core analysis
+
+Alan J. Tepley -
+
+Kristina J. Anderson-Teixeira - Oversaw all aspects of research and manuscript development.
 

### 2021-10-19 RScripts/Pheno_tsensitivity_figure_april_april_tmax.R
change may window to april window
#82

@@ -98,12 +98,12 @@ weatherdata <-
 #  read.csv("results/Climwin_results/Weekly/SCBI/weekly_climwin_results_SCBI_TMAX.csv") %>%
 #  filter(wood_type != "other") %>%
 #  mutate(
-    #median_windowopendate = as.Date(median_windowopendate),
-    #median_windowclosedate = as.Date(median_windowclosedate),
-    #opendoy = yday(median_windowopendate),
-    #closedoy = yday(median_windowclosedate)
-    #winopen = as.Date(refwoy * 7 - winopenwoy * 7, origin = paste0("2011-01-01")),
-    #winclose = as.Date(refwoy * 7 - winclosewoy * 7, origin = paste0("2011-01-01")),
+#median_windowopendate = as.Date(median_windowopendate),
+#median_windowclosedate = as.Date(median_windowclosedate),
+#opendoy = yday(median_windowopendate),
+#closedoy = yday(median_windowclosedate)
+#winopen = as.Date(refwoy * 7 - winopenwoy * 7, origin = paste0("2011-01-01")),
+#winclose = as.Date(refwoy * 7 - winclosewoy * 7, origin = paste0("2011-01-01")),
 #    winopen = as.Date(paste(refwoy-winopenwoy, 1, sep="-"), "%U-%u"),
 #    winclose = as.Date(paste(refwoy-winclosewoy, 1, sep="-"), "%U-%u"),
 #    opendoy = yday(winopen),
@@ -123,11 +123,11 @@ weatherdata <-
 # winclose <- dpcross[i,2]
 
 
-  #climwinmeans_dp <- weatherdata %>%
-  #  filter(doy %in% c(as.numeric(winopen[1,1]):as.numeric(winclose[1,1]))) %>% # 68:135
-  #  group_by(year) %>%
-  #  summarize(climwinmean = mean(cleantmax)) %>%
-  #  mutate(wood_type = "diffuse-porous")
+#climwinmeans_dp <- weatherdata %>%
+#  filter(doy %in% c(as.numeric(winopen[1,1]):as.numeric(winclose[1,1]))) %>% # 68:135
+#  group_by(year) %>%
+#  summarize(climwinmean = mean(cleantmax)) %>%
+#  mutate(wood_type = "diffuse-porous")
 #min <- min(climwinmeans_dp$climwinmean)
 #max <- max(climwinmeans_dp$climwinmean)
 #df <- data.frame(min,max)
@@ -689,14 +689,14 @@ weatherdata_hf <-
[…]

### 2021-10-19 RScripts/Pheno_tsensitivity_figure_april_may_tmax.R

@@ -1,1474 +1,0 @@
-# Setup ------------------------------------------------------------------------
-#
-# To get a quick overview of the sections of this code, go to RStudio menu bar ->
-# Edit -> Folding -> Collapse all.
-#
-
-library(tidyverse)
-library(lubridate)
-library(tidybayes)
-library(patchwork)
-library(gridExtra)
-library(knitr)
-library(scales)
-library(broom.mixed)
-library(tictoc)
-
-# rstanarm stuff
-options(mc.cores = parallel::detectCores())
-library(rstanarm)
-library(patchwork)
-# Number of MCMC chains & number of simulations per chain.
-# Need to increase this at the end
-n_iter <- 10000
-n_chains <- 2
-
-# Objects to keep during memory management clean-up
-objects_to_keep <- c(
-  "objects_to_keep", "n_iter", "n_chains",
-  "fig6_RP", "fig6_DP", "fig6_RP_hf", "fig6_DP_hf",
-  "fig6_RP_mrdoy", "fig6_DP_mrdoy", "fig6_RP_mrdoy_hf", "fig6_DP_mrdoy_hf",
-  "fig6_RP_sl", "fig6_DP_sl", "fig6_RP_sl_hf", "fig6_DP_sl_hf",
-  "fig6_RP_mr", "fig6_DP_mr", "fig6_RP_mr_hf", "fig6_DP_mr_hf",
-  "fig6_RP_tot", "fig6_DP_tot", "fig6_RP_tot_hf", "fig6_DP_tot_hf",
-  "woodtable","Wood_pheno_table_RP","Wood_pheno_table_DP","Wood_pheno_table_RP_hf","Wood_pheno_table_DP_hf", "woodtable_hf",
-  "predictions_RP", "predictions_DP"
-)
-
-# Start timer
-tic()
-
-
-
-# 1. Run analysis on SCBI ---------------------------------------------------------
-## Get growth data --------------------------------------------------------------
-Wood_pheno_table <- read_csv("Data/Wood_pheno_table_SCBI_CLEAN.csv") %>%
-  # Keep only RP and DP for now
-  filter(wood_type != "other") %>%
-  # filter(tot >= 1) %>%
-  # filter(tot <= 12.06)%>%
-  # Rename ring porous to not have a space
-  mutate(wood_type = ifelse(wood_type == "ring porous", "ring-porous", wood_type))
-
[…]

### 2021-10-19 RScripts/Pheno_tsensitivity_figure_april_april_tmax.R
Update Pheno_tsensitivity_figure_april_april_tmax.R

@@ -83,32 +83,32 @@ twosevenfive <- twosevenfive[, c(3, 7, 16)]
 #8. Change windowopen and windowclose values in rp & dp subsets (lines 149 & 167)
 ################################################################
 
-#weatherdata <-
-#  read_csv("climate data/met_tower_data_sensor2_ncdc_supplemented.csv") %>%
-#  filter(!is.na(cleantmax)) %>%
-#  mutate(year = year.x)
-#TMIN
 weatherdata <-
-  read_csv("climate data/SCBI_tmin.csv") %>%
-  filter(!is.na(TMIN))
+  read_csv("climate data/met_tower_data_sensor2_ncdc_supplemented.csv") %>%
+  filter(!is.na(cleantmax)) %>%
+  mutate(year = year.x)
+#TMIN
+#weatherdata <-
+#  read_csv("climate data/SCBI_tmin.csv") %>%
+#  filter(!is.na(TMIN))
 
 # 2. Get climwin data
 #TMAX
-#climwindows <-
-#  read.csv("results/Climwin_results/Weekly/SCBI/weekly_climwin_results_SCBI_TMAX.csv") %>%
-#  filter(wood_type != "other") %>%
-#  mutate(
-#median_windowopendate = as.Date(median_windowopendate),
-#median_windowclosedate = as.Date(median_windowclosedate),
-#opendoy = yday(median_windowopendate),
-#closedoy = yday(median_windowclosedate)
-#winopen = as.Date(refwoy * 7 - winopenwoy * 7, origin = paste0("2011-01-01")),
-#winclose = as.Date(refwoy * 7 - winclosewoy * 7, origin = paste0("2011-01-01")),
-#    winopen = as.Date(paste(refwoy-winopenwoy, 1, sep="-"), "%U-%u"),
-#    winclose = as.Date(paste(refwoy-winclosewoy, 1, sep="-"), "%U-%u"),
-#    opendoy = yday(winopen),
-#    closedoy = yday(winclose)
-#  )
+climwindows <-
+  read.csv("results/Climwin_results/Weekly/SCBI/weekly_climwin_results_SCBI_TMAX.csv") %>%
+  filter(wood_type != "other") %>%
+  mutate(
+    #median_windowopendate = as.Date(median_windowopendate),
+    #median_windowclosedate = as.Date(median_windowclosedate),
[…]

### 2021-10-20 doc/manuscript/Dow_Extended_Data.Rmd
change patchwork figure to april_april version
knit in updated figures

@@
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
-95% credible intervals are represented by bands.](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may.png){ width=90% }
+95% credible intervals are represented by bands.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }
 
@@
 
-![Extended Data Figure 7 | **PLACEHOLDER!!!** Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+![Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmax.png){ width=50% }
+Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmax_extended.png){ width=90% }

### 2021-10-20 doc/manuscript/Dow_manuscript_Nature.Rmd
methods edits

@@
 However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009;@xie_predicting_2018]<!--we can cut some of these references-->, with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are resulting in a lengthening of the period between the slowing of stem growth and the cessation of ecosystem C uptake.
+Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are resulting in a lengthening of the period between the slowing of stem growth and the cessation of ecosystem C uptake.
 As the extra C sequestration in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
@@
 As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis).
-Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]--> and, on longer time scales, species adaptations and shifts are likely to alter forest composition[**REFS**].
+Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]...I'm not sure i've seen much describing trees acclimating to warming temps, only the gessler paper about drought.--> and, on longer time scales, species adaptations and shifts are likely to alter forest composition[**REFS**].
 If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
@@
 The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
-Across years, the number of bands sampled averaged **~124** (range: 91-138) at SCBI and  **~717** (range: **700-755**) at Harvard Forest. <!--Cam, values in bold need to be verified-->
+Across years, the number of bands sampled averaged 129 (range: 91-138) at SCBI and 717 (range: 700-755) at Harvard Forest. <!--verified - these are the pre-cleaning numbers-->
 In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).
@@
 and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote. 
-*We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`) occurs on $DOY_{ip}$ only when $\theta = 1$.*
-The model was fit in R **v XX** using the script developed in Ref.l[@mcmahon_general_2015]. <!--Cam, please fill in details/proper citions.-->
-Using the optimized parameters for each tree-year, we then modeled DBH throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
+*We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ only when $\theta = 1$.*
+The model was fit in R v 1.3.1093 using the functions developed in [@mcmahon_general_2015].
+These functions take the timeseries of manual dendrometer band measurements and return the optimized parameters to predict DBH throughout the year.
+We then modeled DBH throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
 
@@
 
-Sean McMahon -
+Sean McMahon - Provided R functions for dealing with dendroband data (RDendrom)
 

### 2021-10-20 RScripts/Pheno_tsensitivity_figure_april_april_tmax.R
correct placeholder figure

@@ -1223,7 +1223,7 @@ timer$toc - timer$tic
 
 ## Create single figure using patchwork ----------
 png(
-  filename = "doc/manuscript/tables_figures/pheno_Tsensitivity_combo_patchwork_april_april.png", width = 24, height = 20,
+  filename = "doc/manuscript/tables_figures/pheno_Tsensitivity_combo_patchwork_april.png", width = 24, height = 20,
   pointsize = 20, bg = "transparent", units = "in", res = 600
   #restoreConsole = FALSE
 )

### 2021-10-20 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion edits

@@
 `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
-Finally, extra C may be allocated to growth of coarse roots[@zohner_how_2021] or branches<!-- REF?--> and therefore be unapparent in the basal portions of trunks.
+Finally, extra C may be allocated to growth of coarse roots[@zohner_how_2021] or branches[**REF?**]<!-- REF?--> and therefore be unapparent in the basal portions of trunks.
 Warm springs could increase woody productivity through any of these mechanisms, yet none are likely to provide substantial, sustained C sinks under warming spring temperatures.

### 2021-10-20 doc/manuscript/Dow_manuscript_Nature.Rmd
methods edit

@@
 A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, and production of foliage, fine roots[@zohner_how_2021], or root exudates.
-Most of this carbon would have a shorter residence time within the ecosystem than C allocated to woody productivity[**REF**], <!-- [@trumbore_quantifying_2002].?-->but would likely remain in the ecosystem for more than a year<!--ref-->, such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
+Most of this carbon would have a shorter residence time within the ecosystem than C allocated to woody productivity but may remain in the ecosystem for more than a year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
 Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth has a multi-year residence time within the ecosystem.
@@
 As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis).
-Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]...I'm not sure i've seen much describing trees acclimating to warming temps, only the gessler paper about drought.--> and, on longer time scales, species adaptations and shifts are likely to alter forest composition[**REFS**].
+Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]--> and, on longer time scales, species adaptations and shifts in community composition[**REFS**] are likely to alter the phenology of forest C cycling.
 If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
@@
 For SCBI, weather data were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
-Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station[@boose_shaler_2021-1; @boose_fisher_2021-2].
-Climate data for Harvard Forest were complete and used as is.
 The R package *clim.pact*[@climpact_REF; @benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
@@
 For each site, we extracted daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
+Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station[@boose_shaler_2021-1; @boose_fisher_2021-2]. **(Cam, please be sure to sync these in Zotero.)**
+Climate data for Harvard Forest were complete and used as is.
 

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
author contributions
@camerondow35 , this doesn't need to be super specific. I condensed into a typical list.

@@
 
-All authors gave suggestions on improvements to manuscript during development.
-
-Cameron Dow^1^ ( Orcid ID : 0000-0002-8365-598X)
-
-Albert Y. Kim - Developed and assisted in coding the statistical framework for analyses used.
-
-Loïc D’Orangeville - Provided tree core chronologies for additional sites, gave feedback on draft of manuscript.
-
-Erika B. Gonzalez-Akre - Oversees and manages data collection at SCBI.
-
-Ryan Helcoski - Collected and measured cores at SCBI.
-
-Valentine Herrmann - Assisted with coding and interpretation of statistical results.
-
-Grant L. Harley - Provided tree core chronologies for additional sites and assisted with interpretation of results.
-
-Justin T. Maxwell - Provided tree core chronologies for additional sites, detrended crns, and assisted with interpretation of results.
-
-Ian R. McGregor - Assisted with the acquisition and processing of leaf phenology data.
-
-William McShea -
-
-Sean McMahon - Provided R functions for dealing with dendroband data (RDendrom)
-
-David A. Orwig -
-
-Neil Pederson - Provided tree core chronologies for additional sites and gave feedback on tree core analysis
-
-Alan J. Tepley -
-
-Kristina J. Anderson-Teixeira - Oversaw all aspects of research and manuscript development.
+Cameron Dow and Kristina J. Anderson-Teixeira conceived the ideas and designed the study; 
+Cameron Dow, Loïc D’Orangeville, Erika B.
+Gonzalez-Akre, Ryan Helcoski, Grant L.
+Harley, Justin T.
+Maxwell, Ian R.
+McGregor, William McShea, David A.
+Orwig, Neil Pederson, Alan J.
+Tepley, and Kristina J.
+Anderson-Teixeira collected or oversaw collection of data;
+Cameron Dow, Albert Y. Kim, Valentine Herrmann, Justin T. Maxwell, Ian R. McGregor, Sean McMahon analyzed the data or provided analytical tools; 
+Cameron Dow and Kristina J. Anderson-Teixeira led the writing of the manuscript. 
+All authors contributed critically to the drafts and gave final approval for publication.
 

### 2021-10-21 doc/manuscript/Dow_Extended_Data.Rmd
proper citation of supplementary info

@@
 EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"
+
+# Supplementary info
+chron_table= "SI Table 1"
 
@@
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmin.png){ width=90% }
+Chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmin.png){ width=90% }
 
@@
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmax_extended.png){ width=90% }
+Chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=90% }

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd

@@
 Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature.
-Chronology details are given in the Supplementary Information.](tables_figures/tree_rings_tmax.png){ width=100% }
+Chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
 
@@
 To test for potential lag effects effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
-Correlations and significance levels for months April-August are given in **APPENDIX**.
+Correlations and significance levels for months April-August are given in `r chron_table`.
 
@@
 Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using the base lm() function in R.
-Slopes and p-values for each chronology are given in **APPENDIX**.
+Slopes and p-values for each chronology are given in `r chron_table`.
 

### 2021-10-21 doc/manuscript/Dow_Extended_Data.Rmd
added temperature responses to Table S2
@camerondow35 , there's space to fill in numbers for additional parameters. It would also be good to fill in standard error on the slopes, if you have that. Let's list these only for significant relationships.

@@
 EDT_species="Extended Data Table 1"  
-EDT_mean_growth_parameters="Extended Data Table 2" 
+EDT_phenology_parameters="Extended Data Table 2" 
 EDT_tree_ring_summary="Extended Data Table 3"
@@
 library(kableExtra)
-mean_growth_parameters <- read.csv("tables_figures/mean_growth_parameters.csv",  col.names = c("", "ring","diffuse","ring", "diffuse"), stringsAsFactors = FALSE, check.names = FALSE)
+mean_growth_parameters <- read.csv("tables_figures/phenology_parameters.csv",  col.names = c("", "ring","diffuse","ring", "diffuse"), stringsAsFactors = FALSE, check.names = FALSE)
 kable(mean_growth_parameters, caption = mean_growth_parameters_caption ,  format = "latex", booktabs = TRUE,  escape = F) %>%
@@
     pack_rows("Stem Growth", 1, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-    pack_rows("Leaf Phenology (ecosystem level)", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-    #add_indent(9:12)
+    pack_rows("Canopy Foliage (ecosystem level)", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    pack_rows("Phenological sensitivity to mean max. T in critical T window (days/°C)", 13, 22, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
+    add_indent(14:17)%>%
+    add_indent(19:22)%>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
added temperature responses to Table S2
@camerondow35 , there's space to fill in numbers for additional parameters. It would also be good to fill in standard error on the slopes, if you have that. Let's list these only for significant relationships.

@@
 Both leaf phenology and stem growth milestones occurred earlier at SCBI than at Harvard Forest, with stem growth milestones achieved on average 6-10 days earlier at SCBI (Fig.
-`r DOY_timing`, `r  EDT_mean_growth_parameters`).
+`r DOY_timing`, `r  EDT_phenology_parameters`).
 Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
-`r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_mean_growth_parameters`).
-Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_mean_growth_parameters`).
+`r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
+Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
 
@@
 `r schematic_summary`).
-The strongest relationships, including but not limited to the critical temperature window (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).
+The strongest relationships, including but not limited to the critical temperature window (`r  EDT_phenology_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).
 
@@
 In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest).
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_phenology_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest).
 $\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`). 
@@
 This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and difference in Akaike information criterion corrected for small sample size relative to the null model, $\Delta AICc$, of each window.
-Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_mean_growth_parameters`), selecting the critical temperature window as that with the lowest $\Delta AICc$.
+Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`), selecting the critical temperature window as that with the lowest $\Delta AICc$.
 Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable, as opposed to $T_{min}$.
@@
 `r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses defining the critical temperature window as the month of April.
-This was consistent with the periods identified by *climwin* for ring- and diffuse-porous species groups at both sites, all of which included all or part of April (`r  EDT_mean_growth_parameters`).
+This was consistent with the periods identified by *climwin* for ring- and diffuse-porous species groups at both sites, all of which included all or part of April (`r  EDT_phenology_parameters`).
 

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
removed details on phenological advances, now in Table S2.
@camerondow35 , good suggestion!

@@
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF).
-Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
+Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig.
+`r DOY_timing`, `r EDT_phenology_parameters`).
 Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
-Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
-At Harvard Forest, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
-*(This might be better as a table.
-Add to an existing one?
-May be able to add to mean_growth_parameters table)*
+Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 1.1-1.9 days/ $^\circ$C for ring-porous species and 3.5-3.6 days/ $^\circ$C for diffuse-porous species at SCBI, and 2.8-7.2 days/ $^\circ$C for ring-porous species and 6.6-7.9 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).
 

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
"leaf"--> "canopy foliage"
@camerondow35 , I changed this wording throughout because it gives a more precise and accurate definition.

You might also change the y-axis on Fig. 2 to "Canopy Foliage Stage"

@@
 
-While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
+While the responses of canopy foliage phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021].
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology (measured at ecosystem level via remote sensing).
+This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of canopy foliage phenology (measured at ecosystem level via remote sensing).
 
@@
 
-Both leaf phenology and stem growth milestones occurred earlier at SCBI than at Harvard Forest, with stem growth milestones achieved on average 6-10 days earlier at SCBI (Fig.
+Both canopy foliage phenology and stem growth milestones occurred earlier at SCBI than at Harvard Forest, with stem growth milestones achieved on average 6-10 days earlier at SCBI (Fig.
 `r DOY_timing`, `r  EDT_phenology_parameters`).
@@
 
-![**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
@@
 
-Both leaf phenology and stem growth were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
+Both canopy foliage and stem growth phenology were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
 While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-5), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
@@
 `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF).
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of canopy greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF).
 Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig.
@@
 
-Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig.
+Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig.
 `r DOY_timing`), there was no consistent lengthening of $L_{PGS}$.
@@
 `r schematic_summary`- `r tree_rings`).
-The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for leaf development (Fig.
+The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
 `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020] and NEE [@keenan_net_2014; @finzi_carbon_2020].
@@
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn canopy foliage senesence[@zani_increased_2020;@keenan_timing_2015].
 Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with drought stress[@dorangeville_drought_2018].
@@
 
-Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
+canopy foliage phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
 MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019].

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
#84

@@
 
-While the responses of canopy foliage phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
+While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021].
@@
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn canopy foliage senesence[@zani_increased_2020;@keenan_timing_2015].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015].
 Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with drought stress[@dorangeville_drought_2018].
@@
 
-A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, and production of foliage, fine roots[@zohner_how_2021], or root exudates.
-Most of this carbon would have a shorter residence time within the ecosystem than C allocated to woody productivity but may remain in the ecosystem for more than a year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
+A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, fine roots[@zohner_how_2021], or root exudates.
+Much of this carbon would have a relatively short residence time within the ecosystem, and C loss though fall or winter respiration may offset gains from an earlier spring [@oishi_warmer_2018; @buermann_widespread_2018].
+However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
 Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth has a multi-year residence time within the ecosystem.
@@
 
-canopy foliage phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
+Canopy foliage phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
 MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019].

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
working on references
#84

@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
 However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
@@
 These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
-Thus, in contrast to global carbon cycle model expectations[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014], warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
+Thus, in contrast to global carbon cycle model expectations[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
 
@@
 
-<!--Currently 903. Goal: 700-800-->
+<!--Currently 1005. Goal: 700-800-->
 
@@
 The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020] and NEE [@keenan_net_2014; @finzi_carbon_2020].
+`r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
 However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig.
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015].
-Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with drought stress[@dorangeville_drought_2018].
-Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analysis.
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
+Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of a lengthened growing season[@zhang_drought_2021].
+Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.
 
@@
 It is possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], resulting in formation of more carbon-dense cell walls in years with warm springs (**coauthors, do any of you know studies that have looked at this?**).
-Extra carbon could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013]. 
+Extra carbon could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
 Extension of our tree-ring analysis revealed little correlation between April $T_{max}$ and growth the following year (sig.
@@
 `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
-Finally, extra C may be allocated to growth of coarse roots[@zohner_how_2021] or branches[**REF?**]<!-- REF?--> and therefore be unapparent in the basal portions of trunks.
-Warm springs could increase woody productivity through any of these mechanisms, yet none are likely to provide substantial, sustained C sinks under warming spring temperatures.
+Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.
 
-A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, fine roots[@zohner_how_2021], or root exudates.
+A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
 Much of this carbon would have a relatively short residence time within the ecosystem, and C loss though fall or winter respiration may offset gains from an earlier spring [@oishi_warmer_2018; @buermann_widespread_2018].
@@
 Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]--> and, on longer time scales, species adaptations and shifts in community composition[**REFS**] are likely to alter the phenology of forest C cycling.
-If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
+If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
 Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures, which constitutes a positive feedback to climate change.
-
-# content to incorporate or delete:
-
- @fu_variation_2014
 

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
earlywood/ latewood references
@cdow, FYI. cool stuff!

@@
 First, extra C production in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
-It is possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], resulting in formation of more carbon-dense cell walls in years with warm springs (**coauthors, do any of you know studies that have looked at this?**).
+It is theoretically possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more carbon-dense wood in years with warm springs.
+However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs total C content of annual rings is unlikely.
 Extra carbon could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].

### 2021-10-22 doc/manuscript/Dow_manuscript_Nature.Rmd
Added "Peak"

@@
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
-Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
+Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
 

### 2021-11-03 doc/manuscript/Dow_manuscript_Nature.Rmd
update citations

@@
 For each site, we extracted daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
-Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station[@boose_shaler_2021-1; @boose_fisher_2021-2]. **(Cam, please be sure to sync these in Zotero.)**
+Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station[@boose_shaler_2021; @boose_fisher_2021]. 
 Climate data for Harvard Forest were complete and used as is.

### 2021-11-04 doc/manuscript/Dow_manuscript_Nature.Rmd
Fix typos caught by coauthors

@@
 
-Loïc D’Orangeville^3,4^
+Loïc D’Orangeville^3,4^ ( Orcid ID : 0000-0001-7841-7082)
 
@@
 
-Justin T. Maxwell^6^
+Justin T. Maxwell^6^ (Orcid ID: 0000-0001-9195-3146)
 
-Ian R. McGregor^1,7^
+Ian R. McGregor^1,7^ (Orcid ID: 0000-0002-5763-021X)
 
@@
 The expectation is that as growing season length increases, net annual carbon sequestered (*i.e.*, net ecosystem exchange, _NEE_) will increase as well[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and this expectation is represented in models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021] and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018] suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
+However, recent findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
 
@@
 
-![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant increases and decreases at p < 0.05, respectively, 'n.s.' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 
@@
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
-Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
+Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and < one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
 
 ![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
-Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
+Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
 Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }
@@
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of canopy greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF).
-Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig.
+Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, mid-greenup also advanced but to a lesser extent (Fig.
 `r DOY_timing`, `r EDT_phenology_parameters`).
@@
 `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig.
+However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have a negligible effect on total annual growth for most species and locations (Fig.
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
 Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of a lengthened growing season[@zhang_drought_2021].
@@
 It is theoretically possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more carbon-dense wood in years with warm springs.
-However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs total C content of annual rings is unlikely.
+However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
 Extra carbon could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
@@
 
-Dendrometer band were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
+Dendrometer bands were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
 SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km South of Front Royal, Virginia. 
 The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
-Dominate canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
+Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1° C in January and 24° C in July[@anderson-teixeira_joint_2021].
@@
 The forest is secondary and mixed age, having re-established around the beginning of the 20th century following agricultural use and significant hurricane damage in 1938. 
-Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
+Dominant species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2021].
@@
 The model was fit in R v 1.3.1093 using the functions developed in [@mcmahon_general_2015].
-These functions take the timeseries of manual dendrometer band measurements and return the optimized parameters to predict DBH throughout the year.
+These functions take the time-series of manual dendrometer band measurements and return the optimized parameters to predict DBH throughout the year.
 We then modeled DBH throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
@@
 ## Tree-ring analysis {-#Tree} 
-We analyzed tree-ring records for `r n_cores_sites`, including our focal sites.
-All cores had been previously collected, cross-dated, and measured using standard methodology[@stokes_introduction_1968].
+We analyzed tree-ring records for `r n_cores_sites` sites, including our focal sites.
+All cores had been previously collected, cross-dated, and measured using standard collection and processing methodologies[@stokes_introduction_1968].
 
@@
 We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
-To test for potential lag effects effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
+To test for potential lag effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
 Correlations and significance levels for months April-August are given in `r chron_table`.

### 2021-11-09 doc/manuscript/Dow_manuscript_Nature.Rmd
Coauthor feedback
not done - uploading where im at

@@
 
-Albert Y. Kim^1,2^
+Albert Y. Kim^1,2^ ( Orcid ID : 0000-0001-7824-306X)
 
@@
 As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
-Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
+However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014].
+Using dendrometer band measurements from *465* trees across two deciduous mid-Atlantic forests, we show that warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
 The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
 These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
-Thus, in contrast to global carbon cycle model expectations[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
+Such empirical results contradict current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], and imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
 
@@
 In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
-A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019].
+A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019] *Coauthors suggest removing this but i'm unsure.
+I find this distinction relevant, but if more experienced people disagree I'm fine removing - would like your opinion*.
 The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
 
-In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-The expectation is that as growing season length increases, net annual carbon sequestered (*i.e.*, net ecosystem exchange, _NEE_) will increase as well[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and this expectation is represented in models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
+In temperate deciduous forests, spring warming generally lengthens the growing season, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999], and extends the period where the ecosystem is a net CO~2~ sink[@keenan_net_2014].
+Current models assume that longer growing seasons lead to increasing net annual carbon sequestration (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+However, recent experimental and observational findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
 
-While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
+While the responses of leaf phenology and seasonal *ecosystem (?)* CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021].
@@
 
-Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
+Tree-ring records, which can be used to examine correlations of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
 <!-- dorangeville_drought_2018 looks at PET.-->
@@
 
-Here, we describe how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
+Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
+To test the assumption that warmer springs lead to an extension of the growing season we used measurements from 465 dendrometer bands across two decidious mid-atlantic forests.
+To test whether spring temperatrues were consistently positively correlated with annual ring-width Index (RWI; tree-ring width which has been standardized for age) we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.
 
@@
 
-Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
-This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of canopy foliage phenology (measured at ecosystem level via remote sensing).
+This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021].
+These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
 
-![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant increases and decreases at p < 0.05, respectively, 'n.s.' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant (determined by whether the posterior credible interval included 0 or not) increases and decreases, respectively, 'n.s.' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 
@@
 `r DOY_timing`, `r  EDT_phenology_parameters`).
-Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
+Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
@@
 
-![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
@@
 
-Both canopy foliage and stem growth phenology were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
-While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-5), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan.
-1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig.
-`r schematic_summary`).
-The strongest relationships, including but not limited to the critical temperature window (`r  EDT_phenology_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).
+Both MODIS-derived canopy foliage phenology and dendrometerband measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig.
+`r DOY_timing`, `r EDF_pheno_all`).
 
-For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs.
-`r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of canopy greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF).
-Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, mid-greenup also advanced but to a lesser extent (Fig.
+We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the Spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW).
+CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 and $DOY_{25}$ (`r EDF_climwin`).
+The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.
+
+For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`),
+which is consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014]
+Specifically, greenup *(DOY when EVI2 first crossed 15% of the segment EVI2 amplitude)* DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, while mid-greenup also advanced but to a lesser extent (Fig.
 `r DOY_timing`, `r EDT_phenology_parameters`).
@@
 Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig.
-`r DOY_timing`), there was no consistent lengthening of $L_{PGS}$.
-Specifically, because both $DOY_{25}$ and $DOY_{75}$ advanced in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, `r EDF_pheno_all`).
+`r DOY_timing`), there was no consistent lengthening of $L_{PGS}$ (Fig.
+`r schematic_summary`, `r EDF_pheno_all`).
 
@@
 $\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`).
-Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth.
 
@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
-Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of records: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
+Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of chronologies: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
 In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
-$T_{min}$ generally exerted less influence over annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
+$T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
 
-To test whether a positive effect of spring temperatures might be offset by a negative effect of summer temperatures, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
+To test whether the negative effect of summer temperatures might offset the enhancement of growth by warmer spring temperatrues, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatrues were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
 
@@
 Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
-Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of a lengthened growing season[@zhang_drought_2021].
+Rather than lengthening the growing season, the primary effect of warming temperature on annual growth is a negative correlation with $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of an earlier onset of growth in the spring[@zhang_drought_2021].
 Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.
@@
 
+~Add this comment somewhere in here: "What about mentioning "we converted the trw into a dimensionless growth index (thereafter simply referred to as RWI; see Methods) to emphasize interannual variability associated with climate  (Cook & Kairiukstis, 1990)" or something along those lines?"~
+
 We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 

### 2021-11-11 doc/manuscript/Dow_manuscript_Nature.Rmd
second pass at coauthor feedback

@@
 However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are resulting in a lengthening of the period between the slowing of stem growth and the cessation of ecosystem C uptake.
+Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are leading to a lengthening of the period from peak stem growth to the cessation of ecosystem C uptake.
 As the extra C sequestration in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
@@
 
-First, extra C production in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
+One possibility is that extra C production in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
 It is theoretically possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more carbon-dense wood in years with warm springs.
@@
 A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
-Much of this carbon would have a relatively short residence time within the ecosystem, and C loss though fall or winter respiration may offset gains from an earlier spring [@oishi_warmer_2018; @buermann_widespread_2018].
+Much of this carbon would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring [@oishi_warmer_2018; @buermann_widespread_2018].
 However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
-Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth has a multi-year residence time within the ecosystem.
+Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
 However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.
@@
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
-As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis).
-Moreover, acclimation of trees to warming temperatures[**REFS**] <!--prevsioulsy we had these, but I'm not sure they really work: [@zhang_drought_2021; @gessler_way_2020]--> and, on longer time scales, species adaptations and shifts in community composition[**REFS**] are likely to alter the phenology of forest C cycling.
+As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
+Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[**REFS**] are likely to alter the phenology of forest C cycling.
 If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
@@
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, this is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most climate forecasting  models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures, which constitutes a positive feedback to climate change.
@@
 
-Dendrometer bands were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
+Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
 SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km South of Front Royal, Virginia. 
@@
 
-Metal dendrometer bands were installed on a total of 941 trees within the SCBI and Harvard Forest ForestGEO plots.
+Metal dendrometer bands were installed on 941 trees within the SCBI and Harvard Forest ForestGEO plots.
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest (`r EDT_species`).
 Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
-The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
+The number of bands measured at each site fluctuated slightly as trees were added or dropped (e.g., because of tree mortality) from the census. 
 Across years, the number of bands sampled averaged 129 (range: 91-138) at SCBI and 717 (range: 700-755) at Harvard Forest. <!--verified - these are the pre-cleaning numbers-->
@@
 
-The raw dendrometer band data were manually cleaned by visual inspection before analysis. 
-We screened for three classes of mistakes. 
-First, when a measurement was drastically different from previous and following measurements, this was assumed to be an error and the datapoint was removed.
-Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck and then released.
+The raw dendrometer band data were manually inspected before analysis. 
+We screened the data for three classes of erros. 
+First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and the datapoint was removed.
+Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck on the tree bark and then released.
 In these cases, the full annual record for the tree was removed.
-Third, when data points deviated substantially from normal growth patterns, but with unknown causes, they were removed.
+Third, data points that deviated substantially from normal growth patterns, but had unknown causes, were removed.
 If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
 
-We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define phenological dates and growth rates (Fig. `r schematic_summary`): 
+We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define phenological dates and growth rates (Fig.
+`r schematic_summary`).
+In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; i.e.
+julian days) as:
 $$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
-Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to stem diameter at breast height (DBH) at the beginning and end of the year, respectively. 
-$DOY$ are the modeled days of year (*i.e.*, julian days), 
+Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to DBH at the beginning and end of the year, respectively. 
 $DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
@@
 *We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ only when $\theta = 1$.*
-The model was fit in R v 1.3.1093 using the functions developed in [@mcmahon_general_2015].
-These functions take the time-series of manual dendrometer band measurements and return the optimized parameters to predict DBH throughout the year.
-We then modeled DBH throughout the year and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
+The model was fit in R v 4.0 using the functions developed in the Rdendrom package[@mcmahon_general_2015].
+These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year.
+We then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest detailed in Fig. 1.
 
@@
 Modeled fits for tree-years were removed under five conditions:
-(1) single day growth rates were $\ge 2$ 2 standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
-(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for it's xylem architecture group, year, and site (SCBI = 62, Harvard Forest = 106); 
+(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
+(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 62, Harvard Forest = 106); 
 (3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.02$ mm; SCBI = 0, Harvard Forest = 66); 
@@
 At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
-At SCBI a higher proportion of RP trees were removed, the majority falling under condition 4.
+At SCBI a higher proportion of ring-porous trees were removed, the majority falling under condition 4.
 
@@
 Canopy foliage phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
-MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019].
-These data were used to determine timing of greenup, midgreenup, senescence, and enhanced vegetation index (EVI2) values for each year. 
-(**IAN, I think this could use a bit more detail.**)
+MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019]  via Google Earth Engine.
+Using daily MODIS 2-band Enhanced Vegetation Index data (EVI2) at a spatial resolution of 500m, the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
 
  <!--Climate -->
-Climate data corresponding to the measurement periods at both sites were obtained from local weather stations. 
+
+*ADD SOMETHING ON HOW STANDARD PRECIPITAITON EVAPOTRANSPIRATIN INDEX WAS FOUND*!!!!
+
+For the dendrometer band analysis, climate data corresponding to the measurement periods at both sites were obtained from local weather stations. 
 For SCBI, weather data were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
@@
 
+For the tree core analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
+
 <!-- Determining critical temperature windows -->
@@
 The critical temperature window (Fig.
-`r schematic_summary`), defined as the period over which $DOY_{25}$ was most strongly affected by early spring $T_{max}$, was determined using the R package *climwin*[@vandepol_identifying_2016].
-This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and difference in Akaike information criterion corrected for small sample size relative to the null model, $\Delta AICc$, of each window.
-Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`), selecting the critical temperature window as that with the lowest $\Delta AICc$.
-Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable, as opposed to $T_{min}$.
+`r schematic_summary`), defined as the period over which $T_{max}$ was most strongly correlated with $DOY_{25}$, was determined using the R package *climwin*[@vandepol_identifying_2016].
+This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame.
+It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window.
+The variable and time window yielding the biggest $\Delta AICc$ are selected for our critical temperature window.
+Here, we tested for correlation between temperature predictor variables ($T_{max}$, $T_{min}$) and biological outcome variable $DOY_{25}$ over the time-frame from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`).
+We selected the critical temperature window that had the lowest $\Delta AICc$.
+Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
 We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth.
@@
 To ensure that patterns were robust under an alternative definition of critical temperature window, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
-`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses defining the critical temperature window as the month of April.
+`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the critical temperature window to be the month of April.
 This was consistent with the periods identified by *climwin* for ring- and diffuse-porous species groups at both sites, all of which included all or part of April (`r  EDT_phenology_parameters`).
@@
 
-Correlation between growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
+Correlation between the dendroband-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
 `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
 Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
-Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree.
+Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and of individual tree.
 We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
-Models were run using the LME4 package in R [@bates_fitting_2015]. 
 This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020].
@@
 Dominant tree species were cored at both SCBI[@helcoski_growing_2019;@bourg_initial_2013] and Harvard Forest[@alexander_potential_2019; @dye_comparing_2016; @finzi_carbon_2020] following sampling designs that covered a broad range of DBH.
-We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L.
+We analyzed records for the ring- and diffuse- porous species at each site (`r  EDT_species`), but excluded species with other xylem architectures (*Juglans nigra* L.
 at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest).
-In total, we analyzed `r n_cores_focal_sites` cores from `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r  EDT_species`).
+We studied a total of `r n_cores_focal_sites` cores which included `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r  EDT_species`).
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018; @maxwell_increased_2017; @wilde_comparing_2018;@strange_comparing_2019;@au_demographic_2020;@matheus_placing_2018;@leblanc_radial_2020;@maxwell_higher_2019;@maxwell_dendroclimatic_2015;@schmidt_325year_2015] (**JUSTIN, please confirm that we got all the appropriate references here.**).
-Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018].
+Again, records were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).
 
-~Add this comment somewhere in here: "What about mentioning "we converted the trw into a dimensionless growth index (thereafter simply referred to as RWI; see Methods) to emphasize interannual variability associated with climate  (Cook & Kairiukstis, 1990)" or something along those lines?"~
+~*Add this comment somewhere in here: "What about mentioning "we converted the trw into a dimensionless growth index (thereafter simply referred to as RWI; see Methods) to emphasize interannual variability associated with climate  (Cook & Kairiukstis, 1990)" or something along those lines?"*
+
+*KRISTA can you confirm if the following sentence is describing the transformation from trw into RWI? If it is, we can define RWI here*
 
 We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
-*Chronologies were built using the methods described in Ref.[@helcoski_growing_2019].* (**(NEIL/JUSTIN, is this accurate, and do you think it's sufficient?
-If not, could you please provide a brief summary?**)
+Chronologies were built using the methods described in Helcoski et al. (2019)[@helcoski_growing_2019].
 We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
 
-We obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-
 <!--QUILT PLOT METHODS-->
-Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v.1.3.1093 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
+Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
 We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).

### 2021-11-12 doc/manuscript/Dow_manuscript_Nature.Rmd
update Fig. 1 legend
#85

@@
 
-![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant (determined by whether the posterior credible interval included 0 or not) increases and decreases, respectively, 'n.s.' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant (determined by whether the posterior credible interval included 0 or not) increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 

### 2021-11-13 doc/manuscript/Dow_Extended_Data.Rmd
Update Dow_Extended_Data.Rmd
Hopefully this fixes spacing issue. Needs to be re-kinted @teixeirak (it wouldn't knit for me)

@@
 add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
-  pack_rows("", 1, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 10, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 12, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 15, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 1, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 11, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 13, 15, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 16, 23, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2021-11-13 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
More edits

@@
 
-William McShea^1^
+William J. McShea^1^
 
@@
 `r DOY_timing`, `r EDF_pheno_all`).
-
 We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the Spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW).
@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
+We converted the ring-width measurements into a dimensionless growth index (hereafter simply referred to as RWI; see Methods).
 Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of chronologies: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
@@
 # Acknowledgements {-#Acknowledgements} 
-- people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data)
+- people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data), We would like to thank Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Josh Oliver, Karly Schmidt-Simard, Brandon Strange, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie
+for assistance in the field and laboratory.
 
-- Funding: ForestGEO
+- Funding: ForestGEO, the collection of tree-ring samples was  funded by a USDA Agriculture and Food Research Initiative grant 2017-67013-26191 and from the Indiana University Vice Provost for Research Faculty Research Program.
 
+-David Orwig
 **COAUTHORS, PLEASE ADD.**
@@
 # References {-#References}
+
+*I can't figure out how to get this into zotero yet :(*
+Stan Development Team. 2019. Stan Modeling Language Users Guide and Reference Manual, 2.28. https://mc-stan.org

### 2021-11-13 doc/manuscript/Dow_manuscript_Nature.Rmd
coauthor feedback

@@
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~2512 words (1609 intro + results, discussion 903) and 1 modest + 2 large display items.
+We're currently at ~2678 words (1609 intro + results, discussion 1069) and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently ok there)*
@@
 In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
-A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019] *Coauthors suggest removing this but i'm unsure.
-I find this distinction relevant, but if more experienced people disagree I'm fine removing - would like your opinion*.
+A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019] <!-- Coauthors suggest removing this but i'm unsure.
+I find this distinction relevant, but if more experienced people disagree I'm fine removing - would like your opinion*!
+-->.
 The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
@@
 klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
-The climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
+The climate sensitivity of woody growth phenology in deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
 
@@
 Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
-To test the assumption that warmer springs lead to an extension of the growing season we used measurements from 465 dendrometer bands across two decidious mid-atlantic forests.
-To test whether spring temperatrues were consistently positively correlated with annual ring-width Index (RWI; tree-ring width which has been standardized for age) we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.
+To test the assumption that warmer springs lead to an extension of the growing season we used measurements from 465 dendrometer bands across two deciduous mid-atlantic forests.
+To test whether spring temperatures were consistently positively correlated with annual ring-width Index (RWI; tree-ring width which has been standardized for age) we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.
 
@@
 
-Both MODIS-derived canopy foliage phenology and dendrometerband measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig.
+Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig.
 `r DOY_timing`, `r EDF_pheno_all`).
@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
-We converted the ring-width measurements into a dimensionless growth index (hereafter simply referred to as RWI; see Methods).
 Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of chronologies: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
@@
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatrues were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
 
@@
 Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.
+The effect of warm springs specifically in years with higher levels of precipitation during the peak growing season is not elucidated from this study, and should be examined in the future to clarify this relationship.
 

### 2021-11-14 RScripts/SCBI_wood_phenology.R
cleaning up scripts

@@ -30,7 +30,6 @@ source("RScripts/dendroband_functions.R")
 
 
 # Format dendroband data ----------------------------------------------------
-# files <- dir("data", pattern="_201[0-9]*.csv")
 files <- dir("data", pattern = "_20[1-2][0-9]*.csv")
 dates <- c(2011:2020)
 
@@ -79,11 +78,6 @@ make_growth_list(files, dates)
 # Loop to create timeseries of dbh measurements manually -----
 # Taken from Ian's growth_over_time.R
 
-# before we knew Sean was working on a package (RDendrom), Valentine and I tried
-# to create the same functionality manually (#5 and #6 in this script). Since we
-# do have RDendrom, this manual work is deprecated, but I've decided to leave it
-# here in case it's needed later. To run this section, you need to run section
-# #1 above first.
 
 # description of loop
 ## 1. First, assigns the first dbh of the growth column as the first dbh.
@@ -205,12 +199,11 @@ for (e in 1:length(all_stems_intra)) { # Loop to merge all dendroband surveys in
   all_stems <- rbind(all_stems, Stem)
 }
 all_stems <- all_stems[complete.cases(all_stems$dbh2), ]
-all_stems <- all_stems[, c(1:31, 34, 35)]
-
+#all_stems <- all_stems[, c(1:31, 34, 35)]#####ISSUE
 # Added by bert: write all_stems observed data to csv
 write.csv(all_stems, file = "Data/all_stems.csv", row.names = FALSE)
 all_stems <- read.csv("Data/all_stems.csv")
-all_stems <- read_csv("~/GitHub/Dendrobands/data/scbi.dendroAll_2020.csv")
+#all_stems <- read_csv("~/GitHub/Dendrobands/data/scbi.dendroAll_2020.csv")
 
 
 ## Manual check of data ##
@@ -235,8 +228,10 @@ all_stems <- read_csv("~/GitHub/Dendrobands/data/scbi.dendroAll_2020.csv")
 # Loop to create a final "master dataframe" -------
 # The following loop will cycle through years, species, and tags to create a final "master dataframe"
 # It includes two of Sean's functions wrapped with some of my code
-all_stems <- read_csv("data/all_stems.csv")
-all_stems$dbh2 <- all_stems$dbh2/10
+all_stems <- read.csv("data/all_stems.csv")
+
+all_stems$dbh2 <- all_stems$dbh2/10 #?
+
[…]

### 2021-11-14 RScripts/data_cleaning.R

@@ -60,6 +60,8 @@ percent_growth <- Wood_pheno_table_scbi %>%
     dbh_growth_percent_cummulative = cumsum(dbh_growth_percent)
   )
 
+LG5_parameter_values_scbi$tag_year <- paste0(LG5_parameter_values_scbi$tag,LG5_parameter_values_scbi$year)
+
 Wood_pheno_table_scbi <- Wood_pheno_table_scbi %>%
   separate(tag, into = c("tag", "stem"), sep = "_") %>%
   mutate(
@@ -71,7 +73,7 @@ Wood_pheno_table_scbi <- Wood_pheno_table_scbi %>%
 # Break into RP and DP
 percent_growth_RP <- subset(percent_growth, wood_type == "ring-porous")
 
-# Remove models with single days > 2% growth (this is abnormal for RP trees in our study)
+# Remove models with single day growth greater than 2 SD away from mean
 maxgrowthrate <- aggregate(percent_growth_RP$dbh_growth_percent, by = list(percent_growth_RP$tag_year), FUN = max)
 names(maxgrowthrate) <- c("tag_year", "rate") # Maximum growth rate of each tree
 # high_growthday <- subset(maxgrowthrate, rate >= 0.02) #Remove trees where one day had higher than 2% of total growth
@@ -86,7 +88,6 @@ count(wood_pheno_removed, sp)
 hist(wood_pheno_removed$dbh)
 # subset main DF
 percent_growth_RP <- percent_growth_RP[!(percent_growth_RP$tag_year %in% unique(high_growthday$tag_year)), ]
-
 
 # Remove models with peak growth outside of expected season. IE, winter growth peaks
 ratedoy <- percent_growth_RP[percent_growth_RP$dbh_growth_percent %in% maxgrowthrate$rate, ] # DOY where max growth occured
@@ -119,10 +120,10 @@ percent_growth_RP <- percent_growth_RP[percent_growth_RP$tag_year %in% keeptags,
 #DIffuse porous - Same method
 percent_growth_DP <- subset(percent_growth, wood_type == "diffuse-porous")
 
-# Remove models with single days > 4% growth (this is abnormal for DP trees in our study)
+# Remove models with single day growth rate >2 SD away from mean
 maxgrowthrate <- aggregate(percent_growth_DP$dbh_growth_percent, by = list(percent_growth_DP$tag_year), FUN = max)
[…]

### 2021-11-14 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
add statement about 2 additional "weird" tag years we removed

@@
 The raw dendrometer band data were manually inspected before analysis. 
-We screened the data for three classes of erros. 
+We screened the data for three classes of errors. 
 First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and the datapoint was removed.
@@
 (1) single day growth rates were $\ge 2$ standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
-(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 62, Harvard Forest = 106); 
+(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 53, Harvard Forest = 106); 
 (3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.02$ mm; SCBI = 0, Harvard Forest = 66); 
-(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 181, Harvard Forest = 199); 
+(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 150, Harvard Forest = 199); 
 and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3). 
 At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
+An additional two tree-years were removed from SCBI because they were poor fits missed using the previous cleaning criteria. 
 At SCBI a higher proportion of ring-porous trees were removed, the majority falling under condition 4.

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
reviewing highlighted comments in google doc
@camerondow35 , FYI

@@
 
-David A. Orwig^3^
+David A. Orwig^3^ ***
 
@@
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
+As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
 However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014].
@@
 In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
-A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019] <!-- Coauthors suggest removing this but i'm unsure.
-I find this distinction relevant, but if more experienced people disagree I'm fine removing - would like your opinion*!
--->.
+A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >300 Tg C yr^-1^ (>30% of the total forest C sink)[@pugh_role_2019].
+<!-- The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al.
+2019 estimate that ~200 Tg C yr-1 is attributable to demography (forest regrowth), while >100 Tg C yr-1 is controlled by environmental drivers.
+The latter portion seems more likely to be sensitive to spring temperatures, and I originally cited this number, but upon further thought it's probably complex enough that that's a false distinction.
+-->
 The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
@@
 In temperate deciduous forests, spring warming generally lengthens the growing season, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999], and extends the period where the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-Current models assume that longer growing seasons lead to increasing net annual carbon sequestration (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+Current models assume that longer growing seasons lead to increasing net annual carbon uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
 However, recent experimental and observational findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
@@
 klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
-The climate sensitivity of woody growth phenology in deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
+The climate sensitivity of woody growth phenology in temperate deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
 
@@
 
-Both canopy foliage phenology and stem growth milestones occurred earlier at SCBI than at Harvard Forest, with stem growth milestones achieved on average 6-10 days earlier at SCBI (Fig.
-`r DOY_timing`, `r  EDT_phenology_parameters`).
+Growth milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
 Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
clarify statements on general (leaf + stem)  vs leaf phenology

@@
 
-For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`),
-which is consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014]
-Specifically, greenup *(DOY when EVI2 first crossed 15% of the segment EVI2 amplitude)* DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, while mid-greenup also advanced but to a lesser extent (Fig.
-`r DOY_timing`, `r EDT_phenology_parameters`).
+For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of phenological milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
+Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
+`r DOY_timing`, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
 Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
finish reviewing highlights in google doc

@@
 `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have a negligible effect on total annual growth for most species and locations (Fig.
+However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion, thus having negligible effect on total annual growth for most species and locations (Fig.
 `r tree_rings`).
@@
 However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are leading to a lengthening of the period from peak stem growth to the cessation of ecosystem C uptake.
+Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of ecosystem C uptake.
 As the extra C sequestration in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
@@
 As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
-Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[**REFS**] are likely to alter the phenology of forest C cycling.
+Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.
 If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
adjustments to authorship and acknowledgements
Dave Orwig didn't feel he'd contributed enough to be an author.

@@
 Sean McMahon^8,10^ (Orcid ID : 0000-0001-8302-6908)
-
-David A. Orwig^3^ ***
 
@@
 # Acknowledgements {-#Acknowledgements} 
-- people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data), We would like to thank Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Josh Oliver, Karly Schmidt-Simard, Brandon Strange, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie
-for assistance in the field and laboratory.
-
-- Funding: ForestGEO, the collection of tree-ring samples was  funded by a USDA Agriculture and Food Research Initiative grant 2017-67013-26191 and from the Indiana University Vice Provost for Research Faculty Research Program.
-
--David Orwig
-**COAUTHORS, PLEASE ADD.**
+We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly (**ADD SCBI RESEACHERS TO THIS LIST-- see list on GitHub**) Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
+Thanks to David A. Orwig and members of the ForestGEO Ecosystesms & Climate Lab at SCBI for helpful feedback.
+The research was funded by ForestGEO (Smithsonian). 
+Collection of tree-ring samples was funded by a USDA Agriculture and Food Research Initiative grant 2017-67013-26191 and from the Indiana University Vice Provost for Research Faculty Research Program.
 

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
add SCBI researcher acknowledgements

@@
 # Acknowledgements {-#Acknowledgements} 
-We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly (**ADD SCBI RESEACHERS TO THIS LIST-- see list on GitHub**) Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
+We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
 Thanks to David A. Orwig and members of the ForestGEO Ecosystesms & Climate Lab at SCBI for helpful feedback.

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
formatting

@@
 
-*NOTE: We're probably pretty close to the upper length limit.
+<!--NOTE: We're probably pretty close to the upper length limit.
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~2678 words (1609 intro + results, discussion 1069) and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently ok there)*
+We're currently at ~2652 words and 1 modest + 2 large display items.
+Articles typically have 30-50 references (currently ok there)-->
 
-# (Summary paragraph)
-```{r eval=FALSE, echo=FALSE}
-## For Nature:
-# Articles start with a fully referenced summary paragraph, ideally of no more than 200 words, which is separate from the main text and avoids numbers, abbreviations, acronyms or measurements unless essential.
+<!-- Articles start with a fully referenced summary paragraph, ideally of no more than 200 words, which is separate from the main text and avoids numbers, abbreviations, acronyms or measurements unless essential.
 It is aimed at readers outside the discipline.
 This summary paragraph should be structured as follows: 2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
-Please refer to our annotated example  to see how the summary paragraph should be constructed.
+Please refer to our annotated example  to see how the summary paragraph should be constructed.-->
 
-```
+<!--SUMMARY PARAGRAPH-->
 
-As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014].
-Using dendrometer band measurements from *465* trees across two deciduous mid-Atlantic forests, we show that warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
-The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
-These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
-Such empirical results contradict current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], and imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
+**However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014].**
+**Using dendrometer band measurements from `r n_dendro_trees` trees across two deciduous mid-Atlantic forests, we show that warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
+**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
+**These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
+**
+**Such empirical results contradict current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], and imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.**
 
-# (Body)
+<!--BODY-->
 

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
adjust language around NEE/ C sequestration
I avoided NEE in conjunction with "increases/ decreases" to avoid confusion over the sign convention.

@@
 In temperate deciduous forests, spring warming generally lengthens the growing season, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999], and extends the period where the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-Current models assume that longer growing seasons lead to increasing net annual carbon uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+Current models assume that longer growing seasons lead to increasing annual net carbon uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
 However, recent experimental and observational findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
 
-While the responses of leaf phenology and seasonal *ecosystem (?)* CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
+While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
 In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021].
@@
 <!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
-Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ sequestration[@keenan_net_2014]. 
+Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014]. 
 Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
@@
 The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
+`r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_ [@keenan_net_2014; @finzi_carbon_2020].
 However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion, thus having negligible effect on total annual growth for most species and locations (Fig.
@@
 Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig.
-`r tree_rings`) than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
-However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
+`r tree_rings`) than to growing season length aligns with parallel findings for _NEE_ [@zhou_explaining_2016; @fu_maximum_2019].
+However, there is also a disconnect with findings that C uptake increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of ecosystem C uptake.
-As the extra C sequestration in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
+As the extra C uptake in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
@@
 Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.
-If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
+If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_ [@churkina_spatial_2005].
 Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
+n_dendro_trees_SCBI <- 123 #doing this the lazy way. summed from species_table rather than coded. 
+n_dendro_trees_HF <- 340 #doing this the lazy way. summed from species_table rather than coded. 
 n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
@@
 While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021].
+In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site to our knowledge[@dorangeville_peak_2021].
 <!-- I'm not aware of any others, but we do want to make sure we're not missing anything.
@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019] via Google Earth Engine.
+Data were extracted for the pixel containing the NEON tower at each site.
 Using daily MODIS 2-band Enhanced Vegetation Index data (EVI2) at a spatial resolution of 500m, the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
-
  <!--Climate -->

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
Harvard forest is not mid-Atlantic

@@
 **However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014].**
-**Using dendrometer band measurements from `r n_dendro_trees` trees across two deciduous mid-Atlantic forests, we show that warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
+**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 For SCBI, weather data were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
-The R package *clim.pact*[@climpact_REF; @benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
+The R package *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
@@
 We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
-This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020].
+This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019].
 In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults.
@@
 We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
-Chronologies were built using the methods described in Helcoski et al. (2019)[@helcoski_growing_2019].
+~~Chronologies were built using the methods described in Helcoski et al. (2019)[@helcoski_growing_2019].~~
+~~*The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
+Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/ 3rds spline method to adjust changes in variance as series replica- tion decreased towards the earlier portion of each chronology (Jones et al., 1997).
+The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
+the inner 30–50 yr of each record depending on full chronology length) dropped below three trees.
+Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices (Cook, 1985).*~~
 We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
all references resolved
word-smithing in intro

@@
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
+**These findings imply that extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
 **
@@
 In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
-A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >300 Tg C yr^-1^ (>30% of the total forest C sink)[@pugh_role_2019].
+A large portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ (>30% of the total forest C sink)[@pugh_role_2019].
 <!-- The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al.
@@
 
-In temperate deciduous forests, spring warming generally lengthens the growing season, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999], and extends the period where the ecosystem is a net CO~2~ sink[@keenan_net_2014].
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
 Current models assume that longer growing seasons lead to increasing annual net carbon uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings that sink factors can limit annual growth[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
+However, recent experimental and observational findings that annual growth can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
 
 While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site to our knowledge[@dorangeville_peak_2021].
+In fact, only one study has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021].
 <!-- I'm not aware of any others, but we do want to make sure we're not missing anything.
@@
 The climate sensitivity of woody growth phenology in temperate deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
+
+<!-- KAT paused here.-->
 
@@
 # References {-#References}
-
-*I can't figure out how to get this into zotero yet :(*
-Stan Development Team. 2019. Stan Modeling Language Users Guide and Reference Manual, 2.28. https://mc-stan.org

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
add Neil's text on chronology methods

@@
 
-We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
-~~Chronologies were built using the methods described in Helcoski et al. (2019)[@helcoski_growing_2019].~~
-~~*The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
-Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/ 3rds spline method to adjust changes in variance as series replica- tion decreased towards the earlier portion of each chronology (Jones et al., 1997).
-The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
-the inner 30–50 yr of each record depending on full chronology length) dropped below three trees.
-Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices (Cook, 1985).*~~
-We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
+A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985; @cook_methods_1990].
+An adaptive power transformation, a process that also stabilises the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
+Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records [@jones_estimating_1997].
+The 1/3rds spline method was chosen when replication in the inner portion of each chronology (ca. inner 30–50 yr of each record depending on full chronology length) was less than three trees. 
+When replication was greater than that, we used the average correlation between raw ring-width series (rbar) method.
+The robust biweight mean chronology for each species was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
+We defined chronology start year (Extended Data Table 1) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
 

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
edit chronology methods/ RWI

@@
 
-~*Add this comment somewhere in here: "What about mentioning "we converted the trw into a dimensionless growth index (thereafter simply referred to as RWI; see Methods) to emphasize interannual variability associated with climate  (Cook & Kairiukstis, 1990)" or something along those lines?"*
-
-*KRISTA can you confirm if the following sentence is describing the transformation from trw into RWI? If it is, we can define RWI here*
-
+For each species-site combination, we converted tree ring records into a dimensionless ring width index (RWI) to emphasize interannual variability associated with climate [@cook_methods_1990].
 A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985; @cook_methods_1990].
@@
 When replication was greater than that, we used the average correlation between raw ring-width series (rbar) method.
-The robust biweight mean chronology for each species was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
+The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
 We defined chronology start year (Extended Data Table 1) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 
-<!-- KAT paused here.-->
-
 Tree-ring records, which can be used to examine correlations of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
@@
 
+<!-- KAT paused here.-->
+
 Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
-To test the assumption that warmer springs lead to an extension of the growing season we used measurements from 465 dendrometer bands across two deciduous mid-atlantic forests.
+To test the assumption that warmer springs extend the period of stem growth, we used measurements from 465 dendrometer bands across two deciduous forests. 
 To test whether spring temperatures were consistently positively correlated with annual ring-width Index (RWI; tree-ring width which has been standardized for age) we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.
@@
 Using daily MODIS 2-band Enhanced Vegetation Index data (EVI2) at a spatial resolution of 500m, the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
+
  <!--Climate -->
-
-*ADD SOMETHING ON HOW STANDARD PRECIPITAITON EVAPOTRANSPIRATIN INDEX WAS FOUND*!!!!
 
@@
 Climate data for Harvard Forest were complete and used as is.
+
+*ADD SOMETHING ON HOW STANDARD PRECIPITAITON EVAPOTRANSPIRATIN INDEX WAS FOUND*!!!!
 

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 
+Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
+To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on r n_dendro_trees` trees across two deciduous forests. 
+To test whether spring temperatures consistently increased annual growth, we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.
+
 <!-- KAT paused here.-->
-
-Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
-To test the assumption that warmer springs extend the period of stem growth, we used measurements from 465 dendrometer bands across two deciduous forests. 
-To test whether spring temperatures were consistently positively correlated with annual ring-width Index (RWI; tree-ring width which has been standardized for age) we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.
 
@@
 Maxwell, Ian R.
-McGregor, William McShea, David A.
-Orwig, Neil Pederson, Alan J.
+McGregor, William McShea, Neil Pederson, Alan J.
 Tepley, and Kristina J.

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
updated based on set.seed in quilt plot scripts
@teixeirak these values changed because the function to determine correlations uses randomizations at some point. I've manually set the seed so anyone who runs these scripts should get the same results we've got here.

@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
-Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of chronologies: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
-In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
+Across all chronologies, RWI was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring-porous and 4 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
+In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
 $T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
@@
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% (*4%?
+Double check this*) of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
 

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
Note for #87

@@
 <!--
-TODO added by Bert: Whereever we specify model details, include details about prior from here:
+More information on rstanarm's default weakly informative priors:
 https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
tweak abstract

@@
 **As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014].**
+**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
 **Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
+**These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
 **
-**Such empirical results contradict current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], and imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.**
 
@@
 Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
-To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on r n_dendro_trees` trees across two deciduous forests. 
+To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on r n_dendro_trees` trees across two mid-latitude deciduous forests. 
 To test whether spring temperatures consistently increased annual growth, we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
tweaked abstract & intro

--- 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 **As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014].**
-**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
+**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
+**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
+**These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest carbon balance.
 **
-**Such empirical results contradict current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], and imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
 
@@
 
-In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
+In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011; @friedlingstein_global_2020].
 A large portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ (>30% of the total forest C sink)[@pugh_role_2019].
@@
 Current models assume that longer growing seasons lead to increasing annual net carbon uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings that annual growth can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on C sequestration in trees.
+However, recent experimental and observational findings that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
 
-While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
-In fact, only one study has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021].
+While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], very little is known about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
+In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021].
 <!-- I'm not aware of any others, but we do want to make sure we're not missing anything.
@@
 
-Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.
-To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on r n_dendro_trees` trees across two deciduous forests. 
-To test whether spring temperatures consistently increased annual growth, we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.
+Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
+To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
+To test whether spring temperatures consistently increased annual growth, we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
 

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
-To test whether spring temperatures consistently increased annual growth, we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
+To test whether spring temperatures consistently increased annual growth, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
 

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
edit Fig. 1 legend
@camerondow35 , I removed "(determined by whether the posterior credible interval included 0 or not)" from the Fig. 1 legend. I think that's a detail that can be specified in the methods. But let me know if you have reason to feel strongly about this.

@@
 
-![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant (determined by whether the posterior credible interval included 0 or not) increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
tweaks to results text, built out Fig. 3 caption

@@
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
-Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and < one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
+Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
 
@@
 `r DOY_timing`, `r EDF_pheno_all`).
-We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the Spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW).
+We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW).
 CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 and $DOY_{25}$ (`r EDF_climwin`).
@@
 
-For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of phenological milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
+For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of phenological milestones.
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
 `r DOY_timing`, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
-Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
+Similarly, at both sites, the stem growth milestones $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
@@
 
-To test whether the negative effect of summer temperatures might offset the enhancement of growth by warmer spring temperatrues, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
+To test whether the negative effect of summer temperatures might offset an enhancement of growth by warmer spring temperatures, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% (*4%?
-Double check this*) of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 4% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
 
-![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
-Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature.
-Chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
+![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
+Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
 

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
-`r DOY_timing`, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
+`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
 Similarly, at both sites, the stem growth milestones $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
-`r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
+`r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
 Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 1.1-1.9 days/ $^\circ$C for ring-porous species and 3.5-3.6 days/ $^\circ$C for diffuse-porous species at SCBI, and 2.8-7.2 days/ $^\circ$C for ring-porous species and 6.6-7.9 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).
@@
 Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig.
-`r DOY_timing`), there was no consistent lengthening of $L_{PGS}$ (Fig.
+`r DOY_timing`a-b), there was no consistent lengthening of $L_{PGS}$ (Fig.
 `r schematic_summary`, `r EDF_pheno_all`).
@@
 The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_ [@keenan_net_2014; @finzi_carbon_2020].
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_ [@keenan_net_2014; @finzi_carbon_2020].
 However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion, thus having negligible effect on total annual growth for most species and locations (Fig.

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
edit first par of discussion

@@
 The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_ [@keenan_net_2014; @finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion, thus having negligible effect on total annual growth for most species and locations (Fig.
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
+However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
 `r tree_rings`).
 Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
-Rather than lengthening the growing season, the primary effect of warming temperature on annual growth is a negative correlation with $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of an earlier onset of growth in the spring[@zhang_drought_2021].
-Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.
-The effect of warm springs specifically in years with higher levels of precipitation during the peak growing season is not elucidated from this study, and should be examined in the future to clarify this relationship.
+Our tree-ring analysis (Fig.
+`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual tree growth is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018].
+Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on growth phenology elucidated here (Fig.
+`r schematic_summary`) were not attributable to summer drought.
 

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
edit discussion

@@
 
-Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig.
-`r tree_rings`) than to growing season length aligns with parallel findings for _NEE_ [@zhou_explaining_2016; @fu_maximum_2019].
-However, there is also a disconnect with findings that C uptake increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of ecosystem C uptake.
-As the extra C uptake in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
+Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
+However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
+Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
+We show that the extra carbon fixation in years with warm springs does not substantially augment woody growth, but it remains unclear how it is allocated within the ecosystem.
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
 
-One possibility is that extra C production in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
+One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
 It is theoretically possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more carbon-dense wood in years with warm springs.
@@
 A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
-Much of this carbon would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring [@oishi_warmer_2018; @buermann_widespread_2018].
+Much of this carbon would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
 However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
@@
 Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.
-If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_ [@churkina_spatial_2005].
+If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
 Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in carbon sequestration and ensuant negative climate change feedback that is anticipated in most climate forecasting  models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in CO~2~ sequestration and ensuant negative climate change feedback that is anticipated in most climate forecasting  models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures, which constitutes a positive feedback to climate change.

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
abbreviate carbon as C throughout

@@
 **As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
+**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
 **Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest carbon balance.
+**These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
 **
-**Rather, contradicting current projections from global carbon cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
 
@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-Current models assume that longer growing seasons lead to increasing annual net carbon uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
 However, recent experimental and observational findings that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
 
-While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], very little is known about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures.
+While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], very little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
 In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021].
@@
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
 Our tree-ring analysis (Fig.
@@
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
-We show that the extra carbon fixation in years with warm springs does not substantially augment woody growth, but it remains unclear how it is allocated within the ecosystem.
+We show that the extra C fixation in years with warm springs does not substantially augment woody growth, but it remains unclear how it is allocated within the ecosystem.
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
@@
 One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
-It is theoretically possible that extra carbon is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more carbon-dense wood in years with warm springs.
+It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
 However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
-Extra carbon could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
+Extra C could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
 Extension of our tree-ring analysis revealed little correlation between April $T_{max}$ and growth the following year (sig.
@@
 A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
-Much of this carbon would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
+Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
 However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
@@
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
-As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
+As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and C cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
 Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 We're currently at ~2652 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently ok there)-->
+Articles typically have 30-50 references (currently ok there-- n= 48)-->
 
@@
 Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
-The number of bands measured at each site fluctuated slightly as trees were added or dropped (e.g., because of tree mortality) from the census. 
+The number of bands measured at each site fluctuated slightly as trees were added or dropped from the census (e.g., because of tree mortality). 
 Across years, the number of bands sampled averaged 129 (range: 91-138) at SCBI and 717 (range: 700-755) at Harvard Forest. <!--verified - these are the pre-cleaning numbers-->
@@
 Measurements were timed to begin before the beginning of spring growth and to continue through the cessation of growth in the fall. 
-At SCBI, the median start date was April 14, which was adjusted when early leaf-out of understory vegetation was observed, with the earliest start date being March 30 (in 2020). 
+At SCBI, the median start date was April 14, which was adjusted forward when early leaf-out of understory vegetation was observed, with the earliest start date being March 30 (in 2020). 
 Measurements were continued through to fall leaf senescence, with the median end date being October 17 and the latest end date November 26 (2012). 
@@
 In these cases, the full annual record for the tree was removed.
-Third, data points that deviated substantially from normal growth patterns, but had unknown causes, were removed.
+Third, data points that deviated substantially from normal growth patterns, but for unknown causes, were removed. 
 If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
@@
 `r schematic_summary`).
-In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; i.e.
-julian days) as:
+In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as:
 $$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
@@
 $DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
-$r$ *shapes the slope of the curve at the inflection point*,
+$r$ shapes the slope of the curve at the inflection point,
 and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote. 
-*We note that the DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ only when $\theta = 1$.*
-The model was fit in R v 4.0 using the functions developed in the Rdendrom package[@mcmahon_general_2015].
+The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
+The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015].
 These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year.
-We then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest detailed in Fig. 1.
+We then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
 
@@
 At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
-An additional two tree-years were removed from SCBI because they were poor fits missed using the previous cleaning criteria. 
+*An additional two tree-years were removed from SCBI because they were poor fits missed using the previous cleaning criteria.* 
 At SCBI a higher proportion of ring-porous trees were removed, the majority falling under condition 4.

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 (4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 150, Harvard Forest = 199); 
-and (5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3). 
+(5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3); 
+and (6) poorly fit models that did not meet any of the above criteria (SCBI = 2, Harvard Forest = 0).
 At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
-*An additional two tree-years were removed from SCBI because they were poor fits missed using the previous cleaning criteria.* 
-At SCBI a higher proportion of ring-porous trees were removed, the majority falling under condition 4.
+At SCBI, a higher proportion of ring-porous trees were removed, the majority falling under condition 4.
 
@@
 MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019] via Google Earth Engine.
-Data were extracted for the pixel containing the NEON tower at each site.
+Extracted pixels were those containing the NEON tower at each site. 
 Using daily MODIS 2-band Enhanced Vegetation Index data (EVI2) at a spatial resolution of 500m, the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
@@
  <!--Climate -->
-
-For the dendrometer band analysis, climate data corresponding to the measurement periods at both sites were obtained from local weather stations. 
-For SCBI, weather data were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
+For the dendrometer band and leaf phenology analyses, climate data corresponding to the measurement periods were obtained from local weather stations at each focal site. 
+For SCBI, weather data were obtained from a meteorological tower adjacent to the ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
 The R package *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
-For each site, we extracted daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
-Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station[@boose_shaler_2021; @boose_fisher_2021]. 
-Climate data for Harvard Forest were complete and used as is.
+Daily temperature records for Harvard Forest, which had already been gap-filled based on other local records, were obtained from the Harvard Forest weather station[@boose_shaler_2021; @boose_fisher_2021].
+For each site, we used records of daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
 
+To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring tempreatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
+The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020]. 
 *ADD SOMETHING ON HOW STANDARD PRECIPITAITON EVAPOTRANSPIRATIN INDEX WAS FOUND*!!!!

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
rearrange methods text on climate data

@@
 
-To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring tempreatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
-The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020]. 
-*ADD SOMETHING ON HOW STANDARD PRECIPITAITON EVAPOTRANSPIRATIN INDEX WAS FOUND*!!!!
-
-For the tree core analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-
 <!-- Determining critical temperature windows -->
-
 The critical temperature window (Fig.
@@
 <!--Hypothesis testing -->
-
 Correlation between the dendroband-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
@@
 -->
+
+To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring tempreatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
+The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020]. 
+*ADD SOMETHING ON HOW STANDARD PRECIPITAITON EVAPOTRANSPIRATIN INDEX WAS FOUND*!!!!
 
@@
 <!--QUILT PLOT METHODS-->
+For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
 Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
finished manuscript check!! :-)
@camerondow35 , FYI

@@
 
-<!-- KAT paused here.-->
-
 ## Dendrometer band analysis {-#Dendrometer}
@@
 The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
-Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
+Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera* L.), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1° C in January and 24° C in July[@anderson-teixeira_joint_2021].
@@
 The forest is secondary and mixed age, having re-established around the beginning of the 20th century following agricultural use and significant hurricane damage in 1938. 
-Dominant species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
+Dominant species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis* (L.) Carrière), oak (*Quercus spp.*) and red maple (*Acer rubrum* L.). 
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2021].
@@
 <!-- Determining critical temperature windows -->
-The critical temperature window (Fig.
+The critical temperature window (CTW, Fig.
 `r schematic_summary`), defined as the period over which $T_{max}$ was most strongly correlated with $DOY_{25}$, was determined using the R package *climwin*[@vandepol_identifying_2016].
@@
 It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window.
-The variable and time window yielding the biggest $\Delta AICc$ are selected for our critical temperature window.
 Here, we tested for correlation between temperature predictor variables ($T_{max}$, $T_{min}$) and biological outcome variable $DOY_{25}$ over the time-frame from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`).
-We selected the critical temperature window that had the lowest $\Delta AICc$.
+The time period yielding the lowest $\Delta AICc$ was selected as the CTW.
 Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
-We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth.
+We defined CTW for $DOY_{25}$, as opposed to other growth phenology parameters, because spring temperatures should have the most direct influence on this variable.
 
-To ensure that patterns were robust under an alternative definition of critical temperature window, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
-`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the critical temperature window to be the month of April.
+To ensure that patterns were robust under an alternative definition of CTW, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
+`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the CTW to be the month of April.
 This was consistent with the periods identified by *climwin* for ring- and diffuse-porous species groups at both sites, all of which included all or part of April (`r  EDT_phenology_parameters`).
@@
 <!--Hypothesis testing -->
-Correlation between the dendroband-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
+Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
 `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
 Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
-Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and of individual tree.
+Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree.
 We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
@@
 We analyzed records for the ring- and diffuse- porous species at each site (`r  EDT_species`), but excluded species with other xylem architectures (*Juglans nigra* L.
-at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest).
+at SCBI, *Tsuga canadensis* at Harvard Forest).
 We studied a total of `r n_cores_focal_sites` cores which included `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r  EDT_species`).
@@
 An adaptive power transformation, a process that also stabilises the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
-Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records [@jones_estimating_1997].
+Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records[@jones_estimating_1997].
 The 1/3rds spline method was chosen when replication in the inner portion of each chronology (ca. inner 30–50 yr of each record depending on full chronology length) was less than three trees. 

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
fixed up Extended Data Table 1

--- 2021-10-21 doc/manuscript/Dow_Extended_Data.Rmd
@@
 library(kableExtra)
-species_table <- read.csv(file = "tables_figures/species_table.csv", col.names = c("site", "xylem porosity","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
+species_table <- read.csv(file = "tables_figures/species_table.csv", col.names = c("site", " ","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
+species_table <- species_table[c(2:3, 5:8)] #subset(species_table, select = -c(Species code) )#remove columns that shouldn't be included
+
 kable(species_table, caption = species_table_caption ,  format = "latex", booktabs = TRUE,  escape = F) %>%
-add_header_above(c(" ", " ", " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
-  pack_rows("", 1, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 10, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 12, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 15, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+add_header_above(c( " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
+  pack_rows("SCBI", 1, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 11, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Harvard Forest", 13, 15, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 16, 23, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
fix up extended data table 2
#90

@@
 species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
-mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest."
+mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
 tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations to monthly $T_{max}$ in univariate and multivariate analyses."
@@
 library(kableExtra)
-mean_growth_parameters <- read.csv("tables_figures/phenology_parameters.csv",  col.names = c("", "ring","diffuse","ring", "diffuse"), stringsAsFactors = FALSE, check.names = FALSE)
+mean_growth_parameters <- read.csv("tables_figures/phenology_parameters.csv",  col.names = c("", "ring-porous","diffuse-porous","ring-porous", "diffuse-porous"), stringsAsFactors = FALSE, check.names = FALSE)
 kable(mean_growth_parameters, caption = mean_growth_parameters_caption ,  format = "latex", booktabs = TRUE,  escape = F) %>%
@@
     column_spec(1, width = "4cm") %>%
-    pack_rows("Stem Growth", 1, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    pack_rows("Stem Growth", 2, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
     pack_rows("Canopy Foliage (ecosystem level)", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-    pack_rows("Phenological sensitivity to mean max. T in critical T window (days/°C)", 13, 22, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
+    pack_rows("Phenological temperature sensitivity* (days/°C)", 13, 22, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
     add_indent(14:17)%>%
@@
 ```
+*Phenological temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig. 1).
 

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
fix up extended data table 3

@@
 mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
-tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations to monthly $T_{max}$ in univariate and multivariate analyses."
+tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."
 
@@
 add_header_above(c(" " = 2,  "univariate analysis" = 10,"multivariate analysis" = 4)) %>%
-add_header_above(c(" " = 2,  "n. correlations to monthly max. T significant at p=0.05" = 14)) %>%
-    #column_spec(3:18, width = ".3cm") %>%
-pack_rows("Ring Porous", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-pack_rows("Diffuse Porous", 17, 26, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+add_header_above(c(" " = 2,  "n. correlations to monthly mean max. T significant at p=0.05" = 14)) %>%
+    column_spec(3:18, width = ".4cm") %>%
+pack_rows("Ring-porous", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+pack_rows("Diffuse-porous", 17, 26, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
 row_spec(27, bold = T) %>%

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
review Extended Data Fig. 1
standardize caption formatting

@@
                   \usepackage{caption}
-                  \captionsetup[table]{font=footnotesize}
-                  \captionsetup[figure]{font=footnotesize}
                   \captionsetup[figure]{labelformat=empty}
                   \captionsetup[table]{labelformat=empty}
-                  \usepackage{pdflscape}
-                  \newcommand{\blandscape}{\begin{landscape}}
-                  \newcommand{\elandscape}{\end{landscape}}
 ---
@@
 ![**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
-Sampling details are provided in **SI TBALE NAME**](tables_figures/chronology_map.png)
+Sampling details are provided in `r chron_table`.](tables_figures/chronology_map.png)
 
@@
 
-![**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+![**Extended Data Figure 4 | Response of the timing of stem growth phenology to mean maximum temperatures during the critial temperature window (CTW).** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
@@
 
-![**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).
-Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+![**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperatures in April.** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
@@
 
-![Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures of the current year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+![**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures of the current year for 208 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.
@@
 
-![Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1).
+![**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
 Chronologies are grouped by xylem porosity and ordered by mean April T_max.

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
review Extended Data Fig. 2

@@
 
-![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at our two focal study sites** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2) ..., with each line representing a year between 2000 and 2018.
-For stem growth, each line represents one year's growth for a given tree, fit with McMahon model.](tables_figures/growth_curves_all.png){ width=100% }
+![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smoithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
+For stem growth, each line represents the growth of one tree over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.](tables_figures/growth_curves_all.png){ width=100% }
 
@@
 
-![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest.** Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
+![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smoithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
 Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
review extended data fig 3

@@
 
-![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smoithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF).
-Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb.
-19- May 21 at SCBI, March 19 - May 07 at HF).](tables_figures/climwin_figure.png){ width=70% }
+![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smoithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
+Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`).
+Yellow shading indicates neutral relationships, while orange or red shading indicates that $DOY_{25}$ advances with increased $T_{max}$ over the given time window (negative $\beta$).
+Black circles indicate the critical temperature window selected based on minimization of $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model.
+Critical temperature windows are listed in `r EDT_phenology_parameters`.](tables_figures/climwin_figure.png){ width=70% }
 

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
reviewed Extended Data Figures 4-5

@@
 
-![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smoithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
+![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
 For stem growth, each line represents the growth of one tree over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.](tables_figures/growth_curves_all.png){ width=100% }
@@
 
-![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smoithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
+![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
 Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`).
@@
 
-![**Extended Data Figure 4 | Response of the timing of stem growth phenology to mean maximum temperatures during the critial temperature window (CTW).** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+![**Extended Data Figure 4 | Response of stem growth phenology and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
+Shown are relationships between mean $T_max$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
 95% credible intervals are represented by bands.
-The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective CTW.
-For each CTW $T_{max}$.
-Modeled values from dendrometer band measurements are represented by dots.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }
+For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective CTW.
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }
 
@@
 
-![**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperatures in April.** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown.
+![**Extended Data Figure 5 | Response of stem growth phenology and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_max$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
-95% credible intervals are represented by bands.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }
+95% credible intervals are represented by bands.
+For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI.
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }
 

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
improve captions for extended data figures 6-7

@@
 
+###~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+## some stats/ 
+species_table <- read.csv(file = "tables_figures/species_table.csv")
+n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
+n_dendro_trees_SCBI <- 123 #doing this the lazy way. summed from species_table rather than coded. 
+n_dendro_trees_HF <- 340 #doing this the lazy way. summed from species_table rather than coded. 
+n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
+n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
+chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
+n_chronologies <- nrow(chronology_table)  
+n_cores_sites <- length(unique(chronology_table[, "Location"]))
+n_cores_species <- length(unique(chronology_table[, "Species"]))
+n_chronologies_SCBI <- 12
+n_chronologies_HF <- 4
 ```
@@
 
-![**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures of the current year for 208 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
-Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmin.png){ width=90% }
+![**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean minimum temperatures ($T_{min}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the correlation between monthly $T_{min}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April $T_{min}$.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
+Species analyzed and numbers of significant correlations to $T_{min}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmin.png){ width=90% }
 
@@
 
-![**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures of the current and past year for 208 chronologies from 114 sites across eastern North America** (Extended Data Figure 1).
-Chronologies are grouped by xylem porosity and ordered by mean April T_max.
-Chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=90% }
+![**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$) of the current and past year, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
+Figure presents the same results as main manuscript Fig.
+`r tree_rings` but extends back to include the previous year.
+Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=90% }

### 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
further improve EDF 6-7
@camerondow35 , I think the extended data is good now, just need to address issues #90 , #88

@@
 ## some stats/ 
-species_table <- read.csv(file = "tables_figures/species_table.csv")
-n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
-n_dendro_trees_SCBI <- 123 #doing this the lazy way. summed from species_table rather than coded. 
-n_dendro_trees_HF <- 340 #doing this the lazy way. summed from species_table rather than coded. 
-n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
-n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
 chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
@@
 n_cores_species <- length(unique(chronology_table[, "Species"]))
-n_chronologies_SCBI <- 12
-n_chronologies_HF <- 4
+
 ```
@@
 Figure presents the same results as main manuscript Fig.
-`r tree_rings` but extends back to include the previous year.
+3 but extends back to include the previous year.
 Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=90% }

### 2021-11-17 doc/manuscript/Dow_Extended_Data.Rmd
add pheno variables to edt 2
#90

@@
 ```
-*Phenological temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig. 1).
+*Phenological temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
+1).
+Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straight forward.
+We used the critical window identified for diffuse-porous trees as the critical window for leaf phenology calculations because they included a wider range of spring dates.
 

### 2021-11-17 RScripts/DOYtiming_all_years.R

@@ -1,3 +1,5 @@
+#Script to create DOY_timing_allyears.png fig
+
 library(ggplot2)
 library(readr)
 library(tidyverse)
@@ -89,6 +91,29 @@ for(i in 2011:2020){
   aggregates_scbi_dp <- rbind(aggregates_scbi_dp,year_ag)
 }
 
+# #MAX RATE DOY ----
+# aggregates_scbi_rp <- data.frame(1,1,1,1,1)
+# names(aggregates_scbi_rp) <- c("Group.1", "Group.2", "x","year", "window_temp")
+# for(i in 2011:2020){
+#   year <- subset(Wood_pheno_table_scbi, year == i & wood_type == "ring-porous")
+#   year_ag <- aggregate(year$max_rate_DOY, by = list(year$wood_type, year$perc), FUN = mean)
+#   year_ag$Group.2 <- ifelse(year_ag$Group.2 == .25, 25, ifelse(year_ag$Group.2 == .50, 50, ifelse(year_ag$Group.2 == .75, 75, 0)))
+#   year_ag$year <- i
+#   year_ag$window_temp <- unique(year$climwinmean)
+#   aggregates_scbi_rp <- rbind(aggregates_scbi_rp,year_ag)
+# }
+#
+# aggregates_scbi_dp <- data.frame(1,1,1,1,1)
+# names(aggregates_scbi_dp) <- c("Group.1", "Group.2", "x","year","window_temp")
+# for(i in 2011:2020){
+#   year <- subset(Wood_pheno_table_scbi, year == i & wood_type == "diffuse-porous")
+#   year_ag <- aggregate(year$max_rate_DOY, by = list(year$wood_type, year$perc), FUN = mean)
+#   year_ag$Group.2 <- ifelse(year_ag$Group.2 == .25, 25, ifelse(year_ag$Group.2 == .50, 50, ifelse(year_ag$Group.2 == .75, 75, 0)))
+#   year_ag$year <- i
+#   year_ag$window_temp <- unique(year$climwinmean)
+#   aggregates_scbi_dp <- rbind(aggregates_scbi_dp,year_ag)
+# }
+#----
 aggregates_scbi_rp <- aggregates_scbi_rp[-1,]
 
 aggregates_scbi_rp$Temp <- ifelse(aggregates_scbi_rp$window_temp > median(aggregates_scbi_rp$window_temp), "Above Average",
[…]

### 2021-11-17 doc/manuscript/Dow_manuscript_Nature.Rmd
Add spei details

@@
 To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring tempreatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
-The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020]. 
-*ADD SOMETHING ON HOW STANDARD PRECIPITAITON EVAPOTRANSPIRATIN INDEX WAS FOUND*!!!!
+The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2017].
+Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures lead to greater summer drought stress.
 

### 2021-11-17 doc/manuscript/Dow_Extended_Data.Rmd
knit in updated EDT2
looks great; thanks @camerondow35 !
closes #90

--- 2021-11-16 doc/manuscript/Dow_Extended_Data.Rmd
@@
 ```
-*Phenological temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig. 1).
+*Phenological temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
+1).
+Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straightforward.
+We used the critical window identified for diffuse-porous trees as the critical window for leaf phenology calculations because they included a wider range of spring dates.
 

### 2021-11-17 doc/manuscript/Dow_Extended_Data.Rmd
updated tree-ring figures, knit into ED
#88

@@
 Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
-Species analyzed and numbers of significant correlations to $T_{min}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmin.png){ width=90% }
+Species analyzed and numbers of significant correlations to $T_{min}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmin.png){ width=100% }
 
@@
 3 but extends back to include the previous year.
-Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=90% }
+Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=100% }

### 2021-11-17 doc/manuscript/Dow_manuscript_Nature.Rmd
added SPEI refs + a bit of text
@camerondow35 , please verify added sentence.

#89

@@
 To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring tempreatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
-The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2017].
+The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
 Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures lead to greater summer drought stress. 
+No significant correlations were found (all p>0.05).
 

### 2021-11-18 doc/manuscript/Dow_manuscript_Nature.Rmd
add GH ORCID
GLH feedback

@@
 
-Grant L. Harley^5^
+Grant L. Harley^5^ ( Orcid ID : 0000-0003-1557-8465)
 
@@
 
-5. Department of Geography & Geological Sciences, University of Idaho, ID 83844, USA
+5. Department of Earth and Spatial Sciences, University of Idaho, ID 83844, USA
 
@@
 Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
+However, recent experimental and observational findings show that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
+These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
 
-While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], very little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
+While some studies addressing the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures exist[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
 In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021].
@@
 
-Tree-ring records, which can be used to examine correlations of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
+Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
 <!-- dorangeville_drought_2018 looks at PET.-->
@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 
-## Tree ring analysis {-#Tree}
+## Tree-ring analysis {-#Tree}
 
@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
-Across all chronologies, RWI was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring-porous and 4 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
+Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring-porous and 4 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
 In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
@@
 Extra C could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
-Extension of our tree-ring analysis revealed little correlation between April $T_{max}$ and growth the following year (sig.
+Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig.
 pos.
@@
 Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
-SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km South of Front Royal, Virginia. 
+SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km south of Front Royal, Virginia. 
 The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
@@
 
-For each species-site combination, we converted tree ring records into a dimensionless ring width index (RWI) to emphasize interannual variability associated with climate [@cook_methods_1990].
+For each species-site combination, we converted tree-ring records into the dimensionless RWI to emphasize interannual variability associated with climate [@cook_methods_1990].
 A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985; @cook_methods_1990].
@@
 The 1/3rds spline method was chosen when replication in the inner portion of each chronology (ca. inner 30–50 yr of each record depending on full chronology length) was less than three trees. 
-When replication was greater than that, we used the average correlation between raw ring-width series (rbar) method.
+When replication was greater than n = 3 trees, we used the average correlation between raw ring-width series (rbar) method.
 The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
@@
 We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
-To test for potential lag effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
+To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
 Correlations and significance levels for months April-August are given in `r chron_table`.

### 2021-11-18 doc/manuscript/Dow_manuscript_Nature.Rmd
add affiliation
add a bit of info to Fig. 1 caption

@@
 
-Cameron Dow^1^ ( Orcid ID : 0000-0002-8365-598X)
+Cameron Dow^1,11^ ( Orcid ID : 0000-0002-8365-598X)
 
@@
 10. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+
+11. Department of Forestry and Natural Resources, Purdue University, West Lafayette, Indiana, USA
 
@@
 
-![**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 

### 2021-11-18 doc/manuscript/Dow_manuscript_Nature.Rmd
improve wording

@@
 We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW).
-CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 and $DOY_{25}$ (`r EDF_climwin`).
+CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem porosity-site combination (`r EDF_climwin`).
 The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

### 2021-11-18 doc/manuscript/Dow_manuscript_Nature.Rmd
wording change

@@
 
-While some studies addressing the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures exist[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
+While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
 In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021].

### 2021-11-18 doc/manuscript/Dow_manuscript_Nature.Rmd
fix author affiliation order
one-word change to make statement technically correct
Zohner study looks at temperate deciduous species over > 1 year, so "forests" is appropriate here.

@@
 
-Cameron Dow^1,11^ ( Orcid ID : 0000-0002-8365-598X)
+Cameron Dow^1,2^ ( Orcid ID : 0000-0002-8365-598X)
 
-Albert Y. Kim^1,2^ ( Orcid ID : 0000-0001-7824-306X)
+Albert Y. Kim^1,3^ ( Orcid ID : 0000-0001-7824-306X)
 
-Loïc D’Orangeville^3,4^ ( Orcid ID : 0000-0001-7841-7082)
+Loïc D’Orangeville^4,5^ ( Orcid ID : 0000-0001-7841-7082)
 
@@
 
-Grant L. Harley^5^ ( Orcid ID : 0000-0003-1557-8465)
+Grant L. Harley^6^ ( Orcid ID : 0000-0003-1557-8465)
 
-Justin T. Maxwell^6^ (Orcid ID: 0000-0001-9195-3146)
+Justin T. Maxwell^7^ (Orcid ID: 0000-0001-9195-3146)
 
-Ian R. McGregor^1,7^ (Orcid ID: 0000-0002-5763-021X)
+Ian R. McGregor^1,8^ (Orcid ID: 0000-0002-5763-021X)
 
@@
 
-Sean McMahon^8,10^ (Orcid ID : 0000-0001-8302-6908)
+Sean McMahon^9,11^ (Orcid ID : 0000-0001-8302-6908)
 
-Neil Pederson^3^ (Orcid ID : 0000-0003-3830-263X)
+Neil Pederson^4^ (Orcid ID : 0000-0003-3830-263X)
 
-Alan J. Tepley^1,9^ ( Orcid ID : 0000-0002-5701-9613)
+Alan J. Tepley^1,10^ ( Orcid ID : 0000-0002-5701-9613)
 
-Kristina J. Anderson-Teixeira^1,10^* ( Orcid ID : 0000-0001-8461-9713)
+Kristina J. Anderson-Teixeira^1,11^* ( Orcid ID : 0000-0001-8461-9713)
 
@@
 
-2. Statistical & Data Sciences; Smith College; Northampton, MA 01063, USA
+2. Department of Forestry and Natural Resources, Purdue University, West Lafayette, Indiana, USA
 
-3. Harvard Forest, Petersham, MA 01366, USA
+3. Statistical & Data Sciences; Smith College; Northampton, MA 01063, USA
 
-4. Faculty of Forestry and Environmental Management, University of New Brunswick, Fredericton, NB, E3B 5A3, Canada.
+4. Harvard Forest, Petersham, MA 01366, USA
 
-5. Department of Earth and Spatial Sciences, University of Idaho, ID 83844, USA
+5. Faculty of Forestry and Environmental Management, University of New Brunswick, Fredericton, NB, E3B 5A3, Canada.
 
-6. Department of Geography, Indiana University, Bloomington, IN 47405, USA
+6. Department of Earth and Spatial Sciences, University of Idaho, ID 83844, USA
 
-7. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
+7. Department of Geography, Indiana University, Bloomington, IN 47405, USA
 
-8. Smithsonian Environmental Research Center, Edgewater, MD, USA
+8. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
 
-9. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+9. Smithsonian Environmental Research Center, Edgewater, MD, USA
 
-10. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+10. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
 
-11. Department of Forestry and Natural Resources, Purdue University, West Lafayette, Indiana, USA
+11. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
 
@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021].
+In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous forests over multiple years[@dorangeville_peak_2021].
 <!-- I'm not aware of any others, but we do want to make sure we're not missing anything.

### 2021-11-18 doc/manuscript/2021_11_18_Nature/cover letter.Rmd
turns out a cover letter is needed...

---
title: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

November 19, 2021
\newline
\newline

Dear Editor:

We are pleased to submit our manuscript "Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity", by Cameron Dow and co-authors, for consideration for publication in *Nature*.

As the climate warms, it is generally expected that temperate deciduous forests, which accounts for ~30% of the global forest carbon sink, will increase their carbon dioxide (CO~2~) sequestration due to lengthened growing seasons. 
This expectation is based on observations that warm springs result in earlier leaf-out and commencement of net CO~2~ uptake; however, little is known about how spring temperatures affect tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem.
In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous forests over multiple years (and this one did not consider the influence of spring temperatures on phenology).
Here, using dendrometer band measurements from 463 trees across two forests, we show that warmer spring temperatures shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth. 
We confirmed the latter finding on the centennial scale using 207 tree-ring chronologies from 108 forests across eastern North America, where annual growth was far sensitive to temperatures during the peak growing season but not to spring temperatures. 
Our findings imply that extra CO~2~ uptake in years with warmer springs is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
Rather, contradicting current projections from global C cycle models, our empirical results imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.
In short, our findings imply that the potential for temperate deciduous forests to mitigate the climate crisis is probably over-estimated.
This is a significant finding that will be of interest to the broad readership of *Nature*.

We believe our manuscript to be an appropriate length for *Nature*, with 2,597 words, 3 display items, and 48 references in the main article. The article is accompanied by an Extended Data file (3 tables, 7 figures) and a supplementary table.

Thank you for considering our submission for *Nature*.

Sincerely,
\newline
\newline

Kristina Anderson-Teixeira, on behalf of all co-authors.

### 2021-11-29 doc/manuscript/Dow_manuscript_Nature.Rmd
adjust/update C sink statistics
fixes incorrect Pugh statistic (#92), although I'm still not sure we want to cite the 300 Tg C/yr statistic at all.

update from Pan to Harris citation

cites in-review 2021 global C budget

@@
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at ~2652 words and 1 modest + 2 large display items.
+We're currently at 2597 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently ok there-- n= 48)-->
@@
 
-In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011; @friedlingstein_global_2020].
-A large portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ (>30% of the total forest C sink)[@pugh_role_2019].
-<!-- The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al.
+In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@friedlingstein_global_2021; @harris_global_2021].
+A substantive portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ [@pugh_role_2019].
+<!-- See issue # 92: https://github.com/EcoClimLab/growth_phenology/issues/92.
+ALSO: The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al.
 2019 estimate that ~200 Tg C yr-1 is attributable to demography (forest regrowth), while >100 Tg C yr-1 is controlled by environmental drivers.

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
add placeholder for data and code availability statements
#93
Update Dow_manuscript_Nature.Rmd

@@
 `r schematic_summary`).
-This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021].
+This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in growth phenology[@dorangeville_peak_2021].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 Slopes and p-values for each chronology are given in `r chron_table`.
+Although some models may have benefited from data transformations, we determined that assumptions of normality and homoscedasticity were sufficiently met for the purposes of this analysis.
 
@@
 
+# Data Availability Statement {-#Data}
+
+# Code Availability Statement {-#Code}
+
 # Additional Information {-#Additional}

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
draft code availability statement
@camerondow35 , could you please add version numbers for the R packages we used and add any I missed?

@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
+Correlations between monthly climate and $RWI$ were assessed using *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
 We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
@@
 # Code Availability Statement {-#Code}
+Data were analysed in the open source statistical software R (version 4.0).
+We used packages *Rdendrom* (https://github.com/seanmcm/RDendrom/), *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html), *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008], *dplR* **version###**, and *bootRes* **version###**.
+All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: TBD).
 

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 We used packages *Rdendrom* (https://github.com/seanmcm/RDendrom/), *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html), *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008], *dplR* **version###**, and *bootRes* **version###**.
-All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: TBD).
+All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: [TBD]).
 

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@teixeirak rdendrom functions were used, not the package as a whole because it wasn't available on CRAN.  Climpact is an R package or a separate software, I used the separate software. We should probably clarify this throughout the paper. Other than that versions look good

@@
 Data were analysed in the open source statistical software R (version 4.0).
-We used packages *Rdendrom* (https://github.com/seanmcm/RDendrom/), *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html), *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008], *dplR* **version###**, and *bootRes* **version###**.
+We used packages *Rdendrom* (https://github.com/seanmcm/RDendrom/), *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html), *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008], *dplR* v.1.0.2, and *bootRes* v1.2.4.
 All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: [TBD]).

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
working on code/data statements
#93

@@
 
-<!--Climate -->
+<!--Weather data -->
 For the dendrometer band and leaf phenology analyses, climate data corresponding to the measurement periods were obtained from local weather stations at each focal site. 
@@
 We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
-Thanks to David A. Orwig and members of the ForestGEO Ecosystesms & Climate Lab at SCBI for helpful feedback.
+Thanks to David A. Orwig and members of the ForestGEO Ecosystems & Climate Lab at SCBI for helpful feedback.
 The research was funded by ForestGEO (Smithsonian). 
@@
 # Data Availability Statement {-#Data} 
+Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail), downloaded on [DATE].
+Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive) <!--specify link, version-->.
+Climate data were obtained from CRU v.4.04 via the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
 
@@
 We used packages *Rdendrom* (https://github.com/seanmcm/RDendrom/), *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html), *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008], *dplR* v.1.0.2, and *bootRes* v1.2.4.
-All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: [TBD]).
+All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).
 

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
#93
@camerondow35 , please confirm these changes to code statement

@@
 Data were analysed in the open source statistical software R (version 4.0).
-We used packages *Rdendrom* (https://github.com/seanmcm/RDendrom/), *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html), *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008], *dplR* v.1.0.2, and *bootRes* v1.2.4.
+We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
+We used *climpact* software (see www.climpact-sci.org) <!--specify version-->.
 All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
add spei + leaf pheno data statement

@@
 Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail), downloaded on [DATE].
-Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive) <!--specify link, version-->.
+Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive) (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF001 AND https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF000).
 Climate data were obtained from CRU v.4.04 via the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
+The Standardised Precipitation-Evapotranspiration Index was obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
+Canopy foliage phenology data were extracted from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a. MODIS Global Vegetation Phenology product) via Google Earth Engine (link).
 
@@
 We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
-We used *climpact* software (see www.climpact-sci.org) <!--specify version-->.
+We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
 All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@teixeirak added a link for leaf pheno data but not sure if it's the correct one
Update Dow_manuscript_Nature.Rmd
fix link

@@
 Climate data were obtained from CRU v.4.04 via the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
-The Standardised Precipitation-Evapotranspiration Index was obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
-Canopy foliage phenology data were extracted from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a. MODIS Global Vegetation Phenology product) via Google Earth Engine (link).
+The Standardised Precipitation-Evapotranspiration Index was obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/SPEI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
+Canopy foliage phenology data were extracted from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
+MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#bands).
 

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 Data were analysed in the open source statistical software R (version 4.0).
-We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
+We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
 We used *climpact* software v.1.2.8 (see www.climpact-sci.org).

### 2021-12-03 doc/manuscript/Dow_Extended_Data.Rmd
fix to caption subscripts

@@
 ![**Extended Data Figure 4 | Response of stem growth phenology and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
-Shown are relationships between mean $T_max$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
+Shown are relationships between mean $T_{max}$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
@@
 
-![**Extended Data Figure 5 | Response of stem growth phenology and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_max$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
+![**Extended Data Figure 5 | Response of stem growth phenology and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
#96
#93
fix data availability statement
#93 (I previously failed to copy over content I had written to the master.)

@@
 # Data Availability Statement {-#Data} 
+The datasets generated and analysed during the current study are available via GitHub in the *growth_phenology* repository of the ForestGEO Ecosystems & Climate Lab @ SCBI, (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI *[TBD]*).
+Master versions of the dendrometer band data are available for SCBI via GitHub in the *Dendrobands* repository of the Smithsonian Conservation Biology Institute ForestGEO plot (https://github.com/SCBI-ForestGEO/Dendrobands), which is archived in Zenodo (DOI 10.5281/zenodo.5551143), and for Harvard Forest via the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF149).
 Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail), downloaded on [DATE].
-Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive) (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF001 AND https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF000).
+Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF001 AND https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF000).
 Climate data were obtained from CRU v.4.04 via the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
@@
 Canopy foliage phenology data were extracted from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
-MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#bands).
+MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description).
+Original tree cores are archived at the institutions of various members of the author team and will be made available upon reasonable request. 
+<!-- do we need to add ITRDB??-->
 

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
minor

@@
 The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
-Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures lead to greater summer drought stress. 
+Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures were associated with greater summer drought stress in our data set.
 No significant correlations were found (all p>0.05).

### 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
fix spelling error

@@
 
-To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring tempreatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
+To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
 The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].

### 2021-12-03 doc/submissions/2021_11_18_Nature/Dow_manuscript_Nature.Rmd
move folder

--- 2021-12-03 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 
-In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@friedlingstein_global_2021; @harris_global_2021].
-A substantive portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ [@pugh_role_2019].
-<!-- See issue # 92: https://github.com/EcoClimLab/growth_phenology/issues/92.
-ALSO: The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al.
+In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011; @friedlingstein_global_2020].
+A large portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ (>30% of the total forest C sink)[@pugh_role_2019].
+<!-- The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al.
 2019 estimate that ~200 Tg C yr-1 is attributable to demography (forest regrowth), while >100 Tg C yr-1 is controlled by environmental drivers.
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in growth phenology[@dorangeville_peak_2021].
+This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 
-<!--Weather data -->
+<!--Climate -->
 For the dendrometer band and leaf phenology analyses, climate data corresponding to the measurement periods were obtained from local weather stations at each focal site. 
@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and $RWI$ were assessed using *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
+Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
 We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
@@
 Slopes and p-values for each chronology are given in `r chron_table`.
-Although some models may have benefited from data transformations, we determined that assumptions of normality and homoscedasticity were sufficiently met for the purposes of this analysis.
 
@@
 We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
-Thanks to David A. Orwig and members of the ForestGEO Ecosystems & Climate Lab at SCBI for helpful feedback.
+Thanks to David A. Orwig and members of the ForestGEO Ecosystesms & Climate Lab at SCBI for helpful feedback.
 The research was funded by ForestGEO (Smithsonian). 
@@
 
-# Data Availability Statement {-#Data} 
-The datasets generated and analysed during the current study are available via GitHub in the *growth_phenology* repository of the ForestGEO Ecosystems & Climate Lab @ SCBI, (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI *[TBD]*).
-Master versions of the dendrometer band data are available for SCBI via GitHub in the *Dendrobands* repository of the Smithsonian Conservation Biology Institute ForestGEO plot (https://github.com/SCBI-ForestGEO/Dendrobands), which is archived in Zenodo (DOI 10.5281/zenodo.5551143), and for Harvard Forest via the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF149).
-Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail), downloaded on [DATE].
-Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF001 AND https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF000).
-Climate data were obtained from CRU v.4.04 via the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
-The Standardised Precipitation-Evapotranspiration Index was obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/SPEI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
-Canopy foliage phenology data were extracted from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
-MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description).
-Original tree cores are archived at the institutions of various members of the author team and will be made available upon reasonable request. 
-<!-- do we need to add ITRDB??-->
-
-# Code Availability Statement {-#Code}
-Data were analysed in the open source statistical software R (version 4.0).
-We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
-We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
-All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).
-
 # Additional Information {-#Additional}

### 2021-12-10 doc/manuscript/Dow_manuscript_Nature.Rmd
add Etzold reference
#101

@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous forests over multiple years[@dorangeville_peak_2021].
+In fact, we are aware of only two studies that have documented stem-growth phenology of temperate deciduous forests over multiple years[@dorangeville_peak_2021; @etzold_number_2022].
 <!-- I'm not aware of any others, but we do want to make sure we're not missing anything.

### 2021-12-13 doc/manuscript/Dow_manuscript_Nature.Rmd
add references to Etzold
#101

refs added after thorough reading of Etzold

@@
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.
-**
+**These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.**
 **Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
@@
 klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
-The climate sensitivity of woody growth phenology in temperate deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).
+**The climate sensitivity of woody growth phenology in temperate deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).** <!-- revise this statement.
+Etzold et al kind of do this.-->
 
@@
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
 Our tree-ring analysis (Fig.
-`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual tree growth is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018].
+`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual tree growth is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018; @etzold_number_2022].
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on growth phenology elucidated here (Fig.
@@
 
-Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
+Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with *(Etzold)*[@etzold_number_2022] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
 However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].

### 2022-01-24 doc/manuscript/Dow_manuscript_Nature.Rmd
added Tumajer et al. 2022 ref
@camerondow35 , FYI

@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we are aware of only two studies that have documented stem-growth phenology of temperate deciduous forests over multiple years[@dorangeville_peak_2021; @etzold_number_2022].
+In fact, we are aware of only a few studies that have documented stem-growth phenology of temperate deciduous forests over multiple years[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022].
 <!-- I'm not aware of any others, but we do want to make sure we're not missing anything.
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in growth phenology[@dorangeville_peak_2021].
+This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in growth phenology[@dorangeville_peak_2021; @tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

### 2022-02-01 doc/manuscript/Dow_manuscript_Nature.Rmd
small fixes related to R1 comments

@@
 
-Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
-<!-- dorangeville_drought_2018 looks at PET.-->
-with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].
-<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
+Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
 Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
@@
 `r DOY_timing`, `r EDF_pheno_all`).
-We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW).
+We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperature window (CTW).
 CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem porosity-site combination (`r EDF_climwin`).

### 2022-02-01 doc/manuscript/Dow_manuscript_Nature.Rmd
Some rewording
in general, change mentions of woody biomass to stem-growth/ring-width

@@
 **However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
-**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
+**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
 **These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.**
-**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase the woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase stem growth or strengthen the CO~2~ sink of temperate deciduous forests.**
 
@@
 
-While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures.
-In fact, we are aware of only a few studies that have documented stem-growth phenology of temperate deciduous forests over multiple years[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022].
-<!-- I'm not aware of any others, but we do want to make sure we're not missing anything.
+While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures.
+In fact, we are aware of only a few studies that have documented stem-growth timing of temperate deciduous trees over multiple years[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022].
+<!-- Could delete this sentence?
+I'm not aware of any others, but we do want to make sure we're not missing anything.
 klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
-**The climate sensitivity of woody growth phenology in temperate deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).** <!-- revise this statement.
+**However, the climate sensitivity of stem growth timing in temperate deciduous trees and its link to annual growth has seen a recent surge in interest within published literature[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; zweifel_why_2021]** <!-- revise this statement.
 Etzold et al kind of do this.-->
@@
 
-Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs.
+Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on annual ring-width (Figs.
 `r schematic_summary`- `r tree_rings`).
-The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
+The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
+However, inconsistent with the concept that an earlier start to growth would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
 Our tree-ring analysis (Fig.
-`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual tree growth is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018; @etzold_number_2022].
-Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on growth phenology elucidated here (Fig.
+`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018; @etzold_number_2022].
+Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
 `r schematic_summary`) were not attributable to summer drought.
 
-Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with *(Etzold)*[@etzold_number_2022] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
+Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
 However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
-We show that the extra C fixation in years with warm springs does not substantially augment woody growth, but it remains unclear how it is allocated within the ecosystem.
+We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

### 2022-02-01 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
add Sean's response to R2

@@
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
+However, inconsistent with the concept that an earlier start to the growing season would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
 `r tree_rings`).
@@
 Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.
-If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
+If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
 Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.
@@
 and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
+This allows an asymetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
 The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.

### 2022-02-01 RScripts/data_cleaning.R
Using 2.5sd instead of 2

@@ -17,7 +17,7 @@ library(scales)
 library(tidybayes)
 
 # Number of standard deviations used to identify "outliers"
-sd <- 2
+sd <- 2.5
 
 # Load Sean McMahon's Dendroband functions
 source("RScripts/dendroband_functions.R")

### 2022-02-01 RScripts/data_cleaning.R
Using 3sd instead of 2sd

@@ -17,7 +17,7 @@ library(scales)
 library(tidybayes)
 
 # Number of standard deviations used to identify "outliers"
-sd <- 2
+sd <- 3
 
 # Load Sean McMahon's Dendroband functions
 source("RScripts/dendroband_functions.R")

### 2022-02-01 RScripts/Pheno_tsensitivity_figure_april_april_tmax.R
Note for later

@@ -176,7 +176,7 @@ joint_model_climwinmeans <- stan_mvmer(
 bayesian_regression_table <- joint_model_climwinmeans %>%
   summary() %>%
   as_tibble(rownames = "coefficient") %>%
-  # Keep only relevant columns:
+  # Keep only relevant columns: Note for Bert: does this need to be changed to reflect 2.5SD?
   select(coefficient, mean, sd, `2.5%`, `97.5%`) %>%
   # Keep only relevant rows:
   filter(!str_detect(coefficient, "Sigma")) %>%

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
addressing R2
changes addressing R2 comments

@@
 The latest end date was November 11, 2002.
+In our analysis, each band-year was treated independently, with no data overlap from one year to the next.
 
@@
 We screened the data for three classes of errors. 
-First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and the datapoint was removed.
+First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and removed.
 Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck on the tree bark and then released.
@@
 These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year.
-We then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).
+An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015; @herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest.
+Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig.
+`r schematic_summary`).
 

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
More rewording. @teixeirak i deleted one of the sentences around line 164. feel free to change back if you disagree

@@
 Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings show that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
+However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
 These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures.
-In fact, we are aware of only a few studies that have documented stem-growth timing of temperate deciduous trees over multiple years[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022].
-<!-- Could delete this sentence?
-I'm not aware of any others, but we do want to make sure we're not missing anything.
-klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
-**However, the climate sensitivity of stem growth timing in temperate deciduous trees and its link to annual growth has seen a recent surge in interest within published literature[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; zweifel_why_2021]** <!-- revise this statement.
-Etzold et al kind of do this.-->
+This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021].
 
@@
 Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
-Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
+Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
 
-Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
-To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
+Here, we evaluate how early spring temperatures affect peak stem growth timing, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
+To test whether warmer springs extend the period of peak stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
 To test whether spring temperatures consistently increased annual growth, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
@@
 
-Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig.
+Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig.
 `r DOY_timing`, `r EDF_pheno_all`).
-We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperature window (CTW).
+We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW).
 CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem porosity-site combination (`r EDF_climwin`).
@@
 
-Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on annual ring-width (Figs.
+Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs.
 `r schematic_summary`- `r tree_rings`).
@@
 Our tree-ring analysis (Fig.
-`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018; @etzold_number_2022].
+`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018].
+This decoupling between growing season length and annual growth mirrors results from another recent study by Etzold et al[@etzold_number_2022] which found that vapor pressure deficit plays a key role in limiting stem growth *- other studies which found VPD limits growth-*.
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
address an R2 concern

@@
 Dominant tree species were cored at both SCBI[@helcoski_growing_2019;@bourg_initial_2013] and Harvard Forest[@alexander_potential_2019; @dye_comparing_2016; @finzi_carbon_2020] following sampling designs that covered a broad range of DBH.
-We analyzed records for the ring- and diffuse- porous species at each site (`r  EDT_species`), but excluded species with other xylem architectures (*Juglans nigra* L.
-at SCBI, *Tsuga canadensis* at Harvard Forest).
+We analyzed records for the ring- and diffuse- porous species at each site (`r  EDT_species`), but excluded semi-ring porous species (e.g., *Juglans nigra* L.
+at SCBI) and conifers (e.g., *Tsuga canadensis* at Harvard Forest).
 We studied a total of `r n_cores_focal_sites` cores which included `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r  EDT_species`).

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@camerondow35 , I added a bit back here to make sure the paragraph is setting up why our study is unique. Feel free to tweak.

@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures.
-This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021].
+This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
 

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
edit abstract as suggested by R3

@@
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance.**
+**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
 **Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase stem growth or strengthen the CO~2~ sink of temperate deciduous forests.**

### 2022-02-03 RScripts/Pheno_tsensitivity_figure_april_april_tmax.R
remove comment

@@ -176,7 +176,7 @@ joint_model_climwinmeans <- stan_mvmer(
 bayesian_regression_table <- joint_model_climwinmeans %>%
   summary() %>%
   as_tibble(rownames = "coefficient") %>%
-  # Keep only relevant columns: Note for Bert: does this need to be changed to reflect 2.5SD?
+  # Keep only relevant columns:
   select(coefficient, mean, sd, `2.5%`, `97.5%`) %>%
   # Keep only relevant rows:
   filter(!str_detect(coefficient, "Sigma")) %>%

### 2022-02-03 RScripts/data_cleaning.R
Mirror @camerondow35 changes in 2.5SD branch

@@ -305,7 +305,7 @@ fig3_scbi
 
 # Harvard Forest ---------------------------------------------------------------
 ## Load up percent growth DF ----
-sd <- 2
+sd <- 3
 Wood_pheno_table_hf <- read_csv("Data/dendrobands/HF/modeled/Wood_pheno_table_HarvardForest_RAW.csv") %>%
   # Keep only RP and DP for now
   filter(wood_type != "other") %>%

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
reword spei statement
R1 comment

@@
 This decoupling between growing season length and annual growth mirrors results from another recent study by Etzold et al[@etzold_number_2022] which found that vapor pressure deficit plays a key role in limiting stem growth *- other studies which found VPD limits growth-*.
-Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
+Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].
+However, we did not find evidence that higher spring temperatures were correlated with more summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
 `r schematic_summary`) were not attributable to summer drought.

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
add several citations and Place holder text linking stem growth and woody productivity

@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures.
-This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
+This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
+
+*<PLACEHOLDER>* Three cambial processes drive the relationship between stem growth and woody productivity: cell differentiation, cell enlargement, and cell wall thickening.
+Irreversible stem expansion occurs as a result of differentiation (20% of stem growth) and expansion (80% of stem growth), and accounts for ~10% of annual woody productivity[@cuny_woody_2015].
+The remaining 90% of woody productivity is caused by cell wall thickening.
+It would stand to reason that years with identical stem growth have a similar number of cells which are similarly sized (increased latewood can be an issue here, as you acknowledge).
+If we assume that cell wall thickness is controlled by processes within the tree (ie.
+hormonal responses instead of external climate drivers), then we can predict that cell wall thickness is relatively uniform within each cell type (early wood and late wood).
+Therefore, years with a similar amount of stem growth would sequester a similar amount of carbon during cell wall thickening, and we can say that stem growth equates to woody productivity.
 
@@
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].
-However, we did not find evidence that higher spring temperatures were correlated with more summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
+However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
 `r schematic_summary`) were not attributable to summer drought.

### 2022-02-03 RScripts/data_cleaning.R
Heads up @camerondow35 this code was giving me an error so I commented it out

@@ -568,8 +568,8 @@ rel_growth_hf
 
 
 
-fig3_hf <- ggplot(percent_growth_SL, aes(x = doy, y = dbh_growth_percent_cummulative, group = tag_year, col = wood_type)) +
-  geom_line(alpha = 0.2) +
-  scale_y_continuous(labels = percent) +
-  labs(x = "DOY", y = "Cummulative percent of total growth", title = "Cummulative percent of total (modeled) growth in diameter")
-fig3_hf
+# fig3_hf <- ggplot(percent_growth_SL, aes(x = doy, y = dbh_growth_percent_cummulative, group = tag_year, col = wood_type)) +
+#   geom_line(alpha = 0.2) +
+#   scale_y_continuous(labels = percent) +
+#   labs(x = "DOY", y = "Cummulative percent of total growth", title = "Cummulative percent of total (modeled) growth in diameter")
+# fig3_hf

### 2022-02-04 doc/manuscript/Dow_manuscript_Nature.Rmd
reinstate "woody productivity" in abstract
@camerondow35 , I want to hold off on changing this. I'm not sure it's necessary and want to think about it. Note that (1) if this changes, our title needs to change, (2) linking growth to woody productivity is critical to tie this to the C cycle.

@@
 **These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
-**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase stem growth or strengthen the CO~2~ sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
 

### 2022-02-04 doc/manuscript/Dow_manuscript_Nature.Rmd
statement on 2.5 and 3 sd runs
Reviewer 1 suggestions

--- 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 **As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
+**However, less is known about how spring temperatures affect tree stem growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
 **Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
@@
 
-While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures.
-This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
+While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021].
+This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
 
@@
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
-Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }
+Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }
 
@@
 We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW).
-CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem porosity-site combination (`r EDF_climwin`).
+CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
 The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.
@@
 `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018].
-This decoupling between growing season length and annual growth mirrors results from another recent study by Etzold et al[@etzold_number_2022] which found that vapor pressure deficit plays a key role in limiting stem growth *- other studies which found VPD limits growth-*.
+This decoupling between growing season length and annual growth mirrors results from other recent studies which found that vapor pressure deficit plays a key role in limiting stem growth[@etzold_number_2022; @delpierre_wood_2016].
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].
@@
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
+Cite delpierre 2016 around here. Delpierre et al. 2016 found that stem growth was not correlated with carbon inputs, agreeing with the results here. However, the authors 
 We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
@@
 
-We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define phenological dates and growth rates (Fig.
+We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig.
 `r schematic_summary`).
@@
 Modeled fits for tree-years were removed under five conditions:
-(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
+(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each xylem architecture (SCBI = 2, Harvard Forest = 34); 
 (2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 53, Harvard Forest = 106); 
@@
 At SCBI, a higher proportion of ring-porous trees were removed, the majority falling under condition 4.
+This process was repeated using 2.5 and 3 standard deviations and the results of analyses were not qualitatively different.
 

### 2022-02-06 doc/manuscript/Dow_manuscript_Nature.Rmd
add statement on tree agree and a few other wording changes

@@
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to the growing season would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
+However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
+3).
+We demonstrate that warmer springs either have little effect on, or even hasten, the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
 `r tree_rings`).
@@
 `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018].
-This decoupling between growing season length and annual growth mirrors results from other recent studies which found that vapor pressure deficit plays a key role in limiting stem growth[@etzold_number_2022; @delpierre_wood_2016].
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].
@@
 
-Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
+Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
 However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
@@
 Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest (`r EDT_species`).
+While we do not estimate the ages of trees in our sample, bands at both sites were placed on individuals of differing sizes in an attempt to measure trees across a range of ages.
 Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 

### 2022-02-06 doc/manuscript/Dow_manuscript_Nature.Rmd
suggested change to wording of leaf pheno data description

@@
 MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019] via Google Earth Engine.
-Extracted pixels were those containing the NEON tower at each site. 
-Using daily MODIS 2-band Enhanced Vegetation Index data (EVI2) at a spatial resolution of 500m, the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
+For each year at each site, we extracted data from the pixel containing the center of each forest plot (resolution of 500m).
+Using the daily MODIS 2-band Enhanced Vegetation Index data (EVI2) , the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
+Data points were kept if they were flagged as "good" or "best" quality.
 

### 2022-02-06 doc/manuscript/Dow_manuscript_Nature.Rmd
update fig 2 caption

@@
 ![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+Canopy foliage variables were extracted from a single pixel at the center of each forest plot.
+Only data flagged as "good" or "best" quality were included in the calculation of these phenology variables.
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
@@
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
 Our tree-ring analysis (Fig.
-`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018].
+`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].

### 2022-02-16 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
add explanation of a, b, and 1998 at hf

@@
 Timing of measurements at Harvard Forest were similar, with the median start date of April 23 and median end date of October 30. 
-1998 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15. 
+1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15. 
 The latest end date was November 11, 2002.
@@
 $$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
-Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to DBH at the beginning and end of the year, respectively. 
+Here, $L$ and $K$ are lower and upper asymptotes of the model. 
 $DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
@@
 The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
+The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year.
+We used the first and last manual dendrometer band measurements as starting points to calculate these two variables.
+With this method, initial measurements too late in the growing season which miss the start of the growing season can cause incorrect estimations of $a$ and $b$.
+The 1998 survey at Harvard Forest recorded first measurements on May 26, which is less than 2 weeks away from the average DOY25 at this site.
+This late start made the calculation of timing and total growth variables unreliable, so we removed tree years in 1998 from our analysis.
 The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015].
@@
 (2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 53, Harvard Forest = 106); 
-(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.02$ mm; SCBI = 0, Harvard Forest = 66); 
-(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 150, Harvard Forest = 199); 
-(5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3); 
-and (6) poorly fit models that did not meet any of the above criteria (SCBI = 2, Harvard Forest = 0).
-At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
-At SCBI, a higher proportion of ring-porous trees were removed, the majority falling under condition 4.
+(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 0, Harvard Forest = 66); 
+(4) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3); 
+and (5) poorly fit models that did not meet any of the above criteria (SCBI = 2, Harvard Forest = 0).
+At *both sites* the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
 This process was repeated using 2.5 and 3 standard deviations and the results of analyses were not qualitatively different.

### 2022-02-16 doc/manuscript/Dow_manuscript_Nature.Rmd
update number of removed tags
this is with 2 sd for all but condition 2 DOY variables, which use 3 sd. Going with a standard 3 here for now because 2 removed a lot of tags

@@
 Modeled fits for tree-years were removed under five conditions:
-(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each xylem architecture (SCBI = 2, Harvard Forest = 34); 
-(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 53, Harvard Forest = 106); 
-(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 0, Harvard Forest = 66); 
-(4) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3); 
-and (5) poorly fit models that did not meet any of the above criteria (SCBI = 2, Harvard Forest = 0).
-At *both sites* the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
+(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each xylem architecture (SCBI = 26, Harvard Forest = 24); 
+(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 3$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 71, Harvard Forest = 226);
+(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 8, Harvard Forest = 137); 
+(4) tree years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 20, HF = 6).
+At both sites the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
 This process was repeated using 2.5 and 3 standard deviations and the results of analyses were not qualitatively different.

### 2022-02-16 RScripts/data_cleaning.R

@@ -110,19 +110,24 @@ percent_growth <- Wood_pheno_table_scbi %>%
 Wood_pheno_table_scbi <- read_csv("Data/dendrobands/SCBI/modeled/wood_table_scbi_updated.csv")
 Wood_pheno_table_scbi <- Wood_pheno_table_scbi[,-1]
 
-means <- aggregate(Wood_pheno_table_scbi$doy, by = list(Wood_pheno_table_scbi$perc, Wood_pheno_table_scbi$wood_type), FUN = mean)
-SD <- aggregate(Wood_pheno_table_scbi$doy, by = list(Wood_pheno_table_scbi$perc, Wood_pheno_table_scbi$wood_type), FUN = sd)
-names(SD) <- c("Group.1", "Group.2", "sd")
+means <- aggregate(Wood_pheno_table_scbi$doy, by = list(Wood_pheno_table_scbi$perc,Wood_pheno_table_scbi$year, Wood_pheno_table_scbi$wood_type), FUN = mean)
+SD <- aggregate(Wood_pheno_table_scbi$doy, by = list(Wood_pheno_table_scbi$perc,Wood_pheno_table_scbi$year, Wood_pheno_table_scbi$wood_type), FUN = sd)
+names(SD) <- c("Group.1", "Group.2","Group.3", "sd")
 means <- left_join(means,SD)
-names(means) <- c("perc", "wood_type", "mean", "SD")
+names(means) <- c("perc","year", "wood_type", "mean", "SD")
 
-Wood_pheno_table_scbi <- left_join(Wood_pheno_table_scbi, means, by = c("perc","wood_type"))
+Wood_pheno_table_scbi <- left_join(Wood_pheno_table_scbi, means, by = c("perc","year", "wood_type"))
 
-Wood_pheno_table_scbi$rm <- ifelse(Wood_pheno_table_scbi$doy > Wood_pheno_table_scbi$mean+(sd*Wood_pheno_table_scbi$SD) | Wood_pheno_table_scbi$doy < Wood_pheno_table_scbi$mean-(sd*Wood_pheno_table_scbi$SD), "rm", "keep")
+Wood_pheno_table_scbi$rm <- ifelse(Wood_pheno_table_scbi$doy > Wood_pheno_table_scbi$mean+(3*Wood_pheno_table_scbi$SD) | Wood_pheno_table_scbi$doy < Wood_pheno_table_scbi$mean-(3*Wood_pheno_table_scbi$SD), "rm", "keep")
[…]

### 2022-02-19 RScripts/scbiPhenQA.R
filtered leaf pheno data

@@ -18,16 +18,16 @@
 # for details). Note that the full QA flag descriptions are included in a separate
 # product, MCD43A2 (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD43A2).
 #
-# Finally, the tables (scbiPhen and scbiEVI2) used were generated from GEE directly. 
+# Finally, the tables (scbiPhen and scbiEVI2) used were generated from GEE directly.
 # Please see this link for the code: https://code.earthengine.google.com/0b14ff8145db942d54f4ceb1f2bcddc0
 # and this link for the asset: https://code.earthengine.google.com/?asset=users/imcgreg/forestGEOsites
 ## -----------------------------------------------------------------------------
-source("scbi/scbiPhenQA_functions.R")
+source("RScripts/scbiPhenQA_functions.R")
 
-phenFile <- "scbi/scbiPhen.csv"
-phenFileSave <- "scbi/scbiPhenFiltered.csv"
-evi2File <- "scbi/scbiEVI2.csv"
-evi2FileSave <- "scbi/scbiEVI2Filtered.csv"
+phenFile <- "Data/Leaf phenology/scbiPhen.csv"
+phenFileSave <- "Data/Leaf phenology/scbiPhenFiltered.csv"
+evi2File <- "Data/Leaf phenology/scbiEVI2.csv"
+evi2FileSave <- "Data/Leaf phenology/scbiEVI2Filtered.csv"
 
 # 1. Filter phenology data from MCD12Q2
 phen <- filterPhenology(filePath=phenFile, saveFile=TRUE, savePath=phenFileSave)

### 2022-02-19 RScripts/DOYtiming_all_years.R
Update canopy phenology

@@ -36,7 +36,7 @@ climwindows <-
 
 #Method for identifying correct windows is in Pheno_tsensitivity_figure_tmax.R
 climwinmeans_rp <- weatherdata %>%
-  filter(doy %in% c(71:78)) %>% #April 2 - April 8
+  filter(doy %in% c(70:77)) %>% #April 2 - April 8
   group_by(year) %>%
   summarize(climwinmean = mean(cleantmax)) %>%
   mutate(wood_type = "ring-porous")
@@ -175,11 +175,6 @@ summary(lm(doy75_dp$x~doy75_dp$window_temp))
 #DOY75 -3.563 days/degree C (p = 0.04)
 
 #Add leaf Phenology----
-leaf_phenology <- read_csv("Data/Leaf phenology/leaf_phenology.csv") %>%
-  filter(site == "SCBI")
-aggregates_scbi <- left_join(aggregates_scbi, leaf_phenology, by = "year")
-aggregates_scbi <- aggregates_scbi[complete.cases(aggregates_scbi$los),]
-aggregates_scbi$stanlos <- (aggregates_scbi$los-min(aggregates_scbi$los))/(max(aggregates_scbi$los)-min(aggregates_scbi$los))
 
 g <-  ggplot(aggregates_scbi, aes(x=x, y = as.character(Group.2), group = interaction(Group.1, year), color = stanT, linetype = Group.1))+
   geom_point(size = 3)+
@@ -192,9 +187,10 @@ g
 
 #Leaf phenology
 leaf_phenology <- read_csv("Data/Leaf phenology/leaf_phenology.csv") %>%
-  filter(site == "SCBI")
+  filter(pointid == "SCBI")
 library(reshape2)
-names(leaf_phenology) <- c("site","year","Greenup", "Mid-greenup", "Peak","Senescence","los","tmp")
+names(leaf_phenology) <- c("site","year","Greenup", "Mid-greenup", "Peak","maturity", "Senescence","midgreendown", "tmp","los")
+leaf_phenology <- leaf_phenology[,c(-6,-8)]
 leaf_phenology_melt <- melt(leaf_phenology, id.vars = c("site","year","tmp","los"))
 leaf_phenology_melt$value <- yday(leaf_phenology_melt$value)
 aggregate(leaf_phenology_melt$value, by = list(leaf_phenology_melt$variable), FUN = mean)
@@ -471,13 +467,16 @@ h
 #Leaf phenology plot----
[…]

### 2022-02-19 RScripts/Pheno_tsensitivity_figure_tmax.R

@@ -859,7 +859,7 @@ climwindows$date <- paste0(climwindows$monthopen, "/", climwindows$dayopen, "-",
 
 #TMAX
 climwinmeans_rp_hf <- weatherdata_hf %>%
-  filter(DOY %in% c(70:140)) %>% # March 26 - May 13
+  filter(DOY %in% c(climwindows[4,13]:climwindows[4,14])) %>% # March 26 - May 13
   group_by(year) %>%
   summarize(climwinmean = mean(airtmax)) %>%
   mutate(wood_type = "ring-porous")
@@ -873,7 +873,7 @@ climwinmeans_rp_hf <- weatherdata_hf %>%
 
 #TMAX
 climwinmeans_dp_hf <- weatherdata_hf %>%
-  filter(DOY %in% c(126:161)) %>% # March 19 - May 13
+  filter(DOY %in% c(climwindows[1,13]:climwindows[1,13])) %>% # March 19 - May 13
   group_by(year) %>%
   summarize(climwinmean = mean(airtmax)) %>%
   mutate(wood_type = "diffuse-porous")

### 2022-02-19 RScripts/Pheno_tsensitivity_figure_tmax.R
change read_csv to read.csv

@@ -737,7 +737,7 @@ weatherdata_hf <-
 #
 # #TMAX
 climwindows <-
-  weekly_climwin_results_HF_TMAX <- read_csv("results/Climwin_results/Weekly/Harvard Forest/weekly_climwin_results_HF_TMAX.csv") %>%
+  weekly_climwin_results_HF_TMAX <- read.csv("results/Climwin_results/Weekly/Harvard Forest/weekly_climwin_results_HF_TMAX.csv") %>%
   filter(wood_type != "other") %>%
   mutate(
     median_windowopendate = as.Date(median_windowopendate),
@@ -859,7 +859,7 @@ climwindows$date <- paste0(climwindows$monthopen, "/", climwindows$dayopen, "-",
 
 #TMAX
 climwinmeans_rp_hf <- weatherdata_hf %>%
-  filter(DOY %in% c(70:140)) %>% # March 26 - May 13
+  filter(DOY %in% c(climwindows[4,13]:climwindows[4,14])) %>% # March 26 - May 13
   group_by(year) %>%
   summarize(climwinmean = mean(airtmax)) %>%
   mutate(wood_type = "ring-porous")
@@ -873,7 +873,7 @@ climwinmeans_rp_hf <- weatherdata_hf %>%
 
 #TMAX
 climwinmeans_dp_hf <- weatherdata_hf %>%
-  filter(DOY %in% c(126:161)) %>% # March 19 - May 13
+  filter(DOY %in% c(climwindows[1,13]:climwindows[1,14])) %>% # March 19 - May 13
   group_by(year) %>%
   summarize(climwinmean = mean(airtmax)) %>%
   mutate(wood_type = "diffuse-porous")

### 2022-02-19 RScripts/Pheno_tsensitivity_figure_tmax.R
change read_csv to read.csv

@@ -737,7 +737,7 @@ weatherdata_hf <-
 #
 # #TMAX
 climwindows <-
-  weekly_climwin_results_HF_TMAX <- read_csv("results/Climwin_results/Weekly/Harvard Forest/weekly_climwin_results_HF_TMAX.csv") %>%
+  weekly_climwin_results_HF_TMAX <- read.csv("results/Climwin_results/Weekly/Harvard Forest/weekly_climwin_results_HF_TMAX.csv") %>%
   filter(wood_type != "other") %>%
   mutate(
     median_windowopendate = as.Date(median_windowopendate),

### 2022-02-19 doc/manuscript/Dow_manuscript_Nature.Rmd
clarify CTW selection for leaf pheno

@@
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
-Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }
+Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring-porous window at HF and diffuse-porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }
 

### 2022-02-19 RScripts/additional_figures.R
use filtered leaf phenos

@@ -457,15 +457,15 @@ percent_growth_hf <- Wood_pheno_table_hf %>%
   )
 
 #EVI2 plots
-scbiEVI2 <- read_csv("Data/Leaf phenology/scbiEVI2.csv")
+scbiEVI2 <- read_csv("Data/Leaf phenology/scbiEVI2Filtered.csv")
 scbiEVI2$year <- year(scbiEVI2$date)
 scbiEVI2$doy <- yday(scbiEVI2$date)
 #SCBI plot
 scbiEVI <- scbiEVI2[scbiEVI2$pointid %in% "SCBI",]
-scbiEVI <- scbiEVI[scbiEVI$BRDF_Albedo_Band_Mandatory_Quality_Band1 %in% 0,]
+#scbiEVI <- scbiEVI[scbiEVI$BRDF_Albedo_Band_Mandatory_Quality_Band1 %in% 0,]
 #HF plot
 hfEVI <- scbiEVI2[scbiEVI2$pointid %in% "HARV",]
-hfEVI <- hfEVI[hfEVI$BRDF_Albedo_Band_Mandatory_Quality_Band1 %in% 0,]
+#hfEVI <- hfEVI[hfEVI$BRDF_Albedo_Band_Mandatory_Quality_Band1 %in% 0,]
 
 png(filename = "doc/manuscript/tables_figures/growth_curves_all.png", width=10, height=10,
     pointsize=12, bg="transparent", units="in", res=600,

### 2022-02-20 doc/manuscript/Dow_manuscript_Nature.Rmd
edit FIg. 2 caption
added explanation of DOY variables

@@
 
-![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
-Canopy foliage variables were extracted from a single pixel at the center of each forest plot.
-Only data flagged as "good" or "best" quality were included in the calculation of these phenology variables.
+![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
+Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
 Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.

### 2022-02-21 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
start replacing references to growth "phenology"

@@
 
-Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
+Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
 Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
@@
 and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
-This allows an asymetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
+This allows an asymmetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
 The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
@@
 Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
-We defined CTW for $DOY_{25}$, as opposed to other growth phenology parameters, because spring temperatures should have the most direct influence on this variable.
+We defined CTW for $DOY_{25}$, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable.
 

### 2022-02-21 doc/manuscript/Dow_manuscript_Nature.Rmd
new line for each sentence
This makes it easier to track changes

@@
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
-Cite delpierre 2016 around here. Delpierre et al. 2016 found that stem growth was not correlated with carbon inputs, agreeing with the results here. However, the authors 
+**(Cite delpierre 2016 around here.) Delpierre et al. 2016 found that stem growth was not correlated with carbon inputs, agreeing with the results here. However, the authors ...**
 We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
@@
 
-We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig.
-`r schematic_summary`).
+We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig. `r schematic_summary`). 
 In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
@@
 For each year at each site, we extracted data from the pixel containing the center of each forest plot (resolution of 500m). 
-Using the daily MODIS 2-band Enhanced Vegetation Index data (EVI2) , the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
+Using the daily MODIS 2-band Enhanced Vegetation Index data (EVI2), the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
 Data points were kept if they were flagged as "good" or "best" quality.

### 2022-02-21 doc/manuscript/Dow_Extended_Data.Rmd
remove use of growth "phenology"

@@
 species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
-mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
+mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the seasonality and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
 tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."
@@
 
-![**Extended Data Figure 4 | Response of stem growth phenology and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
+![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
 Shown are relationships between mean $T_{max}$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
@@
 
-![**Extended Data Figure 5 | Response of stem growth phenology and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
+![**Extended Data Figure 5 | Response of stem growth timing and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.

### 2022-02-21 doc/manuscript/Dow_manuscript_Nature.Rmd
remove use of growth "phenology"

@@
 `r schematic_summary`).
-This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in growth phenology[@dorangeville_peak_2021; @tumajer_limitation_2022].
+This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021; @tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 
-Growth milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
+Seaonal milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
 Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
@@
 
-![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
@@
 As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and C cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
-Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.
+Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the seasonality of forest C cycling.
 If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
@@
 Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
-Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree.
-We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
+Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree.
+We ran separate models for each species group at each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
 This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019].

### 2022-02-21 doc/manuscript/Dow_Extended_Data.Rmd
Update Dow_Extended_Data.Rmd
#112

@@
 Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straightforward.
-We used the critical window identified for diffuse-porous trees as the critical window for leaf phenology calculations because they included a wider range of spring dates.
+We used the critical window which had the greatest correlation with leaf greenup timing as the critical window for leaf phenology calculations.
 

### 2022-02-22 doc/manuscript/Dow_manuscript_Nature.Rmd
fix spelling
Update Dow_manuscript_Nature.Rmd
#111

@@
 
-Seaonal milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
+Seasonal milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
 Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
@@
 Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree.
-We ran separate models for each species group at each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
+We ran separate models for each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
 This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019].

### 2022-02-22 doc/manuscript/Dow_manuscript_Nature.Rmd
added citations to Delpierre et al. 2016

@@
 Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
+However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
 These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021].
-This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
+The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
 
@@
 `r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016].
 Our tree-ring analysis (Fig.
@@
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
-**(Cite delpierre 2016 around here.) Delpierre et al. 2016 found that stem growth was not correlated with carbon inputs, agreeing with the results here. However, the authors ...**
 We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
@@
 
+<!-- The following two paragraphs--and two final sentences in the paragraph above-- need work (assigned to KAT).
+It may make most sense to condense into a single paragraph making the point that while we don't know where the C goes, we can rule out a substantial, sustained boost to the long-term sink of woody productivity-->
 One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 

### 2022-02-22 doc/manuscript/Dow_manuscript_Nature.Rmd
added missing citations

@@
 **As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
+**However, less is known about how spring temperatures affect tree stem growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].** <!--delpierre_chilling_2019, huang_photoperiod_2020, rossi_critical_2008, savage_coordination_2021 are for conifers!-->
 **Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
@@
 While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021].
+<!--delpierre_chilling_2019, huang_photoperiod_2020, rossi_critical_2008, savage_coordination_2021 are for conifers!-->
 The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing of growth remains unclear.

### 2022-02-22 doc/manuscript/Dow_manuscript_Nature.Rmd
clarified conifer references
#115

--- 2022-02-22 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 **As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
+**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
 **Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
@@
 
+<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
@@
 
-While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021].
-The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing of growth remains unclear.
+While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021] (but see Refs.
+[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).
+The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].
+<!--this sentence is repetitive with with previous.-->
 

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
read/ cite delpierre_temperate_2016
#116 , #115

@camerondow35 , FYI, delpierre_temperate_2016 is a very relevant and interesting review!

@@
 **As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
+**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021; @delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
 **Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
+Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
 However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
@@
 
-While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021] (but see Refs.
+While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021; @delpierre_temperate_2016] (but see Refs.
 [@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).
@@
 The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
-3).
-We demonstrate that warmer springs either have little effect on, or even hasten, the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
-`r tree_rings`).
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016].
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
+**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
+3).**
+**We demonstrate that warmer springs either have little effect on, or even hasten, the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
+`r tree_rings`).** <!--repeated sentence-->
+Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016; @delpierre_temperate_2016].
+<!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.-->
 Our tree-ring analysis (Fig.
@@
 However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
-Extra C could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
+Extra C could also be saved within trees as non-structural carbohydrates (NSC) and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
 Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig.
@@
 
-A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
+A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC production, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
 Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
adjust language related to start/end of growing season
#113

@@
 Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
+However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this (and check if keenan_timing_2015 looks at stem growth)--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
 These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
@@
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
-**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
+**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the deceleration of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
 3).**
-**We demonstrate that warmer springs either have little effect on, or even hasten, the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
+**We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
 `r tree_rings`).** <!--repeated sentence-->
-Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016; @delpierre_temperate_2016].
+Our observations suggest that the deceleration of stem expansion (starting on $DOY_{g_{max}}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016; @delpierre_temperate_2016].
 <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.-->
@@
 
-Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
+Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
 However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
@@
 
-A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC production, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
+A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC storage, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
 Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
- notes for revamping intro
- citation formatting

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021; @delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
 **Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
@@
 **These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
-**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
 
@@
 
-In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@friedlingstein_global_2021; @harris_global_2021].
+In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@friedlingstein_global_2021;@harris_global_2021].
 A substantive portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ [@pugh_role_2019].
@@
 -->
-The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].
+The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020].
+
+~~~~ (rework starting here)
 
@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this (and check if keenan_timing_2015 looks at stem growth)--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
+Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014].
+However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016],<!--I (KAT) want to work on this (and check if keenan_timing_2015 looks at stem growth)--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
 These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
 
-While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021; @delpierre_temperate_2016] (but see Refs.
-[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).
-The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].
+While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016] (but see Refs.
+[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).
+The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].
 <!--this sentence is repetitive with with previous.-->
@@
 Therefore, years with a similar amount of stem growth would sequester a similar amount of carbon during cell wall thickening, and we can say that stem growth equates to woody productivity.
+
+*points to make in new version*:
+
+- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
+- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
+- close coordination of budburst and xylogenesis[@delpierre_temperate_2016]
+
+~~~~ (rework ending here)
 
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021; @tumajer_limitation_2022].
+This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of phenological milestones.
-Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
+Consistent with findings from previous studies[@parmesan_globally_2003;@ibanez_forecasting_2010;@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
@@
 The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014;@finzi_carbon_2020].
 **However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the deceleration of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
@@
 `r tree_rings`).** <!--repeated sentence-->
-Our observations suggest that the deceleration of stem expansion (starting on $DOY_{g_{max}}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016; @delpierre_temperate_2016].
+Our observations suggest that the deceleration of stem expansion (starting on $DOY_{g_{max}}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
 <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.-->
@@
 
-Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
-However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
-Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
+Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022;@tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016;@fu_maximum_2019].
+However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].
+Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
 We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
@@
 It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
-However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
+However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
 Extra C could also be saved within trees as non-structural carbohydrates (NSC) and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
@@
 A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC storage, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
-Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
-However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
+Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
+However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
 Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
@@
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in CO~2~ sequestration and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in CO~2~ sequestration and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures, which constitutes a positive feedback to climate change.
@@
 
-Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
+Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015;@davies_forestgeo_2021].
 SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km south of Front Royal, Virginia. 
@@
 These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
-An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015; @herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest.
+An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015;@herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest.
 Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig.
@@
 Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
-Daily temperature records for Harvard Forest, which had already been gap-filled based on other local records, were obtained from the Harvard Forest weather station[@boose_shaler_2021; @boose_fisher_2021].
+Daily temperature records for Harvard Forest, which had already been gap-filled based on other local records, were obtained from the Harvard Forest weather station[@boose_shaler_2021;@boose_fisher_2021].
 For each site, we used records of daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
@@
 We ran separate models for each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
-This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019].
+This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@stan_development_team_stan_2019].
 In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
@@
 To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
-The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
+The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020;@begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
 Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures were associated with greater summer drought stress in our data set.
@@
 
-Dominant tree species were cored at both SCBI[@helcoski_growing_2019;@bourg_initial_2013] and Harvard Forest[@alexander_potential_2019; @dye_comparing_2016; @finzi_carbon_2020] following sampling designs that covered a broad range of DBH.
+Dominant tree species were cored at both SCBI[@helcoski_growing_2019;@bourg_initial_2013] and Harvard Forest[@alexander_potential_2019;@dye_comparing_2016;@finzi_carbon_2020] following sampling designs that covered a broad range of DBH.
 We analyzed records for the ring- and diffuse- porous species at each site (`r  EDT_species`), but excluded semi-ring porous species (e.g., *Juglans nigra* L.
@@
 For each species-site combination, we converted tree-ring records into the dimensionless RWI to emphasize interannual variability associated with climate [@cook_methods_1990].
-A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985; @cook_methods_1990].
+A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
 An adaptive power transformation, a process that also stabilises the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 

### 2022-02-23 RScripts/Pheno_tsensitivity_figure_april_april_tmax.R
updated bayesian results
2.5 SD

@@ -1004,8 +1004,8 @@ fig6_RP_sl_hf <- ggplot() +
 
 fig6_DP_sl_hf <- ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
-  stat_lineribbon(data = predictions_sl_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = .95,alpha = .10, linetype = "dashed") +
-  stat_lineribbon(data = predictions_sl_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = 0,alpha = .75, linetype = "dashed") +
+  stat_lineribbon(data = predictions_sl_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = .95,alpha = .10, linetype = "solid") +
+  stat_lineribbon(data = predictions_sl_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = 0,alpha = .75, linetype = "solid") +
   geom_point(data = Wood_pheno_table_DP_sl_hf, aes(x = climwinmean, y = seasonlength, col = "blue")) +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   #scale_fill_brewer() +
@@ -1072,8 +1072,8 @@ fig6_RP_mr_hf <- ggplot() +
 
 fig6_DP_mr_hf <-   ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
-  stat_lineribbon(data = predictions_mr_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = .95,alpha = .10, linetype = "dashed") +
-  stat_lineribbon(data = predictions_mr_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = 0,alpha = .75, linetype = "dashed") +
+  stat_lineribbon(data = predictions_mr_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = .95,alpha = .10, linetype = "solid") +
+  stat_lineribbon(data = predictions_mr_DP_hf, aes(x = climwinmean, y = predictions_rstanarm, col = "blue"), .width = 0,alpha = .75, linetype = "solid") +
[…]

### 2022-02-23 RScripts/Pheno_tsensitivity_figure_tmax.R
Update Pheno_tsensitivity_figure_tmax.R

@@ -580,8 +580,8 @@ Wood_pheno_table_DP_mr <- filter(woodtable, wood_type == "diffuse-porous")
 
 fig6_RP_mr <-  ggplot() +
   # geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
-  stat_lineribbon(data = predictions_mr_RP, aes(x = climwinmean, y = predictions_rstanarm, col = "purple"), .width = .95, linetype = "dashed", alpha = .10) +
-  stat_lineribbon(data = predictions_mr_RP, aes(x = climwinmean, y = predictions_rstanarm, col = "purple"), .width = 0, linetype = "dashed", alpha = .95) +
+  stat_lineribbon(data = predictions_mr_RP, aes(x = climwinmean, y = predictions_rstanarm, col = "purple"), .width = .95, linetype = "solid", alpha = .10) +
+  stat_lineribbon(data = predictions_mr_RP, aes(x = climwinmean, y = predictions_rstanarm, col = "purple"), .width = 0, linetype = "solid", alpha = .95) +
   geom_point(data = Wood_pheno_table_RP_mr, aes(x = climwinmean, y = max_rate, col = "purple")) +
   # geom_abline(data = posterior_lines, aes(intercept = `(Intercept)`, slope = marchmean, col = perc), size = 1) +
   #scale_fill_brewer() +

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd

@@
 - stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
-- close coordination of budburst and xylogenesis[@delpierre_temperate_2016]
+- close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016]
+- start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period
 

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
read/cite lu_no_2022
#116, #118

@@
 <!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
 Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014].
@@
 `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].
-Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].
+Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period on annual productivity[@buermann_widespread_2018;@zhang_drought_2021].
 However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.

### 2022-02-23 RScripts/data_cleaning.R
remove "stashed changes" text

@@ -147,11 +147,7 @@ names(means) <- c("perc","year", "wood_type", "mean", "SD")
 
 Wood_pheno_table_scbi <- left_join(Wood_pheno_table_scbi, means, by = c("perc","year", "wood_type"))
 
-<<<<<<< Updated upstream
-Wood_pheno_table_scbi$rm <- ifelse(Wood_pheno_table_scbi$doy > Wood_pheno_table_scbi$mean+(2*Wood_pheno_table_scbi$SD) | Wood_pheno_table_scbi$doy < Wood_pheno_table_scbi$mean-(2*Wood_pheno_table_scbi$SD), "rm", "keep")
-=======
 Wood_pheno_table_scbi$rm <- ifelse(Wood_pheno_table_scbi$doy > Wood_pheno_table_scbi$mean+(sd*Wood_pheno_table_scbi$SD) | Wood_pheno_table_scbi$doy < Wood_pheno_table_scbi$mean-(sd*Wood_pheno_table_scbi$SD), "rm", "keep")
->>>>>>> Stashed changes
 bad_doys <- as.character(unique(Wood_pheno_table_scbi[Wood_pheno_table_scbi$rm %in% "rm",]$tag_year)) #27
 bad_doy_tags <- Wood_pheno_table_scbi[Wood_pheno_table_scbi$tag_year %in% bad_doys,]
 bad_doy_percs <- percent_growth[percent_growth$tag_year %in% bad_doys,]
@@ -693,11 +689,7 @@ names(means) <- c("perc","year", "wood_type", "mean", "SD")
 
 Wood_pheno_table_hf <- left_join(Wood_pheno_table_hf, means, by = c("perc","year", "wood_type"))
 
-<<<<<<< Updated upstream
-Wood_pheno_table_hf$rm <- ifelse(Wood_pheno_table_hf$doy > Wood_pheno_table_hf$mean+(2*Wood_pheno_table_hf$SD) | Wood_pheno_table_hf$doy < Wood_pheno_table_hf$mean-(2*Wood_pheno_table_hf$SD), "rm", "keep")
-=======
 Wood_pheno_table_hf$rm <- ifelse(Wood_pheno_table_hf$doy > Wood_pheno_table_hf$mean+(sd*Wood_pheno_table_hf$SD) | Wood_pheno_table_hf$doy < Wood_pheno_table_hf$mean-(sd*Wood_pheno_table_hf$SD), "rm", "keep")
[…]

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
updating parameters

--- 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
+This analysis was performed separately for ring- (67 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (52 trees at SCBI and 183 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 
-Seasonal milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
-Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig.
+Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
+Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
-Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
+Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only six days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
 
 ![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
-Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003).
+Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
 Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
@@
 Consistent with findings from previous studies[@parmesan_globally_2003;@ibanez_forecasting_2010;@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
-`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
-Similarly, at both sites, the stem growth milestones $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
+`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
+Similarly, at both sites, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 <!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
-Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014].
-However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016],<!--I (KAT) want to work on this (and check if keenan_timing_2015 looks at stem growth)--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
+Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], *but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016]*,<!--I (KAT) need to work on the part in italics, if retained keenan_timing and zani_increased_2020 do not look at stem growth and are largely contradicted by lu_no_2022.--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
 These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
@@
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in CO~2~ sequestration and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures, which constitutes a positive feedback to climate change.

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 - start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period
+- probably include some mention of conifers (for whcih more work has been done)
 

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
updated results section

@@
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
-Similarly, at both sites, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs.
+Similarly, with the exception of ring-porous $DOY_{75}$ at SCBI and ring-porous $DOY_{g_{max}}$ at HF, the stem growth milestones $DOY_{25}$, $DOY_{50}$, $DOY_{75}$,  and $DOY_{g_{max}}$ decreased with mean $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
-Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 1.1-1.9 days/ $^\circ$C for ring-porous species and 3.5-3.6 days/ $^\circ$C for diffuse-porous species at SCBI, and 2.8-7.2 days/ $^\circ$C for ring-porous species and 6.6-7.9 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).
+Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring-porous species and 2.8-2.9 days/ $^\circ$C for diffuse-porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring-porous species and 0.9-4.2 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).
 
@@
 In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_phenology_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest).
-$\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`).
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_phenology_parameters`), displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to mean $T_{max}$ during the critical temperature window for ring- and diffuse- porous species.
+$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with mean $T_{max}$ during the critical temperature window at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
 

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
a couple more fixes to wording around phenology

--- 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 
-For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of phenological milestones.
+For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of seasonal milestones.
 Consistent with findings from previous studies[@parmesan_globally_2003;@ibanez_forecasting_2010;@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
@@
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, an earlier onset of growth in the spring is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
-Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures, which constitutes a positive feedback to climate change.
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, advancement in the timing of stem growth (Fig.
+1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.
 

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@teixeirak see comment added here. We should find a way to work the 1999 HF drought into the discussion. I think we could find a way to discuss how the drought may have led to different results from the total growth analysis. Something like, "Our analysis years at SCBI had no severe drought, so we see positive (biologically insignificant) increases in total growth, while at HF we see that the drought in 1999 is influencing the total growth analysis so much, that the model reports negative relationships between spring temp and total growth. This contrast emphasi

@@
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014;@finzi_carbon_2020].
-**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the deceleration of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig.
+**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise failed to translate extended growing seasons into biologically significant increases in stem growth, and thereby have negligible effect on total annual growth for most species and locations (Fig.
 3).**
 **We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
-`r tree_rings`).** <!--repeated sentence-->
+`r tree_rings`).** <!--repeated sentence/ CD made potential changes to the first one.
+See what you think.
+We should mention how 1999 was a drought year and also the warmest year in the CTW at HF for both wood types.
+This is potentially contributing to the negative correlations between CTW TMAX and total growth at HF.
+Maybe if there wasn't a severe drought in the warmest year we would see positive correlations like we do at SCBI.
+This is also probably worth mentioning when we discuss the influence of climate during peak growing season being greater than the spring climate.
+-->
 Our observations suggest that the deceleration of stem expansion (starting on $DOY_{g_{max}}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
-<!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.-->
+<!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF.
+Maxrate_DOY for RP at HF was actually significantly delayed.
+This could be a consequence of the severe drought in 1999 which we've included in our analysis.
+My brain is completely fried to the point that I barely remember what the literature claims this max rate is related to (water stress, right?).
+So Maybe you're in a better position to add some discussion about this here.
+-->
 Our tree-ring analysis (Fig.

### 2022-02-24 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 **These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
-**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the long-term CO~2~ sink of temperate deciduous forests.**
 

### 2022-02-24 RScripts/data_cleaning.R
Update data_cleaning.R

@@ -17,7 +17,7 @@ library(scales)
 library(tidybayes)
 
 # Number of standard deviations used to identify "outliers"
-sd <- 2
+sd <- 2.5
 
 # Load Sean McMahon's Dendroband functions
 source("RScripts/dendroband_functions.R")
@@ -134,10 +134,11 @@ late_starts <- all_stems[all_stems$tag_year %in% start_days$Group.1,]
 #remove from percent growth and wood_pheno
 percent_growth <- percent_growth[!(percent_growth$tag_year %in% late_starts$tag_year),]
 late_survey <- Wood_pheno_table_scbi[(Wood_pheno_table_scbi$tag_year %in% late_starts$tag_year),]
+late_tags <- unique(late_survey$tag_year)
 
 Wood_pheno_table_scbi <- Wood_pheno_table_scbi[!(Wood_pheno_table_scbi$tag_year %in% late_starts$tag_year),] # 3054 -> 2994
 
-# # #Remove tag years where doy% variable is > 3 SD away from mean----
+# # #Remove tag years where doy% variable is > 2.5 SD away from mean----
 
 means <- aggregate(Wood_pheno_table_scbi$doy, by = list(Wood_pheno_table_scbi$perc,Wood_pheno_table_scbi$year, Wood_pheno_table_scbi$wood_type), FUN = mean)
 SD <- aggregate(Wood_pheno_table_scbi$doy, by = list(Wood_pheno_table_scbi$perc,Wood_pheno_table_scbi$year, Wood_pheno_table_scbi$wood_type), FUN = sd)
@@ -392,9 +393,9 @@ percent_growth <- rbind(percent_growth_DP, percent_growth_RP)
 
 ## Clean both at once now ----
 # Remove small/negligable growth models...Seem to be more of a problem with Harvard Forest but remove here for consistency
-percent_growth_gone <- subset(percent_growth, dbh_total_growth <= 0.05) # remove small growth trees
-SCBI_smallgrowth <- unique(percent_growth_gone$tag_year) # rm 8
-percent_growth <- subset(percent_growth, dbh_total_growth >= 0.05) # remove small growth trees
+# percent_growth_gone <- subset(percent_growth, dbh_total_growth <= 0.05) # remove small growth trees
+# SCBI_smallgrowth <- unique(percent_growth_gone$tag_year) # rm 8
+# percent_growth <- subset(percent_growth, dbh_total_growth >= 0.05) # remove small growth trees
[…]

### 2022-02-24 doc/manuscript/Dow_manuscript_Nature.Rmd
updated method section

@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2021].
+The summer of 1999 was abnormally dry.
+Records indicate that in 1999 only 52mm of precipitation fell per month June-August compared to average monthly precipitation values of 100mm+ [@dorangeville_peak_2021].
 
@@
 Modeled fits for tree-years were removed under five conditions:
-(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each xylem architecture (SCBI = 26, Harvard Forest = 24); 
-(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 3$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 71, Harvard Forest = 226);
-(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 8, Harvard Forest = 137); 
-(4) tree years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 20, HF = 6).
+(1) single day growth rates were $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
+(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101);
+(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253); 
+(4) tree years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
 At both sites the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
-This process was repeated using 2.5 and 3 standard deviations and the results of analyses were not qualitatively different.
+This process was repeated using 2 and 3 standard deviations and the results of analyses were not qualitatively different.
 

### 2022-02-24 doc/manuscript/tables_figures/schematic_summary.Rmd
some cleanup/organization
delete redundant figure
add asterisk /footnote to summary table
@camerondow35 , to your question on commit db49c84 (https://github.com/EcoClimLab/growth_phenology/commit/db49c843b14e8c195d5184124058358ef1bd91c9)

I added an example asterisk and footnote. Please modify both the table and footnote as appropriate.

--- 2021-01-26 doc/manuscript/tables_figures/tables.Rmd
@@
 
-![](schematic_v2.png)
+![](schematic_summary.png)
 
@@
 ```
+*not biologically significant

### 2022-02-24 doc/manuscript/tables_figures/schematic_summary.Rmd
remove DOY_g_max from summary table
#119

@@
   column_spec(3:6, width = "1cm") %>%
-  pack_rows("Timing of growth", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Daily growth rate", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Annual growth", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Timing of growth", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Daily growth rate", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Annual growth", 6, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T)

### 2022-02-24 doc/manuscript/Dow_manuscript_Nature.Rmd
remove DOY_g_max from main text
#119

@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$; occurs on average within 2 days of $DOY_{50}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
-Similarly, with the exception of ring-porous $DOY_{75}$ at SCBI and ring-porous $DOY_{g_{max}}$ at HF, the stem growth milestones $DOY_{25}$, $DOY_{50}$, $DOY_{75}$,  and $DOY_{g_{max}}$ decreased with mean $T_{max}$ during the critical temperature window (Figs.
+Similarly, with the exception of ring-porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
@@
 -->
-Our observations suggest that the deceleration of stem expansion (starting on $DOY_{g_{max}}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
+Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
 <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF.
@@
 This allows an asymmetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth. 
-The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
+When $\theta = 1$, $g_{max}$ occurs on $DOY_{ip}$.
 The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. 
@@
 <!--Hypothesis testing -->
-Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
+Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
 `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.

### 2022-02-24 doc/manuscript/Dow_Extended_Data.Rmd
remove DOY_g_max from SI
#119

@@
     column_spec(1, width = "4cm") %>%
-    pack_rows("Stem Growth", 2, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-    pack_rows("Canopy Foliage (ecosystem level)", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-    pack_rows("Phenological temperature sensitivity* (days/°C)", 13, 22, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
-    add_indent(14:17)%>%
-    add_indent(19:22)%>%
+    pack_rows("Stem Growth", 2, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    pack_rows("Canopy Foliage (ecosystem level)", 8, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+    pack_rows("Phenological temperature sensitivity* (days/°C)", 12, 21, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
+    add_indent(13:16)%>%
+    add_indent(18:21)%>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2022-02-24 doc/manuscript/Dow_Extended_Data.Rmd
oops, missed a DOY_g_max...
#119

@@
     pack_rows("Canopy Foliage (ecosystem level)", 8, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-    pack_rows("Phenological temperature sensitivity* (days/°C)", 12, 21, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
-    add_indent(13:16)%>%
-    add_indent(18:21)%>%
+    pack_rows("Phenological temperature sensitivity* (days/°C)", 12, 20, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
+    add_indent(13:15)%>%
+    add_indent(17:20)%>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2022-02-24 RScripts/additional_figures.R
remove DOY_g_max from schematic #119

@@ -141,17 +141,17 @@ HF_doy_diameter_quartile <- bind_rows(
   mutate(year = as.factor(year))
 
 # Semi-synthetic DOY_g_max dates
-HF_doy_max_rate_offset <- 2
-HF_doy_max_rate <- HF_doy_diameter_quartile %>%
-  filter(perc == 0.5) %>%
-  mutate(doy = doy + HF_doy_max_rate_offset) %>%
-  mutate(year = as.factor(year))
-
-HF_doy_max_rate <- bind_rows(
-  HF_true_values_cold %>% mutate(diff = abs(perc-0.53)) %>% arrange(diff) %>% slice(1) %>% mutate(year = "Cold"),
-  HF_true_values_hot %>% mutate(diff = abs(perc-0.53)) %>% arrange(diff) %>% slice(1) %>% mutate(year = "Hot")
-) %>%
-  mutate(year = as.factor(year))
+# HF_doy_max_rate_offset <- 2
+# HF_doy_max_rate <- HF_doy_diameter_quartile %>%
+#   filter(perc == 0.5) %>%
+#   mutate(doy = doy + HF_doy_max_rate_offset) %>%
+#   mutate(year = as.factor(year))
+#
+# HF_doy_max_rate <- bind_rows(
+#   HF_true_values_cold %>% mutate(diff = abs(perc-0.53)) %>% arrange(diff) %>% slice(1) %>% mutate(year = "Cold"),
+#   HF_true_values_hot %>% mutate(diff = abs(perc-0.53)) %>% arrange(diff) %>% slice(1) %>% mutate(year = "Hot")
+# ) %>%
+#   mutate(year = as.factor(year))
 
 
 
@@ -173,16 +173,16 @@ HF_significant_perc <-
     significant = c(TRUE, TRUE, TRUE)
   )
 
-# Add g_max
-HF_significant_perc <- bind_rows(
-  HF_significant_perc,
-  tibble(
-    doy_hot = HF_doy_max_rate$doy[HF_doy_max_rate$year == "Hot"],
-    perc = 0.53,
-    doy_cold = HF_doy_max_rate$doy[HF_doy_max_rate$year == "Cold"],
-    significant = TRUE
-  )
-)
+# # Add g_max
+# HF_significant_perc <- bind_rows(
+#   HF_significant_perc,
+#   tibble(
[…]

### 2022-02-24 RScripts/additional_figures.R
Shortened CTW

@@ -17,7 +17,7 @@ end_doy <- 365
 
 # Critical temperature window:
 window_open <- 91
-window_close <- 135
+window_close <- 120
 
 # Function for logistic growth model written by Sean. Supposed to be in RDendrom
 # package https://github.com/seanmcm/RDendrom, but function does not seem to be included
@@ -219,10 +219,10 @@ schematic_v2_HF <-
     axis.text.x.top = element_text(color = hot_color),
     axis.ticks.x.top = element_line(color = hot_color),
     axis.ticks.x.bottom = element_line(
-      color = c(rep(cold_color, 4), rep("black", 2), "white")
+      color = c(rep(cold_color, 3), rep("black", 2), "white")
     ),
     axis.text.x.bottom = element_text(
-      color = c(rep(cold_color, 4), rep("black", 3))
+      color = c(rep(cold_color, 3), rep("black", 3))
     )
   ) +
   scale_x_continuous(

### 2022-02-24 doc/manuscript/Dow_Extended_Data.Rmd
add row in edt2 for growth
@teixeirak  Any opposition to adding this row to EDT2? Now that tot growth is significant, we should add the values to one of our tables. This seems like the best one.

@@
     pack_rows("Canopy Foliage (ecosystem level)", 8, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-    pack_rows("Phenological temperature sensitivity* (days/°C)", 12, 20, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
-    add_indent(13:15)%>%
-    add_indent(17:20)%>%
+    pack_rows("Temperature sensitivity* (days/°C)", 12, 20, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
+    add_indent(13:16)%>%
+    add_indent(18:21)%>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
 ```
-*Phenological temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
+*Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
 1).

### 2022-02-25 doc/manuscript/tables_figures/schematic_summary.Rmd
updated schematic table

@@
 ```
-*not biologically significant
+*While this result was statistically significant, effect size was small and considered not biologically significant

### 2022-02-25 doc/manuscript/Dow_manuscript_Nature.Rmd
added explanation for different slopes of total growth

@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$; occurs on average within 2 days of $DOY_{50}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$; occurs on average within 5 days of $DOY_{50}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
-This analysis was performed separately for ring- (67 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (52 trees at SCBI and 183 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
+This analysis was performed separately for ring- (67 trees at SCBI and 138 trees at Harvard Forest) and diffuse porous (52 trees at SCBI and 183 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
-Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
+Consistent with the results of Ref[@dorangeville_peak_2021], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
-Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only six days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).
+Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only six days longer at SCBI for ring porous species and less than one day longer for diffuse porous species (`r  EDT_phenology_parameters`).
 
 ![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
-Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
+Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
 Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
 Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring-porous window at HF and diffuse-porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }
+Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring porous window at HF and diffuse porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }
 
@@
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
-Similarly, with the exception of ring-porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the critical temperature window (Figs.
+Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the critical temperature window (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
-Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring-porous species and 2.8-2.9 days/ $^\circ$C for diffuse-porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring-porous species and 0.9-4.2 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).
+Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).
 
@@
 In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_phenology_parameters`), displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to mean $T_{max}$ during the critical temperature window for ring- and diffuse- porous species.
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to mean $T_{max}$ during the critical temperature window for ring- and diffuse- porous species.
 $\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with mean $T_{max}$ during the critical temperature window at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
-Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring-porous and 4 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
-In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
+Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
+In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
 $T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
@@
 3).**
+We find statistically, but not biologically, significant relationships between total growth and spring temperature at both sites in our dendrometer band analysis. 
+However, the slope of the relationships are in opposite directions, with warming causing decreasing stem growth at Harvard Forest but increasing stem growth at SCBI. 
+We believe this difference is due to severe drought conditions in 1999 which obviously caused abnormal growth patterns in that year (pheno_tsens). 
+Therefore, we suggest that our two sites represent two scenarios: 1.
+A forest which had no significant summer drought, but was only influenced by warming temperatures (SCBI).
+2.
+A forest which experienced spring warming along with severe summer drought (Harvard Forest).
+
+At SCBI, we see a very small increase in total stem growth (on average +0.05mm/$^\circ$C), suggesting that even in the absence of drought or incliment climate conditions during the growing season, stem growth is not enhanced in a meaningful way when spring temperatures are warmer, and peak stem growth is initiated earlier.
+At Harvard Forest, we see a more meaningful decrease in total growth in relation to warmer temperatures in the CTW (-0.15mm/$^\circ$C on average).
+We speculate that this result is likely a product of reductions in total stem growth related to drought damage in 1999.
+Due to the method of selection for CTW, years with significant departures from the norm can influence the analysis to a greater extent.
+Stem growth in 1999 was clearly impacted by climate conditions, and so the CTW selected by climwin maximizes the effect of these impacts (most likely the effects of drought) on stem growth.
+To examine this further, we reran our dendrometer band analysis using the month of May as the CTW.
+When we change the CTW to the month of May, we see results much closer to those found at SCBI (EDT...), with a similar advancement in growth timing but a much less pronounced effect of Tmax on total growth (-0.015mm/$^\circ$C on average, a tenfold decrease compared to the CTW optimized around the drought year).
+Interestingly, we once again see that there was no meaningful increase in total growth.
+The magnitude of the effect of the summer drought at Harvard Forest when compared to the relatively drought free period of SCBI illustrates the point that conditions during the peak growing season are much more influential to stem growth than conditions during the Spring.
 **We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
@@
 Metal dendrometer bands were installed on 941 trees within the SCBI and Harvard Forest ForestGEO plots.
-Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest (`r EDT_species`).
+Bands were placed on dominant species, including two diffuse- and two ring porous species at SCBI and eight diffuse- and three ring porous species at Harvard Forest (`r EDT_species`).
 While we do not estimate the ages of trees in our sample, bands at both sites were placed on individuals of differing sizes in an attempt to measure trees across a range of ages.
@@
 `r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the CTW to be the month of April.
-This was consistent with the periods identified by *climwin* for ring- and diffuse-porous species groups at both sites, all of which included all or part of April (`r  EDT_phenology_parameters`).
+This was consistent with the periods identified by *climwin* for ring- and diffuse porous species groups at both sites, all of which included all or part of April (`r  EDT_phenology_parameters`).
 

### 2022-02-26 RScripts/additional_figures.R
Update additional_figures.R
suggested EDF2 changes

@@ -428,6 +428,100 @@ percent_growth_scbi <-Wood_pheno_table_scbi %>%
     dbh_growth_percent_cummulative = cumsum(dbh_growth_percent)
   )
 
+#percent_try <- subset(percent_growth_scbi,percent_growth_scbi$dbh >= percent_growth_scbi$a & percent_growth_scbi$dbh <= percent_growth_scbi$b)
+# percent_try <- percent_growth_scbi[!(percent_growth_scbi$dbh > percent_growth_scbi$b),]
+# percent_try <- percent_try[!(percent_try$dbh < percent_try$a),]
+
+#PLOT TYPE 2, dotted lines outside of a and b
+percent_growth_scbi$dotted <- ifelse(percent_growth_scbi$dbh < percent_growth_scbi$a | percent_growth_scbi$dbh > percent_growth_scbi$b, "yes", "no")
+
+good_scbi <- ggplot() +
+  coord_cartesian(ylim = c(0, 0.05)) +
+  scale_y_continuous(labels = percent) +
+  labs( x="",y = "Percent of total growth",
+        title = "") +
+  theme_bw() +
+  theme(legend.position = c(.75,.75))+
+  geom_line(data = percent_growth_scbi[percent_growth_scbi$dbh < percent_growth_scbi$a,], aes(x = doy, y = dbh_growth_percent,col = wood_type, group = tag_year), linetype = "dotted", alpha = 0.2) +
+  geom_line(data = percent_growth_scbi[percent_growth_scbi$dbh > percent_growth_scbi$b,], aes(x = doy, y = dbh_growth_percent,col = wood_type, group = tag_year), linetype = "dotted",alpha = 0.2) +
+  geom_line(data = percent_growth_scbi[percent_growth_scbi$dotted %in% "no",], aes(x = doy, y = dbh_growth_percent,col = wood_type, group = tag_year), linetype = "solid",alpha = 0.2) +
+  scale_colour_viridis_d("", end = 2/3)
+
+#Try to get dashed lines by year
+year_ags_scbi <- aggregate(percent_growth_scbi$dbh_growth_percent, by = list(percent_growth_scbi$year,percent_growth_scbi$wood_type, percent_growth_scbi$doy), FUN = mean)
+
+dotted_markers <- data.frame(NULL)
+for (j in c("ring-porous", "diffuse-porous")) {
+
+  for (i in c(2011:2020)) {
[…]

### 2022-02-26 doc/manuscript/Dow_manuscript_Nature.Rmd
add correct table/figure names

@@
 3).**
-We find statistically, but not biologically, significant relationships between total growth and spring temperature at both sites in our dendrometer band analysis. 
-However, the slope of the relationships are in opposite directions, with warming causing decreasing stem growth at Harvard Forest but increasing stem growth at SCBI. 
-We believe this difference is due to severe drought conditions in 1999 which obviously caused abnormal growth patterns in that year (pheno_tsens). 
+We find statistically, but not biologically, significant relationships between total growth and spring temperature at both sites in our dendrometer band analysis (`schematic_summary`). 
+However, the slope of the relationships are in opposite directions, with warming causing decreasing stem growth at Harvard Forest but increasing stem growth at SCBI (`EDT_phenology_parameters`). 
+We believe this difference is due to severe drought conditions in 1999 which obviously caused abnormal growth patterns in that year (`r EDF_pheno_Tmax_climwin`). 
 Therefore, we suggest that our two sites represent two scenarios: 1.
@@
 
-At SCBI, we see a very small increase in total stem growth (on average +0.05mm/$^\circ$C), suggesting that even in the absence of drought or incliment climate conditions during the growing season, stem growth is not enhanced in a meaningful way when spring temperatures are warmer, and peak stem growth is initiated earlier.
-At Harvard Forest, we see a more meaningful decrease in total growth in relation to warmer temperatures in the CTW (-0.15mm/$^\circ$C on average).
-We speculate that this result is likely a product of reductions in total stem growth related to drought damage in 1999.
+At SCBI, we see a very small increase in total stem growth (on average +0.05mm/$^\circ$C), suggesting that even in the absence of drought or inclement climate conditions during the growing season, spring warming does not enhance stem growth in a meaningful way.
+At Harvard Forest, we see a decrease in total growth with warmer temperatures in the CTW (-0.15mm/$^\circ$C on average).
+We speculate that this result is likely a product of reductions in total stem growth related to drought effects in 1999.
 Due to the method of selection for CTW, years with significant departures from the norm can influence the analysis to a greater extent. 
-Stem growth in 1999 was clearly impacted by climate conditions, and so the CTW selected by climwin maximizes the effect of these impacts (most likely the effects of drought) on stem growth.
+Stem growth in 1999 was clearly impacted by climate conditions (`r EDF_pheno_Tmax_climwin`), and so the CTW selected by climwin maximizes the effect of these impacts (most likely the effects of drought) on stem growth.
 To examine this further, we reran our dendrometer band analysis using the month of May as the CTW. 
-When we change the CTW to the month of May, we see results much closer to those found at SCBI (EDT...), with a similar advancement in growth timing but a much less pronounced effect of Tmax on total growth (-0.015mm/$^\circ$C on average, a tenfold decrease compared to the CTW optimized around the drought year).
-Interestingly, we once again see that there was no meaningful increase in total growth.
-The magnitude of the effect of the summer drought at Harvard Forest when compared to the relatively drought free period of SCBI illustrates the point that conditions during the peak growing season are much more influential to stem growth than conditions during the Spring.
+When we change the CTW to the month of May, we see results much closer to those found at SCBI, with a similar advancement in growth timing but a much less pronounced effect of $T_{max}$ on total growth (-0.015mm/$^\circ$C on average, a tenfold decrease compared to the CTW optimized around the drought year;`EDF_pheno_Tmax_AprilMay`).
+Once again see that there was no meaningful influence of spring temperature on total growth. 
+In fact, the magnitude of the effect of the summer drought compared to the effects of spring $T_{max}$ illustrates the point that conditions during the peak growing season are much more influential to stem growth than conditions during the Spring.
 **We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
@@
 See what you think.
-We should mention how 1999 was a drought year and also the warmest year in the CTW at HF for both wood types.
-This is potentially contributing to the negative correlations between CTW TMAX and total growth at HF.
-Maybe if there wasn't a severe drought in the warmest year we would see positive correlations like we do at SCBI.
-This is also probably worth mentioning when we discuss the influence of climate during peak growing season being greater than the spring climate.
 -->

### 2022-02-27 doc/manuscript/Dow_Extended_Data.Rmd
Update Dow_Extended_Data.Rmd
#112

@@
 ```
-*Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
+$\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
 1).
@@
 We used the critical window which had the greatest correlation with leaf greenup timing as the critical window for leaf phenology calculations.
+This resulted in the selection of the diffuse-porous critical window at SCBI and the ring-porous critical window at HF, corresponding to the grouping of the species with greatest woody productivity at each site (diffuse-porous *Liriodendron tulipifera* at SCBI, ring-porious *Quercus rubra* at HF).
 

### 2022-02-28 doc/manuscript/Dow_manuscript_Nature.Rmd
minor changes
@camerondow35 , see edits/ notes in data cleaning section. This requires a bit of attention from you.

@@
 
-![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
@@
 Modeled fits for tree-years were removed under five conditions:
-(1) single day growth rates were $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
+<!--rearrange these in order of how data were removed (3 & 4 come first)-->
+(1) ~~single day growth rates were~~ $g_{max}$ was <!--confirm this change--> $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
 (2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101);
 (3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253); 
-(4) tree years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
+(4) tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
 At both sites the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
@@
 For each year at each site, we extracted data from the pixel containing the center of each forest plot (resolution of 500m). 
-Using the daily MODIS 2-band Enhanced Vegetation Index data (EVI2), the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, and senescence as used in this study.
-Data points were kept if they were flagged as "good" or "best" quality.
+Using the daily MODIS 2-band Enhanced Vegetation Index data (EVI2), the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, peak, and senescence as used in this study.
+Data points were included in the analysis if they were flagged as "good" or "best" quality.
 

### 2022-03-01 doc/manuscript/Dow_manuscript_Nature.Rmd
rearrange data cleaning section

@@
 If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
+We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253), and tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
+These were removed because they reduce the reliability of predicted growth in the modeled curves.
 
@@
 (1) ~~single day growth rates were~~ $g_{max}$ was <!--confirm this change--> $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
-(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101);
-(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253); 
-(4) tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
+(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101).
 At both sites the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.

### 2022-03-01 doc/manuscript/Dow_Extended_Data.Rmd
Update Dow_Extended_Data.Rmd
#120 Might need to revisit the wording of this caption

@@
 ![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
-For stem growth, each line represents the growth of one tree over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.](tables_figures/growth_curves_all.png){ width=100% }
+For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.
+Dashed lines represent modeled DBH change which fell outside of the median DOY where predicted starting DBH (a) and predicted ending DBH (b) were reached.
+Solid lines represent DBH change attributable to stem growth.](tables_figures/growth_curves_all.png){ width=100% }
 

### 2022-03-04 doc/manuscript/Dow_manuscript_Nature.Rmd
tweaked data cleaning section
@camerondow35 , please confirm my change in line 368 to finalize this paragraph.

@@
 After fitting the growth model, we removed tree-years with poor fits.
-Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the logical range. 
-Modeled fits for tree-years were removed under five conditions:
-<!--rearrange these in order of how data were removed (3 & 4 come first)-->
+Models were judged to be poorly fit if modeled growth parameters were outliers, which were commonly indicative of unrealistic fits (e.g., growth occurring outside the growing season or over a very short period) and underlain by very slow tree growth or poor data records.
+Modeled fits for tree-years were removed under two conditions:
 (1) ~~single day growth rates were~~ $g_{max}$ was <!--confirm this change--> $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
 (2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101).
-At both sites the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
-This process was repeated using 2 and 3 standard deviations and the results of analyses were not qualitatively different.
+At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
+This process was repeated using 2 and 3 standard deviations as the cutoff for defining outliers, yielding qualitatively similar results.
 

### 2022-03-04 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@camerondow35 , was the fist outlier test  listed here run just by xylem architecture, or also by site-year-xylem architecture?

@@
 Modeled fits for tree-years were removed under two conditions:
-(1) ~~single day growth rates were~~ $g_{max}$ was <!--confirm this change--> $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
+(1) $g_{max}$ was $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
 (2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101).

### 2022-03-04 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 Modeled fits for tree-years were removed under two conditions:
-(1) $g_{max}$ was $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
+(1) $g_{max}$ was $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11);  <!--run just by xylem architecture, or  by site-year-xylem architecture?-->
 (2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101).

### 2022-03-04 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

--- 2022-03-04 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 Modeled fits for tree-years were removed under two conditions:
-(1) $g_{max}$ was $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
-(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101).
+(1) $g_{max}$ was $\ge 2.5$ standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);  
+(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
 At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.

### 2022-03-04 doc/manuscript/tables_figures/schematic_summary.Rmd
updated schematic summary table
@teixeirak  less than or equal to 3% looks good.

--- 2022-02-24 doc/manuscript/tables_figures/schematic_summary.Rmd
@@
 ```
-*not biologically significant
+*While this result was statistically significant, effect size was small (estimated change $\le 3$%/$^\circ$C) and considered not biologically significant

### 2022-03-04 doc/manuscript/Dow_manuscript_Nature.Rmd
update C sink statistics
#92

@@
 
-In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@friedlingstein_global_2021;@harris_global_2021].
-A substantive portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ [@pugh_role_2019].
-<!-- See issue # 92: https://github.com/EcoClimLab/growth_phenology/issues/92.
-ALSO: The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al.
-2019 estimate that ~200 Tg C yr-1 is attributable to demography (forest regrowth), while >100 Tg C yr-1 is controlled by environmental drivers.
-The latter portion seems more likely to be sensitive to spring temperatures, and I originally cited this number, but upon further thought it's probably complex enough that that's a false distinction.
--->
-The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020].
+In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^ <!--net forest estimate in Harris et al.
+Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
+Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
+Of this important C sink, ~`r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al.
+Table 1)--> occurs in temperate forests[@harris_global_2021], with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019].
+<!-- Pugh et al.
+2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.-->
+The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020].
 

### 2022-03-04 doc/manuscript/Dow_manuscript_Nature.Rmd
update schematic summary + caption
@camerondow35 , this is acceptable now, but another option that could simplify this would be to make a requirement for showing an arrow be that effect size exceeds that defined threshold. I think that would be quite a bit better.

@@
 
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant (p<0.05) increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 

### 2022-03-04 doc/manuscript/tables_figures/schematic_summary.Rmd
update schematic summary + caption
@camerondow35 , this is acceptable now, but another option that could simplify this would be to make a requirement for showing an arrow be that effect size exceeds that defined threshold. I think that would be quite a bit better.

--- 2022-02-24 doc/manuscript/tables_figures/schematic_summary.Rmd
@@
 ```
-*not biologically significant
+*statistically significant, but effect size small ($\le$ 3% $^\circ$C$^{-1}$) and unlikely to be of biological significance

### 2022-03-07 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 - probably include some mention of conifers (for whcih more work has been done)
+- cite @gao_earlier_2022, noting that their result may be driven by conifers
 
@@
 
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of forests to the warming conditions, advancement in the timing of stem growth (Fig.
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
 1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+<!--gao_earlier_2022 demonstrates the same for some other regions.-->
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.

### 2022-03-07 doc/manuscript/Dow_manuscript_Nature.Rmd
remove arrows for non-biologically sig relationships

@@
 
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant (p<0.05) increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/$^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 

### 2022-03-07 doc/manuscript/tables_figures/schematic_summary.Rmd
remove arrows for non-biologically sig relationships

--- 2022-02-25 doc/manuscript/tables_figures/schematic_summary.Rmd
@@
 ```
-*While this result was statistically significant, effect size was small and considered not biologically significant

### 2022-03-07 doc/manuscript/Dow_manuscript_Nature.Rmd
knit in new schematic

--- 2022-03-07 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 - close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016]
-- start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period
+- start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period [@etzold_number_2022]
 - probably include some mention of conifers (for whcih more work has been done)
@@
 
-Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
+Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
 Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
@@
 
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate significant (p<0.05) increases and decreases, respectively, '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%$^\circ$C^-1^), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 

### 2022-03-07 doc/manuscript/Dow_manuscript_Nature.Rmd
first draft of new intro content
@camerondow35 , FYI, and see what you think (when you have a chance). I still need to clean up/ make sure I've captured all the points.

#117

@@
 
-~~~~ (rework starting here)
+~~~~ (new intro content starting here)
+
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
+Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+However, model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- and its seasonality is simplistic relative to *known decoupling of photosynthate production and stem growth [@delpierre_temperate_2016;@delpierre_wood_2016]*.
+<!-- go back and check that these references are appropriate for this statement-->
+Given that the C allocated to woody growth has the longest residence time in the ecosystem[@refs], understanding whether warming spring temperatures will lead to sustained increases in forest C sequestration requires understanding the sensitivity of $ANPP_{woody}$ to spring temperatures.
+
+We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures, particularly for temperate deciduous forests. 
+Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@conifer_refs] and increased annual growth in mesic climates[@gao_earlier_2022;@more_conifer_refs].
+Much less is known about the climate sensitivity of growth of deciduous broadleaf trees, although it has been observed that their growth is more dependent on conditions within the growing season then on growing season length[@etzold_number_2022].
+Furthermore, recent evidence suggests that growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
+Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
+These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014].
+Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
+
+*points to make in new version*:
+
+- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
+- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
+- close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016]
+- start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period [@etzold_number_2022]
+- probably include some mention of conifers (for whcih more work has been done)
+- cite @gao_earlier_2022, noting that their result may be driven by conifers
+
+~~~~ ( old version starting here)
 
@@
 
-*points to make in new version*:
-
-- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
-- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
-- close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016]
-- start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period [@etzold_number_2022]
-- probably include some mention of conifers (for whcih more work has been done)
-- cite @gao_earlier_2022, noting that their result may be driven by conifers
-
-~~~~ (rework ending here)
-
 Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
@@
 Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
+
+~~~~ (end old version)
 

### 2022-03-07 doc/manuscript/Dow_manuscript_Nature.Rmd
working on intro
#117

@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 However, model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- and its seasonality is simplistic relative to *known decoupling of photosynthate production and stem growth [@delpierre_temperate_2016;@delpierre_wood_2016]*.
@@
 
-We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures, particularly for temperate deciduous forests. 
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@conifer_refs] and increased annual growth in mesic climates[@gao_earlier_2022;@more_conifer_refs].
-Much less is known about the climate sensitivity of growth of deciduous broadleaf trees, although it has been observed that their growth is more dependent on conditions within the growing season then on growing season length[@etzold_number_2022].
+We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
+Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs].
+Much less is known about how deciduous broadleaf tree growth responds to spring temperatures.
+While close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the timing of growth earlier alongside leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020], it has been observed that annual growth is more dependent on conditions within the growing season then on growing season length[@etzold_number_2022].
 Furthermore, recent evidence suggests that growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
@@
 These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014].
-Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
+Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms[@refs].
 
@@
 - stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
-- close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016]
-- start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period [@etzold_number_2022]
-- probably include some mention of conifers (for whcih more work has been done)
-- cite @gao_earlier_2022, noting that their result may be driven by conifers
 
@@
 
-While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016] (but see Refs.
-[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).
+~~While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016] (but see Refs.
+[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).~~
 The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].
@@
 
-Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
+~~Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
 Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
-Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
+Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.~~
 
@@
 <!--gao_earlier_2022 demonstrates the same for some other regions.-->
+<!-- cite @gao_earlier_2022, noting that their result may be driven by conifers-->
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.

### 2022-03-08 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 
-In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^ <!--net forest estimate in Harris et al.
+In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
 Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
@@
 Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs].
-Much less is known about how deciduous broadleaf tree growth responds to spring temperatures.
-While close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the timing of growth earlier alongside leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020], it has been observed that annual growth is more dependent on conditions within the growing season then on growing season length[@etzold_number_2022].
-Furthermore, recent evidence suggests that growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
+Much less is known about deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
+However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022].
+Furthermore, *growth of broadleaf deciduous trees may be sink-saturated*, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
 Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
-These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014].
+These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
 Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms[@refs].
@@
 <!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+~~In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
+Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].~~
 However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], *but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016]*,<!--I (KAT) need to work on the part in italics, if retained keenan_timing and zani_increased_2020 do not look at stem growth and are largely contradicted by lu_no_2022.--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
-These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.
+~~These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.~~
 

### 2022-03-08 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@teixeirak I've mad some slight changes to your new intro. Feel free to revert as you like

@@
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+#However, this assumption is overly simplistic given the known decoupling of photosynthate production and stem growth[@delpierre_temperate_2016;@delpierre_wood_2016], suggesting that model representaiton of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- needs to be reworked.#
 However, model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- and its seasonality is simplistic relative to *known decoupling of photosynthate production and stem growth [@delpierre_temperate_2016;@delpierre_wood_2016]*.
@@
 Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs].
-Much less is known about deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
-However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022].
-Furthermore, *growth of broadleaf deciduous trees may be sink-saturated*, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier in deciduous trees alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
+However, an earlier start to growth would not necessarily translate to greater or faster stem growth; instead, growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021] within the growing season rather than to growing season length[@etzold_number_2022].
+Furthermore, *growth of broadleaf deciduous trees may be sink-saturated*, resulting in the lack of more C fixation to stem growth in extended growing seasons[@refs].
 Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].

### 2022-03-08 doc/manuscript/Dow_manuscript_Nature.Rmd
delete old content, misc improvements,  merge conflicted versions
#117

--- 2022-03-08 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 
-~~~~ (new intro content starting here)
-
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
-However, model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- and its seasonality is simplistic relative to *known decoupling of photosynthate production and stem growth [@delpierre_temperate_2016;@delpierre_wood_2016]*.
-<!-- go back and check that these references are appropriate for this statement-->
+However, this assumption is overly simplistic given the known decoupling of photosynthate production and stem growth[@delpierre_temperate_2016;@delpierre_wood_2016], suggesting that model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- needs to be reworked[@delpierre_temperate_2016].
 Given that the C allocated to woody growth has the longest residence time in the ecosystem[@refs], understanding whether warming spring temperatures will lead to sustained increases in forest C sequestration requires understanding the sensitivity of $ANPP_{woody}$ to spring temperatures.
@@
 We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs].
+Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
 Much less is known about deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
 Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
-However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022].
-Furthermore, *growth of broadleaf deciduous trees may be sink-saturated*, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
+However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity, or _NEE_[@delpierre_wood_2016].
+Furthermore, growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
 Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
 These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
-Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms[@refs].
+Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
 
-*points to make in new version*:
+~~~*points to make*:
 
@@
 - stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
+- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]
+- The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].
 
-~~~~ ( old version starting here)
-
-<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
-~~In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].~~
-However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], *but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016]*,<!--I (KAT) need to work on the part in italics, if retained keenan_timing and zani_increased_2020 do not look at stem growth and are largely contradicted by lu_no_2022.--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018].
-~~These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.~~
-
-~~While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016] (but see Refs.
-[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).~~
-The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].
-<!--this sentence is repetitive with with previous.-->
-
-*<PLACEHOLDER>* Three cambial processes drive the relationship between stem growth and woody productivity: cell differentiation, cell enlargement, and cell wall thickening.
-Irreversible stem expansion occurs as a result of differentiation (20% of stem growth) and expansion (80% of stem growth), and accounts for ~10% of annual woody productivity[@cuny_woody_2015].
-The remaining 90% of woody productivity is caused by cell wall thickening.
-It would stand to reason that years with identical stem growth have a similar number of cells which are similarly sized (increased latewood can be an issue here, as you acknowledge).
-If we assume that cell wall thickness is controlled by processes within the tree (ie.
-hormonal responses instead of external climate drivers), then we can predict that cell wall thickness is relatively uniform within each cell type (early wood and late wood).
-Therefore, years with a similar amount of stem growth would sequester a similar amount of carbon during cell wall thickening, and we can say that stem growth equates to woody productivity.
-
-~~Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
-Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
-Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.~~
-
-~~~~ (end old version)
+~~~
 

### 2022-03-08 doc/manuscript/Dow_manuscript_Nature.Rmd
work on intro
#117

@@
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
-However, this assumption is overly simplistic given the known decoupling of photosynthate production and stem growth[@delpierre_temperate_2016;@delpierre_wood_2016], suggesting that model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- needs to be reworked[@delpierre_temperate_2016].
-Given that the C allocated to woody growth has the longest residence time in the ecosystem[@refs], understanding whether warming spring temperatures will lead to sustained increases in forest C sequestration requires understanding the sensitivity of $ANPP_{woody}$ to spring temperatures.
+However, model representation of C allocation to stem growth is overly simplistic in that it does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016].
+As a result, *the spring temperature-sensitivity of* stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016].
+Given that the C allocated to woody growth typically resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], compared to much shorter residence times of other pools such as leaves and fine roots[@refs], understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
 
@@
 Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
-Much less is known about deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
+In contrast, little is known about broadleaf deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
 Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
-However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity, or _NEE_[@delpierre_wood_2016].
-Furthermore, growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
+However, earlier initiation of growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016].
+Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@ref], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
 Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
@@
 - stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
-- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
 - positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
work on intro
#117

@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
-However, model representation of C allocation to stem growth is overly simplistic in that it does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016].
-As a result, *the spring temperature-sensitivity of* stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016].
-Given that the C allocated to woody growth typically resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], compared to much shorter residence times of other pools such as leaves and fine roots[@refs], understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
+Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
+However, model representation of C allocation to stem growth is overly simplistic in that it does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@friend_need_2019].
+As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@friend_need_2019].
+Understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
 
@@
 Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
-However, earlier initiation of growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016].
+However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016].
 Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@ref], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
@@
 
-~~~*points to make*:
-
-- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
-- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]
-- The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].
-
-~~~
-
-Here, we evaluate how early spring temperatures affect peak stem growth timing, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
+Here, we evaluate how spring temperatures affect stem growth timing, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
 To test whether warmer springs extend the period of peak stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
@@
 It may make most sense to condense into a single paragraph making the point that while we don't know where the C goes, we can rule out a substantial, sustained boost to the long-term sink of woody productivity-->
+
+~~~*points to make*:
+
+- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
+- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]
+
+~~~
+
 One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
edit abstract - C cycle implications
#107

@@
 **These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
-**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the long-term CO~2~ sink of temperate deciduous forests.**
+**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**
 

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
references

@@
 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.-->
-The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020].
+The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020;@fatichi_moving_2014].
 
@@
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
-However, model representation of C allocation to stem growth is overly simplistic in that it does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@friend_need_2019].
-As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@friend_need_2019].
+However, model representation of C allocation to stem growth is overly simplistic and does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
+As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014].
 Understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
@@
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016].
-Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@ref], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
+Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
 Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
starting on discussion

@@
 The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014;@finzi_carbon_2020].
-**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise failed to translate extended growing seasons into biologically significant increases in stem growth, and thereby have negligible effect on total annual growth for most species and locations (Fig.
-3).**
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020].
+However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
+`r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig.
+r tree_rings`).
+
+<!--new added text-->
 We find statistically, but not biologically, significant relationships between total growth and spring temperature at both sites in our dendrometer band analysis (`schematic_summary`). 
@@
 In fact, the magnitude of the effect of the summer drought compared to the effects of spring $T_{max}$ illustrates the point that conditions during the peak growing season are much more influential to stem growth than conditions during the Spring.
-**We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig.
-`r tree_rings`).** <!--repeated sentence/ CD made potential changes to the first one.
-See what you think.
--->
+<!--end new added text-->
+
 Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].

### 2022-03-09 doc/manuscript/Dow_Extended_Data.Rmd
a bit of cleanup

@@
 ```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
-path_to_results <- paste0(dirname(dirname(getwd())), "/results/") #I was trying to read directly from results folder, but couldn't get it working.
+path_to_results <- paste0(dirname(dirname(getwd())), "/results/")
 
@@
 EDT_tree_ring_summary="Extended Data Table 3"
-
-sample_size_by_year= "XX"
-#chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions
 

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
remove new added text in discussion
#122

@@
 r tree_rings`).
-
-<!--new added text-->
-We find statistically, but not biologically, significant relationships between total growth and spring temperature at both sites in our dendrometer band analysis (`schematic_summary`). 
-However, the slope of the relationships are in opposite directions, with warming causing decreasing stem growth at Harvard Forest but increasing stem growth at SCBI (`EDT_phenology_parameters`). 
-We believe this difference is due to severe drought conditions in 1999 which obviously caused abnormal growth patterns in that year (`r EDF_pheno_Tmax_climwin`). 
-Therefore, we suggest that our two sites represent two scenarios: 1.
-A forest which had no significant summer drought, but was only influenced by warming temperatures (SCBI).
-2.
-A forest which experienced spring warming along with severe summer drought (Harvard Forest).
-
-At SCBI, we see a very small increase in total stem growth (on average +0.05mm/$^\circ$C), suggesting that even in the absence of drought or inclement climate conditions during the growing season, spring warming does not enhance stem growth in a meaningful way.
-At Harvard Forest, we see a decrease in total growth with warmer temperatures in the CTW (-0.15mm/$^\circ$C on average).
-We speculate that this result is likely a product of reductions in total stem growth related to drought effects in 1999.
-Due to the method of selection for CTW, years with significant departures from the norm can influence the analysis to a greater extent. 
-Stem growth in 1999 was clearly impacted by climate conditions (`r EDF_pheno_Tmax_climwin`), and so the CTW selected by climwin maximizes the effect of these impacts (most likely the effects of drought) on stem growth.
-To examine this further, we reran our dendrometer band analysis using the month of May as the CTW. 
-When we change the CTW to the month of May, we see results much closer to those found at SCBI, with a similar advancement in growth timing but a much less pronounced effect of $T_{max}$ on total growth (-0.015mm/$^\circ$C on average, a tenfold decrease compared to the CTW optimized around the drought year;`EDF_pheno_Tmax_AprilMay`).
-Once again see that there was no meaningful influence of spring temperature on total growth. 
-In fact, the magnitude of the effect of the summer drought compared to the effects of spring $T_{max}$ illustrates the point that conditions during the peak growing season are much more influential to stem growth than conditions during the Spring.
-<!--end new added text-->
-
 Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
@@
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].
-However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
-`r schematic_summary`) were not attributable to summer drought.
+*However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
+`r schematic_summary`) were not attributable to summer drought.* <!--issue #122-->
 

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
a couple small changes requested by R1

@@
 
-To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
+To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010].
 The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020;@begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
@@
 
-For each species-site combination, we converted tree-ring records into the dimensionless RWI to emphasize interannual variability associated with climate [@cook_methods_1990].
+For each species-site combination, we converted tree-ring records into the dimensionless RWI to emphasize interannual variability associated with climate[@cook_methods_1990].
 A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
work on discussion
- delete unneeded notes
- add notes on points to hit

@@
 r tree_rings`).
-Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
-<!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF.
-Maxrate_DOY for RP at HF was actually significantly delayed.
-This could be a consequence of the severe drought in 1999 which we've included in our analysis.
-My brain is completely fried to the point that I barely remember what the literature claims this max rate is related to (water stress, right?).
-So Maybe you're in a better position to add some discussion about this here.
--->
+Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
 Our tree-ring analysis (Fig.
@@
 - positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]
-
+- our study provides further evidence of sink limitation to growth
+- we do not know where the extra C is going.
+If we look at instances where growth is more strongly limited than GPP, we see that NSCs can build up[@refs], and this build up can feed back to reduce photosynthesis[@refs].
+We can also look to studies documenting enhanced GPP with no enhancement of growth.
+In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
 ~~~
@@
 One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
-It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
+It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
 However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
adjust methods to capture inter-annual differences in climate/ measurements
@camerondow35 , these changes are related to our discussion around issue #122 .

@@
 Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera* L.), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
-The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1° C in January and 24° C in July[@anderson-teixeira_joint_2021].
+The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2021].
+Within the study period for the dendrometer band analysis (2011-2020), ...*[some stats on climate]*
 
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2021]. 
-The summer of 1999 was abnormally dry.
-Records indicate that in 1999 only 52mm of precipitation fell per month June-August compared to average monthly precipitation values of 100mm+ [@dorangeville_peak_2021].
+Within the study period for the dendrometer band analysis (1999-2003), ...*[some stats on climate]*
+The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2021].
 
@@
 Measurements were continued through to fall leaf senescence, with the median end date being October 17 and the latest end date November 26 (2012). 
-Timing of measurements at Harvard Forest were similar, with the median start date of April 23 and median end date of October 30. 
-1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15. 
+*Timing of measurements at Harvard Forest were similar, with the median start date of April 23 and median end date of October 30.* <!--Does this change?-->
+All measurements from 1998 were dropped because of a late start date (**DATE**).
+*1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15.* <!--I think we dropped this.Correct?-->
 The latest end date was November 11, 2002.

### 2022-03-09 doc/manuscript/Dow_Extended_Data.Rmd
Update Dow_Extended_Data.Rmd

@@
 ```
-$\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
+$^\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
 1).

### 2022-03-10 doc/manuscript/Dow_Extended_Data.Rmd
update text & figure captions for issue #123

@@
 EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
-EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
+EDF_pheno_Tmax_month = "Extended Data Figure 5"
 EDF_tree_ring_figs= "Extended Data Figure 6-7"
@@
 
-![**Extended Data Figure 5 | Response of stem growth timing and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
+![**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the CTW.
+Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
@@
 For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI.
-Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_month.png){ width=90% }
 

### 2022-03-10 doc/manuscript/Dow_manuscript_Nature.Rmd
update text & figure captions for issue #123

@@
 To ensure that patterns were robust under an alternative definition of CTW, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
-`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the CTW to be the month of April.
-This was consistent with the periods identified by *climwin* for ring- and diffuse porous species groups at both sites, all of which included all or part of April (`r  EDT_phenology_parameters`).
+`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the CTW identified by *climwin* to be the month with the most days in the CTW (`r  EDT_phenology_parameters`) each critical window.
+The months identified were March and April for ring- and diffuse-porous species at SCBI, respectively, and April and May for ring- and diffuse-porous species at Harvard Forest, respectively.
 

### 2022-03-10 doc/manuscript/Dow_manuscript_Nature.Rmd
adjustments related to issue #124

@@
 Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
+There were even fewer significant positive correlations with March and May $T_{max}$: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
 In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
@@
 The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
-We defined chronology start year (Extended Data Table 1) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
+We defined chronology start year (`r EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
 
@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and $RWI$ were assessed using *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
+Correlations between monthly climate and RWI were assessed using *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
 We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
 To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
-Correlations and significance levels for months April-August are given in `r chron_table`.
+Correlations and significance levels for months March-August are given in `r chron_table`.
 
@@
 We used a multivariate model to test for joint effects of April and summer $T_{max}$ on RWI.
+We focused on April to represent spring temperatures because it was the month with greatest overall alignment with the CTW's identified in the dendrometer band analysis and had the highest rate of positive correlations with RWI (`r EDT_tree_ring_summary`).
 We began by testing univariate correlations of $T_{max}$ over three summer windows: June, June-July, and May-August. 

### 2022-03-10 doc/manuscript/Dow_Extended_Data.Rmd
update EDT3 to include March
#124

@@
                                    stringsAsFactors = FALSE, check.names = FALSE)
-TreeRingResultsSummary <- TreeRingResultsSummary[c(2:17)]
+TreeRingResultsSummary <- TreeRingResultsSummary[c(2:19)]
 kable(TreeRingResultsSummary, 
       caption = tree_ring_summary_caption,  
-      col.names = c("species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
+      col.names = c("species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
       format = "latex", booktabs = TRUE,  escape = F) %>%
-add_header_above(c(" " = 2,  "April" = 2, "May" = 2,"June" = 2,"July" = 2,"Aug" = 2,"April" = 2,"June-July" = 2)) %>%
-add_header_above(c(" " = 2,  "univariate analysis" = 10,"multivariate analysis" = 4)) %>%
-add_header_above(c(" " = 2,  "n. correlations to monthly mean max. T significant at p=0.05" = 14)) %>%
+add_header_above(c(" " = 2,  "March" =2 , "April" = 2, "May" = 2,"June" = 2,"July" = 2,"Aug" = 2,"April" = 2,"June-July" = 2)) %>%
+add_header_above(c(" " = 2,  "univariate analysis" = 12,"multivariate analysis" = 4)) %>%
+add_header_above(c(" " = 2,  "n. correlations to monthly mean max. T significant at p=0.05" = 16)) %>%
     column_spec(3:18, width = ".4cm") %>%
@@
 For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI.
-Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_month.png){ width=90% }
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }
 

### 2022-03-10 doc/manuscript/Dow_manuscript_Nature.Rmd
minor

@@
 - stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
-- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]
+- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
 - our study provides further evidence of sink limitation to growth
@@
 pos.
-correlations for 5/142 RP and 3/66 DP species-site combinations, Fig.
-`r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
+correlations for 5/142 RP and 3/66 DP species-site combinations, `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
 Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

### 2022-03-10 doc/manuscript/Dow_manuscript_Nature.Rmd

@@
 We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
+Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
 In contrast, little is known about broadleaf deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
update dendroband counts

@@
 n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
-n_dendro_trees_SCBI <- 123 #doing this the lazy way. summed from species_table rather than coded. 
-n_dendro_trees_HF <- 340 #doing this the lazy way. summed from species_table rather than coded. 
+n_dendro_trees_SCBI <- 119 #doing this the lazy way. summed from species_table rather than coded. 
+n_dendro_trees_HF <- 321 #doing this the lazy way. summed from species_table rather than coded. 
 n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
move presentation of previous year analysis to results

@@
 In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
-$T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
+$T_{min}$ generally exhibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
+
+To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and $T_{max}$ in the previous year (`r EDF_tree_rings_TMAX_extended`).
+This revealed little correlation between spring temperatures and growth the following year, with significant positive correlations of RWI to March or April $T_{max}$ for 5/142 ring porous chronologies and to April or May $T_{max}$ for 7/66 diffuse porous chronologies.
 
@@
 `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].
-Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021].
+Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
 *However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
@@
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
-We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
+We show that the extra C fixation in years with warm springs does not substantially augment stem growth in the current of following year (`r EDF_tree_rings_TMAX_extended`), but it remains unclear how it is allocated within the ecosystem.
 There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
@@
 However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
-Extra C could also be saved within trees as non-structural carbohydrates (NSC) and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014].
-Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig.
-pos.
-correlations for 5/142 RP and 3/66 DP species-site combinations, `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
 Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
remove text on spring T- summer drought correlation from discussion
A reviewer had trouble with this sentence and with the fact that it wasn't described in results. I started to add text on this to results, but don't think it warrants the discussion there.

@@
 
-<!-- Spring warming at SCBI and Harvard Forest in the study years used here did not result in increased $\Delta DBH$ or summer water stress (SPEI; Supp(?) SPEI Figure).
-This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at these sites.
-This could be because the woody growth at these sites are already limited by some mechanism.
--->
-
 ## Tree-ring analysis {-#Tree}
@@
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
-*However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig.
-`r schematic_summary`) were not attributable to summer drought.* <!--issue #122-->
 

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
minor edits

@@
 `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
-In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
+In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April $T_{max}$.
 Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
@@
 
-<!--Currently 1005. Goal: 700-800-->
-
+<!--1. summary-->
 Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs.
@@
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
+However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
 `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig.
-r tree_rings`).
+`r tree_rings`).
 Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
@@
 `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].
+
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
@@
 - positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
-- our study provides further evidence of sink limitation to growth
+- our study adds to a growing body of literature demonstrating a sink limitation of stem growth and $ANPP_{woody}$ [@refs].
+Specifically, [explain].
+The implication is that while elements of global change such as warming spring temperatures or elevated CO~2~[@refs] may enhance GPP, woody growth remains constrained.
 - we do not know where the extra C is going.

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
restructuring discussion

@@
 
-<!--1. summary-->
+<!--summary-->
 Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs.
@@
 
-Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
-
+<!--expectation of extra C, not going to ANPP_woody (sink limitation)-->
 Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022;@tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016;@fu_maximum_2019].
@@
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
-We show that the extra C fixation in years with warm springs does not substantially augment stem growth in the current of following year (`r EDF_tree_rings_TMAX_extended`), but it remains unclear how it is allocated within the ecosystem.
-There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
+We show that the extra C fixation in years with warm springs does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`).
 
-<!-- The following two paragraphs--and two final sentences in the paragraph above-- need work (assigned to KAT).
-It may make most sense to condense into a single paragraph making the point that while we don't know where the C goes, we can rule out a substantial, sustained boost to the long-term sink of woody productivity-->
+Although stem diameter increments are routinely used to calculate $ANPP_{woody}$[@refs], stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016 and refs therein].
+Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
+However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
+*Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that $ANPP_{woody}$ is limited by sink factors, such that the expected increase photosynthate availability[@refs] (observed at Harvard Forest[@finzi_carbon_2020_but_verify]) does not enhance growth.*
+*Indeed, such a phenomenon has been observed when GPP has been increased through lengthening of the growing season[@refs] or elevated CO~2~[@refs].*
 
-~~~*points to make*:
+<!--on the fate of extra C-->
+*notes:*
 
-- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
-- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
 - our study adds to a growing body of literature demonstrating a sink limitation of stem growth and $ANPP_{woody}$ [@refs].
@@
 In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
-~~~
+- lessons learned from the more intense study of CO2 fertilization imply that stimulation of ANPP_woody is likely to be less than that of GPP
 
-One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
-It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
-However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
-Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.
-
+*old content:*
 A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC storage, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
@@
 
+<!--... short vs long-term effects-->
+
+*notes:*
+
+- we show lack of effect on interannual scale 
+- feedbacks such as greater nutrient acquisition could change this.
+
+*old content:*
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
@@
 
+<!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
+
+- Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
+- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
+- our analysis isn't set up for detecting interaction. But, sites are fairly mesic. Mechanisms remain to be understood, but the net result is the same.
+
+<!--conclusions-->
+
+*old content:*
 As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
working on discussion

@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
+Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
 However, model representation of C allocation to stem growth is overly simplistic and does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
@@
 Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
-However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016].
+However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
 Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
@@
 The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020].
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020].
 However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
@@
 <!--expectation of extra C, not going to ANPP_woody (sink limitation)-->
-Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022;@tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016;@fu_maximum_2019].
-However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].
+Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022;@tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for NEE[@zhou_explaining_2016;@fu_maximum_2019].
+However, there is also a disconnect with findings that NEE is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].
 Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
@@
 <!--on the fate of extra C-->
-*notes:*
+Our results add to a growing body of evidence for a sink limitation of stem growth and $ANPP_{woody}$[@refs], wherein global change factors such as warming spring temperatures[@refs] or elevated CO~2~[@refs] enhance GPP, but woody growth is less responsive, if at all[@refs].
+The fate of any additional C fixed during years with warm springs remains unresolved, but possible *destinations* -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times that woody growth.
+Indeed, in a situation where GPP was enhanced (through experimental elevation of CO~2~) but $ANPP_{woody}$ was constrained *(presumably by sink factors)*, additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
+Indeed, it has been observed that C loss through fall or winter respiration can offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
+It is also possible that a buildup of NSCs that are not required for growth or other functions would feed back to reduce the rate of photosynthesis[@refs], such that sink limitation of $ANPP_{woody}$ reduces GPP relative to its potential in years with warm springs.
+Either way, our results demonstrate that warm springs do not increase $ANPP_{woody}$ on an interannual time scale.
 
-- our study adds to a growing body of literature demonstrating a sink limitation of stem growth and $ANPP_{woody}$ [@refs].
-Specifically, [explain].
-The implication is that while elements of global change such as warming spring temperatures or elevated CO~2~[@refs] may enhance GPP, woody growth remains constrained.
-- we do not know where the extra C is going.
-If we look at instances where growth is more strongly limited than GPP, we see that NSCs can build up[@refs], and this build up can feed back to reduce photosynthesis[@refs].
-We can also look to studies documenting enhanced GPP with no enhancement of growth.
-In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
-- lessons learned from the more intense study of CO2 fertilization imply that stimulation of ANPP_woody is likely to be less than that of GPP
-
-*old content:*
-A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC storage, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates.
-Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
-However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
-Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
-However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.
-
-<!--... short vs long-term effects-->
+<!--.short vs long-term effects-->
+It is possible that
 
@@
 *old content:*
+
+However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
+Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
+
 Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
@@
 Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the seasonality of forest C cycling.
-If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
+If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
 Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.
@@
 <!-- cite @gao_earlier_2022, noting that their result may be driven by conifers-->
-Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.
+<!--lessons learned from the more intense study of CO2 fertilization imply that stimulation of ANPP_woody is likely to be less than that of GPP-->
+Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig.
+`r tree_rings`), which constitutes a positive feedback to climate change.
 

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
work on discussion

@@
 It is also possible that a buildup of NSCs that are not required for growth or other functions would feed back to reduce the rate of photosynthesis[@refs], such that sink limitation of $ANPP_{woody}$ reduces GPP relative to its potential in years with warm springs.
-Either way, our results demonstrate that warm springs do not increase $ANPP_{woody}$ on an interannual time scale.
+Either way, our results demonstrate that warm springs do not increase $ANPP_{woody}$ on an interannual time scale, contrasting with their tendency to augment _NEE_[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
 
 <!--.short vs long-term effects-->
-It is possible that
-
-*notes:*
-
-- we show lack of effect on interannual scale 
-- feedbacks such as greater nutrient acquisition could change this.
-
-*old content:*
-
-However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
-Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
-
-Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
-As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and C cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
-Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the seasonality of forest C cycling.
-If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
-Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.
+It is possible that, as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
+Across spatial gradients in spring temperature, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
+Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth, but that potential is not currently being realized on an interannual time frame.
+If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
+Therefore, understanding how warming spring temperatures are influencing C allocation within temperate deciduous forest ecosystems remains an important outstanding question.
 

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion

@@
 <!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
+Irregardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season.
+Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig.
+`r tree_rings`, `r EDT_tree_ring_summary`).
+This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress[@refs] or high temperatures[@refs] during the peak growing season.
 

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
finished restructuring discussion
@camerondow35 , FYI. See what you think. :-)

This does still need some work, but I think the content is much improved.

@@
 
-Although stem diameter increments are routinely used to calculate $ANPP_{woody}$[@refs], stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016 and refs therein].
+Although stem diameter increments are routinely used to calculate $ANPP_{woody}$[@refs], stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016; @and_refs_therein].
 Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
@@
 This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress[@refs] or high temperatures[@refs] during the peak growing season.
-
-- Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
-- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
-- our analysis isn't set up for detecting interaction. But, sites are fairly mesic. Mechanisms remain to be understood, but the net result is the same.
+<!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of $T_{max}$ in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
+Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
+While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@refs].
 
 <!--conclusions-->
-
-*old content:*
 As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
word + citation count
#125
Update Dow_manuscript_Nature.Rmd
removed several sources

@@
 The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2597 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently ok there-- n= 48)-->
+We're currently at 3005 words and 1 modest + 2 large display items.
+Articles typically have 30-50 references (currently at 59)-->
 
@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999-->;@menzel_european_2006;<!--@ibanez_forecasting_2010-->;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010-->;@menzel_european_2006;<!--@menzel_growing_1999-->;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
@@
 For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of seasonal milestones.
-Consistent with findings from previous studies[@parmesan_globally_2003;@ibanez_forecasting_2010;@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig.
+Consistent with findings from previous studies[<!--@parmesan_globally_2003-->;<!--@ibanez_forecasting_2010-->;<!--@friedl_tale_2014-->], leaf phenological milestones advanced at both sites (Fig.
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
@@
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible *destinations* -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times that woody growth.
-Indeed, in a situation where GPP was enhanced (through experimental elevation of CO~2~) but $ANPP_{woody}$ was constrained *(presumably by sink factors)*, additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
+In fact, in a situation where GPP was enhanced (through experimental elevation of CO~2~) but $ANPP_{woody}$ was constrained *(presumably by sink factors)*, additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
 Indeed, it has been observed that C loss through fall or winter respiration can offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
@@
 <!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
-Irregardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season.
+Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season.
 Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig.
@@
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
-While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@refs].
+While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@refs].
 
 <!--conclusions-->
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
+As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier.
+However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
 1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
work on word/citation limit
Down to ~2700 words and 53 citations

@teixeirak I think I got most of the easy citation removals here. The rest might be difficult.

@@
 
-We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
 Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
-In contrast, little is known about broadleaf deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
+<!--In contrast, little is known about broadleaf deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].-->
+In contrast, we have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
@@
 
-Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
+<!--Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig.
+`r DOY_timing`, `r  EDT_phenology_parameters`).--><!--cool to note, but is it necessary?
+Especially considering how broad "seasonal milestones for BOTH..." is.-->
 Consistent with the results of Ref[@dorangeville_peak_2021], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
@@
 Consistent with findings from previous studies[<!--@parmesan_globally_2003-->;<!--@ibanez_forecasting_2010-->;<!--@friedl_tale_2014-->], leaf phenological milestones advanced at both sites (Fig.
-`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
+`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
 Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the critical temperature window (Figs.
@@
 `r DOY_timing`a-b), there was no consistent lengthening of $L_{PGS}$ (Fig.
-`r schematic_summary`, `r EDF_pheno_all`).
+`r schematic_summary`, `r EDF_pheno_all`).<!--Note to Cam: double check this-->
 
@@
 `r tree_rings`).
-Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
+Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014-->;@delpierre_wood_2016;@delpierre_temperate_2016].
 Our tree-ring analysis (Fig.
-`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].
+`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].<!--We speculate that warmer temps lad to more water stress during the summer but can we claim it like we do here?-->
 
@@
 However, there is also a disconnect with findings that NEE is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].
-Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
+<!--Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.-->
 We show that the extra C fixation in years with warm springs does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`).
@@
 Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
-However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
+However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006-->;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006-->;@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
 *Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that $ANPP_{woody}$ is limited by sink factors, such that the expected increase photosynthate availability[@refs] (observed at Harvard Forest[@finzi_carbon_2020_but_verify]) does not enhance growth.*
@@
 <!--.short vs long-term effects-->
-It is possible that, as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
-Across spatial gradients in spring temperature, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
+It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
+Across spatial gradients in spring temperature, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005]<!--Can we replace this churkina_2005 ref with another NEE ref?-->.<!-- this sentence confuses me a bit.
+It seems like you're saying that warmer springs actually are associated wtih greater stem growth.
+Are you saying that warmer springs increase stem growth in some controlled studies, but that isn't being found in field conditions?
+-->
 Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth, but that potential is not currently being realized on an interannual time frame.

### 2022-03-12 RScripts/HF_wood_phenology.R
Update HF_wood_phenology.R
unintentional changes

@@ -76,11 +76,11 @@ all_stems <- all_stems[complete.cases(all_stems$dbh2),]
 #all_stems <- all_stems_clean
 #all_stems <- all_stems[all_stems$plot_tag %in% tagsubset,]
 for (q in 1998:2003) {
-  q = 1998
+
   skip_to_next <- FALSE
   Stem2 <- subset(all_stems, year == q)
   for (w in unique(Stem2$sp)) { # removes trees with less than 10 measurements in each year
-    w = "betual"
+
     Stem3 <- subset(Stem2, sp == w)
     count_df <- count(Stem3$tag_stem)
     count_df <- subset(count_df, freq >= 10)

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
methods work

@@
 Measurements were continued through to fall leaf senescence, with the median end date being October 17 and the latest end date November 26 (2012). 
-*Timing of measurements at Harvard Forest were similar, with the median start date of April 23 and median end date of October 30.* <!--Does this change?-->
-All measurements from 1998 were dropped because of a late start date (**DATE**).
-*1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15.* <!--I think we dropped this.Correct?-->
+*Timing of measurements at Harvard Forest were similar, with the median start date of April 23 and median end date of October 30.* <!--Changes depending on when we calculate median.
+I think we can calculate it before removal of 1998 since we technically incude 1998 in our raw data, just remove it during our data cleaning.
+If you don't like that, it cahges to: start: april 21, end: October 27-->
+All measurements from 1998 were dropped because of a late start date (May 26).
+*1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15.* <!--I think we dropped this.Correct?
+CD: No, left in.
+This could influence growth in 1999 if "a" was selected based on pre-hydration, winter DBH, but I have faith that Sean's process to find "a" can handle this.
+That being said, it is unfortunate that 1999 is the only year with early measurements and very weird timing.
+Justification for my faith: If "a" uses prehydration, winter dbh, then we would expect the "a" in 1999 to be proportionally smaller than in other years.
+This smaller "a" coupled with an unchanged "b" (last measurement was about the same in 1999 as other years) would result in greater total growth (b-a).
+But we actually see that 1999 total growth is the smallest, indicating to me that the hot summer is driving the weirdness.
+-->
 The latest end date was November 11, 2002.

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
reviewed / responded to @camerondow35's edits, misc edits

@@
 
-<!--NOTE: We're probably pretty close to the upper length limit.
-The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
+<!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
 We're currently at 3005 words and 1 modest + 2 large display items.
@@
 Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
-<!--In contrast, little is known about broadleaf deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].-->
+<!--we can potentially cut some of these references, if not needed elsewhere.-->
 In contrast, we have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
@@
 
-<!--Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig.
-`r DOY_timing`, `r  EDT_phenology_parameters`).--><!--cool to note, but is it necessary?
-Especially considering how broad "seasonal milestones for BOTH..." is.-->
+Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig.
+`r DOY_timing`, `r  EDT_phenology_parameters`).<!--cool to note, but is it necessary?
+Especially considering how broad "seasonal milestones for BOTH..." is.
+KAT: I'd like to try to keep it in because it links to I think we should keep it in because it links-->
 Consistent with the results of Ref[@dorangeville_peak_2021], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
@@
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
-Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the critical temperature window (Figs.
+Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the CTW (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
@@
 
-Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig.
+Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the CTW compared to those with cooler temperatures (Fig.
 `r DOY_timing`a-b), there was no consistent lengthening of $L_{PGS}$ (Fig.
@@
 In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to mean $T_{max}$ during the critical temperature window for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with mean $T_{max}$ during the critical temperature window at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to mean $T_{max}$ during the CTW for ring- and diffuse- porous species.
+$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with mean $T_{max}$ during the CTW at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
 
@@
 Our tree-ring analysis (Fig.
-`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].<!--We speculate that warmer temps lad to more water stress during the summer but can we claim it like we do here?-->
+`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to high $T_{max}$ during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].
 
@@
 Our results add to a growing body of evidence for a sink limitation of stem growth and $ANPP_{woody}$[@refs], wherein global change factors such as warming spring temperatures[@refs] or elevated CO~2~[@refs] enhance GPP, but woody growth is less responsive, if at all[@refs].
-The fate of any additional C fixed during years with warm springs remains unresolved, but possible *destinations* -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times that woody growth.
-In fact, in a situation where GPP was enhanced (through experimental elevation of CO~2~) but $ANPP_{woody}$ was constrained *(presumably by sink factors)*, additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
-Indeed, it has been observed that C loss through fall or winter respiration can offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
+The fate of any additional C fixed during years with warm springs remains unresolved, but possible *destinations* -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
+Indeed, in a situation where GPP was enhanced (through experimental elevation of CO~2~) but $ANPP_{woody}$ was constrained *(presumably by sink factors)*, additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
+Consistent with this, it has been observed that C loss through fall or winter respiration can offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
 It is also possible that a buildup of NSCs that are not required for growth or other functions would feed back to reduce the rate of photosynthesis[@refs], such that sink limitation of $ANPP_{woody}$ reduces GPP relative to its potential in years with warm springs.
@@
 It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
-Across spatial gradients in spring temperature, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005]<!--Can we replace this churkina_2005 ref with another NEE ref?-->.<!-- this sentence confuses me a bit.
-It seems like you're saying that warmer springs actually are associated wtih greater stem growth.
-Are you saying that warmer springs increase stem growth in some controlled studies, but that isn't being found in field conditions?
+Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005]<!--Can we replace this churkina_2005 ref with another NEE ref?
+NO, it's making a different point.-->.<!-- The point here is that if you compare across sites in different climates (Harvard Forest and SCBI, for example), you do see greater stem growth associated with the longer growing seasons.
+Hopefully my rewording makes that clear.
 -->
-Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth, but that potential is not currently being realized on an interannual time frame.
+Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
 If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
-Therefore, understanding how warming spring temperatures are influencing C allocation within temperate deciduous forest ecosystems remains an important outstanding question.
+Therefore, understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.
 
@@
 `r tree_rings`, `r EDT_tree_ring_summary`).
-This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress[@refs] or high temperatures[@refs] during the peak growing season.
+This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress[@refs] or high temperatures[@anderson-teixeira_joint_2021] (which cause drought stress[@refs]) during the peak growing season.
 <!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of $T_{max}$ in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
@@
 1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
-<!--gao_earlier_2022 demonstrates the same for some other regions.-->
-<!-- cite @gao_earlier_2022, noting that their result may be driven by conifers-->
-<!--lessons learned from the more intense study of CO2 fertilization imply that stimulation of ANPP_woody is likely to be less than that of GPP-->
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig.
@@
 # Methods {-#Methods} 
-
 <!--Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.-->
-<!--We are currently at ~2000 words -->
 

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
closes #103

@@
 Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
-If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
+If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
 Therefore, understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
refining text, citations

@@
 ```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
-## potential reviewers: Zani, Keenan
 
@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 3005 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently at 59)-->
+We're currently at 2839 words and 1 modest + 2 large display items.
+Articles typically have 30-50 references (currently at 53)-->
 
@@
 
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010-->;@menzel_european_2006;<!--@menzel_growing_1999-->;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;-->@menzel_european_2006;<!--@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
@@
 `r tree_rings`).
-Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014-->;@delpierre_wood_2016;@delpierre_temperate_2016].
+Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
 Our tree-ring analysis (Fig.
@@
 
-Although stem diameter increments are routinely used to calculate $ANPP_{woody}$[@refs], stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016; @and_refs_therein].
+Although stem diameter increments are routinely used to calculate $ANPP_{woody}$, stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].
 Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
-However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006-->;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006-->;@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
-*Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that $ANPP_{woody}$ is limited by sink factors, such that the expected increase photosynthate availability[@refs] (observed at Harvard Forest[@finzi_carbon_2020_but_verify]) does not enhance growth.*
-*Indeed, such a phenomenon has been observed when GPP has been increased through lengthening of the growing season[@refs] or elevated CO~2~[@refs].*
+However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
+Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that growth is sink-limited, such that increased GPP does not enhance $ANPP_{woody}$[@delpierre_wood_2016].
 
 <!--on the fate of extra C-->
-Our results add to a growing body of evidence for a sink limitation of stem growth and $ANPP_{woody}$[@refs], wherein global change factors such as warming spring temperatures[@refs] or elevated CO~2~[@refs] enhance GPP, but woody growth is less responsive, if at all[@refs].
-The fate of any additional C fixed during years with warm springs remains unresolved, but possible *destinations* -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
-Indeed, in a situation where GPP was enhanced (through experimental elevation of CO~2~) but $ANPP_{woody}$ was constrained *(presumably by sink factors)*, additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
-Consistent with this, it has been observed that C loss through fall or winter respiration can offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
-It is also possible that a buildup of NSCs that are not required for growth or other functions would feed back to reduce the rate of photosynthesis[@refs], such that sink limitation of $ANPP_{woody}$ reduces GPP relative to its potential in years with warm springs.
-Either way, our results demonstrate that warm springs do not increase $ANPP_{woody}$ on an interannual time scale, contrasting with their tendency to augment _NEE_[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
+Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, fail to stimulate $ANPP_{woody}$[@jiang_fate_2020;@refs].
+The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
+Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, $ANPP_{woody}$ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
+Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018].
+Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be short lived.
 
@@
 <!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
-Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season.
+Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season, when increased atmospheric demand associated with high temperatures limits growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
 Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig.
 `r tree_rings`, `r EDT_tree_ring_summary`).
-This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress[@refs] or high temperatures[@anderson-teixeira_joint_2021] (which cause drought stress[@refs]) during the peak growing season.
+This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress or high temperatures during the peak growing season[@dorangeville_drought_2018;@helcoski_growing_2019;@anderson-teixeira_joint_2021;@babst_twentieth_2019].
 <!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of $T_{max}$ in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
-While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@refs].
+While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@gao_earlier_2022].
 

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
condensing discussion

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2839 words and 1 modest + 2 large display items.
+We're currently at 2763 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently at 53)-->
@@
 
-<!--summary-->
+<!--summary, setting expectation of extra C-->
 Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs.
-`r schematic_summary`- `r tree_rings`).
+`r schematic_summary`-`r tree_rings`).
 The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
@@
 `r tree_rings`).
-Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
-Our tree-ring analysis (Fig.
-`r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to high $T_{max}$ during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].
+Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
+Combined with observations that warming springs tend to lengthen the season of C sequestration[@refs], our observations imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
+What happens with the extra C fixed during this period?
 
-<!--expectation of extra C, not going to ANPP_woody (sink limitation)-->
-Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022;@tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for NEE[@zhou_explaining_2016;@fu_maximum_2019].
-However, there is also a disconnect with findings that NEE is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].
-<!--Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.-->
-We show that the extra C fixation in years with warm springs does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`).
-
-Although stem diameter increments are routinely used to calculate $ANPP_{woody}$, stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].
+<!--extra C is probably not going to ANPP_woody-->
+Additional C fixed in years with warm springs does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
 Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
minor

@@
 Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
-Combined with observations that warming springs tend to lengthen the season of C sequestration[@refs], our observations imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
+Combined with observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
 What happens with the extra C fixed during this period?
@@
 <!--extra C is probably not going to ANPP_woody-->
-Additional C fixed in years with warm springs does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
-Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
+Additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
+It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
 However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
tweaking

@@
 Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
-Combined with observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
-What happens with the extra C fixed during this period?
+Combined with observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion.
+What happens with is extra C?
 

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
minor

@@
 `r tree_rings`).
-Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
+Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
 Combined with observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion.
-What happens with is extra C?
+What happens with this extra C?
 

### 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
fix refs

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999-->;@menzel_european_2006;<!--@ibanez_forecasting_2010-->;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;-->@menzel_european_2006;<!--@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**

### 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
comment out sample sizes
@camerondow35 , was this change specifically requested by reviewers? It really breaks up the sentence/ makes it difficult to read. I instead added a citation of the table with this info. If we need to present these numbers in the text, I'd find a different way to do it, but I don't see why citing the table wouldn't be sufficient.

@@
 `r schematic_summary`).
-This analysis was performed separately for ring- (67 trees at SCBI and 138 trees at Harvard Forest) and diffuse porous (52 trees at SCBI and 183 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
+This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

### 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
remove results paragraph comparing sites & species groups
@camerondow35 , on second thought, I think you were right that we could remove the comparison of sites, and also species groups. (I'd keep all or none.)

See what you think.

--- 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- (67 trees at SCBI and 138 trees at Harvard Forest) and diffuse porous (52 trees at SCBI and 183 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022].
+This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth(`r  EDT_phenology_parameters`)[@dorangeville_peak_2021;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 
-Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig.
-`r DOY_timing`, `r  EDT_phenology_parameters`).<!--cool to note, but is it necessary?
-Especially considering how broad "seasonal milestones for BOTH..." is.
-KAT: I'd like to try to keep it in because it links to I think we should keep it in because it links-->
-Consistent with the results of Ref[@dorangeville_peak_2021], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
-`r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).
-Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only six days longer at SCBI for ring porous species and less than one day longer for diffuse porous species (`r  EDT_phenology_parameters`).
+<!--Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).-->
+<!--Consistent with the results of Ref[@dorangeville_peak_2021], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
+`r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).-->
+<!--Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only six days longer at SCBI for ring porous species and less than one day longer for diffuse porous species (`r  EDT_phenology_parameters`).-->
+
+Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig.
+`r DOY_timing`, `r EDF_pheno_all`).
+We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW).
+CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
+The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.
 
@@
 Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring porous window at HF and diffuse porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }
-
-Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig.
-`r DOY_timing`, `r EDF_pheno_all`).
-We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW).
-CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
-The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.
 

### 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
update word count
#125
wordsmithing - results

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2763 words and 1 modest + 2 large display items.
+We're currently at 2641 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently at 53)-->
@@
 
-Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig.
-`r DOY_timing`, `r EDF_pheno_all`).
+Both canopy foliage phenology and stem growth timing shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
 We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW).
-CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
-The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.
+CTW was identified by measuring the correlation between weekly $T_{max}$ and $DOY_{25}$ for all combinations of consecutive weeks from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
+The CTW was defined as the weeks for which this correlation was strongest, and mean $T_{max}$ during this period (CTW $T_{max}$) used as the independet variable.
 
@@
 Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
-Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring porous window at HF and diffuse porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }
+Mean maximum temperature ($T_{max}$) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+Leaf phenology years are colored according to the CTW $T_{max}$ with the best correlation with timing of greenup (diffuse porous window at SCBI and ring porous window at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }
 
-For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of seasonal milestones.
+For ring- and diffuse- porous species at both sites, warmer CTW $T_{max}$ resulted in earlier achievement of seasonal milestones.
 Consistent with findings from previous studies[<!--@parmesan_globally_2003-->;<!--@ibanez_forecasting_2010-->;<!--@friedl_tale_2014-->], leaf phenological milestones advanced at both sites (Fig.
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
-Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the CTW (Figs.
+Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with CTW $T_{max}$ (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
@@
 
-Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the CTW compared to those with cooler temperatures (Fig.
-`r DOY_timing`a-b), there was no consistent lengthening of $L_{PGS}$ (Fig.
+Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with CTW $T_{max}$ (Fig.
+`r DOY_timing`a-b)<!--specify which species group used?-->, there was no consistent lengthening of $L_{PGS}$ (Fig.
 `r schematic_summary`, `r EDF_pheno_all`).<!--Note to Cam: double check this-->
 
-In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to mean $T_{max}$ during the CTW for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with mean $T_{max}$ during the CTW at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
+In contrast to the pronounced effects of CTW $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to CTW $T_{max}$ for ring- and diffuse- porous species.
+$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with CTW $T_{max}$ at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
 

### 2022-03-13 doc/manuscript/Dow_Extended_Data.Rmd
update criteria for CTW selection for leaf phenology
@camerondow35 , I think the better criteria to state here is that we used the group containing the dominant species at each site. Liriodendron accounts for almost 1/2 of productivity at SCBI (and probably more than half of canopy productivity), and red oak is probably >90% at HF.

@@
 ```
-$^\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (Fig.
+$^\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (CTW, Fig.
 1).
 Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straightforward.
-We used the critical window which had the greatest correlation with leaf greenup timing as the critical window for leaf phenology calculations.
-This resulted in the selection of the diffuse-porous critical window at SCBI and the ring-porous critical window at HF, corresponding to the grouping of the species with greatest woody productivity at each site (diffuse-porous *Liriodendron tulipifera* at SCBI, ring-porious *Quercus rubra* at HF).
+We used the CTW of the porosity group containing the dominant canopy species at each site: *Liriodendron tulipifera* (diffuse porous) at SCBI and *Quercus rubra* (ring porous) at Harvard Forest.
 

### 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
update criteria for CTW selection for leaf phenology
@camerondow35 , I think the better criteria to state here is that we used the group containing the dominant species at each site. Liriodendron accounts for almost 1/2 of productivity at SCBI (and probably more than half of canopy productivity), and red oak is probably >90% at HF.

@@
 Mean maximum temperature ($T_{max}$) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to the CTW $T_{max}$ with the best correlation with timing of greenup (diffuse porous window at SCBI and ring porous window at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }
+Leaf phenology years are colored according to the CTW $T_{max}$ of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }
 
@@
 
-Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with CTW $T_{max}$ (Fig.
-`r DOY_timing`a-b)<!--specify which species group used?-->, there was no consistent lengthening of $L_{PGS}$ (Fig.
+Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the CTW $T_{max}$ of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig.
+`r DOY_timing`a-b), there was no consistent lengthening of $L_{PGS}$ (Fig.
 `r schematic_summary`, `r EDF_pheno_all`).<!--Note to Cam: double check this-->

### 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
consistently use "Harvard Forest" as opposed to HF

@@
 In contrast to the pronounced effects of CTW $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to CTW $T_{max}$ for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with CTW $T_{max}$ at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
+Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW $T_{max}$ for ring- and diffuse- porous species.
+$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with CTW $T_{max}$ at SCBI and a weak negative relationship (-0.2-0.03 mm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
 
@@
 If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
-We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253), and tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
+We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253), and tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
 These were removed because they reduce the reliability of predicted growth in the modeled curves.

### 2022-03-14 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
refining results & discussion

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2641 words and 1 modest + 2 large display items.
+We're currently at 2671 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently at 53)-->
@@
 
-Both canopy foliage phenology and stem growth timing shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
+Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
 We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW).
 CTW was identified by measuring the correlation between weekly $T_{max}$ and $DOY_{25}$ for all combinations of consecutive weeks from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
-The CTW was defined as the weeks for which this correlation was strongest, and mean $T_{max}$ during this period (CTW $T_{max}$) used as the independet variable.
+The CTW was defined as the weeks for which this correlation was strongest, and mean $T_{max}$ during this period (CTW $T_{max}$) used as our independent variable.
 
@@
 For ring- and diffuse- porous species at both sites, warmer CTW $T_{max}$ resulted in earlier achievement of seasonal milestones.
-Consistent with findings from previous studies[<!--@parmesan_globally_2003-->;<!--@ibanez_forecasting_2010-->;<!--@friedl_tale_2014-->], leaf phenological milestones advanced at both sites (Fig.
+Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig.
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
@@
 
-In contrast to the pronounced effects of CTW $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
+In contrast to the pronounced effects of CTW $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
 Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW $T_{max}$ for ring- and diffuse- porous species.
@@
 
-To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
+To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
 `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
@@
 To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and $T_{max}$ in the previous year (`r EDF_tree_rings_TMAX_extended`).
-This revealed little correlation between spring temperatures and growth the following year, with significant positive correlations of RWI to March or April $T_{max}$ for 5/142 ring porous chronologies and to April or May $T_{max}$ for 7/66 diffuse porous chronologies.
+This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of RWI to previous March or April $T_{max}$ for 5/142 ring porous chronologies and to previous April or May $T_{max}$ for 7/66 diffuse porous chronologies.
 
@@
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 4% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 4% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies (`r EDT_tree_ring_summary`).
 
@@
 `r schematic_summary`-`r tree_rings`).
-The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig.
+The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig.
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020].
@@
 <!--extra C is probably not going to ANPP_woody-->
-Additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
+Additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$.
+Although $ANPP_{woody}$ is routinely calculated based on stem growth, the two can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].
 It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
@@
 If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
-Therefore, understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.
+Understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.
 
 <!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
-Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season, when increased atmospheric demand associated with high temperatures limits growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
+Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season, when increased atmospheric demand associated with high temperatures limits growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
 Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig.
@@
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
-While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@gao_earlier_2022].
+While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), our tree-ring analysis does reveal a higher frequency of negative than positive correlations of annual growth to spring temperatures, particularly for ring porous species in cooler climates (Fig.
+`r tree_rings`, `r EDT_tree_ring_summary`).
+Thus, warm springs can have a net negative effect on growth at some times or places[@gao_earlier_2022].
 

### 2022-03-14 doc/manuscript/Dow_manuscript_Nature.Rmd
#92

@@
 Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
-Of this important C sink, ~`r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al.
-Table 1)--> occurs in temperate forests[@harris_global_2021], with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019].
+<!--@xu_changes_2021 provides an alternate estimate, which is about half of Harris.
+The main difference being that Xu includes savannas and woodlands, which have experienced a lot of loss from fires.
+Given focus on forests here, Harris seems more appropriate.
+-->
+Of this important C sink, ~ `r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021],
+<!-- alternate estimate: `r round(.31/.88,2)*100`% (temperate net / global net, (@xu_changes_2021. Table 2))--> 
+with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019].
 <!-- Pugh et al.

### 2022-03-14 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
-Combined with observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion.
+Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion.
 What happens with this extra C?
@@
 <!--extra C is probably not going to ANPP_woody-->
-Additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$.
+The presumed additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$.
 Although $ANPP_{woody}$ is routinely calculated based on stem growth, the two can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].
@@
 However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
-Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that growth is sink-limited, such that increased GPP does not enhance $ANPP_{woody}$[@delpierre_wood_2016].
+Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that growth is sink-limited[@delpierre_wood_2016;@refs].
 
 <!--on the fate of extra C-->
-Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, fail to stimulate $ANPP_{woody}$[@jiang_fate_2020;@refs].
+Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, not cause a corresponding increase in $ANPP_{woody}$[@jiang_fate_2020;@refs].
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
@@
 It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
-Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005]<!--Can we replace this churkina_2005 ref with another NEE ref?
-NO, it's making a different point.-->.<!-- The point here is that if you compare across sites in different climates (Harvard Forest and SCBI, for example), you do see greater stem growth associated with the longer growing seasons.
-Hopefully my rewording makes that clear.
--->
+Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
 Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.

### 2022-03-18 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 <!--on the fate of extra C-->
-Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, not cause a corresponding increase in $ANPP_{woody}$[@jiang_fate_2020;@refs].
+Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in $ANPP_{woody}$[@jiang_fate_2020;@refs].
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
@@
 `r tree_rings`, `r EDT_tree_ring_summary`).
-Thus, warm springs can have a net negative effect on growth at some times or places[@gao_earlier_2022].
+Thus, warm springs can have a net negative effect on growth, particularly when water is limiting[@gao_earlier_2022].
 

### 2022-03-18 doc/manuscript/Dow_manuscript_Nature.Rmd
limit discussion of extra C to one discussion paragraph

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2671 words and 1 modest + 2 large display items.
+We're currently at 2644 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently at 53)-->
@@
 Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
-Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion.
-What happens with this extra C?
+This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@jiang_fate_2020;@refs].
 
-<!--extra C is probably not going to ANPP_woody-->
-The presumed additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$.
-Although $ANPP_{woody}$ is routinely calculated based on stem growth, the two can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].
-It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
-However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely.
-Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that growth is sink-limited[@delpierre_wood_2016;@refs].
-
-<!--on the fate of extra C-->
-Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in $ANPP_{woody}$[@jiang_fate_2020;@refs].
+<!--discussion of extra C-->
+Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year (`r EDF_tree_rings_TMAX_extended`).
+It remains theoretically possible that warm springs could augment $ANPP_{woody}$, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
+Extra C fixed in years with warm springs could potentially be allocated to formation of more C-dense wood, either through enhanced cell wall thickening (a process that lags behind stem expansion[@cuny_woody_2015]) or to a higher ratio of high-density latewood to lower-density earlywood.
+However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
+Thus, a positive effect of warm springs on the total C content of annual rings is unlikely, suggesting that warm springs do not augment $ANPP_{woody}$. 
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.

### 2022-03-18 doc/manuscript/Dow_manuscript_Nature.Rmd
small changes

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2644 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently at 53)-->
+We're currently at 2635 words and 1 modest + 2 large display items.
+Articles typically have 30-50 references (currently at 54)-->
 
@@
 However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
-Thus, a positive effect of warm springs on the total C content of annual rings is unlikely, suggesting that warm springs do not augment $ANPP_{woody}$. 
+Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or $ANPP_{woody}$. 
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
@@
 <!--conclusions-->
-As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier. 
+As spring temperatures become increasingly warmer, growing seasons will start earlier. 
 However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.

### 2022-03-18 doc/manuscript/Dow_manuscript_Nature.Rmd
#125

@@
 We're currently at 2635 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently at 54)-->
+Articles typically have 30-50 references (currently at 53)-->
 
@@
 
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
+Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
 <!--we can potentially cut some of these references, if not needed elsewhere.-->

### 2022-03-19 doc/manuscript/Dow_manuscript_Nature.Rmd
rename gmax (max daily growth rate --> max growth rate)

@@
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
-**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
+**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$; occurs on average within 5 days of $DOY_{50}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum growth rate ($g_{max}$; occurs on average within 5 days of $DOY_{50}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@jiang_fate_2020;@refs].
+<!--rocha_linking_2006 may work here if needed-->
 

### 2022-03-19 doc/manuscript/tables_figures/schematic_summary.Rmd
rename gmax (max daily growth rate --> max growth rate)

@@
   pack_rows("Timing of growth", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Daily growth rate", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Growth rate", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   pack_rows("Annual growth", 6, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%

### 2022-03-19 doc/manuscript/Dow_manuscript_Nature.Rmd
minor changes related to R1 comments

@@
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
-**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
+**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
@@
 Here, we evaluate how spring temperatures affect stem growth timing, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
-To test whether warmer springs extend the period of peak stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
+To test whether warmer springs advance the timing and extend the duration of stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
 To test whether spring temperatures consistently increased annual growth, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
@@
 The time period yielding the lowest $\Delta AICc$ was selected as the CTW.
+To avoid spurious correlations that could occur using temperature data at the daily resolution, we ran this analysis with weekly resolution, using temperatures averaged over weekly time periods.
 Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
@@
 <!--Hypothesis testing -->
-Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig.
-`r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
+Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$, Fig.
+`r schematic_summary`) and CTW $T_{max}$ (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
 Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 

### 2022-03-19 doc/manuscript/Dow_manuscript_Nature.Rmd
add clim summaries
#127

@@
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2021].
-Within the study period for the dendrometer band analysis (2011-2020), ...*[some stats on climate]*
+Within the study period for the dendrometer band analysis (2011-2020), Spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-2010).
+Prior to the study period, the average spring $T_{max}$ was 16.9 $degreeC$ and summer SPEI was -0.06.
+In the study period, average spring $T_{max}$ was 17.6 $degreeC$ and summer SPEI was 0.51.
 
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2021]. 
-Within the study period for the dendrometer band analysis (1999-2003), ...*[some stats on climate]*
+Within the study period for the dendrometer band analysis (1999-2003), Spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
+Prior to the study period, the average spring $T_{max}$ was 10.9 $degreeC$ and summer SPEI was 0.13.
+In the study period, average spring $T_{max}$ was 11.2 $degreeC$ and summer SPEI was 0.20.
 The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2021].

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
addressing R1 comments
mostly fixing references

@camerondow35 , the quilt plot methods paragraph needs some attention from you

@@
 
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022].
+Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2022; @gao_earlier_2022].
 <!--we can potentially cut some of these references, if not needed elsewhere.-->
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth(`r  EDT_phenology_parameters`)[@dorangeville_peak_2021;@tumajer_limitation_2022].
+This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth(`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 <!--Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).-->
-<!--Consistent with the results of Ref[@dorangeville_peak_2021], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
+<!--Consistent with the results of Ref[@dorangeville_peak_2022], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).-->
@@
 It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
-Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
+Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
 Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
@@
 `r tree_rings`, `r EDT_tree_ring_summary`).
-This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress or high temperatures during the peak growing season[@dorangeville_drought_2018;@helcoski_growing_2019;@anderson-teixeira_joint_2021;@babst_twentieth_2019].
+This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress or high temperatures during the peak growing season[@dorangeville_drought_2018;@helcoski_growing_2019;@anderson-teixeira_joint_2022;@babst_twentieth_2019].
 <!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of $T_{max}$ in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
@@
 
-Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015;@davies_forestgeo_2021].
+Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2022], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015;@davies_forestgeo_2021].
 SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km south of Front Royal, Virginia. 
@@
 Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera* L.), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
-The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2021].
+The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
 Within the study period for the dendrometer band analysis (2011-2020), Spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-2010).
@@
 Dominant species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis* (L.) Carrière), oak (*Quercus spp.*) and red maple (*Acer rubrum* L.). 
-The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2021]. 
+The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2022]. 
 Within the study period for the dendrometer band analysis (1999-2003), Spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
@@
 In the study period, average spring $T_{max}$ was 11.2 $degreeC$ and summer SPEI was 0.20.
-The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2021].
+The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
 
@@
 We ran separate models for each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
-This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@stan_development_team_stan_2019].
+This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@standevelopmentteam_stan_2019].
 In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and RWI were assessed using *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
-We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
+<!-- this sentence needs some work-->Correlations between monthly climate and RWI were assessed in R **v 4.0**[@rcoreteam_language_2020]<!--check reference year--> using the packages *dplR*[@bunn_dendrochronology_2008] <!-- did we actually use this package?--> and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
+`r tree_rings`, Extended Data Figures 6-7) following the methods of Ref.[@biondi_dendroclim2002_2004].
+Our analysis focused on assessing correlations of RWI to months spanning from January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
 To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
minor edits related to R2 comments

@@
 Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, $ANPP_{woody}$ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
-Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018].
-Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be short lived.
+Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
+Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.
 
@@
 
-The raw dendrometer band data were manually inspected before analysis. 
-We screened the data for three classes of errors. 
+The raw dendrometer band data were screened before analysis to remove records or entire tree-years that were either unreliable or inappropriate for our analysis. 
+First, we manually screened the data for three classes of errors. 
 First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and removed.
@@
 After fitting the growth model, we removed tree-years with poor fits.
-Models were judged to be poorly fit if modeled growth parameters were outliers, which were commonly indicative of unrealistic fits (e.g., growth occurring outside the growing season or over a very short period) and underlain by very slow tree growth or poor data records.
+Models were judged to be poorly fit if modeled growth parameters were outliers, which were commonly indicative of unrealistic fits (e.g., growth occurring outside the growing season or over a very short period) and underlain by very slow tree growth or poor data records that passed the initial screening (described above).
 Modeled fits for tree-years were removed under two conditions:

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
minor changes related to R4 comments

@@
 <!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
-Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season, when increased atmospheric demand associated with high temperatures limits growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
+Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season of the current or sometimes previous year (Fig.`r tree_rings`, `r EDF_tree_rings_TMAX_extended`), when increased atmospheric demand associated with high temperatures can limit both leaf-level gas exchange and stem growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
 Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig.

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
acknowledge reviewers
Update Dow_manuscript_Nature.Rmd

@@
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
-Within the study period for the dendrometer band analysis (2011-2020), Spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-2010).
+Within the study period for the dendrometer band analysis (2011-2020), spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-2010).
 Prior to the study period, the average spring $T_{max}$ was 16.9 $degreeC$ and summer SPEI was -0.06.
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2022]. 
-Within the study period for the dendrometer band analysis (1999-2003), Spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
+Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
 Prior to the study period, the average spring $T_{max}$ was 10.9 $degreeC$ and summer SPEI was 0.13.
@@
 We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
-Thanks to David A. Orwig and members of the ForestGEO Ecosystems & Climate Lab at SCBI for helpful feedback.
+Thanks to David A. Orwig, members of the ForestGEO Ecosystems & Climate Lab at SCBI, and four anonymous reviewers for helpful feedback.
 The research was funded by ForestGEO (Smithsonian). 

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
methods cleanup + issue #127

@@
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
-Within the study period for the dendrometer band analysis (2011-2020), spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-2010).
-Prior to the study period, the average spring $T_{max}$ was 16.9 $degreeC$ and summer SPEI was -0.06.
-In the study period, average spring $T_{max}$ was 17.6 $degreeC$ and summer SPEI was 0.51.
+Within the study period for the dendrometer band analysis (2011-2020), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average prior to the study period (1970-2010).
+Prior to the study period, the average spring $T_{max}$ was 16.9 $degreeC$ and summer SPEI was -0.06. In the study period, average spring $T_{max}$ was 17.6 $degreeC$ and summer SPEI was 0.51.
 
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2022]. 
-Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March + April $T_{max}$) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
-Prior to the study period, the average spring $T_{max}$ was 10.9 $degreeC$ and summer SPEI was 0.13.
-In the study period, average spring $T_{max}$ was 11.2 $degreeC$ and summer SPEI was 0.20.
+Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
+Prior to the study period, the average spring $T_{max}$ was 10.9 °C and summer SPEI was 0.13. In the study period, average spring $T_{max}$ was 11.2 °C and summer SPEI was 0.20.
 The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
@@
 Measurements were continued through to fall leaf senescence, with the median end date being October 17 and the latest end date November 26 (2012). 
-*Timing of measurements at Harvard Forest were similar, with the median start date of April 23 and median end date of October 30.* <!--Changes depending on when we calculate median.
-I think we can calculate it before removal of 1998 since we technically incude 1998 in our raw data, just remove it during our data cleaning.
-If you don't like that, it cahges to: start: april 21, end: October 27-->
-All measurements from 1998 were dropped because of a late start date (May 26).
-*1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15.* <!--I think we dropped this.Correct?
-CD: No, left in.
-This could influence growth in 1999 if "a" was selected based on pre-hydration, winter DBH, but I have faith that Sean's process to find "a" can handle this.
-That being said, it is unfortunate that 1999 is the only year with early measurements and very weird timing.
-Justification for my faith: If "a" uses prehydration, winter dbh, then we would expect the "a" in 1999 to be proportionally smaller than in other years.
-This smaller "a" coupled with an unchanged "b" (last measurement was about the same in 1999 as other years) would result in greater total growth (b-a).
-But we actually see that 1999 total growth is the smallest, indicating to me that the hot summer is driving the weirdness.
--->
+At Harvard Forest, all measurements from 1998 were dropped because of a late start date (May 26).
+Among the remaining years,the median start date was April 21 and median end date of October 27.
+1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15.
 The latest end date was November 11, 2002.
@@
 For each site, we used records of daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
+Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values were obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020;@begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
 
@@
 
-To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010].
-The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020;@begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
+To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer SPEI indeces. 
 Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures were associated with greater summer drought stress in our data set.

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
#127

@@
 Within the study period for the dendrometer band analysis (2011-2020), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average prior to the study period (1970-2010).
-Prior to the study period, the average spring $T_{max}$ was 16.9 $degreeC$ and summer SPEI was -0.06. In the study period, average spring $T_{max}$ was 17.6 $degreeC$ and summer SPEI was 0.51.
+Specifically, average spring $T_{max}$ was 16.9°C prior to the study period (range: ) and 17.6°C (range: ) during the study period, while summer SPEI was -0.06 (range: ) prior to the study period and 0.51 (range: ) during.
 
@@
 Dominant species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis* (L.) Carrière), oak (*Quercus spp.*) and red maple (*Acer rubrum* L.). 
-The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5° C in January and 22° C in July[@anderson-teixeira_joint_2022]. 
+The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5°C in January and 22°C in July[@anderson-teixeira_joint_2022]. 
 Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
-Prior to the study period, the average spring $T_{max}$ was 10.9 °C and summer SPEI was 0.13. In the study period, average spring $T_{max}$ was 11.2 °C and summer SPEI was 0.20.
+Specifically, average spring $T_{max}$ was 10.9°C prior to the study period (range: ) and 11.2°C (range: ) during the study period, while summer SPEI was 0.13 (range: ) prior to the study period and 0.20 (range: ) during.
 The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].

### 2022-03-21 RScripts/climate_summaries.R
Create climate_summaries.R
#127 @teixeirak here's the rscript I've been using for this issue. I can get to this tonight if you'd prefer I did it. This script is not annotated and i'm not sure we should include it in the final version of this repo.

@@ -1,0 +1,173 @@
+#Compare our study period to the past 30 years. SPEI, Summer temps, Spring temps
+library(tidyverse)
+library(reshape2)
+library(lubridate)
+tmp <- read_csv("Data/climate data/CRU/tmx.1901.2019-all_sites-3-11.csv")
+scbi_tmp <- tmp[tmp$sites.sitename %in% "SCBI",]
+scbi_tmp <- melt(scbi_tmp, id.vars = "sites.sitename")
+scbi_tmp$variable <- substr(scbi_tmp$variable, 2, nchar(as.character(scbi_tmp$variable)) )
+
+scbi_2020 <- met_tower_data_sensor2_ncdc_supplemented <- read_csv("Data/climate data/SCBI/met_tower_data_sensor2_ncdc_supplemented.csv")
+scbi_2020 <- scbi_2020[scbi_2020$year.x %in% 2020,]
+monthly_2020 <- aggregate(scbi_2020[!is.na(scbi_2020$cleantmax),]$cleantmax, by = list(scbi_2020[!is.na(scbi_2020$cleantmax),]$month), FUN = mean)
+
+scbi_tmp$variable <- as.Date(scbi_tmp$variable, format = "%Y.%m.%d")
+scbi_tmp$year <- year(scbi_tmp$variable)
+scbi_tmp$month <- month(scbi_tmp$variable)
+
+scbi_tmp <- scbi_tmp[scbi_tmp$year %in% c(1970:2019),]
+scbi_tmp <- scbi_tmp[!is.na(scbi_tmp$value),]
+scbi_tmp <- scbi_tmp[scbi_tmp$month %in% c(3,4),]
+scbi_tmp <- scbi_tmp[,c(3,4,5)]
+
+monthly_2020 <- monthly_2020[monthly_2020$Group.1 %in% c(3,4),]
+clim_2020_3 <- data.frame(monthly_2020[1,2], 2020, 3 )
+names(clim_2020_3) <- names(scbi_tmp)
+clim_2020_4 <- data.frame(monthly_2020[2,2], 2020, 4 )
+names(clim_2020_4) <- names(scbi_tmp)
+
+clim_2020 <- rbind(clim_2020_3,clim_2020_4)
+scbi_tmp <- rbind(scbi_tmp, clim_2020)
+mean(scbi_tmp[scbi_tmp$year %in% c(1970:2010),]$value)
[…]

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
add spaces to report SDs #127

@@
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
-Within the study period for the dendrometer band analysis (2011-2020), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average prior to the study period (1970-2010).
-Specifically, average spring $T_{max}$ was 16.9°C prior to the study period (range: ) and 17.6°C (range: ) during the study period, while summer SPEI was -0.06 (range: ) prior to the study period and 0.51 (range: ) during.
+Within the study period for the dendrometer band analysis (2011-2020), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
+Specifically, average spring $T_{max}$ was 16.9$\pm$**##**°C (mean $\pm$ 1SD) prior to the study period (range: **##-##**) and 17.6$\pm$**##**°C (range: **##-##**) during the study period, while summer SPEI was -0.06$\pm$**##** (range: **##-##**) prior to the study period and 0.51$\pm$**##** (range: **##-##**) during.
 
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5°C in January and 22°C in July[@anderson-teixeira_joint_2022]. 
-Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer SPEI values (4-month value of August) were similar to the average prior to the study period (1970-1998).
-Specifically, average spring $T_{max}$ was 10.9°C prior to the study period (range: ) and 11.2°C (range: ) during the study period, while summer SPEI was 0.13 (range: ) prior to the study period and 0.20 (range: ) during.
+Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
+Specifically, average spring $T_{max}$ was 10.9$\pm$**##**°C prior to the study period (range: **##-##**) and 11.2$\pm$**##**°C (range: **##-##**) during the study period, while summer SPEI was 0.13$\pm$**##** (range: **##-##**) prior to the study period and 0.20$\pm$**##** (range: **##-##**) during.
 The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
+<!--this still needs to be put in context-->
 

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
#128

@@
 Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW $T_{max}$ for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with CTW $T_{max}$ at SCBI and a weak negative relationship (-0.2-0.03 mm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
+$\Delta DBH$ displayed a weak positive relationship (+0.0029-0.008 cm/ $^\circ$C) with CTW $T_{max}$ at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
 
@@
 If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
-We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253), and tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
+We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253), and tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
 These were removed because they reduce the reliability of predicted growth in the modeled curves.

### 2022-03-21 doc/manuscript/Dow_manuscript_Nature.Rmd
tiny change
Update Dow_manuscript_Nature.Rmd
#127

@@
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
-Within the study period for the dendrometer band analysis (2011-2020), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
-Specifically, average spring $T_{max}$ was 16.9$\pm$**##**°C (mean $\pm$ 1SD) prior to the study period (range: **##-##**) and 17.6$\pm$**##**°C (range: **##-##**) during the study period, while summer SPEI was -0.06$\pm$**##** (range: **##-##**) prior to the study period and 0.51$\pm$**##** (range: **##-##**) during.
+Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April $T_{max}$; source: CRU, as described below) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
+Specifically, average spring $T_{max}$ was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4 ) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.
 
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5°C in January and 22°C in July[@anderson-teixeira_joint_2022]. 
-Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: **local weather stations or CRU**, as described below) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
-Specifically, average spring $T_{max}$ was 10.9$\pm$**##**°C prior to the study period (range: **##-##**) and 11.2$\pm$**##**°C (range: **##-##**) during the study period, while summer SPEI was 0.13$\pm$**##** (range: **##-##**) prior to the study period and 0.20$\pm$**##** (range: **##-##**) during.
-The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
+Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: CRU, as described below) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
+Specifically, average spring $T_{max}$ was 10.9$\pm$ 1.5°C prior to the study period (range: 8.0-13.2) and 11.2$\pm$ 1.0°C (range: 10.1-12.2) during the study period, while summer SPEI was 0.1$\pm$ 0.9 (range: -1.8-1.7) prior to the study period and 0.2$\pm$ 0.9 (range: -1.0-1.1) during.
+The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022] and a SPEI value of -1.0, which is the 5th lowest in the period 1970-2003.
 <!--this still needs to be put in context-->

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
#127
@camerondow35 , FYI. I copied these updates into the google doc

@@
 
-William J. McShea^1^
+William J. McShea^1^ (Orchid ID: 0000-0002-8102-02000)
 
@@
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
-Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April $T_{max}$; source: CRU, as described below) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
-Specifically, average spring $T_{max}$ was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4 ) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.
+Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April $T_{max}$; source: CRU v.4.04[@harris_version_2020]) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
+Specifically, average spring $T_{max}$ was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.
 
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5°C in January and 22°C in July[@anderson-teixeira_joint_2022]. 
-Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: CRU, as described below) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
+Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: CRU v.4.04[@harris_version_2020]) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
 Specifically, average spring $T_{max}$ was 10.9$\pm$ 1.5°C prior to the study period (range: 8.0-13.2) and 11.2$\pm$ 1.0°C (range: 10.1-12.2) during the study period, while summer SPEI was 0.1$\pm$ 0.9 (range: -1.8-1.7) prior to the study period and 0.2$\pm$ 0.9 (range: -1.0-1.1) during.
-The summer of 1999 was abnormally dry, with precipitation of only 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022] and a SPEI value of -1.0, which is the 5th lowest in the period 1970-2003.
-<!--this still needs to be put in context-->
+The driest summer during the study period (1999) had the 5th lowest SPEI value (-1.0) in the period 1970-2003, with precipitation of 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
 

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
replace latex symbols with non-latex
@camerondow35 , FYI, this allows copy/paste into google docs without losing the symbols, plus works better when the journal prints it.

@@
 However, model representation of C allocation to stem growth is overly simplistic and does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
-As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014].
-Understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
+As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014].
+Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
 
@@
 <!--we can potentially cut some of these references, if not needed elsewhere.-->
-In contrast, we have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
+In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
 Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum growth rate ($g_{max}$; occurs on average within 5 days of $DOY_{50}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*; occurs on average within 5 days of *DOY~50~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over $DOY_{25}$; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%$^\circ$C^-1^), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%$^\circ$C^-1^), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 
 <!--Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).-->
-<!--Consistent with the results of Ref[@dorangeville_peak_2022], ring porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
+<!--Consistent with the results of Ref[@dorangeville_peak_2022], ring porous species began growing earlier, reaching the *DOY~25~* benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average *L~pgs~* 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
 `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).-->
-<!--Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only six days longer at SCBI for ring porous species and less than one day longer for diffuse porous species (`r  EDT_phenology_parameters`).-->
+<!--Peak growing season length was similar across sites, with *L~pgs~* being, on average, only six days longer at SCBI for ring porous species and less than one day longer for diffuse porous species (`r  EDT_phenology_parameters`).-->
 
 Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
-We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW).
-CTW was identified by measuring the correlation between weekly $T_{max}$ and $DOY_{25}$ for all combinations of consecutive weeks from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
-The CTW was defined as the weeks for which this correlation was strongest, and mean $T_{max}$ during this period (CTW $T_{max}$) used as our independent variable.
+We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW).
+CTW was identified by measuring the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
+The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.
 
@@
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
-Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
-Mean maximum temperature ($T_{max}$) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to the CTW $T_{max}$ of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }
+Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
+Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }
 
-For ring- and diffuse- porous species at both sites, warmer CTW $T_{max}$ resulted in earlier achievement of seasonal milestones.
+For ring- and diffuse- porous species at both sites, warmer CTW *T~max~* resulted in earlier achievement of seasonal milestones.
 Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig.
 `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
-Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with CTW $T_{max}$ (Figs.
+Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
-Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).
+Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).
 
-Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the CTW $T_{max}$ of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig.
-`r DOY_timing`a-b), there was no consistent lengthening of $L_{PGS}$ (Fig.
+Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig.
+`r DOY_timing`a-b), there was no consistent lengthening of *L~pgs~* (Fig.
 `r schematic_summary`, `r EDF_pheno_all`).<!--Note to Cam: double check this-->
 
-In contrast to the pronounced effects of CTW $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW $T_{max}$ for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.0029-0.008 cm/ $^\circ$C) with CTW $T_{max}$ at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
+In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were weak and inconsistent (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
+Specifically, *g~max~*, which occurred very close to *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
+$\Delta DBH$ displayed a weak positive relationship (+0.0029-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
 
@@
 `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
-In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April $T_{max}$.
-Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
-There were even fewer significant positive correlations with March and May $T_{max}$: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
-In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
-$T_{min}$ generally exhibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).
+In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
+Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
+There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
+In contrast, RWI was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
+*T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and RWI (`r EDF_tree_rings_TMIN`).
 
-To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and $T_{max}$ in the previous year (`r EDF_tree_rings_TMAX_extended`).
-This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of RWI to previous March or April $T_{max}$ for 5/142 ring porous chronologies and to previous April or May $T_{max}$ for 7/66 diffuse porous chronologies.
+To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and *T~max~* in the previous year (`r EDF_tree_rings_TMAX_extended`).
+This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of RWI to previous March or April *T~max~* for 5/142 ring porous chronologies and to previous April or May *T~max~* for 7/66 diffuse porous chronologies.
 
-To test whether the negative effect of summer temperatures might offset an enhancement of growth by warmer spring temperatures, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
+To test whether the negative effect of summer temperatures might offset an enhancement of growth by warmer spring temperatures, we tested for the joint effects of April and June-July *T~max~* on RWI.
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 4% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (at p = 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
 
-![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
-Colors indicate the correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
-Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$.
+![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.
 Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
-Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
+Species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
 
@@
 `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
+However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
 `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig.
@@
 Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year (`r EDF_tree_rings_TMAX_extended`).
-It remains theoretically possible that warm springs could augment $ANPP_{woody}$, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
+It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
 Extra C fixed in years with warm springs could potentially be allocated to formation of more C-dense wood, either through enhanced cell wall thickening (a process that lags behind stem expansion[@cuny_woody_2015]) or to a higher ratio of high-density latewood to lower-density earlywood.
 However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
-Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or $ANPP_{woody}$. 
+Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or ANPP~woody~. 
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
-Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, $ANPP_{woody}$ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
+Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, ANPP~woody~ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
 Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
@@
 <!--.short vs long-term effects-->
-It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
+It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of ANPP~woody~.
 Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
 Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
-If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
+If extra photosynthate made available through a growing difference between GPP and ANPP~woody~ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase ANPP~woody~ through indirect mechanisms.
 Understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.
@@
 Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season of the current or sometimes previous year (Fig.`r tree_rings`, `r EDF_tree_rings_TMAX_extended`), when increased atmospheric demand associated with high temperatures can limit both leaf-level gas exchange and stem growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
-Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig.
+Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to *T~max~* (Fig.
 `r tree_rings`, `r EDT_tree_ring_summary`).
 This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress or high temperatures during the peak growing season[@dorangeville_drought_2018;@helcoski_growing_2019;@anderson-teixeira_joint_2022;@babst_twentieth_2019].
-<!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of $T_{max}$ in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
+<!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of *T~max~* in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
 Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
@@
 The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
-Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April $T_{max}$; source: CRU v.4.04[@harris_version_2020]) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
-Specifically, average spring $T_{max}$ was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.
+Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April *T~max~*; source: CRU v.4.04[@harris_version_2020]) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
+Specifically, average spring *T~max~* was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.
 
@@
 The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5°C in January and 22°C in July[@anderson-teixeira_joint_2022]. 
-Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April $T_{max}$; source: CRU v.4.04[@harris_version_2020]) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
-Specifically, average spring $T_{max}$ was 10.9$\pm$ 1.5°C prior to the study period (range: 8.0-13.2) and 11.2$\pm$ 1.0°C (range: 10.1-12.2) during the study period, while summer SPEI was 0.1$\pm$ 0.9 (range: -1.8-1.7) prior to the study period and 0.2$\pm$ 0.9 (range: -1.0-1.1) during.
+Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April *T~max~*; source: CRU v.4.04[@harris_version_2020]) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
+Specifically, average spring *T~max~* was 10.9$\pm$ 1.5°C prior to the study period (range: 8.0-13.2) and 11.2$\pm$ 1.0°C (range: 10.1-12.2) during the study period, while summer SPEI was 0.1$\pm$ 0.9 (range: -1.8-1.7) prior to the study period and 0.2$\pm$ 0.9 (range: -1.0-1.1) during.
 The driest summer during the study period (1999) had the 5th lowest SPEI value (-1.0) in the period 1970-2003, with precipitation of 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
@@
 Here, $L$ and $K$ are lower and upper asymptotes of the model. 
-$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
+*DOY~ip~* is the day of year where the inflection point in growth rate occurs,
 $r$ shapes the slope of the curve at the inflection point,
@@
 This allows an asymmetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth. 
-When $\theta = 1$, $g_{max}$ occurs on $DOY_{ip}$.
+When $\theta = 1$, *g~max~* occurs on *DOY~ip~*.
 The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. 
@@
 Modeled fits for tree-years were removed under two conditions:
-(1) $g_{max}$ was $\ge 2.5$ standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);
-(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
+(1) *g~max~* was $\ge 2.5$ standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);
+(2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
 At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
@@
 Daily temperature records for Harvard Forest, which had already been gap-filled based on other local records, were obtained from the Harvard Forest weather station[@boose_shaler_2021;@boose_fisher_2021].
-For each site, we used records of daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).
+For each site, we used records of daily maximum (*T~max~*) and minimum temperatures (*T~min~*).
 Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values were obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020;@begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
@@
 The critical temperature window (CTW, Fig.
-`r schematic_summary`), defined as the period over which $T_{max}$ was most strongly correlated with $DOY_{25}$, was determined using the R package *climwin*[@vandepol_identifying_2016].
+`r schematic_summary`), defined as the period over which *T~max~* was most strongly correlated with *DOY~25~*, was determined using the R package *climwin*[@vandepol_identifying_2016].
 This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame. 
 It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
-Here, we tested for correlation between temperature predictor variables ($T_{max}$, $T_{min}$) and biological outcome variable $DOY_{25}$ over the time-frame from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`).
+Here, we tested for correlation between temperature predictor variables (*T~max~*, *T~min~*) and biological outcome variable *DOY~25~* over the time-frame from January 1 to the mean *DOY~25~* for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`).
 The time period yielding the lowest $\Delta AICc$ was selected as the CTW.
 To avoid spurious correlations that could occur using temperature data at the daily resolution, we ran this analysis with weekly resolution, using temperatures averaged over weekly time periods.
-Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
-We defined CTW for $DOY_{25}$, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable.
+Because *T~max~* proved to have a generally stronger influence over *DOY~25~* and other growth parameters, we focused on this variable in our ultimate model, as opposed to *T~min~*.
+We defined CTW for *DOY~25~*, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable.
 
@@
 <!--Hypothesis testing -->
-Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$, Fig.
-`r schematic_summary`) and CTW $T_{max}$ (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
-Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
-Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree.
-We ran separate models for each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
+Correlation between the dendrometer band-derived growth parameters (*DOY~25~*, *DOY~50~*, *DOY~75~*, *L~pgs~*, *g~max~*, and $\Delta DBH$, Fig.
+`r schematic_summary`) and CTW *T~max~* (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
+Analyses were run for both *T~max~* and *T~min~*, with qualitatively similar results, but we present only results for *T~max~*, which had overall stronger correlation with growth parameters. 
+Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean *T~max~* (or *T~min~*) during the CTW, along with random effects of species and of individual tree.
+We ran separate models for each site, and for the response of all growth parameters to *T~max~* (or *T~min~*). 
 This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@standevelopmentteam_stan_2019].
@@
 To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer SPEI indeces. 
-Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures were associated with greater summer drought stress in our data set.
+Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April *T~max~* to determine if warm spring temperatures were associated with greater summer drought stress in our data set.
 No significant correlations were found (all p>0.05).
@@
 <!--QUILT PLOT METHODS-->
-For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
+For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
 <!-- this sentence needs some work-->Correlations between monthly climate and RWI were assessed in R **v 4.0**[@rcoreteam_language_2020]<!--check reference year--> using the packages *dplR*[@bunn_dendrochronology_2008] <!-- did we actually use this package?--> and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
@@
 <!--MULTIVARIATE ANALYSIS METHODS-->
-We used a multivariate model to test for joint effects of April and summer $T_{max}$ on RWI.
+We used a multivariate model to test for joint effects of April and summer *T~max~* on RWI.
 We focused on April to represent spring temperatures because it was the month with greatest overall alignment with the CTW's identified in the dendrometer band analysis and had the highest rate of positive correlations with RWI (`r EDT_tree_ring_summary`).
-We began by testing univariate correlations of $T_{max}$ over three summer windows: June, June-July, and May-August. 
-Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using the base lm() function in R.
+We began by testing univariate correlations of *T~max~* over three summer windows: June, June-July, and May-August. 
+Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April *T~max~* and June-July *T~max~* on RWI for each chronology independently using the base lm() function in R.
 Slopes and p-values for each chronology are given in `r chron_table`. 

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
work on refs
#125 #129

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;-->@menzel_european_2006;<!--@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;-->@menzel_european_2006;<!--@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[@cook_predicting_1991;@babst_twentieth_2019;@anderson-teixeira_joint_2022; @gao_earlier_2022].
-<!--we can potentially cut some of these references, if not needed elsewhere.-->
+Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022].
 In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
@@
 Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
-This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@jiang_fate_2020;@refs].
+This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@jiang_fate_2020;@refs].
 <!--rocha_linking_2006 may work here if needed-->
@@
 It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of ANPP~woody~.
-Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
+Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], ANPP~woody~[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
 Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
eliminate ref placeholders
#129

@@
 Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
-This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@jiang_fate_2020;@refs].
+This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].
 <!--rocha_linking_2006 may work here if needed-->

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
#125

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;-->@menzel_european_2006;<!--@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;-->@menzel_european_2006;<!--@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
reduced refs down to 50!
#125 #129

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.-->
-The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020;@fatichi_moving_2014].
+The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@ahlstrom_robustness_2012;@friedlingstein_global_2020;@fatichi_moving_2014].
 
@@
 In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;<!--@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
@@
 The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020].
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and NEE[@keenan_net_2014;@finzi_carbon_2020].
 However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
update word & citation count
#125
chip away at word count
#125

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2635 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently at 53)-->
+We're currently at 2672 words and 1 modest + 2 large display items.
+Articles typically have 30-50 references (currently at 50)-->
 
@@
 Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
-However, model representation of C allocation to stem growth is overly simplistic and does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
+However, model representation of C allocation to stem growth does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
 As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014].
@@
 In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;<!--@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
 Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
-Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
+Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
 These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*; occurs on average within 5 days of *DOY~50~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig.
-`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
+`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
 Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs.
@@
 `r DOY_timing`a-b), there was no consistent lengthening of *L~pgs~* (Fig.
-`r schematic_summary`, `r EDF_pheno_all`).<!--Note to Cam: double check this-->
+`r schematic_summary`, `r EDF_pheno_all`).
 
-In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were weak and inconsistent (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
-Specifically, *g~max~*, which occurred very close to *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
+In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were usually weak and inconsistent (Fig.
+`r schematic_summary`, `r EDF_pheno_all`).
+Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
 $\Delta DBH$ displayed a weak positive relationship (+0.0029-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
@@
 <!--discussion of extra C-->
-Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year (`r EDF_tree_rings_TMAX_extended`).
+Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
 It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].

### 2022-03-23 doc/manuscript/Dow_manuscript_Nature.Rmd
update word count
#130

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2672 words and 1 modest + 2 large display items.
+We're currently at 2652 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently at 50)-->
@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
+Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons contribute positively to annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
 However, model representation of C allocation to stem growth does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
@@
 However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
-1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+`r schematic_summary`) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback *that is anticipated in most climate forecasting models*[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig.

### 2022-03-23 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
#129

--- 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-<!-- this sentence needs some work-->Correlations between monthly climate and RWI were assessed in R **v 4.0**[@rcoreteam_language_2020]<!--check reference year--> using the packages *dplR*[@bunn_dendrochronology_2008] <!-- did we actually use this package?--> and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
+<!-- this sentence needs some work.
+CD: maybe Valentine can help with this-->Correlations between monthly climate and RWI were assessed in R **v 4.0**[@rcoreteam_language_2020]<!--check reference year.
+CD: this is the reference year that the function "citation()" give me, so it should be correct.--> using the packages *dplR*[@bunn_dendrochronology_2008] <!-- did we actually use this package?
+CD: Yes, at the very least we used modified functions from this package.--> and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
 `r tree_rings`, Extended Data Figures 6-7) following the methods of Ref.[@biondi_dendroclim2002_2004].

### 2022-03-23 doc/manuscript/Dow_manuscript_Nature.Rmd
follow up on quilt plot methods
thanks, @camerondow35.

@ValentineHerr , could you please check the sentence and paragraph modified in this commit? In particular, whether the info on R packages used is correct.

--- 2022-03-23 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-<!-- this sentence needs some work-->Correlations between monthly climate and RWI were assessed in R **v 4.0**[@rcoreteam_language_2020]<!--check reference year--> using the packages *dplR*[@bunn_dendrochronology_2008] <!-- did we actually use this package?--> and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
+Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
 `r tree_rings`, Extended Data Figures 6-7) following the methods of Ref.[@biondi_dendroclim2002_2004].

### 2022-03-26 doc/manuscript/Dow_manuscript_Nature.Rmd
revise intro paragraph on modeling
#130

@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons contribute positively to annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
-However, model representation of C allocation to stem growth does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
-As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014].
+Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012].
+However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018;@buermann_widespread_2018].
+Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
+As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
 Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
@@
 Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
-However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
+However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
 Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
@@
 However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
-`r schematic_summary`) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback *that is anticipated in most climate forecasting models*[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
+`r schematic_summary`) is unlikely to provide a sustained augmentation of CO~2~ sequestration in woody biomass and ensuant negative climate change feedback *that is anticipated in most climate forecasting models*[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
 Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig.
-`r tree_rings`), which constitutes a positive feedback to climate change.
+`r tree_rings`)[@ahlstrom_robustness_2012], which constitutes a positive feedback to climate change.
 

### 2022-03-26 doc/manuscript/Dow_manuscript_Nature.Rmd
improved content on modeling
#130

@@
 **These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
-**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**
+**Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**
 
@@
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
-Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
+Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021].
 Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
@@
 However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
-`r schematic_summary`) is unlikely to provide a sustained augmentation of CO~2~ sequestration in woody biomass and ensuant negative climate change feedback *that is anticipated in most climate forecasting models*[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
-Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig.
+`r schematic_summary`) is unlikely to provide a sustained augmentation of CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@richardson_terrestrial_2012;@keenan_net_2014;@buermann_widespread_2018].
+Rather, the dominant effect of rising temperatures on temperate deciduous forest woody productivity will be a negative effect of high summer temperatures (Fig.
 `r tree_rings`)[@ahlstrom_robustness_2012], which constitutes a positive feedback to climate change.

### 2022-03-26 doc/manuscript/Dow_Extended_Data.Rmd
Quercus mongolica --> Quercus montana
#131

@@
     column_spec(3:18, width = ".4cm") %>%
-pack_rows("Ring-porous", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-pack_rows("Diffuse-porous", 17, 26, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-row_spec(27, bold = T) %>%
+pack_rows("Ring-porous", 1, 15, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+pack_rows("Diffuse-porous", 16, 25, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+row_spec(26, bold = T) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2022-03-28 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012].
-However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018;@buermann_widespread_2018].
+Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
+However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
 Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
@@
 Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, ANPP~woody~ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
-Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
+Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
 Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.

### 2022-03-28 doc/manuscript/Dow_manuscript_Nature.Rmd
various small changes from google doc

@@
 
-Cameron Dow^1,2^ ( Orcid ID : 0000-0002-8365-598X)
+Cameron Dow^1,2^ (Orcid ID : 0000-0002-8365-598X)
 
-Albert Y. Kim^1,3^ ( Orcid ID : 0000-0001-7824-306X)
+Albert Y. Kim^1,3^ (Orcid ID : 0000-0001-7824-306X)
 
-Loïc D’Orangeville^4,5^ ( Orcid ID : 0000-0001-7841-7082)
+Loïc D’Orangeville^4,5^ (Orcid ID : 0000-0001-7841-7082)
 
-Erika B. Gonzalez-Akre^1^ ( Orcid ID : 0000-0001-8305-6672)
+Erika B. Gonzalez-Akre^1^ (Orcid ID : 0000-0001-8305-6672)
 
-Ryan Helcoski^1^
+Ryan Helcoski^1^ (Orcid ID: 0000-0003-3579-0121)
 
-Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)
+Valentine Herrmann^1^ (Orcid ID : 0000-0002-4519-481X)
 
-Grant L. Harley^6^ ( Orcid ID : 0000-0003-1557-8465)
+Grant L. Harley^6^ (Orcid ID : 0000-0003-1557-8465)
 
@@
 
-Alan J. Tepley^1,10^ ( Orcid ID : 0000-0002-5701-9613)
+Alan J. Tepley^1,10^ (Orcid ID : 0000-0002-5701-9613)
 
@@
 `r schematic_summary`).
-This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth(`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022].
+This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
-Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
+Across all chronologies, the standardized ring-width index (RWI) was significantly (p $\le$ 0.05) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
 There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
-In contrast, RWI was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
+In contrast, RWI was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring- and diffuse-porous species, respectively.
 *T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and RWI (`r EDF_tree_rings_TMIN`).
@@
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (at p = 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (p $\le$0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
 
@@
 Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or ANPP~woody~.
+
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
@@
 If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
-We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253), and tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
+We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253), and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
 These were removed because they reduce the reliability of predicted growth in the modeled curves.
@@
 MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019] via Google Earth Engine.
-For each year at each site, we extracted data from the pixel containing the center of each forest plot (resolution of 500m). 
+For each year at each site, we extracted data from the pixel (500m resolution) containing the center of each forest plot. 
 Using the daily MODIS 2-band Enhanced Vegetation Index data (EVI2), the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, peak, and senescence as used in this study.
@@
 
-To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer SPEI indeces. 
+To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer SPEI indices. 
 Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April *T~max~* to determine if warm spring temperatures were associated with greater summer drought stress in our data set.
@@
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018].
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018].
 Again, records were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).
@@
 A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
-An adaptive power transformation, a process that also stabilises the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
+An adaptive power transformation, a process that also stabilizes the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
 Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records[@jones_estimating_1997].
-The 1/3rds spline method was chosen when replication in the inner portion of each chronology (ca. inner 30–50 yr of each record depending on full chronology length) was less than three trees. 
+The 1/3rds spline method was chosen when replication in the inner portion of each chronology (the earliest ca. 30–50 yr of each record depending on full chronology length) was less than three trees. 
 When replication was greater than n = 3 trees, we used the average correlation between raw ring-width series (rbar) method.
-The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
+The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilization[@cook_time_1985].
 We defined chronology start year (`r EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
-`r tree_rings`, Extended Data Figures 6-7) following the methods of Ref.[@biondi_dendroclim2002_2004].
-Our analysis focused on assessing correlations of RWI to months spanning from January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
+Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] to assess correlations and create summary figures (Fig.
+`r tree_rings`, Extended Data Figures 6-7).
+Our analysis focused on assessing correlations of RWI to months spanning January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
 To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).

### 2022-03-28 doc/manuscript/Dow_manuscript_Nature.Rmd
fix minor error

@@
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (p $\le$0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (p $\le$ 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
 

### 2022-03-28 doc/manuscript/Dow_manuscript_Nature.Rmd
update data availability statement
#100

@@
 Master versions of the dendrometer band data are available for SCBI via GitHub in the *Dendrobands* repository of the Smithsonian Conservation Biology Institute ForestGEO plot (https://github.com/SCBI-ForestGEO/Dendrobands), which is archived in Zenodo (DOI 10.5281/zenodo.5551143), and for Harvard Forest via the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF149).
-Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail), downloaded on [DATE].
+Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
 Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF001 AND https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF000).
@@
 MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description).
-Original tree cores are archived at the institutions of various members of the author team and will be made available upon reasonable request. 
-<!-- do we need to add ITRDB??-->
+In addition to being archived in this project's repository, many tree-ring data sets are archived in the International Tree-Ring Data Bank (ITRDB;https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring), the DendroEcological Network (DEN; https://www.uvm.edu/femc/dendro/), or the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive), as detailed in `r chron_table`.
+Original tree cores are archived at the institutions of various members of the author team (Harvard Forest, Smithsonian Conservation Biology Institute, Indiana University, and University of Idaho) and will be made available upon reasonable request.
 

### 2022-03-28 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
https://github.com/EcoClimLab/growth_phenology/commit/b380b8f9e4d7cfd8ae8e86bc3eb0f16b99314230#commitcomment-69721790

@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] to assess correlations and create summary figures (Fig.
+Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
 `r tree_rings`, Extended Data Figures 6-7).

### 2022-03-28 doc/manuscript/Dow_manuscript_Nature.Rmd
update word and citation count
#125
update citations for tree-ring records

@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2652 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently at 50)-->
+We're currently at 2685 words and 1 modest + 2 large display items.
+Articles typically have 30-50 references (currently at 51)-->
 
@@
 To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
-`r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
+`r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`).
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
@@
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in eastern North America[@maxwell_sampling_2020;@dorangeville_drought_2018].
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in eastern North America[@pederson_climatic_2005; @maxwell_sampling_2020;@dorangeville_drought_2018].
 Again, records were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).

### 2022-03-29 doc/manuscript/Dow_manuscript_Nature.Rmd
screenshot of Fig.1, knit in
#132

@@
 
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%$^\circ$C^-1^), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%$^\circ$C^-1^), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 

### 2022-03-29 doc/manuscript/Dow_manuscript_Nature.Rmd
proofreading edits

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
-However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
+However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
 Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].

### 2022-03-29 doc/manuscript/Dow_Extended_Data.Rmd
update Extended Data Table S3
#133

@@
     column_spec(3:18, width = ".4cm") %>%
-pack_rows("Ring-porous", 1, 15, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-pack_rows("Diffuse-porous", 16, 25, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-row_spec(26, bold = T) %>%
+pack_rows("Ring-porous", 1, 17, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+pack_rows("Diffuse-porous", 18, 27, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+row_spec(28, bold = T) %>%
 kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2022-03-29 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
from google doc
minor proof-reading edits

@@
 
-Sean McMahon^9,11^ (Orcid ID : 0000-0001-8302-6908)
+Sean M. McMahon^9,11^ (Orcid ID : 0000-0001-8302-6908)
 
@@
 n_cores_sites <- length(unique(chronology_table[, "Location"]))
-n_cores_species <- length(unique(chronology_table[, "Species"]))
+n_cores_species <- length(unique(chronology_table[, "Species"])) - 1 #the -1 accounts for the Quercus spp. record (combination of white oak group)
 n_chronologies_SCBI <- 12
@@
 
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%$^\circ$C^-1^), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
 
@@
 We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW).
-CTW was identified by measuring the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
+CTW was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
 The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.
@@
 Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig.
-`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
+`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
 Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs.
@@
 Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.0029-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
+$\Delta DBH$ displayed a weak positive relationship (+0.003-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
 
@@
 
-![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
 Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.
-Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
-Species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`).
+Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
 
@@
 Anderson-Teixeira collected or oversaw collection of data;
-Cameron Dow, Albert Y. Kim, Valentine Herrmann, Justin T. Maxwell, Ian R. McGregor, Sean McMahon analyzed the data or provided analytical tools; 
+Cameron Dow, Albert Y. Kim, Valentine Herrmann, Justin T. Maxwell, Ian R. McGregor, Sean M. McMahon analyzed the data or provided analytical tools; 
 Cameron Dow and Kristina J. Anderson-Teixeira led the writing of the manuscript. 

### 2022-03-29 doc/manuscript/Dow_manuscript_Nature.Rmd
add some more info on tree-ring methodology

@@
 We analyzed tree-ring records for `r n_cores_sites` sites, including our focal sites.
-All cores had been previously collected, cross-dated, and measured using standard collection and processing methodologies[@stokes_introduction_1968].
+All cores had been previously collected, cross-dated, and measured using standard collection and processing methodologies[@stokes_introduction_1968;@speer_fundamentals_2010].
 
@@
 
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in eastern North America[@pederson_climatic_2005; @maxwell_sampling_2020;@dorangeville_drought_2018].
-Again, records were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).
+The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in eastern North America[@pederson_climatic_2005;@maxwell_sampling_2020;@dorangeville_drought_2018].
+For the majority of sampled populations (i.e., site-species combinations), sampling focused on canopy trees (typically >20 trees per population)[@pederson_climatic_2005;@maxwell_sampling_2020;@dorangeville_drought_2018], while ~15% of the total 207 chronologies came from plot-level collections where trees above a certain diameter (typically 10 cm DBH) were censused and cored[@dye_comparing_2016;@helcoski_growing_2019].
+Again, analyses were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).
 

### 2022-03-29 doc/manuscript/Dow_manuscript_Nature.Rmd
tiny change

@@
 MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description).
-In addition to being archived in this project's repository, many tree-ring data sets are archived in the International Tree-Ring Data Bank (ITRDB;https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring), the DendroEcological Network (DEN; https://www.uvm.edu/femc/dendro/), or the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive), as detailed in `r chron_table`.
+In addition to being archived in this project's repository, many tree-ring data sets are archived in the International Tree-Ring Data Bank (ITRDB;https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring), the DendroEcological Network (DEN; https://www.uvm.edu/femc/dendro/), and/or the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive), as detailed in `r chron_table`.
 Original tree cores are archived at the institutions of various members of the author team (Harvard Forest, Smithsonian Conservation Biology Institute, Indiana University, and University of Idaho) and will be made available upon reasonable request.

### 2022-03-29 doc/submissions/2022_03_29_Nature/cover letter.Rmd
load cover letter

---
title: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

March 29, 2022
\newline
\newline

Dear Editor:

We are pleased to submit our manuscript "Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity", by Cameron Dow and co-authors, for consideration for publication in *Nature*. The manuscript demonstrates that although warm spring temperatures shift the timing of temperate deciduous tree growth earlier, they do not augment annual growth. The implication is that, contrary to the expectations imbued in most climate models, warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term carbon sink of temperate deciduous forests.

We have received and responded to the comments of four reviewers. The largest changes changes included the following:

- We redid dendrometer band analyses to provide more stable parameter estimates, a refined data cleaning process, and a more conservative approach to discarding outliers. This did not substantively alter the results (although there were some modest changes) or affect the conclusions.

- We reworked the introduction and discussion to incorporate content and citation suggestions of the reviewers, as well as discussion of relevant papers that have been published since our initial submission. This includes discussion of the literature on wood formation and reworking of the discussion on C cycling.

We believe that the revised manuscript is significantly improved and thank the reviewers for their constructive feedback.

We believe our manuscript to be an appropriate length for *Nature*, with 2,687 words, 3 display items, and 51 references in the main article. The article is accompanied by an Extended Data file (3 tables, 7 figures) and a supplementary table.

Thank you for considering our submission for *Nature*.

Sincerely,
\newline
\newline

Kristina Anderson-Teixeira, on behalf of all co-authors.

### 2022-04-23 doc/submissions/2022_04_Nature/Response to reviews.Rmd
create response to reviews doc

---
title: "Response to reviews"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Referees' comments:

## Referee #1 (Remarks to the Author):

GENERAL COMMENTS

The revised manuscript entitled: “Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity” by Cameron Dow, Albert Y. Kim, Loïc D’Orangeville, Erika B. Gonzalez-Akre, Ryan Helcoski, Valentine Herrmann, Grant L. Harley, Justin T. Maxwell, Ian R. McGregor, William J. McShea, Sean McMahon, Neil Pederson, Alan J. Tepley, Kristina J. Anderson-Teixeira, is well written, well-structured and easy to read. The main message is clear, important and timely.

**Thank you.**

A. Summary of the key results
Using dendrometer band measurements from about 500 trees, over 10 years and across two forest plots, plus remote sensing leaf phenology, the authors wanted to show that warmer spring temperatures shifted the stem growth of deciduous trees earlier but had little effect on peak growing season length, maximum growth rates, or annual radial growth. Using tree-ring data from more than 200 chronologies from 100 forest plots across eastern North America, the authors wanted to confirm that, in this region, tree-ring width from angiosperm trees is not sensitive to spring but to summer temperatures. The authors point out that these findings imply that extra CO2 uptake in years with warmer springs is not sequestered in long-lived woody biomass as expected by vegetation models.

**We agree with this summary.**

B. Originality and significance
I like the manuscript and the approach presented here. I found the revised manuscript even more consistent in showing that despite increasing spring temperature resulting in advance leaf phenology and stem growth timings, the length and rate of peak growing season are not changing much and had little effect on wood production. The revised manuscript is also more cautious when discussing the fate of the captured carbon.

**Thank you.**

C. Data & methodology:
I appreciate the effort made by the authors to clarify their text about phenology. For me this point is fine now.

**Okay.**

I have a minor concern about the use of the word “sequestration” now. I believe this word as to be checked all along the manuscript, starting with the first sentence (L36-37). The authors want to show that despite earlier carbon capture, enhanced sequestration in the stem is not insured, so better to be crystal clear on the use of “sequestration” and “capture”. Also, I find strange “CO2 sequestration”, to my point of view, C only is sequestered in trees, not CO2.

*Response needed.*

D. Appropriate use of statistics and treatment of uncertainties
Statistics are generally fine to me.
I just wanted to point out that CTW is fitted to be related to DOY25, therefore it is rather expected to find a significant relationship between CWT Tmax and DOY25, and unfair to compare it with the relationship between CWT Tmax and ∆DBH.
I appreciated the clarifications done concerning the data cleaning process. However, I didn’t find in the manuscript the quantity of data that have been removed at it stages. I believe it is worth mentioning it.

*Response needed.*

E. Conclusions
I found the revised version very coherent and the discussion section true to the result section.

**Thank you.**

F. Suggested improvements
No more suggestions from my side…

**Okay.**

G. References:
I really appreciate the effort made by the authors to complete the reference with the elements provided by the referees and to improve the discussion with the new information.

**Thank you.**

H. Clarity and context
The abstract is excellent in presenting Author’s ideas, which I fully support now.

**Thank you.**

## Referee #2 (Remarks to the Author):

This is an important paper, as noted by a number of the other reviewers. It has been substantially improved in response to the first round review comments (I note in particular the Discussion paragraph beginning L232 on the “fate” of “additional” C taken up in years with warm springs). The presentation is clear and crisp, and the paper is easy to read and follow.

**Thank you.**

I have only one minor suggestion. Despite the assertion that, “in the 2nd sentence of the summary paragraph… it is clear that we are talking about stem growth” I have lingering concerns about what is meant by “growth”. I harp on this because it is something that I often discuss with my students (because they are not always clear exactly what they mean, either!). My take is that growth can be characterized in terms of wood volume (from diameter increment) or mass, and they don’t always mean the same thing or indicate the same amount of “growth”, because of the variation in density that I mentioned in my initial review, as well as changes in diameter related to water relations. I’m asking that the authors state explicitly (in the abstract as well as the main body) that they are focusing on diameter, just so that this is clear for the reader.

*Response needed.*

## Referee #3 (Remarks to the Author):

I feel the reviewers have adequately addressed all of my concerns and clarified the their use of the MODIS phenology data. I therefore recommend that the manuscript be accepted.

**Thank you.**

## Referee #4 (Remarks to the Author):

The authors have very carefully taken all my comments and criticisms and incorporated them into the manuscript. From my point of view, there are no further points of criticism.
The results of this work contribute an important new piece of knowledge to the understanding of deciduous forests and their response to climate change.

With kind Regards
Roman Zweifel

**Thank you.**

### 2022-04-23 doc/manuscript/Dow_manuscript_Nature.Rmd
update word count
update SCBI name

@@
 
-1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA
+1. Conservation Ecology Center; Smithsonian National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA
 
@@
 <!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2685 words and 1 modest + 2 large display items.
+We're currently at 2687 words and 1 modest + 2 large display items.
 Articles typically have 30-50 references (currently at 51)-->

### 2022-04-25 doc/manuscript/Dow_Extended_Data.Rmd
defined center of 95% ci

@@
 ---
+
 ```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
@@
 library(kableExtra)
+
 species_table <- read.csv(file = "tables_figures/species_table.csv", col.names = c("site", " ","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE)
@@
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
-95% credible intervals are represented by bands.
+The 95% credible intervals are represented by bands centered on the posterior mean for each year.
 For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective CTW.
@@
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
-95% credible intervals are represented by bands.
+The 95% credible intervals are represented by bands centered on the posterior mean for each year.
 For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI.

### 2022-04-26 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
@teixeirak FYI, Valentine and I dug a little deeper into how significance was determined in the quiltplots and found that it used a bootstrapped 95% (0.05) or 99.8% (0.002) CI to determine sig levels. IE, if the 95% CI included 0, it was NS. We're currently set up to report this as a p-value which isn't quite correct. Do you think we need to change how we report significance in the chronology table?

@@
 In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
-Across all chronologies, the standardized ring-width index (RWI) was significantly (p $\le$ 0.05) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
+Across all chronologies, the standardized ring-width index (RWI) was significantly (95% confidence interval did not include 0) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
 There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).

### 2022-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
add Elmore reference
closes #135

@@
 In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
@@
 `r tree_rings`).
-Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
+Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink limitation[@delpierre_wood_2016;@zani_increased_2020].
 This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].
-<!--rocha_linking_2006 may work here if needed-->
 

### 2022-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
create placeholder sections for main and methods references
#137

References at end will need to be copied here manually just before submission
response to R1 sequestration comment, + misc

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 `r tree_rings`).
-Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink limitation[@delpierre_wood_2016;@zani_increased_2020].
+Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink saturation[@delpierre_wood_2016;@zani_increased_2020].
 This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].
@@
 <!--discussion of extra C-->
-Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
+Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
 It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
@@
 However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
-`r schematic_summary`) is unlikely to provide a sustained augmentation of CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@richardson_terrestrial_2012;@keenan_net_2014;@buermann_widespread_2018].
+`r schematic_summary`) is unlikely to provide a sustained augmentation of C sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@richardson_terrestrial_2012;@keenan_net_2014;@buermann_widespread_2018].
 Rather, the dominant effect of rising temperatures on temperate deciduous forest woody productivity will be a negative effect of high summer temperatures (Fig.
 `r tree_rings`)[@ahlstrom_robustness_2012], which constitutes a positive feedback to climate change.
+
+# References {-#References}
 
@@
 
+## Methods References {-#Methods}
+
 # Acknowledgements {-#Acknowledgements} 
@@
 Correspondence and requests for materials should be addressed to Kristina Anderson-Teixeira (teixeirak@si.edu).
-
-# References {-#References}

### 2022-04-27 doc/submissions/2022_04_Nature/Response to reviews.Rmd
response to R1 sequestration comment, + misc
R1 responses

@@
 
-**Okay.**
+**Thank you.**
 
@@
 
-*Response needed.*
+**We have replaced "CO~2~ sequestration" or "C sequestration" with "CO~2~ uptake" when referring to ecosystem-atmosphere exchange of CO~2~ and use "sequestration" only to refer to woody growth.
+Further, we have removed all use of “CO2 sequestration” and now use only  “C sequestration”.**
 
@@
 I just wanted to point out that CTW is fitted to be related to DOY25, therefore it is rather expected to find a significant relationship between CWT Tmax and DOY25, and unfair to compare it with the relationship between CWT Tmax and ∆DBH.
+
+**We agree with this comment and specify our motivation in the methods section: "We defined CTW for *DOY~25~*, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable."**
+
+**We also repeat the analyses using a definition of CTW that is independent of *DOY~25~* and all other response variables.
+As explained in the methods, "To ensure that patterns were robust under an alternative definition of CTW, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
+3, Extended Data Figure 6-7), we also ran analyses where we fixed the CTW identified by climwin to be the month with the most days in the CTW (Extended Data Table 2) each critical window."**
+
 I appreciated the clarifications done concerning the data cleaning process.
@@
 
-*Response needed.*
+*Response needed: [https://github.com/EcoClimLab/growth_phenology/issues/141](https://github.com/EcoClimLab/growth_phenology/issues/141)*
 

### 2022-04-27 doc/submissions/2022_04_Nature/Response to reviews.Rmd
address R3 comment

@@
 
-*Response needed.*
+**We have clarified this in the abstract as follows: "Using dendrometer band measurements from 440 trees across two forests, we show that warmer spring temperatures shifted stem *diameter* growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.
+The latter finding was confirmed on the centennial scale by 207 tree-ring chronologies from 108 forests across eastern North America, where annual *ring width* was far more sensitive to temperatures during the peak growing season than in the spring.**
+
+**We have clarified this in several key places throughout the main body, including the following:**
+
+- **last paragraph of intro: Here, we evaluate how spring temperatures affect *the timing, rates, and annual increments of stem diameter growth* of temperate deciduous trees across eastern North America.
+To test whether warmer springs advance the timing and extend the duration of stem *diameter* growth, we used dendrometer band measurements on 440 trees across two mid-latitude forests.
+To test whether spring temperatures consistently increased *annual radial increments*, we analyzed 207 tree-ring chronologies from 108 forests."**
+- **first sentence of "Dendrometer Bands" section: "we fit a logistic growth model35 to determine the days of year (DOY) when 25, 50, and 75% annual *diameter* growth were achieved (DOY25, DOY50, DOY75), ..."**
+- **first sentence of "Tree-ring analysis" section: "To understand how annual *radial stem* growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies ..."**
+- **first sentence of discussion: "Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem *diameter* growth but have little effect on annual *increments*"**
 

### 2022-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
address R3 comment

@@
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
-**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
-**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
+**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem diameter growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
+**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual ring width was far more sensitive to temperatures during the peak growing season than in the spring.**
 **These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
@@
 
-Here, we evaluate how spring temperatures affect stem growth timing, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
-To test whether warmer springs advance the timing and extend the duration of stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
-To test whether spring temperatures consistently increased annual growth, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
+Here, we evaluate how spring temperatures affect the timing, rates, and annual increments of stem diameter growth of temperate deciduous trees across eastern North America. 
+To test whether warmer springs advance the timing and extend the duration of stem diameter growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
+To test whether spring temperatures consistently increased annual radial increments, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
 
@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
 `r schematic_summary`).
@@
 
-To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
+To understand how annual radial stem growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
 `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`).
@@
 <!--summary, setting expectation of extra C-->
-Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs.
+Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem diameter growth but have little effect on annual increments (Figs.
 `r schematic_summary`-`r tree_rings`).

### 2022-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
new title!
closes #139
remove one reference (gets us to 50 limit)
#137

@@
 
-**Title:** Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity
+**Title:** Warm springs alter timing but not total growth of temperate deciduous trees
 
@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
+In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[<!--@jeong_phenology_2011;--><!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@keenan_net_2014;@finzi_carbon_2020;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
 Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
@@
 In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
+Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].

### 2022-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
add ethics declaration, organization of end matter
Added / organized end matter following template of Cook-Patton et al. 2020 Nature

#137

@@
 
-# References {-#References}
+# References {-}
 
-# Methods {-#Methods} 
+# Methods {-} 
 <!--Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.-->
 
-## Dendrometer band analysis {-#Dendrometer} 
+## Dendrometer band analysis {-} 
 <!--Study sites -->
@@
 
-## Tree-ring analysis {-#Tree} 
+## Tree-ring analysis {-} 
 We analyzed tree-ring records for `r n_cores_sites` sites, including our focal sites.
@@
 
-## Methods References {-#Methods}
+## Methods References {-}
 
-# Acknowledgements {-#Acknowledgements} 
+# Acknowledgements {-} 
 We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
@@
 
-# Author Contributions {-#Author}
+# Author Information{-}
+
+## Contributions {-}
 
@@
 All authors contributed critically to the drafts and gave final approval for publication.
+
+## Corresponding Author {-}
+
+Correspondence to [Kristina J. Anderson-Teixeira](teixeirak@si.edu).
+
+# Ethics declarations {-}
+## Competing interests {-}
+
+The authors declare no competing interests.
 

### 2022-04-29 doc/manuscript/tables_figures/DOYtiming_allyears.Rmd
Create pdf versions of figures in main manuscript

--- 2022-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 ---
-title: ''
+title: ""
+author:
 output: 
-  bookdown::word_document2: 
-    reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+  pdf_document:
     fig_caption: yes
-bibliography: references.bib
-csl: nature.csl
-
+    #toc: true # table of content true
+    #toc_depth: 2  
+header-includes:  \usepackage{float}
+                  \usepackage{caption}
+                  \captionsetup[figure]{labelformat=empty}
+                  \captionsetup[table]{labelformat=empty}
 ---
@@
 
-\raggedright
-
-**Title:** Warm springs alter timing but not total growth of temperate deciduous trees
-
-**Authors:**
-
-Cameron Dow^1,2^ (Orcid ID : 0000-0002-8365-598X)
-
-Albert Y. Kim^1,3^ (Orcid ID : 0000-0001-7824-306X)
-
-Loïc D’Orangeville^4,5^ (Orcid ID : 0000-0001-7841-7082)
-
-Erika B. Gonzalez-Akre^1^ (Orcid ID : 0000-0001-8305-6672)
-
-Ryan Helcoski^1^ (Orcid ID: 0000-0003-3579-0121)
-
-Valentine Herrmann^1^ (Orcid ID : 0000-0002-4519-481X)
-
-Grant L. Harley^6^ (Orcid ID : 0000-0003-1557-8465)
-
-Justin T. Maxwell^7^ (Orcid ID: 0000-0001-9195-3146)
-
-Ian R. McGregor^1,8^ (Orcid ID: 0000-0002-5763-021X)
-
-William J. McShea^1^ (Orchid ID: 0000-0002-8102-02000)
-
-Sean M. McMahon^9,11^ (Orcid ID : 0000-0001-8302-6908)
-
-Neil Pederson^4^ (Orcid ID : 0000-0003-3830-263X)
-
-Alan J. Tepley^1,10^ (Orcid ID : 0000-0002-5701-9613)
-
-Kristina J. Anderson-Teixeira^1,11^* ( Orcid ID : 0000-0001-8461-9713)
-
-**Author Affiliations:**
-
-1. Conservation Ecology Center; Smithsonian National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA
-
-2. Department of Forestry and Natural Resources, Purdue University, West Lafayette, Indiana, USA
-
-3. Statistical & Data Sciences; Smith College; Northampton, MA 01063, USA
-
-4. Harvard Forest, Petersham, MA 01366, USA
-
-5. Faculty of Forestry and Environmental Management, University of New Brunswick, Fredericton, NB, E3B 5A3, Canada.
-
-6. Department of Earth and Spatial Sciences, University of Idaho, ID 83844, USA
-
-7. Department of Geography, Indiana University, Bloomington, IN 47405, USA
-
-8. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
-
-9. Smithsonian Environmental Research Center, Edgewater, MD, USA
-
-10. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
-
-11. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
-
-*corresponding author: teixeirak@si.edu; +1 540 635 6546
-
-```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
-
-## Target journal: Nature
-
-## Nature articles: 3-4 modest display items
-
-#Display items: listing figures and their number here is helpful for planning, and also allows automatic numbering within text -->easy changing of order.
-
-#Figure numbers
-schematic_summary=1 
-DOY_timing = 2
-tree_rings = 3
-
-# EXTENDED DATA  (this chunk must be repeated in Dow_Extended_Data.Rmd)
-##tables
-EDT_species="Extended Data Table 1"  
-EDT_phenology_parameters="Extended Data Table 2" 
-EDT_tree_ring_summary="Extended Data Table 3"
-
-##figures
-EDF_map = "Extended Data Figure 1" 
-EDF_growth_curves_all = "Extended Data Figure 2"  
-EDF_climwin = "Extended Data Figure 3"
-EDF_pheno_all = "Extended Data Figures 4-5"
-EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
-EDF_pheno_Tmax_month = "Extended Data Figure 5"
-EDF_tree_ring_figs= "Extended Data Figure 6-7"
-EDF_tree_rings_TMIN = "Extended Data Figure 6"
-EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"
-
-# Supplementary info
-chron_table= "SI Table 1"
-
-###~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-## some stats/ 
-species_table <- read.csv(file = "tables_figures/species_table.csv")
-n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
-n_dendro_trees_SCBI <- 119 #doing this the lazy way. summed from species_table rather than coded. 
-n_dendro_trees_HF <- 321 #doing this the lazy way. summed from species_table rather than coded. 
-n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
-n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
-chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
-n_chronologies <- nrow(chronology_table)  
-n_cores_sites <- length(unique(chronology_table[, "Location"]))
-n_cores_species <- length(unique(chronology_table[, "Species"])) - 1 #the -1 accounts for the Quercus spp. record (combination of white oak group)
-n_chronologies_SCBI <- 12
-n_chronologies_HF <- 4
-```
-
-\newpage
-
-<!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text).
-We're currently at 2687 words and 1 modest + 2 large display items.
-Articles typically have 30-50 references (currently at 51)-->
-
-<!-- Articles start with a fully referenced summary paragraph, ideally of no more than 200 words, which is separate from the main text and avoids numbers, abbreviations, acronyms or measurements unless essential.
-It is aimed at readers outside the discipline.
-This summary paragraph should be structured as follows: 2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
-Please refer to our annotated example  to see how the summary paragraph should be constructed.-->
-
-<!--SUMMARY PARAGRAPH-->
-
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
-**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
-**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem diameter growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
-**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual ring width was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
-**Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**
-
-<!--BODY-->
-
-In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
-Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
-<!--@xu_changes_2021 provides an alternate estimate, which is about half of Harris.
-The main difference being that Xu includes savannas and woodlands, which have experienced a lot of loss from fires.
-Given focus on forests here, Harris seems more appropriate.
--->
-Of this important C sink, ~ `r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021],
-<!-- alternate estimate: `r round(.31/.88,2)*100`% (temperate net / global net, (@xu_changes_2021. Table 2))--> 
-with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019].
-<!-- Pugh et al.
-2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.-->
-The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@ahlstrom_robustness_2012;@friedlingstein_global_2020;@fatichi_moving_2014].
-
-In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[<!--@jeong_phenology_2011;--><!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@keenan_net_2014;@finzi_carbon_2020;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
-However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
-Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
-As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
-Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
-
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022].
-In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
-Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
-However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
-Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021].
-Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
-These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
-Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
-
-Here, we evaluate how spring temperatures affect the timing, rates, and annual increments of stem diameter growth of temperate deciduous trees across eastern North America. 
-To test whether warmer springs advance the timing and extend the duration of stem diameter growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
-To test whether spring temperatures consistently increased annual radial increments, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.
-
-## Dendrometer band analysis {-#Dendrometer}
-
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
-`r schematic_summary`).
-This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022].
-These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
-
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
-
-<!--Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).-->
-<!--Consistent with the results of Ref[@dorangeville_peak_2022], ring porous species began growing earlier, reaching the *DOY~25~* benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average *L~pgs~* 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig.
-`r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).-->
-<!--Peak growing season length was similar across sites, with *L~pgs~* being, on average, only six days longer at SCBI for ring porous species and less than one day longer for diffuse porous species (`r  EDT_phenology_parameters`).-->
-
-Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
-We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW).
-CTW was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
-The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.
-
 ![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
-Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
+Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in Extended Data Figure 4.
 Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }
-
-For ring- and diffuse- porous species at both sites, warmer CTW *T~max~* resulted in earlier achievement of seasonal milestones.
-Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig.
-`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
-Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs.
-`r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
-Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).
-
-Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig.
-`r DOY_timing`a-b), there was no consistent lengthening of *L~pgs~* (Fig.
-`r schematic_summary`, `r EDF_pheno_all`).
-
-In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were usually weak and inconsistent (Fig.
-`r schematic_summary`, `r EDF_pheno_all`).
-Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.003-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
-
-## Tree-ring analysis {-#Tree}
-
-To understand how annual radial stem growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig.
-`r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`).
-In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
-Across all chronologies, the standardized ring-width index (RWI) was significantly (95% confidence interval did not include 0) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
-There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
-In contrast, RWI was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring- and diffuse-porous species, respectively.
-*T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and RWI (`r EDF_tree_rings_TMIN`).
-
-To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and *T~max~* in the previous year (`r EDF_tree_rings_TMAX_extended`).
-This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of RWI to previous March or April *T~max~* for 5/142 ring porous chronologies and to previous April or May *T~max~* for 7/66 diffuse porous chronologies.
-
-To test whether the negative effect of summer temperatures might offset an enhancement of growth by warmer spring temperatures, we tested for the joint effects of April and June-July *T~max~* on RWI.
-Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (p $\le$ 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
-
-![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
-Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
-Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.
-Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`).
-Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
-
-## Discussion {-#Discussion}
-
-<!--summary, setting expectation of extra C-->
-Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem diameter growth but have little effect on annual increments (Figs.
-`r schematic_summary`-`r tree_rings`).
-The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and NEE[@keenan_net_2014;@finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
-`r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig.
-`r tree_rings`).
-Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink saturation[@delpierre_wood_2016;@zani_increased_2020].
-This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].
-
-<!--discussion of extra C-->
-Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
-It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
-Extra C fixed in years with warm springs could potentially be allocated to formation of more C-dense wood, either through enhanced cell wall thickening (a process that lags behind stem expansion[@cuny_woody_2015]) or to a higher ratio of high-density latewood to lower-density earlywood.
-However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
-Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or ANPP~woody~.
-
-The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
-Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, ANPP~woody~ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
-Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
-Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.
-
-<!--.short vs long-term effects-->
-It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of ANPP~woody~.
-Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], ANPP~woody~[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
-Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
-If extra photosynthate made available through a growing difference between GPP and ANPP~woody~ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase ANPP~woody~ through indirect mechanisms.
-Understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.
-
-<!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
-Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season of the current or sometimes previous year (Fig.`r tree_rings`, `r EDF_tree_rings_TMAX_extended`), when increased atmospheric demand associated with high temperatures can limit both leaf-level gas exchange and stem growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
-Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to *T~max~* (Fig.
-`r tree_rings`, `r EDT_tree_ring_summary`).
-This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress or high temperatures during the peak growing season[@dorangeville_drought_2018;@helcoski_growing_2019;@anderson-teixeira_joint_2022;@babst_twentieth_2019].
-<!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of *T~max~* in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
-Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
-While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), our tree-ring analysis does reveal a higher frequency of negative than positive correlations of annual growth to spring temperatures, particularly for ring porous species in cooler climates (Fig.
-`r tree_rings`, `r EDT_tree_ring_summary`).
-Thus, warm springs can have a net negative effect on growth, particularly when water is limiting[@gao_earlier_2022].
-
-<!--conclusions-->
-As spring temperatures become increasingly warmer, growing seasons will start earlier. 
-However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
-`r schematic_summary`) is unlikely to provide a sustained augmentation of C sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@richardson_terrestrial_2012;@keenan_net_2014;@buermann_widespread_2018].
-Rather, the dominant effect of rising temperatures on temperate deciduous forest woody productivity will be a negative effect of high summer temperatures (Fig.
-`r tree_rings`)[@ahlstrom_robustness_2012], which constitutes a positive feedback to climate change.
-
-# References {-#References}
-
-# Methods {-#Methods} 
-<!--Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.-->
-
-## Dendrometer band analysis {-#Dendrometer} 
-<!--Study sites -->
-
-Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2022], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015;@davies_forestgeo_2021].
-SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km south of Front Royal, Virginia. 
-The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
-Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera* L.), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
-The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
-Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April *T~max~*; source: CRU v.4.04[@harris_version_2020]) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
-Specifically, average spring *T~max~* was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.
-
-Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.) is located near the central Massachusetts town of Petersham. 
-The forest is secondary and mixed age, having re-established around the beginning of the 20th century following agricultural use and significant hurricane damage in 1938. 
-Dominant species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis* (L.) Carrière), oak (*Quercus spp.*) and red maple (*Acer rubrum* L.). 
-The climate is temperate continental, with 1950-2019 mean annual precipitation of 1104 mm and temperatures averaging -5°C in January and 22°C in July[@anderson-teixeira_joint_2022]. 
-Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April *T~max~*; source: CRU v.4.04[@harris_version_2020]) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
-Specifically, average spring *T~max~* was 10.9$\pm$ 1.5°C prior to the study period (range: 8.0-13.2) and 11.2$\pm$ 1.0°C (range: 10.1-12.2) during the study period, while summer SPEI was 0.1$\pm$ 0.9 (range: -1.8-1.7) prior to the study period and 0.2$\pm$ 0.9 (range: -1.0-1.1) during.
-The driest summer during the study period (1999) had the 5th lowest SPEI value (-1.0) in the period 1970-2003, with precipitation of 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
-
-<!--Data collection and preparation -->
-
-Metal dendrometer bands were installed on 941 trees within the SCBI and Harvard Forest ForestGEO plots.
-Bands were placed on dominant species, including two diffuse- and two ring porous species at SCBI and eight diffuse- and three ring porous species at Harvard Forest (`r EDT_species`).
-While we do not estimate the ages of trees in our sample, bands at both sites were placed on individuals of differing sizes in an attempt to measure trees across a range of ages.
-Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
-The number of bands measured at each site fluctuated slightly as trees were added or dropped from the census (e.g., because of tree mortality). 
-Across years, the number of bands sampled averaged 129 (range: 91-138) at SCBI and 717 (range: 700-755) at Harvard Forest. <!--verified - these are the pre-cleaning numbers-->
-In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).
-
-Measurements were timed to begin before the beginning of spring growth and to continue through the cessation of growth in the fall. 
-At SCBI, the median start date was April 14, which was adjusted forward when early leaf-out of understory vegetation was observed, with the earliest start date being March 30 (in 2020). 
-Measurements were continued through to fall leaf senescence, with the median end date being October 17 and the latest end date November 26 (2012). 
-At Harvard Forest, all measurements from 1998 were dropped because of a late start date (May 26).
-Among the remaining years,the median start date was April 21 and median end date of October 27.
-1999 was an anomalous year where initial measurements were taken on January 5, but not taken again until April 15.
-The latest end date was November 11, 2002.
-In our analysis, each band-year was treated independently, with no data overlap from one year to the next.
-
-The raw dendrometer band data were screened before analysis to remove records or entire tree-years that were either unreliable or inappropriate for our analysis. 
-First, we manually screened the data for three classes of errors. 
-First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and removed.
-Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck on the tree bark and then released.
-In these cases, the full annual record for the tree was removed.
-Third, data points that deviated substantially from normal growth patterns, but for unknown causes, were removed. 
-If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
-We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253), and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
-These were removed because they reduce the reliability of predicted growth in the modeled curves.
-
-We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig. `r schematic_summary`). 
-In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
-$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
-Here, $L$ and $K$ are lower and upper asymptotes of the model. 
-*DOY~ip~* is the day of year where the inflection point in growth rate occurs,
-$r$ shapes the slope of the curve at the inflection point,
-and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
-This allows an asymmetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth. 
-When $\theta = 1$, *g~max~* occurs on *DOY~ip~*.
-The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. 
-We used the first and last manual dendrometer band measurements as starting points to calculate these two variables. 
-With this method, initial measurements too late in the growing season which miss the start of the growing season can cause incorrect estimations of $a$ and $b$. 
-The 1998 survey at Harvard Forest recorded first measurements on May 26, which is less than 2 weeks away from the average DOY25 at this site. 
-This late start made the calculation of timing and total growth variables unreliable, so we removed tree years in 1998 from our analysis. 
-The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. 
-These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
-An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015;@herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest.
-Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig.
-`r schematic_summary`).
-
-After fitting the growth model, we removed tree-years with poor fits.
-Models were judged to be poorly fit if modeled growth parameters were outliers, which were commonly indicative of unrealistic fits (e.g., growth occurring outside the growing season or over a very short period) and underlain by very slow tree growth or poor data records that passed the initial screening (described above).
-Modeled fits for tree-years were removed under two conditions:
-(1) *g~max~* was $\ge 2.5$ standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);
-(2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
-At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
-This process was repeated using 2 and 3 standard deviations as the cutoff for defining outliers, yielding qualitatively similar results.
-
-<!-- Leaf phenology -->
-
-Canopy foliage phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
-MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019] via Google Earth Engine.
-For each year at each site, we extracted data from the pixel (500m resolution) containing the center of each forest plot. 
-Using the daily MODIS 2-band Enhanced Vegetation Index data (EVI2), the product yields the timing of phenometrics (vegetation phenology) over each year, including timing of greenup, midgreenup, peak, and senescence as used in this study.
-Data points were included in the analysis if they were flagged as "good" or "best" quality.
-
-<!--Weather data -->
-For the dendrometer band and leaf phenology analyses, climate data corresponding to the measurement periods were obtained from local weather stations at each focal site. 
-For SCBI, weather data were obtained from a meteorological tower adjacent to the ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020].
-The R package *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset.
-Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
-Daily temperature records for Harvard Forest, which had already been gap-filled based on other local records, were obtained from the Harvard Forest weather station[@boose_shaler_2021;@boose_fisher_2021].
-For each site, we used records of daily maximum (*T~max~*) and minimum temperatures (*T~min~*).
-Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values were obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020;@begueria_standardized_2014;@vicente-serrano_multiscalar_2010a].
-
-<!-- Determining critical temperature windows -->
-The critical temperature window (CTW, Fig.
-`r schematic_summary`), defined as the period over which *T~max~* was most strongly correlated with *DOY~25~*, was determined using the R package *climwin*[@vandepol_identifying_2016].
-This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame. 
-It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
-Here, we tested for correlation between temperature predictor variables (*T~max~*, *T~min~*) and biological outcome variable *DOY~25~* over the time-frame from January 1 to the mean *DOY~25~* for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`).
-The time period yielding the lowest $\Delta AICc$ was selected as the CTW.
-To avoid spurious correlations that could occur using temperature data at the daily resolution, we ran this analysis with weekly resolution, using temperatures averaged over weekly time periods.
-Because *T~max~* proved to have a generally stronger influence over *DOY~25~* and other growth parameters, we focused on this variable in our ultimate model, as opposed to *T~min~*.
-We defined CTW for *DOY~25~*, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable.
-
-To ensure that patterns were robust under an alternative definition of CTW, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
-`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the CTW identified by *climwin* to be the month with the most days in the CTW (`r  EDT_phenology_parameters`) each critical window.
-The months identified were March and April for ring- and diffuse-porous species at SCBI, respectively, and April and May for ring- and diffuse-porous species at Harvard Forest, respectively.
-
-<!--Hypothesis testing -->
-Correlation between the dendrometer band-derived growth parameters (*DOY~25~*, *DOY~50~*, *DOY~75~*, *L~pgs~*, *g~max~*, and $\Delta DBH$, Fig.
-`r schematic_summary`) and CTW *T~max~* (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
-Analyses were run for both *T~max~* and *T~min~*, with qualitatively similar results, but we present only results for *T~max~*, which had overall stronger correlation with growth parameters. 
-Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean *T~max~* (or *T~min~*) during the CTW, along with random effects of species and of individual tree.
-We ran separate models for each site, and for the response of all growth parameters to *T~max~* (or *T~min~*). 
-This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@standevelopmentteam_stan_2019].
-In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
-<!--
-More information on rstanarm's default weakly informative priors:
-https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
--->
-
-To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer SPEI indices. 
-Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April *T~max~* to determine if warm spring temperatures were associated with greater summer drought stress in our data set.
-No significant correlations were found (all p>0.05).
-
-## Tree-ring analysis {-#Tree} 
-We analyzed tree-ring records for `r n_cores_sites` sites, including our focal sites.
-All cores had been previously collected, cross-dated, and measured using standard collection and processing methodologies[@stokes_introduction_1968;@speer_fundamentals_2010].
-
-Dominant tree species were cored at both SCBI[@helcoski_growing_2019;@bourg_initial_2013] and Harvard Forest[@alexander_potential_2019;@dye_comparing_2016;@finzi_carbon_2020] following sampling designs that covered a broad range of DBH.
-We analyzed records for the ring- and diffuse- porous species at each site (`r  EDT_species`), but excluded semi-ring porous species (e.g., *Juglans nigra* L.
-at SCBI) and conifers (e.g., *Tsuga canadensis* at Harvard Forest).
-We studied a total of `r n_cores_focal_sites` cores which included `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest (`r  EDT_species`).
-
-The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in eastern North America[@pederson_climatic_2005;@maxwell_sampling_2020;@dorangeville_drought_2018].
-For the majority of sampled populations (i.e., site-species combinations), sampling focused on canopy trees (typically >20 trees per population)[@pederson_climatic_2005;@maxwell_sampling_2020;@dorangeville_drought_2018], while ~15% of the total 207 chronologies came from plot-level collections where trees above a certain diameter (typically 10 cm DBH) were censused and cored[@dye_comparing_2016;@helcoski_growing_2019].
-Again, analyses were limited to broadleaf deciduous species with clearly defined xylem porosity (*i.e.*, excluding semi-ring porous).
-
-For each species-site combination, we converted tree-ring records into the dimensionless RWI to emphasize interannual variability associated with climate[@cook_methods_1990].
-A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
-An adaptive power transformation, a process that also stabilizes the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
-Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records[@jones_estimating_1997].
-The 1/3rds spline method was chosen when replication in the inner portion of each chronology (the earliest ca. 30–50 yr of each record depending on full chronology length) was less than three trees. 
-When replication was greater than n = 3 trees, we used the average correlation between raw ring-width series (rbar) method.
-The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilization[@cook_time_1985].
-We defined chronology start year (`r EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
-
-<!--QUILT PLOT METHODS-->
-For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
-`r tree_rings`, Extended Data Figures 6-7).
-Our analysis focused on assessing correlations of RWI to months spanning January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
-To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
-Correlations and significance levels for months March-August are given in `r chron_table`.
-
-<!--MULTIVARIATE ANALYSIS METHODS-->
-We used a multivariate model to test for joint effects of April and summer *T~max~* on RWI.
-We focused on April to represent spring temperatures because it was the month with greatest overall alignment with the CTW's identified in the dendrometer band analysis and had the highest rate of positive correlations with RWI (`r EDT_tree_ring_summary`).
-We began by testing univariate correlations of *T~max~* over three summer windows: June, June-July, and May-August. 
-Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April *T~max~* and June-July *T~max~* on RWI for each chronology independently using the base lm() function in R.
-Slopes and p-values for each chronology are given in `r chron_table`. 
-Although some models may have benefited from data transformations, we determined that assumptions of normality and homoscedasticity were sufficiently met for the purposes of this analysis.
-
-## Methods References {-#Methods}
-
-# Acknowledgements {-#Acknowledgements} 
-We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
-Thanks to David A. Orwig, members of the ForestGEO Ecosystems & Climate Lab at SCBI, and four anonymous reviewers for helpful feedback.
-The research was funded by ForestGEO (Smithsonian). 
-Collection of tree-ring samples was funded by a USDA Agriculture and Food Research Initiative grant 2017-67013-26191 and from the Indiana University Vice Provost for Research Faculty Research Program.
-
-# Author Contributions {-#Author}
-
-Cameron Dow and Kristina J. Anderson-Teixeira conceived the ideas and designed the study; 
-Cameron Dow, Loïc D’Orangeville, Erika B.
-Gonzalez-Akre, Ryan Helcoski, Grant L.
-Harley, Justin T.
-Maxwell, Ian R.
-McGregor, William McShea, Neil Pederson, Alan J.
-Tepley, and Kristina J.
-Anderson-Teixeira collected or oversaw collection of data;
-Cameron Dow, Albert Y. Kim, Valentine Herrmann, Justin T. Maxwell, Ian R. McGregor, Sean M. McMahon analyzed the data or provided analytical tools; 
-Cameron Dow and Kristina J. Anderson-Teixeira led the writing of the manuscript. 
-All authors contributed critically to the drafts and gave final approval for publication.
-
-# Data Availability Statement {-#Data} 
-The datasets generated and analysed during the current study are available via GitHub in the *growth_phenology* repository of the ForestGEO Ecosystems & Climate Lab @ SCBI, (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI *[TBD]*).
-Master versions of the dendrometer band data are available for SCBI via GitHub in the *Dendrobands* repository of the Smithsonian Conservation Biology Institute ForestGEO plot (https://github.com/SCBI-ForestGEO/Dendrobands), which is archived in Zenodo (DOI 10.5281/zenodo.5551143), and for Harvard Forest via the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF149).
-Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
-Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF001 AND https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF000).
-Climate data were obtained from CRU v.4.04 via the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
-The Standardised Precipitation-Evapotranspiration Index was obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/SPEI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
-Canopy foliage phenology data were extracted from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a.
-MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description).
-In addition to being archived in this project's repository, many tree-ring data sets are archived in the International Tree-Ring Data Bank (ITRDB;https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring), the DendroEcological Network (DEN; https://www.uvm.edu/femc/dendro/), and/or the Harvard Forest Data Archive (https://harvardforest.fas.harvard.edu/harvard-forest-data-archive), as detailed in `r chron_table`.
-Original tree cores are archived at the institutions of various members of the author team (Harvard Forest, Smithsonian Conservation Biology Institute, Indiana University, and University of Idaho) and will be made available upon reasonable request.
-
-# Code Availability Statement {-#Code}
-Data were analysed in the open source statistical software R (version 4.0).
-We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
-We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
-All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).
-
-# Additional Information {-#Additional}
-
-**Supplementary Information** is available for this paper.
-
-Correspondence and requests for materials should be addressed to Kristina Anderson-Teixeira (teixeirak@si.edu).
+Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](DOYtiming_allyears.png){ width=100% }

### 2022-04-29 doc/manuscript/tables_figures/tmax_quiltplot.Rmd
Create pdf versions of figures in main manuscript

---
title: ""
author:
output: 
  pdf_document:
    fig_caption: yes
    #toc: true # table of content true
    #toc_depth: 2  
header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for 207 chronologies from 108 sites across eastern North America**. Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; Extended Data Table 1). Sites included are mapped in Extended Data Figure 1, species analyzed and numbers of significant correlations to *T~max~* are summarized in `Extended Data Table 3, and chronology details are given in SI Table 1.](tmax_quiltplots.jpg){ width=100% }

### 2022-05-07 doc/submissions/2022_04_Nature/Response to reviews.Rmd
draft response
@teixeirak let me know what you think

@@
 
-*Response needed: [https://github.com/EcoClimLab/growth_phenology/issues/141](https://github.com/EcoClimLab/growth_phenology/issues/141)*
+*Our data cleaning methods were split into two sections.
+First, we performed manual inspection of the raw dendrometer band measurements to remove unreliable records.
+Second, we removed models which returned parameters that we determined to be outliers.
+We provide quanitity of data removed for the second section, but not the first.
+When we revisited the first section to provide numbers, we noticed that the manually cleaned dendrometer band measurements we originally planned to feed into the modeling process had been replaced by the raw dendrometer band measurements with no manual cleaning done.
+As such, we have decided to remove the text, "The raw dendrometer band data were screened before analysis to remove records or entire tree-years that were either unreliable or inappropriate for our analysis.
+First, we manually screened the data for three classes of errors.
+First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and removed.
+Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck on the tree bark and then released.
+In these cases, the full annual record for the tree was removed.
+Third, data points that deviated substantially from normal growth patterns, but for unknown causes, were removed.
+If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis." from the methods section.
+*
+
+*We believe that losing this step in the data cleaning process does not significantly alter the quantity or quality of the data we ultimately used for analysis.
+This is because any tree-year which may have been removed using the criteria above would either fail the model fitting process, or would be removed if it produced an outlier model.
+Meaning, since there were not a substantial number of tree-years where manually cleaning was necessary, any model that was potentially compromised by faulty measurements would be removed by our post-model cleaning method.*
 

### 2022-05-07 doc/manuscript/Dow_manuscript_Nature.Rmd
draft response
@teixeirak let me know what you think

@@
 
-The raw dendrometer band data were screened before analysis to remove records or entire tree-years that were either unreliable or inappropriate for our analysis. 
-First, we manually screened the data for three classes of errors. 
-First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and removed.
-Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck on the tree bark and then released.
-In these cases, the full annual record for the tree was removed.
-Third, data points that deviated substantially from normal growth patterns, but for unknown causes, were removed. 
-If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
-We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253), and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
+We removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253), and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
 These were removed because they reduce the reliability of predicted growth in the modeled curves.

### 2022-05-09 doc/submissions/2022_04_Nature/Response to reviews.Rmd
update text on n records removed, response to reviewer on this
#141

@camerondow35 , I've changed this to add direct statements of numbers removed, which I think is what the reviewer was looking for.

--- 2022-04-27 doc/submissions/2022_04_Nature/Response to reviews.Rmd
@@
 
-*Response needed: [https://github.com/EcoClimLab/growth_phenology/issues/141](https://github.com/EcoClimLab/growth_phenology/issues/141)*
+**We simplified the presentation of our data cleaning methods, ensuring that these numbers are clearly presented for both stagges of cleaning:**
+
+- **Paragraph describing screening of raw dendrometer band data: "In total, this process removed 309 of the 2210 available records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest."**
+- **Paragraph describing removal of poorly fit records: "In total, this process removed 189 of the 2467 records deemed appropriate for analysis."**
 

### 2022-05-09 doc/manuscript/Dow_manuscript_Nature.Rmd
update text on n records removed, response to reviewer on this
#141

@camerondow35 , I've changed this to add direct statements of numbers removed, which I think is what the reviewer was looking for.

@@
 
-We removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253), and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
-These were removed because they reduce the reliability of predicted growth in the modeled curves.
+The raw dendrometer band data were screened to remove records or entire tree-years that were inappropriate for our analysis because of reduced reliability of predicted growth in the modeled curves.
+Specifically, we removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253) and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
+In total, this process removed `r 26+253+22+8` of the `r n_tree_years` available records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest.
 
@@
 (2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
+In total, this process removed `r 3+11+74+101` of the `r n_tree_years - 26+253+22+8` records deemed appropriate for analysis.
 At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.

### 2022-05-09 doc/submissions/2022_04_Nature/Response to reviews.Rmd
clarify dendro band n's
#141

--- 2022-04-27 doc/submissions/2022_04_Nature/Response to reviews.Rmd
@@
 
-*Response needed: [https://github.com/EcoClimLab/growth_phenology/issues/141](https://github.com/EcoClimLab/growth_phenology/issues/141)*
+**We simplified the presentation of our data cleaning methods, ensuring that these numbers are clearly presented for both stagges of cleaning:**
+
+- **Paragraph describing screening of raw dendrometer band data: "In total, this process removed 309 of the 2708 available tree-year records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest."**
+- **Paragraph describing removal of poorly fit records: "In total, this process removed 189 of the 2399 tree-year records deemed appropriate for analysis, leaving a total of 2210 tree-years included in the final analysis (Extended Data Table 1)."**
 

### 2022-05-09 doc/manuscript/Dow_Extended_Data.Rmd
add map source
#137

@camerondow35 , FYI. Not sure if we should add any more info here.
clarify dendro band n's
#141

@@
 ## table captions
-species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
+species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in our final analysis."
 mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the seasonality and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
@@
 ![**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
-Sampling details are provided in `r chron_table`.](tables_figures/chronology_map.png)
+Sampling details are provided in `r chron_table`.
+Base map source is ggplot2.](tables_figures/chronology_map.png)
 

### 2022-05-09 doc/manuscript/Dow_manuscript_Nature.Rmd
clarify dendro band n's
#141

@@
 Across years, the number of bands sampled averaged 129 (range: 91-138) at SCBI and 717 (range: 700-755) at Harvard Forest. <!--verified - these are the pre-cleaning numbers-->
-In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).
 
@@
 Specifically, we removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253) and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
-In total, this process removed `r 26+253+22+8` of the `r n_tree_years` available records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest.
+In total, this process removed `r 26+253+22+8` of the `r n_tree_years + (3+11+74+101) + (26+253+22+8)`<!--first set is n removed for poor model fit, 2nd set is n removed here--> available tree-year records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest.
 
@@
 (2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
-In total, this process removed `r 3+11+74+101` of the `r n_tree_years - 26+253+22+8` records deemed appropriate for analysis.
+In total, this process removed `r 3+11+74+101` of the `r n_tree_years + (3+11+74+101)` tree-year records deemed appropriate for analysis, leaving a total of `r n_tree_years` tree-years included in the final analysis (`r  EDT_species`).
 At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.

### 2022-05-09 doc/manuscript/Dow_manuscript_Nature.Rmd
remove figures from article file
#137

figures commented out in .Rmd, captions moved to end

@@
 
-![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }
+<!--![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }-->
 
@@
 
-![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+<!--![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
 Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
@@
 Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }
+Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }-->
 
@@
 
-![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
+<!--![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
 Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
@@
 Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`).
-Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
+Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }-->
 
@@
 
-# Data Availability Statement {-#Data} 
+# Data Availability Statement {-} 
 The datasets generated and analysed during the current study are available via GitHub in the *growth_phenology* repository of the ForestGEO Ecosystems & Climate Lab @ SCBI, (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI *[TBD]*).
@@
 
-# Code Availability Statement {-#Code}
+# Code Availability Statement {-}
 Data were analysed in the open source statistical software R (version 4.0).
@@
 
-# Additional Information {-#Additional}
+# Additional Information {-}
 
@@
 Correspondence and requests for materials should be addressed to Kristina Anderson-Teixeira (teixeirak@si.edu).
+
+\newpage
+# Figure Captions {-}
+
+**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.
+
+**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down).
+Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003).
+Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
+Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).
+
+**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
+Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`).
+Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.
+
+\newpage

### 2022-05-10 doc/manuscript/Dow_manuscript_Nature.Rmd
final read - minor changes
@camerondow35 , please confirm content on line 331. Specifically, we say a and b are constrained by the first and last measurements, but I thought it was min and max? It's not clear to me from looking at Sean's paper what the code would be doing.

@@
 
-Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022].
+Most studies on tree stem growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022].
 In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
@@
 
-In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were usually weak and inconsistent (Fig.
+In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were inconsistent, and usually weak (Fig.
 `r schematic_summary`, `r EDF_pheno_all`).
@@
 
-To test whether the negative effect of summer temperatures might offset an enhancement of growth by warmer spring temperatures, we tested for the joint effects of April and June-July *T~max~* on RWI.
+To test whether there may be an enhancement of growth by warmer spring temperatures that was offset by the negative effect of high summer temperatures, we tested for the joint effects of April and June-July *T~max~* on RWI.
 Results were qualitatively similar to the univariate correlations (Fig.
@@
 If extra photosynthate made available through a growing difference between GPP and ANPP~woody~ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase ANPP~woody~ through indirect mechanisms.
-Understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.
+Understanding how warming spring temperatures are influencing C allocation within ecosystems remains a key outstanding question.
 
 <!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
-Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season of the current or sometimes previous year (Fig.`r tree_rings`, `r EDF_tree_rings_TMAX_extended`), when increased atmospheric demand associated with high temperatures can limit both leaf-level gas exchange and stem growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
+Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season of the current or sometimes previous year (Fig.
+`r tree_rings`, `r EDF_tree_rings_TMAX_extended`), when increased atmospheric demand associated with high temperatures can limit both leaf-level gas exchange and stem growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021].
 Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to *T~max~* (Fig.
@@
 When $\theta = 1$, *g~max~* occurs on *DOY~ip~*.
-The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. 
-We used the first and last manual dendrometer band measurements as starting points to calculate these two variables. 
-With this method, initial measurements too late in the growing season which miss the start of the growing season can cause incorrect estimations of $a$ and $b$. 
-The 1998 survey at Harvard Forest recorded first measurements on May 26, which is less than 2 weeks away from the average DOY25 at this site. 
-This late start made the calculation of timing and total growth variables unreliable, so we removed tree years in 1998 from our analysis. 
+The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year and are constrained by the first and last dendrometer band measurements. 
 The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. 

### 2022-05-16 doc/manuscript/Dow_manuscript_Nature.Rmd
add Cabon reference
closes #145

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;<!--@crabbe_extreme_2016;-->@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 **The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual ring width was far more sensitive to temperatures during the peak growing season than in the spring.**
-**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
+**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011<!--;@crabbe_extreme_2016-->] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
 **Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**
@@
 However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
-Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021].
+Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021;@cabon_crossbiome_2022].
 Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
@@
 Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink saturation[@delpierre_wood_2016;@zani_increased_2020].
-This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].
+This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014;@cabon_crossbiome_2022], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].
 
 <!--discussion of extra C-->
-Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
+Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;<!--@crabbe_extreme_2016;-->@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
 It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
@@
 However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig.
-`r schematic_summary`) is unlikely to provide a sustained augmentation of C sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@richardson_terrestrial_2012;@keenan_net_2014;@buermann_widespread_2018].
+`r schematic_summary`) is unlikely to provide a sustained augmentation of C sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@richardson_terrestrial_2012;@keenan_net_2014;@buermann_widespread_2018;@cabon_crossbiome_2022].
 Rather, the dominant effect of rising temperatures on temperate deciduous forest woody productivity will be a negative effect of high summer temperatures (Fig.

### 2022-05-21 doc/manuscript/Dow_manuscript_Nature.Rmd
Rewording awkward sentence

@@
 For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] to assess correlations, provide bootstrapped confidence intervals for these relationships, and create summary figures (Fig.
+Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013].
+Reported correlations and significance were determined using bootstrapped confidence intervals.
+Summary figures were created using the package *dplR*[@bunn_dendrochronology_2008] (Fig.
 `r tree_rings`, Extended Data Figures 6-7).

### 2022-05-21 doc/manuscript/Dow_Extended_Data.Rmd
clarify analysis used
#147

Valentine and I decided this was a bootstrapped confidence interval. Does this text clarify that? We already changed it in the methods, just not in the legends @teixeirak

@@
 ![**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean minimum temperatures ($T_{min}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
-Colors indicate the correlation between monthly $T_{min}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Colors indicate the bootstrapped correlation between monthly $T_{min}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April $T_{min}$.
@@
 ![**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$) of the current and past year, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
-Colors indicate the correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Colors indicate the bootstrapped correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$.

### 2022-05-21 doc/manuscript/Dow_Extended_Data.Rmd
change table 3 caption
Suggested by valentine (See: https://github.com/EcoClimLab/growth_phenology/commit/d69250697de0fbbcd7cffa98f501ff45f90a0a1d#commitcomment-72315269)

@@
 mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the seasonality and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
-tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."
+tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at significance level = 0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."
 

### 2022-05-21 doc/manuscript/Dow_manuscript_Nature.Rmd
clarify analysis used
#147

Valentine and I decided this was a bootstrapped confidence interval. Does this text clarify that? We already changed it in the methods, just not in the legends @teixeirak
add Rdendrom version

@@
 Data were analysed in the open source statistical software R (version 4.0).
-We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
+We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* v.0.1.0 (https://github.com/seanmcm/RDendrom/).
 We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
@@
 **Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
-Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Colors indicate the bootstrapped correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.

### 2022-05-21 RScripts/Create_final_quiltplot.R
Update Create_final_quiltplot.R

@@ -32,7 +32,7 @@ all_dcc_output_scbi$site <- paste0("SCBI_", all_dcc_output_scbi$Species)
 all.dcc.output_all <- rbind(all_dcc_output_other,all_dcc_output_scbi,all_dcc_output_hf)
 
 #Load in clim means
-clim_means <- read_csv("Results/tree_cores/quiltplots/clim_means_all.csv")#Created in other_quiltplot.R
+clim_means <- read_csv("Results/tree_cores/quiltplots/plot_data/clim_means_all.csv")#Created in other_quiltplot.R
 clim_means$tmx <- ifelse(clim_means$Location == "HF_LyfordPlots", clim_means$tmx+0.01, clim_means$tmx)#added to allow HF_lyford plots to be plotted separately from HF, even though spring means were the same. Does not affect analyses
 #Create porosity lists
 RP <- c("CAGL","CAOV","CATO","CACO","QURU", "QUST", "QUAL","QUPR","QUMO", "FRAM", "QUVE", "FRNI","QUMA", "QUPA")
@@ -137,7 +137,7 @@ X <- X[X$wood_type != "SP",]
 # write.csv(SI_table, file = paste0("doc/manuscript/tables_figures/", "chronology_table.csv"), row.names = FALSE)
   # X$entry_number <- seq(1, nrow(X)/8, 1)
   #
-  write.csv(X, file = paste0("results/tree_cores/quiltlplots/all/", v, "_quilt_plot_data.csv"), row.names = FALSE)
+  write.csv(X, file = paste0("results/tree_cores/quiltplots/plot_data/all/", v, "_quilt_plot_data.csv"), row.names = FALSE)
 }
 
 #TMAX LOOP
@@ -200,7 +200,7 @@ for(v in climate_variables) {
   # write.csv(SI_table, file = paste0("doc/manuscript/tables_figures/", "chronology_table.csv"), row.names = FALSE)
   # X$entry_number <- seq(1, nrow(X)/8, 1)
   #
-  write.csv(X, file = paste0("results/tree_cores/quiltlplots/all/", v, "_quilt_plot_data.csv"), row.names = FALSE)
+  write.csv(X, file = paste0("results/tree_cores/quiltplots/plot_data/all/", v, "_quilt_plot_data.csv"), row.names = FALSE)
 }
 #}
 
[…]

### 2022-05-22 doc/submissions/2022_05_Nature/Response to reviews.Rmd
Add RZ to acknowledgements

--- 2022-05-09 doc/submissions/2022_04_Nature/Response to reviews.Rmd
@@
 
-**We simplified the presentation of our data cleaning methods, ensuring that these numbers are clearly presented for both stagges of cleaning:**
+**We simplified the presentation of our data cleaning methods and ensured that these numbers are clearly presented for both stages of cleaning:**
 
@@
 
-**Thank you.**
+**Thank you, Roman. We added your name to the acknowledgements.**

### 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
Add RZ to acknowledgements
cut abstract 201-->200 words

@@
 
-**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;<!--@crabbe_extreme_2016;-->@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
+**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;<!--@crabbe_extreme_2016;-->@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
 **However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
@@
 We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie.
-Thanks to David A. Orwig, members of the ForestGEO Ecosystems & Climate Lab at SCBI, and four anonymous reviewers for helpful feedback.
+Thanks to David A. Orwig, Roman Zweifel, members of the ForestGEO Ecosystems & Climate Lab at SCBI, and three anonymous reviewers for helpful feedback.
 The research was funded by ForestGEO (Smithsonian). 

### 2022-05-22 doc/manuscript/Dow_Extended_Data.Rmd
italicize variables & avoid latex
-Italic font used for variables, genes and non-English languages; bold font for vectors
- Equations and special characters in Word in ‘normal text’ or Symbol font (Word Equation Editor avoided where possible)

@@
 ```
-$^\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (CTW, Fig.
+$^\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (_CTW_, Fig.
 1).
 Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straightforward.
-We used the CTW of the porosity group containing the dominant canopy species at each site: *Liriodendron tulipifera* (diffuse porous) at SCBI and *Quercus rubra* (ring porous) at Harvard Forest.
+We used the _CTW_ of the porosity group containing the dominant canopy species at each site: *Liriodendron tulipifera* (diffuse porous) at SCBI and *Quercus rubra* (ring porous) at Harvard Forest.
 
@@
 
-![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
-Shown are relationships between mean $T_{max}$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
+![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
+Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
 Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
 The 95% credible intervals are represented by bands centered on the posterior mean for each year.
-For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective CTW.
+For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective _CTW_.
 Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }
@@
 
-![**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the CTW.
+![**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
 Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
@@
 ![**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean minimum temperatures ($T_{min}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
-Colors indicate the bootstrapped correlation between monthly $T_{min}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Colors indicate the bootstrapped correlation between monthly $T_{min}$ and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April $T_{min}$.
@@
 ![**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$) of the current and past year, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
-Colors indicate the bootstrapped correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Colors indicate the bootstrapped correlation between monthly $T_{max}$ and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$.

### 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
italicize variables & avoid latex
-Italic font used for variables, genes and non-English languages; bold font for vectors
- Equations and special characters in Word in ‘normal text’ or Symbol font (Word Equation Editor avoided where possible)

@@
 In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[<!--@jeong_phenology_2011;--><!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@keenan_net_2014;@finzi_carbon_2020;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
-Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
+Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (_GPP_) and net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
 However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
-Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
+Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (*ANPP~woody~*) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
 As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
-Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
+Understanding the sensitivity of *ANPP~woody~* to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
 
 Most studies on tree stem growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022].
-In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
+In contrast, we have little evidence as to how stem growth and *ANPP~woody~* respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
 Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
-However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016].
+However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], _GPP_[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016].
 Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021;@cabon_crossbiome_2022].
@@
 
-Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig.
+Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height (*∆DBH*; Fig.
 `r schematic_summary`).
@@
 
-<!--![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }-->
+<!--![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ °C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }-->
 
@@
 Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
-We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW).
-CTW was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
-The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.
+We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (_CTW_).
+_CTW_ was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
+The _CTW_ was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (_CTW_ *T~max~*) used as our independent variable.
 
@@
 Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
-Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }-->
+Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (_CTW_), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+Leaf phenology years are colored according to the _CTW_ *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }-->
 
-For ring- and diffuse- porous species at both sites, warmer CTW *T~max~* resulted in earlier achievement of seasonal milestones.
+For ring- and diffuse- porous species at both sites, warmer _CTW_ *T~max~* resulted in earlier achievement of seasonal milestones.
 Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig.
-`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
-Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs.
+`r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/°C at SCBI (p=0.016) and 2.4 days/ °C at Harvard Forest (p=0.1).
+Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with _CTW_ *T~max~* (Figs.
 `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
-Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).
+Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ °C for ring porous species and 2.8-2.9 days/ °C for diffuse porous species at SCBI, and 10.3-12.3 days/ °C for ring porous species and 0.9-4.2 days/°C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).
 
-Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig.
+Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the _CTW_ *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig.
 `r DOY_timing`a-b), there was no consistent lengthening of *L~pgs~* (Fig.
@@
 
-In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were inconsistent, and usually weak (Fig.
+In contrast to the pronounced effects of _CTW_ *T~max~* on the timing of growth, its effects on *g~max~* and *∆DBH* were inconsistent, and usually weak (Fig.
 `r schematic_summary`, `r EDF_pheno_all`).
-Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
-$\Delta DBH$ displayed a weak positive relationship (+0.003-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
+Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to _CTW_ *T~max~* for ring- and diffuse- porous species.
+*∆DBH* displayed a weak positive relationship (+0.003-0.008 cm/ °C) with _CTW_ *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ °C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).
 
@@
 `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`).
-In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
-Across all chronologies, the standardized ring-width index (RWI) was significantly (95% confidence interval did not include 0) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
+In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35°N latitude) to Michigan (45.56°N latitude) and spanning a 15 °C range in April *T~max~*.
+Across all chronologies, the standardized ring-width index (_RWI_) was significantly (95% confidence interval did not include 0) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`).
 There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
-In contrast, RWI was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring- and diffuse-porous species, respectively.
-*T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and RWI (`r EDF_tree_rings_TMIN`).
+In contrast, _RWI_ was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring- and diffuse-porous species, respectively.
+*T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and _RWI_ (`r EDF_tree_rings_TMIN`).
 
-To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and *T~max~* in the previous year (`r EDF_tree_rings_TMAX_extended`).
-This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of RWI to previous March or April *T~max~* for 5/142 ring porous chronologies and to previous April or May *T~max~* for 7/66 diffuse porous chronologies.
+To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between _RWI_ and *T~max~* in the previous year (`r EDF_tree_rings_TMAX_extended`).
+This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of _RWI_ to previous March or April *T~max~* for 5/142 ring porous chronologies and to previous April or May *T~max~* for 7/66 diffuse porous chronologies.
 
-To test whether there may be an enhancement of growth by warmer spring temperatures that was offset by the negative effect of high summer temperatures, we tested for the joint effects of April and June-July *T~max~* on RWI.
+To test whether there may be an enhancement of growth by warmer spring temperatures that was offset by the negative effect of high summer temperatures, we tested for the joint effects of April and June-July *T~max~* on _RWI_.
 Results were qualitatively similar to the univariate correlations (Fig.
-`r tree_rings`), with significant (p $\le$ 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
+`r tree_rings`), with significant (p ≤ 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).
 
 <!--![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
-Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.
@@
 The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig.
-`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and NEE[@keenan_net_2014;@finzi_carbon_2020].
-However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
+`r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and _NEE_[@keenan_net_2014;@finzi_carbon_2020].
+However, inconsistent with the concept that an earlier start to growth would increase *ANPP~woody~*, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig.
 `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig.
@@
 Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;<!--@crabbe_extreme_2016;-->@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
-It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
+It remains theoretically possible that warm springs could augment *ANPP~woody~*, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
 Extra C fixed in years with warm springs could potentially be allocated to formation of more C-dense wood, either through enhanced cell wall thickening (a process that lags behind stem expansion[@cuny_woody_2015]) or to a higher ratio of high-density latewood to lower-density earlywood.
 However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
-Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or ANPP~woody~.
+Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or *ANPP~woody~*.
 
 The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
-Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, ANPP~woody~ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
+Indeed, when _GPP_ of a mature forest was increased through experimental CO~2~ enrichment, *ANPP~woody~* remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
 Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
-Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.
+Thus, observed augmentation of _NEE_ by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.
 
 <!--.short vs long-term effects-->
-It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of ANPP~woody~.
-Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], ANPP~woody~[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
+It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of *ANPP~woody~*.
+Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], *ANPP~woody~*[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
 Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
-If extra photosynthate made available through a growing difference between GPP and ANPP~woody~ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase ANPP~woody~ through indirect mechanisms.
+If extra photosynthate made available through a growing difference between _GPP_ and *ANPP~woody~* is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase *ANPP~woody~* through indirect mechanisms.
 Understanding how warming spring temperatures are influencing C allocation within ecosystems remains a key outstanding question.
@@
 Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April *T~max~*; source: CRU v.4.04[@harris_version_2020]) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
-Specifically, average spring *T~max~* was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.
+Specifically, average spring *T~max~* was 16.9 ±1.4°C (mean ± 1SD) prior to the study period (range: 14.1-19.9) and 17.6± 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06± 1.02 (range: -1.7-2.4) prior to the study period and 0.5± 0.8 (range: -0.8-1.6) during.
 
@@
 Within the study period for the dendrometer band analysis (1999-2003), spring temperatures (March and April *T~max~*; source: CRU v.4.04[@harris_version_2020]) and summer SPEI values (4-month value of August) were similar to the average climate prior to the study period (1970-1998).
-Specifically, average spring *T~max~* was 10.9$\pm$ 1.5°C prior to the study period (range: 8.0-13.2) and 11.2$\pm$ 1.0°C (range: 10.1-12.2) during the study period, while summer SPEI was 0.1$\pm$ 0.9 (range: -1.8-1.7) prior to the study period and 0.2$\pm$ 0.9 (range: -1.0-1.1) during.
+Specifically, average spring *T~max~* was 10.9± 1.5°C prior to the study period (range: 8.0-13.2) and 11.2± 1.0°C (range: 10.1-12.2) during the study period, while summer SPEI was 0.1± 0.9 (range: -1.8-1.7) prior to the study period and 0.2± 0.9 (range: -1.0-1.1) during.
 The driest summer during the study period (1999) had the 5th lowest SPEI value (-1.0) in the period 1970-2003, with precipitation of 52 mm mo^-1^ in June-August compared to average monthly precipitation of 100+ mm[@dorangeville_peak_2022].
@@
 The raw dendrometer band data were screened to remove records or entire tree-years that were inappropriate for our analysis because of reduced reliability of predicted growth in the modeled curves.
-Specifically, we removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253) and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
+Specifically, we removed tree-years with small or negligible total growth (*∆DBH* ≤0.005 cm; SCBI = 26, Harvard Forest = 253) and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
 In total, this process removed `r 26+253+22+8` of the `r n_tree_years + (3+11+74+101) + (26+253+22+8)`<!--first set is n removed for poor model fit, 2nd set is n removed here--> available tree-year records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest.
@@
 $$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
-Here, $L$ and $K$ are lower and upper asymptotes of the model. 
+Here, _L_ and _K_ are lower and upper asymptotes of the model. 
 *DOY~ip~* is the day of year where the inflection point in growth rate occurs,
-$r$ shapes the slope of the curve at the inflection point,
+_r_ shapes the slope of the curve at the inflection point,
 and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
@@
 When $\theta = 1$, *g~max~* occurs on *DOY~ip~*.
-The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year and are constrained by the first and last dendrometer band measurements. 
+The model outputs two additional variables, _a_ and _b_, which represent the beginning and end DBH in each model year and are constrained by the first and last dendrometer band measurements. 
 The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. 
@@
 Modeled fits for tree-years were removed under two conditions:
-(1) *g~max~* was $\ge 2.5$ standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);
-(2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
+(1) *g~max~* was ≥2.5 standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);
+(2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were ≥2.5 standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
 In total, this process removed `r 3+11+74+101` of the `r n_tree_years + (3+11+74+101)` tree-year records deemed appropriate for analysis, leaving a total of `r n_tree_years` tree-years included in the final analysis (`r  EDT_species`).
@@
 <!-- Determining critical temperature windows -->
-The critical temperature window (CTW, Fig.
+The critical temperature window (_CTW_, Fig.
 `r schematic_summary`), defined as the period over which *T~max~* was most strongly correlated with *DOY~25~*, was determined using the R package *climwin*[@vandepol_identifying_2016].
 This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame. 
-It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
+It does so by reporting the correlation and *∆AICc*, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
 Here, we tested for correlation between temperature predictor variables (*T~max~*, *T~min~*) and biological outcome variable *DOY~25~* over the time-frame from January 1 to the mean *DOY~25~* for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`).
-The time period yielding the lowest $\Delta AICc$ was selected as the CTW.
+The time period yielding the lowest *∆AICc* was selected as the _CTW_.
 To avoid spurious correlations that could occur using temperature data at the daily resolution, we ran this analysis with weekly resolution, using temperatures averaged over weekly time periods.
 Because *T~max~* proved to have a generally stronger influence over *DOY~25~* and other growth parameters, we focused on this variable in our ultimate model, as opposed to *T~min~*.
-We defined CTW for *DOY~25~*, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable.
+We defined _CTW_ for *DOY~25~*, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable.
 
-To ensure that patterns were robust under an alternative definition of CTW, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
-`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the CTW identified by *climwin* to be the month with the most days in the CTW (`r  EDT_phenology_parameters`) each critical window.
+To ensure that patterns were robust under an alternative definition of _CTW_, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig.
+`r tree_rings`, `r EDF_tree_ring_figs`), we also ran analyses where we fixed the _CTW_ identified by *climwin* to be the month with the most days in the _CTW_ (`r  EDT_phenology_parameters`) each critical window.
 The months identified were March and April for ring- and diffuse-porous species at SCBI, respectively, and April and May for ring- and diffuse-porous species at Harvard Forest, respectively.
@@
 <!--Hypothesis testing -->
-Correlation between the dendrometer band-derived growth parameters (*DOY~25~*, *DOY~50~*, *DOY~75~*, *L~pgs~*, *g~max~*, and $\Delta DBH$, Fig.
-`r schematic_summary`) and CTW *T~max~* (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
+Correlation between the dendrometer band-derived growth parameters (*DOY~25~*, *DOY~50~*, *DOY~75~*, *L~pgs~*, *g~max~*, and *∆DBH*, Fig.
+`r schematic_summary`) and _CTW_ *T~max~* (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
 Analyses were run for both *T~max~* and *T~min~*, with qualitatively similar results, but we present only results for *T~max~*, which had overall stronger correlation with growth parameters. 
-Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean *T~max~* (or *T~min~*) during the CTW, along with random effects of species and of individual tree.
+Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean *T~max~* (or *T~min~*) during the _CTW_, along with random effects of species and of individual tree.
 We ran separate models for each site, and for the response of all growth parameters to *T~max~* (or *T~min~*). 
@@
 
-For each species-site combination, we converted tree-ring records into the dimensionless RWI to emphasize interannual variability associated with climate[@cook_methods_1990].
+For each species-site combination, we converted tree-ring records into the dimensionless _RWI_ to emphasize interannual variability associated with climate[@cook_methods_1990].
 A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
@@
 When replication was greater than n = 3 trees, we used the average correlation between raw ring-width series (rbar) method.
-The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilization[@cook_time_1985].
-We defined chronology start year (`r EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.
+The robust biweight mean chronology (_RWI_) for each species-site combination was calculated from the ring-width indices following variance stabilization[@cook_time_1985].
+We defined chronology start year (`r EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where ≥80% of the population signal was captured in the chronology.
 
 <!--QUILT PLOT METHODS-->
-For the analysis of correlation between RWI and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
-Correlations between monthly climate and RWI were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013].
+For the analysis of correlation between _RWI_ and climate variables, we obtained monthly *T~max~* and *T~min~* data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
+Correlations between monthly climate and _RWI_ were assessed in R v 4.0[@rcoreteam_language_2020] using the packages *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013].
 Reported correlations and significance were determined using bootstrapped confidence intervals.
@@
 `r tree_rings`, Extended Data Figures 6-7).
-Our analysis focused on assessing correlations of RWI to months spanning January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
+Our analysis focused on assessing correlations of _RWI_ to months spanning January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
 To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
@@
 <!--MULTIVARIATE ANALYSIS METHODS-->
-We used a multivariate model to test for joint effects of April and summer *T~max~* on RWI.
-We focused on April to represent spring temperatures because it was the month with greatest overall alignment with the CTW's identified in the dendrometer band analysis and had the highest rate of positive correlations with RWI (`r EDT_tree_ring_summary`).
+We used a multivariate model to test for joint effects of April and summer *T~max~* on _RWI_.
+We focused on April to represent spring temperatures because it was the month with greatest overall alignment with the _CTW_'s identified in the dendrometer band analysis and had the highest rate of positive correlations with _RWI_ (`r EDT_tree_ring_summary`).
 We began by testing univariate correlations of *T~max~* over three summer windows: June, June-July, and May-August. 
-Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April *T~max~* and June-July *T~max~* on RWI for each chronology independently using the base lm() function in R.
+Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April *T~max~* and June-July *T~max~* on _RWI_ for each chronology independently using the base lm() function in R.
 Slopes and p-values for each chronology are given in `r chron_table`. 
@@
 
-**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.
+**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ °C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.
 
@@
 Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`.
-Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
-Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).
+Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (_CTW_), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.
+Leaf phenology years are colored according to the _CTW_ *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).
 
 **Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**.
-Colors indicate the bootstrapped correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Colors indicate the bootstrapped correlation between monthly *T~max~* and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
 Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.

### 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
add legends for extended data figures to main file
@camerondow35 , any updates in Extended Data file need to be made here too

#142

@@
 \newpage
+# Legends for Extended Data Figures {-}
+
+**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
+Sampling details are provided in `r chron_table`.
+Base map source is ggplot2.
+
+**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
+For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.
+Dashed lines represent modeled DBH change which fell outside of the median DOY where predicted starting DBH (a) and predicted ending DBH (b) were reached.
+Solid lines represent DBH change attributable to stem growth.
+
+**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
+Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`).
+Yellow shading indicates neutral relationships, while orange or red shading indicates that $DOY_{25}$ advances with increased $T_{max}$ over the given time window (negative $\beta$).
+Black circles indicate the critical temperature window selected based on minimization of $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model.
+Critical temperature windows are listed in `r EDT_phenology_parameters`.
+
+**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
+Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+The 95% credible intervals are represented by bands centered on the posterior mean for each year.
+For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective _CTW_.
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.
+
+**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
+Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+The 95% credible intervals are represented by bands centered on the posterior mean for each year.
+For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI.
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.
+
+**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean minimum temperatures ($T_{min}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the bootstrapped correlation between monthly $T_{min}$ and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April $T_{min}$.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
+Species analyzed and numbers of significant correlations to $T_{min}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.
+
+**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$) of the current and past year, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the bootstrapped correlation between monthly $T_{max}$ and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
+Figure presents the same results as main manuscript Fig.
+3 but extends back to include the previous year.
+Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.
+
+\newpage

### 2022-05-22 doc/manuscript/Dow_Extended_Data.Rmd
reorder extended data figures to match order of presentation in text

@@
 ##figures
-EDF_map = "Extended Data Figure 1" 
-EDF_growth_curves_all = "Extended Data Figure 2"  
-EDF_climwin = "Extended Data Figure 3"
-EDF_pheno_all = "Extended Data Figures 4-5"
-EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
-EDF_pheno_Tmax_month = "Extended Data Figure 5"
+EDF_growth_curves_all = "Extended Data Figure 1"  
+EDF_climwin = "Extended Data Figure 2"
+EDF_pheno_all = "Extended Data Figures 3-4"
+EDF_pheno_Tmax_climwin = "Extended Data Figure 3"
+EDF_pheno_Tmax_month = "Extended Data Figure 4"
+EDF_map = "Extended Data Figure 5" 
 EDF_tree_ring_figs= "Extended Data Figure 6-7"
@@
 
-![**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
-Sampling details are provided in `r chron_table`.
-Base map source is ggplot2.](tables_figures/chronology_map.png)
-
-\newpage
-
-![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
+![**Extended Data Figure 1 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
 For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.
@@
 
-![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
+![**Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
 Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`).
@@
 
-![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
+![**Extended Data Figure 3 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
 Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
@@
 
-![**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
+![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
 Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
@@
 Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }
+
+\newpage
+
+![**Extended Data Figure 5 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
+Sampling details are provided in `r chron_table`.
+Base map source is ggplot2.](tables_figures/chronology_map.png)
 

### 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
reorder extended data figures to match order of presentation in text

@@
 `r schematic_summary`).
-This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022].
+This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`, `r EDF_growth_curves_all`)[@dorangeville_peak_2022;@tumajer_limitation_2022].
 These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).
@@
 
-Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
+Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`). 
 We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (_CTW_).
@@
 
-**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
-Sampling details are provided in `r chron_table`.
-Base map source is ggplot2.
-
-**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
+**Extended Data Figure 1 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
 For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.
@@
 
-**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
+**Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$.
 Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`).
@@
 
-**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
+**Extended Data Figure 3 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
 Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
@@
 
-**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
+**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
 Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
@@
 Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.
+
+**Extended Data Figure 5 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
+Sampling details are provided in `r chron_table`.
+Base map source is ggplot2.
 

### 2022-05-22 doc/manuscript/Dow_Extended_Data.Rmd
fixed a few typos, but this is no longer the master copy

@@
 
-![**Extended Data Figure 1 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
+![**Extended Data Figure 1 | Seasonal patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
 For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.
@@
 
-![**Extended Data Figure 3 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
+![**Extended Data Figure 3 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`).
 Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).
@@
 
-![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
+![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
 Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row).

### 2022-05-22 doc/manuscript/Dow_Extended_Data_noLegends.Rmd
Extended Data Legends go in main manuscript file
Removed these from ED file, added in main file. Also, incorporated footnote of EDT2 into legend. And fixed up some typos, etc.

---
title: ""
author:
output: 
  pdf_document:
    fig_caption: yes
    #toc: true # table of content true
    #toc_depth: 2  
header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}
---

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
path_to_results <- paste0(dirname(dirname(getwd())), "/results/")

# EXTENDED DATA 
##tables
EDT_species="Extended Data Table 1"  
EDT_phenology_parameters="Extended Data Table 2" 
EDT_tree_ring_summary="Extended Data Table 3"

##figures
EDF_growth_curves_all = "Extended Data Figure 1"  
EDF_climwin = "Extended Data Figure 2"
EDF_pheno_all = "Extended Data Figures 3-4"
EDF_pheno_Tmax_climwin = "Extended Data Figure 3"
EDF_pheno_Tmax_month = "Extended Data Figure 4"
EDF_map = "Extended Data Figure 5" 
EDF_tree_ring_figs= "Extended Data Figure 6-7"
EDF_tree_rings_TMIN = "Extended Data Figure 6"
EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"

###~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## some stats/ 
chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
n_chronologies <- nrow(chronology_table)  
n_cores_sites <- length(unique(chronology_table[, "Location"]))
n_cores_species <- length(unique(chronology_table[, "Species"]))

```

\newpage

**Extended Data Table 1**

```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

species_table <- read.csv(file = "tables_figures/species_table.csv", col.names = c("site", " ","species", "species code", "n trees", "n tree-years", "n cores", "date range"),  stringsAsFactors = FALSE, check.names = FALSE) 
species_table <- species_table[c(2:3, 5:8)] #subset(species_table, select = -c(Species code) )#remove columns that shouldn't be included

kable(species_table,  format = "latex", booktabs = TRUE,  escape = F) %>%
add_header_above(c( " ", " ", "dendrometer bands" = 2, "tree cores"=2 )) %>%
  pack_rows("SCBI", 1, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("", 11, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Harvard Forest", 13, 15, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("", 16, 23, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

\newpage

**Extended Data Table 2**

```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
mean_growth_parameters <- read.csv("tables_figures/phenology_parameters.csv",  col.names = c("", "ring-porous","diffuse-porous","ring-porous", "diffuse-porous"), stringsAsFactors = FALSE, check.names = FALSE)
kable(mean_growth_parameters, format = "latex", booktabs = TRUE,  escape = F) %>%
add_header_above(c(" ",  "SCBI" = 2, "Harvard Forest"=2 )) %>%
    column_spec(1, width = "4cm") %>%
    pack_rows("Stem Growth", 2, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
    pack_rows("Canopy Foliage (ecosystem level)", 8, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
    pack_rows("Temperature sensitivity (days/°C)", 12, 20, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
    add_indent(13:16)%>%
    add_indent(18:21)%>%
kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

\newpage

**Extended Data Table 3**

```{r tree_ring_summary,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
TreeRingResultsSummary <- read.csv("tables_figures/TreeRingResultsSummary.csv",  
                                   stringsAsFactors = FALSE, check.names = FALSE)
TreeRingResultsSummary <- TreeRingResultsSummary[c(2:19)]
kable(TreeRingResultsSummary, 
      col.names = c("species","n","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-","+", "-"), 
      format = "latex", booktabs = TRUE,  escape = F) %>%
add_header_above(c(" " = 2,  "March" =2 , "April" = 2, "May" = 2,"June" = 2,"July" = 2,"Aug" = 2,"April" = 2,"June-July" = 2)) %>%
add_header_above(c(" " = 2,  "univariate analysis" = 12,"multivariate analysis" = 4)) %>%
add_header_above(c(" " = 2,  "n. significant bootstrapped correlations of RWI to monthly mean max. T" = 16)) %>%
    column_spec(3:18, width = ".4cm") %>%
pack_rows("Ring-porous", 1, 17, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
pack_rows("Diffuse-porous", 18, 27, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
row_spec(28, bold = T) %>%
kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

\newpage

**Extended Data Figure 1**

![](tables_figures/growth_curves_all.png){ width=100% }

\newpage

**Extended Data Figure 2**

![](tables_figures/climwin_figure.png){ width=70% }

\newpage

**Extended Data Figure 3**

![](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

\newpage

**Extended Data Figure 4**

![](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }

\newpage

**Extended Data Figure 5**

![](tables_figures/chronology_map.png)

\newpage

**Extended Data Figure 6**

![](tables_figures/tree_rings_tmin.png){ width=100% }

\newpage

**Extended Data Figure 7**

![](tables_figures/tree_rings_tmax_extended.png){ width=100% }

### 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
Extended Data Legends go in main manuscript file
Removed these from ED file, added in main file. Also, incorporated footnote of EDT2 into legend. And fixed up some typos, etc.

--- 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 \newpage
+# Legends for Extended Data Table & Figures {-}
+
+**Extended Data Table 1 | Dominant broadleaf deciduous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in our final analysis.**
+
+**Extended Data Table 2 | Summary of parameters describing the seasonality and temperature sensitivity of broadleaf deciduous tree woody growth and canopy phenology at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature (*T~max~*) during the critical temperature window (_CTW_, Fig.
+1).
+Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straightforward.
+We used the _CTW_ of the porosity group containing the dominant canopy species at each site: *Liriodendron tulipifera* (diffuse porous) at SCBI and *Quercus rubra* (ring porous) at Harvard Forest.
+
+**Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at significance level = 0.05) positive or negative correlations of ring width index to monthly *T~max~* in univariate and multivariate analyses.**
+
+**Extended Data Figure 1 | Seasonal patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018.
+For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data.
+Dashed lines represent modeled DBH change which fell outside of the median DOY where predicted starting DBH (a) and predicted ending DBH (b) were reached.
+Solid lines represent DBH change attributable to stem growth.
+
+**Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved (*DOY~25~*) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature (*T~max~*) and *DOY~25~*.
+Window Open and Window Close indicate number of weeks prior to *DOY~25~* (listed in `r EDT_phenology_parameters`).
+Yellow shading indicates neutral relationships, while orange or red shading indicates that *DOY~25~* advances with increased *T~max~* over the given time window (negative $\beta$).
+Black circles indicate the critical temperature window selected based on minimization of *∆AICc*, the difference in Akaike Information Criterion corrected for small sample size relative to a null model.
+Critical temperature windows are listed in `r EDT_phenology_parameters`.
+
+**Extended Data Figure 3 | Response of stem growth timing and rates to mean maximum temperatures (*T~max~*) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which *T~max~* was most strongly correlated with the day of year on which 25% of annual growth was achieved (*DOY~25~*; `r EDT_phenology_parameters`, `r EDF_climwin`).
+Shown are relationships between mean *T~max~* over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*, respectively; first row); the length of the peak growing season (*L~pgs~*; second row); maximum growth rate (*g~max~*; third row); and total seasonal radial stem growth (*∆DBH*; fourth row).
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+The 95% credible intervals are represented by bands centered on the posterior mean for each year.
+For both species groups at both sites, *DOY~25~*, *DOY~50~*, and *DOY~75~* all declined significantly with mean *T~max~* during their respective _CTW_.
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.
+
+**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures (*T~max~*) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which *T~max~* was most strongly correlated with the day of year on which 25% of annual growth was achieved (*DOY~25~*; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_.
+Shown are relationships between monthly *T~max~* and days of the year on which 25%, 50%, and 75% total stem growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*, respectively; first row); the length of the peak growing season (*L~pgs~*; second row); maximum growth rate (*g~max~*; third row); and total seasonal radial stem growth (*∆DBH*; fourth row).
+Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines.
+The 95% credible intervals are represented by bands centered on the posterior mean for each year.
+For both species groups at both sites, *DOY~25~*, *DOY~50~*, and *DOY~75~* declined significantly with April *T~max~*, with the exception of *DOY~25~* for ring-porous species at SCBI.
+Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.
+
+**Extended Data Figure 5 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both.
+Sampling details are provided in `r chron_table`.
+Base map source is ggplot2.
+
+**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean minimum temperatures (*T~min~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the bootstrapped correlation between monthly *T~min~* and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April *T~min~*.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
+Species analyzed and numbers of significant correlations to *T~min~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.
+
+**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*) of the current and past year, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `).
+Colors indicate the bootstrapped correlation between monthly *T~max~* and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate.
+Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*.
+Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`).
+Figure presents the same results as main manuscript Fig.
+3 but extends back to include the previous year.
+Species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.
+
+\newpage

### 2022-06-01 doc/manuscript/Dow_manuscript_Nature.Rmd
@teixeirak adding version no. & code location

@@
 We ran separate models for each site, and for the response of all growth parameters to *T~max~* (or *T~min~*). 
-This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@standevelopmentteam_stan_2019].
+This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` version 2.21.3 R interface to the Stan programming language (source code available at https://github.com/EcoClimLab/growth_phenology#steps-to-replicate-the-analysis)[@gabry_rstanarm_2020;@standevelopmentteam_stan_2019].
 In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 

### 2022-06-04 doc/manuscript/Dow_manuscript_Nature.Rmd
update code version numbers and URLs -- still need ARSTAN v number
#148

--- 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
@@
 We ran separate models for each site, and for the response of all growth parameters to *T~max~* (or *T~min~*). 
-This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@standevelopmentteam_stan_2019].
+This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` version 2.21.3 R interface to the Stan programming language [@gabry_rstanarm_2020;@standevelopmentteam_stan_2019].
 In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
@@
 For each species-site combination, we converted tree-ring records into the dimensionless _RWI_ to emphasize interannual variability associated with climate[@cook_methods_1990].
-A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
+A 2/3rds _n_ spline was applied to each core using ARSTAN **v.###** to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
 An adaptive power transformation, a process that also stabilizes the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
@@
 # Code Availability Statement {-}
+<!--this statement to be replicated in "nr-reporting-summary.pdf"-->
 Data were analysed in the open source statistical software R (version 4.0).
@@
 We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
+Mixed effects models were run with mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` version 2.21.3 R interface to the Stan programming language (source code available at https://github.com/EcoClimLab/growth_phenology#steps-to-replicate-the-analysis).
+Tree-ring chronologies were developed using the program ARSTAN **v.###** (https://www.geog.cam.ac.uk/research/projects/dendrosoftware/).
 All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).

### 2022-06-06 doc/manuscript/Dow_manuscript_Nature.Rmd
add ARSTAN version
#148

@@
 For each species-site combination, we converted tree-ring records into the dimensionless _RWI_ to emphasize interannual variability associated with climate[@cook_methods_1990].
-A 2/3rds _n_ spline was applied to each core using ARSTAN **v.###** to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
+A 2/3rds _n_ spline was applied to each core using ARSTAN V49_1b to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
 An adaptive power transformation, a process that also stabilizes the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
@@
 Mixed effects models were run with mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` version 2.21.3 R interface to the Stan programming language (source code available at https://github.com/EcoClimLab/growth_phenology#steps-to-replicate-the-analysis).
-Tree-ring chronologies were developed using the program ARSTAN **v.###** (https://www.geog.cam.ac.uk/research/projects/dendrosoftware/).
+Tree-ring chronologies were developed using the program ARSTAN V49_1b (https://www.geog.cam.ac.uk/research/projects/dendrosoftware/).
 All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).

### 2022-06-08 doc/submissions/2022_06_Nature/Cover letter.Rmd
update folder name

---
title: "Cover letter"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Anna Armstrong:

We are pleased to provide a revised version of Nature manuscript 2021-11-18307A, now entitled "Warm springs alter timing but not total growth of temperate deciduous trees".

We have made the following minor revisions to the manuscript and associated materials:

- changed the title to one meeting the length constraints
- responded to all reviewer comments, as detailed in our Response to Reviews (no major changes required)
- responded to all editorial comments in the email and editorial request form (no major changes required)
- archived all raw tree-ring data to the GitHub repository
- created a new Zenodo release, and provided Zenodo DOI in the required locations
- provided a supplementary information guide
- completed all related manuscript forms
- provided potential cover images

We hope that you will find all materials acceptable for publication, and will be happy to address any required revisions.

Best regards,

Kristina Anderson-Teixeira (on behalf of full author team)

### 2022-06-10 doc/manuscript/Dow_manuscript_Nature.Rmd
update DOI in data & code availability statements
#150

@@
 # Data Availability Statement {-} 
-The datasets generated and analysed during the current study are available via GitHub in the *growth_phenology* repository of the ForestGEO Ecosystems & Climate Lab @ SCBI, (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI *[TBD]*).
+The datasets generated and analysed during the current study are available via GitHub in the *growth_phenology* repository of the ForestGEO Ecosystems & Climate Lab @ SCBI, (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: 10.5281/zenodo.6632090).
 Master versions of the dendrometer band data are available for SCBI via GitHub in the *Dendrobands* repository of the Smithsonian Conservation Biology Institute ForestGEO plot (https://github.com/SCBI-ForestGEO/Dendrobands), which is archived in Zenodo (DOI 10.5281/zenodo.5551143), and for Harvard Forest via the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF149).
@@
 Tree-ring chronologies were developed using the program ARSTAN V49_1b (https://www.geog.cam.ac.uk/research/projects/dendrosoftware/).
-All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).
+All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: 10.5281/zenodo.6632090).
 

### 2022-06-10 doc/manuscript/Dow_manuscript_Nature.Rmd
update funding statement

@@
 Thanks to David A. Orwig, Roman Zweifel, members of the ForestGEO Ecosystems & Climate Lab at SCBI, and three anonymous reviewers for helpful feedback.
-The research was funded by ForestGEO (Smithsonian). 
-Collection of tree-ring samples was funded by a USDA Agriculture and Food Research Initiative grant 2017-67013-26191 and from the Indiana University Vice Provost for Research Faculty Research Program.
+Analyses and SCBI data collection were funded by the Smithsonian Institution (ForestGEO - Smithsonian Tropical Research Institute, Smithsonian's National Zoo & Conservation Biology Institute, and Scholarly Studies, Competitive Grants Program for Science, and Grand Challenges Award Program grants to KJAT).
+Collection of tree-ring samples was funded by USDA Agriculture and Food Research Initiative grant 2017-67013-26191 (to JTM), National Science Foundation grants #1805276 (to GLH) and #1805617 (to JTM), and from the Indiana University Vice Provost for Research Faculty Research Program (to JTM).
+LD received funding from the Natural Sciences and Engineering Research Council of Canada (DG RGPIN-2019-04353) and New Brunswick Innovation Foundation (RIF 2019-029).
+SMM was partially funded by NSF RAPID-2030862.
+IRM was supported by a NASA FINESST grant 80NSSC20K1661P00001.
 

### 2022-08-02 doc/manuscript/Dow_manuscript_Nature.Rmd
fix typo in equation
removed extra closing parentheses

@@
 In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
-$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
+$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta] ^ \theta}$$ 
 Here, _L_ and _K_ are lower and upper asymptotes of the model. 
